### 2016-02-05 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
Initial commit

---
title: "A Meta Analysis of the Bacterial Microbiome and Human Obesity"
output:
  html_document:
    keep_md: yes
  word_document:
    reference_docx: manuscript_template.docx
bibliography: metaAnalysis.obesity.microbiome.bib
---

Marc A Sze<sup>1</sup> and Patrick D Schloss<sup>1</sup>

<sup>1</sup> Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI, USA  
<br>
<br>
<br>
<br>
<br>
<br>
Corresponding author: Patrick D Schloss  
*Current Address:*  
1526 MSRB I  
Ann Arbor, MI  
USA 48109-5620  
Email: pschloss@umich.edu  
Phone: 734-936-2951    
<br>
<br>
**Contributions:** MAS (planned, designed, and performed experiments, completed data analysis, wrote first draft), PDS (planned and designed the study, data interpretation, and edited subsequent drafts).  
<br>
**Running Title:** Bacterial Microbiome and Obesity
<br>

*******
##### **Abstract:**
**Rationale:** It has been well accepted within the literature that changes to the bacterial microbiome are related to obesity.  However, recent reports from two different research groups re-examining existing data suggests that there may not be a strong association between the bacterial microbiome and obesity.  These two newer studies, however, did not look at the results of pooling all the data together.  In this meta analysis, we drastically increase the number of data sets examined and investigate the pooled data to provide clearity on the association between the bacterial microbiome and obesity.  
**Methods:** A total of 11 data sets with BMI data were analyzed and processed through the standard mothur pipeline for either 454 or MiSeq data.  The data included were from Baxter et al. 2015, Ross et al. 2015, Goodrich et al. 2014, Escobar et al. 2014, Zupancic et al. 2012, Yatsunenko et al. 2012, Human Microbiome Project, Nam et al. 2011, Wu et al. 2011, MetaHit, and Turnbaugh et al. 2009.  Three analysis piplines were pursued for the data.  First, each data set was analyzed separately for relationships with BMI, similar to previous reports on the re-analysis of the bacterial microbiome and obesity.  Second, a classical apporach for pooling data in a meta analysis was used to invesitigate the over all relative risk (RR) ratio from all studies of being obese based on the bacterial microbiome.  Finally, data normalization was used to investigate all data sets together for relationships between the microbiome and obesity.  
**Results:** For the first part of the analysis there was no significant difference (P > 0.05) between non-obese and obese individuals in Bacteroidetes, Firmicutes, B/F ratio, Shannon diversity, OTU richness, relative risk, or the NMDS of the Bray-Curtis distance matrix analyzed by PERMANOVA.  For the second part of the analysis there was no pooled difference in RR of obesity based on Shannon diversity (RR = 1.05, CI = 0.90, 1.23).  There was also no difference in RR based on the B/F ratio for the poole data (RR = 0.85, CI = 0.73, 1.00).  
**Conclusion:** Our meta-analysis supports the null hypothesis that there is no difference between non-obese and obese individuals, based on the currently avialable bacterial microbiome measures.

*******
##### **Introduction:**  
Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014].  This number increases to approximately 35% in adults (aged 20 or older) and these statisitics have seen little change since 2003 [@ogden_prevalence_2014].  Traditionally the body mass index (BMI) has been used as the traditional method of classifying indviduals as non-obese or obese [@lichtash_body_2013].  However, other methods do exist, such as body adiposity index (BAI) but do not always correlate with important molecular measures of traditionally associated with obesity [@lichtash_body_2013].  Recently, there has been a lot of interest in the bacteiral microbiome and it's potential ability to modulate obesity [@brahe_can_2016; @dror_microbiota_2016].  If this modulation potential is true then the bacterial microbiome could have a enormous impact and role to play in the future treatment of obesity and helping to stem the rising obesity epidemic.

There has been a continuous stream of studies that report that there is a link between the bacterial microbiome and obesity.  The first research article to solidify the bacterial microbiomes' role in obesity was the landmark study by Turnbaugh et al. [-@turnbaugh_obesity-associated_2006] which provided direct evidence that by modulating the microbiome one could have an effect on the obesity phenotype of mice.  This was quickly followed up by human studies that found observable differences in the bacterial microbiome between non-obese and obese individuals [@ley_microbial_2006; @turnbaugh_core_2009].  Since the publication of these studies there have been a steady stream of publications on the bacterial microbiome with associated BMI data but without clear explicit mention of any significant correlations between the bacterial community and obesity [@ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; arumugam_enterotypes_2011; @goodrich_human_2014; @yatsunenko_human_2012; @wu_linking_2011; @escobar_gut_2014].  However, two recent publications [@walters_meta-analyses_2014; @finucane_taxonomic_2014] re-analyzed some of the previously publishd data and came to the shocking conclusion that the human bacterial microbiome did not correlate with obesity.  One common critique of these previous studies was that they never looked at the data pooled together and that this could result in a significant result where each individual study alone would not.

The purpose of this study is to perform an extensive meta-analysis and systematic review of the bacterial microbiome and obesity by analyzing more data sets than the previous studies and including both seperate and pooled analysis to assess the bacterial microbiome and obesity. By including these two components this study will build on the previous two re-analysis studies and provide a stronger assesment of the direction of effect or lack there of on the bacterial microbiome and obesity.

Although there have been numerous RT-PCR and qPCR studies directed at examining the bacterial microbiome and obesity [@guo_development_2008; @schwiertz_microbiota_2010].  The main method of choice to obtain the most complete picture of the overall microbial community is through 16S sequencing of a specific hypervariable region on the 16S rRNA gene.  The sequencers used to complete this task have varied depending on research group with the most popular being the Roche 454, Illumina MiSeq, and Illumina HiSeq.  For this report we limit ourselves to analyzing studies were 16S sequencing data is avaliable along with BMI information.  We also analyzed every study with a uniform approach utilizing the mothur software package for sequence processing and clean up and R statistical software for analysis.

*******
##### **Methods:**  
*Literature Review and Study Inclusion*
<br><br>
A literature review was performed in pubmed for bacterial microbiome studies that also contained 16S rRNA gene sequencing data using 454, MiSeq, or HiSeq sequencing technologies.  A total of 11 studies (combined n = 2133) were identified as fullfiling this criteria.  One final study published late in 2015 used the Illumina NextSeq platform for sequencing [@zeevi_personalized_2015].  However, the BMI data for each individual was not included in the published study or the supplement.  An e-mail was sent to inquire but was never returned.  Thus it was decided not to include this data set in the subsequent meta analysis.

*Sequence Analysis Pipeline*
<br><br>
All data was publically available and the sequence data was downloaded from the NCBI Sequence Read Archive, MG-RAST, or the European Nucleotide Archive.  A total of 7/11 [@turnbaugh_core_2009; @ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @wu_linking_2011; @escobar_gut_2014; @conlan_species-level_2012] studies were processed by staying as close as possible to the mothur 454 standard operating procedure, which can be found at http://www.mothur.org/wiki/454_SOP.  A total of 3/11 [@goodrich_human_2014; @yatsunenko_human_2012] of the studies were processed by staying as close as possible to the mothur MiSeq standard operating procedure found at http://www.mothur.org/wiki/MiSeq_SOP.  The final data set [@arumugam_enterotypes_2011] that was used took advantage of the Metaphlan2 package [@truong_metaphlan2_2015] to generate the necessary data for this analysis.  A comparison of the similarity of Metaphlan2 to 16S sequencing using the Human Microbiome Project (HMP) data set can be found in the online supplement [Figure S1 & S2].  For a few of the studies after sequence processing there was a large decrease in the number of samples available for analysis [Table 1].  A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/marcsze/MOM-Analysis.

*Data Analysis*
<br><br>
All analysis was performed using R version 3.2.2 and R studio version 0.99.491.  The overall analysis was split into three general categories.  The first category involved following the same method employed by Finucane et al. and Walters et al. [-@finucane_taxonomic_2014; -@walters_meta-analyses_2014] in which each study was re-analyzed seperately for associations with BMI.  The main associations with BMI that were explored in each seperate data set were Bacteroidetes phyla, Firmicutes phyla, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, NMDS visualization of Bray-Curtis distance matrix PERMANOVA, and relative risk (RR).  A two-way T-test was performed for comparison for non-obese versus obese.  An ANOVA was used to investigate all BMI categories (if available) with a Tukey post hoc test. Third, a Pearson test for linear correlation between the variables was used.  Finally, Random forest analysis to identify the best variables for classification of obese and non-obese was performed to look if there were similarities between data sets that we may have missed.  The R package AUCRF (v1.1) was used for this.
<br><br>
The second category involved pooling all the data from all the studies together to asses whether the combined direction of all the studies increased an indivdiual's RR for being obese.  The two measurements that were investigated in detail for this part were Shannon diversity and the B/F ratio.  To generate the RR the median value of each study was taken and the number of those that were non-obese or obese in each group was recorded and a Fisher exact-test was performed.  The relative risk from the the Fisher exact-test was obtained using the epiR (v0.9-6.9) package in R.  To obtain the pooled RR and confidence intervals (CI) the metafor (v1.9-8) package available for the R statistics software was utilized.  If the pooled analysis CI does not touch or cross 1.00 the pooled RR had a P-value < 0.05. 
<br><br>
The third and final category involved using mostly Zscore normalizations to pool all the data together to investigate the correlation between BMI and Shannon diversity, Bacteroidetes, Firmicutes, and B/F ratio.  Using the Zscore values a two-way t-test was used to investigate if there was a difference between non-obese and obese individuals.  Differences in BMI categories (if data was available) was investigated using ANOVA with a tukey post hoc test.  Linear correlations between these variables and BMI were explored using the Pearson test.  Where appropriate the Bonferroni correction was applied for multiple comparisons.  In all other cases a P-value under 0.05 was considered to be significant.

*******
##### Results:

*******
##### Discussion:

*******
**Table 1:Summary Information of Studies used in Meta-analysis.**     
![Table 1](C:\Users\marc\Desktop\obesity2\SummaryTableStudies.tif)

*******
##### References:

### 2016-02-05 code/StatisticsProcessing.Rmd
Initial commit

---
title: "Bacterial Microbiome and Obesity Meta-analysis Statistics Pipeline"
author: "Marc Sze"
date: "January 29, 2016"
output: 
  html_document:
    keep_md: yes

---
<br>

#### **Preamble**  
The following workflow was used to generate all the statistical analysis of our meta-analysis.  It is organized by part of which there are three.  Code chunks will be displayed as such below:

```{r,  eval=FALSE}
print("Hello World!")
x = 1 + 1
print(x)
```
Data tables are provided in this repository for your use if you do not want to have to repeat the somewhat laborious sequence analysis pipeline from start to finish.  BMI data if publicly available has been kept in this repository for your ease of reference.  However, there are a few data sets in which the BMI data is not public and as such you will need to request this information from the respective sources if you would like to use them.  The two data sets this mostly concerns are the Wu, et al study and the HMP study.
<br><br>

#### **Part 1: Seperate Analysis of each Study**  
The statistical analysis will follow a general order of data loading, alpha diversity and BMI analysis, PERMANOVA of Bray-Curtis distance matrix on obese versus non-obese and BMI classifications (where available), relative risk, and then classification using AUCRF.
<br><br>

##### **Baxter et al. 2016:**  
```{r,eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

# Reading in the necessary Data
setwd("C:/Users/marc/Desktop/obesity2/data1")
demographics <- read.csv("demographics.v2.csv")
microbiome <- read.csv("data1.subsample.otus.csv")
phylogenetic.info <- read.csv("data1.summary.taxonomy.csv")
taxonomy <- read.csv("data1.taxonomy.csv")

# Minor modifications and Rownames adustments of data tables
rownames(demographics) <- demographics[,1]
rownames(microbiome) <- microbiome[,2]
demographics <- demographics[,-1]
microbiome <- microbiome[,-2]
test.samples <- rownames(demographics)
microbiome <- microbiome[,-c(1:2)]

# Create Table for phyla information
phylogenetic.info <- phylogenetic.info[-c(1:12),]
phyla <- which(phylogenetic.info$taxlevel == 2)
phyla.table <- phylogenetic.info[phyla, ]
phyla.table <- phyla.table[,-c(1:2,4:5)]
phyla.names <- as.character(phyla.table[,1])
phyla.table <- phyla.table[,-1]
phyla.table <- as.data.frame(t(phyla.table))
colnames(phyla.table) <- phyla.names
rownames(phyla.table) <- rownames(demographics)

#Add phyla together that are not very abundant and delete them from the table
phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deferribacteres", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "Cyanobacteria_Chloroplast", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(1, 4:7, 9:11, 13, 15)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- phyla.table.rel.abund[, -8]

#Generate Alpha Diversity Measures and alpha diversity table
library(vegan)
H <- diversity(microbiome)
S <- specnumber(microbiome)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)

#Create age quartiles

demographics$age.quartile <- with(demographics, cut(Age, breaks=quantile(Age, probs=seq(0,1, by=0.25)), include.lowest=TRUE))

#Create Obese Yes/No groups
demographics$obese[demographics$BMI.classification=="Normal" | demographics$BMI.classification=="Overweight"] <- "No"
demographics$obese[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Yes"

#Create Obese.num groups
demographics$obese.num[demographics$obese=="No"] <- 0
demographics$obese.num[demographics$obese=="Yes"] <- 1
obese <- factor(demographics$obese.num)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

#Create a column with obese and extreme obese as single entity
demographics$BMIclass2[demographics$BMI.classification=="Normal"] <- "Normal"
demographics$BMIclass2[demographics$BMI.classification=="Overweight"] <- "Overweight"                     
demographics$BMIclass2[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Obese"

##Test BMI versus alpha diversity and phyla

bmi <- demographics$BMI
bmi.class <- factor(demographics$BMI.classification)
bmi.class2 <- factor(demographics$BMIclass2)

anova(lm(H ~ obese)) #P-value=0.01643
anova(lm(H ~ bmi.class)) #P-value=0.06
anova(lm(H ~ bmi.class2)) #P-value=0.04471
a2 <- aov(H ~ bmi.class2)
TukeyHSD(x=a2, 'bmi.class2', conf.level=0.95)
# Obese-Norm (p-value=0.201), Over-Norm (p-value=0.769), Over-Obese (p-value=0.036)

anova(lm(S ~ obese)) #P-value=0.01571
anova(lm(S ~ bmi.class)) #P-value=0.06
anova(lm(S ~ bmi.class2)) #P-value=0.03
a2 <- aov(S ~ bmi.class2)
TukeyHSD(x=a2, 'bmi.class2', conf.level=0.95)
# Obese-Norm (p-value=0.025), Over-Norm (p-value=0.546), Over-Obese (p-value=0.178)

anova(lm(J ~ obese)) #P-value=0.03241
anova(lm(J ~ bmi.class)) #P-value=0.0272
a2 <- aov(J ~ bmi.class)
TukeyHSD(x=a2, 'bmi.class', conf.level=0.95)
# Obese-Norm (p-value=0.02), Over-Norm (p-value=0.047), Over-Obese (p-value=0.06)
anova(lm(J ~ bmi.class2)) #P-value=0.02
a2 <- aov(J ~ bmi.class2)
TukeyHSD(x=a2, 'bmi.class2', conf.level=0.95)
# Obese-Norm (p-value=0.630), Over-Norm (p-value=0.159), Over-Obese (p-value=0.019)

##linear correlations
H.bmi.test <- lm(H ~ bmi)
summary(H.bmi.test)  
#R2 = 0.01246, P-value=0.0774

S.bmi.test <- lm(S ~ bmi)
summary(S.bmi.test)
#R2 = 0.03222, P-value=0.0105

J.bmi.test <- lm(J ~ bmi)
summary(J.bmi.test)
#R2 = -0.0003052, P-value=0.332

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.3175
anova(lm(bacter ~ bmi.class)) #P-value=0.6457
anova(lm(bacter ~ bmi.class2)) #P-value=0.5993

anova(lm(firm ~ obese)) #P-value=0.6817
anova(lm(firm ~ bmi.class)) #P-value=0.9293
anova(lm(firm ~ bmi.class2)) #P-value=0.8936

anova(lm(BFratio ~ obese)) #P-value=0.6305
anova(lm(BFratio ~ bmi.class)) #P-value=0.8334
anova(lm(BFratio ~ bmi.class2)) #P-value=0.881

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

table(demographics[,23])
###Normal=54, Overweight=71, Obese=45, Extreme Obesity=2
###Need to omit extreme obesity
#Removing rows with extreme obesity
test <- which(demographics$BMI.classification == "Extreme Obesity")
BMI.class.demo <- demographics[-test,]
BMI.microbiome <- microbiome[-test,]
bmi.class.group <- factor(BMI.class.demo$BMI.classification)

set.seed(3)
adonis(BMI.microbiome ~ bmi.class.group, permutations=1000)
#Not significant PERMANOVA = 0.2278, pseudo-F = 1.1323

set.seed(3)
adonis(microbiome ~ bmi.class, permutations=1000)
#Not significant PERMANOVA = 0.1848, pseudo-F = 1.1368

set.seed(3)
adonis(microbiome ~ bmi.class2, permutations=1000)
#Not significant PERMANOVA = 0.1359, pseudo-F = 1.1937

###########################################################################
############ Relative Risk#################################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe

#Shannon diversity
median(H) # 3.67469
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 198
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.692766
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Run the RR for Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:86), 4])
table(test3[c(87:172), 4])
#Group1 (Higher than median), obese = 22 and non-obese = 64
#Group2 (Lower than median), obese = 25 and non-obese = 61

group1 <- c(22, 64)
group2 <- c(25, 61)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.14
## CI = 0.70, 1.85
## p-value = 0.608

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 0.3621609
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:86), 2]) #higher group
table(test4[c(87:172), 2]) #lower group
#Group1 (Higher than median), obese = 23 and non-obese = 63
#Group2 (Lower than median), obese = 24 and non-obese = 62

group1 <- c(23, 63)
group2 <- c(24, 62)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.04
## CI = 0.64, 1.70
## p-value = 0.864

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

library(AUCRF)

#generate test set
  # get rid of values that only have 0 and something else
testset <- Filter(function(x)(length(unique(x))>2), microbiome)
  # get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), microbiome)

#Need to add phyla and alpha diversity measures to the dataset

testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
testset <- cbind(obese, testset)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 14 Measures, AUCopt = 0.7553
plot(fit)

```
<br><br>

##### **Ross et al. 2015:**  
```{r, eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/hispanic/")

hispanic.microb <- read.table("hispanic.subsample.shared", header=T)

metadata <- read.csv("s40168-015-0072-y-s1.csv")
sample.match <- read.csv("Hispanic_dataset.csv")

#Create a microbiome data table with sample names from metadata
test <- cbind(as.character(sample.match$Run_s), as.character(sample.match$Library_Name_s))
test <- test[order(test[, 1]), ]

test2 <- cbind(test, hispanic.microb)
rownames(test2) <- test2[, 2]
his.microb.edit <- test2[, -c(1:5)]
edit.metadata <- metadata[, -c(2:5)]
rownames(edit.metadata) <- edit.metadata[, 1]
edit.metadata <- edit.metadata[, -1]

#Create a metadata file in the order of the microbiome data
order1 <- rownames(his.microb.edit)
edit.metadata2 <- edit.metadata[order1, ]

#Get alpha diversity of the samples
library(vegan)

H <- diversity(his.microb.edit)
S <- specnumber(his.microb.edit)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.csv("phyla.data.csv")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(his.microb.edit)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- his.microb.edit
colnames(phyla.table) <- phyla.names
  #add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
  #combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Deinococcus-Thermus", "Elusimicrobia", "Fusobacteria", "Lentisphaerae", "Synergistetes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(3:4, 6:7, 9:10)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- as.data.frame(phyla.table.rel.abund[, -8])

#Create BMI groups
edit.metadata2$BMI.class[edit.metadata2$BMI<=24] <- "Normal"
edit.metadata2$BMI.class[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
edit.metadata2$BMI.class[edit.metadata2$BMI>=30 & edit.metadata2$BMI<40] <- "Obese"
edit.metadata2$BMI.class[edit.metadata2$BMI>=40] <- "Extreme Obesity"

#Create Obese Yes/No groups
edit.metadata2$obese[edit.metadata2$BMI.class=="Normal" | edit.metadata2$BMI.class=="Overweight"] <- "No"
edit.metadata2$obese[edit.metadata2$BMI.class=="Obese" | edit.metadata2$BMI.class=="Extreme Obesity"] <- "Yes"

#Create BMI class 2 groups
edit.metadata2$BMI.class2[edit.metadata2$BMI<=24] <- "Normal"
edit.metadata2$BMI.class2[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
edit.metadata2$BMI.class2[edit.metadata2$BMI>=30] <- "Obese"

#Get paitent demographics to be tested

bmi <- edit.metadata2$BMI
obese <- factor(edit.metadata2$obese)
bmi.class <- factor(edit.metadata2$BMI.class)
bmi.class2 <- factor(edit.metadata2$BMI.class2)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ obese)) #P-value=0.1657
anova(lm(H ~ bmi.class)) #P-value=0.5448
anova(lm(H ~ bmi.class2)) #P-value=0.3615

anova(lm(S ~ obese)) #P-value=0.1841
anova(lm(S ~ bmi.class)) #P-value=0.61
anova(lm(S ~ bmi.class2)) #P-value=0.4

anova(lm(J ~ obese)) #P-value=0.2538
anova(lm(J ~ bmi.class)) #P-value=0.6431
anova(lm(J ~ bmi.class2)) #P-value=0.492

##linear correlations
summary(lm(H ~ bmi)) #P-value=0.343, R2=0.01476
summary(lm(S ~ bmi)) #P-value=0.345, R2=0.01466
summary(lm(J ~ bmi)) #P-value=0.472, R2=0.008502

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.2092
anova(lm(bacter ~ bmi.class)) #P-value=0.5436
anova(lm(bacter ~ bmi.class2)) #P-value=0.4374

anova(lm(firm ~ obese)) #P-value=0.2569
anova(lm(firm ~ bmi.class)) #P-value=0.1741
anova(lm(firm ~ bmi.class2)) #P-value=0.321

anova(lm(BFratio ~ obese)) #P-value=0.2508
anova(lm(BFratio ~ bmi.class)) #P-value=0.4966
anova(lm(BFratio ~ bmi.class2)) #P-value=0.506

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
adonis(his.microb.edit ~ obese, permutations=1000) 
  #PERMANOVA=0.7153, pseudo-F=0.73249

set.seed(3)
adonis(his.microb.edit ~ bmi.class, permutations=1000) 
  #PERMANOVA=0.3357, pseudo-F=1.0772

set.seed(3)
adonis(his.microb.edit ~ bmi.class2, permutations=1000) 
  #PERMANOVA=0.3017, pseudo-F=1.1061

###########################################################################
############ Relative Risk#################################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity
median(H) # 2.747097
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 56
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.6819666
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:31), 4])
table(test3[c(32:63), 4])
#Group1 (Higher than median), obese = 18 and non-obese = 13
#Group2 (Lower than median), obese = 20 and non-obese = 12

group1 <- c(18, 13)
group2 <- c(20, 12)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group1", "group2")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.08
## CI = 0.72, 1.61
## p-value = 0.719

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 1.543046
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:31), 2]) #higher group
table(test4[c(32:63), 2]) #lower group
#Group1 (Higher than median), obese = 22 and non-obese = 9
#Group2 (Lower than median), obese = 16 and non-obese = 16

group1 <- c(22, 9)
group2 <- c(16, 16)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 0.70
## CI = 0.47, 1.07
## p-value = 0.089

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

library(AUCRF)

#Create Obese.num group
edit.metadata2$obese.num[edit.metadata2$obese=="No"] <- 0
edit.metadata2$obese.num[edit.metadata2$obese=="Yes"] <- 1
obese <- factor(edit.metadata2$obese.num)

#generate test set
  # get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), his.microb.edit)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 3 Measures, AUCopt = 0.7526
plot(fit)

```
<br><br>

##### **Goodrich et al. 2014:**  
```{r}

```
<br><br>

##### **Escobar et al. 2014:**  
```{r, eval=FALSE}

###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/columbian/")

columbian.microb <- read.table("columbian.0.03.subsample.shared", header=T)
metadata <- read.csv("columbian_dataset.csv")

#Organize the microbiome shared data
rownames(columbian.microb) <- columbian.microb[, 2]
columbian.microb <- columbian.microb[, -c(1:3)]

#Organize the metadata
rownames(metadata) <- metadata[, 5]

#Get only data that we are interested in
edit.metadata <- metadata[, -c(1:7, 9, 13:15, 16:52)]

#Sort metadata into the same order as the microbiome data
order1 <- rownames(columbian.microb)
edit.metadata2 <- edit.metadata[order1, ]

#Get alpha diversity of the samples
library(vegan)

H <- diversity(columbian.microb)
S <- specnumber(columbian.microb)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.csv("phyla.data.csv")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(columbian.microb)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- columbian.microb
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(4:5, 7:9)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- phyla.table.rel.abund[, -8]

#Create Obese Yes/No groups
edit.metadata2$obese[edit.metadata2$description_s=="Adequate weight male stool sample" | edit.metadata2$description_s=="Overweight male stool sample" | edit.metadata2$description_s=="Adequate weight female stool sample" | edit.metadata2$description_s=="Overweight female stool sample"] <- "No"
edit.metadata2$obese[edit.metadata2$description_s=="Obese male stool sample" | edit.metadata2$description_s=="Obese female stool sample"] <- "Yes"

#Create BMI groups
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s<=24] <- "Normal"
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=30 & edit.metadata2$body_mass_index_s<40] <- "Obese"
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=40] <- "Extreme Obesity"

#Create BMI groups2
edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s<=24] <- "Normal"
edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>=30] <- "Obese"

#Get paitent demographics data to be tested 
bmi <- edit.metadata2$body_mass_index_s
obese <- factor(edit.metadata2$obese)
bmi.class <- factor(edit.metadata2$BMI.class)
bmi.class2 <- factor(edit.metadata2$BMI.class2)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ obese)) #P-value=0.8636
anova(lm(H ~ bmi.class)) #P-value=0.9584
anova(lm(H ~ bmi.class2)) #P-value=0.9584

anova(lm(S ~ obese)) #P-value=0.1947
anova(lm(S ~ bmi.class)) #P-value=0.1432
anova(lm(S ~ bmi.class2)) #P-value=0.1432

anova(lm(J ~ obese)) #P-value=0.4762
anova(lm(J ~ bmi.class)) #P-value=0.764
anova(lm(J ~ bmi.class2)) #P-value=0.764

##linear correlations
summary(lm(H ~ bmi)) #P-value=0.7895, R2=0.002587
summary(lm(S ~ bmi)) #P-value=0.4474, R2=0.02076
summary(lm(J ~ bmi)) #P-value=0.9794, R2=2.423x10-5

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.09253
anova(lm(bacter ~ bmi.class)) #P-value=0.1742
anova(lm(bacter ~ bmi.class2)) #P-value=0.1742

anova(lm(firm ~ obese)) #P-value=0.3602
anova(lm(firm ~ bmi.class)) #P-value=0.5409
anova(lm(firm ~ bmi.class2)) #P-value=0.5409

anova(lm(BFratio ~ obese)) #P-value=0.4663
anova(lm(BFratio ~ bmi.class)) #P-value=0.7545
anova(lm(BFratio ~ bmi.class2)) #P-value=0.7545

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
adonis(columbian.microb ~ obese, permutations=1000) 
#PERMANOVA=0.08891, pseudo-F=1.3797

set.seed(3)
adonis(columbian.microb ~ bmi.class, permutations=1000) 
  #PERMANOVA=0.2358, pseudo-F=1.1142

set.seed(3)
adonis(columbian.microb ~ bmi.class2, permutations=1000) 
  #PERMANOVA=0.2358, pseudo-F=1.1142

###########################################################################
############ Relative Risk#################################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity
median(H) # 4.080262
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 362
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.6896595
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:15), 4])
table(test3[c(16:30), 4])
#Group1 (Higher than median), obese = 4 and non-obese = 11
#Group2 (Lower than median), obese = 6 and non-obese = 9

group1 <- c(4, 11)
group2 <- c(6, 9)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group1", "group2")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.50
## CI = 0.53, 4.26
## p-value = 0.439

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 0.2516729
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:15), 2]) #higher group
table(test4[c(16:30), 2]) #lower group
#Group1 (Higher than median), obese = 7 and non-obese = 8
#Group2 (Lower than median), obese = 3 and non-obese = 12

group1 <- c(7, 8)
group2 <- c(3, 12)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

library(epiR)
epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 0.43
## CI = 0.14, 1.35
## p-value = 0.121

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

library(AUCRF)

#Create Obese.num group
edit.metadata2$obese.num[edit.metadata2$obese=="No"] <- 0
edit.metadata2$obese.num[edit.metadata2$obese=="Yes"] <- 1
obese <- factor(edit.metadata2$obese.num)

#generate test set
  # get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), columbian.microb)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 3 Measures, AUCopt = 0.95
plot(fit)

```

##### **Zupancic et al. 2012:**

```{r, eval=FALSE}

###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/Amish/updatedGOOD")

#Read in and match metadata to microbiome data
metadata <- read.csv("amish_obesity_table2.csv")
test <- metadata[!duplicated(metadata$submitted_sample_id_s), ]
rownames(test) <- test[, 4]
shared.data <- read.table("combined.0.03.subsample.shared", header=T)
rownames(shared.data) <- shared.data[, 2]
shared.data <- shared.data[, -c(1:3)]
keep  <- rownames(shared.data)

test2 <- test[keep, ]
good.metadata <- test2
rm(metadata, test, test2)

metadata <- good.metadata[!grepl("_s2", good.metadata$submitted_sample_id_s),]
keep <- rownames(metadata)
microbiome <- shared.data[keep, ]
rm(good.metadata, shared.data, keep)

#generate alpha diversity measures with vegan
library(vegan)
H <- diversity(microbiome)
S <- specnumber(microbiome)
J <- H/log(S)
select.alpha.diversity <- as.data.frame(cbind(H, S, J))
s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
alpha.test <- s1.alpha.diversity

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.csv("phyla.data.csv")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(microbiome)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- microbiome
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Spirochaetes")], 1, sum)
phyla.table <- phyla.table[, -c(4, 6)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:6)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100

#Create Obese Yes/No groups
# One problem is that we are assuming those with "not provided" are defaulting to the not affected group
metadata$obese[metadata$subject_is_affected_s=="<not provided>" | metadata$subject_is_affected_s=="No"] <- "No"
metadata$obese[metadata$subject_is_affected_s=="Yes"] <- "Yes"

#Create Obese.num group
metadata$obese.num[metadata$obese=="No"] <- 0
metadata$obese.num[metadata$obese=="Yes"] <- 1

#create groups to be used
obese <- factor(metadata$obese)

####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
###########################################################################
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ obese)) #P-value=0.2145
anova(lm(S ~ obese)) #P-value=0.05925
anova(lm(J ~ obese)) #P-value=0.5093

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.9485
anova(lm(firm ~ obese)) #P-value=0.2379
anova(lm(BFratio ~ obese)) #P-value=0.9277

####################################################################################### NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
adonis(microbiome ~ obese, permutations=1000) 
#PERMANOVA=0.5504, pseudo-F=0.97273

###########################################################################
############ Relative Risk#################################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity

median(H) # 3.073974
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 93
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.6790902
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:100), 4])
table(test3[c(101:201), 4])
#Group1 (Higher than median), obese = 29 and non-obese = 71
#Group2 (Lower than median), obese = 45 and non-obese = 56

group1 <- c(29, 71)
group2 <- c(45, 56)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.54
## CI = 1.05, 2.24
## p-value = 0.022

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 0.537415
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:100), 2]) #higher group
table(test4[c(101:201), 2]) #lower group
#Group1 (Higher than median), obese = 38 and non-obese = 62
#Group2 (Lower than median), obese = 36 and non-obese = 65

group1 <- c(38, 62)
group2 <- c(36, 65)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 0.94
## CI = 0.65, 1.35
## p-value = 0.729

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

library(AUCRF)

#Create Obese.num group
obese <- factor(metadata$obese.num)

#generate test set
# get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), microbiome)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 32 Measures, AUCopt = 0.6679613
plot(fit)
```

##### **Yatsunenko et al. 2012:**  
```{r}

```

##### **Human Microbiome Project (HMP) 2011:**  
```{r, eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

##Read in relevant data

setwd("C:/users/marc/Desktop/obesity2/HMP.analysis")
microbiome <- read.table("Stool.an.0.03.subsample.shared", header=T)
meta.cat <- read.table("categorical.metadata", header=T)
meta.cont <- read.table("continuous.metadata", header=T)

#Only interested in first visit microbiome data
#need to create microbiome data set for only that

test <- microbiome[grep("\\.01\\.", microbiome$Group), ] #selecting by .01.

microb.rownames <- gsub("([0-9]+).*", "\\1", test$Group) #extracting only numeric before first "."
rownames(test) <- microb.rownames

#Get rid of information not used in downstram analysis
test <- test[, -c(1:3)]

#Subset data to the microbiome total n
select.meta.cat <- meta.cat[microb.rownames, ]
select.meta.cont <- meta.cont[microb.rownames, ]

#generate alpha diversity measures with vegan
library(vegan)
H <- diversity(test)
S <- specnumber(test)
J <- H/log(S)
select.alpha.diversity <- as.data.frame(cbind(H, S, J))
alpha.test <- select.alpha.diversity

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.csv("phyla.data.csv")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(test)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- test
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)

#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(1, 4, 6, 7, 9, 10:12)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- phyla.table.rel.abund[, -8]

#Create Obese Yes/No groups
select.meta.cat$obese[select.meta.cat$BMI_C=="normal" | select.meta.cat$BMI_C=="overweight"] <- "No"
select.meta.cat$obese[select.meta.cat$BMI_C=="obese" | select.meta.cat$BMI_C=="extreme obesity"] <- "Yes"

#Create BMI groups2
select.meta.cat$BMI.class2[select.meta.cont$DTPBMI<=24] <- "Normal"
select.meta.cat$BMI.class2[select.meta.cont$DTPBMI>24 & select.meta.cont$DTPBMI<30] <- "Overweight"
select.meta.cat$BMI.class2[select.meta.cont$DTPBMI>=30] <- "Obese"

#Get paitent demographics data to be tested 
bmi <- select.meta.cont$DTPBMI
obese <- factor(select.meta.cat$obese)
bmi.class <- factor(select.meta.cat$BMI_C)
bmi.class2 <- factor(select.meta.cat$BMI.class2)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ obese)) #P-value=0.585
anova(lm(H ~ bmi.class)) #P-value=0.322
anova(lm(H ~ bmi.class2)) #P-value=0.322

anova(lm(S ~ obese)) #P-value=0.845
anova(lm(S ~ bmi.class)) #P-value=0.2051
anova(lm(S ~ bmi.class2)) #P-value=0.2051

anova(lm(J ~ obese)) #P-value=0.3559
anova(lm(J ~ bmi.class)) #P-value=0.3501
anova(lm(J ~ bmi.class2)) #P-value=0.3501

##linear correlations
summary(lm(H ~ bmi)) #P-value=0.7041, R2=0.0005689
summary(lm(S ~ bmi)) #P-value=0.2304, R2=0.005657
summary(lm(J ~ bmi)) #P-value=0.962, R2=8.932x10-6

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.3518
anova(lm(bacter ~ bmi.class)) #P-value=0.1343
anova(lm(bacter ~ bmi.class2)) #P-value=0.1343

anova(lm(firm ~ obese)) #P-value=0.636
anova(lm(firm ~ bmi.class)) #P-value=0.07813
anova(lm(firm ~ bmi.class2)) #P-value=0.07813

anova(lm(BFratio ~ obese)) #P-value=0.5688
anova(lm(BFratio ~ bmi.class)) #P-value=0.6807
anova(lm(BFratio ~ bmi.class2)) #P-value=0.6807

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
adonis(test ~ obese, permutations=1000) 
#PERMANOVA=0.8112, pseudo-F=0.7024

set.seed(3)
adonis(test ~ bmi.class, permutations=1000) 
  #PERMANOVA=0.8022, pseudo-F=0.77896

set.seed(3)
adonis(test ~ bmi.class2, permutations=1000) 
  #PERMANOVA=0.8022, pseudo-F=77896

######################################################################################## Relative Risk ###############################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity
median(H) # 2.541824
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 57
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.6305742
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:128), 4])
table(test3[c(129:256), 4])
#Group1 (Higher than median), obese = 10 and non-obese = 118
#Group2 (Lower than median), obese = 16 and non-obese = 112

group1 <- c(10, 118)
group2 <- c(16, 112)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group1", "group2")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.60
## CI = 0.75, 3.39
## p-value = 0.214

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 2.529168
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:128), 2]) #higher group
table(test4[c(129:256), 2]) #lower group
#Group1 (Higher than median), obese = 13 and non-obese = 115
#Group2 (Lower than median), obese = 13 and non-obese = 115

group1 <- c(13, 115)
group2 <- c(13, 115)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

library(epiR)
epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.00
## CI = 0.48, 2.07
## p-value = 1.00

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

library(AUCRF)

#Create Obese.num group
select.meta.cat$obese.num[select.meta.cat$obese=="No"] <- 0
select.meta.cat$obese.num[select.meta.cat$obese=="Yes"] <- 1
obese <- factor(select.meta.cat$obese.num)

#generate test set
  # get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), test)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 6 Measures, AUCopt = 0.7031773
plot(fit)

```

##### **Nam, et al. 2011:**  
<br>
It should be noted that for the grouping part of this analysis it was done in relationship to being overweight instead of obese. But for the combined portion of the analysis or third arm (all data sets taken together) it was used.  Conclusions on the other arms are not dependent on this specific data set. 
```{r, eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/korean/")

microb <- read.table("DRR000776.0.03.subsample.shared", header=T)
rownames(microb) <- microb[, 2]
microb <- microb[,-c(1:3)]

metadata <- read.csv("korean.metadata.csv")

#Need to get rid of the repeated samples and the children
  #First the metadata
metadata2 <- metadata[-c(2:3, 5:6, 8:9, 11:12, 14:15, 17:18, 19:24), ]
rownames(metadata2) <- metadata2[, 2]
metadata2 <- metadata2[, -c(1:2)]

  #Second the microbiome data set
microb2 <- microb[-c(2:3, 5:6, 8:9, 11:12, 14:15, 17:18, 19:24), ]
rownames(microb2) <- rownames(metadata2)

#Get alpha diversity of the samples
library(vegan)

H <- diversity(microb2)
S <- specnumber(microb2)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)
# Seems to look okay....

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.csv("phyla.data.csv")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(microb2)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- microb2
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Synergistetes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(4:5, 7:8)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- phyla.table.rel.abund[, -8]

#Create BMI groups
metadata2$BMI.class[metadata2$BMI<=24] <- "Normal"
metadata2$BMI.class[metadata2$BMI>24 & metadata2$BMI<30] <- "Overweight"
metadata2$BMI.class[metadata2$BMI>=30 & metadata2$BMI<40] <- "Obese"
metadata2$BMI.class[metadata2$BMI>=40] <- "Extreme Obesity"

#Create Overweight Yes/No groups
metadata2$overweight[metadata2$BMI.class=="Normal"] <- "No"
metadata2$overweight[metadata2$BMI.class=="Obese" | metadata2$BMI.class=="Extreme Obesity" | metadata2$BMI.class=="Overweight"] <- "Yes"

#Get paitent demographics data to be tested 
bmi <- metadata2$BMI
overweight <- factor(metadata2$overweight)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ overweight)) #P-value=0.961
anova(lm(S ~ overweight)) #P-value=0.5165
anova(lm(J ~ overweight)) #P-value=0.6547

##linear correlations
summary(lm(H ~ bmi)) #P-value=0.8838, R2=0.001376
summary(lm(S ~ bmi)) #P-value=0.3153, R2=0.06293
summary(lm(J ~ bmi)) #P-value=0.7022, R2=0.009386

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ overweight)) #P-value=0.4382
anova(lm(firm ~ overweight)) #P-value=0.2609
anova(lm(BFratio ~ overweight)) #P-value=0.4089

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
adonis(microb2 ~ overweight, permutations=1000) 
#PERMANOVA=0.4236, pseudo-F=0.97425

######################################################################################## Relative Risk ###############################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity
median(H) # 3.888484
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 156.5
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.7715805
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(overweight)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:9), 4])
table(test3[c(10:18), 4])
#Group1 (Higher than median), overweight = 4 and normal = 5
#Group2 (Lower than median), overweight = 2 and normal = 7

group1 <- c(4, 5)
group2 <- c(2, 7)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Overweight", "normal")
rownames(r.test) <- c("group1", "group2")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 0.50
## CI = 0.12, 2.08
## p-value = 0.317

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 0.4243408
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, overweight)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:9), 2]) #higher group
table(test4[c(10:18), 2]) #lower group
#Group1 (Higher than median), overweight = 2 and normal = 7
#Group2 (Lower than median), overweight = 4 and normal = 5

group1 <- c(2, 7)
group2 <- c(4, 5)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

library(epiR)
epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 2.00
## CI = 0.48, 8.31
## p-value = 0.317

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

library(AUCRF)

#Create Obese.num group
metadata2$over.num[metadata2$overweight=="No"] <- 0
metadata2$over.num[metadata2$overweight=="Yes"] <- 1
overweight <- factor(metadata2$over.num)

#generate test set
  # get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), microb2)
testset <- cbind(overweight, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=500, nodesize=1)
  #Since this data set was so small had to modify the standardized        
  #parameters.
summary(fit) # list of 2 Measures, AUCopt = 1.00
plot(fit)

```

##### **Wu, et al. 2011:**

```{r, eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/COMBO/")

microbiome <- read.table("combined.good.subsample.shared", header=T)
rownames(microbiome) <- microbiome$Group
microbiome <- microbiome[, -c(1:3)]
seqData <- read.csv("COMBO_data_table.csv", header=T)
seqData <- seqData[-1, ]
rownames(seqData) <- seqData$Run_s
metadata <- read.table("BMI_and_Counts.txt", header=T)
metadata <- metadata[, c(1, 2)]

#get seqData set in line with microbiome data
namesToKeep <- rownames(microbiome)
test <- seqData[namesToKeep, ]
seqData <- test
rm(test)

#change all rownames to match sample IDs
rownames(seqData) <- seqData$submitted_subject_id_s
rownames(microbiome) <- rownames(seqData)
rownames(metadata) <- metadata[, 1]

#Match the metadata now with the microbiome data
namesToKeep <- rownames(microbiome)
test <- metadata[namesToKeep, ]
metadata <- test
rm(test)
rm(seqData, keep, namesToKeep)

#Get alpha diversity of the samples
library(vegan)

H <- diversity(microbiome)
S <- specnumber(microbiome)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.table("combined.good.phyla.txt", header=T)
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(microbiome)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- microbiome
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Synergistetes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(4, 6:7)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:6)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
rm(keep, phyla.names, phyla.total, phylogenetic.info, phyla.table, phyla.good)

#Create BMI groups
metadata$BMI.class[metadata$BMI<=24] <- "Normal"
metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"

#Create Obese Yes/No groups
metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"

#Create a column with obese and extreme obese as single entity
metadata$BMIclass2[metadata$BMI.class=="Normal"] <- "Normal"
metadata$BMIclass2[metadata$BMI.class=="Overweight"] <- "Overweight"                     
metadata$BMIclass2[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Obese"

#Get paitent demographics to be tested
bmi <- metadata$BMI
obese <- factor(metadata$obese)
bmi.class <- factor(metadata$BMI.class)
bmi.class2 <- factor(metadata$BMIclass2)

###########################################################################
######### First Level Analysis & Alpha Diversity with BMI #################
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ obese)) #P-value=0.7493
anova(lm(H ~ bmi.class)) #P-value=0.2959
anova(lm(H ~ bmi.class2)) #P-value=0.2089

anova(lm(S ~ obese)) #P-value=0.7511
anova(lm(S ~ bmi.class)) #P-value=0.5128
anova(lm(S ~ bmi.class2)) #P-value=0.3891

anova(lm(J ~ obese)) #P-value=0.5557
anova(lm(J ~ bmi.class)) #P-value=0.2647
anova(lm(J ~ bmi.class2)) #P-value=0.1818

##linear correlations
summary(lm(H ~ bmi)) #P-value=0.8276, R2=0.0008541
summary(lm(S ~ bmi)) #P-value=0.8195, R2=0.0009377
summary(lm(J ~ bmi)) #P-value=0.8521, R2=0.0006262

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.9173
anova(lm(bacter ~ bmi.class)) #P-value=0.5554
anova(lm(bacter ~ bmi.class2)) #P-value=0.6846

anova(lm(firm ~ obese)) #P-value=0.5765
anova(lm(firm ~ bmi.class)) #P-value=0.2157
anova(lm(firm ~ bmi.class2)) #P-value=0.2817

anova(lm(BFratio ~ obese)) #P-value=0.6735
anova(lm(BFratio ~ bmi.class)) #P-value=0.6208
anova(lm(BFratio ~ bmi.class2)) #P-value=0.4681

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
adonis(microbiome ~ obese, permutations=1000) 
#PERMANOVA=0.8921, pseudo-F=0.66065

set.seed(3)
adonis(microbiome ~ bmi.class, permutations=1000) 
#PERMANOVA=0.9321, pseudo-F=0.73921

set.seed(3)
adonis(microbiome ~ bmi.class2, permutations=1000) 
#PERMANOVA=0.75922, pseudo-F=0.8791

###########################################################################
############ Relative Risk#################################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity
median(H) # 3.000655
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 98.5
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.6432727
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:29), 4])
table(test3[c(30:58), 4])
#Group1 (Higher than median), obese = 3 and non-obese = 26
#Group2 (Lower than median), obese = 2 and non-obese = 27

group1 <- c(3, 26)
group2 <- c(2, 27)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 0.67
## CI = 0.12, 3.70
## p-value = 0.64

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 2.438335
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:29), 2]) #higher group
table(test4[c(30:58), 2]) #lower group
#Group1 (Higher than median), obese = 3 and non-obese = 26
#Group2 (Lower than median), obese = 2 and non-obese = 27

group1 <- c(3, 26)
group2 <- c(2, 27)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 0.67
## CI = 0.12, 3.70
## p-value = 0.64

###########################################################################
############ Classification using AUCRF####################################
###########################################################################

library(AUCRF)

#Create Obese.num group
metadata$obese.num[metadata$obese=="No"] <- 0
metadata$obese.num[metadata$obese=="Yes"] <- 1
obese <- factor(metadata$obese.num)

#generate test set
# get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), microbiome)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 15 Measures, AUCopt = 0.7075472
plot(fit)

```

##### **Arumugam, et al. 2010 (MetaHit):**  
<br>
This first R analysis chunk looks at how similar metaphlan2 is with 16S sequencing in the HMP data set.
```{r, eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

#Phyla first then will explore other measures if data is promising

setwd("C:/Users/marc/Desktop/obesity2/HMP.analysis")

metaphlanData <- read.table("HMP_metaphlan2_merged_abundance_table.txt", header=T)
IDconversion <- read.csv("HMP.stool.IDs.csv")
microbiome <- read.table("Stool.an.0.03.subsample.shared", header=T)

#Only interested in first visit microbiome data
#need to create microbiome data set for only that

Data16S <- microbiome[grep("\\.01\\.", microbiome$Group), ] #selecting by .01.

microb.rownames <- gsub("([0-9]+).*", "\\1", Data16S$Group) #extracting only numeric before first "."
rownames(Data16S) <- microb.rownames

#Get rid of information not used in downstram analysis
Data16S <- Data16S[, -c(1:3)]

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.csv("phyla.data.csv")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(Data16S)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- Data16S
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing, keep, microb.rownames, phyla.names)

#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(1, 4, 6, 7, 9, 10:12)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
rm(phyla.total)

# Create comparable data set with 
metaphlanPhlyaData <- read.csv("HMP.metaphlan.phyla.data.csv", header=T)
rownames(metaphlanPhlyaData) <- metaphlanPhlyaData$ID
metaphlanPhlyaData <- metaphlanPhlyaData[,-1]
metaphlanPhlyaData <- as.data.frame(t(metaphlanPhlyaData))

#combine phyla that are not that abundant
metaphlanPhlyaData$other <- apply(metaphlanPhlyaData[, c("Acidobacteria", "Candidatus_Saccharibacteria", "Deinococcus_Thermus", "Fusobacteria", "Spirochaetes", "Synergistetes", "Tenericutes")], 1, sum)
metaphlanPhlyaData <- metaphlanPhlyaData[, -c(1, 4:5, 7, 9:11)]

#Create a relative abundance table for phyla
metaphlanPhlyaData.total <- apply(metaphlanPhlyaData[, c(1:6)], 1, sum)
metaphlanPhlyaData.rel.abund <- (metaphlanPhlyaData/metaphlanPhlyaData.total)*100

#Need to get sampleIDs to match 16S data
metaphlanNames <- rownames(metaphlanPhlyaData)
rownames(IDconversion) <- IDconversion[,1]
matchedData <- IDconversion[metaphlanNames, c(1,3)]
IndIDmatched <- matchedData[,2]

#Need to find and remove duplicated values (No idea which sample goes with which)
test1 <- as.data.frame(table(matchedData$Individual.ID))
good <- which(test1$Freq == 1)
unique <- test1[good, ]
uniqueNames <- as.character(unique$Var1)

test <- phyla.table.rel.abund[rownames(phyla.table.rel.abund) %in% uniqueNames, ]
finalUniqueNames <- rownames(test)
UpdatedMatchedData <- matchedData[matchedData$Individual.ID %in% finalUniqueNames, ]

goodMetaphlanPhylaData.rel.abund <- metaphlanPhlyaData.rel.abund[rownames(metaphlanPhlyaData.rel.abund) %in% rownames(UpdatedMatchedData), ]
rownames(goodMetaphlanPhylaData.rel.abund) <- UpdatedMatchedData[, 2]
testing <- goodMetaphlanPhylaData.rel.abund[order(rownames(goodMetaphlanPhylaData.rel.abund)), ]

phyla16SData <- test
phylaMetaphlanData <- testing
rm(finalUniqueNames, good, IndIDmatched, metaphlanNames, metaphlanPhlyaData.total, uniqueNames, goodMetaphlanPhylaData.rel.abund, matchedData, metaphlanPhlyaData, metaphlanPhlyaData.rel.abund, phyla.good, phyla.table, phyla.table.rel.abund, phylogenetic.info, test, test1, testing, unique, UpdatedMatchedData)

###########################################################################
############ Let's Get to the Actual Comparison ###########################
###########################################################################

# Can start off with a general bar plot of the two side by side 
library(qdap)
library(plyr)
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(gridExtra) #using grid.arrange() can add different plots together
library(plotrix)

#phyla16SData <- phyla16SData[,-5] only if want to ignore unclassified
phylaMetaphlanData$unclassified <- 0
phylaMetaphlanData <- subset(phylaMetaphlanData, select=c(Actinobacteria:Proteobacteria, unclassified, Verrucomicrobia, other))
phyla16SData$dataset <- 1 
phylaMetaphlanData$dataset <- 2

testset <- rbind(phyla16SData, phylaMetaphlanData)
testset$dataset <- factor(testset$dataset)
melt.testset <- melt(testset, id.vars=c("dataset"))
colnames(melt.testset)[2] <- "phyla"
barplotAnalysisData <- ddply(melt.testset, c("dataset", "phyla"), summarise, mean = mean(value), sd=sd(value), sem=sd(value/sqrt(length(value))))

bpg1 <- ggplot(barplotAnalysisData, aes(x=dataset, y=mean, fill=phyla)) + geom_bar(position = "fill", stat="identity")
bpg2 <- bpg1 + scale_fill_brewer(palette="Set1") + scale_y_continuous(label=percent)
bpg2

# Now can look at direct linear correlation between same sample
# Between metaphlan2 and 16S amplicon sequencing

#Actinobacteria
summary(lm(phyla16SData$Actinobacteria ~ phylaMetaphlanData$Actinobacteria)) 
#P-value 0.5543, R2= 0.006899
plot(phyla16SData$Actinobacteria ~ phylaMetaphlanData$Actinobacteria)
abline(lm(phyla16SData$Actinobacteria ~ phylaMetaphlanData$Actinobacteria), col="red")

#Bacteroidetes
summary(lm(phyla16SData$Bacteroidetes ~ phylaMetaphlanData$Bacteroidetes)) 
#P-value 0.0002277, R2= 0.2358
plot(phyla16SData$Bacteroidetes ~ phylaMetaphlanData$Bacteroidetes)
abline(lm(phyla16SData$Bacteroidetes ~ phylaMetaphlanData$Bacteroidetes), col="red")

#Firmicutes
summary(lm(phyla16SData$Firmicutes ~ phylaMetaphlanData$Firmicutes)) 
#P-value 0.1223, R2= 0.04617
plot(phyla16SData$Firmicutes ~ phylaMetaphlanData$Firmicutes)
abline(lm(phyla16SData$Firmicutes ~ phylaMetaphlanData$Firmicutes), col="red")

#Proteobacteria
summary(lm(phyla16SData$Proteobacteria ~ phylaMetaphlanData$Proteobacteria)) 
#P-value 0.004252, R2= 0.1494
plot(phyla16SData$Proteobacteria ~ phylaMetaphlanData$Proteobacteria)
abline(lm(phyla16SData$Proteobacteria ~ phylaMetaphlanData$Proteobacteria), col="red")

#Verrucomicrobia
summary(lm(phyla16SData$Verrucomicrobia ~ phylaMetaphlanData$Verrucomicrobia)) 
#P-value 0.1761, R2= 0.03559
plot(phyla16SData$Verrucomicrobia ~ phylaMetaphlanData$Verrucomicrobia)
abline(lm(phyla16SData$Verrucomicrobia ~ phylaMetaphlanData$Verrucomicrobia), col="red")

#other
summary(lm(phyla16SData$other ~ phylaMetaphlanData$other)) 
#P-value 0.8054, R2= 0.001201
plot(phyla16SData$other ~ phylaMetaphlanData$other)
abline(lm(phyla16SData$other ~ phylaMetaphlanData$other), col="red")

# Create a bland Altman of the best correlation
  # bias (difference of metaphlan) based on 16S analysis  
library(BlandAltmanLeh)
bland.altman.plot(phyla16SData$Bacteroidetes, phylaMetaphlanData$Bacteroidetes, main="Bland Altman Plot of Bacteroidetes", xlab="Means", ylab="Differences")

```

<br>
This second R analysis chunk uses metaphlan2 to look at the questions we were tackling with the other data sets indivdiually.  
```{r, eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/MetaHit/")

microb <- read.csv("finalized_metaHit_shared.csv")
rownames(microb) <- microb[, 1]
microb <- microb[,-1]

metadata <- read.csv("MetaHit_metadata.csv")
rownames(metadata) <- metadata[, 1]
metadata <- metadata[, -1]

phyla <- read.csv("phyla.csv")
rownames(phyla) <- phyla[, 1]
phyla <- phyla[, -1]

# Generate Overall Relative Abundance - normalize to a value of 100
# Needs to be done since not all measures have 100 bacteria from metaphlan2
overall <- rowSums(microb)
microb.norm <- (microb / overall) * 100

phyla.table <- read.csv("phyla.csv", header = T)
rownames(phyla.table) <- phyla.table[, 1]
phyla.table <- phyla.table[, -1]
phyla.table <- as.data.frame(t(phyla.table))

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:10)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
rm(phyla.table, phyla.info)

#Get alpha diversity of the samples
library(vegan)

H <- diversity(microb.norm)
S <- specnumber(microb.norm)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)
# Seems to look okay....

#Create BMI groups
metadata$BMI.class[metadata$BMI<=24] <- "Normal"
metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"

#Create Obese Yes/No groups
metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"

#Create Obese.num group
metadata$obese.num[metadata$obese=="No"] <- 0
metadata$obese.num[metadata$obese=="Yes"] <- 1

#Create a column with obese and extreme obese as single entity
metadata$BMIclass2[metadata$BMI<=24] <- "Normal"
metadata$BMIclass2[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"      
metadata$BMIclass2[metadata$BMI>=30] <- "Obese"

#Get paitent demographics to be tested
bmi <- metadata$BMI
obese <- factor(metadata$obese)
bmi.class <- factor(metadata$BMI.class)
bmi.class2 <- factor(metadata$BMIclass2)

####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
###########################################################################
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ obese)) #P-value=0.2174
anova(lm(H ~ bmi.class)) #P-value=0.5339
anova(lm(H ~ bmi.class2)) #P-value=0.3628

anova(lm(S ~ obese)) #P-value=0.9878
anova(lm(S ~ bmi.class)) #P-value=0.7083
anova(lm(S ~ bmi.class2)) #P-value=0.4985

anova(lm(J ~ obese)) #P-value=0.1874
anova(lm(J ~ bmi.class)) #P-value=0.3772
anova(lm(J ~ bmi.class2)) #P-value=0.2326

##linear correlations
summary(lm(H ~ bmi)) #P-value=0.1022, R2=0.03185
summary(lm(S ~ bmi)) #P-value=0.7642, R2=0.001091
summary(lm(J ~ bmi)) #P-value=0.06682, R2=0.0399

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.1911
anova(lm(bacter ~ bmi.class)) #P-value=0.2975
anova(lm(bacter ~ bmi.class2)) #P-value=0.2263

anova(lm(firm ~ obese)) #P-value=0.2925
anova(lm(firm ~ bmi.class)) #P-value=0.3654
anova(lm(firm ~ bmi.class2)) #P-value=0.2998

anova(lm(BFratio ~ obese)) #P-value=0.05254
anova(lm(BFratio ~ bmi.class)) #P-value=0.1808
anova(lm(BFratio ~ bmi.class2)) #P-value=0.1284

####################################################################################### NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
adonis(microb.norm ~ obese, permutations=1000) 
#PERMANOVA=0.05495, pseudo-F=1.6047

set.seed(3)
adonis(microb.norm ~ bmi.class, permutations=1000) 
#PERMANOVA=0.1798, pseudo-F=1.176

set.seed(3)
adonis(microb.norm ~ bmi.class2, permutations=1000) 
#PERMANOVA=0.2138, pseudo-F=1.1686

###########################################################################
############ Relative Risk#################################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity

median(H) # 2.886851
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 85
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.664365
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:42), 4])
table(test3[c(43:85), 4])
#Group1 (Higher than median), obese = 17 and non-obese = 25
#Group2 (Lower than median), obese = 20 and non-obese = 23

group1 <- c(17, 25)
group2 <- c(20, 23)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.15
## CI = 0.71, 1.87
## p-value = 0.575

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 0.9224021
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:42), 2]) #higher group
table(test4[c(43:85), 2]) #lower group
#Group1 (Higher than median), obese = 20 and non-obese = 22
#Group2 (Lower than median), obese = 17 and non-obese = 26

group1 <- c(20, 22)
group2 <- c(17, 26)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 0.83
## CI = 0.51, 1.35
## p-value = 0.452

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

library(AUCRF)

#Create Obese.num group
metadata$obese.num[metadata$obese=="No"] <- 0
metadata$obese.num[metadata$obese=="Yes"] <- 1
obese <- factor(metadata$obese.num)

#generate test set
# get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), microb.norm)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 19 Measures, AUCopt = 0.7649212
plot(fit)
```

##### **Turnbaugh, et al. 2008:**  
```{r, eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/turnbaugh.twins")
shared.data <- read.table("test.unique.good.filter.unique.precluster.pick.pick.an.shared", header=T)
rownames(shared.data) <- shared.data[, 2]
shared.data <- shared.data[, -c(1:3)]

subsample.data <- read.table("test.unique.good.filter.unique.precluster.pick.pick.an.0.03.subsample.shared", header=T)
rownames(subsample.data) <- subsample.data[, 2]
subsample.data <- subsample.data[, -c(1:3)]

metadata <- read.csv("turnbaugh.metadata.csv")
rownames(metadata) <- metadata[, 4]
metadata <- metadata[, -4]

keep1 <- which(metadata$Sample == 1) # n =146

#use the first sampling
#subset data for only the first sampling
s1.metadata <- metadata[keep1, ]
s1.subsample.data <- subsample.data[keep1, ]

#generate alpha diversity measures with vegan
library(vegan)
H <- diversity(s1.subsample.data)
S <- specnumber(s1.subsample.data)
J <- H/log(S)
select.alpha.diversity <- as.data.frame(cbind(H, S, J))
s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
alpha.test <- s1.alpha.diversity

#Generate phyla table data

phyla.info <- read.csv("phyla.csv")
phyla.table <- shared.data
colnames(phyla.table) <- phyla.info[, 2]

#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)

#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(4:5, 7:9)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
s1.phyla.rel.abund <- phyla.table.rel.abund[rownames(shared.data), ]
s1.phyla.rel.abund <- s1.phyla.rel.abund[keep1, ]

#Create Obese Yes/No groups
s1.metadata$obese[s1.metadata$BMI.category=="Lean" | s1.metadata$BMI.category=="Overweight"] <- "No"
s1.metadata$obese[s1.metadata$BMI.category=="Obese"] <- "Yes"

#Create Obese.num group
s1.metadata$obese.num[s1.metadata$obese=="No"] <- 0
s1.metadata$obese.num[s1.metadata$obese=="Yes"] <- 1

#create groups to be used
obese <- factor(s1.metadata$obese)
bmi <- s1.metadata$BMI.category

####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
###########################################################################
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ obese)) #P-value=0.06976
anova(lm(S ~ obese)) #P-value=0.02435
anova(lm(J ~ obese)) #P-value=0.1253

##linear correlations
summary(lm(H ~ bmi)) #P-value=0.1926, R2=0.02277
summary(lm(S ~ bmi)) #P-value=0.03434, R2=0.04606
summary(lm(J ~ bmi)) #P-value=0.2988, R2=0.01675

#B and F tests against obesity
bacter <- s1.phyla.rel.abund$Bacteroidetes
firm <- s1.phyla.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.5771
anova(lm(firm ~ obese)) #P-value=0.4073
anova(lm(BFratio ~ obese)) #P-value=0.2216

####################################################################################### NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
adonis(s1.subsample.data ~ obese, permutations=1000) 
#PERMANOVA=0.09491, pseudo-F=1.2114

###########################################################################
############ Relative Risk#################################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity

median(H) # 4.327308
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 290.5
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.7611711
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:73), 4])
table(test3[c(74:146), 4])
#Group1 (Higher than median), obese = 47 and non-obese = 26
#Group2 (Lower than median), obese = 52 and non-obese = 21

group1 <- c(47, 26)
group2 <- c(52, 21)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.11
## CI = 0.88, 1.38
## p-value = 0.376

##Run the RR for B/F ratio
Bacter = s1.phyla.rel.abund$Bacteroidetes
Firm = s1.phyla.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 0.4764057
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:73), 2]) #higher group
table(test4[c(74:146), 2]) #lower group
#Group1 (Higher than median), obese = 54 and non-obese = 19
#Group2 (Lower than median), obese = 45 and non-obese = 28

group1 <- c(54, 19)
group2 <- c(45, 28)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 0.83
## CI = 0.66, 1.05
## p-value = 0.111

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

library(AUCRF)

#Create Obese.num group
s1.metadata$obese.num[s1.metadata$obese=="No"] <- 0
s1.metadata$obese.num[s1.metadata$obese=="Yes"] <- 1
obese <- factor(s1.metadata$obese.num)

#generate test set
# get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), s1.subsample.data)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, s1.phyla.rel.abund)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 11 Measures, AUCopt = 0.7764883
plot(fit)

```

#### **Part 2: Pooled Meta-analysis RR of Obesity for Shannon Diversity and B/F Ratio**    
<br>
This first R chunk shows the code used to generate the RR for obesity and Shannon diversity.  
```{r, eval=FALSE}
#Load required packages
library(meta)
library(metafor) #This package is used mostly for the more in-depth graph
library(rmeta)  #This package was used for the less intensive graph

#Change working directory
setwd("C:/users/marc/Desktop/obesity2")

StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8)
Study <- c("Baxter", "HMP", "Ross", "Escobar", "Argumugam", "Turnbaugh", "Zupancic", "Wu")
Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012")
Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=201", "n=58")
tpos <- c(25, 16, 20, 6, 20, 52, 45, 2) #obese in low diversity group
tneg <- c(61, 112, 12, 9, 23, 21, 56, 27) #nonobese in low diversity group
cpos <- c(22, 10, 18, 4, 17, 47, 29, 3) #obese in high diversity group
cneg <- c(64, 118, 13, 11, 25, 26, 71, 26) #nonobese in high diversity group
RR <- c(1.14, 1.6, 1.08, 1.50, 1.15, 1.11, 1.54, 0.67)
low <- c(0.70, 0.75, 0.72, 0.53, 0.71, 0.88, 1.05, 0.12)
high <- c(1.85, 3.39, 1.61, 4.26, 1.87, 1.38, 2.24, 3.70)

numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))

dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
k <- length(dat$StudyNo)
dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))

escalc(out ~ grp | study, weights = freq, data = dat.fm, measure = "RR")

#Used to generate the RE Model estimate based on all the studies
res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")

#To create a graph visualize the data
forest(res, at=log(c(0.3, 1, 5)), atransf=exp, xlab="RR of Obesity in high versus low Shannon diversity based on Median", cex=0.8)
par(font=2) #Switch to bold

# RR Title ################################################
#text(3.2, 3.10, "Relative Risk [95% CI]", pos=2, cex=0.8)#
  #large scale (Plot in R studio at full)                 #
                                                          #
text(5.5, 9.3, "Relative Risk [95% CI]", pos=2, cex=0.8)  #
  #smaller scale (Plot in R studia at normal)             #
###########################################################

# Group Title#############################################
                                                         #
text(-6.5, 9.3, "Study, Year, and n", pos=4, cex=0.8)    #
                                                         #
##########################################################

dev.off()

## Alternative way - 
# Seems easier to follow plus I am not trying to do
# a meta analysis on a model

data <-as.data.frame(cbind(RR, low, high))
raw.overall2 <- as.data.frame(cbind(Study,Year, data))

raw.overall2$logRR <- log(raw.overall2$RR)
raw.overall2$SE <- (log(raw.overall2$high) - log(raw.overall2$low))/(2*1.96)
attach(raw.overall2)
metaplot(logRR, SE, labels=Study, nn=1, logeffect=TRUE, col=meta.colors("black"), xlab="RR of Obesity in high versus low Shannon diversity based on Median", ylab="")
  #Graphical output is not as informative

##Calculate combined P-value
m = 8
T = -2*log(0.608)-2*log(0.214)-2*log(0.236)-2*log(0.439)-2*log(0.575)-2*log(0.376)-2*log(0.022)-2*log(0.64)
T.pval = 1-pchisq(T,m) #P-value = 0.01

##Alternatively can use built in package (I trust this more)
anova.rma(res)
# QM = 5.0973, p-value = 0.0240

```
<br>
This second R chunk shows the code used to generate the RR for obesity and B/F ratio.  
```{r, eval=FALSE}
#Load required packages
library(meta)
library(metafor) #This package is used mostly for the more in-depth graph

#Change working directory
setwd("C:/users/marc/Desktop/obesity2")

StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8)
Study <- c("Baxter", "HMP", "Ross", "Escobar", "Argumugam", "Turnbaugh", "Zupancic", "Wu")
Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012")
Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=201", "n=58")
tpos <- c(24, 13, 16, 3, 17, 45, 36, 2) #obese in low BF group
tneg <- c(62, 115, 16, 12, 26, 28, 65, 27) #nonobese in low BF group
cpos <- c(23, 13, 22, 7, 20, 54, 38, 3) #obese in high BF group
cneg <- c(63, 115, 9, 8, 22, 19, 62, 26) #nonobese in high BF group
RR <- c(1.04, 1.00, 0.70, 0.43, 0.83, 0.83, 0.94, 0.67)
low <- c(0.64, 0.48, 0.47, 0.14, 0.51, 0.66, 0.65, 0.12)
high <- c(1.70, 2.07, 1.07, 1.35, 1.35, 1.05, 1.35, 3.70)

numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))

dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
k <- length(dat$StudyNo)
dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))

escalc(out ~ grp | study, weights = freq, data = dat.fm, measure = "RR")

#Used to generate the RE Model estimate based on all the studies
res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")

#To create a graph visualize the data
forest(res, at=log(c(0.1, 1, 3)), atransf=exp, xlab="RR of Obesity in high versus low B/F ratio based on Median", cex=0.8)
par(font=2) #Switch to bold

# RR Title ################################################
#text(3.2, 3.10, "Relative Risk [95% CI]", pos=2, cex=0.8)#
#large scale (Plot in R studio at full)                 #
#
text(5.5, 9.3, "Relative Risk [95% CI]", pos=2, cex=0.8)  #
#smaller scale (Plot in R studia at normal)             #
###########################################################

# Group Title#############################################
#
text(-6.0, 9.3, "Study, Year, and n", pos=4, cex=0.8)    #
#
##########################################################

dev.off()

##Calculate combined P-value
m = 8
T = -2*log(0.864)-2*log(1.00)-2*log(0.089)-2*log(0.121)-2*log(0.452)-2*log(0.111)-2*log(0.729)-2*log(0.64)
T.pval = 1-pchisq(T,m) #P-value = 0.03155827

##Alternatively can use built in package (I trust this more)
anova.rma(res)
  # QM = 4.8452, p-value = 0.0277

```

#### **Part 3: Pooled Analysis Using ZScore Normalizations**

```{r, eval=FALSE}

## Load in libraries to be used as well as directory with the data
library(ggplot2)
library(plyr)
library(reshape2)
library(RColorBrewer)
library(scales)
library(pheatmap)
library(gridExtra)

setwd("C:/Users/marc/Desktop/obesity2/BMI.normalization")

#Load in BMI data
data1 <- read.csv("combined.H.corr.data.csv")
#Load in Turnbaugh data
turn.data <- read.csv("Turnbaugh.normalized.csv")
#Load in Amish data
amish.data <- read.csv("Amish.normalized.csv")

#Amish STD = 0.6010608
amish.ZScore.H <- amish.data$H.corr/0.6010608
hist(amish.ZScore.H)
#Turnbaugh STD = 0.4818751
turn.ZScore.H <- turn.data$H.corr/0.4818751
hist(turn.ZScore.H)
#MetaHit-Den STD = 0.4038558
keepmetahit <- which(data1$dataset == "MetaHit")
metahitdata <- data1[keepmetahit, ]
metahit.ZScore.H <- metahitdata$H.corr/0.4038558
hist(metahit.ZScore.H)
#Korean STD = 0.5400615
keepKorean <- which(data1$dataset == "Korean")
Koreandata <- data1[keepKorean, ]
Korean.ZScore.H <- Koreandata$H.corr/0.5400615
hist(Korean.ZScore.H)
#HMP STD = 0.6774514
keepHMP <- which(data1$dataset == "HMP")
hmpdata <- data1[keepHMP, ]
HMP.ZScore.H <- hmpdata$H.corr/0.6774514
hist(HMP.ZScore.H)
#CCHC STD = 0.5784522
keepCCHC <- which(data1$dataset == "CCHC")
CCHCdata <- data1[keepCCHC, ]
CCHC.ZScore.H <- CCHCdata$H.corr/0.5784522
hist(CCHC.ZScore.H)
#GLNE07 STD = 0.4663871
keepGLNE <- which(data1$dataset == "GLNE")
GLNEdata <- data1[keepGLNE, ]
GLNE.ZScore.H <- GLNEdata$H.corr/0.4663871
hist(GLNE.ZScore.H)
#Escobar STD = 0.5378935
keepescobar <- which(data1$dataset == "Escobar")
escobardata <- data1[keepescobar, ]
Escobar.ZScore.H <- escobardata$H.corr/0.5378935
hist(Escobar.ZScore.H)
#COMBO STD = 0.6870763
keepCOMBO <- which(data1$dataset == "COMBO")
COMBOdata <- data1[keepCOMBO, ]
COMBO.ZScore.H <- COMBOdata$H.corr/0.6870763
hist(COMBO.ZScore.H)

#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
HZScore <- c(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, COMBO.ZScore.H)
data1 <- cbind(HZScore, data1)

amish.data <- cbind(amish.ZScore.H, amish.data)
turn.data <- cbind(turn.ZScore.H, turn.data)
rm(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, amish.ZScore.H, turn.ZScore.H, COMBO.ZScore.H)
rm(keepmetahit, keepCCHC, keepescobar, keepGLNE, keepHMP, keepKorean)

#Create BMI groups
data1$BMI.class[data1$bmi<=24] <- "Lean"
data1$BMI.class[data1$bmi>24 & data1$bmi<30] <- "Overweight"
data1$BMI.class[data1$bmi>=30] <- "Obese"

#Create Obese Yes/No groups
data1$obese[data1$BMI.class=="Lean" | data1$BMI.class=="Overweight"] <- "No"
data1$obese[data1$BMI.class=="Obese"] <- "Yes"

#Create group analysis table to harness full data set
#For ZScore normalization
d1.obese <- data1$obese
t.obese <- as.character(turn.data$obese)
am.obese <- as.character(amish.data$obese)

d1.bmi.class <- data1$BMI.class
t.bmi.class <- as.character(turn.data$BMI.category)

d1.H.Zscore <- data1$HZScore
t.H.Zscore <- turn.data$turn.ZScore.H
am.H.Zscore<- amish.data$amish.ZScore.H

#Combine everything together
comb1.obese <- c(d1.obese, t.obese)
comb2.obese <- c(d1.obese, t.obese, am.obese)
comb1.bmi.class <- c(d1.bmi.class, t.bmi.class)
comb1.H.Zscore <- as.data.frame(c(d1.H.Zscore, t.H.Zscore))
comb2.H.Zscore <- as.data.frame(c(d1.H.Zscore, t.H.Zscore, am.H.Zscore))

group.data.test1 <- cbind(comb1.H.Zscore, comb1.obese, comb1.bmi.class)
group.data.test2 <- cbind(comb2.H.Zscore, comb2.obese)
colnames(group.data.test1) <- c("H.Zscore", "obese", "BMI.class")
colnames(group.data.test2) <- c("H.Zscore", "obese")
#Look at Obese versus non-obese
t.test(H.Zscore ~ obese, data=group.data.test2) #P-value = 0.002792
table(group.data.test2$obese) # No = 693 | Yes = 336
p1 <- ggplot(group.data.test2, aes(obese, H.Zscore)) + geom_boxplot()

#A more in-depth look, breaking down the groups
anova(lm(H.Zscore ~ BMI.class, data=group.data.test1)) #P-value = 0.01783
a2 <- aov(H.Zscore ~ BMI.class, data=group.data.test1)
posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
  # Obese-Lean (p-value=0.070), Over-Norm (p-value=0.791), Over-Obese (p-value=0.021)

table(group.data.test1$BMI.class) 
  # Lean = 326 | Overweight = 240 | Obese = 262
ddply(group.data.test1, ~BMI.class, summarise, mean=mean(H.Zscore), sd=sd(H.Zscore))
  # Lean = 0.042 +\- 0.940 | Overweight = 0.097 +/- 0.993 | Obese = -0.140 +\- 1.052

#Let's change the order so it makes sense
group.data.test1$BMI.class <- factor(group.data.test1$BMI.class, levels = c("Lean", "Overweight", "Obese"))

p2 <- ggplot(group.data.test1, aes(BMI.class, H.Zscore)) + geom_boxplot()

summary(lm(HZScore ~ bmi, data=data1)) #P-value 0.06558, R2= 0.004977
plot(HZScore ~ bmi, data=data1)
abline(lm(HZScore ~ bmi, data=data1), col="red")

#Make a better graph with ggplot2
p1 <- ggplot(data1, aes(x=bmi, y=HZScore))
p2 <- p1 + geom_point(aes(color=dataset))
p3 <- p2 + scale_colour_brewer(palette="Set1")
p4 <- p3 + geom_smooth(method="lm", se=TRUE)

#Probably want to add the Bacteroidetes / Firmicute ratio
#Large variation in data sets observed so will have to normalize (zscore)
#The data looks funny with either a log numerator with log sd denominator 
#or just sd as the denominator
#Think about using only Zscore but this produces a very large tail

setwd("C:/Users/marc/Desktop/obesity2/BF.normalization")

#Load in data
data2 <- read.csv("combined.BF.corr.csv")

#Create BMI groups
data2$BMI.class[data2$bmi<=24] <- "Lean"
data2$BMI.class[data2$bmi>24 & data2$bmi<30] <- "Overweight"
data2$BMI.class[data2$bmi>=30] <- "Obese"

#Create Obese Yes/No groups
data2$obese[data2$BMI.class=="Lean" | data2$BMI.class=="Overweight"] <- "No"
data2$obese[data2$BMI.class=="Obese"] <- "Yes"

#Amish SD = 0.7990372
am.BF.corr <- as.character(amish.data$BF.ratio.corr)
am.BF.corr <- as.numeric(am.BF.corr) # Value 19 has no value
amish.norm.BF <- am.BF.corr/0.7990372
hist(amish.norm.BF)
#Turnbaugh SD = 0.4578937
turn.norm.BF <- turn.data$BF.corr/0.4578937
hist(turn.norm.BF)
#MetaHit SD = 0.9707577
keepmetahit <- which(data2$dataset == "MetaHit")
metahitdata <- data2[keepmetahit, ]
metahit.norm.BF <- metahitdata$BF.ratio.corr/0.9707577
hist(metahit.norm.BF)
#Korean SD = 0.6048195
keepKorean <- which(data2$dataset == "Korean")
Koreandata <- data2[keepKorean, ]
Korean.norm.BF <- Koreandata$BF.ratio.corr/0.6048195
hist(Korean.norm.BF)
#HMP SD = 8.301864
keepHMP <- which(data2$dataset == "HMP")
hmpdata <- data2[keepHMP, ]
HMP.norm.BF <- hmpdata$BF.ratio.corr/8.301864
hist(HMP.norm.BF)
#CCHC SD = 2.00527
keepCCHC <- which(data2$dataset == "CCHC")
CCHCdata <- data2[keepCCHC, ]
CCHC.norm.BF <- CCHCdata$BF.ratio.corr/2.00527
hist(CCHC.norm.BF)
#GLNE07 SD = 0.4451074
keepGLNE <- which(data2$dataset == "GLNE")
GLNEdata <- data2[keepGLNE, ]
GLNE.norm.BF <- GLNEdata$BF.ratio.corr/0.4451074
hist(GLNE.norm.BF)
#Escobar SD = 0.4232106
keepescobar <- which(data2$dataset == "Escobar")
escobardata <- data2[keepescobar, ]
Escobar.norm.BF <- escobardata$BF.ratio.corr/0.4232106
hist(Escobar.norm.BF)
#COMBO SD = 5.928664
keepCOMBO <- which(data2$dataset == "COMBO")
COMBOdata <- data2[keepCOMBO, ]
COMBO.norm.BF <- COMBOdata$BF.ratio.corr/5.928664
hist(COMBO.norm.BF)

#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
BFScore <- c(Escobar.norm.BF, GLNE.norm.BF, CCHC.norm.BF, HMP.norm.BF, Korean.norm.BF, metahit.norm.BF, COMBO.norm.BF)
data2 <- cbind(BFScore, data2)

amish.data <- cbind(amish.norm.BF, amish.data)
turn.data <- cbind(turn.norm.BF, turn.data)
rm(Escobar.norm.BF, GLNE.norm.BF, CCHC.norm.BF, HMP.norm.BF, Korean.norm.BF, metahit.norm.BF, amish.norm.BF, turn.norm.BF, COMBO.norm.BF)
rm(keepmetahit, keepCCHC, keepescobar, keepGLNE, keepHMP, keepKorean)

d1.BF.corr <- data2$BFScore
t.BF.corr <- turn.data$turn.norm.BF
am.BF.corr <- amish.data$amish.norm.BF

comb.BF.corr <- as.data.frame(c(d1.BF.corr, t.BF.corr))
group.data.test <- cbind(comb.BF.corr, group.data.test1)
colnames(group.data.test)[1] <- "BF.ratio.corr"

comb2.BF.corr <- as.data.frame(c(d1.BF.corr, t.BF.corr, am.BF.corr))
group.data.test2 <- cbind(comb2.BF.corr, group.data.test2)
colnames(group.data.test2)[1] <- "BF.ratio.corr"

#Look at Obese versus non-obese
t.test(BF.ratio.corr ~ obese, data=group.data.test2) 
  # P-value = 0.05066
table(group.data.test2$obese)
  # No = 693 | Yes = 336
bf1 <- ggplot(group.data.test2, aes(obese, BF.ratio.corr)) + geom_boxplot()

#A more in-depth look, breaking down the groups
anova(lm(BF.ratio.corr ~ BMI.class, data=group.data.test)) 
  # P-value = 0.0227
a2 <- aov(BF.ratio.corr ~ BMI.class, data=group.data.test)
posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
  # Obese-Norm (p-value=0.026) | Over-Norm (p-value=0.117), Over-Obese (p-value=0.863)
table(group.data.test$BMI.class)
  # Lean = 326 | Overweight = 240 | Obese = 262
ddply(group.data.test, ~BMI.class, summarise, mean=mean(BF.ratio.corr), sd=sd(BF.ratio.corr))
  # Lean = -0.361 +\- 1.431 | Overweight = -0.695 +/- 2.032 | Obese = -0.787 +\- 2.461
bf2 <- ggplot(group.data.test, aes(BMI.class, BF.ratio.corr)) + geom_boxplot()

summary(lm(BFScore ~ bmi, data=data2)) #P-value 0.07859, R2= 0.004543
plot(BFScore ~ bmi, data=data2)
abline(lm(BFScore ~ bmi, data=data2), col="red")

#Make a better graph with ggplot2
b1 <- ggplot(data2, aes(x=bmi, y=BFScore))
b2 <- b1 + geom_point(aes(color=dataset))
b3 <- b2 + scale_colour_brewer(palette="Set1")
b4 <- b3 + geom_smooth(method="lm", se=TRUE)

##Run Zscore normalized Bacter and Firm overall analysis
#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
d1.bacter.corr <- data2$Zbacter
d1.firm.corr <- data2$Zfirm
d1.zbf.corr <- data2$ZBF
t.bacter.corr <- turn.data$Zbacter
t.firm.corr <- turn.data$Zfirm
t.zbf.corr <- turn.data$ZBF
am.bacter.corr <- amish.data$Zbacter
am.firm.corr <- amish.data$Zfirm
am.zbf.corr <- amish.data$ZBF

#Bacteroidetes Datasets
comb.bacter.corr <- as.data.frame(c(d1.bacter.corr, t.bacter.corr))
group.data.test <- cbind(comb.bacter.corr, group.data.test)
colnames(group.data.test)[1] <- "Zbacter"

comb2.bacter.corr <- as.data.frame(c(d1.bacter.corr, t.bacter.corr, am.bacter.corr))
group.data.test2 <- cbind(comb2.bacter.corr, group.data.test2)
colnames(group.data.test2)[1] <- "Zbacter"

#Firmicutes Datasets
comb.firm.corr <- as.data.frame(c(d1.firm.corr, t.firm.corr))
group.data.test <- cbind(comb.firm.corr, group.data.test)
colnames(group.data.test)[1] <- "Zfirm"

comb2.firm.corr <- as.data.frame(c(d1.firm.corr, t.firm.corr, am.firm.corr))
group.data.test2 <- cbind(comb2.firm.corr, group.data.test2)
colnames(group.data.test2)[1] <- "Zfirm"

#Bacteroidetes/Firmicutes Datasets
comb.zbf.corr <- as.data.frame(c(d1.zbf.corr, t.zbf.corr))
group.data.test <- cbind(comb.zbf.corr, group.data.test)
colnames(group.data.test)[1] <- "zbf"

comb2.zbf.corr <- as.data.frame(c(d1.zbf.corr, t.zbf.corr, am.zbf.corr))
group.data.test2 <- cbind(comb2.zbf.corr, group.data.test2)
colnames(group.data.test2)[1] <- "zbf"

###################################Bacteroidetes#######################################
#Look at Obese versus non-obese
t.test(Zbacter ~ obese, data=group.data.test2) 
  # P-value = 0.3161
table(group.data.test2$obese)
  # No = 693 | Yes = 336
bacter1 <- ggplot(group.data.test2, aes(obese, Zbacter)) + geom_boxplot()

#A more in-depth look, breaking down the groups
anova(lm(Zbacter ~ BMI.class, data=group.data.test)) #P-value = 0.0957
a2 <- aov(Zbacter ~ BMI.class, data=group.data.test)
posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
  # Obese-Norm (p-value=0.979) | Over-Norm (p-value=0.151) | Over-Obese (p-value=0.125)
table(group.data.test$BMI.class)
  # Lean = 326 | Overweight = 240 | Obese = 262
ddply(group.data.test, ~BMI.class, summarise, mean=mean(Zbacter), sd=sd(Zbacter))
  # Lean = 0.041 +/- 0.966 | Overweight = -0.117 +/- 1.000 | Obese = 0.057 +/- 1.02

bacter2 <- ggplot(group.data.test, aes(BMI.class, Zbacter)) + geom_boxplot()

summary(lm(Zbacter ~ bmi, data=data2)) #P-value 0.7168, R2= 0.0001936
plot(Zbacter ~ bmi, data=data2)
abline(lm(Zbacter ~ bmi, data=data2), col="red")

#Make a better graph with ggplot2
b1 <- ggplot(data2, aes(x=bmi, y=Zbacter))
b2 <- b1 + geom_point(aes(color=dataset))
b3 <- b2 + scale_colour_brewer(palette="Set1")
b4 <- b3 + geom_smooth(method="lm", se=TRUE)

###################################Firmicutes#######################################
#Look at Obese versus non-obese
t.test(Zfirm ~ obese, data=group.data.test2) #P-value = 0.8298
table(group.data.test2$obese)
  # No = 693 | Yes = 336
firm1 <- ggplot(group.data.test2, aes(obese, Zfirm)) + geom_boxplot()

#A more in-depth look, breaking down the groups
anova(lm(Zfirm ~ BMI.class, data=group.data.test)) 
  #P-value = 0.05521
a2 <- aov(Zfirm ~ BMI.class, data=group.data.test)
posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
  # Obese-Norm (p-value=0.976), Over-Norm (p-value=0.063), Over-Obese (p-value=0.1252)
table(group.data.test$BMI.class)
  # Lean = 326 | Overweight = 240 | Obese = 262
ddply(group.data.test, ~BMI.class, summarise, mean=mean(Zfirm), sd=sd(Zfirm))
  # Lean = -0.061 +/- 0.959 | Overweight = 0.130 +/- 1.018 | Obese = -0.043 +/- 1.013

firm2 <- ggplot(group.data.test, aes(BMI.class, Zfirm)) + geom_boxplot()

summary(lm(Zfirm ~ bmi, data=data2)) #P-value 0.8682, R2= 4.06x10-5
plot(Zfirm ~ bmi, data=data2)
abline(lm(Zfirm ~ bmi, data=data2), col="red")

#Make a better graph with ggplot2
b1 <- ggplot(data2, aes(x=bmi, y=Zfirm))
b2 <- b1 + geom_point(aes(color=dataset))
b3 <- b2 + scale_colour_brewer(palette="Set1")
b4 <- b3 + geom_smooth(method="lm", se=TRUE)

###################################Bacteroidetes/Firmicutes#######################################
#Look at Obese versus non-obese
t.test(zbf ~ obese, data=group.data.test2) #P-value = 0.4167
table(group.data.test2$obese)
  # No = 693 | Yes = 336
zbf1 <- ggplot(group.data.test2, aes(obese, zbf)) + geom_boxplot()

#A more in-depth look, breaking down the groups
anova(lm(zbf ~ BMI.class, data=group.data.test)) 
  # P-value = 0.4769
a2 <- aov(zbf ~ BMI.class, data=group.data.test)
posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
  # Obese-Norm (p-value=0.892) | Over-Norm (p-value=0.692) | Over-Obese (p-value=0.452)
table(group.data.test$BMI.class)
  # Lean = 326 | Overweight = 240 | Obese = 262
ddply(group.data.test, ~BMI.class, summarise, mean=mean(zbf), sd=sd(zbf))
  # Lean = 0.0081 +/- 0.954 | Overweight = -0.064 +/- 0.951 | Obese = 0.046 +/- 1.084
zbf2 <- ggplot(group.data.test, aes(BMI.class, zbf)) + geom_boxplot()

summary(lm(ZBF ~ bmi, data=data2)) #P-value 0.7118, R2= 0.0002008
plot(ZBF ~ bmi, data=data2)
abline(lm(ZBF ~ bmi, data=data2), col="red")

#Make a better graph with ggplot2
b1 <- ggplot(data2, aes(x=bmi, y=ZBF))
b2 <- b1 + geom_point(aes(color=dataset))
b3 <- b2 + scale_colour_brewer(palette="Set1")
b4 <- b3 + geom_smooth(method="lm", se=TRUE)

```

##### **References:**

### 2016-02-05 code/sequenceProcessing.Rmd
Initial commit

---
title: "Bacterial Microbiome and Obesity Meta-analysis Sequence Processing Pipeline"
author: "Marc Sze"
date: "January 26, 2016"
output: 
  html_document:
    keep_md: yes
bibliography: metaAnalysis.obesity.microbiome.bib
---
#### **Preamble**  
Each study is brought up one at a time.  This recount starts from time of download all the way to generation of the final shared file used for the meta-analysis in the study.  You will also need the latest version of mothur, SRA Toolkit, Metaphlan2, and Python 2.7 installed on your system.  There are also a few reference files that you will need.  The first is the [SILVA reference](http://www.mothur.org/wiki/Silva_reference_files) database and the second is the [RDP database](http://www.mothur.org/wiki/RDP_reference_files) of which we used release 102 and version 9 respectively.  Finally, any entered code is denoted will be surronded with ''.  As an example, ''print("Hello World)''.
<br>
<br>

##### **Baxter et al. 2016:**
This data set consisted of controls without colorectal cancer (CRC) and those with CRC.  The sequence reads of only the controls who did not have CRC were obtained directly from the first author since the data was generated within the Schloss lab.  However, the sequences can be found on the SRA under the accession number SRP062005. Only the raw 16S sequence reads for normal controls were processed  from the Sequence Read Archive (SRA) and subsequently used in the analysis.  So here we go....
<br>
<br>
I'm sure a smarter person could seperate the specific files that are normal only but I'm lazy and we had the storage space so I downloaded the entire data set.  
<br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP062/SRP062005/''
<br>
''fastq-dump \*.sra''
<br><br>
The metadata can be found and downloaded by searching SRP062005 at the following site: http://www.ncbi.nlm.nih.gov/Traces/study/
<br><br>
The simple python program (Python 3.0) used to generate the .file file needed to make contigs can be found in this repository titled: *MSze.makefile.good.trial2.py* and the resulting file is titled *mothur.data.file*.  All you need to make this work is to have only the necessary fastq files in the respective folder.
<br><br>
Before getting started enter mothur and curate the silva.fasta for the v4 region with the following commands:
<br><br>
''pcr.seqs(fasta=silva.bacteria.fasta, start=11894, end=25319, keepdots=F)''
<br>
''system(mv silva.bacteria.pcr.fasta silva.v4.fasta)''
<br><br>

Next we utilized the batch mode process and can come back the next day to see everything completed.  Our command was the following:
<br><br>
''/share/scratch/schloss/mothur/mothur data1.batch''
<br>
<br>
If you thought this was difficult to get data just wait until later on in this pipeline to see just how crazy it can get trying to get the data to a useable point.
<br>
<br>

##### **Ross et al. 2015:**
This data set consisted of normal individuals from the Cameron Count Hispanic Cohort (CCHC).  The paper for this data set can be found [here](http://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-015-0072-y).  The project accession number in the SRA is SRP053023.  We can download the entire data set again with the code mentioned above but with one small substitution at the ending.  Here we go......
<br><br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP053/SRP053023/''
<br>
<br>
Now since there was no sff file provided in this data set even though it is a 454 data set (I found this out when I tried to run the command -> sff-dump \*.sra).  We have to use the fastq function of the SRA Tool kit.
<br>
<br>
''fastq-dump \*.sra''
<br>
<br>
Because of this difference there are a few changes in this processing versus the standardized mothur 454 protocol.  First, instead of the sff.info() command we have to use the fastq.info() command.  Second, and perhaps more importantly, because we don't have the sff file we have to use the quality scores.  We used the author specified defaults of *qwindowaverage=25* and *qwindowsize=50*.  It should also be noted that we had *maxambig=5* and *maxhomop=8*.  Since the data come seperated already we do not need an oligo file and all we have to do is loop the mothur *trim.seqs* command. You can run a bash or shell program to get all of this done with either *#!/bin/sh* or *#!/bin/bash*.  So in a text editor:

''#!/bin/sh  
for sample in \$(ls \*sra)  
do  
fastq-dump \$sample  
done
<br><br>
for sample2 in \$(ls \*fastq)  
do  
/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=\$sample2)"  
rm \*logfile  
done
<br><br>
for TRIM in \$(ls ERR\*.fasta)  
do  
QFILE=\${TRIM//fasta/qual}  
/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIM, qfile=\$QFILE, maxambig=5, maxhomop=8, qwindowaverage=25, qwindowsize=50, processors=8)"  
done  
rm \*.fastq''
<br><br>
The next lines of code are used to combine the fasta file, create a names file with the mothur unique command, and create a combined group file for subsequent processing.
<br><br>
''cat \*trim.fasta > combined.trim.fasta  
/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"  
grep -e '^>' combined.trim.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/\\./\_/g' test2.fasta > test3.fasta  
sed 's/_.\*//' test3.fasta > test4.fasta  
paste -d" " test2.fasta test4.fasta > combined.groups  
exit 0''
<br><br>
After this little program is run we follow the mothur 454 SOP as written and can use the following command to complete the processing using the put together *Ross.batch* file and running mothur.  One final difference from the standardized mothur approach that was made was during the screen.seqs command after alignment a hard cutoff *minlength=200* and *maxlength=250* was used.
<br><br>
''/share/scratch/schloss/mothur/mothur Ross.batch''
<br><br>
That completes the processing for this data set.  Simple so far I hope.  On to the next data set.
<br><br>

##### **Goodrich et al. 2014:**  
This is a second larger twin data set that builds on the Turnbaugh et al. findings.  The paper can be found [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4255478/).  The individuals that participated in the study were part of the TwinsUK registry and is somewhat heavily skewed toward women.  The raw sequence data can be found on the SRA under ERP006339.  So using wget we can download the data with the following code:
<br><br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/ERP/ERP006/ERP006339/''
<br><br>
 I started off with analyzing the data separately while sticking the the default mothur SOP for MiSeq processing.....
<br><br>
''#!/bin/sh  
\# Unpack sra files  
\# mkdir sra.complete  
for start in \$(ls \*sra)  
do  
fastq-dump \$start  
mv \$start sra.complete  
done
<br><br>
\# Unpack fastq files  
mkdir fastq.complete  
mkdir logfile.complete  
for info in \$(ls \*fastq)  
do  
/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=\$info)"  
mv \$info fastq.complete  
mv \*.logfile logfile.complete  
done
<br><br>
\# Screen the fasta file that came from the fastq  
mkdir fasta.complete  
for next in \$(ls \*fasta)  
do  
/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$next, maxambig=0, maxlength=275, processors=8)"  
mv \$next fasta.complete \# should store silva file in here and then move it working directory  
mv \*.logfile logfile.complete  
done
<br><br>
mv ~/twinsUK/fasta.complete/silva.v4.fasta ~/twinsUK  
\# Align the screened fasta file to V4 region of SILVA  
mkdir good.fasta.complete  
for sample2 in \$(ls \*good.fasta)  
do  
/share/scratch/schloss/mothur/mothur "#align.seqs(fasta=\$sample2, reference=silva.v4.fasta, processors=8)"  
mv \$sample2 good.fasta.complete  
mv \*logfile logfile.complete  
done
<br><br>
\#Screen the aligned sequences to specific location  
mkdir align.complete  
for sample3 in \$(ls \*.align)  
do  
/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$sample3, start=1968, end=11546, maxhomop=8, processors=8)"  
mv \$sample3 align.complete  
done
<br><br>
\# Filter the screened sequences to get a fasta file back  
mkdir align.good.complete  
for sample4 in \$(ls \*good.align)  
do  
/share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=\$sample4, vertical=T, trump=., processors=8)"  
mv \$sample4 align.good.complete  
mv \*logfile logfile.complete  
done
<br><br>
\# Make the combined fasta file  
mkdir filter.fasta.complete  
cat \*filter.fasta > combined.fasta  
mv \*filter.fasta filter.fasta.complete
<br><br>
\# Make the combined group file  
grep -e '^>' combined.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/\\..\*\\$//' test2.fasta > test3.fasta  
paste -d"\\t" test2.fasta test3.fasta > combined.groups''
<br><br>
After this program was run I thought perhaps naively that I could simply run a set a batch file and be done with this data set.  Well not quite.  The sheer size and the way our server was configured meant that if I called much more than 130GB of RAM my job would run the risk of starting over regularly. So I did use the batch method but only upto *pre.cluster*.  After this the *chimera.uchime* step was just taking too long so I seperated everything again and did the *chimera.uchime* separately.  This also meant having to redo the unique seqs step.  The code for this is below:
<br><br>
''/share/scratch/schloss/mothur/mothur Goodrich.batch''
<br><br>
So now the biggest challenge that I have found so far is getting the data into a distance matrix not based on the closed-reference approach.  The file is just too big for cluster.split so I'm going to have to find an alternative way.

<br><br>

##### **Escobar et al. 2014:**
This data set consisted of normal individuals from Medellin, Colombia South America.  The paper for this data set can be found [here](http://bmcmicrobiol.biomedcentral.com/articles/10.1186/s12866-014-0311-6).  The project accession number in the ENA is ERP003466.  We can download the entire data set again with the code mentioned above but with a few substitutions versus the previous two studies.  Here we go......
<br><br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/ERP/ERP003/ERP003466/''
<br>
<br>
Now just like the previous data set there was no sff file provided in this data set even though it is a 454 data set (I found this out, again, when I tried to run the command -> sff-dump \*.sra).  We have to use the fastq function of the SRA Tool kit.
<br>
<br>
''fastq-dump \*.sra''
<br>
<br>
Because of this difference there are a few changes in this processing versus the standardized mothur 454 protocol.  First, instead of the sff.info() command we have to use the fastq.info() command.  Second, and perhaps more importantly, because we don't have the sff file we have to use the quality scores.  Unlike the Ross, et al study we used the mothur specified defaults of *qwindowaverage=35* and *qwindowsize=50*.  It should also be noted that we had *maxambig=0* and *maxhomop=8*.  Since the data come seperated already we do not need an oligo file and all we have to do is loop the mothur *trim.seqs* command. You can run a bash or shell program to get all of this done with either *#!/bin/sh* or *#!/bin/bash*.  So in a text editor:

''#!/bin/sh  
for sample in \$(ls \*sra)  
do  
fastq-dump \$sample  
done
<br><br>
for sample2 in \$(ls \*fastq)  
do  
/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=\$sample2)"  
rm \*logfile  
done
<br><br>
for TRIM in \$(ls SRR\*.fasta)  
do  
QFILE=\${TRIM//fasta/qual}  
/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIM, qfile=\$QFILE, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50, processors=8)"  
done  
rm \*.fastq''
<br><br>
The next lines of code are used to combine the fasta file, create a names file with the mothur unique command, and create a combined group file for subsequent processing.
<br><br>
''cat \*trim.fasta > combined.trim.fasta  
/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"  
grep -e '^>' combined.trim.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/\\./\_/g' test2.fasta > test3.fasta  
sed 's/_.\*//' test3.fasta > test4.fasta  
paste -d" " test2.fasta test4.fasta > combined.groups  
exit 0''
<br><br>
After this little program is run we follow the mothur 454 SOP as written and can use the following command to complete the processing using the put together *Escobar.batch* file and running mothur.  Again like the Ross et al data set one final difference from the standardized mothur approach that was made was during the screen.seqs command after alignment a hard cutoff *minlength=200* and *maxlength=250* was used.
<br><br>
''/share/scratch/schloss/mothur/mothur Escobar.batch''
<br><br>
That completes the processing for this data set.  So far so good, no major problems.  On to the next data set.
<br><br>

##### **Zupancic et al. 2012:**
This data set consisted of normal individuals from an Amish community living in Lancaster County, Pennsylvania.  The paper for this data set can be found [here](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.00430526).  The project accession number in the SRA is SRP002465.  We can download the entire data set again with the code mentioned above but with a few substitutions.  Here we go......
<br><br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP002/SRP002465/''
<br><br>
Now this data set included the sff files so we can use sffinfo function and start from the very start of the mothur 454 pipleine.  Plus one for this data set so far.  The below program was used and you can use either *#!/bin/sh* or *#!/bin/bash* and text editor to create it and then run it.  Also you should be aware that this part of the processing will take a LONG time to get through.
<br><br>
''#!/bin/sh
for sample in \$(ls \*sra)  
do  
sff-dump \$sample  
done
<br><br>
for sample2 in \$(ls \*sff)  
do  
/share/scratch/schloss/mothur/mothur "#sffinfo(sff=\$sample2, flow=T)"  
rm \*logfile  
done
<br><br>
for FLOW in \$(ls \*flow)  
do  
/share/scratch/schloss/mothur/mothur "#trim.flows(flow=\$FLOW, processors=8)"  
rm \*logfile  
done
<br><br>
for SHHH in \$(ls \*files)  
do  
/share/scratch/schloss/mothur/mothur "#shhh.flows(file=\$SHHH, processors=8)"  
rm \*logfile  
done
<br><br>
for TRIMS in \$(ls \*shhh.fasta)  
do  
NAMES=\${TRIMS//fasta/names}  
/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIMS, name=\$NAMES, maxhomop=8, maxambig=0, minlength=200, processors=8)"  
rm \*logfile  
done  
exit 0''
<br><br>
Now, I had a few extra lines that combined the finalized files together but when I went to analyze these they gave some weird errors.  So I needed to make a second program to sort through these errors and analyze the data separately still.  One thing I noticed was that by using the standard workflow a large number of files ended up with 0 or very few sequences passing the quality filters.  The other problem which I delayed dealing with in this data set was that many samples were done sequenced more than once. This in itself is not a problem but the sequence ID associated with these sequences from different files were the exact same and so were carbon copies.  This wreaked havoc when I tried to combined the data together.  But I'll deal with that problem later.  I'm just going to continue on with the pipeline with the file seperated.  I will however, have to deal with this before the preclustering step.
<br><br>
So here is the second program used to keep processing the sequences separately rather than as one combined fasta file.....
<br><br>
''#!/bin/sh  
\#Align the fasta file to the SILVA database and create directories to move files into    
mkdir good.fasta.complete  
mkdir reference.files  
mkdir logifle.complete  
mv silva* reference files  
mv trainset9* reference files  
for sample2 in \$(ls \*.fasta)  
do  
/share/scratch/schloss/mothur/mothur "#align.seqs(fasta=\$sample2, reference=reference.files/silva.bacteria.fasta, flip=t, processors=8)"  
mv \$sample2 good.fasta.complete  
mv \*logfile logfile.complete  
done
<br><br>
\#Screen the align sequences  
mkdir align.complete  
for sample3 in \$(ls \*.align)  
do  
/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$sample3, end=6453, optimize=start, criteria=95, processors=8)"  
mv \$sample3 align.complete
mv \*logfile logfile.complete  
done
<br><br>
\#filter the screen sequences  
mkdir align.good.complete  
for sample4 in \$(ls \*good.align)  
do  
/share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=\$sample4, vertical=T, trump=., processors=8)"  
mv \$sample4 align.good.complete  
mv \*logfile logfile.complete  
done  
exit 0''
<br><br>
Right, now I need to pre cluster the data and I can't get away with analyzing each file separately anymore.  However, remember from up above there is something wierd going on in that the unique sequence ID are duplicated for many samples and there were a whole bunch of samples that did not survive quality filtering.  Since some samples had two samples others had 6 samples to be fair I decided to take the first sample from an individual that had sequences in them (so greater than 0 MB).  Now, this was done when I was still truly a novice in using shell scripting so I did this the "hard" way (i.e. the following method is pretty manual with very little automation so someone can make this more automated to improve on this process).  Essentially if a sample had sequence in it the indivdiual study ID was assigned to the fasta file and stored in a new directory.  Those that either had 0 MB after this we carried over and the next sequence ID file was used.  This was done until all samples had an indivdiual ID.  In my novice approach the only thing I automated was the original copying of the sequence ID file to a individual study ID file.  I then visually checked the size and manually sorted the files into directories.  From this I created a new list to run through.  The lists that were used were *combined2*, *combined3*, *combined4*, *combined5*, *combined6*, *combined7*, and *combined8* and can be found in this same repository.
<br><br>
As an aside I have learned a lot about coding doing this project and if I were to do this analysis again for this study I would automate this by first generated the list of sequence IDs for each respective Indivdiual and then adding an argument that if the file size is larger than 0 MB to stop and move it.  If it was 0 MB the program would move on to the next sequence ID file for that Indivdiual.
<br><br>
Now that this little problem is out of the way and I have been perfectly upfront with how I dealt with this lets move on to the final part of the processing pipeline for this data set.
<br><br>
For this step I took the individual sample ID named fasta files and created a list which I looped with a shell script to create a group file which can then be combined along with the fasta file.  The code I used for this was:
<br><br>
''#!/bin/sh  
mkdir trim.fasta.complete2  
\#Make loop to create a group file  
groupN="106 107 108 113 116 11 126 128 12 13 143 14 150 156 157 160 162 163 169 172 177 181 18 190 191 192 19 1 201 202 212 214 216 217 219 225 226 230 242 24 269 26 275 283 287 289 291 292 297 299 29 2 313 323 327 333 33 346 347 349 351 352 354 358 359 361 367 373 376 388 38 392 397 398 401 406 410 411 415 416 421 423 424 427 429 444 44 453 455 459 463 464 467 46 471 473 474 475 476 481 482 483 484 488 48 49 50 51 523 52 53 54 59 66 68 6 74 77 81 84 87 92 99 p0 p105 p108 p111 p112 p113 p114 p115 p116 p117 p118 p119 p120 p121 p122 p123 p124 p125 p126 p127 p130 p132 p133 p134 p135 p136 p137 p138 p139 p13 p140 p142 p145 p148 p14 p150 p151 p152 p153 p154 p155 p156 p158 p159 p15 p160 p16 p17 p22 p23 p24 p25 p26 p33 p34 p36 p37 p38 p3 p41 p42 p43 p44 p46 p49 p4 p57 p58 p63 p68 p69 p71 p72 p73 p74 p77 p79 p7 p81 p82 p85 p86 p87 p8 p90 p91 p95 p98"  
for V in \${groupN}  
do  
grep -e '^>' \${V}.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed "s/$/\\t\$V/" test2.fasta > \${V}.groups  
mv \${V}.fasta trim.fasta.complete2  
rm test.fasta  
rm test2.fasta  
done
<br><br>
cd trim.fasta.complete2  
cat \*fasta > combined.fasta  
cat \*groups > combined.groups  
/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.fasta)"''
<br><br>
Now that this has been completed we can create a batch file for the remaining commands and analyze the data altogether.  One final comment is that for the *screen.seq* command I did not set a *maxhomop=8*.  Due to the large loss (which you will see from the subsampling) from this data set I did not want to be picky.
<br><br>
''/share/scratch/schloss/mothur/mothur Zupancic.batch''
<br><br>
With that you should now be able to generate the subsample shared file that was used for the downstream analysis.  This was a headache of a data set to work with.  I think the next one will be better. At least I hope so....
<br><br>

##### **Yatsunenko et al. 2012:**  
This data set was another famous one looking at the microbiome of indivdiuals at different geographical locations and pretty different diets.  The link to the paper is [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3376388/).  A couple problems from a data processing and sequence clean up point of view is that they used a HiSeq, single end, 100bp total length, sequencing run.  This wreaked absolute havoc with a lot of the downstream steps mostly with trying to figure out a way to reduce the artificial diversity within the files.  Oh... and the data was not on a public database but stored on MG-RAST.  I had a lot trouble downloading the data on command line all at once (maybe I'm just not smart enough to do it).  So I had to manually download the sequence files.Thats right I am just a glutton for punishment.    After you have downloaded the files the first part of the code is:
<br><br>
''#!/bin/sh  
mkdir fna.done  
mkdir logfiles.done  
mkdir bad.accnos.done
<br><br>
mkdir good.fna.done  
for sample in \$(ls \*.fna)  
do  
/share/scratch/schloss/mothur/mothur "#align.seqs(fasta=\$sample, reference=silva.bacteria.fasta, processors=8)"  
mv \$sample good.fna.done  
mv \*logfile logfiles.done  
done''
<br><br>
For the next code it might be best to just double check that the start and end are what is written here.  To make it go faster I made a custom pcr.seqs based on the start and end of the sequences on the aligned file.  It shouldn't make a difference but.....you never know.
<br><br>
''#!/bin/sh  
mkdir align.done  
for sample2 in \$(ls \*.align)  
do  
/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$sample2, start=8, end=2448, maxhomop=8, processors=8)"  
mv \$sample2 align.done  
mv \*logfile logfiles.done  
done
<br><br>
mkdir good.align.done  
for sample3 in \$(ls \*good.align)  
do  
/share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=\$sample3, vertical=T, trump=., processors=8)"  
mv \$sample3 good.align.done  
mv \*logfile logfiles.done  
done
<br><br>
mkdir completed.good.groups  
for sample in \$(ls \*.good.filter.fasta)  
do  
grep -e '^>' \$sample > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/_.\*//' test2.fasta > test3.fasta  
paste -d"\\t" test2.fasta test3.fasta > \${sample}.groups  
rm test.fasta  
rm test2.fasta  
rm test3.fasta  
mv ~/Yatsunenko.Nature/${sample}.groups ~/Yatsunenko.Nature/completed.good.groups   
done
<br><br>
From here I thought I might be able to get away with combining the data set and running everything together.  Well that was not the case since it is just too large for the memory I have readily available without having the job restarting.  The plan is to seperate everything back out again and run the *chimera.uchime* function.  Then combine them and run the *pre.cluster* function.  The code below shows what was used for this part:
<br><br>
''#!/bin/sh  
cat *fasta > combined.fasta  
cat *groups > combined.groups  
/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.fasta)"    
/share/scratch/schloss/mothur/mothur Yatsunenko.batch  
done
<br><br>

<br><br>

##### **Human Microbiome Project (HMP) 2011:**  
This data set consisted of healthy individuals that were sampled at multiple body sites and underwent extensive medical work up for metadata collection.  In order to be included in this study you could not have any underlying medical conditions.  The project website can be found [here](http://hmpdacc.org/).  The project accession number in the SRA is SRP002860.  The sequence processing pipeline can be found in an ipython notebook located [here](http://nbviewer.jupyter.org/gist/pschloss/9815766/notebook.ipynb).  The one exception to the workflow provided in the notebook is that you will need to make sure that you do not use a closed reference data base approach for OTU clustering and instead use an open reference approach.
<br><br>

##### **Nam, et al. 2011:**  
This study recruited 20 individuals that were from the area surronding Kyung Hee University in South Korea and can be found [here](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0022109).  The data from this project can be found under the accession number PRJDA6057 and we downloaded the data from DRR000776.  There is some small difference to the wget, which is mostly associated with the directory where the data is stored.  From here you can run the SRA tool kit command to get the sff file.  Once this is done you can then run mothur with the batch file and get the sub sample shared file that was used in this analysis.  We have also provided the oligo file in this repository *korean6.oligos* that was used so that you can easily use it on this data set.
<br><br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByRun/sra/DRR/DRR000/DRR000776
<br><br>
sff-dump DRR000776.sra
<br><br>
/share/scratch/schloss/mothur/mothur Nam.batch''
<br><br>
Compared to the previous data sets this was one of the most straight forward to actually process.
<br><br>

##### **Wu, et al. 2011:**  
This study was published in Science using the COMBO data set and can be found [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3368382/).  The investigators were mostly interested in investigating dietary patterns its influence of gur microbial enterotypes. The data for this project can be found on the SRA website under the Accession number SRP002424.  So with a little modification we can use wget again to get the data.
<br><br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP002/SRP002424/''
<br><br>
From here there was a program that was run to loop through a few of the mothur functions on each individual file.  Since they are already seperated by indivdiual we don't need the oligo file.  So here we go......
<br><br>
''#!/bin/sh  
mkdir sra.complete  
\#run first loop which will extract sff files  
for sample in \$(ls \*sra)  
do  
sff-dump \$sample  
mv \$sample sra.complete
<br><br>
\#create new directories for logfiles and sff's that are complete.  
mkdir logfile.complete  
mkdir sff.complete  
\#Extract info from sff files  
for sample2 in \$(ls \*sff)  
do  
/share/scratch/schloss/mothur/mothur "#sffinfo(sff=\$sample2, flow=T)"  
mv \$sample2 sff.complete  
mv \*logfile logfile.complete  
done
<br><br>
mkdir flow.complete  
\#Trim respective flow files  
for FLOW in $(ls \*flow)  
do  
/share/scratch/schloss/mothur/mothur "#trim.flows(flow=\$FLOW, processors=8)"  
mv $FLOW flow.complete  
mv \*logfile logfile.complete  
done
<br><br>
mkdir file.complete  
\#Make loop to run the shhh.flow and move completed sample to another directory  
for SHHH in \$(ls \*files)  
do  
/share/scratch/schloss/mothur/mothur "#shhh.flows(file=\$SHHH, processors=8)"  
mv \$SHHH file.complete  
mv \*logfile logfile.complete  
done
<br><br>
mkdir shhh.fasta.complete  
for TRIMS in \$(ls \*shhh.fasta)  
do  
NAMES=\${TRIMS//fasta/names}  
/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIMS, name=\$NAMES, maxhomop=8, minlength=200, processors=8)"  
mv \$TRIMS shhh.fasta.complete  
mv \*logfile logfile.complete''  
<br><br>
So for the next part I needed to make a group file so I renamed the fasta file to represent the sampleID label so that I could with more ease create the labels I needed within the group file. The code for it is....
<br><br>
''samples="SRR327637 SRR327638 SRR327639 SRR327640 SRR327641 SRR327642 SRR327643 SRR327644 SRR327645 SRR327646 SRR327647 SRR327648 SRR327649 SRR327650 SRR327651 SRR327652 SRR327653 SRR327654 SRR327655 SRR327656 SRR327657 SRR327658 SRR327659 SRR327660 SRR327661 SRR327662 SRR327663 SRR327664 SRR327665 SRR327666 SRR327667 SRR327668 SRR327669 SRR327670 SRR327671 SRR327672 SRR327673 SRR327674 SRR327675 SRR327676 SRR327677 SRR327678 SRR327679 SRR327680 SRR327681 SRR327682 SRR327683 SRR327684 SRR327685 SRR327686 SRR327687 SRR327688 SRR327689 SRR327690 SRR327691 SRR327692 SRR327693 SRR327694 SRR327695 SRR327696 SRR327697 SRR327698 SRR327699 SRR327700 SRR327701"
<br><br>
for V in \${samples}  
do  
grep -e '^>' \${V}.shhh.trim.good.filter.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed "s/\$/\\t\$V/" test2.fasta > \${V}.groups  
mv \${V}.shhh.trim.good.filter.fasta filter.fasta.complete  
rm test.fasta  
rm test2.fasta  
echo \$V  
done''
<br><br>
From here you can go to the individual folders and find the final version of the respective fasta, names, and groups files for every indivdiual and combine them into a single larger file.  In essence you don't need the names file since mothur will re-generate one when you run the *unique.seqs* command.  This was the route that I opted for since it meant slightly less typing.  So the only files that I needed to combine are the fasta and groups files.
<br><br>
''cat \*trim.good.filter.fasta > combined.fasta''  
''cat \*groups > combined.groups''
<br><br>
After you have done this you can run a batch file with mothur and that should generate the subsample shared file that we used for this anlaysis.  We lose a large number of samples like in the Zupancic data set since we are subsampling to 1000.  We choose 1000 since it is an arbitrary standard many use but also because we wanted to set a standardized lowest vaule to subsample for that we can hold every data set to.  Anyways by running the final command you should get the files needed to all the downstream analysis.
<br><br>
''/share/scratch/schloss/mothur/mothur Wu.batch''
<br><br>

##### **Arumugam, et al. 2010 (MetaHit):**  
We only used the Danish individuals for this study and left the others alone since many had IBD or other disease conditions.  For this data set Metaphlan2 was used to generate the abundance data.  Metaphlan2 and how well it associates with 16S data is also discussed in the supplemental of the draft manuscript we are writing.  To test this we downloaded ~40 samples from the HMP data set which had both metagenome and 16S data.  We ran the same code that we used with the MetaHit data to generate the abundance data using Metaphlan2 and compared that result with the obtained 16S values.  So here is code that was used below:
<br><br>
''#!/bin/sh  
PATH=\$PATH:~/metaphlan2  
export mpa_dir=~/metaphlan2  
mkdir completed.combined.fq  
test_samples="MH0001 MH0002 MH0003 MH0004 MH0005 MH0006 MH0007 MH0008 
MH0009 MH0010 MH0011 MH0012 MH0013 MH0014 MH0015 MH0016 MH0017 MH0018 
MH0019 MH0020 MH0021 MH0022 MH0023 MH0024 MH0025 MH0026 MH0027 MH0028 
MH0030 MH0031 MH0032 MH0033 MH0034 MH0035 MH0036 MH0037 MH0038 MH0039 
MH0040 MH0041 MH0042 MH0043 MH0044 MH0045 MH0046 MH0047 MH0048 MH0049 
MH0050 MH0051 MH0052 MH0053 MH0054 MH0055 MH0056 MH0057 MH0058 MH0059 
MH0060 MH0061 MH0062 MH0063 MH0064 MH0065 MH0066 MH0067 MH0068 MH0069 
MH0070 MH0071 MH0072 MH0073 MH0074 MH0075 MH0076 MH0077 MH0078 MH0079 
MH0080 MH0081 MH0082 MH0083 MH0084 MH0085 MH0086"
<br><br>
for s in \${test_samples}  
do  
cat \${s}\*.1.fq > \${s}.combined.1.fq  
cat \${s}\*.2.fq > \${s}.combined.2.fq  
mv \${s}\*combined.1.fq completed.combined.fq  
mv \${s}\*combined.2.fq completed.combined.fq  
done''
<br><br>
This code was used to combine the fastq files since there are multiple files per sample.  To actually implement the metaphlan2 program for these files see below:
<br><br>
''for s in \${test_samples}  
do  
metaphlan2.py /mnt/EXT/Schloss-data/msze/test/danish.obesity.metagenome/raw.data.sep/completed.combined.fq/\${s}.combined.1.fq,/mnt/EXT/Schloss-data/msze/test/danish.obesity.metagenome/raw.data.sep/completed.combined.fq/\${s}.combined.2.fq --mpa_pkl db_v20/mpa_v20_m200.pkl --bowtie2db db_v20/mpa_v20_m200 --input_type fastq --bowtie2out profiled_samples/\${s}.bowtie2.bz2 -o profiled_samples/\${s}.txt  
done''
<br><br>
In order to combine all the files into a single data table the following code was run:
<br><br>
''python ~/metaphlan2/utils/merge\_metaphlan\_tables.py profiled\_samples/\*.txt > profiled\_samples/output/merged\_abundance\_table.txt''
<br><br>
For the 16S and metaphlan2 comparison the following code was run to obtain the necessary data tables:
<br><br>
''#!/bin/sh  
PATH=\$PATH:~/metaphlan2  
export mpa_dir=~/metaphlan2
<br><br>
test_samples="SRS012849 SRS012969 SRS013098 SRS013216 SRS013638 SRS013639 SRS016437 SRS016438 SRS016990 SRS017916 SRS020622 SRS021153 SRS021219 SRS022093 SRS024663 SRS043667 SRS045528 SRS045739 SRS049402 SRS049896 SRS050026 SRS053356 SRS053573 SRS053649 SRS054352 SRS055533 SRS056273 SRS058070 SRS063489 SRS063518 SRS064973 SRS074670 SRS074964 SRS075078 SRS075341 SRS075963 SRS076929 SRS077086 SRS077194 SRS077294 SRS077335 SRS077502 SRS077552 SRS077753 SRS077849 SRS078242 SRS078419 SRS078665 SRS097889 SRS098514 SRS098571 SRS098644 SRS098717 SRS098827 SRS0100021 SRS101376 SRS101433 SRS103987 SRS0104197 SRS104311 SRS104400 SRS104485 SRS105153 SRS140492 SRS140513 SRS140645 SRS142503 SRS142505 SRS142599 SRS142712 SRS142890 SRS143070 SRS143181 SRS143342 SRS143417 SRS143598 SRS143780 SRS143876 SRS143991 SRS144362 SRS144506 SRS144537 SRS145497 SRS146764 SRS146812 SRS146813 SRS147022 SRS147088 SRS147139 SRS147346 SRS147445 SRS147652 SRS147766 SRS147919 SRS148196 SRS148424 SRS148721" <br><br>
for s in \${test_samples}  
do  
\#Download and unpack necessary files  
wget -r -np -nd -k ftp://public-ftp.hmpdacc.org/Illumina/PHASEII/stool/${s}.tar.bz2  
tar xvjf \${s}.tar.bz2  
mv \${s}/\${s}.denovo\_duplicates\_marked.trimmed.1.fastq ~/metaphlan2  
mv \${s}/\${s}.denovo\_duplicates\_marked.trimmed.2.fastq ~/metaphlan2  
metaphlan2.py \${s}.denovo\_duplicates\_marked.trimmed.1.fastq,\${s}.denovo\_\duplicates_\marked.trimmed.2.fastq --mpa\_pkl db\_v20/mpa\_v20\_m200.pkl --bowtie2db db\_v20/mpa\_v20\_m200 --input\_type fastq --bowtie2out profiled\_samples/\${s}.bowtie2.bz2 -o profiled\_samples/\${s}.txt  
rm -r \${s}  
rm \${s}.denovo_duplicates_marked.trimmed.1.fastq  
rm \${s}.denovo_duplicates_marked.trimmed.2.fastq  
done''
<br><br>
You can then go into the directory with all the seperate data tables and combined them into one with the following command:
<br><br>
''python ~/metaphlan2/utils/merge\_metaphlan\_tables.py profiled\_samples/\*.txt > profiled\_samples/output/merged\_abundance\_table.txt''
<br><br>
Now I decided, perhaps foolishly, to manually curate the combined file.  What I mean by this is that I manually deleted viruses and eukaryotic cells that were identified instead of using a computer to do this.  I also then manually created data tables with phyla and the lowest taxonomic identification.  The data table with the lowest taxonomic identification information was used to generate alpha diversity measures and to complete and PERMANOVA analysis that was used.  These manually curated files are provided in this repository for you to use as you want.  Finally, I used relative abundance instead of subsampling in these samples.  So I had to redo the relative abundance after removing the viruses and eukaryotic cells.  This part was done in R.
<br><br>

##### **Turnbaugh, et al. 2008:**  
This is the very famous twin study of the microbiome and obesity.  The other two re-analysis articles [@finucane_taxonomic_2014; @walters_meta-analyses_2014] also use this data set.  The raw data is not avialable on the SRA but you can download a fasta file for either the V2 or V6 16S rRNA gene [here](https://gordonlab.wustl.edu/NatureTwins_2008/TurnbaughNature_11_30_08.html).  We used the V2 region.  For the most part it was relatively straight forward, just be aware that this took a lot of RAM to create a distance matrix.  The code I used is below:
<br><br>
''grep -e '^>' combined.trim.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta    
sed 's/\\s.\*\$//' test2.fasta > test3.fasta  
paste -d" " test2.fasta test3.fasta > V2.groups''
<br><br>
Once we have a groups file we can just create a batch file and put it into mothur to run and come back when it is completed.
<br><br>
''/share/scratch/schloss/mothur/mothur Turnbaugh.batch''
<br><br>

##### **References:**

### 2016-02-05 code/Turnbaugh.Feb32016.data.analysis.R
Turnbaugh R file

@@ -1,0 +1,205 @@
+##Turnbaugh Twinset data
+# Feb 3rd, 2016
+
+#Put into desired format for StatisticsProcessing.Rmd
+
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+setwd("C:/users/marc/Desktop/obesity2/turnbaugh.twins")
+shared.data <- read.table("test.unique.good.filter.unique.precluster.pick.pick.an.shared", header=T)
+rownames(shared.data) <- shared.data[, 2]
+shared.data <- shared.data[, -c(1:3)]
+
+subsample.data <- read.table("test.unique.good.filter.unique.precluster.pick.pick.an.0.03.subsample.shared", header=T)
+rownames(subsample.data) <- subsample.data[, 2]
+subsample.data <- subsample.data[, -c(1:3)]
+
+
+metadata <- read.csv("turnbaugh.metadata.csv")
+rownames(metadata) <- metadata[, 4]
+metadata <- metadata[, -4]
+
+keep1 <- which(metadata$Sample == 1) # n =146
+
+#use the first sampling
+#subset data for only the first sampling
+s1.metadata <- metadata[keep1, ]
+s1.subsample.data <- subsample.data[keep1, ]
+
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(s1.subsample.data)
+S <- specnumber(s1.subsample.data)
+J <- H/log(S)
+select.alpha.diversity <- as.data.frame(cbind(H, S, J))
+s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
+alpha.test <- s1.alpha.diversity
+
+#Generate phyla table data
+
+phyla.info <- read.csv("phyla.csv")
+phyla.table <- shared.data
+colnames(phyla.table) <- phyla.info[, 2]
+
+
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(4:5, 7:9)]
+
+#Create a relative abundance table for phyla
[…]

### 2016-02-05 code/Feb052016.good.COMBO.obesity.R
Individual R scripts

@@ -1,0 +1,257 @@
+##COMBO Obesity data analysis
+##Feb 5, 2016
+
+setwd("C:/users/marc/Desktop/obesity2/COMBO/")
+
+microbiome <- read.table("combined.good.subsample.shared", header=T)
+rownames(microbiome) <- microbiome$Group
+microbiome <- microbiome[, -c(1:3)]
+seqData <- read.csv("COMBO_data_table.csv", header=T)
+seqData <- seqData[-1, ]
+rownames(seqData) <- seqData$Run_s
+metadata <- read.table("BMI_and_Counts.txt", header=T)
+metadata <- metadata[, c(1, 2)]
+
+#get seqData set in line with microbiome data
+namesToKeep <- rownames(microbiome)
+test <- seqData[namesToKeep, ]
+seqData <- test
+rm(test)
+
+#change all rownames to match sample IDs
+rownames(seqData) <- seqData$submitted_subject_id_s
+rownames(microbiome) <- rownames(seqData)
+rownames(metadata) <- metadata[, 1]
+
+#Match the metadata now with the microbiome data
+namesToKeep <- rownames(microbiome)
+test <- metadata[namesToKeep, ]
+metadata <- test
+rm(test)
+rm(seqData, keep, namesToKeep)
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(microbiome)
+S <- specnumber(microbiome)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.table("combined.good.phyla.txt", header=T)
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(microbiome)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- microbiome
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
[…]

### 2016-02-05 code/Feb12016.Data.Analysis.R

@@ -1,0 +1,222 @@
+##Amish Obesity data analysis
+## Using a subsample of 500 instead (Go from 68 to 80 total samples)
+##Feb 1, 2016
+
+setwd("C:/users/marc/Desktop/obesity2/Amish")
+
+#Read in and match metadata to microbiome data
+metadata <- read.csv("amish_obesity_table2.csv")
+test <- metadata[!duplicated(metadata$submitted_sample_id_s), ]
+rownames(test) <- test[, 5]
+shared.data <- read.table("Amish.subsample500.shared", header=T)
+rownames(shared.data) <- shared.data[, 2]
+shared.data <- shared.data[, -c(1:3)]
+keep  <- rownames(shared.data)
+
+test2 <- test[keep, ]
+good.metadata <- test2
+rm(metadata, test, test2)
+
+#Create alpha diversity table
+library(vegan)
+H <- diversity(shared.data)
+S <- specnumber(shared.data)
+J <- H/log(S)
+select.alpha.diversity <- as.data.frame(cbind(H, S, J))
+alpha.diversity <- as.data.frame(select.alpha.diversity)
+alpha.test <- as.data.frame(alpha.diversity)
+rm(select.alpha.diversity)
+
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.info.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(shared.data)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- shared.data
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Deferribacteres", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(3, 5:6, 8:10)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
[…]

### 2016-02-05 code/HMP.BF.dataAnalysis.Jan132016.R

@@ -1,0 +1,94 @@
+##Data Analysis
+##HMP and Obesity
+##Nov 23, 2015
+##Looking at prediction
+
+
+
+##Read in relevant data
+
+setwd("C:/users/marc/Desktop/obesity2/HMP.analysis")
+microbiome <- read.table("Stool.an.0.03.subsample.shared", header=T)
+meta.cat <- read.table("categorical.metadata", header=T)
+meta.cont <- read.table("continuous.metadata", header=T)
+
+#Only interested in first visit microbiome data
+#need to create microbiome data set for only that
+
+test <- microbiome[grep("\\.01\\.", microbiome$Group), ] #selecting by .01. 
+
+microb.rownames <- gsub("([0-9]+).*", "\\1", test$Group) #extracting only numeric before first "."
+rownames(test) <- microb.rownames
+
+#Get rid of information not used in downstram analysis
+test <- test[, -c(1:3)]
+
+#Subset data to the microbiome total n
+select.meta.cat <- meta.cat[microb.rownames, ]
+select.meta.cont <- meta.cont[microb.rownames, ]
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(test)
+S <- specnumber(test)
+J <- H/log(S)
+select.alpha.diversity <- as.data.frame(cbind(H, S, J))
+alpha.diversity <- as.data.frame(select.alpha.diversity)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(test)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- test
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+
[…]

### 2016-02-05 code/HMP.obesity.Nov32015.R

@@ -1,0 +1,122 @@
+##Data Analysis
+##HMP and Obesity
+##Nov 3, 2015
+##Investigating Relative Risk as a means of looking at obesity and microbiome
+
+
+
+##Read in relevant data
+
+setwd("C:/users/marc/Desktop/HMP.analysis")
+microbiome <- read.table("Stool.an.0.03.subsample.shared", header=T)
+meta.cat <- read.table("categorical.metadata", header=T)
+meta.cont <- read.table("continuous.metadata", header=T)
+
+#Only interested in first visit microbiome data
+#need to create microbiome data set for only that
+
+test <- microbiome[grep("\\.01\\.", microbiome$Group), ] #selecting by .01. 
+
+microb.rownames <- gsub("([0-9]+).*", "\\1", test$Group) #extracting only numeric before first "."
+rownames(test) <- microb.rownames
+
+#Get rid of information not used in downstram analysis
+test <- test[, -c(1:3)]
+
+#Subset data to the microbiome total n
+select.meta.cat <- meta.cat[microb.rownames, ]
+select.meta.cont <- meta.cont[microb.rownames, ]
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(test)
+S <- specnumber(test)
+J <- H/log(S)
+select.alpha.diversity <- as.data.frame(cbind(H, S, J))
+
+#generate categorical variables of interest
+bmi <- select.meta.cont$DTPBMI
+bmi.cat <- select.meta.cat$BMI_C
+caucasian <- select.meta.cat$WHITE_C
+asian <-select.meta.cat$ASIAN_C
+black <- select.meta.cat$BLACK_C
+
+#Add bacterial microbiome categorical variables based on median values
+
+#Shannon Diversity
+median(H) #2.541824
+alpha.test <- within(select.alpha.diversity, {shannon.cat = ifelse(H <= 2.541824, "less", "higher")})
+
+#OTU Richness
+median(S) #57
+alpha.test <- within(alpha.test, {S.cat = ifelse(S <= 57, "less", "higher")})
+
+#Evenness
+median(J)
+alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
+
+##Shannon Diversity
[…]

### 2016-02-05 code/HMP.pred.obesity.Nov232015.R

@@ -1,0 +1,177 @@
+##Data Analysis
+##HMP and Obesity
+##Nov 23, 2015
+##Looking at prediction
+
+
+
+##Read in relevant data
+
+setwd("C:/users/marc/Desktop/obesity2/HMP.analysis")
+microbiome <- read.table("Stool.an.0.03.subsample.shared", header=T)
+meta.cat <- read.table("categorical.metadata", header=T)
+meta.cont <- read.table("continuous.metadata", header=T)
+
+#Only interested in first visit microbiome data
+#need to create microbiome data set for only that
+
+test <- microbiome[grep("\\.01\\.", microbiome$Group), ] #selecting by .01. 
+
+microb.rownames <- gsub("([0-9]+).*", "\\1", test$Group) #extracting only numeric before first "."
+rownames(test) <- microb.rownames
+
+#Get rid of information not used in downstram analysis
+test <- test[, -c(1:3)]
+
+#Subset data to the microbiome total n
+select.meta.cat <- meta.cat[microb.rownames, ]
+select.meta.cont <- meta.cont[microb.rownames, ]
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(test)
+S <- specnumber(test)
+J <- H/log(S)
+select.alpha.diversity <- as.data.frame(cbind(H, S, J))
+alpha.diversity <- as.data.frame(select.alpha.diversity)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(test)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- test
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
[…]

### 2016-02-05 code/Jan122016.Columbian.BF.obesity.R

@@ -1,0 +1,100 @@
+##Columbian Obesity data analysis
+##Nov 10, 2015
+
+setwd("C:/users/marc/Desktop/obesity2/columbian/")
+
+columbian.microb <- read.table("columbian.0.03.subsample.shared", header=T)
+metadata <- read.csv("columbian_dataset.csv")
+
+#Organize the microbiome shared data
+rownames(columbian.microb) <- columbian.microb[, 2]
+columbian.microb <- columbian.microb[, -c(1:3)]
+
+#Organize the metadata
+rownames(metadata) <- metadata[, 5]
+
+#Get only data that we are interested in
+edit.metadata <- metadata[, -c(1:7, 9, 13:15, 16:52)]
+
+#Sort metadata into the same order as the microbiome data
+order1 <- rownames(columbian.microb)
+edit.metadata2 <- edit.metadata[order1, ]
+
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(columbian.microb)
+S <- specnumber(columbian.microb)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(columbian.microb)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- columbian.microb
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(4:5, 7:9)]
+
[…]

### 2016-02-05 code/Jan122016.Hispanic.BF.obesity.R

@@ -1,0 +1,114 @@
+##Hispanic Obesity data analysis
+##Nov 11, 2015
+
+setwd("C:/users/marc/Desktop/obesity2/hispanic/")
+
+hispanic.microb <- read.table("hispanic.subsample.shared", header=T)
+
+metadata <- read.csv("s40168-015-0072-y-s1.csv")
+sample.match <- read.csv("Hispanic_dataset.csv")
+
+#Create a microbiome data table with sample names from metadata
+test <- cbind(as.character(sample.match$Run_s), as.character(sample.match$Library_Name_s))
+test <- test[order(test[, 1]), ]
+
+test2 <- cbind(test, hispanic.microb)
+rownames(test2) <- test2[, 2]
+his.microb.edit <- test2[, -c(1:5)]
+edit.metadata <- metadata[, -c(2:5)]
+rownames(edit.metadata) <- edit.metadata[, 1]
+edit.metadata <- edit.metadata[, -1]
+
+#Create a metadata file in the order of the microbiome data
+order1 <- rownames(his.microb.edit)
+edit.metadata2 <- edit.metadata[order1, ]
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(his.microb.edit)
+S <- specnumber(his.microb.edit)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(his.microb.edit)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- his.microb.edit
+colnames(phyla.table) <- phyla.names
+  #add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
[…]

### 2016-02-05 code/Jan122016.MetaHIT.obesity.R

@@ -1,0 +1,162 @@
+##MetaHit (Danish) Obesity data analysis
+##Nov 17, 2015
+
+setwd("C:/users/marc/Desktop/obesity2/MetaHit/")
+
+microb <- read.csv("finalized_metaHit_shared.csv")
+rownames(microb) <- microb[, 1]
+microb <- microb[,-1]
+
+metadata <- read.csv("MetaHit_metadata.csv")
+rownames(metadata) <- metadata[, 1]
+metadata <- metadata[, -1]
+
+# Generate Overall Relative Abundance - normalize to a value of 100
+# Needs to be done since not all measures have 100 bacteria from metaphlan2
+overall <- rowSums(microb)
+microb.norm <- (microb / overall) * 100
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(microb.norm)
+S <- specnumber(microb.norm)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+# Seems to look okay....
+
+#Create BMI groups
+metadata$BMI.class[metadata$BMI<=24] <- "Normal"
+metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
+metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
+metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
+
+#Create Obese Yes/No groups
+metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
+metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
+
+#Plot NMDS and test with PERMANOVA
+#Non-Obese versus Obese
+
+obese <- factor(metadata$obese)
+cols.obese <- c('blue', 'red')[obese]
+obese.nmds <- capscale(microb.norm ~ 1, distance = "bray")
+plot(obese.nmds, type="n")
+points(obese.nmds, display="sites", pch=16, col=cols.obese)
+legend("bottomright", c("Yes", "No"), pch=16, col=c('red', 'blue'))
+set.seed(3)
+adonis(microb.norm ~ obese, permutations=1000) 
[…]

### 2016-02-05 code/Jan132016.Columbian.BF.dataAnalysis.R

@@ -1,0 +1,89 @@
+##Columbian Obesity data analysis
+##Nov 10, 2015
+
+setwd("C:/users/marc/Desktop/obesity2/columbian/")
+
+columbian.microb <- read.table("columbian.0.03.subsample.shared", header=T)
+metadata <- read.csv("columbian_dataset.csv")
+
+#Organize the microbiome shared data
+rownames(columbian.microb) <- columbian.microb[, 2]
+columbian.microb <- columbian.microb[, -c(1:3)]
+
+#Organize the metadata
+rownames(metadata) <- metadata[, 5]
+
+#Get only data that we are interested in
+edit.metadata <- metadata[, -c(1:7, 9, 13:15, 16:52)]
+
+#Sort metadata into the same order as the microbiome data
+order1 <- rownames(columbian.microb)
+edit.metadata2 <- edit.metadata[order1, ]
+
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(columbian.microb)
+S <- specnumber(columbian.microb)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(columbian.microb)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- columbian.microb
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(4:5, 7:9)]
+
[…]

### 2016-02-05 code/Jan132016.Korean.BF.overweight.R

@@ -1,0 +1,97 @@
+##Korean Obesity (PLOS One 2011) data analysis
+##Nov 23, 2015
+##There are no obese Koreans in this data set
+##So have to look at normal vs overweight
+##Look at prediction
+
+setwd("C:/users/marc/Desktop/obesity2/korean/")
+
+microb <- read.table("DRR000776.0.03.subsample.shared", header=T)
+rownames(microb) <- microb[, 2]
+microb <- microb[,-c(1:3)]
+
+metadata <- read.csv("korean.metadata.csv")
+
+#Need to get rid of the repeated samples and the children
+  #First the metadata
+metadata2 <- metadata[-c(2:3, 5:6, 8:9, 11:12, 14:15, 17:18, 19:24), ]
+rownames(metadata2) <- metadata2[, 2]
+metadata2 <- metadata2[, -c(1:2)]
+
+  #Second the microbiome data set
+microb2 <- microb[-c(2:3, 5:6, 8:9, 11:12, 14:15, 17:18, 19:24), ]
+rownames(microb2) <- rownames(metadata2)
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(microb2)
+S <- specnumber(microb2)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+# Seems to look okay....
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(microb2)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- microb2
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
[…]

### 2016-02-05 code/Nov102015.Columbian.obesity.R

@@ -1,0 +1,129 @@
+##Columbian Obesity data analysis
+##Nov 10, 2015
+
+setwd("C:/users/marc/Desktop/obesity2/columbian/")
+
+columbian.microb <- read.table("columbian.0.03.subsample.shared", header=T)
+metadata <- read.csv("columbian_dataset.csv")
+
+#Organize the microbiome shared data
+rownames(columbian.microb) <- columbian.microb[, 2]
+columbian.microb <- columbian.microb[, -c(1:3)]
+
+#Organize the metadata
+rownames(metadata) <- metadata[, 5]
+
+#Get only data that we are interested in
+edit.metadata <- metadata[, -c(1:7, 9, 13:15, 16:52)]
+
+#Sort metadata into the same order as the microbiome data
+order1 <- rownames(columbian.microb)
+edit.metadata2 <- edit.metadata[order1, ]
+
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(columbian.microb)
+S <- specnumber(columbian.microb)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+
+#Get paitent demographics data to be tested 
+
+bmi <- edit.metadata2$body_mass_index_s
+
+#Create Obese Yes/No groups
+edit.metadata2$obese[edit.metadata2$description_s=="Adequate weight male stool sample" | edit.metadata2$description_s=="Overweight male stool sample" | edit.metadata2$description_s=="Adequate weight female stool sample" | edit.metadata2$description_s=="Overweight female stool sample"] <- "No"
+edit.metadata2$obese[edit.metadata2$description_s=="Obese male stool sample" | edit.metadata2$description_s=="Obese female stool sample"] <- "Yes"
+
+
+#Plot NMDS and test with PERMANOVA
+#Non-Obese versus Obese
+
+obese <- factor(edit.metadata2$obese)
+cols.obese <- c('blue', 'red')[obese]
+obese.nmds <- capscale(columbian.microb ~ 1, distance = "bray")
+plot(obese.nmds, type="n")
+points(obese.nmds, display="sites", pch=16, col=cols.obese)
[…]

### 2016-02-05 code/Nov102015.Hispanic.obesity.R

@@ -1,0 +1,150 @@
+##Hispanic Obesity data analysis
+##Nov 11, 2015
+
+setwd("C:/users/marc/Desktop/obesity2/hispanic/")
+
+hispanic.microb <- read.table("hispanic.subsample.shared", header=T)
+
+metadata <- read.csv("s40168-015-0072-y-s1.csv")
+sample.match <- read.csv("Hispanic_dataset.csv")
+
+#Create a microbiome data table with sample names from metadata
+test <- cbind(as.character(sample.match$Run_s), as.character(sample.match$Library_Name_s))
+test <- test[order(test[, 1]), ]
+
+test2 <- cbind(test, hispanic.microb)
+rownames(test2) <- test2[, 2]
+his.microb.edit <- test2[, -c(1:5)]
+edit.metadata <- metadata[, -c(2:5)]
+rownames(edit.metadata) <- edit.metadata[, 1]
+edit.metadata <- edit.metadata[, -1]
+
+#Create a metadata file in the order of the microbiome data
+order1 <- rownames(his.microb.edit)
+edit.metadata2 <- edit.metadata[order1, ]
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(his.microb.edit)
+S <- specnumber(his.microb.edit)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Get paitent demographics to be tested
+
+bmi <- edit.metadata2$BMI
+
+#Create BMI groups
+edit.metadata2$BMI.class[edit.metadata2$BMI<=24] <- "Normal"
+edit.metadata2$BMI.class[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
+edit.metadata2$BMI.class[edit.metadata2$BMI>=30 & edit.metadata2$BMI<40] <- "Obese"
+edit.metadata2$BMI.class[edit.metadata2$BMI>=40] <- "Extreme Obesity"
+
+
+#Create Obese Yes/No groups
+edit.metadata2$obese[edit.metadata2$BMI.class=="Normal" | edit.metadata2$BMI.class=="Overweight"] <- "No"
+edit.metadata2$obese[edit.metadata2$BMI.class=="Obese" | edit.metadata2$BMI.class=="Extreme Obesity"] <- "Yes"
+
[…]

### 2016-02-05 code/Nov172015.MetaHIT.obesity.R

@@ -1,0 +1,152 @@
+##MetaHit (Danish) Obesity data analysis
+##Nov 17, 2015
+
+setwd("C:/users/marc/Desktop/obesity2/MetaHit/")
+
+microb <- read.csv("finalized_metaHit_shared.csv")
+rownames(microb) <- microb[, 1]
+microb <- microb[,-1]
+
+metadata <- read.csv("MetaHit_metadata.csv")
+rownames(metadata) <- metadata[, 1]
+metadata <- metadata[, -1]
+
+# Generate Overall Relative Abundance - normalize to a value of 100
+# Needs to be done since not all measures have 100 bacteria from metaphlan2
+overall <- rowSums(microb)
+microb.norm <- (microb / overall) * 100
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(microb.norm)
+S <- specnumber(microb.norm)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+# Seems to look okay....
+
+#Create BMI groups
+metadata$BMI.class[metadata$BMI<=24] <- "Normal"
+metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
+metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
+metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
+
+#Create Obese Yes/No groups
+metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
+metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
+
+#Plot NMDS and test with PERMANOVA
+#Non-Obese versus Obese
+
+obese <- factor(metadata$obese)
+cols.obese <- c('blue', 'red')[obese]
+obese.nmds <- capscale(microb.norm ~ 1, distance = "bray")
+plot(obese.nmds, type="n")
+points(obese.nmds, display="sites", pch=16, col=cols.obese)
+legend("bottomright", c("Yes", "No"), pch=16, col=c('red', 'blue'))
+set.seed(3)
+adonis(microb.norm ~ obese, permutations=1000) 
+#PERMANOVA=0.05495, pseudo-F=1.6047
[…]

### 2016-02-05 code/Nov192015.Korean.overweight.R

@@ -1,0 +1,130 @@
+##Korean Obesity (PLOS One 2011) data analysis
+##Nov 17, 2015
+##There are no obese Koreans in this data set
+##So have to look at normal vs overweight
+
+setwd("C:/users/marc/Desktop/obesity2/korean/")
+
+microb <- read.table("DRR000776.0.03.subsample.shared", header=T)
+rownames(microb) <- microb[, 2]
+microb <- microb[,-c(1:3)]
+
+metadata <- read.csv("korean.metadata.csv")
+
+#Need to get rid of the repeated samples and the children
+  #First the metadata
+metadata2 <- metadata[-c(2:3, 5:6, 8:9, 11:12, 14:15, 17:18, 19:24), ]
+rownames(metadata2) <- metadata2[, 2]
+metadata2 <- metadata2[, -c(1:2)]
+
+  #Second the microbiome data set
+microb2 <- microb[-c(2:3, 5:6, 8:9, 11:12, 14:15, 17:18, 19:24), ]
+rownames(microb2) <- rownames(metadata2)
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(microb2)
+S <- specnumber(microb2)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+# Seems to look okay....
+
+#Create BMI groups
+metadata2$BMI.class[metadata2$BMI<=24] <- "Normal"
+metadata2$BMI.class[metadata2$BMI>24 & metadata2$BMI<30] <- "Overweight"
+metadata2$BMI.class[metadata2$BMI>=30 & metadata2$BMI<40] <- "Obese"
+metadata2$BMI.class[metadata2$BMI>=40] <- "Extreme Obesity"
+
+#Create Obese Yes/No groups
+metadata2$obese[metadata2$BMI.class=="Normal" | metadata2$BMI.class=="Overweight"] <- "No"
+metadata2$obese[metadata2$BMI.class=="Obese" | metadata2$BMI.class=="Extreme Obesity"] <- "Yes"
+
+#Create Overweight Yes/No groups
+metadata2$overweight[metadata2$BMI.class=="Normal"] <- "No"
+metadata2$overweight[metadata2$BMI.class=="Obese" | metadata2$BMI.class=="Extreme Obesity" | metadata2$BMI.class=="Overweight"] <- "Yes"
+
[…]

### 2016-02-05 code/Nov232015.Columbian.pred.obesity.R

@@ -1,0 +1,116 @@
+##Columbian Obesity data analysis
+##Nov 10, 2015
+
+setwd("C:/users/marc/Desktop/obesity2/columbian/")
+
+columbian.microb <- read.table("columbian.0.03.subsample.shared", header=T)
+metadata <- read.csv("columbian_dataset.csv")
+
+#Organize the microbiome shared data
+rownames(columbian.microb) <- columbian.microb[, 2]
+columbian.microb <- columbian.microb[, -c(1:3)]
+
+#Organize the metadata
+rownames(metadata) <- metadata[, 5]
+
+#Get only data that we are interested in
+edit.metadata <- metadata[, -c(1:7, 9, 13:15, 16:52)]
+
+#Sort metadata into the same order as the microbiome data
+order1 <- rownames(columbian.microb)
+edit.metadata2 <- edit.metadata[order1, ]
+
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(columbian.microb)
+S <- specnumber(columbian.microb)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(columbian.microb)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- columbian.microb
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(4:5, 7:9)]
+
[…]

### 2016-02-05 code/Nov232015.Hispanic.pred.obesity.R

@@ -1,0 +1,130 @@
+##Hispanic Obesity data analysis
+##Nov 11, 2015
+
+setwd("C:/users/marc/Desktop/obesity2/hispanic/")
+
+hispanic.microb <- read.table("hispanic.subsample.shared", header=T)
+
+metadata <- read.csv("s40168-015-0072-y-s1.csv")
+sample.match <- read.csv("Hispanic_dataset.csv")
+
+#Create a microbiome data table with sample names from metadata
+test <- cbind(as.character(sample.match$Run_s), as.character(sample.match$Library_Name_s))
+test <- test[order(test[, 1]), ]
+
+test2 <- cbind(test, hispanic.microb)
+rownames(test2) <- test2[, 2]
+his.microb.edit <- test2[, -c(1:5)]
+edit.metadata <- metadata[, -c(2:5)]
+rownames(edit.metadata) <- edit.metadata[, 1]
+edit.metadata <- edit.metadata[, -1]
+
+#Create a metadata file in the order of the microbiome data
+order1 <- rownames(his.microb.edit)
+edit.metadata2 <- edit.metadata[order1, ]
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(his.microb.edit)
+S <- specnumber(his.microb.edit)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(his.microb.edit)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- his.microb.edit
+colnames(phyla.table) <- phyla.names
+  #add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
[…]

### 2016-02-05 code/Nov232015.Korean.pred.overweight.R

@@ -1,0 +1,105 @@
+##Korean Obesity (PLOS One 2011) data analysis
+##Nov 23, 2015
+##There are no obese Koreans in this data set
+##So have to look at normal vs overweight
+##Look at prediction
+
+setwd("C:/users/marc/Desktop/obesity2/korean/")
+
+microb <- read.table("DRR000776.0.03.subsample.shared", header=T)
+rownames(microb) <- microb[, 2]
+microb <- microb[,-c(1:3)]
+
+metadata <- read.csv("korean.metadata.csv")
+
+#Need to get rid of the repeated samples and the children
+  #First the metadata
+metadata2 <- metadata[-c(2:3, 5:6, 8:9, 11:12, 14:15, 17:18, 19:24), ]
+rownames(metadata2) <- metadata2[, 2]
+metadata2 <- metadata2[, -c(1:2)]
+
+  #Second the microbiome data set
+microb2 <- microb[-c(2:3, 5:6, 8:9, 11:12, 14:15, 17:18, 19:24), ]
+rownames(microb2) <- rownames(metadata2)
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(microb2)
+S <- specnumber(microb2)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+# Seems to look okay....
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(microb2)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- microb2
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
[…]

### 2016-02-05 code/Nov232015.MetaHIT.pred.obesity.R

@@ -1,0 +1,118 @@
+##MetaHit (Danish) Obesity data analysis
+##Nov 23, 2015
+##Invesitgating prediction
+
+setwd("C:/users/marc/Desktop/obesity2/MetaHit/")
+
+microb <- read.csv("finalized_metaHit_shared.csv")
+rownames(microb) <- microb[, 1]
+microb <- microb[,-1]
+
+metadata <- read.csv("MetaHit_metadata.csv")
+rownames(metadata) <- metadata[, 1]
+metadata <- metadata[, -1]
+
+phyla <- read.csv("phyla.csv")
+rownames(phyla) <- phyla[, 1]
+phyla <- phyla[, -1]
+
+# Generate Overall Relative Abundance - normalize to a value of 100
+# Needs to be done since not all measures have 100 bacteria from metaphlan2
+overall <- rowSums(microb)
+microb.norm <- (microb / overall) * 100
+
+# Generate Overall Relative Abundance for phyla - normalize to a value of 100
+# Needs to be done since not all measures have 100 bacteria from metaphlan2
+t.phyla <- as.data.frame(t(phyla))
+t.phyla$other <- apply(t.phyla[, c("Acidobacteria", "Candidatus_Saccharibacteria", "Chlorobi", "Fusobacteria", "Spirochaetes")], 1, sum)
+t.phyla <- t.phyla[, -c(1, 4:5, 7, 9)]
+overall2 <- rowSums(t.phyla)
+phyla.norm <- (t.phyla / overall) * 100
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(microb.norm)
+S <- specnumber(microb.norm)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+# Seems to look okay....
+
+#Create BMI groups
+metadata$BMI.class[metadata$BMI<=24] <- "Normal"
+metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
+metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
+metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
+
+#Create Obese Yes/No groups
+metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
[…]

### 2016-02-05 code/Zupancic.Feb52016.dataAnalysis.R
deletions

@@ -1,196 +1,0 @@
-#Zupancic et al.
-# Single study analysis
-#Feb 5th, 2016
-
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-
-setwd("C:/users/marc/Desktop/obesity2/Amish/updatedGOOD")
-
-#Read in and match metadata to microbiome data
-metadata <- read.csv("amish_obesity_table2.csv")
-test <- metadata[!duplicated(metadata$submitted_sample_id_s), ]
-rownames(test) <- test[, 4]
-shared.data <- read.table("combined.0.03.subsample.shared", header=T)
-rownames(shared.data) <- shared.data[, 2]
-shared.data <- shared.data[, -c(1:3)]
-keep  <- rownames(shared.data)
-
-test2 <- test[keep, ]
-good.metadata <- test2
-rm(metadata, test, test2)
-
-metadata <- good.metadata[!grepl("_s2", good.metadata$submitted_sample_id_s),]
-keep <- rownames(metadata)
-microbiome <- shared.data[keep, ]
-rm(good.metadata, shared.data, keep)
-
-
-#generate alpha diversity measures with vegan
-library(vegan)
-H <- diversity(microbiome)
-S <- specnumber(microbiome)
-J <- H/log(S)
-select.alpha.diversity <- as.data.frame(cbind(H, S, J))
-s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
-alpha.test <- s1.alpha.diversity
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(microbiome)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- microbiome
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-#combine phyla that are not that abundant
[…]

### 2016-02-05 code/StatisticsProcessing.Rmd
updated statistics

@@
 good.metadata <- test2
-rm(metadata, test, test2)
-
-metadata <- good.metadata[!grepl("_s2", good.metadata$submitted_sample_id_s),]
-keep <- rownames(metadata)
+test3 <- test2[!duplicated(test2$submitted_subject_id_s), ]
+keep <- rownames(test3)
 microbiome <- shared.data[keep, ]
-rm(good.metadata, shared.data, keep)
+metadata <- test3
+
+rm(good.metadata, shared.data, keep, test, test2, test3)
 
@@
 
-anova(lm(H ~ obese)) #P-value=0.2145
-anova(lm(S ~ obese)) #P-value=0.05925
-anova(lm(J ~ obese)) #P-value=0.5093
+anova(lm(H ~ obese)) #P-value=0.2034
+anova(lm(S ~ obese)) #P-value=0.03516
+anova(lm(J ~ obese)) #P-value=0.5369
 
@@
 
-anova(lm(bacter ~ obese)) #P-value=0.9485
-anova(lm(firm ~ obese)) #P-value=0.2379
-anova(lm(BFratio ~ obese)) #P-value=0.9277
+anova(lm(bacter ~ obese)) #P-value=0.9582
+anova(lm(firm ~ obese)) #P-value=0.2654
+anova(lm(BFratio ~ obese)) #P-value=0.9899
 
@@
 adonis(microbiome ~ obese, permutations=1000) 
-#PERMANOVA=0.5504, pseudo-F=0.97273
+#PERMANOVA=0.5734, pseudo-F=0.95626
 
@@
 test3 <- test3[order(H.cat), ]
-table(test3[c(1:100), 4])
-table(test3[c(101:201), 4])
-#Group1 (Higher than median), obese = 29 and non-obese = 71
-#Group2 (Lower than median), obese = 45 and non-obese = 56
-
-group1 <- c(29, 71)
-group2 <- c(45, 56)
+table(test3[c(1:103), 4])
+table(test3[c(104:207), 4])
+#Group1 (Higher than median), obese = 31 and non-obese = 72
+#Group2 (Lower than median), obese = 47 and non-obese = 57
+
+group1 <- c(31, 72)
+group2 <- c(47, 57)
 r.test <- rbind(group2, group1)
@@
 epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 1.54
-## CI = 1.05, 2.24
-## p-value = 0.022
+## Risk Ratio = 1.50
+## CI = 1.04, 2.16
+## p-value = 0.025
 
@@
 
-median(BFRatio$BFRatio) # 0.537415
+median(BFRatio$BFRatio) # 0.5414258
 BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
@@
 test4 <- test4[order(BFRatio.cat), ]
-table(test4[c(1:100), 2]) #higher group
-table(test4[c(101:201), 2]) #lower group
-#Group1 (Higher than median), obese = 38 and non-obese = 62
-#Group2 (Lower than median), obese = 36 and non-obese = 65
-
-group1 <- c(38, 62)
-group2 <- c(36, 65)
+table(test4[c(1:103), 2]) #higher group
+table(test4[c(104:207), 2]) #lower group
+#Group1 (Higher than median), obese = 41 and non-obese = 62
+#Group2 (Lower than median), obese = 37 and non-obese = 67
+
+group1 <- c(41, 62)
+group2 <- c(37, 67)
 r.test <- rbind(group2, group1)
@@
 epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 0.94
-## CI = 0.65, 1.35
-## p-value = 0.729
+## Risk Ratio = 0.89
+## CI = 0.66, 1.27
+## p-value = 0.53
 
@@
 fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-summary(fit) # list of 32 Measures, AUCopt = 0.6679613
+summary(fit) # list of 42 Measures, AUCopt = 0.6365534
 plot(fit)
@@
 StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8)
-Study <- c("Baxter", "HMP", "Ross", "Escobar", "Argumugam", "Turnbaugh", "Zupancic", "Wu")
+Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu")
 Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012")
-Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=201", "n=58")
-tpos <- c(25, 16, 20, 6, 20, 52, 45, 2) #obese in low diversity group
-tneg <- c(61, 112, 12, 9, 23, 21, 56, 27) #nonobese in low diversity group
-cpos <- c(22, 10, 18, 4, 17, 47, 29, 3) #obese in high diversity group
-cneg <- c(64, 118, 13, 11, 25, 26, 71, 26) #nonobese in high diversity group
-RR <- c(1.14, 1.6, 1.08, 1.50, 1.15, 1.11, 1.54, 0.67)
-low <- c(0.70, 0.75, 0.72, 0.53, 0.71, 0.88, 1.05, 0.12)
-high <- c(1.85, 3.39, 1.61, 4.26, 1.87, 1.38, 2.24, 3.70)
+Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=207", "n=58")
+tpos <- c(25, 16, 20, 6, 20, 52, 47, 2) #obese in low diversity group
+tneg <- c(61, 112, 12, 9, 23, 21, 57, 27) #nonobese in low diversity group
+cpos <- c(22, 10, 18, 4, 17, 47, 31, 3) #obese in high diversity group
+cneg <- c(64, 118, 13, 11, 25, 26, 72, 26) #nonobese in high diversity group
+RR <- c(1.14, 1.6, 1.08, 1.50, 1.15, 1.11, 1.50, 0.67)
+low <- c(0.70, 0.75, 0.72, 0.53, 0.71, 0.88, 1.04, 0.12)
+high <- c(1.85, 3.39, 1.61, 4.26, 1.87, 1.38, 2.16, 3.70)
 
@@
 
-## Alternative way - 
-# Seems easier to follow plus I am not trying to do
-# a meta analysis on a model
-
-data <-as.data.frame(cbind(RR, low, high))
-raw.overall2 <- as.data.frame(cbind(Study,Year, data))
-
-raw.overall2$logRR <- log(raw.overall2$RR)
-raw.overall2$SE <- (log(raw.overall2$high) - log(raw.overall2$low))/(2*1.96)
-attach(raw.overall2)
-metaplot(logRR, SE, labels=Study, nn=1, logeffect=TRUE, col=meta.colors("black"), xlab="RR of Obesity in high versus low Shannon diversity based on Median", ylab="")
-  #Graphical output is not as informative
-
 ##Calculate combined P-value
 m = 8
-T = -2*log(0.608)-2*log(0.214)-2*log(0.236)-2*log(0.439)-2*log(0.575)-2*log(0.376)-2*log(0.022)-2*log(0.64)
+T = -2*log(0.608)-2*log(0.214)-2*log(0.236)-2*log(0.439)-2*log(0.575)-2*log(0.376)-2*log(0.025)-2*log(0.64)
 T.pval = 1-pchisq(T,m) #P-value = 0.01
@@
 anova.rma(res)
-# QM = 5.0973, p-value = 0.0240
+# QM = 5.1101, p-value = 0.0238
 
@@
 StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8)
-Study <- c("Baxter", "HMP", "Ross", "Escobar", "Argumugam", "Turnbaugh", "Zupancic", "Wu")
+Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu")
 Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012")
-Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=201", "n=58")
-tpos <- c(24, 13, 16, 3, 17, 45, 36, 2) #obese in low BF group
-tneg <- c(62, 115, 16, 12, 26, 28, 65, 27) #nonobese in low BF group
-cpos <- c(23, 13, 22, 7, 20, 54, 38, 3) #obese in high BF group
+Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=207", "n=58")
+tpos <- c(24, 13, 16, 3, 17, 45, 37, 2) #obese in low BF group
+tneg <- c(62, 115, 16, 12, 26, 28, 67, 27) #nonobese in low BF group
+cpos <- c(23, 13, 22, 7, 20, 54, 41, 3) #obese in high BF group
 cneg <- c(63, 115, 9, 8, 22, 19, 62, 26) #nonobese in high BF group
-RR <- c(1.04, 1.00, 0.70, 0.43, 0.83, 0.83, 0.94, 0.67)
-low <- c(0.64, 0.48, 0.47, 0.14, 0.51, 0.66, 0.65, 0.12)
-high <- c(1.70, 2.07, 1.07, 1.35, 1.35, 1.05, 1.35, 3.70)
+RR <- c(1.04, 1.00, 0.70, 0.43, 0.83, 0.83, 0.89, 0.67)
+low <- c(0.64, 0.48, 0.47, 0.14, 0.51, 0.66, 0.66, 0.12)
+high <- c(1.70, 2.07, 1.07, 1.35, 1.35, 1.05, 1.27, 3.70)
 
@@
 m = 8
-T = -2*log(0.864)-2*log(1.00)-2*log(0.089)-2*log(0.121)-2*log(0.452)-2*log(0.111)-2*log(0.729)-2*log(0.64)
+T = -2*log(0.864)-2*log(1.00)-2*log(0.089)-2*log(0.121)-2*log(0.452)-2*log(0.111)-2*log(0.53)-2*log(0.64)
 T.pval = 1-pchisq(T,m) #P-value = 0.03155827
@@
 anova.rma(res)
-  # QM = 4.8452, p-value = 0.0277
+  # QM = 5.3494, p-value = 0.0207
 
@@
 
-## Load in libraries to be used as well as directory with the data
 library(ggplot2)
@@
 
-#Amish STD = 0.6010608
-amish.ZScore.H <- amish.data$H.corr/0.6010608
+#Amish STD = 0.5893173
+amish.ZScore.H <- amish.data$H.corr/0.5893173
 hist(amish.ZScore.H)
@@
 #Look at Obese versus non-obese
-t.test(H.Zscore ~ obese, data=group.data.test2) #P-value = 0.002792
-table(group.data.test2$obese) # No = 693 | Yes = 336
+t.test(H.Zscore ~ obese, data=group.data.test2) #P-value = 0.002583
+table(group.data.test2$obese) # No = 695 | Yes = 340
 p1 <- ggplot(group.data.test2, aes(obese, H.Zscore)) + geom_boxplot()
@@
 
-#Amish SD = 0.7990372
+#Amish SD = 0.7918527
 am.BF.corr <- as.character(amish.data$BF.ratio.corr)
 am.BF.corr <- as.numeric(am.BF.corr) # Value 19 has no value
-amish.norm.BF <- am.BF.corr/0.7990372
+amish.norm.BF <- am.BF.corr/0.7918527
 hist(amish.norm.BF)
@@
 t.test(BF.ratio.corr ~ obese, data=group.data.test2) 
-  # P-value = 0.05066
+  # P-value = 0.0525
 table(group.data.test2$obese)
-  # No = 693 | Yes = 336
+  # No = 695 | Yes = 340
 bf1 <- ggplot(group.data.test2, aes(obese, BF.ratio.corr)) + geom_boxplot()
@@
 t.test(Zbacter ~ obese, data=group.data.test2) 
-  # P-value = 0.3161
+  # P-value = 0.3202
 table(group.data.test2$obese)
-  # No = 693 | Yes = 336
+  # No = 695 | Yes = 340
 bacter1 <- ggplot(group.data.test2, aes(obese, Zbacter)) + geom_boxplot()
@@
 #Look at Obese versus non-obese
-t.test(Zfirm ~ obese, data=group.data.test2) #P-value = 0.8298
+t.test(Zfirm ~ obese, data=group.data.test2) #P-value = 0.8149
 table(group.data.test2$obese)
-  # No = 693 | Yes = 336
+  # No = 695 | Yes = 340
 firm1 <- ggplot(group.data.test2, aes(obese, Zfirm)) + geom_boxplot()
@@
 #Look at Obese versus non-obese
-t.test(zbf ~ obese, data=group.data.test2) #P-value = 0.4167
+t.test(zbf ~ obese, data=group.data.test2) #P-value = 0.4379
 table(group.data.test2$obese)
-  # No = 693 | Yes = 336
+  # No = 695 | Yes = 340
 zbf1 <- ggplot(group.data.test2, aes(obese, zbf)) + geom_boxplot()

### 2016-02-05 code/Zupancic.Feb52016.dataAnalysis.R

@@ -1,0 +1,196 @@
+#Zupancic et al.
+# Single study analysis
+#Feb 5th, 2016
+
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+
+setwd("C:/users/marc/Desktop/obesity2/Amish/updatedGOOD")
+
+#Read in and match metadata to microbiome data
+metadata <- read.csv("amish_obesity_table2.csv")
+test <- metadata[!duplicated(metadata$submitted_sample_id_s), ]
+rownames(test) <- test[, 4]
+shared.data <- read.table("combined.0.03.subsample.shared", header=T)
+rownames(shared.data) <- shared.data[, 2]
+shared.data <- shared.data[, -c(1:3)]
+keep  <- rownames(shared.data)
+
+test2 <- test[keep, ]
+good.metadata <- test2
+test3 <- test2[!duplicated(test2$submitted_subject_id_s), ]
+keep <- rownames(test3)
+microbiome <- shared.data[keep, ]
+metadata <- test3
+
+rm(good.metadata, shared.data, keep, test, test2, test3)
+
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(microbiome)
+S <- specnumber(microbiome)
+J <- H/log(S)
+select.alpha.diversity <- as.data.frame(cbind(H, S, J))
+s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
+alpha.test <- s1.alpha.diversity
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(microbiome)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- microbiome
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
[…]

### 2016-02-06 code/StatisticsProcessing.Rmd
twinsUK update

@@
 ##### **Goodrich et al. 2014:**  
-```{r}
-
+```{r, eval=FALSE}
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+setwd("C:/users/marc/Desktop/obesity2/twinsUK")
+
+#Read in and match metadata to microbiome data
+metadata <- read.csv("TwinsUKStudy_metadata.csv")
+rownames(metadata) <- metadata[, 1]
+shared.data <- read.table("combined.tx.1.subsample.shared", header=T)
+rownames(shared.data) <- shared.data[, 2]
+microbiome <- shared.data[, -c(1:3)]
+keep  <- rownames(metadata)
+
+microbiome<- microbiome[keep, ]
+
+rm(good.metadata, shared.data, keep)
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(microbiome)
+S <- specnumber(microbiome)
+J <- H/log(S)
+alpha.diversity <- as.data.frame(cbind(H, S, J))
+alpha.test <- alpha.diversity
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.table("phyla.txt", header=T)
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(microbiome)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- microbiome
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Chloroflexi", "Deferribacteres", "Elusimicrobia", "Fusobacteria", "Gemmatimonadetes", "Lentisphaerae", "OD1", "Planctomycetes", "Spirochaetes", "SR1", "Synergistetes", "Tenericutes")], 1, sum)
+phyla.table <- phyla.table[, -c(1, 4:6, 8:12, 14:17)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
+
+#Create BMI groups
+metadata$BMI.class[metadata$body_mass_index_s<=24] <- "Normal"
+metadata$BMI.class[metadata$body_mass_index_s>24 & metadata$body_mass_index_s<30] <- "Overweight"
+metadata$BMI.class[metadata$body_mass_index_s>=30 & metadata$body_mass_index_s<40] <- "Obese"
+metadata$BMI.class[metadata$body_mass_index_s>=40] <- "Extreme Obesity"
+
+#Create Obese Yes/No groups
+metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
+metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
+
+#Create a column with obese and extreme obese as single entity
+metadata$BMIclass2[metadata$BMI.class=="Normal"] <- "Normal"
+metadata$BMIclass2[metadata$BMI.class=="Overweight"] <- "Overweight"                     
+metadata$BMIclass2[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Obese"
+
+#Get paitent demographics to be tested
+bmi <- metadata$body_mass_index_s
+obese <- factor(metadata$obese)
+bmi.class <- factor(metadata$BMI.class)
+bmi.class2 <- factor(metadata$BMIclass2)
+
+###########################################################################
+######### First Level Analysis & Alpha Diversity with BMI #################
+###########################################################################
+
+##Test BMI versus alpha diversity and phyla
+
+anova(lm(H ~ obese)) #P-value=0.6671
+anova(lm(H ~ bmi.class)) #P-value=0.9732
+anova(lm(H ~ bmi.class2)) #P-value=0.8987
+
+anova(lm(S ~ obese)) #P-value=0.9073
+anova(lm(S ~ bmi.class)) #P-value=0.9124
+anova(lm(S ~ bmi.class2)) #P-value=0.8202
+
+anova(lm(J ~ obese)) #P-value=0.6319
+anova(lm(J ~ bmi.class)) #P-value=0.943
+anova(lm(J ~ bmi.class2)) #P-value=0.8394
+
+##linear correlations
+summary(lm(H ~ bmi)) #P-value=0.8662, R2=5.63x10-5
+summary(lm(S ~ bmi)) #P-value=0.9142, R2=2.302x10-5
+summary(lm(J ~ bmi)) #P-value=0.8582, R2=6.332x10-5
+
+#B and F tests against obesity
+bacter <- phyla.table.rel.abund$Bacteroidetes
+firm <- phyla.table.rel.abund$Firmicutes
+BFratio <- bacter/firm
+
+anova(lm(bacter ~ obese)) #P-value=0.6661
+anova(lm(bacter ~ bmi.class)) #P-value=0.3184
+anova(lm(bacter ~ bmi.class2)) #P-value=0.3184
+
+anova(lm(firm ~ obese)) #P-value=0.9688
+anova(lm(firm ~ bmi.class)) #P-value=0.399
+anova(lm(firm ~ bmi.class2)) #P-value=0.2455
+
+anova(lm(BFratio ~ obese)) #P-value=0.8709
+anova(lm(BFratio ~ bmi.class)) #P-value=0.7915
+anova(lm(BFratio ~ bmi.class2)) #P-value=0.5976
+
+###########################################################################
+############ NMDS and PERMANOVA Analysis###################################
+###########################################################################
+
+set.seed(3)
+adonis(microbiome ~ obese, permutations=1000) 
+#PERMANOVA=0.004995, pseudo-F=2.9308
+
+set.seed(3)
+adonis(microbiome ~ bmi.class, permutations=1000) 
+#PERMANOVA=0.005994, pseudo-F=1.8853
+
+set.seed(3)
+adonis(microbiome ~ bmi.class2, permutations=1000) 
+#PERMANOVA=0.007992, pseudo-F=2.1453
+
+###########################################################################
+############ Relative Risk#################################################
+###########################################################################
+
+library(epiR)
+
+#Generate median values and put them into existing alpha.test dataframe
+#Shannon diversity
+median(H) # 2.725646
+alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
+
+#OTU Richness
+median(S) # 73
+alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})
+
+#Evenness
+median(J) # 0.6311328
+alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
+
+##Shannon Diversity
+H.cat <- alpha.test$shannon.cat
+S.cat <- alpha.test$S.cat
+J.cat <- alpha.test$J.cat
+bmi.cat <- as.character(obese)
+test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
+test3 <- test3[order(H.cat), ]
+table(test3[c(1:253), 4])
+table(test3[c(254:507), 4])
+#Group1 (Higher than median), obese = 47 and non-obese = 206
+#Group2 (Lower than median), obese = 56 and non-obese = 198
+
+group1 <- c(47, 206)
+group2 <- c(56, 198)
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+epi.2by2(r.test, method="cohort.count")
+## Risk Ratio = 1.19
+## CI = 0.84, 1.68
+## p-value = 0.332
+
+##Run the RR for B/F ratio
+Bacter = phyla.table.rel.abund$Bacteroidetes
+Firm = phyla.table.rel.abund$Firmicutes
+BFRatio = Bacter/Firm
+BFRatio <- as.data.frame(BFRatio)
+
+median(BFRatio$BFRatio) # 0.5304444
+BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
+
+BFRatio.cat <- BFRatio$BFRatio.cat
+test4 <- cbind(BFRatio.cat, obese)
+test4 <- test4[order(BFRatio.cat), ]
+table(test4[c(1:253), 2]) #higher group
+table(test4[c(254:507), 2]) #lower group
+#Group1 (Higher than median), obese = 52 and non-obese = 201
+#Group2 (Lower than median), obese = 51 and non-obese = 203
+
+group1 <- c(52, 201)
+group2 <- c(51, 203)
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+epi.2by2(r.test, method="cohort.count")
+## Risk Ratio = 0.98
+## CI = 0.69, 1.38
+## p-value = 0.894
+
+###########################################################################
+############ Classification using AUCRF####################################
+###########################################################################
+
+library(AUCRF)
+
+#Create Obese.num group
+metadata$obese.num[metadata$obese=="No"] <- 0
+metadata$obese.num[metadata$obese=="Yes"] <- 1
+obese <- factor(metadata$obese.num)
+
+#generate test set
+# get rid of those with 0 and only 4 other values
+testset <- Filter(function(x)(length(unique(x))>5), microbiome)
+testset <- cbind(obese, testset)
+colnames(testset)[1] <- "obese"
+testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
+
+#Try AUCRF with default measures provided in readme
+set.seed(3)
+fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+summary(fit) # list of 8 Measures, AUCopt = 0.6767159
+plot(fit)
+
+#Try AUCRF with only phyla and alpha diversity measures in the testset
+testset <- cbind(obese, H, S, J, phyla.table.rel.abund)
+set.seed(3)
+fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+summary(fit) 
+# list of 7 Measures, AUCopt = 0.538138
+# Proteobacteria, unclassified, Verrucomicrobia
+plot(fit)
 ```
@@
 
-StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8)
-Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu")
-Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012")
-Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=207", "n=58")
-tpos <- c(25, 16, 20, 6, 20, 52, 47, 2) #obese in low diversity group
-tneg <- c(61, 112, 12, 9, 23, 21, 57, 27) #nonobese in low diversity group
-cpos <- c(22, 10, 18, 4, 17, 47, 31, 3) #obese in high diversity group
-cneg <- c(64, 118, 13, 11, 25, 26, 72, 26) #nonobese in high diversity group
-RR <- c(1.14, 1.6, 1.08, 1.50, 1.15, 1.11, 1.50, 0.67)
-low <- c(0.70, 0.75, 0.72, 0.53, 0.71, 0.88, 1.04, 0.12)
-high <- c(1.85, 3.39, 1.61, 4.26, 1.87, 1.38, 2.16, 3.70)
+StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
+Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu", "Goodrich")
+Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012", "2014")
+Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=207", "n=58", "n=507")
+tpos <- c(25, 16, 20, 6, 20, 52, 47, 2, 56) #obese in low diversity group
+tneg <- c(61, 112, 12, 9, 23, 21, 57, 27, 198) #nonobese in low diversity group
+cpos <- c(22, 10, 18, 4, 17, 47, 31, 3, 47) #obese in high diversity group
+cneg <- c(64, 118, 13, 11, 25, 26, 72, 26, 206) #nonobese in high diversity group
+RR <- c(1.14, 1.6, 1.08, 1.50, 1.15, 1.11, 1.50, 0.67, 1.19)
+low <- c(0.70, 0.75, 0.72, 0.53, 0.71, 0.88, 1.04, 0.12, 0.84)
+high <- c(1.85, 3.39, 1.61, 4.26, 1.87, 1.38, 2.16, 3.70, 1.68)
 
@@
                                                           #
-text(5.5, 9.3, "Relative Risk [95% CI]", pos=2, cex=0.8)  #
+text(5.5, 10.3, "Relative Risk [95% CI]", pos=2, cex=0.8)  #
   #smaller scale (Plot in R studia at normal)             #
@@
                                                          #
-text(-6.5, 9.3, "Study, Year, and n", pos=4, cex=0.8)    #
+text(-6.5, 10.3, "Study, Year, and n", pos=4, cex=0.8)    #
                                                          #
@@
 
+## Alternative way - 
+# Seems easier to follow plus I am not trying to do
+# a meta analysis on a model
+
+data <-as.data.frame(cbind(RR, low, high))
+raw.overall2 <- as.data.frame(cbind(Study,Year, data))
+
+raw.overall2$logRR <- log(raw.overall2$RR)
+raw.overall2$SE <- (log(raw.overall2$high) - log(raw.overall2$low))/(2*1.96)
+attach(raw.overall2)
+metaplot(logRR, SE, labels=Study, nn=1, logeffect=TRUE, col=meta.colors("black"), xlab="RR of Obesity in high versus low Shannon diversity based on Median", ylab="")
+  #Graphical output is not as informative
+
 ##Calculate combined P-value
 m = 8
-T = -2*log(0.608)-2*log(0.214)-2*log(0.236)-2*log(0.439)-2*log(0.575)-2*log(0.376)-2*log(0.025)-2*log(0.64)
-T.pval = 1-pchisq(T,m) #P-value = 0.01
+T = -2*log(0.608)-2*log(0.214)-2*log(0.236)-2*log(0.439)-2*log(0.575)-2*log(0.376)-2*log(0.025)-2*log(0.64)-2*log(0.332)
+T.pval = 1-pchisq(T,m) #P-value = 0.00464
 
@@
 anova.rma(res)
-# QM = 5.1101, p-value = 0.0238
+# QM = 6.0488, p-value = 0.0139
 
@@
 
-StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8)
-Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu")
-Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012")
-Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=207", "n=58")
-tpos <- c(24, 13, 16, 3, 17, 45, 37, 2) #obese in low BF group
-tneg <- c(62, 115, 16, 12, 26, 28, 67, 27) #nonobese in low BF group
-cpos <- c(23, 13, 22, 7, 20, 54, 41, 3) #obese in high BF group
-cneg <- c(63, 115, 9, 8, 22, 19, 62, 26) #nonobese in high BF group
-RR <- c(1.04, 1.00, 0.70, 0.43, 0.83, 0.83, 0.89, 0.67)
-low <- c(0.64, 0.48, 0.47, 0.14, 0.51, 0.66, 0.66, 0.12)
-high <- c(1.70, 2.07, 1.07, 1.35, 1.35, 1.05, 1.27, 3.70)
+StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
+Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu", "Goodrich")
+Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012", "2014")
+Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=201", "n=58", "n=507")
+tpos <- c(24, 13, 16, 3, 17, 45, 37, 2, 51) #obese in low BF group
+tneg <- c(62, 115, 16, 12, 26, 28, 67, 27, 203) #nonobese in low BF group
+cpos <- c(23, 13, 22, 7, 20, 54, 41, 3, 52) #obese in high BF group
+cneg <- c(63, 115, 9, 8, 22, 19, 62, 26, 201) #nonobese in high BF group
+RR <- c(1.04, 1.00, 0.70, 0.43, 0.83, 0.83, 0.89, 0.67, 0.98)
+low <- c(0.64, 0.48, 0.47, 0.14, 0.51, 0.66, 0.66, 0.12, 0.69)
+high <- c(1.70, 2.07, 1.07, 1.35, 1.35, 1.05, 1.27, 3.70, 1.38)
 
@@
 #
-text(5.5, 9.3, "Relative Risk [95% CI]", pos=2, cex=0.8)  #
+text(5.5, 10.3, "Relative Risk [95% CI]", pos=2, cex=0.8)  #
 #smaller scale (Plot in R studia at normal)             #
@@
 #
-text(-6.0, 9.3, "Study, Year, and n", pos=4, cex=0.8)    #
+text(-6.0, 10.3, "Study, Year, and n", pos=4, cex=0.8)    #
 #
@@
 m = 8
-T = -2*log(0.864)-2*log(1.00)-2*log(0.089)-2*log(0.121)-2*log(0.452)-2*log(0.111)-2*log(0.53)-2*log(0.64)
-T.pval = 1-pchisq(T,m) #P-value = 0.03155827
+T = -2*log(0.864)-2*log(1.00)-2*log(0.089)-2*log(0.121)-2*log(0.452)-2*log(0.111)-2*log(0.53)-2*log(0.64)-2*log(0.894)
+T.pval = 1-pchisq(T,m) #P-value = 0.02338081
 
@@
 anova.rma(res)
-  # QM = 5.3494, p-value = 0.0207
+  # QM = 4.7318, p-value = 0.0296
 
@@
 hist(COMBO.ZScore.H)
-
-#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
-HZScore <- c(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, COMBO.ZScore.H)
+#COMBO STD = 0.3489582
+keeptwinsUK <- which(data1$dataset == "twinsUK")
+twinsUKdata <- data1[keeptwinsUK, ]
+twinsUK.ZScore.H <- twinsUKdata$H.corr/0.3489582
+hist(twinsUK.ZScore.H)
+
+#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit, COMBO, twinsUK
+HZScore <- c(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, COMBO.ZScore.H, twinsUK.ZScore.H)
 data1 <- cbind(HZScore, data1)
@@
 turn.data <- cbind(turn.ZScore.H, turn.data)
-rm(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, amish.ZScore.H, turn.ZScore.H, COMBO.ZScore.H)
+rm(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, amish.ZScore.H, turn.ZScore.H, COMBO.ZScore.H, twinsUK.ZScore.H)
 rm(keepmetahit, keepCCHC, keepescobar, keepGLNE, keepHMP, keepKorean)
@@
 #Look at Obese versus non-obese
-t.test(H.Zscore ~ obese, data=group.data.test2) #P-value = 0.002583
-table(group.data.test2$obese) # No = 695 | Yes = 340
+t.test(H.Zscore ~ obese, data=group.data.test2) #P-value = 0.00493
+table(group.data.test2$obese) # No = 1099 | Yes = 443
 p1 <- ggplot(group.data.test2, aes(obese, H.Zscore)) + geom_boxplot()
@@
 #A more in-depth look, breaking down the groups
-anova(lm(H.Zscore ~ BMI.class, data=group.data.test1)) #P-value = 0.01783
+anova(lm(H.Zscore ~ BMI.class, data=group.data.test1)) #P-value = 0.03826
 a2 <- aov(H.Zscore ~ BMI.class, data=group.data.test1)
 posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-  # Obese-Lean (p-value=0.070), Over-Norm (p-value=0.791), Over-Obese (p-value=0.021)
+  # Obese-Lean (p-value=0.107), Over-Norm (p-value=0.872), Over-Obese (p-value=0.0394479)
 
 table(group.data.test1$BMI.class) 
-  # Lean = 326 | Overweight = 240 | Obese = 262
+  # Lean = 509 | Overweight = 365 | Obese = 461
 ddply(group.data.test1, ~BMI.class, summarise, mean=mean(H.Zscore), sd=sd(H.Zscore))
-  # Lean = 0.042 +\- 0.940 | Overweight = 0.097 +/- 0.993 | Obese = -0.140 +\- 1.052
+  # Lean = 0.027 +\- 0.993| Overweight = 0.059 +/- 0.982 | Obese = -0.111 +\- 1.016
 
@@
 
-summary(lm(HZScore ~ bmi, data=data1)) #P-value 0.06558, R2= 0.004977
+summary(lm(HZScore ~ bmi, data=data1)) #P-value 0.1272, R2= 0.001959
 plot(HZScore ~ bmi, data=data1)
@@
 hist(COMBO.norm.BF)
+#twinsUK SD = 0.9802656
+keeptwinsUK <- which(data2$dataset == "twinsUK")
+twinsUKdata <- data2[keeptwinsUK, ]
+twinsUK.norm.BF <- twinsUKdata$BF.ratio.corr/0.9802656
+hist(twinsUK.norm.BF)
 
 #data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
-BFScore <- c(Escobar.norm.BF, GLNE.norm.BF, CCHC.norm.BF, HMP.norm.BF, Korean.norm.BF, metahit.norm.BF, COMBO.norm.BF)
+BFScore <- c(Escobar.norm.BF, GLNE.norm.BF, CCHC.norm.BF, HMP.norm.BF, Korean.norm.BF, metahit.norm.BF, COMBO.norm.BF, twinsUK.norm.BF)
 data2 <- cbind(BFScore, data2)
@@
 turn.data <- cbind(turn.norm.BF, turn.data)
-rm(Escobar.norm.BF, GLNE.norm.BF, CCHC.norm.BF, HMP.norm.BF, Korean.norm.BF, metahit.norm.BF, amish.norm.BF, turn.norm.BF, COMBO.norm.BF)
+rm(Escobar.norm.BF, GLNE.norm.BF, CCHC.norm.BF, HMP.norm.BF, Korean.norm.BF, metahit.norm.BF, amish.norm.BF, turn.norm.BF, COMBO.norm.BF, twinsUK.norm.BF)
 rm(keepmetahit, keepCCHC, keepescobar, keepGLNE, keepHMP, keepKorean)
@@
 t.test(BF.ratio.corr ~ obese, data=group.data.test2) 
-  # P-value = 0.0525
+  # P-value = 0.07283
 table(group.data.test2$obese)
-  # No = 695 | Yes = 340
+  # No = 1099 | Yes = 443
 bf1 <- ggplot(group.data.test2, aes(obese, BF.ratio.corr)) + geom_boxplot()
@@
 anova(lm(BF.ratio.corr ~ BMI.class, data=group.data.test)) 
-  # P-value = 0.0227
+  # P-value = 0.005442
 a2 <- aov(BF.ratio.corr ~ BMI.class, data=group.data.test)
 posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-  # Obese-Norm (p-value=0.026) | Over-Norm (p-value=0.117), Over-Obese (p-value=0.863)
+  # Obese-Norm (p-value=0.016) | Over-Norm (p-value=0.017), Over-Obese (p-value=0.979)
 table(group.data.test$BMI.class)
@@
 ddply(group.data.test, ~BMI.class, summarise, mean=mean(BF.ratio.corr), sd=sd(BF.ratio.corr))
-  # Lean = -0.361 +\- 1.431 | Overweight = -0.695 +/- 2.032 | Obese = -0.787 +\- 2.461
+  # Lean = -0.392 +\- 1.319 | Overweight = -0.695 +/- 1.705 | Obese = -0.718 +\- 2.180
 bf2 <- ggplot(group.data.test, aes(BMI.class, BF.ratio.corr)) + geom_boxplot()
 
-summary(lm(BFScore ~ bmi, data=data2)) #P-value 0.07859, R2= 0.004543
+summary(lm(BFScore ~ bmi, data=data2)) #P-value 0.06589, R2= 0.002847
 plot(BFScore ~ bmi, data=data2)
@@
 t.test(Zbacter ~ obese, data=group.data.test2) 
-  # P-value = 0.3202
+  # P-value = 0.2892
 table(group.data.test2$obese)
-  # No = 695 | Yes = 340
+  # No = 1099 | Yes = 443
 bacter1 <- ggplot(group.data.test2, aes(obese, Zbacter)) + geom_boxplot()
@@
 #A more in-depth look, breaking down the groups
-anova(lm(Zbacter ~ BMI.class, data=group.data.test)) #P-value = 0.0957
+anova(lm(Zbacter ~ BMI.class, data=group.data.test)) #P-value = 0.03564
 a2 <- aov(Zbacter ~ BMI.class, data=group.data.test)
 posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-  # Obese-Norm (p-value=0.979) | Over-Norm (p-value=0.151) | Over-Obese (p-value=0.125)
+  # Obese-Norm (p-value=0.999) | Over-Norm (p-value=0.054) | Over-Obese (p-value=0.085)
 table(group.data.test$BMI.class)
-  # Lean = 326 | Overweight = 240 | Obese = 262
+  # Lean = 509 | Overweight = 461 | Obese = 365
 ddply(group.data.test, ~BMI.class, summarise, mean=mean(Zbacter), sd=sd(Zbacter))
-  # Lean = 0.041 +/- 0.966 | Overweight = -0.117 +/- 1.000 | Obese = 0.057 +/- 1.02
+  # Lean = 0.051 +/- 0.970 | Overweight = -0.097 +/- 1.000 | Obese = 0.051 +/- 1.02
 
@@
 
-summary(lm(Zbacter ~ bmi, data=data2)) #P-value 0.7168, R2= 0.0001936
+summary(lm(Zbacter ~ bmi, data=data2)) #P-value 0.8789, R2= 1.957x10-5
 plot(Zbacter ~ bmi, data=data2)
@@
 #Look at Obese versus non-obese
-t.test(Zfirm ~ obese, data=group.data.test2) #P-value = 0.8149
+t.test(Zfirm ~ obese, data=group.data.test2) #P-value = 0.8582
 table(group.data.test2$obese)
-  # No = 695 | Yes = 340
+  # No = 1099 | Yes = 443
 firm1 <- ggplot(group.data.test2, aes(obese, Zfirm)) + geom_boxplot()
@@
 anova(lm(Zfirm ~ BMI.class, data=group.data.test)) 
-  #P-value = 0.05521
+  #P-value = 0.01847
 a2 <- aov(Zfirm ~ BMI.class, data=group.data.test)
 posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-  # Obese-Norm (p-value=0.976), Over-Norm (p-value=0.063), Over-Obese (p-value=0.1252)
+  # Obese-Norm (p-value=0.812), Over-Norm (p-value=0.017), Over-Obese (p-value=0.134)
 table(group.data.test$BMI.class)
-  # Lean = 326 | Overweight = 240 | Obese = 262
+  # Lean = 509 | Overweight = 461 | Obese = 365
 ddply(group.data.test, ~BMI.class, summarise, mean=mean(Zfirm), sd=sd(Zfirm))
-  # Lean = -0.061 +/- 0.959 | Overweight = 0.130 +/- 1.018 | Obese = -0.043 +/- 1.013
+  # Lean = -0.072 +/- 0.965 | Overweight = 0.104 +/- 1.008 | Obese = -0.030 +/- 1.020
 
@@
 
-summary(lm(Zfirm ~ bmi, data=data2)) #P-value 0.8682, R2= 4.06x10-5
+summary(lm(Zfirm ~ bmi, data=data2)) #P-value 0.7604, R2= 7.835x10-5
 plot(Zfirm ~ bmi, data=data2)
@@
 #Look at Obese versus non-obese
-t.test(zbf ~ obese, data=group.data.test2) #P-value = 0.4379
+t.test(zbf ~ obese, data=group.data.test2) #P-value = 0.4516
 table(group.data.test2$obese)
-  # No = 695 | Yes = 340
+  # No = 1099 | Yes = 443
 zbf1 <- ggplot(group.data.test2, aes(obese, zbf)) + geom_boxplot()
@@
 anova(lm(zbf ~ BMI.class, data=group.data.test)) 
-  # P-value = 0.4769
+  # P-value = 0.3317
 a2 <- aov(zbf ~ BMI.class, data=group.data.test)
 posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-  # Obese-Norm (p-value=0.892) | Over-Norm (p-value=0.692) | Over-Obese (p-value=0.452)
+  # Obese-Norm (p-value=0.979) | Over-Norm (p-value=0.434) | Over-Obese (p-value=0.383)
 table(group.data.test$BMI.class)
-  # Lean = 326 | Overweight = 240 | Obese = 262
+  # Lean = 509 | Overweight = 461 | Obese = 365
 ddply(group.data.test, ~BMI.class, summarise, mean=mean(zbf), sd=sd(zbf))
-  # Lean = 0.0081 +/- 0.954 | Overweight = -0.064 +/- 0.951 | Obese = 0.046 +/- 1.084
+  # Lean = 0.0024 +/- 0.979 | Overweight = -0.055 +/- 0.988 | Obese = 0.037 +/- 1.033
 zbf2 <- ggplot(group.data.test, aes(BMI.class, zbf)) + geom_boxplot()
 
-summary(lm(ZBF ~ bmi, data=data2)) #P-value 0.7118, R2= 0.0002008
+summary(lm(ZBF ~ bmi, data=data2)) #P-value 0.8311, R2= 3.836x10-5
 plot(ZBF ~ bmi, data=data2)

### 2016-02-06 code/Goodrich.Feb62016.dataAnalysis.R

@@ -1,0 +1,265 @@
+#Goodrich et al.
+# Single study analysis
+#Feb 6th, 2016
+
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+
+setwd("C:/users/marc/Desktop/obesity2/twinsUK")
+
+#Read in and match metadata to microbiome data
+metadata <- read.csv("TwinsUKStudy_metadata.csv")
+rownames(metadata) <- metadata[, 1]
+shared.data <- read.table("combined.tx.1.subsample.shared", header=T)
+rownames(shared.data) <- shared.data[, 2]
+microbiome <- shared.data[, -c(1:3)]
+keep  <- rownames(metadata)
+
+microbiome<- microbiome[keep, ]
+
+rm(good.metadata, shared.data, keep)
+
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(microbiome)
+S <- specnumber(microbiome)
+J <- H/log(S)
+alpha.diversity <- as.data.frame(cbind(H, S, J))
+alpha.test <- alpha.diversity
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.table("phyla.txt", header=T)
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(microbiome)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- microbiome
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
[…]

### 2016-02-12 code/sequenceProcessing.Rmd
sequence processing updates

@@
 This recount starts from time of download all the way to generation of the final shared file used for the meta-analysis in the study.
-You will also need the latest version of mothur, SRA Toolkit, Metaphlan2, and Python 2.7 installed on your system.
+You will also need the latest version of mothur, SRA Toolkit, Metaphlan2, Python 2.7, and Python 3+ installed on your system.
 There are also a few reference files that you will need.
 The first is the [SILVA reference](http://www.mothur.org/wiki/Silva_reference_files) database and the second is the [RDP database](http://www.mothur.org/wiki/RDP_reference_files) of which we used release 102 and version 9 respectively.
-Finally, any entered code is denoted will be surronded with ''.
-As an example, ''print("Hello World)''.
+In general, we tried to keep close to the Schloss SOP for processsing sequence reads.
+However, where this was not possible (mostly due to a high loss of reads resulting in a high loss of samples) we defaulted to the processing used by the published authors.
+Where it was not possible to use a *de novo* approach (mostly due to high numbers of unique sequences) and the authors used a closed-reference approach we used the RDP database instead of greengenes.
+
 <br>
@@
 <br>
-''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP062/SRP062005/''
-<br>
-''fastq-dump \*.sra''
+```{r, engine='bash', eval=FALSE}
+wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP062/SRP062005/
+
+fastq-dump \*.sra
+```
 <br><br>
@@
 Before getting started enter mothur and curate the silva.fasta for the v4 region with the following commands:
-<br><br>
-''pcr.seqs(fasta=silva.bacteria.fasta, start=11894, end=25319, keepdots=F)''
-<br>
-''system(mv silva.bacteria.pcr.fasta silva.v4.fasta)''
-<br><br>
+
+```{r, engine='bash', eval=FALSE}
+pcr.seqs(fasta=silva.bacteria.fasta, start=11894, end=25319, keepdots=F)
+system(mv silva.bacteria.pcr.fasta silva.v4.fasta)
+```
 
 Next we utilized the batch mode process and can come back the next day to see everything completed.  Our command was the following:
-<br><br>
-''/share/scratch/schloss/mothur/mothur data1.batch''
-<br>
-<br>
+
+```{r, engine='bash', eval=FALSE}
+/share/scratch/schloss/mothur/mothur data1.batch
+```
+
 If you thought this was difficult to get data just wait until later on in this pipeline to see just how crazy it can get trying to get the data to a useable point.
@@
 Here we go......
-<br><br>
-''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP053/SRP053023/''
-<br>
-<br>
+
+```{r, engine='bash', eval=FALSE}
+wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP053/SRP053023/
+```
+
 Now since there was no sff file provided in this data set even though it is a 454 data set (I found this out when I tried to run the command -> sff-dump \*.sra).
@@
 <br>
-<br>
-''fastq-dump \*.sra''
-<br>
+```{r, engine='bash', eval=FALSE}
+fastq-dump \*.sra
+```
 <br>
@@
 
-''#!/bin/sh  
-for sample in \$(ls \*sra)  
+```{r, engine='sh', eval=FALSE}
+#!/bin/sh  
+for sample in $(ls *sra)  
 do  
-fastq-dump \$sample  
+fastq-dump $sample  
 done
-<br><br>
-for sample2 in \$(ls \*fastq)  
+
+for sample2 in $(ls *fastq)  
 do  
-/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=\$sample2)"  
-rm \*logfile  
+/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"  
+rm *logfile  
 done
-<br><br>
-for TRIM in \$(ls ERR\*.fasta)  
+
+for TRIM in $(ls ERR*.fasta)  
 do  
-QFILE=\${TRIM//fasta/qual}  
-/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIM, qfile=\$QFILE, maxambig=5, maxhomop=8, qwindowaverage=25, qwindowsize=50, processors=8)"  
+QFILE=${TRIM//fasta/qual}  
+/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=5, maxhomop=8, qwindowaverage=25, qwindowsize=50, processors=8)"  
 done
-rm \*.fastq''
-<br><br>
+
+rm *.fastq
+```
+
 The next lines of code are used to combine the fasta file, create a names file with the mothur unique command, and create a combined group file for subsequent processing.
-<br><br>
-''cat \*trim.fasta > combined.trim.fasta  
+
+```{r, engine='bash', eval=FALSE}
+cat \*trim.fasta > combined.trim.fasta  
 /share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"
+
 grep -e '^>' combined.trim.fasta > test.fasta  
 sed 's/>//' test.fasta > test2.fasta  
-sed 's/\\./\_/g' test2.fasta > test3.fasta  
-sed 's/_.\*//' test3.fasta > test4.fasta  
+sed 's/\./_/g' test2.fasta > test3.fasta  
+sed 's/_.*//' test3.fasta > test4.fasta  
 paste -d" " test2.fasta test4.fasta > combined.groups
-exit 0''
-<br><br>
+
+exit 0
+```
+
 After this little program is run we follow the mothur 454 SOP as written and can use the following command to complete the processing using the put together *Ross.batch* file and running mothur.
 One final difference from the standardized mothur approach that was made was during the screen.seqs command after alignment a hard cutoff *minlength=200* and *maxlength=250* was used.
-<br><br>
-''/share/scratch/schloss/mothur/mothur Ross.batch''
-<br><br>
+
+```{r, engine='bash', eval=FALSE}
+/share/scratch/schloss/mothur/mothur Ross.batch
+```
+
 That completes the processing for this data set.  Simple so far I hope.  On to the next data set.
@@
 So using wget we can download the data with the following code:
-<br><br>
-''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/ERP/ERP006/ERP006339/''
-<br><br>
+
+```{r, engine='bash', eval=FALSE}
+wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/ERP/ERP006/ERP006339/
+
+```
+
 I started off with analyzing the data separately while sticking the the default mothur SOP for MiSeq processing.....
-<br><br>
-''#!/bin/sh  
-\# Unpack sra files  
-\# mkdir sra.complete  
+
+```{r, engine='sh', eval=FALSE}
+#!/bin/sh  
+# Unpack sra files  
+# mkdir sra.complete  
 for start in \$(ls \*sra)  
 do  
-fastq-dump \$start  
-mv \$start sra.complete  
+fastq-dump $start  
+mv $start sra.complete  
 done
-<br><br>
-\# Unpack fastq files  
+
+# Unpack fastq files  
 mkdir fastq.complete  
 mkdir logfile.complete  
-for info in \$(ls \*fastq)  
+for info in $(ls *fastq)  
 do  
-/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=\$info)"  
-mv \$info fastq.complete  
-mv \*.logfile logfile.complete  
+/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$info)"  
+mv $info fastq.complete  
+mv *.logfile logfile.complete  
 done
-<br><br>
-\# Screen the fasta file that came from the fastq  
+
+# Screen the fasta file that came from the fastq  
 mkdir fasta.complete  
-for next in \$(ls \*fasta)  
+for next in $(ls *fasta)  
 do  
-/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$next, maxambig=0, maxlength=275, processors=8)"  
-mv \$next fasta.complete \# should store silva file in here and then move it working directory  
-mv \*.logfile logfile.complete  
+/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$next, maxambig=0, maxlength=275, processors=8)"  
+mv $next fasta.complete # should store silva file in here and then move it to working directory  
+mv *.logfile logfile.complete  
 done
-<br><br>
+
 mv ~/twinsUK/fasta.complete/silva.v4.fasta ~/twinsUK  
-\# Align the screened fasta file to V4 region of SILVA  
+# Align the screened fasta file to V4 region of SILVA  
 mkdir good.fasta.complete  
-for sample2 in \$(ls \*good.fasta)  
+for sample2 in $(ls *good.fasta)  
 do  
-/share/scratch/schloss/mothur/mothur "#align.seqs(fasta=\$sample2, reference=silva.v4.fasta, processors=8)"  
-mv \$sample2 good.fasta.complete  
-mv \*logfile logfile.complete  
+/share/scratch/schloss/mothur/mothur "#align.seqs(fasta=$sample2, reference=silva.v4.fasta, processors=8)"  
+mv $sample2 good.fasta.complete  
+mv *logfile logfile.complete  
 done
-<br><br>
-\#Screen the aligned sequences to specific location  
+
+#Screen the aligned sequences to specific location  
 mkdir align.complete  
-for sample3 in \$(ls \*.align)  
+for sample3 in $(ls *.align)  
 do  
-/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$sample3, start=1968, end=11546, maxhomop=8, processors=8)"  
-mv \$sample3 align.complete  
+/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$sample3, start=1968, end=11546, maxhomop=8, processors=8)"  
+mv $sample3 align.complete  
 done
-<br><br>
-\# Filter the screened sequences to get a fasta file back  
+
+# Filter the screened sequences to get a fasta file back  
 mkdir align.good.complete  
-for sample4 in \$(ls \*good.align)  
+for sample4 in $(ls *good.align)  
 do  
-/share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=\$sample4, vertical=T, trump=., processors=8)"  
-mv \$sample4 align.good.complete  
-mv \*logfile logfile.complete  
+/share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=$sample4, vertical=T, trump=., processors=8)"  
+mv $sample4 align.good.complete  
+mv *logfile logfile.complete  
 done
-<br><br>
-\# Make the combined fasta file  
+
+# Make the combined fasta file  
 mkdir filter.fasta.complete  
-cat \*filter.fasta > combined.fasta  
-mv \*filter.fasta filter.fasta.complete
-<br><br>
-\# Make the combined group file  
+cat *filter.fasta > combined.fasta  
+mv *filter.fasta filter.fasta.complete
+
+# Make the combined group file  
 grep -e '^>' combined.fasta > test.fasta  
 sed 's/>//' test.fasta > test2.fasta  
-sed 's/\\..\*\\$//' test2.fasta > test3.fasta  
-paste -d"\\t" test2.fasta test3.fasta > combined.groups''
-<br><br>
+sed 's/\..*$//' test2.fasta > test3.fasta  
+paste -d"\t" test2.fasta test3.fasta > combined.groups
+
+```
+
 After this program was run I thought perhaps naively that I could simply run a set a batch file and be done with this data set.
@@
 The code for this is below:
-<br><br>
-''/share/scratch/schloss/mothur/mothur Goodrich.batch''
-<br><br>
+
+```{r, engine='bash', eval=FALSE}
+/share/scratch/schloss/mothur/mothur Goodrich.batch
+```
+
 So now the biggest challenge that I have found so far is getting the data into a distance matrix not based on the closed-reference approach.
-The file is just too big for cluster.split so I'm going to have to find an alternative way.
+The file is just too big for cluster.split so we have had to use the closed-reference approach for this data set.
+
+```{r, engine='bash', eval=FALSE}
+/share/scratch/schloss/mothur/mothur Goodrich2.batch
+```
 
@@
 Here we go......
-<br><br>
-''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/ERP/ERP003/ERP003466/''
-<br>
-<br>
+
+```{r, engine='bash', eval=FALSE}
+wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/ERP/ERP003/ERP003466/
+```
+
 Now just like the previous data set there was no sff file provided in this data set even though it is a 454 data set (I found this out, again, when I tried to run the command -> sff-dump \*.sra).
 We have to use the fastq function of the SRA Tool kit.
-<br>
-<br>
-''fastq-dump \*.sra''
-<br>
-<br>
+
+```{r, engine='bash', eval=FALSE}
+fastq-dump *.sra
+```
+
 Because of this difference there are a few changes in this processing versus the standardized mothur 454 protocol.
@@
 
-''#!/bin/sh  
-for sample in \$(ls \*sra)  
+```{r, engine='sh', eval=FALSE}
+#!/bin/sh  
+for sample in $(ls *sra)  
 do  
-fastq-dump \$sample  
+fastq-dump $sample  
 done
 <br><br>
-for sample2 in \$(ls \*fastq)  
+for sample2 in $(ls *fastq)  
 do  
-/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=\$sample2)"  
-rm \*logfile  
+/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"  
+rm *logfile  
 done
 <br><br>
-for TRIM in \$(ls SRR\*.fasta)  
+for TRIM in $(ls SRR*.fasta)  
 do  
 QFILE=\${TRIM//fasta/qual}  
-/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIM, qfile=\$QFILE, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50, processors=8)"  
+/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50, processors=8)"  
 done  
-rm \*.fastq''
-<br><br>
+rm *.fastq
+```
+
 The next lines of code are used to combine the fasta file, create a names file with the mothur unique command, and create a combined group file for subsequent processing.
-<br><br>
-''cat \*trim.fasta > combined.trim.fasta  
+
+```{r, engine='bash', eval=FALSE}
+cat *trim.fasta > combined.trim.fasta
+
 /share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"
+
 grep -e '^>' combined.trim.fasta > test.fasta  
 sed 's/>//' test.fasta > test2.fasta  
-sed 's/\\./\_/g' test2.fasta > test3.fasta  
-sed 's/_.\*//' test3.fasta > test4.fasta  
+sed 's/\./_/g' test2.fasta > test3.fasta  
+sed 's/_.*//' test3.fasta > test4.fasta  
 paste -d" " test2.fasta test4.fasta > combined.groups  
-exit 0''
-<br><br>
+exit 0
+```
+
 After this little program is run we follow the mothur 454 SOP as written and can use the following command to complete the processing using the put together *Escobar.batch* file and running mothur.
 Again like the Ross et al data set one final difference from the standardized mothur approach that was made was during the screen.seqs command after alignment a hard cutoff *minlength=200* and *maxlength=250* was used.
-<br><br>
-''/share/scratch/schloss/mothur/mothur Escobar.batch''
-<br><br>
+
+```{r, engine='bash', eval=FALSE}
+/share/scratch/schloss/mothur/mothur Escobar.batch
+```
+
 That completes the processing for this data set.  So far so good, no major problems.  On to the next data set.
@@
 Here we go......
-<br><br>
-''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP002/SRP002465/''
-<br><br>
+```{r, engine='bash', eval=FALSE}
+wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP002/SRP002465/
+```
+
 Now this data set included the sff files so we can use sffinfo function and start from the very start of the mothur 454 pipleine.
@@
 Also you should be aware that this part of the processing will take a LONG time to get through.
-<br><br>
-''#!/bin/sh
-for sample in \$(ls \*sra)  
+
+```{r, engine='bash', eval=FALSE}
+#!/bin/bash
+
+# Make a directory to store the reference files
+mkdir reference.files
+mv silva.bacteria.fasta reference.files
+mv trainset9_032012.pds.tax reference.files
+mv trainset9_032012.pds.fasta reference.files
+
+# Read in the sra and make a fastq instead of a sff.
+mkdir completed.redo.sra
+mkdir logfile.completed
+
+for sample in $(ls *sra)
 do
-sff-dump \$sample  
+	fastq-dump $sample
+	mv $sample completed.redo.sra
+	mv *logfile logfile.completed
+
 done
-<br><br>
-for sample2 in \$(ls \*sff)  
+
+# Get the quality file and fasta information from the fastq
+mkdir completed.fastq
+
+for sample2 in $(ls *fastq)
 do
-/share/scratch/schloss/mothur/mothur "#sffinfo(sff=\$sample2, flow=T)"  
-rm \*logfile  
+	/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"	
+	mv $sample2 completed.fastq
+	mv *logfile logfile.completed
+
 done
-<br><br>
-for FLOW in \$(ls \*flow)  
+
+# Trim the fastas according to quality 
+mkdir first.trim.completed
+for Trimming in $(ls *fasta)
 do
-/share/scratch/schloss/mothur/mothur "#trim.flows(flow=\$FLOW, processors=8)"  
-rm \*logfile  
+	Quality=${Trimming//fasta/qual}
+	/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
+	mv $Trimming first.trim.completed
+	mv $Quality first.trim.completed
+	mv *logfile logfile.completed
 done
-<br><br>
-for SHHH in \$(ls \*files)  
+
+mkdir scrap.fasta.completed
+mkdir scrap.qual.completed
+mkdir trim.qual.completed
+mv *scrap.qual scrap.qual.completed
+mv * scrap.fasta.completed
+mv *trim.qual trim.qual.completed
+```
+
+Now I realized moving through this data set that they had SRR's for an individual sample so I decided to combine all these SRR's together based on their sample ID.
+I used a python script that would make this possible.
+It takes to columns seperated by a tab and looks for common names from one column and inserts a user defined command at the beginning as well as any special symbols like /* or .fasta to the end of each column as needed.
+If you want to use the program it is found in my personal repository [here](https://github.com/marcsze/pythonPrograms) called RepeatEdit.py.
+At any rate if you just want to copy and paste the command (it is long) you can see it below.
+The output of the RepeatEdit.py program will basically spit this out for you.
+
+```{r, engine='bash', eval=FALSE}
+#!/bin/bash
+
+cat SRR050640* SRR050641* > p0_s2.fasta
+cat SRR050550* SRR050551* SRR050806* SRR050807* > p0_s1.fasta
+cat SRR050558* SRR050559* SRR050814* SRR050815* SRR050732* SRR050733* > p10_s1.fasta
+cat SRR050808* SRR050809* SRR050552* SRR050553* > p100_s1.fasta
+cat SRR050724* SRR050725* > p103_s1.fasta
+cat SRR050452* SRR050453* SRR050650* SRR050651* > p115_s1.fasta
+cat SRR050456* SRR050457* > p118_s1.fasta
+cat SRR050460* SRR050461* > p120_s2.fasta
+cat SRR050556* SRR050557* SRR050812* SRR050813* > p108_s1.fasta
+cat SRR050560* SRR050561* SRR050816* SRR050817* > p111_s1.fasta
+cat SRR050642* SRR050643* > p106_s2.fasta
+cat SRR050644* SRR050645* > p112_s1.fasta
+cat SRR050648* SRR050649* > p114_s1.fasta
+cat SRR050652* SRR050653* > p119_s1.fasta
+cat SRR050728* SRR050729* > p105_s1.fasta
+cat SRR050730* SRR050731* > p105_s2.fasta
+cat SRR050554* SRR050555* SRR050810* SRR050811* > p106_s1.fasta
+cat SRR050458* SRR050459* > p120_s1.fasta
+cat SRR050464* SRR050465* > p124_s1.fasta
+cat SRR050466* SRR050467* > p125_s1.fasta
+cat SRR050470* SRR050471* > p129_s1.fasta
+cat SRR050472* SRR050473* > p130_s1.fasta
+cat SRR050476* SRR050477* > p133_s1.fasta
+cat SRR050478* SRR050479* > p134_s1.fasta
+cat SRR050482* SRR050483* > p136_s1.fasta
+cat SRR050484* SRR050485* > p137_s1.fasta
+cat SRR050488* SRR050489* > p140_s1.fasta
+cat SRR050490* SRR050491* > p142_s1.fasta
+cat SRR050494* SRR050495* > p148_s1.fasta
+cat SRR050496* SRR050497* > p149_s1.fasta
+cat SRR050500* SRR050501* > p151_s1.fasta
+cat SRR050502* SRR050503* > p152_s1.fasta
+cat SRR050522* SRR050523* > p36_s1.fasta
+cat SRR050524* SRR050525* > p39_s2.fasta
+cat SRR050528* SRR050529* > p42_s1.fasta
+cat SRR050530* SRR050531* SRR050682* SRR050683* SRR095127* SRR095128* > p44_s2.fasta
+cat SRR050564* SRR050565* SRR050820* SRR050821* > p14_s1.fasta
+cat SRR050568* SRR050569* SRR050824* SRR050825* > p17_s1.fasta
+cat SRR050572* SRR050573* SRR050828* SRR050829* > p20_s1.fasta
+cat SRR050576* SRR050577* SRR050832* SRR050833* > p3_s1.fasta
+cat SRR050580* SRR050581* SRR050764* SRR050765* SRR050836* SRR050837* > p48_s1.fasta
+cat SRR050584* SRR050585* SRR050586* SRR050587* SRR050840* SRR050841* > p4_s1.fasta
+cat SRR050660* SRR050661* > p16_s2.fasta
+cat SRR050662* SRR050663* > p18_s2.fasta
+cat SRR050736* SRR050737* > p13_s1.fasta
+cat SRR050738* SRR050739* > p13_s2.fasta
+cat SRR050538* SRR050539* > p6_s2.fasta
+cat SRR050540* SRR050541* > p72_s2.fasta
+cat SRR050582* SRR050583* SRR050766* SRR050767* SRR050838* SRR050839* > p49_s1.fasta
+cat SRR050588* SRR050589* SRR050842* SRR050843* > p50_s1.fasta
+cat SRR050592* SRR050593* SRR050846* SRR050847* > p54_s1.fasta
+cat SRR050594* SRR050595* SRR050848* SRR050849* > p55_s1.fasta
+cat SRR050598* SRR050599* SRR050852* SRR050853* > p57_s1.fasta
+cat SRR050600* SRR050601* SRR050854* SRR050855* > p61_s1.fasta
+cat SRR050604* SRR050605* SRR050858* SRR050859* > p64_s1.fasta
+cat SRR050606* SRR050607* SRR050860* SRR050861* > p6_s1.fasta
+cat SRR050610* SRR050611* SRR050864* SRR050865* > p79_s1.fasta
+cat SRR050612* SRR050613* SRR050866* SRR050867* > p7_s1.fasta
+cat SRR050616* SRR050617* SRR050870* SRR050871* > p82_s1.fasta
+cat SRR050626* SRR050627* SRR050882* SRR050883* > p8_s1.fasta
+cat SRR050666* SRR050667* > p24_s1.fasta
+cat SRR050668* SRR050669* > p25_s1.fasta
+cat SRR050672* SRR050673* > p27_s1.fasta
+cat SRR050674* SRR050675* > p37_s1.fasta
+cat SRR050678* SRR050679* > p39_s1.fasta
+cat SRR050680* SRR050681* > p43_s1.fasta
+cat SRR050602* SRR050603* SRR050686* SRR050687* SRR050856* SRR050857* > p63_s1.fasta
+cat SRR050690* SRR050691* SRR050782* SRR050783* > p69_s1.fasta
+cat SRR050694* SRR050695* > p71_s2.fasta
+cat SRR050696* SRR050697* SRR050788* SRR050789* > p72_s1.fasta
+cat SRR050700* SRR050701* > p74_s2.fasta
+cat SRR050702* SRR050703* SRR050796* SRR050797* > p77_s1.fasta
+cat SRR050706* SRR050707* > p81_s2.fasta
+cat SRR050708* SRR050709* > p82_s2.fasta
+cat SRR050740* SRR050741* > p20_s2.fasta
+cat SRR050742* SRR050743* > p22_s1.fasta
+cat SRR050746* SRR050747* > p29_s1.fasta
+cat SRR050748* SRR050749* > p29_s2.fasta
+cat SRR050752* SRR050753* > p33_s1.fasta
+cat SRR050754* SRR050755* > p38_s1.fasta
+cat SRR050758* SRR050759* > p41_s1.fasta
+cat SRR050578* SRR050579* SRR050760* SRR050761* SRR050834* SRR050835* > p46_s1.fasta
+cat SRR050772* SRR050773* > p62_s1.fasta
+cat SRR050774* SRR050775* > p62_s2.fasta
+cat SRR050778* SRR050779* > p64_s2.fasta
+cat SRR050688* SRR050689* SRR050780* SRR050781* > p68_s1.fasta
+cat SRR050698* SRR050699* SRR050790* SRR050791* > p73_s1.fasta
+cat SRR050794* SRR050795* > p75_s1.fasta
+cat SRR050570* SRR050571* SRR050826* SRR050827* > p18_s1.fasta
+cat SRR050504* SRR050505* > p153_s1.fasta
+cat SRR050506* SRR050507* > p154_s1.fasta
+cat SRR050510* SRR050511* > p156_s1.fasta
+cat SRR050512* SRR050513* > p158_s1.fasta
+cat SRR050518* SRR050519* > p161_s1.fasta
+cat SRR050534* SRR050535* > p48_s2.fasta
+cat SRR050546* SRR050547* > p87_s2.fasta
+cat SRR050548* SRR050549* > p90_s2.fasta
+cat SRR050620* SRR050621* SRR050876* SRR050877* > p86_s1.fasta
+cat SRR050622* SRR050623* SRR050800* SRR050801* SRR050878* SRR050879* > p87_s1.fasta
+cat SRR050628* SRR050629* SRR050884* SRR050885* > p90_s1.fasta
+cat SRR050630* SRR050631* SRR050886* SRR050887* > p91_s1.fasta
+cat SRR050634* SRR050635* SRR050890* SRR050891* > p95_s1.fasta
+cat SRR050636* SRR050637* SRR050892* SRR050893* > p97_s1.fasta
+cat SRR050544* SRR050545* SRR050710* SRR050711* > p85_s2.fasta
+cat SRR050716* SRR050717* > p91_s2.fasta
+cat SRR050720* SRR050721* > p98_s1.fasta
+cat SRR050722* SRR050723* > p98_s2.fasta
+cat SRR050804* SRR050805* > p94_s2.fasta
+cat SRR050608* SRR050609* SRR050862* SRR050863* > p78_s1.fasta
+cat SRR050872* SRR050873* > p83_s1.fasta
+cat SRR050624* SRR050625* SRR050880* SRR050881* > p88_s1.fasta
+cat SRR050638* SRR050639* SRR050894* SRR050895* > p99_s1.fasta
+cat SRR095005* SRR095006* SRR095053* SRR095054* SRR095125* SRR095126* > 177.fasta
+cat SRR095007* SRR095008* SRR095055* SRR095056* SRR095101* SRR095102* > 186.fasta
+cat SRR095011* SRR095012* SRR095059* SRR095060* > 211.fasta
+cat SRR095013* SRR095014* SRR095061* SRR095062* > 213.fasta
+cat SRR095017* SRR095018* SRR095065* SRR095066* > 224.fasta
+cat SRR095021* SRR095022* SRR095069* SRR095070* SRR095105* SRR095106* > 238.fasta
+cat SRR095001* SRR095002* SRR095047* SRR095048* > 102.fasta
+cat SRR095009* SRR095010* SRR095057* SRR095058* > 208.fasta
+cat SRR095095* SRR095096* > 168.fasta
+cat SRR095099* SRR095100* > 179.fasta
+cat SRR095107* SRR095108* > 259.fasta
+cat SRR095109* SRR095110* SRR095147* SRR095148* > 283.fasta
+cat SRR095133* SRR095134* > 116.fasta
+cat SRR095135* SRR095136* > 126.fasta
+cat SRR095139* SRR095140* > 162.fasta
+cat SRR095141* SRR095142* > 269.fasta
+cat SRR095145* SRR095146* > 275.fasta
+cat SRR095149* SRR095150* > 297.fasta
+cat SRR095253* SRR095254* > 128.fasta
+cat SRR095257* SRR095258* > 157.fasta
+cat SRR095261* SRR095262* > 169.fasta
+cat SRR095269* SRR095270* > 212.fasta
+cat SRR095273* SRR095274* > 226.fasta
+cat SRR095275* SRR095276* > 242.fasta
+cat SRR095369* SRR095370* SRR095456* SRR095457* > 107.fasta
+cat SRR095371* SRR095372* > 113.fasta
+cat SRR095375* SRR095376* SRR095424* SRR095425* > 11.fasta
+cat SRR095377* SRR095378* SRR095434* SRR095435* > 219.fasta
+cat SRR095381* SRR095382* SRR095482* SRR095483* > 12.fasta
+cat SRR095373* SRR095374* SRR095387* SRR095388* > 13.fasta
+cat SRR095393* SRR095394* SRR095610* SRR095611* > 150.fasta
+cat SRR095395* SRR095396* SRR095492* SRR095493* > 156.fasta
+cat SRR095401* SRR095402* SRR095458* SRR095459* > 287.fasta
+cat SRR095412* SRR095413* SRR095440* SRR095441* > 192.fasta
+cat SRR095414* SRR095415* SRR095430* SRR095431* > 217.fasta
+cat SRR095418* SRR095419* SRR095438* SRR095439* > 1.fasta
+cat SRR095420* SRR095421* SRR095598* SRR095599* > 201.fasta
+cat SRR095428* SRR095429* SRR095452* SRR095453* > 216.fasta
+cat SRR095436* SRR095437* SRR095502* SRR095503* > 230.fasta
+cat SRR095379* SRR095380* SRR095460* SRR095461* > 289.fasta
+cat SRR095442* SRR095443* SRR095466* SRR095467* > 2.fasta
+cat SRR095426* SRR095427* SRR095532* SRR095533* > 214.fasta
+cat SRR095391* SRR095392* SRR095620* SRR095621* > 14.fasta
+cat SRR095031* SRR095032* SRR095079* SRR095080* > 392.fasta
+cat SRR095033* SRR095034* SRR095081* SRR095082* SRR095121* SRR095122* > 397.fasta
+cat SRR095037* SRR095038* SRR095085* SRR095086* SRR095123* SRR095124* > 453.fasta
+cat SRR095027* SRR095028* SRR095075* SRR095076* > 327.fasta
+cat SRR095111* SRR095112* > 313.fasta
+cat SRR095115* SRR095116* > 323.fasta
+cat SRR095119* SRR095120* > 388.fasta
+cat SRR095155* SRR095156* > 333.fasta
+cat SRR095159* SRR095160* > 358.fasta
+cat SRR095161* SRR095162* > 381.fasta
+cat SRR095167* SRR095168* > 459.fasta
+cat SRR095169* SRR095170* > 46.fasta
+cat SRR095173* SRR095174* > 481.fasta
+cat SRR095175* SRR095176* SRR095177* SRR095178* SRR095179* SRR095180* SRR095181* SRR095182* SRR095183* SRR095184* > 482.fasta
+cat SRR095185* SRR095186* SRR095187* SRR095188* SRR095189* SRR095190* SRR095191* SRR095192* SRR095193* SRR095194* SRR095195* SRR095196* SRR095197* SRR095198* SRR095199* SRR095200* > 49.fasta
+cat SRR095283* SRR095284* > 354.fasta
+cat SRR095291* SRR095292* > 406.fasta
+cat SRR095293* SRR095294* > 416.fasta
+cat SRR095299* SRR095300* > 444.fasta
+cat SRR095301* SRR095302* > 44.fasta
+cat SRR095307* SRR095308* SRR095309* SRR095310* SRR095311* SRR095312* SRR095313* SRR095314* SRR095315* SRR095316* SRR095317* SRR095318* SRR095319* SRR095320* > 488.fasta
+cat SRR095321* SRR095322* SRR095323* SRR095324* SRR095325* SRR095326* SRR095327* SRR095328* SRR095329* SRR095330* SRR095331* SRR095332* SRR095333* SRR095334* SRR095335* SRR095336* SRR095337* SRR095338* SRR095339* SRR095340* SRR095341* SRR095342* SRR095343* SRR095344* SRR095345* SRR095346* SRR095347* SRR095348* SRR095349* SRR095350* SRR095351* SRR095352* SRR095353* SRR095354* SRR095355* SRR095356* SRR095357* SRR095358* SRR095359* SRR095360* SRR095361* SRR095362* > 50.fasta
+cat SRR095403* SRR095404* SRR095572* SRR095573* SRR095574* SRR095575* SRR095576* SRR095577* > 48.fasta
+cat SRR095405* SRR095406* SRR095560* SRR095561* > 473.fasta
+cat SRR095432* SRR095433* SRR095540* SRR095541* > 445.fasta
+cat SRR095446* SRR095447* SRR095514* SRR095515* > 395.fasta
+cat SRR095450* SRR095451* SRR095550* SRR095551* > 463.fasta
+cat SRR095454* SRR095455* SRR095518* SRR095519* > 401.fasta
+cat SRR095470* SRR095471* SRR095500* SRR095501* > 367.fasta
+cat SRR095448* SRR095449* SRR095472* SRR095473* > 308.fasta
+cat SRR095476* SRR095477* SRR095478* SRR095479* > 316.fasta
+cat SRR095486* SRR095487* > 346.fasta
+cat SRR095490* SRR095491* SRR095568* SRR095569* > 349.fasta
+cat SRR095494* SRR095495* SRR095548* SRR095549* > 352.fasta
+cat SRR095480* SRR095481* SRR095504* SRR095505* > 373.fasta
+cat SRR095383* SRR095384* SRR095508* SRR095509* > 376.fasta
+cat SRR095522* SRR095523* > 410.fasta
+cat SRR095524* SRR095525* > 411.fasta
+cat SRR095528* SRR095529* SRR095542* SRR095543* > 424.fasta
+cat SRR095530* SRR095531* SRR095536* SRR095537* > 427.fasta
+cat SRR095534* SRR095535* SRR095538* SRR095539* > 429.fasta
+cat SRR095546* SRR095547* SRR095590* SRR095591* > 455.fasta
+cat SRR095516* SRR095517* SRR095556* SRR095557* > 398.fasta
+cat SRR095564* SRR095565* SRR095633* SRR095634* > 475.fasta
+cat SRR095474* SRR095475* SRR095570* SRR095571* > 483.fasta
+cat SRR095562* SRR095563* SRR095600* SRR095601* > 474.fasta
+cat SRR050450* SRR050451* SRR095255* SRR095256* > p113_s1.fasta
+cat SRR050516* SRR050517* SRR095289* SRR095290* > p160_s1.fasta
+cat SRR050574* SRR050575* SRR095297* SRR095298* > p34_s1.fasta
+cat SRR095019* SRR095020* SRR095067* SRR095068* > 234.fasta
+cat SRR095039* SRR095040* SRR095087* SRR095088* > 59.fasta
+cat SRR095041* SRR095042* SRR095089* SRR095090* > 66.fasta
+cat SRR095045* SRR095046* SRR095093* SRR095094* > 84.fasta
+cat SRR050566* SRR050567* SRR095051* SRR095052* SRR095097* SRR095098* > p15_s1.fasta
+cat SRR095113* SRR095114* SRR095151* SRR095152* > p58_s1.fasta
+cat SRR095129* SRR095130* > 523.fasta
+cat SRR095165* SRR095166* > 423.fasta
+cat SRR095201* SRR095202* SRR095203* SRR095204* SRR095205* SRR095206* SRR095207* SRR095208* SRR095209* SRR095210* SRR095211* SRR095212* > 51.fasta
+cat SRR095213* SRR095214* SRR095215* SRR095216* SRR095217* SRR095218* SRR095219* SRR095220* SRR095221* SRR095222* SRR095223* SRR095224* SRR095225* SRR095226* SRR095227* SRR095228* SRR095229* SRR095230* SRR095231* SRR095232* SRR095233* SRR095234* SRR095235* SRR095236* SRR095237* SRR095238* SRR095239* SRR095240* SRR095241* SRR095242* SRR095243* SRR095244* SRR095245* SRR095246* > 52.fasta
+cat SRR095249* SRR095250* > 87.fasta
+cat SRR095265* SRR095266* > 190.fasta
+cat SRR095267* SRR095268* > p127_s1.fasta
+cat SRR095287* SRR095288* > p138_s1.fasta
+cat SRR095303* SRR095304* > p123_s1.fasta
+cat SRR095363* SRR095364* > p121_s1.fasta
+cat SRR095365* SRR095366* > 68.fasta
+cat SRR095397* SRR095398* > 160.fasta
+cat SRR095410* SRR095411* SRR095510* SRR095511* > 191.fasta
+cat SRR095444* SRR095445* > 24.fasta
+cat SRR095468* SRR095469* SRR095631* SRR095632* > 92.fasta
+cat SRR095552* SRR095553* SRR095558* SRR095559* > 464.fasta
+cat SRR095580* SRR095581* SRR095602* SRR095603* > 53.fasta
+cat SRR095582* SRR095583* SRR095584* SRR095585* SRR095586* SRR095587* SRR095594* SRR095595* > 54.fasta
+cat SRR095596* SRR095597* SRR095604* SRR095605* > 61.fasta
+cat SRR095608* SRR095609* SRR095624* SRR095625* > 83.fasta
+cat SRR095612* SRR095613* SRR095630* > 77.fasta
+cat SRR095614* SRR095615* SRR095618* SRR095619* > 7.fasta
+cat SRR095622* SRR095623* SRR095628* SRR095629* > 81.fasta
+
+mkdir trim.fasta.renamed.complete
+mv *trim.fasta trim.fasta.renamed.complete
+```
+
+So after this I need to make a group file and then combine the fastas and group files together.
+```{r, engine='bash', eval=FALSE}
+#!/bin/bash
+
+groupN="102 107 113 116 11 126 128 12 13 14 150 156 157 160 162 168 169 177 179 186 190 191 192 1 201 208 211 212 213 214 216 217 219 224 226 230 234 238 242 24 259 269 275 283 287 289 297 2 308 313 316 323 327 333 346 349 352 354 358 367 373 376 381 388 392 395 397 398 401 406 410 411 416 423 424 427 429 444 445 44 453 455 459 463 464 46 473 474 475 481 482 483 488 48 49 50 51 523 52 53 54 59 61 66 68 77 7 81 83 84 87 92 p0_s1 p0_s2 p100_s1 p103_s1 p105_s1 p105_s2 p106_s1 p106_s2 p108_s1 p10_s1 p111_s1 p112_s1 p113_s1 p114_s1 p115_s1 p118_s1 p119_s1 p120_s1 p120_s2 p121_s1 p123_s1 p124_s1 p125_s1 p127_s1 p129_s1 p130_s1 p133_s1 p134_s1 p136_s1 p137_s1 p138_s1 p13_s1 p13_s2 p140_s1 p142_s1 p148_s1 p149_s1 p14_s1 p151_s1 p152_s1 p153_s1 p154_s1 p156_s1 p158_s1 p15_s1 p160_s1 p161_s1 p16_s2 p17_s1 p18_s1 p18_s2 p20_s1 p20_s2 p22_s1 p24_s1 p25_s1 p27_s1 p29_s1 p29_s2 p33_s1 p34_s1 p36_s1 p37_s1 p38_s1 p39_s1 p39_s2 p3_s1 p41_s1 p42_s1 p43_s1 p44_s2 p46_s1 p48_s1 p48_s2 p49_s1 p4_s1 p50_s1 p54_s1 p55_s1 p57_s1 p58_s1 p61_s1 p62_s1 p62_s2 p63_s1 p64_s1 p64_s2 p68_s1 p69_s1 p6_s1 p6_s2 p71_s2 p72_s1 p72_s2 p73_s1 p74_s2 p75_s1 p77_s1 p78_s1 p79_s1 p7_s1 p81_s2 p82_s1 p82_s2 p83_s1 p85_s2 p86_s1 p87_s1 p87_s2 p88_s1 p8_s1 p90_s1 p90_s2 p91_s1 p91_s2 p94_s2 p95_s1 p97_s1 p98_s1 p98_s2 p99_s1"
+
+for V in ${groupN}
 do
-/share/scratch/schloss/mothur/mothur "#shhh.flows(file=\$SHHH, processors=8)"  
-rm \*logfile  
-done
-<br><br>
-for TRIMS in \$(ls \*shhh.fasta)  
-do  
-NAMES=\${TRIMS//fasta/names}  
-/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIMS, name=\$NAMES, maxhomop=8, maxambig=0, minlength=200, processors=8)"  
-rm \*logfile  
-done  
-exit 0''
-<br><br>
-Now, I had a few extra lines that combined the finalized files together but when I went to analyze these they gave some weird errors.
-So I needed to make a second program to sort through these errors and analyze the data separately still.
-One thing I noticed was that by using the standard workflow a large number of files ended up with 0 or very few sequences passing the quality filters.
-The other problem which I delayed dealing with in this data set was that many samples were done sequenced more than once.
-This in itself is not a problem but the sequence ID associated with these sequences from different files were the exact same and so were carbon copies.
-This wreaked havoc when I tried to combined the data together.
-But I'll deal with that problem later.
-I'm just going to continue on with the pipeline with the file seperated.
-I will however, have to deal with this before the preclustering step.
-<br><br>
-So here is the second program used to keep processing the sequences separately rather than as one combined fasta file.....
-<br><br>
-''#!/bin/sh  
-\#Align the fasta file to the SILVA database and create directories to move files into    
-mkdir good.fasta.complete  
-mkdir reference.files  
-mkdir logifle.complete  
-mv silva* reference files  
-mv trainset9* reference files  
-for sample2 in \$(ls \*.fasta)  
-do  
-/share/scratch/schloss/mothur/mothur "#align.seqs(fasta=\$sample2, reference=reference.files/silva.bacteria.fasta, flip=t, processors=8)"  
-mv \$sample2 good.fasta.complete  
-mv \*logfile logfile.complete  
-done
-<br><br>
-\#Screen the align sequences  
-mkdir align.complete  
-for sample3 in \$(ls \*.align)  
-do  
-/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$sample3, end=6453, optimize=start, criteria=95, processors=8)"  
-mv \$sample3 align.complete
-mv \*logfile logfile.complete  
-done
-<br><br>
-\#filter the screen sequences  
-mkdir align.good.complete  
-for sample4 in \$(ls \*good.align)  
-do  
-/share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=\$sample4, vertical=T, trump=., processors=8)"  
-mv \$sample4 align.good.complete  
-mv \*logfile logfile.complete  
-done  
-exit 0''
-<br><br>
-Right, now I need to pre cluster the data and I can't get away with analyzing each file separately anymore.
-However, remember from up above there is something wierd going on in that the unique sequence ID are duplicated for many samples and there were a whole bunch of samples that did not survive quality filtering.
-Since some samples had two samples others had 6 samples to be fair I decided to take the first sample from an individual that had sequences in them (so greater than 0 MB).
-Now, this was done when I was still truly a novice in using shell scripting so I did this the "hard" way (i.e.
-the following method is pretty manual with very little automation so someone can make this more automated to improve on this process).
-Essentially if a sample had sequence in it the indivdiual study ID was assigned to the fasta file and stored in a new directory.
-Those that either had 0 MB after this we carried over and the next sequence ID file was used.
-This was done until all samples had an indivdiual ID.
-In my novice approach the only thing I automated was the original copying of the sequence ID file to a individual study ID file.
-I then visually checked the size and manually sorted the files into directories.
-From this I created a new list to run through.
-The lists that were used were *combined2*, *combined3*, *combined4*, *combined5*, *combined6*, *combined7*, and *combined8* and can be found in this same repository.
-<br><br>
-As an aside I have learned a lot about coding doing this project and if I were to do this analysis again for this study I would automate this by first generated the list of sequence IDs for each respective Indivdiual and then adding an argument that if the file size is larger than 0 MB to stop and move it.
-If it was 0 MB the program would move on to the next sequence ID file for that Indivdiual.
-<br><br>
-Now that this little problem is out of the way and I have been perfectly upfront with how I dealt with this lets move on to the final part of the processing pipeline for this data set.
-<br><br>
-For this step I took the individual sample ID named fasta files and created a list which I looped with a shell script to create a group file which can then be combined along with the fasta file.
-The code I used for this was:
-<br><br>
-''#!/bin/sh  
-mkdir trim.fasta.complete2  
-\#Make loop to create a group file  
-groupN="106 107 108 113 116 11 126 128 12 13 143 14 150 156 157 160 162 163 169 172 177 181 18 190 191 192 19 1 201 202 212 214 216 217 219 225 226 230 242 24 269 26 275 283 287 289 291 292 297 299 29 2 313 323 327 333 33 346 347 349 351 352 354 358 359 361 367 373 376 388 38 392 397 398 401 406 410 411 415 416 421 423 424 427 429 444 44 453 455 459 463 464 467 46 471 473 474 475 476 481 482 483 484 488 48 49 50 51 523 52 53 54 59 66 68 6 74 77 81 84 87 92 99 p0 p105 p108 p111 p112 p113 p114 p115 p116 p117 p118 p119 p120 p121 p122 p123 p124 p125 p126 p127 p130 p132 p133 p134 p135 p136 p137 p138 p139 p13 p140 p142 p145 p148 p14 p150 p151 p152 p153 p154 p155 p156 p158 p159 p15 p160 p16 p17 p22 p23 p24 p25 p26 p33 p34 p36 p37 p38 p3 p41 p42 p43 p44 p46 p49 p4 p57 p58 p63 p68 p69 p71 p72 p73 p74 p77 p79 p7 p81 p82 p85 p86 p87 p8 p90 p91 p95 p98"
-for V in \${groupN}  
-do  
-grep -e '^>' \${V}.fasta > test.fasta  
+	grep -e '^>' ${V}.fasta > test.fasta
 	sed 's/>//' test.fasta > test2.fasta
-sed "s/$/\\t\$V/" test2.fasta > \${V}.groups  
-mv \${V}.fasta trim.fasta.complete2  
+	sed "s/$/\t$V/" test2.fasta > ${V}.groups
 	rm test.fasta
@@
 done
-<br><br>
-cd trim.fasta.complete2  
-cat \*fasta > combined.fasta  
-cat \*groups > combined.groups  
-/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.fasta)"''
-<br><br>
-Now that this has been completed we can create a batch file for the remaining commands and analyze the data altogether.
-One final comment is that for the *screen.seq* command I did not set a *maxhomop=8*.
-Due to the large loss (which you will see from the subsampling) from this data set I did not want to be picky.
-<br><br>
-''/share/scratch/schloss/mothur/mothur Zupancic.batch''
-<br><br>
+
+cat *fasta > combined.fasta
+cat *groups > combined.groups
+
+```
+
+From here things get pretty simple.  All you now need to do is run mothur with the provided batch file and everything should work....
+
+```{r, engine='bash', eval=FALSE}
+/share/scratch/schloss/mothur/mothur Zupancic.batch
+```
+
 With that you should now be able to generate the subsample shared file that was used for the downstream analysis.
@@
 After you have downloaded the files the first part of the code is:
-<br><br>
-''#!/bin/sh  
+
+```{r, engine='bash', eval=FALSE}
+#!/bin/sh  
 mkdir fna.done  
@@
 mkdir bad.accnos.done
-<br><br>
+
+# Algin the sequences
 mkdir good.fna.done  
-for sample in \$(ls \*.fna)  
+for sample in $(ls *.fna)  
 do  
 /share/scratch/schloss/mothur/mothur "#align.seqs(fasta=\$sample, reference=silva.bacteria.fasta, processors=8)"  
-mv \$sample good.fna.done  
-mv \*logfile logfiles.done  
-done''
-<br><br>
+mv $sample good.fna.done  
+mv *logfile logfiles.done  
+done
+
+```
+
 For the next code it might be best to just double check that the start and end are what is written here.
@@
 It shouldn't make a difference but.....you never know.
-<br><br>
-''#!/bin/sh  
+
+```{r, engine='bash', eval=FALSE}
+#!/bin/sh
+
+# Screen the sequences to the same start and end
 mkdir align.done  
-for sample2 in \$(ls \*.align)  
+for sample2 in $(ls *.align)  
 do  
-/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$sample2, start=8, end=2448, maxhomop=8, processors=8)"  
-mv \$sample2 align.done  
-mv \*logfile logfiles.done  
+/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$sample2, start=8, end=2448, maxhomop=8, processors=8)"  
+mv $sample2 align.done  
+mv *logfile logfiles.done  
 done
-<br><br>
+
+# Filter the sequences
 mkdir good.align.done  
@@
 done
-<br><br>
+
+# Create a groups file and combine the fasta and groups together
 mkdir completed.good.groups  
@@
 sed 's/>//' test.fasta > test2.fasta  
-sed 's/_.\*//' test2.fasta > test3.fasta  
-paste -d"\\t" test2.fasta test3.fasta > \${sample}.groups  
+sed 's/_.*//' test2.fasta > test3.fasta  
+paste -d"\t" test2.fasta test3.fasta > \${sample}.groups
+
 rm test.fasta  
@@
 mv ~/Yatsunenko.Nature/${sample}.groups ~/Yatsunenko.Nature/completed.good.groups
+
 done
-<br><br>
+```
+
 From here I thought I might be able to get away with combining the data set and running everything together.
@@
 The code below shows what was used for this part:
-<br><br>
-''#!/bin/sh  
+
+```{r, engine='bash', eval=FALSE}
+#!/bin/sh  
 cat *fasta > combined.fasta  
 cat *groups > combined.groups
+
 /share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.fasta)"    
 /share/scratch/schloss/mothur/mothur Yatsunenko.batch
+
 done
-<br><br>
+```
 
@@
 We have also provided the oligo file in this repository *korean6.oligos* that was used so that you can easily use it on this data set.
-<br><br>
-''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByRun/sra/DRR/DRR000/DRR000776
-<br><br>
+
+```{r, engine='bash', eval=FALSE}
+wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByRun/sra/DRR/DRR000/DRR000776
+
 sff-dump DRR000776.sra
-<br><br>
+
 /share/scratch/schloss/mothur/mothur Nam.batch''
-<br><br>
+
+```
+
 Compared to the previous data sets this was one of the most straight forward to actually process.
@@
 So with a little modification we can use wget again to get the data.
-<br><br>
-''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP002/SRP002424/''
-<br><br>
+
+```{r, engine='bash', eval=FALSE}
+wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP002/SRP002424/
+```
+
 From here there was a program that was run to loop through a few of the mothur functions on each individual file.
@@
 So here we go......
-<br><br>
-''#!/bin/sh  
-mkdir sra.complete  
-\#run first loop which will extract sff files  
-for sample in \$(ls \*sra)  
+
+```{r, engine='bash', eval=FALSE}
+#!/bin/bash
+
+# Make a directory to store the reference files
+mkdir reference.files
+mv silva.bacteria.fasta reference.files
+mv trainset9_032012.pds.tax reference.files
+mv trainset9_032012.pds.fasta reference.files
+
+# Read in the sra and make a fastq instead of a sff.
+mkdir completed.redo.sra
+mkdir logfile.completed
+
+for sample in $(ls *sra)
 do
-sff-dump \$sample  
-mv \$sample sra.complete
-<br><br>
-\#create new directories for logfiles and sff's that are complete.  
-mkdir logfile.complete  
-mkdir sff.complete  
-\#Extract info from sff files  
-for sample2 in \$(ls \*sff)  
+	fastq-dump $sample
+	mv $sample completed.redo.sra
+	mv *logfile logfile.completed
+
+done
+
+# Get the quality file and fasta information from the fastq
+mkdir completed.fastq
+
+for sample2 in $(ls *fastq)
 do
-/share/scratch/schloss/mothur/mothur "#sffinfo(sff=\$sample2, flow=T)"  
-mv \$sample2 sff.complete  
-mv \*logfile logfile.complete  
+	/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"	
+	mv $sample2 completed.fastq
+	mv *logfile logfile.completed
+
 done
-<br><br>
-mkdir flow.complete  
-\#Trim respective flow files  
-for FLOW in $(ls \*flow)  
+
+# Trim the fastas according to quality 
+mkdir first.trim.completed
+for Trimming in $(ls *fasta)
 do
-/share/scratch/schloss/mothur/mothur "#trim.flows(flow=\$FLOW, processors=8)"  
-mv $FLOW flow.complete  
-mv \*logfile logfile.complete  
+	Quality=${Trimming//fasta/qual}
+	/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
+	mv $Trimming first.trim.completed
+	mv $Quality first.trim.completed
+	mv *logfile logfile.completed
 done
-<br><br>
-mkdir file.complete  
-\#Make loop to run the shhh.flow and move completed sample to another directory  
-for SHHH in \$(ls \*files)  
-do  
-/share/scratch/schloss/mothur/mothur "#shhh.flows(file=\$SHHH, processors=8)"  
-mv \$SHHH file.complete  
-mv \*logfile logfile.complete  
-done
-<br><br>
-mkdir shhh.fasta.complete  
-for TRIMS in \$(ls \*shhh.fasta)  
-do  
-NAMES=\${TRIMS//fasta/names}  
-/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIMS, name=\$NAMES, maxhomop=8, minlength=200, processors=8)"  
-mv \$TRIMS shhh.fasta.complete  
-mv \*logfile logfile.complete''  
-<br><br>
+
+mkdir scrap.fasta.completed
+mkdir scrap.qual.completed
+mkdir trim.qual.completed
+mv *scrap.qual scrap.qual.completed
+mv *scrap.fasta scrap.fasta.completed
+mv *trim.qual trim.qual.completed
+
+```
+
 So for the next part I needed to make a group file so I renamed the fasta file to represent the sampleID label so that I could with more ease create the labels I needed within the group file.
 The code for it is....
-<br><br>
-''samples="SRR327637 SRR327638 SRR327639 SRR327640 SRR327641 SRR327642 SRR327643 SRR327644 SRR327645 SRR327646 SRR327647 SRR327648 SRR327649 SRR327650 SRR327651 SRR327652 SRR327653 SRR327654 SRR327655 SRR327656 SRR327657 SRR327658 SRR327659 SRR327660 SRR327661 SRR327662 SRR327663 SRR327664 SRR327665 SRR327666 SRR327667 SRR327668 SRR327669 SRR327670 SRR327671 SRR327672 SRR327673 SRR327674 SRR327675 SRR327676 SRR327677 SRR327678 SRR327679 SRR327680 SRR327681 SRR327682 SRR327683 SRR327684 SRR327685 SRR327686 SRR327687 SRR327688 SRR327689 SRR327690 SRR327691 SRR327692 SRR327693 SRR327694 SRR327695 SRR327696 SRR327697 SRR327698 SRR327699 SRR327700 SRR327701"
-<br><br>
-for V in \${samples}  
+
+```{r, engine='bash', eval=FALSE}
+#!/bin/bash
+
+groupN="SRR327637 SRR327638 SRR327639 SRR327640 SRR327641 SRR327642 SRR327643 SRR327644 SRR327645 SRR327646 SRR327647 SRR327648 SRR327649 SRR327650 SRR327651 SRR327652 SRR327653 SRR327654 SRR327655 SRR327656 SRR327657 SRR327658 SRR327659 SRR327660 SRR327661 SRR327662 SRR327663 SRR327664 SRR327665 SRR327666 SRR327667 SRR327668 SRR327669 SRR327670 SRR327671 SRR327672 SRR327673 SRR327674 SRR327675 SRR327676 SRR327677 SRR327678 SRR327679 SRR327680 SRR327681 SRR327682 SRR327683 SRR327684 SRR327685 SRR327686 SRR327687 SRR327688 SRR327689 SRR327690 SRR327691 SRR327692 SRR327693 SRR327694 SRR327695 SRR327696 SRR327697 SRR327698 SRR327699 SRR327700 SRR327701"
+
+for V in ${groupN}
 do
-grep -e '^>' \${V}.shhh.trim.good.filter.fasta > test.fasta  
+	grep -e '^>' ${V}.trim.fasta > test.fasta
 	sed 's/>//' test.fasta > test2.fasta
-sed "s/\$/\\t\$V/" test2.fasta > \${V}.groups  
-mv \${V}.shhh.trim.good.filter.fasta filter.fasta.complete  
+	sed "s/$/\t$V/" test2.fasta > ${V}.groups
 	rm test.fasta
 	rm test2.fasta
-echo \$V  
-done''
-<br><br>
-From here you can go to the individual folders and find the final version of the respective fasta, names, and groups files for every indivdiual and combine them into a single larger file.
-In essence you don't need the names file since mothur will re-generate one when you run the *unique.seqs* command.
-This was the route that I opted for since it meant slightly less typing.
-So the only files that I needed to combine are the fasta and groups files.
-<br><br>
-''cat \*trim.good.filter.fasta > combined.fasta''  
-''cat \*groups > combined.groups''
-<br><br>
+done
+
+cat *fasta > combined.fasta
+cat *groups > combined.groups
+
+```
+
 After you have done this you can run a batch file with mothur and that should generate the subsample shared file that we used for this anlaysis.
-We lose a large number of samples like in the Zupancic data set since we are subsampling to 1000.
-We choose 1000 since it is an arbitrary standard many use but also because we wanted to set a standardized lowest vaule to subsample for that we can hold every data set to.
-Anyways by running the final command you should get the files needed to all the downstream analysis.
-<br><br>
-''/share/scratch/schloss/mothur/mothur Wu.batch''
+
+```{r, engine='bash', eval=FALSE}
+/share/scratch/schloss/mothur/mothur Wu.batch
+```
 <br><br>
@@
 So here is code that was used below:
-<br><br>
-''#!/bin/sh  
+
+```{r, engine='bash', eval=FALSE}
+#!/bin/sh
+
 PATH=\$PATH:~/metaphlan2  
@@
 mkdir completed.combined.fq
+
 test_samples="MH0001 MH0002 MH0003 MH0004 MH0005 MH0006 MH0007 MH0008 
@@
 MH0080 MH0081 MH0082 MH0083 MH0084 MH0085 MH0086"
-<br><br>
-for s in \${test_samples}  
+
+for s in ${test_samples}  
 do  
-cat \${s}\*.1.fq > \${s}.combined.1.fq  
-cat \${s}\*.2.fq > \${s}.combined.2.fq  
-mv \${s}\*combined.1.fq completed.combined.fq  
-mv \${s}\*combined.2.fq completed.combined.fq  
-done''
-<br><br>
+cat ${s}*.1.fq > ${s}.combined.1.fq  
+cat ${s}*.2.fq > ${s}.combined.2.fq  
+mv ${s}*combined.1.fq completed.combined.fq  
+mv ${s}*combined.2.fq completed.combined.fq  
+done
+
+```
+
 This code was used to combine the fastq files since there are multiple files per sample.  To actually implement the metaphlan2 program for these files see below:
-<br><br>
-''for s in \${test_samples}  
+
+```{r, engine='bash', eval=FALSE}
+for s in ${test_samples}  
 do  
-metaphlan2.py /mnt/EXT/Schloss-data/msze/test/danish.obesity.metagenome/raw.data.sep/completed.combined.fq/\${s}.combined.1.fq,/mnt/EXT/Schloss-data/msze/test/danish.obesity.metagenome/raw.data.sep/completed.combined.fq/\${s}.combined.2.fq --mpa_pkl db_v20/mpa_v20_m200.pkl --bowtie2db db_v20/mpa_v20_m200 --input_type fastq --bowtie2out profiled_samples/\${s}.bowtie2.bz2 -o profiled_samples/\${s}.txt
-done''
-<br><br>
+metaphlan2.py /mnt/EXT/Schloss-data/msze/test/danish.obesity.metagenome/raw.data.sep/completed.combined.fq/${s}.combined.1.fq,/mnt/EXT/Schloss-data/msze/test/danish.obesity.metagenome/raw.data.sep/completed.combined.fq/${s}.combined.2.fq --mpa_pkl db_v20/mpa_v20_m200.pkl --bowtie2db db_v20/mpa_v20_m200 --input_type fastq --bowtie2out profiled_samples/${s}.bowtie2.bz2 -o profiled_samples/${s}.txt
+
+done
+```
+
 In order to combine all the files into a single data table the following code was run:
-<br><br>
-''python ~/metaphlan2/utils/merge\_metaphlan\_tables.py profiled\_samples/\*.txt > profiled\_samples/output/merged\_abundance\_table.txt''
-<br><br>
+```{r, engine='bash', eval=FALSE}
+python ~/metaphlan2/utils/merge_metaphlan_tables.py profiled_samples*.txt > profiled_samples/output/merged_abundance_table.txt
+```
+
 For the 16S and metaphlan2 comparison the following code was run to obtain the necessary data tables:
-<br><br>
-''#!/bin/sh  
+
+```{r, engine='bash', eval=FALSE}
+#!/bin/sh  
 PATH=\$PATH:~/metaphlan2  
 export mpa_dir=~/metaphlan2
-<br><br>
-test_samples="SRS012849 SRS012969 SRS013098 SRS013216 SRS013638 SRS013639 SRS016437 SRS016438 SRS016990 SRS017916 SRS020622 SRS021153 SRS021219 SRS022093 SRS024663 SRS043667 SRS045528 SRS045739 SRS049402 SRS049896 SRS050026 SRS053356 SRS053573 SRS053649 SRS054352 SRS055533 SRS056273 SRS058070 SRS063489 SRS063518 SRS064973 SRS074670 SRS074964 SRS075078 SRS075341 SRS075963 SRS076929 SRS077086 SRS077194 SRS077294 SRS077335 SRS077502 SRS077552 SRS077753 SRS077849 SRS078242 SRS078419 SRS078665 SRS097889 SRS098514 SRS098571 SRS098644 SRS098717 SRS098827 SRS0100021 SRS101376 SRS101433 SRS103987 SRS0104197 SRS104311 SRS104400 SRS104485 SRS105153 SRS140492 SRS140513 SRS140645 SRS142503 SRS142505 SRS142599 SRS142712 SRS142890 SRS143070 SRS143181 SRS143342 SRS143417 SRS143598 SRS143780 SRS143876 SRS143991 SRS144362 SRS144506 SRS144537 SRS145497 SRS146764 SRS146812 SRS146813 SRS147022 SRS147088 SRS147139 SRS147346 SRS147445 SRS147652 SRS147766 SRS147919 SRS148196 SRS148424 SRS148721" <br><br>
-for s in \${test_samples}  
+
+test_samples="SRS012849 SRS012969 SRS013098 SRS013216 SRS013638 SRS013639 SRS016437 SRS016438 SRS016990 SRS017916 SRS020622 SRS021153 SRS021219 SRS022093 SRS024663 SRS043667 SRS045528 SRS045739 SRS049402 SRS049896 SRS050026 SRS053356 SRS053573 SRS053649 SRS054352 SRS055533 SRS056273 SRS058070 SRS063489 SRS063518 SRS064973 SRS074670 SRS074964 SRS075078 SRS075341 SRS075963 SRS076929 SRS077086 SRS077194 SRS077294 SRS077335 SRS077502 SRS077552 SRS077753 SRS077849 SRS078242 SRS078419 SRS078665 SRS097889 SRS098514 SRS098571 SRS098644 SRS098717 SRS098827 SRS0100021 SRS101376 SRS101433 SRS103987 SRS0104197 SRS104311 SRS104400 SRS104485 SRS105153 SRS140492 SRS140513 SRS140645 SRS142503 SRS142505 SRS142599 SRS142712 SRS142890 SRS143070 SRS143181 SRS143342 SRS143417 SRS143598 SRS143780 SRS143876 SRS143991 SRS144362 SRS144506 SRS144537 SRS145497 SRS146764 SRS146812 SRS146813 SRS147022 SRS147088 SRS147139 SRS147346 SRS147445 SRS147652 SRS147766 SRS147919 SRS148196 SRS148424 SRS148721"
+
+for s in ${test_samples}  
 do  
-\#Download and unpack necessary files  
+#Download and unpack necessary files  
 wget -r -np -nd -k ftp://public-ftp.hmpdacc.org/Illumina/PHASEII/stool/${s}.tar.bz2  
-tar xvjf \${s}.tar.bz2  
-mv \${s}/\${s}.denovo\_duplicates\_marked.trimmed.1.fastq ~/metaphlan2  
-mv \${s}/\${s}.denovo\_duplicates\_marked.trimmed.2.fastq ~/metaphlan2  
-metaphlan2.py \${s}.denovo\_duplicates\_marked.trimmed.1.fastq,\${s}.denovo\_\duplicates_\marked.trimmed.2.fastq --mpa\_pkl db\_v20/mpa\_v20\_m200.pkl --bowtie2db db\_v20/mpa\_v20\_m200 --input\_type fastq --bowtie2out profiled\_samples/\${s}.bowtie2.bz2 -o profiled\_samples/\${s}.txt
-rm -r \${s}  
-rm \${s}.denovo_duplicates_marked.trimmed.1.fastq  
-rm \${s}.denovo_duplicates_marked.trimmed.2.fastq  
-done''
-<br><br>
+tar xvjf ${s}.tar.bz2  
+mv ${s}/${s}.denovo_duplicates_marked.trimmed.1.fastq ~/metaphlan2  
+mv ${s}/${s}.denovo_duplicates_marked.trimmed.2.fastq ~/metaphlan2  
+metaphlan2.py ${s}.denovo_duplicates_marked.trimmed.1.fastq,${s}.denovo_duplicates_marked.trimmed.2.fastq --mpa_pkl db_v20/mpa_v20_m200.pkl --bowtie2db db_v20/mpa_v20_m200 --input_type fastq --bowtie2out profiled_samples/${s}.bowtie2.bz2 -o profiled_samples/${s}.txt
+rm -r ${s}  
+rm ${s}.denovo_duplicates_marked.trimmed.1.fastq  
+rm ${s}.denovo_duplicates_marked.trimmed.2.fastq  
+done
+```
+
 You can then go into the directory with all the seperate data tables and combined them into one with the following command:
-<br><br>
-''python ~/metaphlan2/utils/merge\_metaphlan\_tables.py profiled\_samples/\*.txt > profiled\_samples/output/merged\_abundance\_table.txt''
-<br><br>
+
+```{r, engine='bash', eval=FALSE}
+python ~/metaphlan2/utils/merge_metaphlan_tables.py profiled_samples*.txt > profiled_samples/output/merged_abundance_table.txt
+```
+
 Now I decided, perhaps foolishly, to manually curate the combined file.
@@
 The code I used is below:
-<br><br>
-''grep -e '^>' combined.trim.fasta > test.fasta  
+
+```{r, engine='bash', eval=FALSE}
+grep -e '^>' combined.trim.fasta > test.fasta  
 sed 's/>//' test.fasta > test2.fasta    
-sed 's/\\s.\*\$//' test2.fasta > test3.fasta  
+sed 's/\s.*$//' test2.fasta > test3.fasta  
 paste -d" " test2.fasta test3.fasta > V2.groups''
-<br><br>
+```
+
 Once we have a groups file we can just create a batch file and put it into mothur to run and come back when it is completed.
-<br><br>
-''/share/scratch/schloss/mothur/mothur Turnbaugh.batch''
-<br><br>
+```{r, engine='bash', eval=FALSE}
+/share/scratch/schloss/mothur/mothur Turnbaugh.batch
+```
 

### 2016-02-12 code/sequenceProcessing.Rmd
sequence processing update again

@@
 done
-<br><br>
+
 for sample2 in $(ls *fastq)  
@@
 done
+
 rm *.fastq

### 2016-02-12 code/Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
manuscript update 2

--- 2016-02-05 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
@@
 For a few of the studies after sequence processing there was a large decrease in the number of samples available for analysis [Table 1].
-A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/marcsze/MOM-Analysis.
+A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/blob/master/code/sequenceProcessing.md.
 
@@
 <br><br>
-The third and final category involved using mostly Zscore normalizations to pool all the data together to investigate the correlation between BMI and Shannon diversity, Bacteroidetes, Firmicutes, and B/F ratio.
+The third and final category involved using  Zscore normalizations to pool all the data together to investigate the correlation between BMI and Shannon diversity, Bacteroidetes, Firmicutes, and B/F ratio.
 Using the Zscore values a two-way t-test was used to investigate if there was a difference between non-obese and obese individuals.
@@
 *******
+```{r data_loading, echo=FALSE, include=FALSE}
+
+```
+
 **Table 1:Summary Information of Studies used in Meta-analysis.**     
-![Table 1](C:\Users\marc\Desktop\obesity2\SummaryTableStudies.tif)
+```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
+library(knitr)
+
+Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "Yatsunenko", "HMP", "Nam", "Wu", "Arumugam", "Turnbaugh")
+
+Year <- c(2016, 2015, 2014, 2014, 2012, 2012, 2011, 2011, 2011, 2010, 2009)
+
+Total <- c(172, 63, 525, 30, 227, 531, 256, 18, 65, 85, 154)
+
+Processed <- c(172, 63, 507, 30, 207, "In Progress", 256, 18, 58, 85, 146)
+
+table1 <- cbind(Study, Year, Total, Processed)
+
+colnames(table1) <- c("Study", "Year", "Total n", "n After Sequence Processing")
+
+kable(table1, format = "markdown", align = 'c', padding = 2)
+
+```
 

### 2016-02-12 code/microbiomeAndobesitySupplemental.Rmd
manuscript update 2

--- 2016-02-05 code/StatisticsProcessing.Rmd
@@
 ---
-title: "Bacterial Microbiome and Obesity Meta-analysis Statistics Pipeline"
+title: "Bacterial Microbiome and Obesity Meta-analysis Supplemental"
 author: "Marc Sze"
-date: "January 29, 2016"
+date: "January 26, 2016"
 output: 
@@
     keep_md: yes
-
+bibliography: metaAnalysis.obesity.microbiome.bib
 ---
-<br>
-
-#### **Preamble**  
-The following workflow was used to generate all the statistical analysis of our meta-analysis.  It is organized by part of which there are three.  Code chunks will be displayed as such below:
-
-```{r,  eval=FALSE}
-print("Hello World!")
-x = 1 + 1
-print(x)
-```
-Data tables are provided in this repository for your use if you do not want to have to repeat the somewhat laborious sequence analysis pipeline from start to finish.
-BMI data if publicly available has been kept in this repository for your ease of reference.
-However, there are a few data sets in which the BMI data is not public and as such you will need to request this information from the respective sources if you would like to use them.
-The two data sets this mostly concerns are the Wu, et al study and the HMP study.
 <br><br>
 
-#### **Part 1: Seperate Analysis of each Study**  
-The statistical analysis will follow a general order of data loading, alpha diversity and BMI analysis, PERMANOVA of Bray-Curtis distance matrix on obese versus non-obese and BMI classifications (where available), relative risk, and then classification using AUCRF.
-<br><br>
-
-##### **Baxter et al. 2016:**  
-```{r,eval=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-# Reading in the necessary Data
-setwd("C:/Users/marc/Desktop/obesity2/data1")
-demographics <- read.csv("demographics.v2.csv")
-microbiome <- read.csv("data1.subsample.otus.csv")
-phylogenetic.info <- read.csv("data1.summary.taxonomy.csv")
-taxonomy <- read.csv("data1.taxonomy.csv")
-
-# Minor modifications and Rownames adustments of data tables
-rownames(demographics) <- demographics[,1]
-rownames(microbiome) <- microbiome[,2]
-demographics <- demographics[,-1]
-microbiome <- microbiome[,-2]
-test.samples <- rownames(demographics)
-microbiome <- microbiome[,-c(1:2)]
-
-# Create Table for phyla information
-phylogenetic.info <- phylogenetic.info[-c(1:12),]
-phyla <- which(phylogenetic.info$taxlevel == 2)
-phyla.table <- phylogenetic.info[phyla, ]
-phyla.table <- phyla.table[,-c(1:2,4:5)]
-phyla.names <- as.character(phyla.table[,1])
-phyla.table <- phyla.table[,-1]
-phyla.table <- as.data.frame(t(phyla.table))
-colnames(phyla.table) <- phyla.names
-rownames(phyla.table) <- rownames(demographics)
-
-#Add phyla together that are not very abundant and delete them from the table
-phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deferribacteres", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "Cyanobacteria_Chloroplast", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(1, 4:7, 9:11, 13, 15)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
-
-#Generate Alpha Diversity Measures and alpha diversity table
-library(vegan)
-H <- diversity(microbiome)
-S <- specnumber(microbiome)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-
-#Create age quartiles
-
-demographics$age.quartile <- with(demographics, cut(Age, breaks=quantile(Age, probs=seq(0,1, by=0.25)), include.lowest=TRUE))
-
-#Create Obese Yes/No groups
-demographics$obese[demographics$BMI.classification=="Normal" | demographics$BMI.classification=="Overweight"] <- "No"
-demographics$obese[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Yes"
-
-#Create Obese.num groups
-demographics$obese.num[demographics$obese=="No"] <- 0
-demographics$obese.num[demographics$obese=="Yes"] <- 1
-obese <- factor(demographics$obese.num)
-
-######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-
-#Create a column with obese and extreme obese as single entity
-demographics$BMIclass2[demographics$BMI.classification=="Normal"] <- "Normal"
-demographics$BMIclass2[demographics$BMI.classification=="Overweight"] <- "Overweight"                     
-demographics$BMIclass2[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Obese"
-
-##Test BMI versus alpha diversity and phyla
-
-bmi <- demographics$BMI
-bmi.class <- factor(demographics$BMI.classification)
-bmi.class2 <- factor(demographics$BMIclass2)
-
-anova(lm(H ~ obese)) #P-value=0.01643
-anova(lm(H ~ bmi.class)) #P-value=0.06
-anova(lm(H ~ bmi.class2)) #P-value=0.04471
-a2 <- aov(H ~ bmi.class2)
-TukeyHSD(x=a2, 'bmi.class2', conf.level=0.95)
-# Obese-Norm (p-value=0.201), Over-Norm (p-value=0.769), Over-Obese (p-value=0.036)
-
-anova(lm(S ~ obese)) #P-value=0.01571
-anova(lm(S ~ bmi.class)) #P-value=0.06
-anova(lm(S ~ bmi.class2)) #P-value=0.03
-a2 <- aov(S ~ bmi.class2)
-TukeyHSD(x=a2, 'bmi.class2', conf.level=0.95)
-# Obese-Norm (p-value=0.025), Over-Norm (p-value=0.546), Over-Obese (p-value=0.178)
-
-anova(lm(J ~ obese)) #P-value=0.03241
-anova(lm(J ~ bmi.class)) #P-value=0.0272
-a2 <- aov(J ~ bmi.class)
-TukeyHSD(x=a2, 'bmi.class', conf.level=0.95)
-# Obese-Norm (p-value=0.02), Over-Norm (p-value=0.047), Over-Obese (p-value=0.06)
-anova(lm(J ~ bmi.class2)) #P-value=0.02
-a2 <- aov(J ~ bmi.class2)
-TukeyHSD(x=a2, 'bmi.class2', conf.level=0.95)
-# Obese-Norm (p-value=0.630), Over-Norm (p-value=0.159), Over-Obese (p-value=0.019)
-
-##linear correlations
-H.bmi.test <- lm(H ~ bmi)
-summary(H.bmi.test)  
-#R2 = 0.01246, P-value=0.0774
-
-S.bmi.test <- lm(S ~ bmi)
-summary(S.bmi.test)
-#R2 = 0.03222, P-value=0.0105
-
-J.bmi.test <- lm(J ~ bmi)
-summary(J.bmi.test)
-#R2 = -0.0003052, P-value=0.332
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-anova(lm(bacter ~ obese)) #P-value=0.3175
-anova(lm(bacter ~ bmi.class)) #P-value=0.6457
-anova(lm(bacter ~ bmi.class2)) #P-value=0.5993
-
-anova(lm(firm ~ obese)) #P-value=0.6817
-anova(lm(firm ~ bmi.class)) #P-value=0.9293
-anova(lm(firm ~ bmi.class2)) #P-value=0.8936
-
-anova(lm(BFratio ~ obese)) #P-value=0.6305
-anova(lm(BFratio ~ bmi.class)) #P-value=0.8334
-anova(lm(BFratio ~ bmi.class2)) #P-value=0.881
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-table(demographics[,23])
-###Normal=54, Overweight=71, Obese=45, Extreme Obesity=2
-###Need to omit extreme obesity
-#Removing rows with extreme obesity
-test <- which(demographics$BMI.classification == "Extreme Obesity")
-BMI.class.demo <- demographics[-test,]
-BMI.microbiome <- microbiome[-test,]
-bmi.class.group <- factor(BMI.class.demo$BMI.classification)
-
-set.seed(3)
-adonis(BMI.microbiome ~ bmi.class.group, permutations=1000)
-#Not significant PERMANOVA = 0.2278, pseudo-F = 1.1323
-
-set.seed(3)
-adonis(microbiome ~ bmi.class, permutations=1000)
-#Not significant PERMANOVA = 0.1848, pseudo-F = 1.1368
-
-set.seed(3)
-adonis(microbiome ~ bmi.class2, permutations=1000)
-#Not significant PERMANOVA = 0.1359, pseudo-F = 1.1937
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-library(epiR)
-
-#Generate median values and put them into existing alpha.test dataframe
-
-#Shannon diversity
-median(H) # 3.67469
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-#OTU Richness
-median(S) # 198
-alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})
-
-#Evenness
-median(J) # 0.692766
-alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
-
-##Run the RR for Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-S.cat <- alpha.test$S.cat
-J.cat <- alpha.test$J.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-table(test3[c(1:86), 4])
-table(test3[c(87:172), 4])
-#Group1 (Higher than median), obese = 22 and non-obese = 64
-#Group2 (Lower than median), obese = 25 and non-obese = 61
-
-group1 <- c(22, 64)
-group2 <- c(25, 61)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 1.14
-## CI = 0.70, 1.85
-## p-value = 0.608
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-median(BFRatio$BFRatio) # 0.3621609
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-table(test4[c(1:86), 2]) #higher group
-table(test4[c(87:172), 2]) #lower group
-#Group1 (Higher than median), obese = 23 and non-obese = 63
-#Group2 (Lower than median), obese = 24 and non-obese = 62
-
-group1 <- c(23, 63)
-group2 <- c(24, 62)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 1.04
-## CI = 0.64, 1.70
-## p-value = 0.864
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-library(AUCRF)
-
-#generate test set
-  # get rid of values that only have 0 and something else
-testset <- Filter(function(x)(length(unique(x))>2), microbiome)
-  # get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), microbiome)
-
-#Need to add phyla and alpha diversity measures to the dataset
-
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-testset <- cbind(obese, testset)
-
-#Try AUCRF with default measures provided in readme
-set.seed(3)
-fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-summary(fit) # list of 14 Measures, AUCopt = 0.7553
-plot(fit)
-
-```
-<br><br>
-
-##### **Ross et al. 2015:**  
-```{r, eval=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/hispanic/")
-
-hispanic.microb <- read.table("hispanic.subsample.shared", header=T)
-
-metadata <- read.csv("s40168-015-0072-y-s1.csv")
-sample.match <- read.csv("Hispanic_dataset.csv")
-
-#Create a microbiome data table with sample names from metadata
-test <- cbind(as.character(sample.match$Run_s), as.character(sample.match$Library_Name_s))
-test <- test[order(test[, 1]), ]
-
-test2 <- cbind(test, hispanic.microb)
-rownames(test2) <- test2[, 2]
-his.microb.edit <- test2[, -c(1:5)]
-edit.metadata <- metadata[, -c(2:5)]
-rownames(edit.metadata) <- edit.metadata[, 1]
-edit.metadata <- edit.metadata[, -1]
-
-#Create a metadata file in the order of the microbiome data
-order1 <- rownames(his.microb.edit)
-edit.metadata2 <- edit.metadata[order1, ]
-
-#Get alpha diversity of the samples
-library(vegan)
-
-H <- diversity(his.microb.edit)
-S <- specnumber(his.microb.edit)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(his.microb.edit)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- his.microb.edit
-colnames(phyla.table) <- phyla.names
-  #add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-  #combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Deinococcus-Thermus", "Elusimicrobia", "Fusobacteria", "Lentisphaerae", "Synergistetes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(3:4, 6:7, 9:10)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-phyla.table.rel.abund <- as.data.frame(phyla.table.rel.abund[, -8])
-
-#Create BMI groups
-edit.metadata2$BMI.class[edit.metadata2$BMI<=24] <- "Normal"
-edit.metadata2$BMI.class[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
-edit.metadata2$BMI.class[edit.metadata2$BMI>=30 & edit.metadata2$BMI<40] <- "Obese"
-edit.metadata2$BMI.class[edit.metadata2$BMI>=40] <- "Extreme Obesity"
-
-#Create Obese Yes/No groups
-edit.metadata2$obese[edit.metadata2$BMI.class=="Normal" | edit.metadata2$BMI.class=="Overweight"] <- "No"
-edit.metadata2$obese[edit.metadata2$BMI.class=="Obese" | edit.metadata2$BMI.class=="Extreme Obesity"] <- "Yes"
-
-#Create BMI class 2 groups
-edit.metadata2$BMI.class2[edit.metadata2$BMI<=24] <- "Normal"
-edit.metadata2$BMI.class2[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
-edit.metadata2$BMI.class2[edit.metadata2$BMI>=30] <- "Obese"
-
-#Get paitent demographics to be tested
-
-bmi <- edit.metadata2$BMI
-obese <- factor(edit.metadata2$obese)
-bmi.class <- factor(edit.metadata2$BMI.class)
-bmi.class2 <- factor(edit.metadata2$BMI.class2)
-
-######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-anova(lm(H ~ obese)) #P-value=0.1657
-anova(lm(H ~ bmi.class)) #P-value=0.5448
-anova(lm(H ~ bmi.class2)) #P-value=0.3615
-
-anova(lm(S ~ obese)) #P-value=0.1841
-anova(lm(S ~ bmi.class)) #P-value=0.61
-anova(lm(S ~ bmi.class2)) #P-value=0.4
-
-anova(lm(J ~ obese)) #P-value=0.2538
-anova(lm(J ~ bmi.class)) #P-value=0.6431
-anova(lm(J ~ bmi.class2)) #P-value=0.492
-
-##linear correlations
-summary(lm(H ~ bmi)) #P-value=0.343, R2=0.01476
-summary(lm(S ~ bmi)) #P-value=0.345, R2=0.01466
-summary(lm(J ~ bmi)) #P-value=0.472, R2=0.008502
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-anova(lm(bacter ~ obese)) #P-value=0.2092
-anova(lm(bacter ~ bmi.class)) #P-value=0.5436
-anova(lm(bacter ~ bmi.class2)) #P-value=0.4374
-
-anova(lm(firm ~ obese)) #P-value=0.2569
-anova(lm(firm ~ bmi.class)) #P-value=0.1741
-anova(lm(firm ~ bmi.class2)) #P-value=0.321
-
-anova(lm(BFratio ~ obese)) #P-value=0.2508
-anova(lm(BFratio ~ bmi.class)) #P-value=0.4966
-anova(lm(BFratio ~ bmi.class2)) #P-value=0.506
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-adonis(his.microb.edit ~ obese, permutations=1000) 
-  #PERMANOVA=0.7153, pseudo-F=0.73249
-
-set.seed(3)
-adonis(his.microb.edit ~ bmi.class, permutations=1000) 
-  #PERMANOVA=0.3357, pseudo-F=1.0772
-
-set.seed(3)
-adonis(his.microb.edit ~ bmi.class2, permutations=1000) 
-  #PERMANOVA=0.3017, pseudo-F=1.1061
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-library(epiR)
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-median(H) # 2.747097
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-#OTU Richness
-median(S) # 56
-alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})
-
-#Evenness
-median(J) # 0.6819666
-alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-S.cat <- alpha.test$S.cat
-J.cat <- alpha.test$J.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-table(test3[c(1:31), 4])
-table(test3[c(32:63), 4])
-#Group1 (Higher than median), obese = 18 and non-obese = 13
-#Group2 (Lower than median), obese = 20 and non-obese = 12
-
-group1 <- c(18, 13)
-group2 <- c(20, 12)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group1", "group2")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 1.08
-## CI = 0.72, 1.61
-## p-value = 0.719
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-median(BFRatio$BFRatio) # 1.543046
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-table(test4[c(1:31), 2]) #higher group
-table(test4[c(32:63), 2]) #lower group
-#Group1 (Higher than median), obese = 22 and non-obese = 9
-#Group2 (Lower than median), obese = 16 and non-obese = 16
-
-group1 <- c(22, 9)
-group2 <- c(16, 16)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 0.70
-## CI = 0.47, 1.07
-## p-value = 0.089
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-library(AUCRF)
-
-#Create Obese.num group
-edit.metadata2$obese.num[edit.metadata2$obese=="No"] <- 0
-edit.metadata2$obese.num[edit.metadata2$obese=="Yes"] <- 1
-obese <- factor(edit.metadata2$obese.num)
-
-#generate test set
-  # get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), his.microb.edit)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-set.seed(3)
-fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-summary(fit) # list of 3 Measures, AUCopt = 0.7526
-plot(fit)
-
-```
-<br><br>
-
-##### **Goodrich et al. 2014:**  
-```{r}
-
-```
-<br><br>
-
-##### **Escobar et al. 2014:**  
-```{r, eval=FALSE}
-
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/columbian/")
-
-columbian.microb <- read.table("columbian.0.03.subsample.shared", header=T)
-metadata <- read.csv("columbian_dataset.csv")
-
-#Organize the microbiome shared data
-rownames(columbian.microb) <- columbian.microb[, 2]
-columbian.microb <- columbian.microb[, -c(1:3)]
-
-#Organize the metadata
-rownames(metadata) <- metadata[, 5]
-
-#Get only data that we are interested in
-edit.metadata <- metadata[, -c(1:7, 9, 13:15, 16:52)]
-
-#Sort metadata into the same order as the microbiome data
-order1 <- rownames(columbian.microb)
-edit.metadata2 <- edit.metadata[order1, ]
-
-#Get alpha diversity of the samples
-library(vegan)
-
-H <- diversity(columbian.microb)
-S <- specnumber(columbian.microb)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(columbian.microb)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- columbian.microb
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(4:5, 7:9)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
-
-#Create Obese Yes/No groups
-edit.metadata2$obese[edit.metadata2$description_s=="Adequate weight male stool sample" | edit.metadata2$description_s=="Overweight male stool sample" | edit.metadata2$description_s=="Adequate weight female stool sample" | edit.metadata2$description_s=="Overweight female stool sample"] <- "No"
-edit.metadata2$obese[edit.metadata2$description_s=="Obese male stool sample" | edit.metadata2$description_s=="Obese female stool sample"] <- "Yes"
-
-#Create BMI groups
-edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s<=24] <- "Normal"
-edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
-edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=30 & edit.metadata2$body_mass_index_s<40] <- "Obese"
-edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=40] <- "Extreme Obesity"
-
-#Create BMI groups2
-edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s<=24] <- "Normal"
-edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
-edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>=30] <- "Obese"
-
-#Get paitent demographics data to be tested 
-bmi <- edit.metadata2$body_mass_index_s
-obese <- factor(edit.metadata2$obese)
-bmi.class <- factor(edit.metadata2$BMI.class)
-bmi.class2 <- factor(edit.metadata2$BMI.class2)
-
-######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-anova(lm(H ~ obese)) #P-value=0.8636
-anova(lm(H ~ bmi.class)) #P-value=0.9584
-anova(lm(H ~ bmi.class2)) #P-value=0.9584
-
-anova(lm(S ~ obese)) #P-value=0.1947
-anova(lm(S ~ bmi.class)) #P-value=0.1432
-anova(lm(S ~ bmi.class2)) #P-value=0.1432
-
-anova(lm(J ~ obese)) #P-value=0.4762
-anova(lm(J ~ bmi.class)) #P-value=0.764
-anova(lm(J ~ bmi.class2)) #P-value=0.764
-
-##linear correlations
-summary(lm(H ~ bmi)) #P-value=0.7895, R2=0.002587
-summary(lm(S ~ bmi)) #P-value=0.4474, R2=0.02076
-summary(lm(J ~ bmi)) #P-value=0.9794, R2=2.423x10-5
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-anova(lm(bacter ~ obese)) #P-value=0.09253
-anova(lm(bacter ~ bmi.class)) #P-value=0.1742
-anova(lm(bacter ~ bmi.class2)) #P-value=0.1742
-
-anova(lm(firm ~ obese)) #P-value=0.3602
-anova(lm(firm ~ bmi.class)) #P-value=0.5409
-anova(lm(firm ~ bmi.class2)) #P-value=0.5409
-
-anova(lm(BFratio ~ obese)) #P-value=0.4663
-anova(lm(BFratio ~ bmi.class)) #P-value=0.7545
-anova(lm(BFratio ~ bmi.class2)) #P-value=0.7545
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-adonis(columbian.microb ~ obese, permutations=1000) 
-#PERMANOVA=0.08891, pseudo-F=1.3797
-
-set.seed(3)
-adonis(columbian.microb ~ bmi.class, permutations=1000) 
-  #PERMANOVA=0.2358, pseudo-F=1.1142
-
-set.seed(3)
-adonis(columbian.microb ~ bmi.class2, permutations=1000) 
-  #PERMANOVA=0.2358, pseudo-F=1.1142
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-library(epiR)
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-median(H) # 4.080262
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-#OTU Richness
-median(S) # 362
-alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})
-
-#Evenness
-median(J) # 0.6896595
-alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-S.cat <- alpha.test$S.cat
-J.cat <- alpha.test$J.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-table(test3[c(1:15), 4])
-table(test3[c(16:30), 4])
-#Group1 (Higher than median), obese = 4 and non-obese = 11
-#Group2 (Lower than median), obese = 6 and non-obese = 9
-
-group1 <- c(4, 11)
-group2 <- c(6, 9)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group1", "group2")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 1.50
-## CI = 0.53, 4.26
-## p-value = 0.439
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-median(BFRatio$BFRatio) # 0.2516729
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-table(test4[c(1:15), 2]) #higher group
-table(test4[c(16:30), 2]) #lower group
-#Group1 (Higher than median), obese = 7 and non-obese = 8
-#Group2 (Lower than median), obese = 3 and non-obese = 12
-
-group1 <- c(7, 8)
-group2 <- c(3, 12)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-library(epiR)
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 0.43
-## CI = 0.14, 1.35
-## p-value = 0.121
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-library(AUCRF)
-
-#Create Obese.num group
-edit.metadata2$obese.num[edit.metadata2$obese=="No"] <- 0
-edit.metadata2$obese.num[edit.metadata2$obese=="Yes"] <- 1
-obese <- factor(edit.metadata2$obese.num)
-
-#generate test set
-  # get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), columbian.microb)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-set.seed(3)
-fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-summary(fit) # list of 3 Measures, AUCopt = 0.95
-plot(fit)
-
-```
-
-##### **Zupancic et al. 2012:**
-
-```{r, eval=FALSE}
-
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/Amish/updatedGOOD")
-
-#Read in and match metadata to microbiome data
-metadata <- read.csv("amish_obesity_table2.csv")
-test <- metadata[!duplicated(metadata$submitted_sample_id_s), ]
-rownames(test) <- test[, 4]
-shared.data <- read.table("combined.0.03.subsample.shared", header=T)
-rownames(shared.data) <- shared.data[, 2]
-shared.data <- shared.data[, -c(1:3)]
-keep  <- rownames(shared.data)
-
-test2 <- test[keep, ]
-good.metadata <- test2
-test3 <- test2[!duplicated(test2$submitted_subject_id_s), ]
-keep <- rownames(test3)
-microbiome <- shared.data[keep, ]
-metadata <- test3
-
-rm(good.metadata, shared.data, keep, test, test2, test3)
-
-#generate alpha diversity measures with vegan
-library(vegan)
-H <- diversity(microbiome)
-S <- specnumber(microbiome)
-J <- H/log(S)
-select.alpha.diversity <- as.data.frame(cbind(H, S, J))
-s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
-alpha.test <- s1.alpha.diversity
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(microbiome)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- microbiome
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Spirochaetes")], 1, sum)
-phyla.table <- phyla.table[, -c(4, 6)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:6)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-
-#Create Obese Yes/No groups
-# One problem is that we are assuming those with "not provided" are defaulting to the not affected group
-metadata$obese[metadata$subject_is_affected_s=="<not provided>" | metadata$subject_is_affected_s=="No"] <- "No"
-metadata$obese[metadata$subject_is_affected_s=="Yes"] <- "Yes"
-
-#Create Obese.num group
-metadata$obese.num[metadata$obese=="No"] <- 0
-metadata$obese.num[metadata$obese=="Yes"] <- 1
-
-#create groups to be used
-obese <- factor(metadata$obese)
-
-####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-anova(lm(H ~ obese)) #P-value=0.2034
-anova(lm(S ~ obese)) #P-value=0.03516
-anova(lm(J ~ obese)) #P-value=0.5369
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-anova(lm(bacter ~ obese)) #P-value=0.9582
-anova(lm(firm ~ obese)) #P-value=0.2654
-anova(lm(BFratio ~ obese)) #P-value=0.9899
-
-####################################################################################### NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-adonis(microbiome ~ obese, permutations=1000) 
-#PERMANOVA=0.5734, pseudo-F=0.95626
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-library(epiR)
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-median(H) # 3.073974
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-#OTU Richness
-median(S) # 93
-alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})
-
-#Evenness
-median(J) # 0.6790902
-alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-S.cat <- alpha.test$S.cat
-J.cat <- alpha.test$J.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-table(test3[c(1:103), 4])
-table(test3[c(104:207), 4])
-#Group1 (Higher than median), obese = 31 and non-obese = 72
-#Group2 (Lower than median), obese = 47 and non-obese = 57
-
-group1 <- c(31, 72)
-group2 <- c(47, 57)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 1.50
-## CI = 1.04, 2.16
-## p-value = 0.025
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-median(BFRatio$BFRatio) # 0.5414258
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-table(test4[c(1:103), 2]) #higher group
-table(test4[c(104:207), 2]) #lower group
-#Group1 (Higher than median), obese = 41 and non-obese = 62
-#Group2 (Lower than median), obese = 37 and non-obese = 67
-
-group1 <- c(41, 62)
-group2 <- c(37, 67)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 0.89
-## CI = 0.66, 1.27
-## p-value = 0.53
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-library(AUCRF)
-
-#Create Obese.num group
-obese <- factor(metadata$obese.num)
-
-#generate test set
-# get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), microbiome)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-set.seed(3)
-fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-summary(fit) # list of 42 Measures, AUCopt = 0.6365534
-plot(fit)
-```
-
-##### **Yatsunenko et al. 2012:**  
-```{r}
-
-```
-
-##### **Human Microbiome Project (HMP) 2011:**  
-```{r, eval=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-##Read in relevant data
-
-setwd("C:/users/marc/Desktop/obesity2/HMP.analysis")
-microbiome <- read.table("Stool.an.0.03.subsample.shared", header=T)
-meta.cat <- read.table("categorical.metadata", header=T)
-meta.cont <- read.table("continuous.metadata", header=T)
-
-#Only interested in first visit microbiome data
-#need to create microbiome data set for only that
-
-test <- microbiome[grep("\\.01\\.", microbiome$Group), ] #selecting by .01.
-
-microb.rownames <- gsub("([0-9]+).*", "\\1", test$Group) #extracting only numeric before first "."
-rownames(test) <- microb.rownames
-
-#Get rid of information not used in downstram analysis
-test <- test[, -c(1:3)]
-
-#Subset data to the microbiome total n
-select.meta.cat <- meta.cat[microb.rownames, ]
-select.meta.cont <- meta.cont[microb.rownames, ]
-
-#generate alpha diversity measures with vegan
-library(vegan)
-H <- diversity(test)
-S <- specnumber(test)
-J <- H/log(S)
-select.alpha.diversity <- as.data.frame(cbind(H, S, J))
-alpha.test <- select.alpha.diversity
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(test)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- test
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(1, 4, 6, 7, 9, 10:12)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
-
-#Create Obese Yes/No groups
-select.meta.cat$obese[select.meta.cat$BMI_C=="normal" | select.meta.cat$BMI_C=="overweight"] <- "No"
-select.meta.cat$obese[select.meta.cat$BMI_C=="obese" | select.meta.cat$BMI_C=="extreme obesity"] <- "Yes"
-
-#Create BMI groups2
-select.meta.cat$BMI.class2[select.meta.cont$DTPBMI<=24] <- "Normal"
-select.meta.cat$BMI.class2[select.meta.cont$DTPBMI>24 & select.meta.cont$DTPBMI<30] <- "Overweight"
-select.meta.cat$BMI.class2[select.meta.cont$DTPBMI>=30] <- "Obese"
-
-#Get paitent demographics data to be tested 
-bmi <- select.meta.cont$DTPBMI
-obese <- factor(select.meta.cat$obese)
-bmi.class <- factor(select.meta.cat$BMI_C)
-bmi.class2 <- factor(select.meta.cat$BMI.class2)
-
-######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-anova(lm(H ~ obese)) #P-value=0.585
-anova(lm(H ~ bmi.class)) #P-value=0.322
-anova(lm(H ~ bmi.class2)) #P-value=0.322
-
-anova(lm(S ~ obese)) #P-value=0.845
-anova(lm(S ~ bmi.class)) #P-value=0.2051
-anova(lm(S ~ bmi.class2)) #P-value=0.2051
-
-anova(lm(J ~ obese)) #P-value=0.3559
-anova(lm(J ~ bmi.class)) #P-value=0.3501
-anova(lm(J ~ bmi.class2)) #P-value=0.3501
-
-##linear correlations
-summary(lm(H ~ bmi)) #P-value=0.7041, R2=0.0005689
-summary(lm(S ~ bmi)) #P-value=0.2304, R2=0.005657
-summary(lm(J ~ bmi)) #P-value=0.962, R2=8.932x10-6
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-anova(lm(bacter ~ obese)) #P-value=0.3518
-anova(lm(bacter ~ bmi.class)) #P-value=0.1343
-anova(lm(bacter ~ bmi.class2)) #P-value=0.1343
-
-anova(lm(firm ~ obese)) #P-value=0.636
-anova(lm(firm ~ bmi.class)) #P-value=0.07813
-anova(lm(firm ~ bmi.class2)) #P-value=0.07813
-
-anova(lm(BFratio ~ obese)) #P-value=0.5688
-anova(lm(BFratio ~ bmi.class)) #P-value=0.6807
-anova(lm(BFratio ~ bmi.class2)) #P-value=0.6807
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-adonis(test ~ obese, permutations=1000) 
-#PERMANOVA=0.8112, pseudo-F=0.7024
-
-set.seed(3)
-adonis(test ~ bmi.class, permutations=1000) 
-  #PERMANOVA=0.8022, pseudo-F=0.77896
-
-set.seed(3)
-adonis(test ~ bmi.class2, permutations=1000) 
-  #PERMANOVA=0.8022, pseudo-F=77896
-
-######################################################################################## Relative Risk ###############################################
-###########################################################################
-
-library(epiR)
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-median(H) # 2.541824
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-#OTU Richness
-median(S) # 57
-alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})
-
-#Evenness
-median(J) # 0.6305742
-alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-S.cat <- alpha.test$S.cat
-J.cat <- alpha.test$J.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-table(test3[c(1:128), 4])
-table(test3[c(129:256), 4])
-#Group1 (Higher than median), obese = 10 and non-obese = 118
-#Group2 (Lower than median), obese = 16 and non-obese = 112
-
-group1 <- c(10, 118)
-group2 <- c(16, 112)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group1", "group2")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 1.60
-## CI = 0.75, 3.39
-## p-value = 0.214
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-median(BFRatio$BFRatio) # 2.529168
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-table(test4[c(1:128), 2]) #higher group
-table(test4[c(129:256), 2]) #lower group
-#Group1 (Higher than median), obese = 13 and non-obese = 115
-#Group2 (Lower than median), obese = 13 and non-obese = 115
-
-group1 <- c(13, 115)
-group2 <- c(13, 115)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-library(epiR)
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 1.00
-## CI = 0.48, 2.07
-## p-value = 1.00
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-library(AUCRF)
-
-#Create Obese.num group
-select.meta.cat$obese.num[select.meta.cat$obese=="No"] <- 0
-select.meta.cat$obese.num[select.meta.cat$obese=="Yes"] <- 1
-obese <- factor(select.meta.cat$obese.num)
-
-#generate test set
-  # get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), test)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-set.seed(3)
-fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-summary(fit) # list of 6 Measures, AUCopt = 0.7031773
-plot(fit)
-
-```
-
-##### **Nam, et al. 2011:**  
-<br>
-It should be noted that for the grouping part of this analysis it was done in relationship to being overweight instead of obese.
-But for the combined portion of the analysis or third arm (all data sets taken together) it was used.
-Conclusions on the other arms are not dependent on this specific data set.
-```{r, eval=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/korean/")
-
-microb <- read.table("DRR000776.0.03.subsample.shared", header=T)
-rownames(microb) <- microb[, 2]
-microb <- microb[,-c(1:3)]
-
-metadata <- read.csv("korean.metadata.csv")
-
-#Need to get rid of the repeated samples and the children
-  #First the metadata
-metadata2 <- metadata[-c(2:3, 5:6, 8:9, 11:12, 14:15, 17:18, 19:24), ]
-rownames(metadata2) <- metadata2[, 2]
-metadata2 <- metadata2[, -c(1:2)]
-
-  #Second the microbiome data set
-microb2 <- microb[-c(2:3, 5:6, 8:9, 11:12, 14:15, 17:18, 19:24), ]
-rownames(microb2) <- rownames(metadata2)
-
-#Get alpha diversity of the samples
-library(vegan)
-
-H <- diversity(microb2)
-S <- specnumber(microb2)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-# Seems to look okay....
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(microb2)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- microb2
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Synergistetes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(4:5, 7:8)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
-
-#Create BMI groups
-metadata2$BMI.class[metadata2$BMI<=24] <- "Normal"
-metadata2$BMI.class[metadata2$BMI>24 & metadata2$BMI<30] <- "Overweight"
-metadata2$BMI.class[metadata2$BMI>=30 & metadata2$BMI<40] <- "Obese"
-metadata2$BMI.class[metadata2$BMI>=40] <- "Extreme Obesity"
-
-#Create Overweight Yes/No groups
-metadata2$overweight[metadata2$BMI.class=="Normal"] <- "No"
-metadata2$overweight[metadata2$BMI.class=="Obese" | metadata2$BMI.class=="Extreme Obesity" | metadata2$BMI.class=="Overweight"] <- "Yes"
-
-#Get paitent demographics data to be tested 
-bmi <- metadata2$BMI
-overweight <- factor(metadata2$overweight)
-
-######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-anova(lm(H ~ overweight)) #P-value=0.961
-anova(lm(S ~ overweight)) #P-value=0.5165
-anova(lm(J ~ overweight)) #P-value=0.6547
-
-##linear correlations
-summary(lm(H ~ bmi)) #P-value=0.8838, R2=0.001376
-summary(lm(S ~ bmi)) #P-value=0.3153, R2=0.06293
-summary(lm(J ~ bmi)) #P-value=0.7022, R2=0.009386
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-anova(lm(bacter ~ overweight)) #P-value=0.4382
-anova(lm(firm ~ overweight)) #P-value=0.2609
-anova(lm(BFratio ~ overweight)) #P-value=0.4089
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-adonis(microb2 ~ overweight, permutations=1000) 
-#PERMANOVA=0.4236, pseudo-F=0.97425
-
-######################################################################################## Relative Risk ###############################################
-###########################################################################
-
-library(epiR)
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-median(H) # 3.888484
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-#OTU Richness
-median(S) # 156.5
-alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})
-
-#Evenness
-median(J) # 0.7715805
-alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-S.cat <- alpha.test$S.cat
-J.cat <- alpha.test$J.cat
-bmi.cat <- as.character(overweight)
-test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-table(test3[c(1:9), 4])
-table(test3[c(10:18), 4])
-#Group1 (Higher than median), overweight = 4 and normal = 5
-#Group2 (Lower than median), overweight = 2 and normal = 7
-
-group1 <- c(4, 5)
-group2 <- c(2, 7)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Overweight", "normal")
-rownames(r.test) <- c("group1", "group2")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 0.50
-## CI = 0.12, 2.08
-## p-value = 0.317
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-median(BFRatio$BFRatio) # 0.4243408
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, overweight)
-test4 <- test4[order(BFRatio.cat), ]
-table(test4[c(1:9), 2]) #higher group
-table(test4[c(10:18), 2]) #lower group
-#Group1 (Higher than median), overweight = 2 and normal = 7
-#Group2 (Lower than median), overweight = 4 and normal = 5
-
-group1 <- c(2, 7)
-group2 <- c(4, 5)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-library(epiR)
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 2.00
-## CI = 0.48, 8.31
-## p-value = 0.317
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-library(AUCRF)
-
-#Create Obese.num group
-metadata2$over.num[metadata2$overweight=="No"] <- 0
-metadata2$over.num[metadata2$overweight=="Yes"] <- 1
-overweight <- factor(metadata2$over.num)
-
-#generate test set
-  # get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), microb2)
-testset <- cbind(overweight, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-set.seed(3)
-fit <- AUCRF(obese ~ ., data=testset, ntree=500, nodesize=1)
-  #Since this data set was so small had to modify the standardized        
-  #parameters.
-summary(fit) # list of 2 Measures, AUCopt = 1.00
-plot(fit)
-
-```
-
-##### **Wu, et al. 2011:**
-
-```{r, eval=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/COMBO/")
-
-microbiome <- read.table("combined.good.subsample.shared", header=T)
-rownames(microbiome) <- microbiome$Group
-microbiome <- microbiome[, -c(1:3)]
-seqData <- read.csv("COMBO_data_table.csv", header=T)
-seqData <- seqData[-1, ]
-rownames(seqData) <- seqData$Run_s
-metadata <- read.table("BMI_and_Counts.txt", header=T)
-metadata <- metadata[, c(1, 2)]
-
-#get seqData set in line with microbiome data
-namesToKeep <- rownames(microbiome)
-test <- seqData[namesToKeep, ]
-seqData <- test
-rm(test)
-
-#change all rownames to match sample IDs
-rownames(seqData) <- seqData$submitted_subject_id_s
-rownames(microbiome) <- rownames(seqData)
-rownames(metadata) <- metadata[, 1]
-
-#Match the metadata now with the microbiome data
-namesToKeep <- rownames(microbiome)
-test <- metadata[namesToKeep, ]
-metadata <- test
-rm(test)
-rm(seqData, keep, namesToKeep)
-
-#Get alpha diversity of the samples
-library(vegan)
-
-H <- diversity(microbiome)
-S <- specnumber(microbiome)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.table("combined.good.phyla.txt", header=T)
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(microbiome)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- microbiome
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Synergistetes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(4, 6:7)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:6)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-rm(keep, phyla.names, phyla.total, phylogenetic.info, phyla.table, phyla.good)
-
-#Create BMI groups
-metadata$BMI.class[metadata$BMI<=24] <- "Normal"
-metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
-metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
-metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
-
-#Create Obese Yes/No groups
-metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
-metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
-
-#Create a column with obese and extreme obese as single entity
-metadata$BMIclass2[metadata$BMI.class=="Normal"] <- "Normal"
-metadata$BMIclass2[metadata$BMI.class=="Overweight"] <- "Overweight"                     
-metadata$BMIclass2[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Obese"
-
-#Get paitent demographics to be tested
-bmi <- metadata$BMI
-obese <- factor(metadata$obese)
-bmi.class <- factor(metadata$BMI.class)
-bmi.class2 <- factor(metadata$BMIclass2)
-
-###########################################################################
-######### First Level Analysis & Alpha Diversity with BMI #################
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-anova(lm(H ~ obese)) #P-value=0.7493
-anova(lm(H ~ bmi.class)) #P-value=0.2959
-anova(lm(H ~ bmi.class2)) #P-value=0.2089
-
-anova(lm(S ~ obese)) #P-value=0.7511
-anova(lm(S ~ bmi.class)) #P-value=0.5128
-anova(lm(S ~ bmi.class2)) #P-value=0.3891
-
-anova(lm(J ~ obese)) #P-value=0.5557
-anova(lm(J ~ bmi.class)) #P-value=0.2647
-anova(lm(J ~ bmi.class2)) #P-value=0.1818
-
-##linear correlations
-summary(lm(H ~ bmi)) #P-value=0.8276, R2=0.0008541
-summary(lm(S ~ bmi)) #P-value=0.8195, R2=0.0009377
-summary(lm(J ~ bmi)) #P-value=0.8521, R2=0.0006262
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-anova(lm(bacter ~ obese)) #P-value=0.9173
-anova(lm(bacter ~ bmi.class)) #P-value=0.5554
-anova(lm(bacter ~ bmi.class2)) #P-value=0.6846
-
-anova(lm(firm ~ obese)) #P-value=0.5765
-anova(lm(firm ~ bmi.class)) #P-value=0.2157
-anova(lm(firm ~ bmi.class2)) #P-value=0.2817
-
-anova(lm(BFratio ~ obese)) #P-value=0.6735
-anova(lm(BFratio ~ bmi.class)) #P-value=0.6208
-anova(lm(BFratio ~ bmi.class2)) #P-value=0.4681
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-adonis(microbiome ~ obese, permutations=1000) 
-#PERMANOVA=0.8921, pseudo-F=0.66065
-
-set.seed(3)
-adonis(microbiome ~ bmi.class, permutations=1000) 
-#PERMANOVA=0.9321, pseudo-F=0.73921
-
-set.seed(3)
-adonis(microbiome ~ bmi.class2, permutations=1000) 
-#PERMANOVA=0.75922, pseudo-F=0.8791
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-library(epiR)
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-median(H) # 3.000655
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-#OTU Richness
-median(S) # 98.5
-alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})
-
-#Evenness
-median(J) # 0.6432727
-alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-S.cat <- alpha.test$S.cat
-J.cat <- alpha.test$J.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-table(test3[c(1:29), 4])
-table(test3[c(30:58), 4])
-#Group1 (Higher than median), obese = 3 and non-obese = 26
-#Group2 (Lower than median), obese = 2 and non-obese = 27
-
-group1 <- c(3, 26)
-group2 <- c(2, 27)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 0.67
-## CI = 0.12, 3.70
-## p-value = 0.64
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-median(BFRatio$BFRatio) # 2.438335
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-table(test4[c(1:29), 2]) #higher group
-table(test4[c(30:58), 2]) #lower group
-#Group1 (Higher than median), obese = 3 and non-obese = 26
-#Group2 (Lower than median), obese = 2 and non-obese = 27
-
-group1 <- c(3, 26)
-group2 <- c(2, 27)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 0.67
-## CI = 0.12, 3.70
-## p-value = 0.64
-
-###########################################################################
-############ Classification using AUCRF####################################
-###########################################################################
-
-library(AUCRF)
-
-#Create Obese.num group
-metadata$obese.num[metadata$obese=="No"] <- 0
-metadata$obese.num[metadata$obese=="Yes"] <- 1
-obese <- factor(metadata$obese.num)
-
-#generate test set
-# get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), microbiome)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-set.seed(3)
-fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-summary(fit) # list of 15 Measures, AUCopt = 0.7075472
-plot(fit)
-
-```
-
-##### **Arumugam, et al. 2010 (MetaHit):**  
-<br>
-This first R analysis chunk looks at how similar metaphlan2 is with 16S sequencing in the HMP data set.
-```{r, eval=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-#Phyla first then will explore other measures if data is promising
-
+```{r data_table_loading, include=FALSE}
 setwd("C:/Users/marc/Desktop/obesity2/HMP.analysis")
@@
 
-###########################################################################
-############ Let's Get to the Actual Comparison ###########################
-###########################################################################
-
 # Can start off with a general bar plot of the two side by side 
@@
 library(plotrix)
+library(ggExtra)
+library(BlandAltmanLeh)
 
@@
 
+```
+
+**Figure S1:Comparison of Phyla between 16S Sequencing and Metaphlan2.**
+
+```{r Figure_S1, echo=FALSE}
 testset <- rbind(phyla16SData, phylaMetaphlanData)
@@
 bpg2 <- bpg1 + scale_fill_brewer(palette="Set1") + scale_y_continuous(label=percent)
-bpg2
+bpg3 <- bpg2 + theme(axis.text=element_text(size=15), axis.title=element_text(size=20, face="bold"))
+bpg4 <- bpg3 + xlab("Analysis Used") + ylab("Mean Relative Abundance (%)")
+bpg5 <- bpg4 + guides(fill=guide_legend(title="Phyla")) + theme(legend.title = element_text(size=16, face="bold"))
+bpg6 <- bpg5 + theme(axis.line=element_line(colour="black"), 
+                 panel.grid.major = element_blank(), 
+                 panel.grid.minor = element_blank(), 
+                 panel.border = element_blank(), 
+                 panel.background = element_blank())
+bpg7 <- bpg6 + scale_x_discrete(breaks = 1:2, labels=c("16S", "Metaphlan2"))
 
-# Now can look at direct linear correlation between same sample
-# Between metaphlan2 and 16S amplicon sequencing
+bpg7
+```
 
-#Actinobacteria
-summary(lm(phyla16SData$Actinobacteria ~ phylaMetaphlanData$Actinobacteria)) 
-#P-value 0.5543, R2= 0.006899
-plot(phyla16SData$Actinobacteria ~ phylaMetaphlanData$Actinobacteria)
-abline(lm(phyla16SData$Actinobacteria ~ phylaMetaphlanData$Actinobacteria), col="red")
+<br><br><br><br>
 
-#Bacteroidetes
-summary(lm(phyla16SData$Bacteroidetes ~ phylaMetaphlanData$Bacteroidetes)) 
-#P-value 0.0002277, R2= 0.2358
-plot(phyla16SData$Bacteroidetes ~ phylaMetaphlanData$Bacteroidetes)
-abline(lm(phyla16SData$Bacteroidetes ~ phylaMetaphlanData$Bacteroidetes), col="red")
+**Figure S2: Bacteroidetes Linear Correlation between 16S Sequencing and Metaphlan2.**  Bacteroidetes has the best correlation of all the phyla (P-value = 0.0002277, R^2^ = 0.2358).
+The histogram for the 16S samples are on the x-axis while the histogram for Metaphlan2 is on the y-axis.
 
-#Firmicutes
-summary(lm(phyla16SData$Firmicutes ~ phylaMetaphlanData$Firmicutes)) 
-#P-value 0.1223, R2= 0.04617
-plot(phyla16SData$Firmicutes ~ phylaMetaphlanData$Firmicutes)
-abline(lm(phyla16SData$Firmicutes ~ phylaMetaphlanData$Firmicutes), col="red")
+```{r Figure_S2, echo=FALSE, warning=FALSE, message=FALSE}
+graphData <- as.data.frame(cbind(phyla16SData$Bacteroidetes, phylaMetaphlanData$Bacteroidetes))
+colnames(graphData) <- c("rRNA_16S", "Metaphlan2")
 
-#Proteobacteria
-summary(lm(phyla16SData$Proteobacteria ~ phylaMetaphlanData$Proteobacteria)) 
-#P-value 0.004252, R2= 0.1494
-plot(phyla16SData$Proteobacteria ~ phylaMetaphlanData$Proteobacteria)
-abline(lm(phyla16SData$Proteobacteria ~ phylaMetaphlanData$Proteobacteria), col="red")
+p1 <- ggplot(graphData, aes(x=rRNA_16S, y=Metaphlan2)) + geom_point(colour="blue", size=4)
+p2 <- p1 + geom_smooth(method="lm", colour="black", se=TRUE)
+p3 <- p2 + theme(axis.text=element_text(size=15), axis.title=element_text(size=20, face="bold"))
+p4 <- p3 + xlim(0, 100) + xlab("16S") + ylim(0, 100)
+p5 <- p4 + theme(axis.line=element_line(colour="black"), 
+                 panel.grid.major = element_blank(), 
+                 panel.grid.minor = element_blank(), 
+                 panel.border = element_blank(), 
+                 panel.background = element_blank())
 
-#Verrucomicrobia
-summary(lm(phyla16SData$Verrucomicrobia ~ phylaMetaphlanData$Verrucomicrobia)) 
-#P-value 0.1761, R2= 0.03559
-plot(phyla16SData$Verrucomicrobia ~ phylaMetaphlanData$Verrucomicrobia)
-abline(lm(phyla16SData$Verrucomicrobia ~ phylaMetaphlanData$Verrucomicrobia), col="red")
-
-#other
-summary(lm(phyla16SData$other ~ phylaMetaphlanData$other)) 
-#P-value 0.8054, R2= 0.001201
-plot(phyla16SData$other ~ phylaMetaphlanData$other)
-abline(lm(phyla16SData$other ~ phylaMetaphlanData$other), col="red")
-
-# Create a bland Altman of the best correlation
-  # bias (difference of metaphlan) based on 16S analysis  
-library(BlandAltmanLeh)
-bland.altman.plot(phyla16SData$Bacteroidetes, phylaMetaphlanData$Bacteroidetes, main="Bland Altman Plot of Bacteroidetes", xlab="Means", ylab="Differences")
+ggExtra::ggMarginal(p5, type = "histogram")
 
@@
 
-<br>
-This second R analysis chunk uses metaphlan2 to look at the questions we were tackling with the other data sets indivdiually.  
-```{r, eval=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
+<br><br><br><br>
 
-setwd("C:/users/marc/Desktop/obesity2/MetaHit/")
+**Figure S3: Bland-Altman plot of Bacteroidetes measurements for Metaphlan2 and 16S sequencing.**
 
-microb <- read.csv("finalized_metaHit_shared.csv")
-rownames(microb) <- microb[, 1]
-microb <- microb[,-1]
+```{r Figure_S3, echo=FALSE}
 
-metadata <- read.csv("MetaHit_metadata.csv")
-rownames(metadata) <- metadata[, 1]
-metadata <- metadata[, -1]
+ba <- bland.altman.plot(phyla16SData$Bacteroidetes, phylaMetaphlanData$Bacteroidetes, graph.sys="ggplot2", main="Bland Altman Plot of Bacteroidetes", xlab="Means", ylab="Differences")
 
-phyla <- read.csv("phyla.csv")
-rownames(phyla) <- phyla[, 1]
-phyla <- phyla[, -1]
+ba1 <- ba + theme(axis.text=element_text(size=15), axis.title=element_text(size=25, face="bold"))
+ba2 <- ba1 + ylab("Difference") + xlab("Mean of Measurements")
+ba3 <- ba2 + geom_point(size=4)
+ba4 <- ba3 + theme(axis.line=element_line(colour="black"), 
+                     panel.grid.major = element_blank(), 
+                     panel.grid.minor = element_blank(), 
+                     panel.border = element_blank(), 
+                     panel.background = element_blank())
 
-# Generate Overall Relative Abundance - normalize to a value of 100
-# Needs to be done since not all measures have 100 bacteria from metaphlan2
-overall <- rowSums(microb)
-microb.norm <- (microb / overall) * 100
-
-phyla.table <- read.csv("phyla.csv", header = T)
-rownames(phyla.table) <- phyla.table[, 1]
-phyla.table <- phyla.table[, -1]
-phyla.table <- as.data.frame(t(phyla.table))
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:10)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-rm(phyla.table, phyla.info)
-
-#Get alpha diversity of the samples
-library(vegan)
-
-H <- diversity(microb.norm)
-S <- specnumber(microb.norm)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-# Seems to look okay....
-
-#Create BMI groups
-metadata$BMI.class[metadata$BMI<=24] <- "Normal"
-metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
-metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
-metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
-
-#Create Obese Yes/No groups
-metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
-metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
-
-#Create Obese.num group
-metadata$obese.num[metadata$obese=="No"] <- 0
-metadata$obese.num[metadata$obese=="Yes"] <- 1
-
-#Create a column with obese and extreme obese as single entity
-metadata$BMIclass2[metadata$BMI<=24] <- "Normal"
-metadata$BMIclass2[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"      
-metadata$BMIclass2[metadata$BMI>=30] <- "Obese"
-
-#Get paitent demographics to be tested
-bmi <- metadata$BMI
-obese <- factor(metadata$obese)
-bmi.class <- factor(metadata$BMI.class)
-bmi.class2 <- factor(metadata$BMIclass2)
-
-####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-anova(lm(H ~ obese)) #P-value=0.2174
-anova(lm(H ~ bmi.class)) #P-value=0.5339
-anova(lm(H ~ bmi.class2)) #P-value=0.3628
-
-anova(lm(S ~ obese)) #P-value=0.9878
-anova(lm(S ~ bmi.class)) #P-value=0.7083
-anova(lm(S ~ bmi.class2)) #P-value=0.4985
-
-anova(lm(J ~ obese)) #P-value=0.1874
-anova(lm(J ~ bmi.class)) #P-value=0.3772
-anova(lm(J ~ bmi.class2)) #P-value=0.2326
-
-##linear correlations
-summary(lm(H ~ bmi)) #P-value=0.1022, R2=0.03185
-summary(lm(S ~ bmi)) #P-value=0.7642, R2=0.001091
-summary(lm(J ~ bmi)) #P-value=0.06682, R2=0.0399
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-anova(lm(bacter ~ obese)) #P-value=0.1911
-anova(lm(bacter ~ bmi.class)) #P-value=0.2975
-anova(lm(bacter ~ bmi.class2)) #P-value=0.2263
-
-anova(lm(firm ~ obese)) #P-value=0.2925
-anova(lm(firm ~ bmi.class)) #P-value=0.3654
-anova(lm(firm ~ bmi.class2)) #P-value=0.2998
-
-anova(lm(BFratio ~ obese)) #P-value=0.05254
-anova(lm(BFratio ~ bmi.class)) #P-value=0.1808
-anova(lm(BFratio ~ bmi.class2)) #P-value=0.1284
-
-####################################################################################### NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-adonis(microb.norm ~ obese, permutations=1000) 
-#PERMANOVA=0.05495, pseudo-F=1.6047
-
-set.seed(3)
-adonis(microb.norm ~ bmi.class, permutations=1000) 
-#PERMANOVA=0.1798, pseudo-F=1.176
-
-set.seed(3)
-adonis(microb.norm ~ bmi.class2, permutations=1000) 
-#PERMANOVA=0.2138, pseudo-F=1.1686
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-library(epiR)
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-median(H) # 2.886851
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-#OTU Richness
-median(S) # 85
-alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})
-
-#Evenness
-median(J) # 0.664365
-alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-S.cat <- alpha.test$S.cat
-J.cat <- alpha.test$J.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-table(test3[c(1:42), 4])
-table(test3[c(43:85), 4])
-#Group1 (Higher than median), obese = 17 and non-obese = 25
-#Group2 (Lower than median), obese = 20 and non-obese = 23
-
-group1 <- c(17, 25)
-group2 <- c(20, 23)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 1.15
-## CI = 0.71, 1.87
-## p-value = 0.575
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-median(BFRatio$BFRatio) # 0.9224021
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-table(test4[c(1:42), 2]) #higher group
-table(test4[c(43:85), 2]) #lower group
-#Group1 (Higher than median), obese = 20 and non-obese = 22
-#Group2 (Lower than median), obese = 17 and non-obese = 26
-
-group1 <- c(20, 22)
-group2 <- c(17, 26)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 0.83
-## CI = 0.51, 1.35
-## p-value = 0.452
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-library(AUCRF)
-
-#Create Obese.num group
-metadata$obese.num[metadata$obese=="No"] <- 0
-metadata$obese.num[metadata$obese=="Yes"] <- 1
-obese <- factor(metadata$obese.num)
-
-#generate test set
-# get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), microb.norm)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-set.seed(3)
-fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-summary(fit) # list of 19 Measures, AUCopt = 0.7649212
-plot(fit)
+ba4
 ```
-
-##### **Turnbaugh, et al. 2008:**  
-```{r, eval=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/turnbaugh.twins")
-shared.data <- read.table("test.unique.good.filter.unique.precluster.pick.pick.an.shared", header=T)
-rownames(shared.data) <- shared.data[, 2]
-shared.data <- shared.data[, -c(1:3)]
-
-subsample.data <- read.table("test.unique.good.filter.unique.precluster.pick.pick.an.0.03.subsample.shared", header=T)
-rownames(subsample.data) <- subsample.data[, 2]
-subsample.data <- subsample.data[, -c(1:3)]
-
-metadata <- read.csv("turnbaugh.metadata.csv")
-rownames(metadata) <- metadata[, 4]
-metadata <- metadata[, -4]
-
-keep1 <- which(metadata$Sample == 1) # n =146
-
-#use the first sampling
-#subset data for only the first sampling
-s1.metadata <- metadata[keep1, ]
-s1.subsample.data <- subsample.data[keep1, ]
-
-#generate alpha diversity measures with vegan
-library(vegan)
-H <- diversity(s1.subsample.data)
-S <- specnumber(s1.subsample.data)
-J <- H/log(S)
-select.alpha.diversity <- as.data.frame(cbind(H, S, J))
-s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
-alpha.test <- s1.alpha.diversity
-
-#Generate phyla table data
-
-phyla.info <- read.csv("phyla.csv")
-phyla.table <- shared.data
-colnames(phyla.table) <- phyla.info[, 2]
-
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(4:5, 7:9)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-s1.phyla.rel.abund <- phyla.table.rel.abund[rownames(shared.data), ]
-s1.phyla.rel.abund <- s1.phyla.rel.abund[keep1, ]
-
-#Create Obese Yes/No groups
-s1.metadata$obese[s1.metadata$BMI.category=="Lean" | s1.metadata$BMI.category=="Overweight"] <- "No"
-s1.metadata$obese[s1.metadata$BMI.category=="Obese"] <- "Yes"
-
-#Create Obese.num group
-s1.metadata$obese.num[s1.metadata$obese=="No"] <- 0
-s1.metadata$obese.num[s1.metadata$obese=="Yes"] <- 1
-
-#create groups to be used
-obese <- factor(s1.metadata$obese)
-bmi <- s1.metadata$BMI.category
-
-####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-anova(lm(H ~ obese)) #P-value=0.06976
-anova(lm(S ~ obese)) #P-value=0.02435
-anova(lm(J ~ obese)) #P-value=0.1253
-
-##linear correlations
-summary(lm(H ~ bmi)) #P-value=0.1926, R2=0.02277
-summary(lm(S ~ bmi)) #P-value=0.03434, R2=0.04606
-summary(lm(J ~ bmi)) #P-value=0.2988, R2=0.01675
-
-#B and F tests against obesity
-bacter <- s1.phyla.rel.abund$Bacteroidetes
-firm <- s1.phyla.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-anova(lm(bacter ~ obese)) #P-value=0.5771
-anova(lm(firm ~ obese)) #P-value=0.4073
-anova(lm(BFratio ~ obese)) #P-value=0.2216
-
-####################################################################################### NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-adonis(s1.subsample.data ~ obese, permutations=1000) 
-#PERMANOVA=0.09491, pseudo-F=1.2114
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-library(epiR)
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-median(H) # 4.327308
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-#OTU Richness
-median(S) # 290.5
-alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})
-
-#Evenness
-median(J) # 0.7611711
-alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-S.cat <- alpha.test$S.cat
-J.cat <- alpha.test$J.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-table(test3[c(1:73), 4])
-table(test3[c(74:146), 4])
-#Group1 (Higher than median), obese = 47 and non-obese = 26
-#Group2 (Lower than median), obese = 52 and non-obese = 21
-
-group1 <- c(47, 26)
-group2 <- c(52, 21)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 1.11
-## CI = 0.88, 1.38
-## p-value = 0.376
-
-##Run the RR for B/F ratio
-Bacter = s1.phyla.rel.abund$Bacteroidetes
-Firm = s1.phyla.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-median(BFRatio$BFRatio) # 0.4764057
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-table(test4[c(1:73), 2]) #higher group
-table(test4[c(74:146), 2]) #lower group
-#Group1 (Higher than median), obese = 54 and non-obese = 19
-#Group2 (Lower than median), obese = 45 and non-obese = 28
-
-group1 <- c(54, 19)
-group2 <- c(45, 28)
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-epi.2by2(r.test, method="cohort.count")
-## Risk Ratio = 0.83
-## CI = 0.66, 1.05
-## p-value = 0.111
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-library(AUCRF)
-
-#Create Obese.num group
-s1.metadata$obese.num[s1.metadata$obese=="No"] <- 0
-s1.metadata$obese.num[s1.metadata$obese=="Yes"] <- 1
-obese <- factor(s1.metadata$obese.num)
-
-#generate test set
-# get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), s1.subsample.data)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, s1.phyla.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-set.seed(3)
-fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-summary(fit) # list of 11 Measures, AUCopt = 0.7764883
-plot(fit)
-
-```
-
-#### **Part 2: Pooled Meta-analysis RR of Obesity for Shannon Diversity and B/F Ratio**    
-<br>
-This first R chunk shows the code used to generate the RR for obesity and Shannon diversity.  
-```{r, eval=FALSE}
-#Load required packages
-library(meta)
-library(metafor) #This package is used mostly for the more in-depth graph
-library(rmeta)  #This package was used for the less intensive graph
-
-#Change working directory
-setwd("C:/users/marc/Desktop/obesity2")
-
-StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8)
-Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu")
-Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012")
-Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=207", "n=58")
-tpos <- c(25, 16, 20, 6, 20, 52, 47, 2) #obese in low diversity group
-tneg <- c(61, 112, 12, 9, 23, 21, 57, 27) #nonobese in low diversity group
-cpos <- c(22, 10, 18, 4, 17, 47, 31, 3) #obese in high diversity group
-cneg <- c(64, 118, 13, 11, 25, 26, 72, 26) #nonobese in high diversity group
-RR <- c(1.14, 1.6, 1.08, 1.50, 1.15, 1.11, 1.50, 0.67)
-low <- c(0.70, 0.75, 0.72, 0.53, 0.71, 0.88, 1.04, 0.12)
-high <- c(1.85, 3.39, 1.61, 4.26, 1.87, 1.38, 2.16, 3.70)
-
-numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
-raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
-
-dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
-k <- length(dat$StudyNo)
-dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
-dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
-dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
-dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
-
-escalc(out ~ grp | study, weights = freq, data = dat.fm, measure = "RR")
-
-#Used to generate the RE Model estimate based on all the studies
-res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
-
-#To create a graph visualize the data
-forest(res, at=log(c(0.3, 1, 5)), atransf=exp, xlab="RR of Obesity in high versus low Shannon diversity based on Median", cex=0.8)
-par(font=2) #Switch to bold
-
-# RR Title ################################################
-#text(3.2, 3.10, "Relative Risk [95% CI]", pos=2, cex=0.8)#
-  #large scale (Plot in R studio at full)                 #
-                                                          #
-text(5.5, 9.3, "Relative Risk [95% CI]", pos=2, cex=0.8)  #
-  #smaller scale (Plot in R studia at normal)             #
-###########################################################
-
-# Group Title#############################################
-                                                         #
-text(-6.5, 9.3, "Study, Year, and n", pos=4, cex=0.8)    #
-                                                         #
-##########################################################
-
-dev.off()
-
-##Calculate combined P-value
-m = 8
-T = -2*log(0.608)-2*log(0.214)-2*log(0.236)-2*log(0.439)-2*log(0.575)-2*log(0.376)-2*log(0.025)-2*log(0.64)
-T.pval = 1-pchisq(T,m) #P-value = 0.01
-
-##Alternatively can use built in package (I trust this more)
-anova.rma(res)
-# QM = 5.1101, p-value = 0.0238
-
-```
-<br>
-This second R chunk shows the code used to generate the RR for obesity and B/F ratio.  
-```{r, eval=FALSE}
-#Load required packages
-library(meta)
-library(metafor) #This package is used mostly for the more in-depth graph
-
-#Change working directory
-setwd("C:/users/marc/Desktop/obesity2")
-
-StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8)
-Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu")
-Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012")
-Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=207", "n=58")
-tpos <- c(24, 13, 16, 3, 17, 45, 37, 2) #obese in low BF group
-tneg <- c(62, 115, 16, 12, 26, 28, 67, 27) #nonobese in low BF group
-cpos <- c(23, 13, 22, 7, 20, 54, 41, 3) #obese in high BF group
-cneg <- c(63, 115, 9, 8, 22, 19, 62, 26) #nonobese in high BF group
-RR <- c(1.04, 1.00, 0.70, 0.43, 0.83, 0.83, 0.89, 0.67)
-low <- c(0.64, 0.48, 0.47, 0.14, 0.51, 0.66, 0.66, 0.12)
-high <- c(1.70, 2.07, 1.07, 1.35, 1.35, 1.05, 1.27, 3.70)
-
-numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
-raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
-
-dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
-k <- length(dat$StudyNo)
-dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
-dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
-dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
-dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
-
-escalc(out ~ grp | study, weights = freq, data = dat.fm, measure = "RR")
-
-#Used to generate the RE Model estimate based on all the studies
-res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
-
-#To create a graph visualize the data
-forest(res, at=log(c(0.1, 1, 3)), atransf=exp, xlab="RR of Obesity in high versus low B/F ratio based on Median", cex=0.8)
-par(font=2) #Switch to bold
-
-# RR Title ################################################
-#text(3.2, 3.10, "Relative Risk [95% CI]", pos=2, cex=0.8)#
-#large scale (Plot in R studio at full)                 #
-#
-text(5.5, 9.3, "Relative Risk [95% CI]", pos=2, cex=0.8)  #
-#smaller scale (Plot in R studia at normal)             #
-###########################################################
-
-# Group Title#############################################
-#
-text(-6.0, 9.3, "Study, Year, and n", pos=4, cex=0.8)    #
-#
-##########################################################
-
-dev.off()
-
-##Calculate combined P-value
-m = 8
-T = -2*log(0.864)-2*log(1.00)-2*log(0.089)-2*log(0.121)-2*log(0.452)-2*log(0.111)-2*log(0.53)-2*log(0.64)
-T.pval = 1-pchisq(T,m) #P-value = 0.03155827
-
-##Alternatively can use built in package (I trust this more)
-anova.rma(res)
-  # QM = 5.3494, p-value = 0.0207
-
-```
-
-#### **Part 3: Pooled Analysis Using ZScore Normalizations**
-
-```{r, eval=FALSE}
-
-library(ggplot2)
-library(plyr)
-library(reshape2)
-library(RColorBrewer)
-library(scales)
-library(pheatmap)
-library(gridExtra)
-
-setwd("C:/Users/marc/Desktop/obesity2/BMI.normalization")
-
-#Load in BMI data
-data1 <- read.csv("combined.H.corr.data.csv")
-#Load in Turnbaugh data
-turn.data <- read.csv("Turnbaugh.normalized.csv")
-#Load in Amish data
-amish.data <- read.csv("Amish.normalized.csv")
-
-#Amish STD = 0.5893173
-amish.ZScore.H <- amish.data$H.corr/0.5893173
-hist(amish.ZScore.H)
-#Turnbaugh STD = 0.4818751
-turn.ZScore.H <- turn.data$H.corr/0.4818751
-hist(turn.ZScore.H)
-#MetaHit-Den STD = 0.4038558
-keepmetahit <- which(data1$dataset == "MetaHit")
-metahitdata <- data1[keepmetahit, ]
-metahit.ZScore.H <- metahitdata$H.corr/0.4038558
-hist(metahit.ZScore.H)
-#Korean STD = 0.5400615
-keepKorean <- which(data1$dataset == "Korean")
-Koreandata <- data1[keepKorean, ]
-Korean.ZScore.H <- Koreandata$H.corr/0.5400615
-hist(Korean.ZScore.H)
-#HMP STD = 0.6774514
-keepHMP <- which(data1$dataset == "HMP")
-hmpdata <- data1[keepHMP, ]
-HMP.ZScore.H <- hmpdata$H.corr/0.6774514
-hist(HMP.ZScore.H)
-#CCHC STD = 0.5784522
-keepCCHC <- which(data1$dataset == "CCHC")
-CCHCdata <- data1[keepCCHC, ]
-CCHC.ZScore.H <- CCHCdata$H.corr/0.5784522
-hist(CCHC.ZScore.H)
-#GLNE07 STD = 0.4663871
-keepGLNE <- which(data1$dataset == "GLNE")
-GLNEdata <- data1[keepGLNE, ]
-GLNE.ZScore.H <- GLNEdata$H.corr/0.4663871
-hist(GLNE.ZScore.H)
-#Escobar STD = 0.5378935
-keepescobar <- which(data1$dataset == "Escobar")
-escobardata <- data1[keepescobar, ]
-Escobar.ZScore.H <- escobardata$H.corr/0.5378935
-hist(Escobar.ZScore.H)
-#COMBO STD = 0.6870763
-keepCOMBO <- which(data1$dataset == "COMBO")
-COMBOdata <- data1[keepCOMBO, ]
-COMBO.ZScore.H <- COMBOdata$H.corr/0.6870763
-hist(COMBO.ZScore.H)
-
-#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
-HZScore <- c(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, COMBO.ZScore.H)
-data1 <- cbind(HZScore, data1)
-
-amish.data <- cbind(amish.ZScore.H, amish.data)
-turn.data <- cbind(turn.ZScore.H, turn.data)
-rm(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, amish.ZScore.H, turn.ZScore.H, COMBO.ZScore.H)
-rm(keepmetahit, keepCCHC, keepescobar, keepGLNE, keepHMP, keepKorean)
-
-#Create BMI groups
-data1$BMI.class[data1$bmi<=24] <- "Lean"
-data1$BMI.class[data1$bmi>24 & data1$bmi<30] <- "Overweight"
-data1$BMI.class[data1$bmi>=30] <- "Obese"
-
-#Create Obese Yes/No groups
-data1$obese[data1$BMI.class=="Lean" | data1$BMI.class=="Overweight"] <- "No"
-data1$obese[data1$BMI.class=="Obese"] <- "Yes"
-
-#Create group analysis table to harness full data set
-#For ZScore normalization
-d1.obese <- data1$obese
-t.obese <- as.character(turn.data$obese)
-am.obese <- as.character(amish.data$obese)
-
-d1.bmi.class <- data1$BMI.class
-t.bmi.class <- as.character(turn.data$BMI.category)
-
-d1.H.Zscore <- data1$HZScore
-t.H.Zscore <- turn.data$turn.ZScore.H
-am.H.Zscore<- amish.data$amish.ZScore.H
-
-#Combine everything together
-comb1.obese <- c(d1.obese, t.obese)
-comb2.obese <- c(d1.obese, t.obese, am.obese)
-comb1.bmi.class <- c(d1.bmi.class, t.bmi.class)
-comb1.H.Zscore <- as.data.frame(c(d1.H.Zscore, t.H.Zscore))
-comb2.H.Zscore <- as.data.frame(c(d1.H.Zscore, t.H.Zscore, am.H.Zscore))
-
-group.data.test1 <- cbind(comb1.H.Zscore, comb1.obese, comb1.bmi.class)
-group.data.test2 <- cbind(comb2.H.Zscore, comb2.obese)
-colnames(group.data.test1) <- c("H.Zscore", "obese", "BMI.class")
-colnames(group.data.test2) <- c("H.Zscore", "obese")
-#Look at Obese versus non-obese
-t.test(H.Zscore ~ obese, data=group.data.test2) #P-value = 0.002583
-table(group.data.test2$obese) # No = 695 | Yes = 340
-p1 <- ggplot(group.data.test2, aes(obese, H.Zscore)) + geom_boxplot()
-
-#A more in-depth look, breaking down the groups
-anova(lm(H.Zscore ~ BMI.class, data=group.data.test1)) #P-value = 0.01783
-a2 <- aov(H.Zscore ~ BMI.class, data=group.data.test1)
-posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-  # Obese-Lean (p-value=0.070), Over-Norm (p-value=0.791), Over-Obese (p-value=0.021)
-
-table(group.data.test1$BMI.class) 
-  # Lean = 326 | Overweight = 240 | Obese = 262
-ddply(group.data.test1, ~BMI.class, summarise, mean=mean(H.Zscore), sd=sd(H.Zscore))
-  # Lean = 0.042 +\- 0.940 | Overweight = 0.097 +/- 0.993 | Obese = -0.140 +\- 1.052
-
-#Let's change the order so it makes sense
-group.data.test1$BMI.class <- factor(group.data.test1$BMI.class, levels = c("Lean", "Overweight", "Obese"))
-
-p2 <- ggplot(group.data.test1, aes(BMI.class, H.Zscore)) + geom_boxplot()
-
-summary(lm(HZScore ~ bmi, data=data1)) #P-value 0.06558, R2= 0.004977
-plot(HZScore ~ bmi, data=data1)
-abline(lm(HZScore ~ bmi, data=data1), col="red")
-
-#Make a better graph with ggplot2
-p1 <- ggplot(data1, aes(x=bmi, y=HZScore))
-p2 <- p1 + geom_point(aes(color=dataset))
-p3 <- p2 + scale_colour_brewer(palette="Set1")
-p4 <- p3 + geom_smooth(method="lm", se=TRUE)
-
-#Probably want to add the Bacteroidetes / Firmicute ratio
-#Large variation in data sets observed so will have to normalize (zscore)
-#The data looks funny with either a log numerator with log sd denominator 
-#or just sd as the denominator
-#Think about using only Zscore but this produces a very large tail
-
-setwd("C:/Users/marc/Desktop/obesity2/BF.normalization")
-
-#Load in data
-data2 <- read.csv("combined.BF.corr.csv")
-
-#Create BMI groups
-data2$BMI.class[data2$bmi<=24] <- "Lean"
-data2$BMI.class[data2$bmi>24 & data2$bmi<30] <- "Overweight"
-data2$BMI.class[data2$bmi>=30] <- "Obese"
-
-#Create Obese Yes/No groups
-data2$obese[data2$BMI.class=="Lean" | data2$BMI.class=="Overweight"] <- "No"
-data2$obese[data2$BMI.class=="Obese"] <- "Yes"
-
-#Amish SD = 0.7918527
-am.BF.corr <- as.character(amish.data$BF.ratio.corr)
-am.BF.corr <- as.numeric(am.BF.corr) # Value 19 has no value
-amish.norm.BF <- am.BF.corr/0.7918527
-hist(amish.norm.BF)
-#Turnbaugh SD = 0.4578937
-turn.norm.BF <- turn.data$BF.corr/0.4578937
-hist(turn.norm.BF)
-#MetaHit SD = 0.9707577
-keepmetahit <- which(data2$dataset == "MetaHit")
-metahitdata <- data2[keepmetahit, ]
-metahit.norm.BF <- metahitdata$BF.ratio.corr/0.9707577
-hist(metahit.norm.BF)
-#Korean SD = 0.6048195
-keepKorean <- which(data2$dataset == "Korean")
-Koreandata <- data2[keepKorean, ]
-Korean.norm.BF <- Koreandata$BF.ratio.corr/0.6048195
-hist(Korean.norm.BF)
-#HMP SD = 8.301864
-keepHMP <- which(data2$dataset == "HMP")
-hmpdata <- data2[keepHMP, ]
-HMP.norm.BF <- hmpdata$BF.ratio.corr/8.301864
-hist(HMP.norm.BF)
-#CCHC SD = 2.00527
-keepCCHC <- which(data2$dataset == "CCHC")
-CCHCdata <- data2[keepCCHC, ]
-CCHC.norm.BF <- CCHCdata$BF.ratio.corr/2.00527
-hist(CCHC.norm.BF)
-#GLNE07 SD = 0.4451074
-keepGLNE <- which(data2$dataset == "GLNE")
-GLNEdata <- data2[keepGLNE, ]
-GLNE.norm.BF <- GLNEdata$BF.ratio.corr/0.4451074
-hist(GLNE.norm.BF)
-#Escobar SD = 0.4232106
-keepescobar <- which(data2$dataset == "Escobar")
-escobardata <- data2[keepescobar, ]
-Escobar.norm.BF <- escobardata$BF.ratio.corr/0.4232106
-hist(Escobar.norm.BF)
-#COMBO SD = 5.928664
-keepCOMBO <- which(data2$dataset == "COMBO")
-COMBOdata <- data2[keepCOMBO, ]
-COMBO.norm.BF <- COMBOdata$BF.ratio.corr/5.928664
-hist(COMBO.norm.BF)
-
-#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
-BFScore <- c(Escobar.norm.BF, GLNE.norm.BF, CCHC.norm.BF, HMP.norm.BF, Korean.norm.BF, metahit.norm.BF, COMBO.norm.BF)
-data2 <- cbind(BFScore, data2)
-
-amish.data <- cbind(amish.norm.BF, amish.data)
-turn.data <- cbind(turn.norm.BF, turn.data)
-rm(Escobar.norm.BF, GLNE.norm.BF, CCHC.norm.BF, HMP.norm.BF, Korean.norm.BF, metahit.norm.BF, amish.norm.BF, turn.norm.BF, COMBO.norm.BF)
-rm(keepmetahit, keepCCHC, keepescobar, keepGLNE, keepHMP, keepKorean)
-
-d1.BF.corr <- data2$BFScore
-t.BF.corr <- turn.data$turn.norm.BF
-am.BF.corr <- amish.data$amish.norm.BF
-
-comb.BF.corr <- as.data.frame(c(d1.BF.corr, t.BF.corr))
-group.data.test <- cbind(comb.BF.corr, group.data.test1)
-colnames(group.data.test)[1] <- "BF.ratio.corr"
-
-comb2.BF.corr <- as.data.frame(c(d1.BF.corr, t.BF.corr, am.BF.corr))
-group.data.test2 <- cbind(comb2.BF.corr, group.data.test2)
-colnames(group.data.test2)[1] <- "BF.ratio.corr"
-
-#Look at Obese versus non-obese
-t.test(BF.ratio.corr ~ obese, data=group.data.test2) 
-  # P-value = 0.0525
-table(group.data.test2$obese)
-  # No = 695 | Yes = 340
-bf1 <- ggplot(group.data.test2, aes(obese, BF.ratio.corr)) + geom_boxplot()
-
-#A more in-depth look, breaking down the groups
-anova(lm(BF.ratio.corr ~ BMI.class, data=group.data.test)) 
-  # P-value = 0.0227
-a2 <- aov(BF.ratio.corr ~ BMI.class, data=group.data.test)
-posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-  # Obese-Norm (p-value=0.026) | Over-Norm (p-value=0.117), Over-Obese (p-value=0.863)
-table(group.data.test$BMI.class)
-  # Lean = 326 | Overweight = 240 | Obese = 262
-ddply(group.data.test, ~BMI.class, summarise, mean=mean(BF.ratio.corr), sd=sd(BF.ratio.corr))
-  # Lean = -0.361 +\- 1.431 | Overweight = -0.695 +/- 2.032 | Obese = -0.787 +\- 2.461
-bf2 <- ggplot(group.data.test, aes(BMI.class, BF.ratio.corr)) + geom_boxplot()
-
-summary(lm(BFScore ~ bmi, data=data2)) #P-value 0.07859, R2= 0.004543
-plot(BFScore ~ bmi, data=data2)
-abline(lm(BFScore ~ bmi, data=data2), col="red")
-
-#Make a better graph with ggplot2
-b1 <- ggplot(data2, aes(x=bmi, y=BFScore))
-b2 <- b1 + geom_point(aes(color=dataset))
-b3 <- b2 + scale_colour_brewer(palette="Set1")
-b4 <- b3 + geom_smooth(method="lm", se=TRUE)
-
-##Run Zscore normalized Bacter and Firm overall analysis
-#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
-d1.bacter.corr <- data2$Zbacter
-d1.firm.corr <- data2$Zfirm
-d1.zbf.corr <- data2$ZBF
-t.bacter.corr <- turn.data$Zbacter
-t.firm.corr <- turn.data$Zfirm
-t.zbf.corr <- turn.data$ZBF
-am.bacter.corr <- amish.data$Zbacter
-am.firm.corr <- amish.data$Zfirm
-am.zbf.corr <- amish.data$ZBF
-
-#Bacteroidetes Datasets
-comb.bacter.corr <- as.data.frame(c(d1.bacter.corr, t.bacter.corr))
-group.data.test <- cbind(comb.bacter.corr, group.data.test)
-colnames(group.data.test)[1] <- "Zbacter"
-
-comb2.bacter.corr <- as.data.frame(c(d1.bacter.corr, t.bacter.corr, am.bacter.corr))
-group.data.test2 <- cbind(comb2.bacter.corr, group.data.test2)
-colnames(group.data.test2)[1] <- "Zbacter"
-
-#Firmicutes Datasets
-comb.firm.corr <- as.data.frame(c(d1.firm.corr, t.firm.corr))
-group.data.test <- cbind(comb.firm.corr, group.data.test)
-colnames(group.data.test)[1] <- "Zfirm"
-
-comb2.firm.corr <- as.data.frame(c(d1.firm.corr, t.firm.corr, am.firm.corr))
-group.data.test2 <- cbind(comb2.firm.corr, group.data.test2)
-colnames(group.data.test2)[1] <- "Zfirm"
-
-#Bacteroidetes/Firmicutes Datasets
-comb.zbf.corr <- as.data.frame(c(d1.zbf.corr, t.zbf.corr))
-group.data.test <- cbind(comb.zbf.corr, group.data.test)
-colnames(group.data.test)[1] <- "zbf"
-
-comb2.zbf.corr <- as.data.frame(c(d1.zbf.corr, t.zbf.corr, am.zbf.corr))
-group.data.test2 <- cbind(comb2.zbf.corr, group.data.test2)
-colnames(group.data.test2)[1] <- "zbf"
-
-###################################Bacteroidetes#######################################
-#Look at Obese versus non-obese
-t.test(Zbacter ~ obese, data=group.data.test2) 
-  # P-value = 0.3202
-table(group.data.test2$obese)
-  # No = 695 | Yes = 340
-bacter1 <- ggplot(group.data.test2, aes(obese, Zbacter)) + geom_boxplot()
-
-#A more in-depth look, breaking down the groups
-anova(lm(Zbacter ~ BMI.class, data=group.data.test)) #P-value = 0.0957
-a2 <- aov(Zbacter ~ BMI.class, data=group.data.test)
-posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-  # Obese-Norm (p-value=0.979) | Over-Norm (p-value=0.151) | Over-Obese (p-value=0.125)
-table(group.data.test$BMI.class)
-  # Lean = 326 | Overweight = 240 | Obese = 262
-ddply(group.data.test, ~BMI.class, summarise, mean=mean(Zbacter), sd=sd(Zbacter))
-  # Lean = 0.041 +/- 0.966 | Overweight = -0.117 +/- 1.000 | Obese = 0.057 +/- 1.02
-
-bacter2 <- ggplot(group.data.test, aes(BMI.class, Zbacter)) + geom_boxplot()
-
-summary(lm(Zbacter ~ bmi, data=data2)) #P-value 0.7168, R2= 0.0001936
-plot(Zbacter ~ bmi, data=data2)
-abline(lm(Zbacter ~ bmi, data=data2), col="red")
-
-#Make a better graph with ggplot2
-b1 <- ggplot(data2, aes(x=bmi, y=Zbacter))
-b2 <- b1 + geom_point(aes(color=dataset))
-b3 <- b2 + scale_colour_brewer(palette="Set1")
-b4 <- b3 + geom_smooth(method="lm", se=TRUE)
-
-###################################Firmicutes#######################################
-#Look at Obese versus non-obese
-t.test(Zfirm ~ obese, data=group.data.test2) #P-value = 0.8149
-table(group.data.test2$obese)
-  # No = 695 | Yes = 340
-firm1 <- ggplot(group.data.test2, aes(obese, Zfirm)) + geom_boxplot()
-
-#A more in-depth look, breaking down the groups
-anova(lm(Zfirm ~ BMI.class, data=group.data.test)) 
-  #P-value = 0.05521
-a2 <- aov(Zfirm ~ BMI.class, data=group.data.test)
-posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-  # Obese-Norm (p-value=0.976), Over-Norm (p-value=0.063), Over-Obese (p-value=0.1252)
-table(group.data.test$BMI.class)
-  # Lean = 326 | Overweight = 240 | Obese = 262
-ddply(group.data.test, ~BMI.class, summarise, mean=mean(Zfirm), sd=sd(Zfirm))
-  # Lean = -0.061 +/- 0.959 | Overweight = 0.130 +/- 1.018 | Obese = -0.043 +/- 1.013
-
-firm2 <- ggplot(group.data.test, aes(BMI.class, Zfirm)) + geom_boxplot()
-
-summary(lm(Zfirm ~ bmi, data=data2)) #P-value 0.8682, R2= 4.06x10-5
-plot(Zfirm ~ bmi, data=data2)
-abline(lm(Zfirm ~ bmi, data=data2), col="red")
-
-#Make a better graph with ggplot2
-b1 <- ggplot(data2, aes(x=bmi, y=Zfirm))
-b2 <- b1 + geom_point(aes(color=dataset))
-b3 <- b2 + scale_colour_brewer(palette="Set1")
-b4 <- b3 + geom_smooth(method="lm", se=TRUE)
-
-###################################Bacteroidetes/Firmicutes#######################################
-#Look at Obese versus non-obese
-t.test(zbf ~ obese, data=group.data.test2) #P-value = 0.4379
-table(group.data.test2$obese)
-  # No = 695 | Yes = 340
-zbf1 <- ggplot(group.data.test2, aes(obese, zbf)) + geom_boxplot()
-
-#A more in-depth look, breaking down the groups
-anova(lm(zbf ~ BMI.class, data=group.data.test)) 
-  # P-value = 0.4769
-a2 <- aov(zbf ~ BMI.class, data=group.data.test)
-posthoc <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-  # Obese-Norm (p-value=0.892) | Over-Norm (p-value=0.692) | Over-Obese (p-value=0.452)
-table(group.data.test$BMI.class)
-  # Lean = 326 | Overweight = 240 | Obese = 262
-ddply(group.data.test, ~BMI.class, summarise, mean=mean(zbf), sd=sd(zbf))
-  # Lean = 0.0081 +/- 0.954 | Overweight = -0.064 +/- 0.951 | Obese = 0.046 +/- 1.084
-zbf2 <- ggplot(group.data.test, aes(BMI.class, zbf)) + geom_boxplot()
-
-summary(lm(ZBF ~ bmi, data=data2)) #P-value 0.7118, R2= 0.0002008
-plot(ZBF ~ bmi, data=data2)
-abline(lm(ZBF ~ bmi, data=data2), col="red")
-
-#Make a better graph with ggplot2
-b1 <- ggplot(data2, aes(x=bmi, y=ZBF))
-b2 <- b1 + geom_point(aes(color=dataset))
-b3 <- b2 + scale_colour_brewer(palette="Set1")
-b4 <- b3 + geom_smooth(method="lm", se=TRUE)
-
-```
-
-##### **References:**

### 2016-02-12 code/GWAS9_master.sh

@@ -1,0 +1,13 @@
+## master script to submit all prephasing jobs in parallel (per chromosome)
+## will vary walltime request depending on scaling from chromosome length
+## GWAS model.. to be saved in directory /g01/home/nfishb/LHS/GWAS_9
+## slope ~ geno + PC1-5 + sex + age + fev0 + o'connor slope + BMI
+## in continuous smokers only
+## FEV0 is first non-missing observation 
+
+cd ~
+hours=($(<Jobs/SNPTEST_times.txt))
+for i in `seq 1 22`; do
+	let "ix=$i-1"
+	qsub -l walltime=${hours[$ix]},nodes=1:ppn=6,pmem=2gb -q nestor -N gwas9_chr${i} -v chr=${i},dir=/g01/home/nfishb/LHS/GWAS_9,samp=phenotypes_GWAS9.sample,pheno=slope Jobs/GWAS_snptest.sh
+done

### 2016-02-12 code/axiomStuff/COMBO1.sh

@@ -1,0 +1,50 @@
+#!/bin/bash
+
+# Make a directory to store the reference files
+#mkdir reference.files
+#mv silva.bacteria.fasta reference.files
+#mv trainset9_032012.pds.tax reference.files
+#mv trainset9_032012.pds.fasta reference.files
+
+# Read in the sra and make a fastq instead of a sff.
+mkdir completed.redo.sra
+mkdir logfile.completed
+
+for sample in $(ls *sra)
+do
+	fastq-dump $sample
+	mv $sample completed.redo.sra
+	mv *logfile logfile.completed 
+	
+done
+
+# Get the quality file and fasta information from the fastq
+mkdir completed.fastq
+
+for sample2 in $(ls *fastq)
+do
+	/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"	
+	mv $sample2 completed.fastq
+	mv *logfile logfile.completed
+	
+done
+
+# Trim the fastas according to quality 
+mkdir first.trim.completed
+for Trimming in $(ls *fasta)
+do
+	Quality=${Trimming//fasta/qual}
+	/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
+	mv $Trimming first.trim.completed
+	mv $Quality first.trim.completed
+	mv *logfile logfile.completed
+done
+
+mkdir scrap.fasta.completed
+mkdir scrap.qual.completed
+mkdir trim.qual.completed
+mv *scrap.qual scrap.qual.completed
+mv *scrap.fasta scrap.fasta.completed
+mv *trim.qual trim.qual.completed
+
+

### 2016-02-12 code/axiomStuff/COMBO2.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+
+
+groupN="SRR327637 SRR327638 SRR327639 SRR327640 SRR327641 SRR327642 SRR327643 SRR327644 SRR327645 SRR327646 SRR327647 SRR327648 SRR327649 SRR327650 SRR327651 SRR327652 SRR327653 SRR327654 SRR327655 SRR327656 SRR327657 SRR327658 SRR327659 SRR327660 SRR327661 SRR327662 SRR327663 SRR327664 SRR327665 SRR327666 SRR327667 SRR327668 SRR327669 SRR327670 SRR327671 SRR327672 SRR327673 SRR327674 SRR327675 SRR327676 SRR327677 SRR327678 SRR327679 SRR327680 SRR327681 SRR327682 SRR327683 SRR327684 SRR327685 SRR327686 SRR327687 SRR327688 SRR327689 SRR327690 SRR327691 SRR327692 SRR327693 SRR327694 SRR327695 SRR327696 SRR327697 SRR327698 SRR327699 SRR327700 SRR327701"
+
+
+for V in ${groupN}
+do
+	grep -e '^>' ${V}.trim.fasta > test.fasta
+	sed 's/>//' test.fasta > test2.fasta
+	sed "s/$/\t$V/" test2.fasta > ${V}.groups
+	rm test.fasta
+	rm test2.fasta
+done 
+
+cat *fasta > combined.fasta
+cat *groups > combined.groups
+
+

### 2016-02-12 code/axiomStuff/TwinsCustomChimeraProgram.sh

@@ -1,0 +1,19 @@
+$ nano test
+
+#Create new directories for chimera completed files
+mkdir chimera.completed
+mkdir mothurLogfiles
+
+#Make loop to run the chimera step and move completed samples to respective directory
+for chim in $(ls *.fasta)
+do
+	CFile=${chim//fasta/count_table}
+	/share/scratch/schloss/mothur/mothur "#chimera.uchime(fasta=$chim, count=$CFile, dereplicate=t, processors=8)"
+	
+	mv $chim chimera.completed
+	mv $CFile chimera.completed
+	mv *logfile mothurLogfiles
+	
+done
+
+exit 0

### 2016-02-12 code/axiomStuff/TwinsCustomChimeraProgram3.sh

@@ -1,0 +1,21 @@
+#!/bin/sh
+
+#Create new directories for chimera completed files
+mv *.chimeras chimera.completed
+mkdir fasta.completed
+mkdir accnos.completed
+
+#Make loop to run the chimera step and move completed samples to respective directory
+for chim in $(ls *.fasta)
+do
+	AccFile=${chim//fasta/denovo.uchime.accnos}
+
+	/share/scratch/schloss/mothur/mothur "#remove.seqs(fasta=$chim, accnos=$AccFile)"
+	
+	mv $chim fasta.completed
+	mv $AccFile accnos.completed
+	mv *logfile mothurLogfiles
+	
+done
+
+exit 0

### 2016-02-12 code/axiomStuff/TwinsCustomChimeraProgram4.sh

@@ -1,0 +1,60 @@
+#!/bin/sh
+
+Create new directories for chimera completed files
+mkdir deunique.fasta.completed
+mkdir deunique.countTable.completed
+
+Make loop to run the chimera step and move completed samples to respective directory
+for chim in $(ls *pick.fasta)
+do
+	CTable=${chim//pick.fasta/denovo.uchime.pick.count_table}
+
+	/share/scratch/schloss/mothur/mothur "#deunique.seqs(fasta=$chim, count=$CTable)"
+	
+	mv $chim deunique.fasta.completed
+	mv $CTable deunique.countTable.completed
+	mv *logfile mothurLogfiles
+done
+	
+mkdir redundant.fasta.completed
+mkdir redundant.groups.completed
+	
+for RFasta in $(ls *pick.redundant.fasta)
+do
+	RGroups=${RFasta//pick.redundant.fasta/denovo.uchime.pick.redundant.groups}
+	
+	/share/scratch/schloss/mothur/mothur "#rename.seqs(fasta=$RFasta, group=$RGroups)"
+
+	mv $RFasta redundant.fasta.completed
+	mv $RGroups redundant.groups.completed
+	mv *renamed_map redundant.groups.completed
+	mv *logfile mothurLogfiles
+done
+
+export var=1
+mkdir redundant.renamed.groups.completed
+mkdir redundant.renamed.fasta.completed
+echo ${var}
+for renamedF in $(ls *pick.redundant.renamed.fasta)
+do
+	renamedG=${renamedF//pick.redundant.renamed.fasta/denovo.uchime.pick.redundant.renamed.groups}
+	echo ${renamedF}
+	echo ${renamedG}
+	echo ${var}
+	
+	cp $renamedF ${var}.fasta
+	cp $renamedG ${var}.groups
+	
+	mv $renamedF redundant.renamed.fasta.completed
+	mv $renamedG redundant.renamed.groups.completed
+	var=$((var+1))
+done	
+	
+	
+/share/scratch/schloss/mothur/mothur "#merge.files(input=1.fasta-2.fasta-3.fasta-4.fasta-5.fasta-6.fasta-7.fasta-8.fasta-9.fasta-10.fasta-11.fasta-12.fasta-13.fasta-14.fasta-15.fasta-16.fasta-17.fasta-18.fasta-19.fasta-20.fasta-21.fasta, output=group16.fasta)"
[…]

### 2016-02-12 code/axiomStuff/YatsuProgram.sh

@@ -1,0 +1,41 @@
+#!/bin/sh
+
+#Create new directories for chimera completed files
+#mkdir precluster.completed
+#mkdir map.completed
+#mkdir mothurLogfiles
+
+#Make loop to run the chimera step and move completed samples to respective directory
+#for Pclust in $(ls *.fasta)
+#do
+#	tempfile=${Pclust//fasta/count_table}
+#	CFile=${tempfile//filter.unique/filter}
+		
+#	/share/scratch/schloss/mothur/mothur "#pre.cluster(fasta=$Pclust, count=$CFile, diffs=1)"
+	
+#	mv $Pclust precluster.completed
+#	mv $CFile precluster.completed
+#	mv *logfile mothurLogfiles
+#	mv *map map.completed
+	
+#done
+
+
+#Create new directories for chimera completed files
+#mkdir chimera.completed
+
+#Make loop to run the chimera step and move completed samples to respective directory
+for chim in $(ls *.fasta)
+do
+	CFile=${chim//fasta/count_table}
+	
+	
+	/share/scratch/schloss/mothur/mothur "#chimera.uchime(fasta=$chim, count=$CFile, dereplicate=t)"
+	
+	mv $chim chimera.completed
+	mv $CFile chimera.completed
+	mv *logfile mothurLogfiles
+	
+done
+
+exit 0

### 2016-02-12 code/axiomStuff/YatsuProgram2.sh

@@ -1,0 +1,27 @@
+#!/bin/sh
+
+#Create new directories for chimera completed files
+mv *.chimeras chimera.completed
+mkdir fasta.completed
+mkdir accnos.completed
+
+#Move into chimera.completed and take fasta files to group number folder
+
+cd chimera.completed
+mv *.fasta /mnt/EXT/Schloss-data/msze/test/Yatsunenko.Nature/combined.final/32
+cd ..
+
+#Make loop to run the chimera step and move completed samples to respective directory
+for chim in $(ls *.fasta)
+do
+	AccFile=${chim//fasta/denovo.uchime.accnos}
+
+	/share/scratch/schloss/mothur/mothur "#remove.seqs(fasta=$chim, accnos=$AccFile)"
+	
+	mv $chim fasta.completed
+	mv $AccFile accnos.completed
+	mv *logfile mothurLogfiles
+	
+done
+
+exit 0

### 2016-02-12 code/axiomStuff/YatsuProgram3.sh

@@ -1,0 +1,60 @@
+#!/bin/sh
+
+Create new directories for chimera completed files
+mkdir deunique.fasta.completed
+mkdir deunique.countTable.completed
+
+Make loop to run the chimera step and move completed samples to respective directory
+for chim in $(ls *pick.fasta)
+do
+	CTable=${chim//pick.fasta/denovo.uchime.pick.count_table}
+
+	/share/scratch/schloss/mothur/mothur "#deunique.seqs(fasta=$chim, count=$CTable)"
+	
+	mv $chim deunique.fasta.completed
+	mv $CTable deunique.countTable.completed
+	mv *logfile mothurLogfiles
+done
+	
+mkdir redundant.fasta.completed
+mkdir redundant.groups.completed
+	
+for RFasta in $(ls *pick.redundant.fasta)
+do
+	RGroups=${RFasta//pick.redundant.fasta/denovo.uchime.pick.redundant.groups}
+	
+	/share/scratch/schloss/mothur/mothur "#rename.seqs(fasta=$RFasta, group=$RGroups)"
+
+	mv $RFasta redundant.fasta.completed
+	mv $RGroups redundant.groups.completed
+	mv *renamed_map redundant.groups.completed
+	mv *logfile mothurLogfiles
+done
+
+export var=1
+mkdir redundant.renamed.groups.completed
+mkdir redundant.renamed.fasta.completed
+echo ${var}
+for renamedF in $(ls *pick.redundant.renamed.fasta)
+do
+	renamedG=${renamedF//pick.redundant.renamed.fasta/denovo.uchime.pick.redundant.renamed.groups}
+	echo ${renamedF}
+	echo ${renamedG}
+	echo ${var}
+	
+	cp $renamedF ${var}.fasta
+	cp $renamedG ${var}.groups
+	
+	mv $renamedF redundant.renamed.fasta.completed
+	mv $renamedG redundant.renamed.groups.completed
+	var=$((var+1))
+done	
+	
+	
+/share/scratch/schloss/mothur/mothur "#merge.files(input=1.fasta-2.fasta-3.fasta-4.fasta-5.fasta-6.fasta-7.fasta-8.fasta-9.fasta-10.fasta-11.fasta-12.fasta, output=group42.fasta)"
[…]

### 2016-02-12 code/axiomStuff/YatsuProgram4.sh

@@ -1,0 +1,23 @@
+#!/bin/sh
+
+Create new directories for chimera completed files
+mkdir uniqueSeqs.completed
+
+Make loop to run the chimera step and move completed samples to respective directory
+for unique in $(ls *fasta)
+do
+	#CTable=${chim//pick.fasta/denovo.uchime.pick.count_table}
+
+	/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=$unique)"
+	
+	mv $unique uniqueSeqs.completed
+	mv *logfile mothurLogfiles
+done
+		
+/share/scratch/schloss/mothur/mothur "#merge.files(input=group1.unique.fasta-group2.unique.fasta-group3.unique.fasta-group4.unique.fasta-group10.unique.fasta-group11.unique.fasta-group12.unique.fasta-group13.unique.fasta-group14.unique.fasta-group15.unique.fasta-group16.unique.fasta-group17.unique.fasta-group18.unique.fasta-group19.unique.fasta-group20.unique.fasta-group21.unique.fasta-group22.unique.fasta-group23.unique.fasta-group24.unique.fasta-group25.unique.fasta-group26.unique.fasta-group27.unique.fasta-group28.unique.fasta-group29.unique.fasta-group30.unique.fasta-group31.unique.fasta-group32.unique.fasta-group33.unique.fasta-group34.unique.fasta-group35.unique.fasta-group36.unique.fasta-group37.unique.fasta-group38.unique.fasta-group39.unique.fasta-group40.unique.fasta-group41.unique.fasta-group42.unique.fasta-group43.unique.fasta-group44.unique.fasta-group45.unique.fasta-group46.unique.fasta-group47.unique.fasta-group48.unique.fasta-group49.unique.fasta, output=combined.unique.fasta)"
[…]

### 2016-02-12 code/axiomStuff/amish.test2.sh

@@ -1,0 +1,50 @@
+#!/bin/bash
+
+# Make a directory to store the reference files
+#mkdir reference.files
+#mv silva.bacteria.fasta reference.files
+#mv trainset9_032012.pds.tax reference.files
+#mv trainset9_032012.pds.fasta reference.files
+
+# Read in the sra and make a fastq instead of a sff.
+#mkdir completed.redo.sra
+#mkdir logfile.completed
+
+#for sample in $(ls *sra)
+#do
+#	fastq-dump $sample
+#	mv $sample completed.redo.sra
+#	mv *logfile logfile.completed 
+	
+#done
+
+# Get the quality file and fasta information from the fastq
+mkdir completed.fastq
+
+for sample2 in $(ls *fastq)
+do
+	/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"	
+	mv $sample2 completed.fastq
+	mv *logfile logfile.completed
+	
+done
+
+# Trim the fastas according to quality 
+mkdir first.trim.completed
+for Trimming in $(ls *fasta)
+do
+	Quality=${Trimming//fasta/qual}
+	/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
+	mv $Trimming first.trim.completed
+	mv $Quality first.trim.completed
+	mv *logfile logfile.completed
+done
+
+mkdir scrap.fasta.completed
+mkdir scrap.qual.completed
+mkdir trim.qual.completed
+mv *scrap.qual scrap.qual.completed
+mv * scrap.fasta.completed
+mv *trim.qual trim.qual.completed
+
+

### 2016-02-12 code/axiomStuff/amish.test3.sh

@@ -1,0 +1,239 @@
+#!/bin/bash
+
+cat SRR050640* SRR050641* > p0_s2.fasta
+cat SRR050550* SRR050551* SRR050806* SRR050807* > p0_s1.fasta
+cat SRR050558* SRR050559* SRR050814* SRR050815* SRR050732* SRR050733* > p10_s1.fasta
+cat SRR050808* SRR050809* SRR050552* SRR050553* > p100_s1.fasta
+cat SRR050724* SRR050725* > p103_s1.fasta
+cat SRR050452* SRR050453* SRR050650* SRR050651* > p115_s1.fasta
+cat SRR050456* SRR050457* > p118_s1.fasta
+cat SRR050460* SRR050461* > p120_s2.fasta
+cat SRR050556* SRR050557* SRR050812* SRR050813* > p108_s1.fasta
+cat SRR050560* SRR050561* SRR050816* SRR050817* > p111_s1.fasta
+cat SRR050642* SRR050643* > p106_s2.fasta
+cat SRR050644* SRR050645* > p112_s1.fasta
+cat SRR050648* SRR050649* > p114_s1.fasta
+cat SRR050652* SRR050653* > p119_s1.fasta
+cat SRR050728* SRR050729* > p105_s1.fasta
+cat SRR050730* SRR050731* > p105_s2.fasta
+cat SRR050554* SRR050555* SRR050810* SRR050811* > p106_s1.fasta
+cat SRR050458* SRR050459* > p120_s1.fasta
+cat SRR050464* SRR050465* > p124_s1.fasta
+cat SRR050466* SRR050467* > p125_s1.fasta
+cat SRR050470* SRR050471* > p129_s1.fasta
+cat SRR050472* SRR050473* > p130_s1.fasta
+cat SRR050476* SRR050477* > p133_s1.fasta
+cat SRR050478* SRR050479* > p134_s1.fasta
+cat SRR050482* SRR050483* > p136_s1.fasta
+cat SRR050484* SRR050485* > p137_s1.fasta
+cat SRR050488* SRR050489* > p140_s1.fasta
+cat SRR050490* SRR050491* > p142_s1.fasta
+cat SRR050494* SRR050495* > p148_s1.fasta
+cat SRR050496* SRR050497* > p149_s1.fasta
+cat SRR050500* SRR050501* > p151_s1.fasta
+cat SRR050502* SRR050503* > p152_s1.fasta
+cat SRR050522* SRR050523* > p36_s1.fasta
[…]

### 2016-02-12 code/axiomStuff/amish.test4.sh

@@ -1,0 +1,19 @@
+#!/bin/bash
+
+
+
+groupN="102 107 113 116 11 126 128 12 13 14 150 156 157 160 162 168 169 177 179 186 190 191 192 1 201 208 211 212 213 214 216 217 219 224 226 230 234 238 242 24 259 269 275 283 287 289 297 2 308 313 316 323 327 333 346 349 352 354 358 367 373 376 381 388 392 395 397 398 401 406 410 411 416 423 424 427 429 444 445 44 453 455 459 463 464 46 473 474 475 481 482 483 488 48 49 50 51 523 52 53 54 59 61 66 68 77 7 81 83 84 87 92 p0_s1 p0_s2 p100_s1 p103_s1 p105_s1 p105_s2 p106_s1 p106_s2 p108_s1 p10_s1 p111_s1 p112_s1 p113_s1 p114_s1 p115_s1 p118_s1 p119_s1 p120_s1 p120_s2 p121_s1 p123_s1 p124_s1 p125_s1 p127_s1 p129_s1 p130_s1 p133_s1 p134_s1 p136_s1 p137_s1 p138_s1 p13_s1 p13_s2 p140_s1 p142_s1 p148_s1 p149_s1 p14_s1 p151_s1 p152_s1 p153_s1 p154_s1 p156_s1 p158_s1 p15_s1 p160_s1 p161_s1 p16_s2 p17_s1 p18_s1 p18_s2 p20_s1 p20_s2 p22_s1 p24_s1 p25_s1 p27_s1 p29_s1 p29_s2 p33_s1 p34_s1 p36_s1 p37_s1 p38_s1 p39_s1 p39_s2 p3_s1 p41_s1 p42_s1 p43_s1 p44_s2 p46_s1 p48_s1 p48_s2 p49_s1 p4_s1 p50_s1 p54_s1 p55_s1 p57_s1 p58_s1 p61_s1 p62_s1 p62_s2 p63_s1 p64_s1 p64_s2 p68_s1 p69_s1 p6_s1 p6_s2 p71_s2 p72_s1 p72_s2 p73_s1 p74_s2 p75_s1 p77_s1 p78_s1 p79_s1 p7_s1 p81_s2 p82_s1 p82_s2 p83_s1 p85_s2 p86_s1 p87_s1 p87_s2 p88_s1 p8_s1 p90_s1 p90_s2 p91_s1 p91_s2 p94_s2 p95_s1 p97_s1 p98_s1 p98_s2 p99_s1"
+
[…]

### 2016-02-12 code/axiomStuff/reCombineProgram.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+
+/share/scratch/schloss/mothur/mothur "#merge.files(input=group1.fasta-group2.fasta-group3.fasta-group4.fasta-group5.fasta-group6.fasta-group7.fasta-group8.fasta-group9.fasta-group10.fasta-group11.fasta-group12.fasta-group13.fasta-group14.fasta-group15.fasta-group16.fasta-group17.fasta-group18.fasta-group19.fasta-group20.fasta-group21.fasta-group22.fasta-group23.fasta-group24.fasta-group25.fasta, output=combined.fasta)"
+/share/scratch/schloss/mothur/mothur "#merge.files(input=group1.groups-group2.groups-group3.groups-group4.groups-group5.groups-group6.groups-group7.groups-group8.groups-group9.groups-group10.groups-group11.groups-group12.groups-group13.groups-group14.groups-group15.groups-group16.groups-group17.groups-group18.groups-group19.groups-group20.groups-group21.groups-group22.groups-group23.groups-group24.groups-group25.groups, output=combined.groups)"
+/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.fasta)"
+/share/scratch/schloss/mothur/mothur "#count.seqs(name=combined.names, group=combined.groups)"
+
+mv *logfile mothurLogfiles 
+
+
+exit 0

### 2016-02-12 code/axiomStuff/reCombineProgram2.sh

@@ -1,0 +1,12 @@
+#!/bin/sh
+
+/share/scratch/schloss/mothur/mothur "#merge.files(input=group1.fasta-group2.fasta-group3.fasta-group4.fasta-group10.fasta-group11.fasta-group12.fasta-group13.fasta-group14.fasta-group15.fasta-group16.fasta-group17.fasta-group18.fasta-group19.fasta-group20.fasta-group21.fasta-group22.fasta-group23.fasta-group24.fasta-group25.fasta-group26.fasta-group27.fasta-group28.fasta-group29.fasta-group30.fasta-group31.fasta-group32.fasta-group33.fasta-group34.fasta-group35.fasta-group36.fasta-group37.fasta-group38.fasta-group39.fasta-group40.fasta-group41.fasta-group42.fasta-group43.fasta-group44.fasta-group45.fasta-group46.fasta-group47.fasta-group48.fasta-group49.fasta, output=combined.fasta)"
+/share/scratch/schloss/mothur/mothur "#merge.files(input=group1.groups-group2.groups-group3.groups-group4.groups-group10.groups-group11.groups-group12.groups-group13.groups-group14.groups-group15.groups-group16.groups-group17.groups-group18.groups-group19.groups-group20.groups-group21.groups-group22.groups-group23.groups-group24.groups-group25.groups-group26.groups-group27.groups-group28.groups-group29.groups-group30.groups-group31.groups-group32.groups-group33.groups-group34.groups-group35.groups-group36.groups-group37.groups-group38.groups-group39.groups-group40.groups-group41.groups-group42.groups-group43.groups-group44.groups-group45.groups-group46.groups-group47.groups-group48.groups-group49.groups, output=combined.groups)"
+/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.fasta)"
+/share/scratch/schloss/mothur/mothur "#count.seqs(name=combined.names, group=combined.groups)"
+/share/scratch/schloss/mothur/mothur "#pre.cluster(fasta=combined.unique.fasta, count=combined.count_table, diffs=1)"
+
+mv *logfile mothurLogfiles 
+
+
+exit 0

### 2016-02-19 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
manuscript draft1 completed

@@
 ---
-title: "A Meta Analysis of the Bacterial Microbiome and Human Obesity"
+title: "Looking for a Signal in the Noise: Revisiting Obesity and the Bacterial Microbiome"
 output:
@@
 In this meta analysis, we drastically increase the number of data sets examined and investigate the pooled data to provide clearity on the association between the bacterial microbiome and obesity.
-**Methods:** A total of 11 data sets with BMI data were analyzed and processed through the standard mothur pipeline for either 454 or MiSeq data.
-The data included were from Baxter et al.
-2015, Ross et al.
+**Methods:** A total of 196 data sets were initially identified for inclusion in this meta-analysis.
+A total of 9 data sets were included once all the search criteria were completed (controls from the GLNE07 cohort, Ross et al.
 2015, Goodrich et al.
@@
 2014, Zupancic et al.
-2012, Yatsunenko et al.
-2012, Human Microbiome Project, Nam et al.
-2011, Wu et al.
+2012, Human Microbiome Project, Wu et al.
 2011, MetaHit, and Turnbaugh et al.
@@
 Finally, data normalization was used to investigate all data sets together for relationships between the microbiome and obesity.
-**Results:** For the first part of the analysis there was no significant difference (P > 0.05) between non-obese and obese individuals in Bacteroidetes, Firmicutes, B/F ratio, Shannon diversity, OTU richness, relative risk, or the NMDS of the Bray-Curtis distance matrix analyzed by PERMANOVA.
-For the second part of the analysis there was no pooled difference in RR of obesity based on Shannon diversity (RR = 1.05, CI = 0.90, 1.23).
-There was also no difference in RR based on the B/F ratio for the poole data (RR = 0.85, CI = 0.73, 1.00).
-**Conclusion:** Our meta-analysis supports the null hypothesis that there is no difference between non-obese and obese individuals, based on the currently avialable bacterial microbiome measures.
+**Results:** For the first part of the analysis there was no significant difference (P > 0.05) between non-obese and obese individuals in Bacteroidetes, Firmicutes, or B/F ratio.
+Shannon diversity, OTU richness, relative risk, or the NMDS of the Bray-Curtis distance matrix analyzed by PERMANOVA was significant in at least one study (P<0.05).
+For the second part of the analysis there was a signifcant pooled difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04, 1.36, (P-value = 0.0139)).
+There was also a significant difference in RR based in the B/F ratio for the pooled data (RR = 0.86, CI = 0.75, 0.99, (P-value = 0.0296)).
+**Conclusion:** Our meta-analysis supports the  hypothesis that there is a difference between non-obese and obese individuals in Shannon diversity and B/F ratio, based on the currently avialable bacterial microbiome data included in this meta-analysis.
 
@@
 
-The purpose of this study is to perform an extensive meta-analysis and systematic review of the bacterial microbiome and obesity by analyzing more data sets than the previous studies and including both seperate and pooled analysis to assess the bacterial microbiome and obesity.
+The purpose of this study is to perform an extensive meta-analysis of the bacterial microbiome and obesity by analyzing applying a more sytematic approach than has been used by previous studies in this area.
+Additionaly, this study will include both seperate and pooled analysis to assess the bacterial microbiome and obesity.
 By including these two components this study will build on the previous two re-analysis studies and provide a stronger assesment of the direction of effect or lack there of on the bacterial microbiome and obesity.
+In particular, the goal of this study is to provide additional clarity on the role of both bacterial diversity and the Bacteroidetes/Firmicutes ratio impact on obesity.
 
@@
 <br><br>
-A literature review was performed in pubmed for bacterial microbiome studies that also contained 16S rRNA gene sequencing data using 454, MiSeq, or HiSeq sequencing technologies.
-A total of 11 studies (combined n = 2133) were identified as fullfiling this criteria.
-One final study published late in 2015 used the Illumina NextSeq platform for sequencing [@zeevi_personalized_2015].
-However, the BMI data for each individual was not included in the published study or the supplement.
-An e-mail was sent to inquire but was never returned.
-Thus it was decided not to include this data set in the subsequent meta analysis.
+We attempted to stick to the original PRISMA guidelines for performing a meta-analysis as published by Moher et al [-@moher_preferred_2010].
+A detailed description of the literature seach and the exact formula used for the initial PubMed search can be found in the online supplement along with the PRISMA flow diagram [Figure 1].
+In brief, a total of 196 studies made it through the initial identification of studies.
+Screening removed a total of 184 studies leaving a total of 12 eligible studies for this meta-analysis.
+Upon a full reading of these specific manuscripts 2 were exculded and the reasons can be found in the online supplement.
+A total of 10 studies were ultimately deemed eligible for qualitative synthesis.
+One final study was removed from the quantitative analysis since no individual in the study was found to be able to be classified in the obese category based on the classical cutoffs.
+A total of 3 studies were from records identified by database searching [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014] and the other 6 were from records identified by other sources [@ross_16s_2015; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @conlan_species-level_2012].
+One of these studies (Baxter) has not been published at the time of this writing.
+
+*Subject Demographics and Study Outcomes*
+<br><br>
+The overall demographics for indivdiuals used in this meta-analysis split by their respective study can be found in table 1.
+A summary of the study type, study population, whether the data set is published, and the outcomes reported can be found in table 2.
+There is a large variation in reported outcomes with two studies finding bacterial community associations with obesity and three studies finding some association between either Firmicutes, Bacteroidetes, or the Bacteroidetes/Firmicutes ratio and obesity.
+However, a total of three manuscripts found no correlations between the bacterial microbiome and obesity.
+This is in addition to the two re-analysis studies published in 2014.
+The cutoffs used in this study for the BMI groups were lean (BMI less than or equal to 24), overweight (BMI was greater than 24 but less than 30), obese (BMI greater than or equal to 30).
+For analysis involving obese and non-obese groups the cutoffs used were a BMI less than 30 (lean/non-obese) and a BMI of equal to or greater than 30 (obese).
 
@@
 All data was publically available and the sequence data was downloaded from the NCBI Sequence Read Archive, MG-RAST, or the European Nucleotide Archive.
-A total of 7/11 [@turnbaugh_core_2009; @ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @wu_linking_2011; @escobar_gut_2014; @conlan_species-level_2012] studies were processed by staying as close as possible to the mothur 454 standard operating procedure, which can be found at http://www.mothur.org/wiki/454_SOP.
-A total of 3/11 [@goodrich_human_2014; @yatsunenko_human_2012] of the studies were processed by staying as close as possible to the mothur MiSeq standard operating procedure found at http://www.mothur.org/wiki/MiSeq_SOP.
+A total of 6/9 [@turnbaugh_core_2009; @ross_16s_2015; @zupancic_analysis_2012; @wu_linking_2011; @escobar_gut_2014; @conlan_species-level_2012] studies were processed by staying as close as possible to the mothur 454 standard operating procedure, which can be found at http://www.mothur.org/wiki/454_SOP.
+A total of 2/9 [@goodrich_human_2014] of the studies were processed by staying as close as possible to the mothur MiSeq standard operating procedure found at http://www.mothur.org/wiki/MiSeq_SOP.
 The final data set [@arumugam_enterotypes_2011] that was used took advantage of the Metaphlan2 package [@truong_metaphlan2_2015] to generate the necessary data for this analysis.
-A comparison of the similarity of Metaphlan2 to 16S sequencing using the Human Microbiome Project (HMP) data set can be found in the online supplement [Figure S1 & S2].
-For a few of the studies after sequence processing there was a large decrease in the number of samples available for analysis [Table 1].
-A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/marcsze/MOM-Analysis.
+A comparison of the similarity of Metaphlan2 to 16S sequencing using the Human Microbiome Project (HMP) data set can be found in the online supplement [Figure S1-S3].
+For the studies that utilized 16S rRNA gene sequencing three studies used primers targeting the V1-V3 region [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], two studies used the V4 region [@goodrich_human_2014], and one study used the V1-V2 region [@wu_linking_2011].
+Where there were two choices of data sets to use we used the V3-V5 data of the HMP [@conlan_species-level_2012] and the V2 data from the Turnbaugh study [-@turnbaugh_core_2009].
+A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/blob/master/code/sequenceProcessing.md.
 
@@
 To obtain the pooled RR and confidence intervals (CI) the metafor (v1.9-8) package available for the R statistics software was utilized.
-If the pooled analysis CI does not touch or cross 1.00 the pooled RR had a P-value < 0.05.
+Finally, to assess potential bias we also analyzed the two pooled analysis using a funnel plot.
 <br><br>
-The third and final category involved using mostly Zscore normalizations to pool all the data together to investigate the correlation between BMI and Shannon diversity, Bacteroidetes, Firmicutes, and B/F ratio.
+The third and final category involved using  Zscore normalizations to pool all the data together to investigate the correlation between BMI and Shannon diversity, Bacteroidetes, Firmicutes, and B/F ratio.
 Using the Zscore values a two-way t-test was used to investigate if there was a difference between non-obese and obese individuals.
@@
 
+*Seperate Analysis:*
+
+We used the categorical variable of obese or non-obese to analyze each data set seperately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
+Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 3 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix.
+There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio.
+The full summary of the exact P-values for each respective study and variable test can be found in table 3.
+
+When investigating how well the OTUs in each study were able to classify obese and non-obese individuals the average Out Of Bag (OOB) Area Under the Curve (AUC) was 0.7553 +/- 0.0888.
+The total range of the total number of variables used for the classification ranged from 3 (Ross and Escobar) to 32 (Zupancic).
+The total number of trees and node size was standardized to 1000 and 20 respectively for each data set.
+Perhaps surprisingly there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study.
+The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/9 studies), an OTU that could be classified to the family Lachnospiraceae (5/9 studies), and the Shannon diversity metric (4/9 studies).
+
+*Classical Pooled Meta-analysis:*
+
+When the Shannon diversity RR, all studies but Wu et al had those with lower Shannon diversity then the median as being at higher risk for obesity [Figure 2].
+This was confirmed with the pooled analysis showing an increased RR of 1.19 (CI 1.04, 1.36) for obesity in the lower then median Shannon diversity group (P-value = 0.0139).
+For the B/F ratio RR most studies showed that a low B/F ratio tended to have less obese indivdiuals than those with a high B/F ratio [Figure 3].
+A notable exception to this was the HMP data set and the Baxter data set [Figure 3].
+The pooled RR was 0.86 (CI 0.75, 0.99) suggesting that a low B/F ratio does indeed have less obese individuals (P-value = 0.0296).
+
+Using a funnel plot to asses the bias of the 9 data sets used in this meta-analysis we can observe that there does not seem to be any significant  bias [Figure 4].
+Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.
+
+*ZScore Normalization Pooled Analysis:*
+
+Using this approach we found that there was a significant difference in the normalized Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A].
+Further there was also a significant difference in the normalized Shannon diversity variance when the data was split between the three BMI groups (P-value = 0.03826) [Figure 5B].
+Using a Tukey post hoc test this difference was driven by the overweight versus the obese group (P-value = 0.03945).
+With respect to the B/F ratio using the ZScore normalization we found that there was no significant difference between obese and non-obese indivdiuals (P-value = 0.4516) [Figure 5C].
+Additionally, we did not observe a difference when the indivdiuals were seperated by BMI groups (P-value = 0.3317) [Figure 5D].
+
 *******
@@
 
+This study helps to provide a little more clarity to the on going debate of whether or not there are specific bacterial microbiome variables that are influenced by obesity.
+Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a bacterial microbiome difference between non-obese and obese individuals.
+However, neither study really made an attempt to pool the existing data together to try and harness the additional power that this would give.
+Here we perform an extensive literature review of the existing studies on the bacterial microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.
+We analyze the data using three different approaches.
+The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses ZScore normalizations and does not use RR.
+
+Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the bacterial microbiome and obestiy could be observed.
+However, the other two approaches using the pooled data was able to identify a correlation between measures of the bacterial microbiome and obesity [Figure 2 & 3].
+For Shannon diversity both approaches using the pooled data found a correlation while only one of the approaches for the B/F ratio found a correlation.
+Taken together this analysis provides a preliminary finding that there is a small but detectable difference between non-obese and obese individuals with respect to both Shannon diversity and the B/F ratio.
+However, our analysis would indicate that the signal for Shannon diversity may be more robust than that of the B/F ratio since both methods of pooled analysis found Shannon diversity significant while only one did for the B/F ratio.
+
+This study does have a few limitations that are worth mentioning.
+First,  it has been documented that in a number of groups BMI is not a good classification of obesity [@who_expert_consultation_appropriate_2004; @rahman_accuracy_2010].
+Of particular note, the asian population has a seperate set of guidelines for obesity [@who_expert_consultation_appropriate_2004].
+Although the asain population in the studies analyzed is very small [Table 1] it is possible that this may contribute to some of the noise in the data.
+However, for the measures, specifically the RR, there does not seem to be a study bias associated with this [Figure 4].
+We are also limited in our BMI classifications by what the previous studies published and performed.
+In general, not enough studies have been done investigating the bacterial microbiome in the context of other markers for obesity or for different ethnicities to be able to perform the analysis that we did.
+Thus as more specific studies are published that address ethnicity and better markers of obesity it is possible that the noise associated with these measurements will decrease.
+
+Second, although we make attempts to normalize the data for the pooled analysis so that we can compare the bacterial microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.
+However, what our analysis shows is that despite these differences, we are still able to identify an overall signal from the data.
+
+Third, this meta-analysis and review of the data to date is still rather preliminary.
+Only a total of 1542 individuals were included in the pooled analysis which is relatively small compared to other meta-analysis in non-microbiome fields of study.
+It will be necessary to continuously update this analysis, as more studies become available, in order to obtain a more accurate estimate of the true effect that obesity has on the bacterial microbiome within humans.
+
+Fourth, although a significant result exists for the RR for both Shannon diversity and the B/F ratio, it is small.
+This is also the case for the ZScore normalization analysis of the Shannon diversity.
+As an example, a lower than median Shannon diversity only has an overall 19% increase of obesity in it versus the higher than median Shannon diversity group.
+So even though there is a detectable signal it remains to be seen if this is a biologically significant phenomenon that could have a tangible effect on obesity.
+
+Despite the many problems that could add more noise to the measurements used, this study provides a reasonable first pass attempt to pool this data together in the effort to find a correlation between the bacterial microbiome and obesity.
+In doing so it provides a tanatalizing result that despite the large differences between studies it is still possible to detect a correlation between the bacterial microbiome and obesity.
+However, it should be cautioned that there are still many aspects of this analysis that could be improved upon and it truly is only a first attempt at trying to summerize and analyze the pooled data that is available.
+We hope that this study helps to add important information to the on-going dialogue of the bacterial microbiome's role in obesity.
+
 *******
-**Table 1:Summary Information of Studies used in Meta-analysis.**     
-![Table 1](C:\Users\marc\Desktop\obesity2\SummaryTableStudies.tif)
+##### Acknowledgements:
+
+The authors would like to thank Nielson Baxter for his help and suggestions during the sequence analysis portion.
+We would also like to thank Shawn Whitefield for her help with making sure the manuscript adhered to the guidelines set out for a meta-analysis.
+Finally, we would like to thank Rick Bushman, Hongzhe Li, and Pixu Shi in agreeing to release and helping to get a hold of the subject specific metadata from the Wu et al study.
+
+*******
+##### Tables:
+
+**Table 1:Summary Demographics of Individuals used in the Meta-analysis.**     
+```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
+library(knitr)
+
+Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic*", "HMP", "Wu", "Arumugam", "Turnbaugh**")
+
+Age <- c("54.3 +/- 9.9", "57.0 +/- 11.2", "61.5 +/- 8.9", "38.1 +/- 11.1", "48.2 +/- 13.2", "26.1 +/- 5.0", "26.4 +/- 9.4", "56.5 +/- 7.7", "21 - 32")
+
+Sex <- c("111|61", "48|15", "505|2", "14|16", "198|112", "127|129", "31|27", "45|40", "Not Available")
+
+Ancestry <- c(87.8, 0, "Not Available", 0, 100, 83, "Not Available", "Not Available", 51.4)
+
+BMI <- c("27.0 +/- 5.3", "31.6 +/- 5.3", "26.3 +/- 4.9", "27.4 +/- 4.5", "29.2 +/- 5.2", "24.1 +/- 3.4", "24.6 +/- 4.8", "27.8 +/- 5.8", "Not Available")
+
+Min <- c(17.5, 22.1, 16.2, 19.5, 16.7, 19, 14.0, 18.6, "Not Available")
+
+Max <- c(46.9, 47.9, 44.8, 37.6, 51.1, 34, 41.3, 40.2, "Not Available")
+
+table1 <- cbind(Study, Age, Sex, Ancestry, BMI, Min, Max)
+
+colnames(table1) <- c("Study", "Age (Mean +/- SD)", "Sex (F|M)", "European Ancestry (%)", "BMI (Mean +/- SD)", "Min BMI","Max BMI")
+
+kable(table1, format = "markdown", align = 'c', padding = 2)
+
+```
+
+*If numerical BMI information was not available then ranges and demographic information was taken from the authors published manuscript for this table.
+
+**Only BMI group information was provided and only an age range was given.
+
+**Table 2:Summary of Important Study Characteristics**
+
+```{r Table_2, echo=FALSE, message=FALSE, warning=FALSE}
+
+Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "HMP", "Wu", "Arumugam", "Turnbaugh")
+
+Type <- c("Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational")
+
+Population <- c("Adults without Cancer from Texas, Ontario, Michigan, Massachusetts", "Hispanic adults with and without type 2 diabetes from Texas
+", "Adult twins and mothers from the United Kingdom", "Healthy adults from Columbia", "Amish Adults with and without metabolic syndrome from Pennsylvania", "Healthy Adults from Texas and Missouri", "Healthy Adults from Pennsylvania", "Healthy Adults from Denmark", "Adult twins and mothers from Missouri")
+
+Published <- c("No", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes","Yes")
+
+Outcomes <- c("N/A (Unplublished)", "Mentions obesity, no specific correlations made though", "Christensenellaceae higher in low BMI individuals.
+No other correlations found.", "Firmicutes less abundant at higher BMI", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "No significant correlations", "Unweighted UniFrac Correlation with BMI", "No significant  correlations", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ")
+
+table2 <- cbind(Study, Type, Population, Published, Outcomes)
+
+colnames(table2) <- c("Study", "Study Type", "Population", "Published", "Outcomes Reported")
+
+kable(table2, format = "markdown", align = 'c', padding = 2)
+```
+
+**Table 3:Summary of P-values for Measurements of Interest for each Individual Study for Obese versus Normal**
+
+```{r Table_3, echo=FALSE, message=FALSE, warning=FALSE}
+
+Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "HMP", "Wu", "Arumugam", "Turnbaugh")
+
+Bacteroidetes <- c(0.318, 0.209, 0.666, 0.093, 0.958, 0.352, 0.917, 0.191, 0.577)
+
+Firmicutes <- c(0.682, 0.257, 0.967, 0.360, 0.265, 0.636, 0.577, 0.293, 0.407)
+
+BFRatio <- c(0.631, 0.251, 0.871, 0.466, 0.990, 0.569, 0.674, 0.053, 0.222)
+
+Shannon <- c(0.016, 0.166, 0.667, 0.864, 0.203, 0.585, 0.749, 0.217, 0.070)
+
+OTURich <- c(0.016, 0.184, 0.907, 0.195, 0.035, 0.845, 0.751, 0.988, 0.024)
+
+Evenness <- c(0.032, 0.254, 0.632, 0.476, 0.537, 0.356, 0.556, 0.187, 0.125)
+
+BrayC <- c(0.064, 0.715, 0.005, 0.089, 0.573, 0.811, 0.892, 0.055, 0.095)
+
+table3 <- cbind(Study, Bacteroidetes, Firmicutes, BFRatio, Shannon, OTURich, Evenness, BrayC)
+
+colnames(table3) <- c("Study", "Bacteroidetes", "Firmicutes", "B/F Ratio", "Shannon Diversity", "OTU Richness", "Evenness", "Bray Curtis")
+
+kable(table3, format = "markdown", align = 'c', padding = 2)
+```
+
+*******
+##### Figures:
+
+![Figure 1](Sze_ObesityMicrobiomeMeta_PeerJ_2016_files/figure-html/PrismaFlowGraph.png)\
+
+**Figure 1: PRISMA Flow Diagram of Total Records Searched.**  This was adapted from:  Moher D, Liberati A, Tetzlaff J, Altman DG, The PRISMA Group (2009).
+Preferred Reporting Items for Systematic Reviews and Meta-Analyses: The PRISMA Statement.
+PLoS Med 6(7): e1000097.
+doi:10.1371/journal.pmed1000097.
+
+```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE}
+
+library(meta)
+library(metafor)
+library(ggplot2)
+library(plyr)
+library(reshape2)
+library(RColorBrewer)
+library(scales)
+library(pheatmap)
+library(gridExtra)
+
+StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
+Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu", "Goodrich")
+Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012", "2014")
+Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=207", "n=58", "n=507")
+tpos <- c(25, 16, 20, 6, 20, 52, 47, 2, 56) #obese in low diversity group
+tneg <- c(61, 112, 12, 9, 23, 21, 57, 27, 198) #nonobese in low diversity group
+cpos <- c(22, 10, 18, 4, 17, 47, 31, 3, 47) #obese in high diversity group
+cneg <- c(64, 118, 13, 11, 25, 26, 72, 26, 206) #nonobese in high diversity group
+RR <- c(1.14, 1.6, 1.08, 1.50, 1.15, 1.11, 1.50, 0.67, 1.19)
+low <- c(0.70, 0.75, 0.72, 0.53, 0.71, 0.88, 1.04, 0.12, 0.84)
+high <- c(1.85, 3.39, 1.61, 4.26, 1.87, 1.38, 2.16, 3.70, 1.68)
+
+numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
+raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
+
+dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
+k <- length(dat$StudyNo)
+dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
+dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
+dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
+dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
+
+#Used to generate the RE Model estimate based on all the studies
+res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
+
+#To create a graph visualize the data
+forest(res, at=log(c(0.3, 1, 5)), atransf=exp, xlab="RR of Obesity in high versus low Shannon diversity based on Median", cex=0.8)
+par(font=2) #Switch to bold
+
+text(5.6, 10.3, "Relative Risk [95% CI]", pos=2, cex=0.8)
+text(-6.5, 10.3, "Study, Year, and n", pos=4, cex=0.8)
+
+```
+
+**Figure 2: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity.**  Groups were divided for each study on high and low Shannon Diversity groups based on the median for that study.
+The overall pooled relative risk was 1.19 for the low diversity group (P-value = 0.0139).
+
+```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE}
+
+StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
+Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu", "Goodrich")
+Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012", "2014")
+Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=201", "n=58", "n=507")
+tpos <- c(24, 13, 16, 3, 17, 45, 37, 2, 51) #obese in low BF group
+tneg <- c(62, 115, 16, 12, 26, 28, 67, 27, 203) #nonobese in low BF group
+cpos <- c(23, 13, 22, 7, 20, 54, 41, 3, 52) #obese in high BF group
+cneg <- c(63, 115, 9, 8, 22, 19, 62, 26, 201) #nonobese in high BF group
+RR <- c(1.04, 1.00, 0.70, 0.43, 0.83, 0.83, 0.89, 0.67, 0.98)
+low <- c(0.64, 0.48, 0.47, 0.14, 0.51, 0.66, 0.66, 0.12, 0.69)
+high <- c(1.70, 2.07, 1.07, 1.35, 1.35, 1.05, 1.27, 3.70, 1.38)
+
+numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
+raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
+
+dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
+k <- length(dat$StudyNo)
+dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
+dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
+dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
+dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
+
+#Used to generate the RE Model estimate based on all the studies
+res2 <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
+
+#To create a graph visualize the data
+forest(res2, at=log(c(0.1, 1, 3)), atransf=exp, xlab="RR of Obesity in high versus low B/F ratio based on Median", cex=0.8)
+par(font=2) #Switch to bold
+
+text(5.6, 10.3, "Relative Risk [95% CI]", pos=2, cex=0.8)
+text(-6.0, 10.3, "Study, Year, and n", pos=4, cex=0.8)
+```
+
+**Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low B/F ratio groups based on the median for that specific study.
+The overall pooled relative risk was 0.86 for the low diversity group (P-value = 0.0296).
+
+```{r Figure_4, echo=FALSE}
+par(mfrow=c(1,2))
+funnel(res, main="A")
+funnel(res2, main="B")
+```
+
+**Figure 4: Funnel Plot of the Shannon Diversity and Bacteroidetes/Firmicutes Ratio Relative Risk.**  **A)** Overall there does not seem to be any bias associated with the studies selected for the Shannon Diversity analysis with studies falling on either side of the predicted value and those with smaller total n falling further away from this.
+**B)** For the Bacteroidetes/Firmicutes ratio there does not seem to be any associated bias.
+The overall pattern is similar to that observed for the Shannon Diversity analysis.
+
+```{r Data_Load_Fig4, echo=FALSE, message=FALSE, warning=FALSE}
+
+setwd("C:/Users/marc/Desktop/obesity2/BMI.normalization")
+
+#Load in BMI data
+data1 <- read.csv("combined.H.corr.data.csv")
+#Load in Turnbaugh data
+turn.data <- read.csv("Turnbaugh.normalized.csv")
+#Load in Amish data
+amish.data <- read.csv("Amish.normalized.csv")
+
+#Amish STD = 0.5893173
+amish.ZScore.H <- amish.data$H.corr/0.5893173
+#Turnbaugh STD = 0.4818751
+turn.ZScore.H <- turn.data$H.corr/0.4818751
+#MetaHit-Den STD = 0.4038558
+keepmetahit <- which(data1$dataset == "MetaHit")
+metahitdata <- data1[keepmetahit, ]
+metahit.ZScore.H <- metahitdata$H.corr/0.4038558
+#Korean STD = 0.5400615
+keepKorean <- which(data1$dataset == "Korean")
+Koreandata <- data1[keepKorean, ]
+Korean.ZScore.H <- Koreandata$H.corr/0.5400615
+#HMP STD = 0.6774514
+keepHMP <- which(data1$dataset == "HMP")
+hmpdata <- data1[keepHMP, ]
+HMP.ZScore.H <- hmpdata$H.corr/0.6774514
+#CCHC STD = 0.5784522
+keepCCHC <- which(data1$dataset == "CCHC")
+CCHCdata <- data1[keepCCHC, ]
+CCHC.ZScore.H <- CCHCdata$H.corr/0.5784522
+#GLNE07 STD = 0.4663871
+keepGLNE <- which(data1$dataset == "GLNE")
+GLNEdata <- data1[keepGLNE, ]
+GLNE.ZScore.H <- GLNEdata$H.corr/0.4663871
+#Escobar STD = 0.5378935
+keepescobar <- which(data1$dataset == "Escobar")
+escobardata <- data1[keepescobar, ]
+Escobar.ZScore.H <- escobardata$H.corr/0.5378935
+#COMBO STD = 0.6870763
+keepCOMBO <- which(data1$dataset == "COMBO")
+COMBOdata <- data1[keepCOMBO, ]
+COMBO.ZScore.H <- COMBOdata$H.corr/0.6870763
+#COMBO STD = 0.3489582
+keeptwinsUK <- which(data1$dataset == "twinsUK")
+twinsUKdata <- data1[keeptwinsUK, ]
+twinsUK.ZScore.H <- twinsUKdata$H.corr/0.3489582
+
+#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit, COMBO, twinsUK
+HZScore <- c(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, COMBO.ZScore.H, twinsUK.ZScore.H)
+data1 <- cbind(HZScore, data1)
+
+amish.data <- cbind(amish.ZScore.H, amish.data)
+turn.data <- cbind(turn.ZScore.H, turn.data)
+rm(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, amish.ZScore.H, turn.ZScore.H, COMBO.ZScore.H, twinsUK.ZScore.H)
+rm(keepmetahit, keepCCHC, keepescobar, keepGLNE, keepHMP, keepKorean)
+
+#Create BMI groups
+data1$BMI.class[data1$bmi<=24] <- "Lean"
+data1$BMI.class[data1$bmi>24 & data1$bmi<30] <- "Overweight"
+data1$BMI.class[data1$bmi>=30] <- "Obese"
+
+#Create Obese Yes/No groups
+data1$obese[data1$BMI.class=="Lean" | data1$BMI.class=="Overweight"] <- "No"
+data1$obese[data1$BMI.class=="Obese"] <- "Yes"
+
+#Create group analysis table to harness full data set
+#For ZScore normalization
+d1.obese <- data1$obese
+t.obese <- as.character(turn.data$obese)
+am.obese <- as.character(amish.data$obese)
+
+d1.bmi.class <- data1$BMI.class
+t.bmi.class <- as.character(turn.data$BMI.category)
+
+d1.H.Zscore <- data1$HZScore
+t.H.Zscore <- turn.data$turn.ZScore.H
+am.H.Zscore<- amish.data$amish.ZScore.H
+
+#Combine everything together
+comb1.obese <- c(d1.obese, t.obese)
+comb2.obese <- c(d1.obese, t.obese, am.obese)
+comb1.bmi.class <- c(d1.bmi.class, t.bmi.class)
+comb1.H.Zscore <- as.data.frame(c(d1.H.Zscore, t.H.Zscore))
+comb2.H.Zscore <- as.data.frame(c(d1.H.Zscore, t.H.Zscore, am.H.Zscore))
+
+group.data.test1 <- cbind(comb1.H.Zscore, comb1.obese, comb1.bmi.class)
+group.data.test2 <- cbind(comb2.H.Zscore, comb2.obese)
+colnames(group.data.test1) <- c("H.Zscore", "obese", "BMI.class")
+colnames(group.data.test2) <- c("H.Zscore", "obese")
+
+setwd("C:/Users/marc/Desktop/obesity2/BF.normalization")
+
+#Let's change the order so it makes sense
+group.data.test1$BMI.class <- factor(group.data.test1$BMI.class, levels = c("Lean", "Overweight", "Obese"))
+
+group.data.test <- group.data.test1
+
+#Load in data
+data2 <- read.csv("combined.BF.corr.csv")
+
+#Create BMI groups
+data2$BMI.class[data2$bmi<=24] <- "Lean"
+data2$BMI.class[data2$bmi>24 & data2$bmi<30] <- "Overweight"
+data2$BMI.class[data2$bmi>=30] <- "Obese"
+
+#Create Obese Yes/No groups
+data2$obese[data2$BMI.class=="Lean" | data2$BMI.class=="Overweight"] <- "No"
+data2$obese[data2$BMI.class=="Obese"] <- "Yes"
+
+##Run Zscore normalized Bacter and Firm overall analysis
+#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
+d1.zbf.corr <- data2$ZBF
+t.zbf.corr <- turn.data$ZBF
+am.zbf.corr <- amish.data$ZBF
+
+#Bacteroidetes/Firmicutes Datasets
+comb.zbf.corr <- as.data.frame(c(d1.zbf.corr, t.zbf.corr))
+group.data.test <- cbind(comb.zbf.corr, group.data.test)
+colnames(group.data.test)[1] <- "zbf"
+
+comb2.zbf.corr <- as.data.frame(c(d1.zbf.corr, t.zbf.corr, am.zbf.corr))
+group.data.test2 <- cbind(comb2.zbf.corr, group.data.test2)
+colnames(group.data.test2)[1] <- "zbf"
+
+```
+
+```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE}
+# Section A
+p1 <- ggplot(group.data.test2, aes(obese, H.Zscore)) 
+p2 <- p1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
+p3 <- p2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
+p4 <- p3 + xlab("Obese Category") + ylab("ZScore Normalized H") 
+p5 <- p4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
+p6 <- p5 + theme(axis.line=element_line(colour="black"), 
+                 panel.grid.major = element_blank(), 
+                 panel.grid.minor = element_blank(), 
+                 panel.border = element_blank(), 
+                 panel.background = element_blank())
+p7 <- p6 + ggtitle('A') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
+
+# Section B
+q1 <- ggplot(group.data.test1, aes(BMI.class, H.Zscore)) 
+q2 <- q1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
+q3 <- q2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
+q4 <- q3 + xlab("BMI Classification") + ylab("ZScore Normalized H") 
+q5 <- q4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
+q6 <- q5 + theme(axis.line=element_line(colour="black"), 
+                 panel.grid.major = element_blank(), 
+                 panel.grid.minor = element_blank(), 
+                 panel.border = element_blank(), 
+                 panel.background = element_blank())
+q7 <- q6 + ggtitle('B') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
+
+# Section C
+bf1 <- ggplot(group.data.test2, aes(obese, zbf)) 
+bf2 <- bf1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
+bf3 <- bf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
+bf4 <- bf3 + xlab("Obese Category") + ylab("ZScore Normalized B/F Ratio") 
+bf5 <- bf4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
+bf6 <- bf5 + theme(axis.line=element_line(colour="black"), 
+                   panel.grid.major = element_blank(), 
+                   panel.grid.minor = element_blank(), 
+                   panel.border = element_blank(), 
+                   panel.background = element_blank())
+bf7 <- bf6 + ggtitle('C') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
+
+# Section D
+qbf1 <- ggplot(group.data.test, aes(BMI.class, zbf)) 
+qbf2 <- qbf1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
+qbf3 <- qbf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
+qbf4 <- qbf3 + xlab("BMI Classification") + ylab("ZScore Normalized B/F Ratio") 
+qbf5 <- qbf4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
+qbf6 <- qbf5 + theme(axis.line=element_line(colour="black"), 
+                     panel.grid.major = element_blank(), 
+                     panel.grid.minor = element_blank(), 
+                     panel.border = element_blank(), 
+                     panel.background = element_blank())
+qbf7 <- qbf6 + ggtitle('D') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
+
+grid.arrange(p7, q7, bf7, qbf7)
+```
+
+**Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** ZScore Normalized Shannon Diversity and Obesity.
+There was a significant difference between non-obese (n=1099) and obese groups (n=443) (P-value = 0.00493).
+**B)** ZScore Normalized Shannon Diversity by BMI Group.
+Total n for lean, overweight, and obese groups are 509, 365, 461 respectively.
+Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = 0.03944) but no difference between lean and obese group (P-value = 0.107).
+**C)** ZScore Normalized B/F Ratio and Obesity.
+There was no significant difference between the non-obese (n=1099) and obese (n=443) group based on B/F ratio (P-value = 0.4516).
+**D)** ZScore Normalized B/F Ratio by BMI Group.
+Total n for lean, overweight, and obese groups are 509, 365, 461 respectively.
+There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = 0.3317).
 

### 2016-02-19 microbiomeAndobesitySupplemental.Rmd
Completed supplemental draft1

--- 2016-02-12 code/microbiomeAndobesitySupplemental.Rmd
@@
 
-**Figure S1:Comparison of Phyla between 16S Sequencing and Metaphlan2.**
+**In-Depth Overview of Search Strategy:**
+
+The inital search strategy included looking for all papers that initially fit under the below NCBI PubMed advanced search criteria.
+The terms included in this criteria were that the manuscript had to have "Bacterial Microbiome" and "Obesity, BMI, bmi, obesity" in their manuscript criteria, it was not published more than 10 years ago, they were not review articles, and it contained research on humans only.
+The below formula when put into PubMed should recapitulate our initial search on the website.
+
+```{bash, eval=FALSE}
+(((((((((Bacterial Microbiome) AND (Obesity or bmi or body mass index or BMI or obesity) AND "last 10 years"[PDat] AND Humans[Mesh])) NOT review[ptyp]) AND "last 10 years"[PDat] AND Humans[Mesh])) AND "last 10 years"[PDat] AND Humans[Mesh])) AND "last 10 years"[PDat] AND Humans[Mesh])
+```
+
+This search yielded a total of 187 manuscripts.
+From two previous other reviews of obesity and the bacterial microbiome along with knowledge of two other published papers that investigated obesity but were missed by the database search we obtained a total of 7 more articles.
+We also had access to normal healthy individuals from an unpublished dataset.
+This brought our total number of records to 196.
+
+From this total we browsed abstracts for mention of stool or feces examination, that did not involve children, was not a clinical trial for probiotics or other diet related treatments, did not only have participants with inflammatory bowel disease, the articles were in English, did not only use PCR, qPCR, or RT-PCR only for their analysis, and sequecning that used only clone libraries.
+This ultimately excluded all but a total of 11 studies.
+
+From this total of 12 studies the full text was reviewed for whether or not sequencing data was publicly available, BMI information (either categorical or continuous) was available in a supplement or, if it was not available, whether authors upon contact were willing to share this information or direct us to repositories that stored this specific information.
+One study was excluded [@yatsunenko_human_2012] because it contained children and their sequencing of the 16S rRNA gene involved amplicons of only 100bp in length.
+They also did not have obesity as part of their results in the actual published manuscript.
+The second study was excluded because they did not have BMI information available and when contacted the authors never returned any correspondance [@zeevi_personalized_2015].
+
+This included a total of 10 studies in the qualitative synthesis of the analysis.
+Because we decided a prioi to use the standard definition for BMI group classification one study from this ten did not have any individuals who were obese by this criteria [@nam_comparative_2011] and was excluded from the final quantitative synthesis and analysis.
+
+*Inclusion Criteria:*
+
+* Contains mention of Bacterial Microbiome and Obesity
+* BMI, bmi, or obesity could be referenced instead of Obesity
+* Not published more than 10 years ago
+* Research on Humans only
+* At least one specific result examining obesity and a bacterial microbiome measure
+* Participants did not have Inflammatory Bowel Disease or Cancer
+* Greater than 100bp single or dual end reads for 16S Sequencing
+* DNA obtained from stool or feces
+
+*Exclusion Criteria:*
+
+* PCR, qPCR, or RT-PCR used as main analysis Tool
+* TRFLP or clone sequencing used to asses the bacterial community
+* Utilization of 100bp or less single end reads for sequencing
+* Sequencing Data not publicly available for download
+* BMI not available and authors do not return correspondance
+* Samples were not stool or feces
+* Study contained children
+* Study was a review
+
+**A Comparison of 16S Amplicon Sequencing and Metaphlan2:**
 
@@
 
-<br><br><br><br>
-
-**Figure S2: Bacteroidetes Linear Correlation between 16S Sequencing and Metaphlan2.**  Bacteroidetes has the best correlation of all the phyla (P-value = 0.0002277, R^2^ = 0.2358).
-The histogram for the 16S samples are on the x-axis while the histogram for Metaphlan2 is on the y-axis.
+**Figure S1:Comparison of Phyla between 16S Sequencing and Metaphlan2.**
 
@@
 
-<br><br><br><br>
-
-**Figure S3: Bland-Altman plot of Bacteroidetes measurements for Metaphlan2 and 16S sequencing.**
+**Figure S2: Bacteroidetes Linear Correlation between 16S Sequencing and Metaphlan2.**  Bacteroidetes has the best correlation of all the phyla (P-value = 0.0002277, R^2^ = 0.2358).
+The histogram for the 16S samples are on the x-axis while the histogram for Metaphlan2 is on the y-axis.
 
@@
 ```
+
+**Figure S3: Bland-Altman plot of Bacteroidetes measurements for Metaphlan2 and 16S sequencing.**
+
+*******
+##### References:

### 2016-02-19 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
manuscript draft1 completed

@@
 Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.
+
+We also investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC performed from each study performed when analyzed with RR [Figure S4].
+There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.98 (CI 0.63, 1.53), P-value = 0.9225).
+Thus it would seem that although Ruminococcaceae is important for classification the directionality is study dependent [Figure S4].
 

### 2016-02-19 microbiomeAndobesitySupplemental.Rmd
manuscript draft1 completed

@@
 
-**In-Depth Overview of Search Strategy:**
+###In-Depth Overview of Search Strategy:
 
@@
 
-**A Comparison of 16S Amplicon Sequencing and Metaphlan2:**
+###A Comparison of 16S Amplicon Sequencing and Metaphlan2:
 
@@
 
+###Supplemental Results:
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+#Load required packages
+library(meta)
+library(metafor) #This package is used mostly for the more in-depth graph
+
+#Change working directory
+setwd("C:/users/marc/Desktop/obesity2")
+
+StudyNo <- c(1, 2, 3, 4, 5, 6, 7)
+Study <- c("Baxter", "HMP", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu")
+Year <- c("2015", "2011", "2014", "2010", "2009", "2012", "2012")
+Total <- c("n=172", "n=256", "n=30", "n=85", "n=146", "n=201", "n=58")
+tpos <- c(36, 13, 7, 17, 22, 55, 2) #obese in low Rumi group
+tneg <- c(59, 140, 8, 26, 56, 103, 37) #nonobese in low Rumi group
+cpos <- c(11, 13, 3, 20, 25, 23, 3) #obese in high Rumi group
+cneg <- c(66, 90, 12, 22, 43, 26, 16) #nonobese in high Rumi group
+RR <- c(2.65, 0.67, 2.33, 0.83, 0.77, 0.74, 0.32)
+low <- c(1.45, 0.33, 0.74, 0.51, 0.48, 0.51, 0.06)
+high <- c(4.86, 1.39, 7.35, 1.35, 1.23, 1.07, 1.78)
+
+numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
+raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
+
+dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
+k <- length(dat$StudyNo)
+dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
+dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
+dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
+dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
+
+#escalc(out ~ grp | study, weights = freq, data = dat.fm, measure = "RR")
+
+#Used to generate the RE Model estimate based on all the studies
+res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
+
+#To create a graph visualize the data
+forest(res, at=log(c(0.01, 1, 10)), atransf=exp, xlab="RR of Obesity in high versus low Ruminococcaceae based on Median", cex=0.8)
+par(font=2) #Switch to bold
+
+text(7.4, 8.3, "Relative Risk [95% CI]", pos=2, cex=0.8)  
+text(-8.0, 8.3, "Study, Year, and n", pos=4, cex=0.8)
+
+```
+
+**Figure S4: Forest Plot of the Relative Risk based on the Ruminococcaceae classified OTU.**
+
 *******

### 2016-02-23 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd

@@
 ##### **Abstract:**
-**Rationale:** It has been well accepted within the literature that changes to the bacterial microbiome are related to obesity.
-However, recent reports from two different research groups re-examining existing data suggests that there may not be a strong association between the bacterial microbiome and obesity.
-These two newer studies, however, did not look at the results of pooling all the data together.
-In this meta analysis, we drastically increase the number of data sets examined and investigate the pooled data to provide clearity on the association between the bacterial microbiome and obesity.
-**Methods:** A total of 196 data sets were initially identified for inclusion in this meta-analysis.
-A total of 9 data sets were included once all the search criteria were completed (controls from the GLNE07 cohort, Ross et al.
-2015, Goodrich et al.
-2014, Escobar et al.
-2014, Zupancic et al.
-2012, Human Microbiome Project, Wu et al.
-2011, MetaHit, and Turnbaugh et al.
-2009.
-Three analysis piplines were pursued for the data.
-First, each data set was analyzed separately for relationships with BMI, similar to previous reports on the re-analysis of the bacterial microbiome and obesity.
-Second, a classical apporach for pooling data in a meta analysis was used to invesitigate the over all relative risk (RR) ratio from all studies of being obese based on the bacterial microbiome.
+**Rationale:** The hypothesis that changes in the microbiome and obesity are associated has been widely accepted.
+However, two recent reports that re-examined existing data suggest that acceptance of this hypothesis is unwarranted.
+These studies, however, studied the original studies in isolation from each other.
+We hypothesized that by statistically pooling the individual results, the statistical power to identify a significant association would increase.
+In this meta analysis, we tested the significance of the association between the microbiome and obesity by increasing the number of data sets examined and investigating the effect of pooling the results across the individual studies.
+**Methods:** We identified 196 studies that tested the microbiome-obesity association hypothesis; however when we limited the studies to those that did X, Y, Z (fill in criteria), a total of 9 data sets were identified.
+We applied three analysis piplines to the data from each study.
+First, we tested the association between the community diversity, Bacteroides to Firmicutes ratio (B/F), and XXXX with the subjects' body mass index (BMI) and obesity category.
+Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each studies.
 Finally, data normalization was used to investigate all data sets together for relationships between the microbiome and obesity.
-**Results:** For the first part of the analysis there was no significant difference (P > 0.05) between non-obese and obese individuals in Bacteroidetes, Firmicutes, or B/F ratio.
-Shannon diversity, OTU richness, relative risk, or the NMDS of the Bray-Curtis distance matrix analyzed by PERMANOVA was significant in at least one study (P<0.05).
-For the second part of the analysis there was a signifcant pooled difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04, 1.36, (P-value = 0.0139)).
-There was also a significant difference in RR based in the B/F ratio for the pooled data (RR = 0.86, CI = 0.75, 0.99, (P-value = 0.0296)).
+[aren't these two points the same thing?
+didn't you have to normalize the data to pool the results?
+Also, what about the RF stuff?]
+**Results:** There were no significant differences (P > 0.05) between the relative abundance of Bacteroidetes, Firmicutes, or thier ratio in non-obese and obese individuals.
+The Shannon diversity, OTU richness, relative risk, or oveall community dissimilarity was significant in at least one study (P<0.05).
+When we pooled the results across studies, there was a signifcant difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04, 1.36, (P-value = 0.0139)) and B/F ratio (RR = 0.86, CI = 0.75, 0.99, (P-value = 0.0296)).
+[RF results?]
 **Conclusion:** Our meta-analysis supports the  hypothesis that there is a difference between non-obese and obese individuals in Shannon diversity and B/F ratio, based on the currently avialable bacterial microbiome data included in this meta-analysis.
+[this repeats the previous sentence.
+why not say something about how the approach of pooling results offers an opportunity to test the generalizability of microbiome results by effectively increasing sample size or something?]
 
@@
 Where there were two choices of data sets to use we used the V3-V5 data of the HMP [@conlan_species-level_2012] and the V2 data from the Turnbaugh study [-@turnbaugh_core_2009].
-A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/blob/master/code/sequenceProcessing.md.
+A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/.
 

### 2016-02-23 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
adding Pat's suggestions to manuscript

@@
 In this meta analysis, we tested the significance of the association between the microbiome and obesity by increasing the number of data sets examined and investigating the effect of pooling the results across the individual studies.
-**Methods:** We identified 196 studies that tested the microbiome-obesity association hypothesis; however when we limited the studies to those that did X, Y, Z (fill in criteria), a total of 9 data sets were identified.
+
+**Methods:** We identified 196 studies that tested the microbiome-obesity association hypothesis; however when we limited the studies to those that did microbiome sequencing, had publicly available sequences, and were either willing to share the BMI information for every individual or were already published, a total of 9 data sets were identified.
 We applied three analysis piplines to the data from each study.
-First, we tested the association between the community diversity, Bacteroides to Firmicutes ratio (B/F), and XXXX with the subjects' body mass index (BMI) and obesity category.
+First, we tested the association between the community diversity, Bacteroides to Firmicutes ratio (B/F), and relative risk (RR) with the subjects' body mass index (BMI) and obesity category.
+We also used random forest to identify variables for classification of obese and non-obese indivdiuals.
 Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each studies.
 Finally, data normalization was used to investigate all data sets together for relationships between the microbiome and obesity.
-[aren't these two points the same thing?
-didn't you have to normalize the data to pool the results?
-Also, what about the RF stuff?]
+
 **Results:** There were no significant differences (P > 0.05) between the relative abundance of Bacteroidetes, Firmicutes, or thier ratio in non-obese and obese individuals.
 The Shannon diversity, OTU richness, relative risk, or oveall community dissimilarity was significant in at least one study (P<0.05).
+Overall, the random forest results for Area Under the Curve (AUC) was 0.7553 +/- 0.0888 and 7/9 studies contained an OTU that could be classified to the Ruminococcaceae family.
 When we pooled the results across studies, there was a signifcant difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04, 1.36, (P-value = 0.0139)) and B/F ratio (RR = 0.86, CI = 0.75, 0.99, (P-value = 0.0296)).
-[RF results?]
-**Conclusion:** Our meta-analysis supports the  hypothesis that there is a difference between non-obese and obese individuals in Shannon diversity and B/F ratio, based on the currently avialable bacterial microbiome data included in this meta-analysis.
-[this repeats the previous sentence.
-why not say something about how the approach of pooling results offers an opportunity to test the generalizability of microbiome results by effectively increasing sample size or something?]
+
+**Conclusion:** We show that by pooling the results of many indivdiual studies we can test the generalizability of the bacterial microbiome.
+Our meta-analysis supports the  hypothesis that there is a difference between non-obese and obese individuals in Shannon diversity and B/F ratio, based on the currently avialable bacterial microbiome data included in this meta-analysis.
 
@@
 This number increases to approximately 35% in adults (aged 20 or older) and these statisitics have seen little change since 2003 [@ogden_prevalence_2014].
-Traditionally the body mass index (BMI) has been used as the traditional method of classifying indviduals as non-obese or obese [@lichtash_body_2013].
-However, other methods do exist, such as body adiposity index (BAI) but do not always correlate with important molecular measures of traditionally associated with obesity [@lichtash_body_2013].
+Traditionally the body mass index (BMI) has been used as the traditional method of classifying individuals as non-obese or obese [@lichtash_body_2013].
+However, other methods do exist, such as body adiposity index (BAI) but do not always correlate with important molecular measures traditionally associated with obesity [@lichtash_body_2013].
 Recently, there has been a lot of interest in the bacteiral microbiome and it's potential ability to modulate obesity [@brahe_can_2016; @dror_microbiota_2016].
-If this modulation potential is true then the bacterial microbiome could have a enormous impact and role to play in the future treatment of obesity and helping to stem the rising obesity epidemic.
+If this modulation potential is true then the bacterial microbiome could have an enormous impact and role to play in the future treatment of obesity and in helping to stem the rising obesity epidemic.
 
@@
 
-The purpose of this study is to perform an extensive meta-analysis of the bacterial microbiome and obesity by analyzing applying a more sytematic approach than has been used by previous studies in this area.
-Additionaly, this study will include both seperate and pooled analysis to assess the bacterial microbiome and obesity.
-By including these two components this study will build on the previous two re-analysis studies and provide a stronger assesment of the direction of effect or lack there of on the bacterial microbiome and obesity.
-In particular, the goal of this study is to provide additional clarity on the role of both bacterial diversity and the Bacteroidetes/Firmicutes ratio impact on obesity.
-
 Although there have been numerous RT-PCR and qPCR studies directed at examining the bacterial microbiome and obesity [@guo_development_2008; @schwiertz_microbiota_2010].
 The main method of choice to obtain the most complete picture of the overall microbial community is through 16S sequencing of a specific hypervariable region on the 16S rRNA gene.
-The sequencers used to complete this task have varied depending on research group with the most popular being the Roche 454, Illumina MiSeq, and Illumina HiSeq.
-For this report we limit ourselves to analyzing studies were 16S sequencing data is avaliable along with BMI information.
+The sequencers used to complete this task have varied depending on research group with the most popular being the Roche 454 and Illumina MiSeq.
+A priori we decided that for this report we would limit ourselves to analyzing studies were 16S sequencing data is avaliable along with BMI information.
 We also analyzed every study with a uniform approach utilizing the mothur software package for sequence processing and clean up and R statistical software for analysis.
+
+The purpose of this study is to perform an extensive meta-analysis of the bacterial microbiome and obesity by analyzing and applying a more sytematic approach than has been used by previous studies in this area.
+Additionaly, we will include both seperate and pooled analysis to assess the bacterial microbiome and obesity.
+By including these two components we will build on the previous two re-analysis studies and provide a stronger assesment of the direction of effect or lack there of on the bacterial microbiome and obesity.
+In particular, our goal is to provide additional clarity on the role of both bacterial diversity and the Bacteroidetes/Firmicutes ratio impact on obesity.
 
@@
 *Literature Review and Study Inclusion*
-<br><br>
+
 We attempted to stick to the original PRISMA guidelines for performing a meta-analysis as published by Moher et al [-@moher_preferred_2010].
-A detailed description of the literature seach and the exact formula used for the initial PubMed search can be found in the online supplement along with the PRISMA flow diagram [Figure 1].
+A detailed description of the literature seach and the exact formula we used for the initial PubMed search can be found in the online supplement.
 In brief, a total of 196 studies made it through the initial identification of studies.
@@
 Upon a full reading of these specific manuscripts 2 were exculded and the reasons can be found in the online supplement.
-A total of 10 studies were ultimately deemed eligible for qualitative synthesis.
+A total of 10 studies were ultimately deemed eligible for qualitative synthesis [Figure 1].
 One final study was removed from the quantitative analysis since no individual in the study was found to be able to be classified in the obese category based on the classical cutoffs.
@@
 *Subject Demographics and Study Outcomes*
-<br><br>
-The overall demographics for indivdiuals used in this meta-analysis split by their respective study can be found in table 1.
+
+The overall demographics for individuals used in this meta-analysis split by their respective study can be found in table 1.
 A summary of the study type, study population, whether the data set is published, and the outcomes reported can be found in table 2.
-There is a large variation in reported outcomes with two studies finding bacterial community associations with obesity and three studies finding some association between either Firmicutes, Bacteroidetes, or the Bacteroidetes/Firmicutes ratio and obesity.
+We found a large variation in reported outcomes with two studies finding bacterial community associations with obesity and three studies finding some association between either Firmicutes, Bacteroidetes, or the Bacteroidetes/Firmicutes ratio and obesity.
 However, a total of three manuscripts found no correlations between the bacterial microbiome and obesity.
 This is in addition to the two re-analysis studies published in 2014.
-The cutoffs used in this study for the BMI groups were lean (BMI less than or equal to 24), overweight (BMI was greater than 24 but less than 30), obese (BMI greater than or equal to 30).
-For analysis involving obese and non-obese groups the cutoffs used were a BMI less than 30 (lean/non-obese) and a BMI of equal to or greater than 30 (obese).
+The cutoffs we used in this study for the BMI groups were lean (BMI less than or equal to 24), overweight (BMI was greater than 24 but less than 30), and obese (BMI greater than or equal to 30).
+For analysis involving obese and non-obese groups the cutoffs we used were a BMI less than 30 (lean/non-obese) and a BMI of equal to or greater than 30 (obese).
 
 *Sequence Analysis Pipeline*
-<br><br>
+
 All data was publically available and the sequence data was downloaded from the NCBI Sequence Read Archive, MG-RAST, or the European Nucleotide Archive.
@@
 *Data Analysis*
-<br><br>
-All analysis was performed using R version 3.2.2 and R studio version 0.99.491.
-The overall analysis was split into three general categories.
+
+We performed the analysis using R version 3.2.2 and R studio version 0.99.491.
+We split the overall analysis into three general categories.
 The first category involved following the same method employed by Finucane et al.
@@
 [-@finucane_taxonomic_2014; -@walters_meta-analyses_2014] in which each study was re-analyzed seperately for associations with BMI.
-The main associations with BMI that were explored in each seperate data set were Bacteroidetes phyla, Firmicutes phyla, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, NMDS visualization of Bray-Curtis distance matrix PERMANOVA, and relative risk (RR).
+The main associations that we explored for BMI in each seperate data set were the Bacteroidetes phyla, Firmicutes phyla, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, NMDS visualization of Bray-Curtis distance matrix PERMANOVA, and relative risk (RR).
 A two-way T-test was performed for comparison for non-obese versus obese.
-An ANOVA was used to investigate all BMI categories (if available) with a Tukey post hoc test.
-Third, a Pearson test for linear correlation between the variables was used.
 Finally, Random forest analysis to identify the best variables for classification of obese and non-obese was performed to look if there were similarities between data sets that we may have missed.
 The R package AUCRF (v1.1) was used for this.
-<br><br>
-The second category involved pooling all the data from all the studies together to asses whether the combined direction of all the studies increased an indivdiual's RR for being obese.
-The two measurements that were investigated in detail for this part were Shannon diversity and the B/F ratio.
-To generate the RR the median value of each study was taken and the number of those that were non-obese or obese in each group was recorded and a Fisher exact-test was performed.
+
+For the second category of analysis we pooled all the data from each seperate study together to asses whether the combined direction of all the studies increased an individual's RR for being obese.
+We investigated both Shannon diversity and the B/F ratio in detail for this part.
+In order to generate the RR we used the median value of each study and the number of those that were non-obese or obese in each group (above or below the median respectively) was recorded and a Fisher exact-test was performed.
 The relative risk from the the Fisher exact-test was obtained using the epiR (v0.9-6.9) package in R.
-To obtain the pooled RR and confidence intervals (CI) the metafor (v1.9-8) package available for the R statistics software was utilized.
-Finally, to assess potential bias we also analyzed the two pooled analysis using a funnel plot.
-<br><br>
-The third and final category involved using  Zscore normalizations to pool all the data together to investigate the correlation between BMI and Shannon diversity, Bacteroidetes, Firmicutes, and B/F ratio.
-Using the Zscore values a two-way t-test was used to investigate if there was a difference between non-obese and obese individuals.
-Differences in BMI categories (if data was available) was investigated using ANOVA with a tukey post hoc test.
-Linear correlations between these variables and BMI were explored using the Pearson test.
-Where appropriate the Bonferroni correction was applied for multiple comparisons.
+We obtained the pooled RR and confidence intervals (CI) from the metafor (v1.9-8) package in R.
+Finally, we also assessed the potential bias in the two pooled analysis by using a funnel plot.
+
+The third and final category involved using  Zscore normalizations to pool all the data together to investigate the correlation between either BMI and Shannon diversity or BMI and the B/F ratio.
+We used a two-way t-test on the Zscore values to investigate if there was a difference between non-obese and obese individuals.
+In order to asses differences between the BMI categories (if data was available) we used ANOVA with a tukey post hoc test.
+Where appropriate we applied a Bonferroni correction for multiple comparisons.
 In all other cases a P-value under 0.05 was considered to be significant.
@@
 
-We used the categorical variable of obese or non-obese to analyze each data set seperately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
+We used the categorical variable of obese or non-obese to analyze each data set seperately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
 Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 3 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix.
 There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio.
-The full summary of the exact P-values for each respective study and variable test can be found in table 3.
-
-When investigating how well the OTUs in each study were able to classify obese and non-obese individuals the average Out Of Bag (OOB) Area Under the Curve (AUC) was 0.7553 +/- 0.0888.
-The total range of the total number of variables used for the classification ranged from 3 (Ross and Escobar) to 32 (Zupancic).
+The full summary of the exact P-values for each respective study and variable tested can be found in table 3.
+
+Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.
+We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was 0.7553 +/- 0.0888.
+We found that the total range for the total number of variables used for the classification ranged from 3 (Ross and Escobar) to 32 (Zupancic).
 The total number of trees and node size was standardized to 1000 and 20 respectively for each data set.
-Perhaps surprisingly there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study.
+One aspect we found surprising was that there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study.
 The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/9 studies), an OTU that could be classified to the family Lachnospiraceae (5/9 studies), and the Shannon diversity metric (4/9 studies).
@@
 
-When the Shannon diversity RR, all studies but Wu et al had those with lower Shannon diversity then the median as being at higher risk for obesity [Figure 2].
+We next moved on to the pooled meta-analysis.
+For the Shannon diversity RR, all studies but Wu et al had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 2].
 This was confirmed with the pooled analysis showing an increased RR of 1.19 (CI 1.04, 1.36) for obesity in the lower then median Shannon diversity group (P-value = 0.0139).
-For the B/F ratio RR most studies showed that a low B/F ratio tended to have less obese indivdiuals than those with a high B/F ratio [Figure 3].
+When investigated the B/F ratio RR most studies showed that a low B/F ratio tended to have less obese individuals than those with a high B/F ratio [Figure 3].
 A notable exception to this was the HMP data set and the Baxter data set [Figure 3].
@@
 
-Using a funnel plot to asses the bias of the 9 data sets used in this meta-analysis we can observe that there does not seem to be any significant  bias [Figure 4].
+We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis.
+From the graph we can observe that there does not seem to be any significant  bias [Figure 4].
 Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.
 
-We also investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC performed from each study performed when analyzed with RR [Figure S4].
+The final forest plot that we used for this analysis investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC in the the seperate study analysis looked when pooled together to assess the RR [Figure S4].
 There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.98 (CI 0.63, 1.53), P-value = 0.9225).
@@
 
-Using this approach we found that there was a significant difference in the normalized Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A].
-Further there was also a significant difference in the normalized Shannon diversity variance when the data was split between the three BMI groups (P-value = 0.03826) [Figure 5B].
+Using the third approach with Zscore normalization we found that there was a significant difference in the Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A].
+Further there was also a significant difference in this normalized Shannon diversity value when the data was split between the three BMI groups (P-value = 0.03826) [Figure 5B].
 Using a Tukey post hoc test this difference was driven by the overweight versus the obese group (P-value = 0.03945).
-With respect to the B/F ratio using the ZScore normalization we found that there was no significant difference between obese and non-obese indivdiuals (P-value = 0.4516) [Figure 5C].
+However, in contrast to this result we found that the B/F ratio, using the ZScore normalization, was not significant different between obese and non-obese indivdiuals (P-value = 0.4516) [Figure 5C].
 Additionally, we did not observe a difference when the indivdiuals were seperated by BMI groups (P-value = 0.3317) [Figure 5D].
@@
 
-This study helps to provide a little more clarity to the on going debate of whether or not there are specific bacterial microbiome variables that are influenced by obesity.
+Our meta-analysis helps to provide a little more clarity to the on going debate of whether or not there are specific bacterial microbiome variables that are influenced by obesity.
 Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a bacterial microbiome difference between non-obese and obese individuals.
-However, neither study really made an attempt to pool the existing data together to try and harness the additional power that this would give.
+However, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give.
 Here we perform an extensive literature review of the existing studies on the bacterial microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.
@@
 Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the bacterial microbiome and obestiy could be observed.
-However, the other two approaches using the pooled data was able to identify a correlation between measures of the bacterial microbiome and obesity [Figure 2 & 3].
+However, we found that the other two approaches, that used the pooled data, was able to identify a correlation between measures of the bacterial microbiome and obesity [Figure 2 & 3].
 For Shannon diversity both approaches using the pooled data found a correlation while only one of the approaches for the B/F ratio found a correlation.
-Taken together this analysis provides a preliminary finding that there is a small but detectable difference between non-obese and obese individuals with respect to both Shannon diversity and the B/F ratio.
-However, our analysis would indicate that the signal for Shannon diversity may be more robust than that of the B/F ratio since both methods of pooled analysis found Shannon diversity significant while only one did for the B/F ratio.
-
-This study does have a few limitations that are worth mentioning.
+Taken together we believe that this analysis provides a strong preliminary finding that there is a small but detectable difference between non-obese and obese individuals with respect to both Shannon diversity and the B/F ratio.
+However, it should be noted that our analysis would indicate that the signal for Shannon diversity may be more robust than that of the B/F ratio since the two different pooled analysis approaches found Shannon diversity significant while only one did for the B/F ratio.
+
+We do acknowledge that this study does have a few limitations that are worth mentioning.
 First,  it has been documented that in a number of groups BMI is not a good classification of obesity [@who_expert_consultation_appropriate_2004; @rahman_accuracy_2010].
 Of particular note, the asian population has a seperate set of guidelines for obesity [@who_expert_consultation_appropriate_2004].
-Although the asain population in the studies analyzed is very small [Table 1] it is possible that this may contribute to some of the noise in the data.
-However, for the measures, specifically the RR, there does not seem to be a study bias associated with this [Figure 4].
-We are also limited in our BMI classifications by what the previous studies published and performed.
-In general, not enough studies have been done investigating the bacterial microbiome in the context of other markers for obesity or for different ethnicities to be able to perform the analysis that we did.
-Thus as more specific studies are published that address ethnicity and better markers of obesity it is possible that the noise associated with these measurements will decrease.
+Although the asain population in the studies analyzed are very small [Table 1] it is possible that this may contribute to some of the noise in the data.
+However, it is important to keep in mind that for these measures, specifically the RR, there does not seem to be any study bias due to studies having more ethnic diversity [Figure 4].
+Additionally, we are also limited in our BMI classifications by what the previous studies published and performed.
+In general, not enough studies have been done to investigate the bacterial microbiome in the context of other markers for obesity or on different ethnicities to be able to perform the analysis that we did.
+Thus as more specific studies are published that address ethnicity and studies on the microbiome that use better markers of obesity are completed it is entirely possible that the noise associated with these measurements will decrease.
 
@@
 However, what our analysis shows is that despite these differences, we are still able to identify an overall signal from the data.
-
 Third, this meta-analysis and review of the data to date is still rather preliminary.
@@
 
-Despite the many problems that could add more noise to the measurements used, this study provides a reasonable first pass attempt to pool this data together in the effort to find a correlation between the bacterial microbiome and obesity.
-In doing so it provides a tanatalizing result that despite the large differences between studies it is still possible to detect a correlation between the bacterial microbiome and obesity.
-However, it should be cautioned that there are still many aspects of this analysis that could be improved upon and it truly is only a first attempt at trying to summerize and analyze the pooled data that is available.
-We hope that this study helps to add important information to the on-going dialogue of the bacterial microbiome's role in obesity.
+Despite the many problems that could add more noise to the measurements used, this study provides a reasonable first pass attempt to pool this data together in an effort to find a correlation between the bacterial microbiome and obesity.
+In doing so we believe that it provides a tanatalizing result that despite the large differences between studies it is still possible to detect a correlation between the bacterial microbiome and obesity.
+However, we emphaszie that it should still be viewed with caution since there are still many aspects of this analysis that could be improved upon.
+It truly is only a first attempt at trying to summerize and analyze the pooled data that is available.
+We hope that this study helps to add important information and clarity to the on-going dialogue of the bacterial microbiome's role in obesity.
 
@@
 
+```{r Data_Baxter, echo=FALSE, message=FALSE, warning=FALSE}
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+# Reading in the necessary Data
+setwd("C:/Users/marc/Desktop/obesity2/data1")
+demographics <- read.csv("demographics.v2.csv")
+microbiome <- read.csv("data1.subsample.otus.csv")
+phylogenetic.info <- read.csv("data1.summary.taxonomy.csv")
+taxonomy <- read.csv("data1.taxonomy.csv")
+
+# Minor modifications and Rownames adustments of data tables
+rownames(demographics) <- demographics[,1]
+rownames(microbiome) <- microbiome[,2]
+demographics <- demographics[,-1]
+microbiome <- microbiome[,-2]
+test.samples <- rownames(demographics)
+microbiome <- microbiome[,-c(1:2)]
+
+# Create Table for phyla information
+phylogenetic.info <- phylogenetic.info[-c(1:12),]
+phyla <- which(phylogenetic.info$taxlevel == 2)
+phyla.table <- phylogenetic.info[phyla, ]
+phyla.table <- phyla.table[,-c(1:2,4:5)]
+phyla.names <- as.character(phyla.table[,1])
+phyla.table <- phyla.table[,-1]
+phyla.table <- as.data.frame(t(phyla.table))
+colnames(phyla.table) <- phyla.names
+rownames(phyla.table) <- rownames(demographics)
+
+#Add phyla together that are not very abundant and delete them from the table
+phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deferribacteres", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "Cyanobacteria_Chloroplast", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(1, 4:7, 9:11, 13, 15)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
+phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
+
+#Generate Alpha Diversity Measures and alpha diversity table
+library(vegan)
+H <- diversity(microbiome)
+S <- specnumber(microbiome)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Create age quartiles
+
+demographics$age.quartile <- with(demographics, cut(Age, breaks=quantile(Age, probs=seq(0,1, by=0.25)), include.lowest=TRUE))
+
+#Create Obese Yes/No groups
+demographics$obese[demographics$BMI.classification=="Normal" | demographics$BMI.classification=="Overweight"] <- "No"
+demographics$obese[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Yes"
+
+#Create Obese.num groups
+demographics$obese.num[demographics$obese=="No"] <- 0
+demographics$obese.num[demographics$obese=="Yes"] <- 1
+obese <- factor(demographics$obese.num)
+
+######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
+###########################################################################
+
+#Create a column with obese and extreme obese as single entity
+demographics$BMIclass2[demographics$BMI.classification=="Normal"] <- "Normal"
+demographics$BMIclass2[demographics$BMI.classification=="Overweight"] <- "Overweight"                     
+demographics$BMIclass2[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Obese"
+
+##Test BMI versus alpha diversity and phyla
+
+bmi <- demographics$BMI
+
+baxterH <- anova(lm(H ~ obese)) #P-value=0.01643
+baxterS <- anova(lm(S ~ obese)) #P-value=0.01571
+baxterJ <- anova(lm(J ~ obese)) #P-value=0.03241
+
+#B and F tests against obesity
+bacter <- phyla.table.rel.abund$Bacteroidetes
+firm <- phyla.table.rel.abund$Firmicutes
+BFratio <- bacter/firm
+
+baxterBacter <- anova(lm(bacter ~ obese)) #P-value=0.3175
+baxterFirm <- anova(lm(firm ~ obese)) #P-value=0.6817
+baxterBF <- anova(lm(BFratio ~ obese)) #P-value=0.6305
+
+###########################################################################
+############ NMDS and PERMANOVA Analysis###################################
+###########################################################################
+
+set.seed(3)
+baxter2 <- adonis(microbiome ~ obese, permutations=1000)
+baxterPERM <- baxter2$aov.tab
+#Not significant PERMANOVA = 0.06394, pseudo-F = 1.4246
+
+###########################################################################
+############ Relative Risk#################################################
+###########################################################################
+
+library(epiR)
+
+#Generate median values and put them into existing alpha.test dataframe
+
+#Shannon diversity
+alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
+
+##Run the RR for Shannon Diversity
+H.cat <- alpha.test$shannon.cat
+bmi.cat <- as.character(obese)
+test3 <- cbind(H.cat, bmi.cat)
+test3 <- test3[order(H.cat), ]
+orderedHCat <- test3[, 1]
+BaxtotalN <- length(orderedHCat)
+BaxHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
+BaxHighShannonGroup <- as.data.frame(table(test3[c(1:BaxHHighTotal), 2]))
+BaxLowShannonGroup <- as.data.frame(table(test3[c((BaxHHighTotal + 1):BaxtotalN), 2]))
+#Group1 (Higher than median), obese = 22 and non-obese = 64
+#Group2 (Lower than median), obese = 25 and non-obese = 61
+
+group1 <- c(BaxHighShannonGroup[2, 2], BaxHighShannonGroup[1, 2])
+group2 <- c(BaxLowShannonGroup[2, 2], BaxLowShannonGroup[1, 2])
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+baxterHEpi <- epi.2by2(r.test, method="cohort.count")
+baxterHMassoc <- baxterHEpi$massoc
+baxterHRR <- baxterHMassoc$RR.strata.score
+baxterHRRsig <- baxterHMassoc$chisq.strata
+## Risk Ratio = 1.14
+## CI = 0.70, 1.85
+## p-value = 0.608
+
+##Run the RR for B/F ratio
+Bacter = phyla.table.rel.abund$Bacteroidetes
+Firm = phyla.table.rel.abund$Firmicutes
+BFRatio = Bacter/Firm
+BFRatio <- as.data.frame(BFRatio)
+
+BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
+
+BFRatio.cat <- BFRatio$BFRatio.cat
+test4 <- cbind(BFRatio.cat, obese)
+test4 <- test4[order(BFRatio.cat), ]
+orderedBFCat <- test4[, 1]
+BaxBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
+BaxHighBFGroup <- as.data.frame(table(test4[c(1:BaxBFHighTotal), 2]))
+BaxLowBFGroup <- as.data.frame(table(test4[c((BaxBFHighTotal + 1):BaxtotalN), 2]))
+#Group1 (Higher than median), obese = 23 and non-obese = 63
+#Group2 (Lower than median), obese = 24 and non-obese = 62
+
+group1 <- c(BaxHighBFGroup[2, 2], BaxHighBFGroup[1, 2])
+group2 <- c(BaxLowBFGroup[2, 2], BaxLowBFGroup[1, 2])
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+baxterBFEpi <- epi.2by2(r.test, method="cohort.count")
+baxterBFMassoc <- baxterBFEpi$massoc
+baxterBFRR <- baxterBFMassoc$RR.strata.score
+baxterBFRRsig <- baxterBFMassoc$chisq.strata
+## Risk Ratio = 1.04
+## CI = 0.64, 1.70
+## p-value = 0.864
+
+###########################################################################
+############ Classification using AUCRF ###################################
+###########################################################################
+
+library(AUCRF)
+
+#generate test set
+  # get rid of values that only have 0 and something else
+testset <- Filter(function(x)(length(unique(x))>2), microbiome)
+  # get rid of those with 0 and only 4 other values
+testset <- Filter(function(x)(length(unique(x))>5), microbiome)
+
+#Need to add phyla and alpha diversity measures to the dataset
+
+testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
+testset <- cbind(obese, testset)
+
+#Try AUCRF with default measures provided in readme
+#set.seed(3)
+#baxterAUCfit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+
+```
+
+```{r Data_Ross, echo=FALSE, message=FALSE, warning=FALSE}
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+setwd("C:/users/marc/Desktop/obesity2/hispanic/")
+
+hispanic.microb <- read.table("hispanic.subsample.shared", header=T)
+
+metadata <- read.csv("s40168-015-0072-y-s1.csv")
+sample.match <- read.csv("Hispanic_dataset.csv")
+
+#Create a microbiome data table with sample names from metadata
+test <- cbind(as.character(sample.match$Run_s), as.character(sample.match$Library_Name_s))
+test <- test[order(test[, 1]), ]
+
+test2 <- cbind(test, hispanic.microb)
+rownames(test2) <- test2[, 2]
+his.microb.edit <- test2[, -c(1:5)]
+edit.metadata <- metadata[, -c(2:5)]
+rownames(edit.metadata) <- edit.metadata[, 1]
+edit.metadata <- edit.metadata[, -1]
+
+#Create a metadata file in the order of the microbiome data
+order1 <- rownames(his.microb.edit)
+edit.metadata2 <- edit.metadata[order1, ]
+
+#Get alpha diversity of the samples
+H <- diversity(his.microb.edit)
+S <- specnumber(his.microb.edit)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(his.microb.edit)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- his.microb.edit
+colnames(phyla.table) <- phyla.names
+  #add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+  #combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Deinococcus-Thermus", "Elusimicrobia", "Fusobacteria", "Lentisphaerae", "Synergistetes", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(3:4, 6:7, 9:10)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
+phyla.table.rel.abund <- as.data.frame(phyla.table.rel.abund[, -8])
+
+#Create BMI groups
+edit.metadata2$BMI.class[edit.metadata2$BMI<=24] <- "Normal"
+edit.metadata2$BMI.class[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
+edit.metadata2$BMI.class[edit.metadata2$BMI>=30 & edit.metadata2$BMI<40] <- "Obese"
+edit.metadata2$BMI.class[edit.metadata2$BMI>=40] <- "Extreme Obesity"
+
+#Create Obese Yes/No groups
+edit.metadata2$obese[edit.metadata2$BMI.class=="Normal" | edit.metadata2$BMI.class=="Overweight"] <- "No"
+edit.metadata2$obese[edit.metadata2$BMI.class=="Obese" | edit.metadata2$BMI.class=="Extreme Obesity"] <- "Yes"
+
+#Create BMI class 2 groups
+edit.metadata2$BMI.class2[edit.metadata2$BMI<=24] <- "Normal"
+edit.metadata2$BMI.class2[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
+edit.metadata2$BMI.class2[edit.metadata2$BMI>=30] <- "Obese"
+
+#Get paitent demographics to be tested
+
+bmi <- edit.metadata2$BMI
+obese <- factor(edit.metadata2$obese)
+
+######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
+###########################################################################
+
+rossH <- anova(lm(H ~ obese)) #P-value=0.1657
+rossS <- anova(lm(S ~ obese)) #P-value=0.1841
+rossJ <- anova(lm(J ~ obese)) #P-value=0.2538
+
+#B and F tests against obesity
+bacter <- phyla.table.rel.abund$Bacteroidetes
+firm <- phyla.table.rel.abund$Firmicutes
+BFratio <- bacter/firm
+
+rossBacter <- anova(lm(bacter ~ obese)) #P-value=0.2092
+rossFirm <- anova(lm(firm ~ obese)) #P-value=0.2569
+rossBF <- anova(lm(BFratio ~ obese)) #P-value=0.2508
+
+###########################################################################
+############ NMDS and PERMANOVA Analysis###################################
+###########################################################################
+
+set.seed(3)
+ross2 <- adonis(his.microb.edit ~ obese, permutations=1000)
+rossPERM <- ross2$aov.tab
+#PERMANOVA=0.7153, pseudo-F=0.73249
+
+###########################################################################
+############ Relative Risk#################################################
+###########################################################################
+
+library(epiR)
+
+#Generate median values and put them into existing alpha.test dataframe
+#Shannon diversity
+
+alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
+
+##Shannon Diversity
+H.cat <- alpha.test$shannon.cat
+bmi.cat <- as.character(obese)
+test3 <- cbind(H.cat, bmi.cat)
+test3 <- test3[order(H.cat), ]
+orderedHCat <- test3[, 1]
+RosstotalN <- length(orderedHCat)
+RossHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
+RossHighShannonGroup <- as.data.frame(table(test3[c(1:RossHHighTotal), 2]))
+RossLowShannonGroup <- as.data.frame(table(test3[c((RossHHighTotal + 1):RosstotalN), 2]))
+group1 <- c(RossHighShannonGroup[2, 2], RossHighShannonGroup[1, 2])
+group2 <- c(RossLowShannonGroup[2, 2], RossLowShannonGroup[1, 2])
+#Group1 (Higher than median), obese = 18 and non-obese = 13
+#Group2 (Lower than median), obese = 20 and non-obese = 12
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group1", "group2")
+
+rossHEpi <- epi.2by2(r.test, method="cohort.count")
+rossHMassoc <- rossHEpi$massoc
+rossHRR <- rossHMassoc$RR.strata.score
+rossHRRsig <- rossHMassoc$chisq.strata
+## Risk Ratio = 1.08
+## CI = 0.72, 1.61
+## p-value = 0.719
+
+##Run the RR for B/F ratio
+Bacter = phyla.table.rel.abund$Bacteroidetes
+Firm = phyla.table.rel.abund$Firmicutes
+BFRatio = Bacter/Firm
+BFRatio <- as.data.frame(BFRatio)
+
+BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
+
+BFRatio.cat <- BFRatio$BFRatio.cat
+test4 <- cbind(BFRatio.cat, obese)
+test4 <- test4[order(BFRatio.cat), ]
+orderedBFCat <- test4[, 1]
+RossBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
+RossHighBFGroup <- as.data.frame(table(test4[c(1:RossBFHighTotal), 2]))
+RossLowBFGroup <- as.data.frame(table(test4[c((RossBFHighTotal + 1):RosstotalN), 2]))
+group1 <- c(RossHighBFGroup[2, 2], RossHighBFGroup[1, 2])
+group2 <- c(RossLowBFGroup[2, 2], RossLowBFGroup[1, 2])
+#Group1 (Higher than median), obese = 22 and non-obese = 9
+#Group2 (Lower than median), obese = 16 and non-obese = 16
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+rossBFEpi <- epi.2by2(r.test, method="cohort.count")
+rossBFMassoc <- rossBFEpi$massoc
+rossBFRR <- rossBFMassoc$RR.strata.score
+rossBFRRsig <- rossBFMassoc$chisq.strata
+## Risk Ratio = 0.70
+## CI = 0.47, 1.07
+## p-value = 0.089
+
+###########################################################################
+############ Classification using AUCRF ###################################
+###########################################################################
+
+#Create Obese.num group
+edit.metadata2$obese.num[edit.metadata2$obese=="No"] <- 0
+edit.metadata2$obese.num[edit.metadata2$obese=="Yes"] <- 1
+obese <- factor(edit.metadata2$obese.num)
+
+#generate test set
+  # get rid of those with 0 and only 4 other values
+testset <- Filter(function(x)(length(unique(x))>5), his.microb.edit)
+testset <- cbind(obese, testset)
+colnames(testset)[1] <- "obese"
+testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
+
+#Try AUCRF with default measures provided in readme
+#set.seed(3)
+#rossAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+# list of 3 Measures, AUCopt = 0.7526
+
+```
+
+```{r Goodrich_data, echo=FALSE, message=FALSE, warning=FALSE}
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+setwd("C:/users/marc/Desktop/obesity2/twinsUK")
+
+#Read in and match metadata to microbiome data
+metadata <- read.csv("TwinsUKStudy_metadata.csv")
+rownames(metadata) <- metadata[, 1]
+shared.data <- read.table("combined.tx.1.subsample.shared", header=T)
+rownames(shared.data) <- shared.data[, 2]
+microbiome <- shared.data[, -c(1:3)]
+keep  <- rownames(metadata)
+
+microbiome<- microbiome[keep, ]
+
+rm(good.metadata, shared.data, keep)
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(microbiome)
+S <- specnumber(microbiome)
+J <- H/log(S)
+alpha.diversity <- as.data.frame(cbind(H, S, J))
+alpha.test <- alpha.diversity
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.table("phyla.txt", header=T)
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(microbiome)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- microbiome
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Chloroflexi", "Deferribacteres", "Elusimicrobia", "Fusobacteria", "Gemmatimonadetes", "Lentisphaerae", "OD1", "Planctomycetes", "Spirochaetes", "SR1", "Synergistetes", "Tenericutes")], 1, sum)
+phyla.table <- phyla.table[, -c(1, 4:6, 8:12, 14:17)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
+
+#Create BMI groups
+metadata$BMI.class[metadata$body_mass_index_s<=24] <- "Normal"
+metadata$BMI.class[metadata$body_mass_index_s>24 & metadata$body_mass_index_s<30] <- "Overweight"
+metadata$BMI.class[metadata$body_mass_index_s>=30 & metadata$body_mass_index_s<40] <- "Obese"
+metadata$BMI.class[metadata$body_mass_index_s>=40] <- "Extreme Obesity"
+
+#Create Obese Yes/No groups
+metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
+metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
+
+#Create a column with obese and extreme obese as single entity
+metadata$BMIclass2[metadata$BMI.class=="Normal"] <- "Normal"
+metadata$BMIclass2[metadata$BMI.class=="Overweight"] <- "Overweight"                     
+metadata$BMIclass2[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Obese"
+
+#Get paitent demographics to be tested
+bmi <- metadata$body_mass_index_s
+obese <- factor(metadata$obese)
+
+###########################################################################
+######### First Level Analysis & Alpha Diversity with BMI #################
+###########################################################################
+
+##Test BMI versus alpha diversity and phyla
+
+goodrichH <- anova(lm(H ~ obese)) #P-value=0.6671
+goodrichS <- anova(lm(S ~ obese)) #P-value=0.9073
+goodrichJ <- anova(lm(J ~ obese)) #P-value=0.6319
+
+#B and F tests against obesity
+bacter <- phyla.table.rel.abund$Bacteroidetes
+firm <- phyla.table.rel.abund$Firmicutes
+BFratio <- bacter/firm
+
+goodrichBacter <- anova(lm(bacter ~ obese)) #P-value=0.6661
+goodrichFirm <- anova(lm(firm ~ obese)) #P-value=0.9688
+goodrichBF <- anova(lm(BFratio ~ obese)) #P-value=0.8709
+
+###########################################################################
+############ NMDS and PERMANOVA Analysis###################################
+###########################################################################
+
+set.seed(3)
+goodrich2 <- adonis(microbiome ~ obese, permutations=1000) 
+goodrichPERM <- goodrich2$aov.tab
+#PERMANOVA=0.004995, pseudo-F=2.9308
+
+###########################################################################
+############ Relative Risk#################################################
+###########################################################################
+
+#Generate median values and put them into existing alpha.test dataframe
+#Shannon diversity
+
+alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
+
+##Shannon Diversity
+H.cat <- alpha.test$shannon.cat
+bmi.cat <- as.character(obese)
+test3 <- cbind(H.cat, bmi.cat)
+test3 <- test3[order(H.cat), ]
+orderedHCat <- test3[, 1]
+GoodtotalN <- length(orderedHCat)
+GoodHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
+GoodHighShannonGroup <- as.data.frame(table(test3[c(1:GoodHHighTotal), 2]))
+GoodLowShannonGroup <- as.data.frame(table(test3[c((GoodHHighTotal + 1):GoodtotalN), 2]))
+group1 <- c(GoodHighShannonGroup[2, 2], GoodHighShannonGroup[1, 2])
+group2 <- c(GoodLowShannonGroup[2, 2], GoodLowShannonGroup[1, 2])
+#Group1 (Higher than median), obese = 47 and non-obese = 206
+#Group2 (Lower than median), obese = 56 and non-obese = 198
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+goodrichHEpi <- epi.2by2(r.test, method="cohort.count")
+goodrichHMassoc <- goodrichHEpi$massoc
+goodrichHRR <- goodrichHMassoc$RR.strata.score
+goodrichHRRsig <- goodrichHMassoc$chisq.strata
+## Risk Ratio = 1.19
+## CI = 0.84, 1.68
+## p-value = 0.332
+
+##Run the RR for B/F ratio
+Bacter = phyla.table.rel.abund$Bacteroidetes
+Firm = phyla.table.rel.abund$Firmicutes
+BFRatio = Bacter/Firm
+BFRatio <- as.data.frame(BFRatio)
+
+BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
+
+BFRatio.cat <- BFRatio$BFRatio.cat
+test4 <- cbind(BFRatio.cat, obese)
+test4 <- test4[order(BFRatio.cat), ]
+orderedBFCat <- test4[, 1]
+GoodBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
+GoodHighBFGroup <- as.data.frame(table(test4[c(1:GoodBFHighTotal), 2]))
+GoodLowBFGroup <- as.data.frame(table(test4[c((GoodBFHighTotal + 1):GoodtotalN), 2]))
+group1 <- c(GoodHighBFGroup[2, 2], GoodHighBFGroup[1, 2])
+group2 <- c(GoodLowBFGroup[2, 2], GoodLowBFGroup[1, 2])
+#Group1 (Higher than median), obese = 52 and non-obese = 201
+#Group2 (Lower than median), obese = 51 and non-obese = 203
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+goodrichBFEpi <- epi.2by2(r.test, method="cohort.count")
+goodrichBFMassoc <- goodrichBFEpi$massoc
+goodrichBFRR <- goodrichBFMassoc$RR.strata.score
+goodrichBFRRsig <- goodrichBFMassoc$chisq.strata
+## Risk Ratio = 0.98
+## CI = 0.69, 1.38
+## p-value = 0.894
+
+###########################################################################
+############ Classification using AUCRF####################################
+###########################################################################
+
+#Create Obese.num group
+metadata$obese.num[metadata$obese=="No"] <- 0
+metadata$obese.num[metadata$obese=="Yes"] <- 1
+obese <- factor(metadata$obese.num)
+
+#generate test set
+# get rid of those with 0 and only 4 other values
+testset <- Filter(function(x)(length(unique(x))>5), microbiome)
+testset <- cbind(obese, testset)
+colnames(testset)[1] <- "obese"
+testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
+
+#Try AUCRF with default measures provided in readme
+#set.seed(3)
+#goodrichAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+# list of 8 Measures, AUCopt = 0.6767159
+
+```
+
+```{r escobar_data, echo=FALSE, message=FALSE, warning=FALSE}
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+setwd("C:/users/marc/Desktop/obesity2/columbian/")
+
+columbian.microb <- read.table("columbian.0.03.subsample.shared", header=T)
+metadata <- read.csv("columbian_dataset.csv")
+
+#Organize the microbiome shared data
+rownames(columbian.microb) <- columbian.microb[, 2]
+columbian.microb <- columbian.microb[, -c(1:3)]
+
+#Organize the metadata
+rownames(metadata) <- metadata[, 5]
+
+#Get only data that we are interested in
+edit.metadata <- metadata[, -c(1:7, 9, 13:15, 16:52)]
+
+#Sort metadata into the same order as the microbiome data
+order1 <- rownames(columbian.microb)
+edit.metadata2 <- edit.metadata[order1, ]
+
+#Get alpha diversity of the samples
+H <- diversity(columbian.microb)
+S <- specnumber(columbian.microb)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(columbian.microb)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- columbian.microb
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(4:5, 7:9)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
+phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
+
+#Create Obese Yes/No groups
+edit.metadata2$obese[edit.metadata2$description_s=="Adequate weight male stool sample" | edit.metadata2$description_s=="Overweight male stool sample" | edit.metadata2$description_s=="Adequate weight female stool sample" | edit.metadata2$description_s=="Overweight female stool sample"] <- "No"
+edit.metadata2$obese[edit.metadata2$description_s=="Obese male stool sample" | edit.metadata2$description_s=="Obese female stool sample"] <- "Yes"
+
+#Create BMI groups
+edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s<=24] <- "Normal"
+edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
+edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=30 & edit.metadata2$body_mass_index_s<40] <- "Obese"
+edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=40] <- "Extreme Obesity"
+
+#Create BMI groups2
+edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s<=24] <- "Normal"
+edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
+edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>=30] <- "Obese"
+
+#Get paitent demographics data to be tested 
+bmi <- edit.metadata2$body_mass_index_s
+obese <- factor(edit.metadata2$obese)
+
+######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
+###########################################################################
+
+##Test BMI versus alpha diversity and phyla
+
+escobarH <- anova(lm(H ~ obese)) #P-value=0.8636
+escobarS <- anova(lm(S ~ obese)) #P-value=0.1947
+escobarJ <- anova(lm(J ~ obese)) #P-value=0.4762
+
+#B and F tests against obesity
+bacter <- phyla.table.rel.abund$Bacteroidetes
+firm <- phyla.table.rel.abund$Firmicutes
+BFratio <- bacter/firm
+
+escobarBacter <- anova(lm(bacter ~ obese)) #P-value=0.09253
+escobarFirm <- anova(lm(firm ~ obese)) #P-value=0.3602
+escobarBF <- anova(lm(BFratio ~ obese)) #P-value=0.4663
+
+###########################################################################
+############ NMDS and PERMANOVA Analysis###################################
+###########################################################################
+
+set.seed(3)
+escobar2 <- adonis(columbian.microb ~ obese, permutations=1000) 
+escobarPERM <- escobar2$aov.tab
+#PERMANOVA=0.08891, pseudo-F=1.3797
+
+###########################################################################
+############ Relative Risk#################################################
+###########################################################################
+
+#Generate median values and put them into existing alpha.test dataframe
+#Shannon diversity
+
+alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
+
+##Shannon Diversity
+H.cat <- alpha.test$shannon.cat
+bmi.cat <- as.character(obese)
+test3 <- cbind(H.cat, bmi.cat)
+test3 <- test3[order(H.cat), ]
+orderedHCat <- test3[, 1]
+EscototalN <- length(orderedHCat)
+EscoHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
+EscoHighShannonGroup <- as.data.frame(table(test3[c(1:EscoHHighTotal), 2]))
+EscoLowShannonGroup <- as.data.frame(table(test3[c((EscoHHighTotal + 1):EscototalN), 2]))
+group1 <- c(EscoHighShannonGroup[2, 2], EscoHighShannonGroup[1, 2])
+group2 <- c(EscoLowShannonGroup[2, 2], EscoLowShannonGroup[1, 2])
+#Group1 (Higher than median), obese = 4 and non-obese = 11
+#Group2 (Lower than median), obese = 6 and non-obese = 9
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group1", "group2")
+
+escobarHEpi <- epi.2by2(r.test, method="cohort.count")
+escobarHMassoc <- escobarHEpi$massoc
+escobarHRR <- escobarHMassoc$RR.strata.score
+escobarHRRsig <- escobarHMassoc$chisq.strata
+## Risk Ratio = 1.50
+## CI = 0.53, 4.26
+## p-value = 0.439
+
+##Run the RR for B/F ratio
+Bacter = phyla.table.rel.abund$Bacteroidetes
+Firm = phyla.table.rel.abund$Firmicutes
+BFRatio = Bacter/Firm
+BFRatio <- as.data.frame(BFRatio)
+
+BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
+
+BFRatio.cat <- BFRatio$BFRatio.cat
+test4 <- cbind(BFRatio.cat, obese)
+test4 <- test4[order(BFRatio.cat), ]
+orderedBFCat <- test4[, 1]
+EscoBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
+EscoHighBFGroup <- as.data.frame(table(test4[c(1:EscoBFHighTotal), 2]))
+EscoLowBFGroup <- as.data.frame(table(test4[c((EscoBFHighTotal + 1):EscototalN), 2]))
+group1 <- c(EscoHighBFGroup[2, 2], EscoHighBFGroup[1, 2])
+group2 <- c(EscoLowBFGroup[2, 2], EscoLowBFGroup[1, 2])
+#Group1 (Higher than median), obese = 7 and non-obese = 8
+#Group2 (Lower than median), obese = 3 and non-obese = 12
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+escobarBFEpi <- epi.2by2(r.test, method="cohort.count")
+escobarBFMassoc <- escobarBFEpi$massoc
+escobarBFRR <- escobarBFMassoc$RR.strata.score
+escobarBFRRsig <- escobarBFMassoc$chisq.strata
+## Risk Ratio = 0.43
+## CI = 0.14, 1.35
+## p-value = 0.121
+
+###########################################################################
+############ Classification using AUCRF ###################################
+###########################################################################
+
+#Create Obese.num group
+edit.metadata2$obese.num[edit.metadata2$obese=="No"] <- 0
+edit.metadata2$obese.num[edit.metadata2$obese=="Yes"] <- 1
+obese <- factor(edit.metadata2$obese.num)
+
+#generate test set
+  # get rid of those with 0 and only 4 other values
+testset <- Filter(function(x)(length(unique(x))>5), columbian.microb)
+testset <- cbind(obese, testset)
+colnames(testset)[1] <- "obese"
+testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
+
+#Try AUCRF with default measures provided in readme
+#set.seed(3)
+#escobarAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+# list of 3 Measures, AUCopt = 0.95
+
+```
+
+```{r Zupancic_data, echo=FALSE, message=FALSE, warning=FALSE}
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+setwd("C:/users/marc/Desktop/obesity2/Amish/updatedGOOD")
+
+#Read in and match metadata to microbiome data
+metadata <- read.csv("amish_obesity_table2.csv")
+test <- metadata[!duplicated(metadata$submitted_sample_id_s), ]
+rownames(test) <- test[, 4]
+shared.data <- read.table("combined.0.03.subsample.shared", header=T)
+rownames(shared.data) <- shared.data[, 2]
+shared.data <- shared.data[, -c(1:3)]
+keep  <- rownames(shared.data)
+
+test2 <- test[keep, ]
+good.metadata <- test2
+test3 <- test2[!duplicated(test2$submitted_subject_id_s), ]
+keep <- rownames(test3)
+microbiome <- shared.data[keep, ]
+metadata <- test3
+
+rm(good.metadata, shared.data, keep, test, test2, test3)
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(microbiome)
+S <- specnumber(microbiome)
+J <- H/log(S)
+select.alpha.diversity <- as.data.frame(cbind(H, S, J))
+s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
+alpha.test <- s1.alpha.diversity
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(microbiome)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- microbiome
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Spirochaetes")], 1, sum)
+phyla.table <- phyla.table[, -c(4, 6)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:6)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
+
+#Create Obese Yes/No groups
+# One problem is that we are assuming those with "not provided" are defaulting to the not affected group
+metadata$obese[metadata$subject_is_affected_s=="<not provided>" | metadata$subject_is_affected_s=="No"] <- "No"
+metadata$obese[metadata$subject_is_affected_s=="Yes"] <- "Yes"
+
+#Create Obese.num group
+metadata$obese.num[metadata$obese=="No"] <- 0
+metadata$obese.num[metadata$obese=="Yes"] <- 1
+
+#create groups to be used
+obese <- factor(metadata$obese)
+
+####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
+###########################################################################
+###########################################################################
+
+##Test BMI versus alpha diversity and phyla
+
+zupancicH <- anova(lm(H ~ obese)) #P-value=0.2034
+zupancicS <- anova(lm(S ~ obese)) #P-value=0.03516
+zupancicJ <- anova(lm(J ~ obese)) #P-value=0.5369
+
+#B and F tests against obesity
+bacter <- phyla.table.rel.abund$Bacteroidetes
+firm <- phyla.table.rel.abund$Firmicutes
+BFratio <- bacter/firm
+
+zupancicBacter <- anova(lm(bacter ~ obese)) #P-value=0.9582
+zupancicFirm <- anova(lm(firm ~ obese)) #P-value=0.2654
+zupancicBF <- anova(lm(BFratio ~ obese)) #P-value=0.9899
+
+####################################################################################### NMDS and PERMANOVA Analysis###################################
+###########################################################################
+
+set.seed(3)
+zupancic2 <- adonis(microbiome ~ obese, permutations=1000) 
+zupancicPERM <- zupancic2$aov.tab
+#PERMANOVA=0.5734, pseudo-F=0.95626
+
+###########################################################################
+############ Relative Risk#################################################
+###########################################################################
+
+#Generate median values and put them into existing alpha.test dataframe
+#Shannon diversity
+
+alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
+
+##Shannon Diversity
+H.cat <- alpha.test$shannon.cat
+bmi.cat <- as.character(obese)
+test3 <- cbind(H.cat, bmi.cat)
+test3 <- test3[order(H.cat), ]
+orderedHCat <- test3[, 1]
+ZupatotalN <- length(orderedHCat)
+ZupaHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
+ZupaHighShannonGroup <- as.data.frame(table(test3[c(1:ZupaHHighTotal), 2]))
+ZupaLowShannonGroup <- as.data.frame(table(test3[c((ZupaHHighTotal + 1):ZupatotalN), 2]))
+group1 <- c(ZupaHighShannonGroup[2, 2], ZupaHighShannonGroup[1, 2])
+group2 <- c(ZupaLowShannonGroup[2, 2], ZupaLowShannonGroup[1, 2])
+#Group1 (Higher than median), obese = 31 and non-obese = 72
+#Group2 (Lower than median), obese = 47 and non-obese = 57
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+zupancicHEpi <- epi.2by2(r.test, method="cohort.count")
+zupancicHMassoc <- zupancicHEpi$massoc
+zupancicHRR <- zupancicHMassoc$RR.strata.score
+zupancicHRRsig <- zupancicHMassoc$chisq.strata
+## Risk Ratio = 1.50
+## CI = 1.04, 2.16
+## p-value = 0.025
+
+##Run the RR for B/F ratio
+Bacter = phyla.table.rel.abund$Bacteroidetes
+Firm = phyla.table.rel.abund$Firmicutes
+BFRatio = Bacter/Firm
+BFRatio <- as.data.frame(BFRatio)
+
+BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
+
+BFRatio.cat <- BFRatio$BFRatio.cat
+test4 <- cbind(BFRatio.cat, obese)
+test4 <- test4[order(BFRatio.cat), ]
+orderedBFCat <- test4[, 1]
+ZupaBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
+ZupaHighBFGroup <- as.data.frame(table(test4[c(1:ZupaBFHighTotal), 2]))
+ZupaLowBFGroup <- as.data.frame(table(test4[c((ZupaBFHighTotal + 1):ZupatotalN), 2]))
+group1 <- c(ZupaHighBFGroup[2, 2], ZupaHighBFGroup[1, 2])
+group2 <- c(ZupaLowBFGroup[2, 2], ZupaLowBFGroup[1, 2])
+#Group1 (Higher than median), obese = 41 and non-obese = 62
+#Group2 (Lower than median), obese = 37 and non-obese = 67
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+zupancicBFEpi <- epi.2by2(r.test, method="cohort.count")
+zupancicBFMassoc <- zupancicBFEpi$massoc
+zupancicBFRR <- zupancicBFMassoc$RR.strata.score
+zupancicBFRRsig <- zupancicBFMassoc$chisq.strata
+## Risk Ratio = 0.89
+## CI = 0.63, 1.27
+## p-value = 0.53
+
+###########################################################################
+############ Classification using AUCRF ###################################
+###########################################################################
+
+#Create Obese.num group
+obese <- factor(metadata$obese.num)
+
+#generate test set
+# get rid of those with 0 and only 4 other values
+testset <- Filter(function(x)(length(unique(x))>5), microbiome)
+testset <- cbind(obese, testset)
+colnames(testset)[1] <- "obese"
+testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
+
+#Try AUCRF with default measures provided in readme
+#set.seed(3)
+#zupancicAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+# list of 42 Measures, AUCopt = 0.6365534
+
+```
+
+```{r HMP_data, echo=FALSE, message=FALSE, warning=FALSE}
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+##Read in relevant data
+
+setwd("C:/users/marc/Desktop/obesity2/HMP.analysis")
+microbiome <- read.table("Stool.an.0.03.subsample.shared", header=T)
+meta.cat <- read.table("categorical.metadata", header=T)
+meta.cont <- read.table("continuous.metadata", header=T)
+
+#Only interested in first visit microbiome data
+#need to create microbiome data set for only that
+
+test <- microbiome[grep("\\.01\\.", microbiome$Group), ] #selecting by .01.
+
+microb.rownames <- gsub("([0-9]+).*", "\\1", test$Group) #extracting only numeric before first "."
+rownames(test) <- microb.rownames
+
+#Get rid of information not used in downstram analysis
+test <- test[, -c(1:3)]
+
+#Subset data to the microbiome total n
+select.meta.cat <- meta.cat[microb.rownames, ]
+select.meta.cont <- meta.cont[microb.rownames, ]
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(test)
+S <- specnumber(test)
+J <- H/log(S)
+select.alpha.diversity <- as.data.frame(cbind(H, S, J))
+alpha.test <- select.alpha.diversity
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.csv("phyla.data.csv")
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(test)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- test
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(1, 4, 6, 7, 9, 10:12)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
+phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
+
+#Create Obese Yes/No groups
+select.meta.cat$obese[select.meta.cat$BMI_C=="normal" | select.meta.cat$BMI_C=="overweight"] <- "No"
+select.meta.cat$obese[select.meta.cat$BMI_C=="obese" | select.meta.cat$BMI_C=="extreme obesity"] <- "Yes"
+
+#Create BMI groups2
+select.meta.cat$BMI.class2[select.meta.cont$DTPBMI<=24] <- "Normal"
+select.meta.cat$BMI.class2[select.meta.cont$DTPBMI>24 & select.meta.cont$DTPBMI<30] <- "Overweight"
+select.meta.cat$BMI.class2[select.meta.cont$DTPBMI>=30] <- "Obese"
+
+#Get paitent demographics data to be tested 
+bmi <- select.meta.cont$DTPBMI
+obese <- factor(select.meta.cat$obese)
+
+######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
+###########################################################################
+
+##Test BMI versus alpha diversity and phyla
+
+HMPH <- anova(lm(H ~ obese)) #P-value=0.585
+HMPS <- anova(lm(S ~ obese)) #P-value=0.845
+HMPJ <- anova(lm(J ~ obese)) #P-value=0.3559
+
+#B and F tests against obesity
+bacter <- phyla.table.rel.abund$Bacteroidetes
+firm <- phyla.table.rel.abund$Firmicutes
+BFratio <- bacter/firm
+
+HMPBacter <- anova(lm(bacter ~ obese)) #P-value=0.3518
+HMPFirm <- anova(lm(firm ~ obese)) #P-value=0.636
+HMPBF <- anova(lm(BFratio ~ obese)) #P-value=0.5688
+
+###########################################################################
+############ NMDS and PERMANOVA Analysis###################################
+###########################################################################
+
+set.seed(3)
+HMP2 <- adonis(test ~ obese, permutations=1000) 
+HMPPERM <- HMP2$aov.tab
+#PERMANOVA=0.8112, pseudo-F=0.7024
+
+######################################################################################## Relative Risk ###############################################
+###########################################################################
+
+#Generate median values and put them into existing alpha.test dataframe
+#Shannon diversity
+
+alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
+
+##Shannon Diversity
+H.cat <- alpha.test$shannon.cat
+bmi.cat <- as.character(obese)
+test3 <- cbind(H.cat, bmi.cat)
+test3 <- test3[order(H.cat), ]
+orderedHCat <- test3[, 1]
+HMPtotalN <- length(orderedHCat)
+HMPHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
+HMPHighShannonGroup <- as.data.frame(table(test3[c(1:HMPHHighTotal), 2]))
+HMPLowShannonGroup <- as.data.frame(table(test3[c((HMPHHighTotal + 1):HMPtotalN), 2]))
+group1 <- c(HMPHighShannonGroup[2, 2], HMPHighShannonGroup[1, 2])
+group2 <- c(HMPLowShannonGroup[2, 2], HMPLowShannonGroup[1, 2])
+#Group1 (Higher than median), obese = 10 and non-obese = 118
+#Group2 (Lower than median), obese = 16 and non-obese = 112
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group1", "group2")
+
+HMPHEpi <- epi.2by2(r.test, method="cohort.count")
+HMPHMassoc <- HMPHEpi$massoc
+HMPHRR <- HMPHMassoc$RR.strata.score
+HMPHRRsig <- HMPHMassoc$chisq.strata
+## Risk Ratio = 1.60
+## CI = 0.75, 3.39
+## p-value = 0.214
+
+##Run the RR for B/F ratio
+Bacter = phyla.table.rel.abund$Bacteroidetes
+Firm = phyla.table.rel.abund$Firmicutes
+BFRatio = Bacter/Firm
+BFRatio <- as.data.frame(BFRatio)
+
+BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
+
+BFRatio.cat <- BFRatio$BFRatio.cat
+test4 <- cbind(BFRatio.cat, obese)
+test4 <- test4[order(BFRatio.cat), ]
+orderedBFCat <- test4[, 1]
+HMPBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
+HMPHighBFGroup <- as.data.frame(table(test4[c(1:HMPBFHighTotal), 2]))
+HMPLowBFGroup <- as.data.frame(table(test4[c((HMPBFHighTotal + 1):HMPtotalN), 2]))
+group1 <- c(HMPHighBFGroup[2, 2], HMPHighBFGroup[1, 2])
+group2 <- c(HMPLowBFGroup[2, 2], HMPLowBFGroup[1, 2])
+#Group1 (Higher than median), obese = 13 and non-obese = 115
+#Group2 (Lower than median), obese = 13 and non-obese = 115
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+HMPBFEpi <- epi.2by2(r.test, method="cohort.count")
+HMPBFMassoc <- HMPBFEpi$massoc
+HMPBFRR <- HMPBFMassoc$RR.strata.score
+HMPBFRRsig <- HMPBFMassoc$chisq.strata
+## Risk Ratio = 1.00
+## CI = 0.48, 2.07
+## p-value = 1.00
+
+###########################################################################
+############ Classification using AUCRF ###################################
+###########################################################################
+
+#Create Obese.num group
+select.meta.cat$obese.num[select.meta.cat$obese=="No"] <- 0
+select.meta.cat$obese.num[select.meta.cat$obese=="Yes"] <- 1
+obese <- factor(select.meta.cat$obese.num)
+
+#generate test set
+  # get rid of those with 0 and only 4 other values
+testset <- Filter(function(x)(length(unique(x))>5), test)
+testset <- cbind(obese, testset)
+colnames(testset)[1] <- "obese"
+testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
+
+#Try AUCRF with default measures provided in readme
+#set.seed(3)
+#HMPAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+# list of 6 Measures, AUCopt = 0.7031773
+
+```
+
+```{r Wu_data, echo=FALSE, message=FALSE, warning=FALSE}
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+setwd("C:/users/marc/Desktop/obesity2/COMBO/")
+
+microbiome <- read.table("combined.good.subsample.shared", header=T)
+rownames(microbiome) <- microbiome$Group
+microbiome <- microbiome[, -c(1:3)]
+seqData <- read.csv("COMBO_data_table.csv", header=T)
+seqData <- seqData[-1, ]
+rownames(seqData) <- seqData$Run_s
+metadata <- read.table("BMI_and_Counts.txt", header=T)
+metadata <- metadata[, c(1, 2)]
+
+#get seqData set in line with microbiome data
+namesToKeep <- rownames(microbiome)
+test <- seqData[namesToKeep, ]
+seqData <- test
+rm(test)
+
+#change all rownames to match sample IDs
+rownames(seqData) <- seqData$submitted_subject_id_s
+rownames(microbiome) <- rownames(seqData)
+rownames(metadata) <- metadata[, 1]
+
+#Match the metadata now with the microbiome data
+namesToKeep <- rownames(microbiome)
+test <- metadata[namesToKeep, ]
+metadata <- test
+rm(test)
+rm(seqData, keep, namesToKeep)
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(microbiome)
+S <- specnumber(microbiome)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+
+#Get phyla information
+#Edited out non phyla information first with sed in linux
+#combined new labels with previous taxonomy file with excel
+phylogenetic.info <- read.table("combined.good.phyla.txt", header=T)
+rownames(phylogenetic.info) <- phylogenetic.info[,1]
+phylogenetic.info <- phylogenetic.info[,-c(1)]
+phyla.names <- as.character(phylogenetic.info$Taxonomy)
+keep <- colnames(microbiome)
+phyla.good <- phylogenetic.info[keep, ]
+phyla.names <- as.character(phyla.good[,2])
+phyla.table <- microbiome
+colnames(phyla.table) <- phyla.names
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Synergistetes", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(4, 6:7)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:6)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
+rm(keep, phyla.names, phyla.total, phylogenetic.info, phyla.table, phyla.good)
+
+#Create BMI groups
+metadata$BMI.class[metadata$BMI<=24] <- "Normal"
+metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
+metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
+metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
+
+#Create Obese Yes/No groups
+metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
+metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
+
+#Create a column with obese and extreme obese as single entity
+metadata$BMIclass2[metadata$BMI.class=="Normal"] <- "Normal"
+metadata$BMIclass2[metadata$BMI.class=="Overweight"] <- "Overweight"                     
+metadata$BMIclass2[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Obese"
+
+#Get paitent demographics to be tested
+bmi <- metadata$BMI
+obese <- factor(metadata$obese)
+
+###########################################################################
+######### First Level Analysis & Alpha Diversity with BMI #################
+###########################################################################
+
+##Test BMI versus alpha diversity and phyla
+
+WuH <- anova(lm(H ~ obese)) #P-value=0.7493
+WuS <- anova(lm(S ~ obese)) #P-value=0.7511
+WuJ <- anova(lm(J ~ obese)) #P-value=0.5557
+
+#B and F tests against obesity
+bacter <- phyla.table.rel.abund$Bacteroidetes
+firm <- phyla.table.rel.abund$Firmicutes
+BFratio <- bacter/firm
+
+WuBacter <- anova(lm(bacter ~ obese)) #P-value=0.9173
+WuFirm <- anova(lm(firm ~ obese)) #P-value=0.5765
+WuBF <- anova(lm(BFratio ~ obese)) #P-value=0.6735
+
+###########################################################################
+############ NMDS and PERMANOVA Analysis###################################
+###########################################################################
+
+set.seed(3)
+Wu2 <- adonis(microbiome ~ obese, permutations=1000)
+WuPERM <- Wu2$aov.tab
+#PERMANOVA=0.8921, pseudo-F=0.66065
+
+###########################################################################
+############ Relative Risk#################################################
+###########################################################################
+
+#Generate median values and put them into existing alpha.test dataframe
+#Shannon diversity
+
+alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
+
+##Shannon Diversity
+H.cat <- alpha.test$shannon.cat
+bmi.cat <- as.character(obese)
+test3 <- cbind(H.cat, bmi.cat)
+test3 <- test3[order(H.cat), ]
+orderedHCat <- test3[, 1]
+WutotalN <- length(orderedHCat)
+WuHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
+WuHighShannonGroup <- as.data.frame(table(test3[c(1:WuHHighTotal), 2]))
+WuLowShannonGroup <- as.data.frame(table(test3[c((WuHHighTotal + 1):WutotalN), 2]))
+group1 <- c(WuHighShannonGroup[2, 2], WuHighShannonGroup[1, 2])
+group2 <- c(WuLowShannonGroup[2, 2], WuLowShannonGroup[1, 2])
+#Group1 (Higher than median), obese = 3 and non-obese = 26
+#Group2 (Lower than median), obese = 2 and non-obese = 27
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+WuHEpi <- epi.2by2(r.test, method="cohort.count")
+WuHMassoc <- WuHEpi$massoc
+WuHRR <- WuHMassoc$RR.strata.score
+WuHRRsig <- WuHMassoc$chisq.strata
+## Risk Ratio = 0.67
+## CI = 0.12, 3.70
+## p-value = 0.64
+
+##Run the RR for B/F ratio
+Bacter = phyla.table.rel.abund$Bacteroidetes
+Firm = phyla.table.rel.abund$Firmicutes
+BFRatio = Bacter/Firm
+BFRatio <- as.data.frame(BFRatio)
+
+BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
+
+BFRatio.cat <- BFRatio$BFRatio.cat
+test4 <- cbind(BFRatio.cat, obese)
+test4 <- test4[order(BFRatio.cat), ]
+orderedBFCat <- test4[, 1]
+WuBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
+WuHighBFGroup <- as.data.frame(table(test4[c(1:WuBFHighTotal), 2]))
+WuLowBFGroup <- as.data.frame(table(test4[c((WuBFHighTotal + 1):WutotalN), 2]))
+group1 <- c(WuHighBFGroup[2, 2], WuHighBFGroup[1, 2])
+group2 <- c(WuLowBFGroup[2, 2], WuLowBFGroup[1, 2])
+#Group1 (Higher than median), obese = 3 and non-obese = 26
+#Group2 (Lower than median), obese = 2 and non-obese = 27
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+WuBFEpi <- epi.2by2(r.test, method="cohort.count")
+WuBFMassoc <- WuBFEpi$massoc
+WuBFRR <- WuBFMassoc$RR.strata.score
+WuBFRRsig <- WuBFMassoc$chisq.strata
+## Risk Ratio = 0.67
+## CI = 0.12, 3.70
+## p-value = 0.64
+
+###########################################################################
+############ Classification using AUCRF####################################
+###########################################################################
+
+#Create Obese.num group
+metadata$obese.num[metadata$obese=="No"] <- 0
+metadata$obese.num[metadata$obese=="Yes"] <- 1
+obese <- factor(metadata$obese.num)
+
+#generate test set
+# get rid of those with 0 and only 4 other values
+testset <- Filter(function(x)(length(unique(x))>5), microbiome)
+testset <- cbind(obese, testset)
+colnames(testset)[1] <- "obese"
+testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
+
+#Try AUCRF with default measures provided in readme
+#set.seed(3)
+#WuAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+# list of 15 Measures, AUCopt = 0.7075472
+
+```
+
+```{r Arumugam_data, echo=FALSE, message=FALSE, warning=FALSE}
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+setwd("C:/users/marc/Desktop/obesity2/MetaHit/")
+
+microb <- read.csv("finalized_metaHit_shared.csv")
+rownames(microb) <- microb[, 1]
+microb <- microb[,-1]
+
+metadata <- read.csv("MetaHit_metadata.csv")
+rownames(metadata) <- metadata[, 1]
+metadata <- metadata[, -1]
+
+phyla <- read.csv("phyla.csv")
+rownames(phyla) <- phyla[, 1]
+phyla <- phyla[, -1]
+
+# Generate Overall Relative Abundance - normalize to a value of 100
+# Needs to be done since not all measures have 100 bacteria from metaphlan2
+overall <- rowSums(microb)
+microb.norm <- (microb / overall) * 100
+
+phyla.table <- read.csv("phyla.csv", header = T)
+rownames(phyla.table) <- phyla.table[, 1]
+phyla.table <- phyla.table[, -1]
+phyla.table <- as.data.frame(t(phyla.table))
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:10)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
+rm(phyla.table, phyla.info)
+
+#Get alpha diversity of the samples
+library(vegan)
+
+H <- diversity(microb.norm)
+S <- specnumber(microb.norm)
+J <- H/log(S)
+alpha.diversity.shannon <- cbind(H,S,J)
+alpha.test <- as.data.frame(alpha.diversity.shannon)
+# Seems to look okay....
+
+#Create BMI groups
+metadata$BMI.class[metadata$BMI<=24] <- "Normal"
+metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
+metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
+metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
+
+#Create Obese Yes/No groups
+metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
+metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
+
+#Create Obese.num group
+metadata$obese.num[metadata$obese=="No"] <- 0
+metadata$obese.num[metadata$obese=="Yes"] <- 1
+
+#Create a column with obese and extreme obese as single entity
+metadata$BMIclass2[metadata$BMI<=24] <- "Normal"
+metadata$BMIclass2[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"      
+metadata$BMIclass2[metadata$BMI>=30] <- "Obese"
+
+#Get paitent demographics to be tested
+bmi <- metadata$BMI
+obese <- factor(metadata$obese)
+
+####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
+###########################################################################
+###########################################################################
+
+##Test BMI versus alpha diversity and phyla
+
+ArumugamH <- anova(lm(H ~ obese)) #P-value=0.2174
+ArumugamS <- anova(lm(S ~ obese)) #P-value=0.9878
+ArumugamJ <- anova(lm(J ~ obese)) #P-value=0.1874
+
+#B and F tests against obesity
+bacter <- phyla.table.rel.abund$Bacteroidetes
+firm <- phyla.table.rel.abund$Firmicutes
+BFratio <- bacter/firm
+
+ArumugamBacter <- anova(lm(bacter ~ obese)) #P-value=0.1911
+ArumugamFirm <- anova(lm(firm ~ obese)) #P-value=0.2925
+ArumugamBF <- anova(lm(BFratio ~ obese)) #P-value=0.05254
+
+####################################################################################### NMDS and PERMANOVA Analysis###################################
+###########################################################################
+
+set.seed(3)
+Arumugam2 <- adonis(microb.norm ~ obese, permutations=1000) 
+ArumugamPERM <- Arumugam2$aov.tab
+#PERMANOVA=0.05495, pseudo-F=1.6047
+
+###########################################################################
+############ Relative Risk#################################################
+###########################################################################
+
+#Generate median values and put them into existing alpha.test dataframe
+#Shannon diversity
+
+alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
+
+##Shannon Diversity
+H.cat <- alpha.test$shannon.cat
+bmi.cat <- as.character(obese)
+test3 <- cbind(H.cat, bmi.cat)
+test3 <- test3[order(H.cat), ]
+orderedHCat <- test3[, 1]
+ArutotalN <- length(orderedHCat)
+AruHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
+AruHighShannonGroup <- as.data.frame(table(test3[c(1:AruHHighTotal), 2]))
+AruLowShannonGroup <- as.data.frame(table(test3[c((AruHHighTotal + 1):ArutotalN), 2]))
+group1 <- c(AruHighShannonGroup[2, 2], AruHighShannonGroup[1, 2])
+group2 <- c(AruLowShannonGroup[2, 2], AruLowShannonGroup[1, 2])
+#Group1 (Higher than median), obese = 17 and non-obese = 25
+#Group2 (Lower than median), obese = 20 and non-obese = 23
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+ArumugamHEpi <- epi.2by2(r.test, method="cohort.count")
+ArumugamHMassoc <- ArumugamHEpi$massoc
+ArumugamHRR <- ArumugamHMassoc$RR.strata.score
+ArumugamHRRsig <- ArumugamHMassoc$chisq.strata
+## Risk Ratio = 1.15
+## CI = 0.71, 1.87
+## p-value = 0.575
+
+##Run the RR for B/F ratio
+Bacter = phyla.table.rel.abund$Bacteroidetes
+Firm = phyla.table.rel.abund$Firmicutes
+BFRatio = Bacter/Firm
+BFRatio <- as.data.frame(BFRatio)
+
+BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
+
+BFRatio.cat <- BFRatio$BFRatio.cat
+test4 <- cbind(BFRatio.cat, obese)
+test4 <- test4[order(BFRatio.cat), ]
+orderedBFCat <- test4[, 1]
+AruBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
+AruHighBFGroup <- as.data.frame(table(test4[c(1:AruBFHighTotal), 2]))
+AruLowBFGroup <- as.data.frame(table(test4[c((AruBFHighTotal + 1):ArutotalN), 2]))
+group1 <- c(AruHighBFGroup[2, 2], AruHighBFGroup[1, 2])
+group2 <- c(AruLowBFGroup[2, 2], AruLowBFGroup[1, 2])
+#Group1 (Higher than median), obese = 20 and non-obese = 22
+#Group2 (Lower than median), obese = 17 and non-obese = 26
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+ArumugamBFEpi <- epi.2by2(r.test, method="cohort.count")
+ArumugamBFMassoc <- ArumugamBFEpi$massoc
+ArumugamBFRR <- ArumugamBFMassoc$RR.strata.score
+ArumugamBFRRsig <- ArumugamBFMassoc$chisq.strata
+## Risk Ratio = 0.83
+## CI = 0.51, 1.35
+## p-value = 0.452
+
+###########################################################################
+############ Classification using AUCRF ###################################
+###########################################################################
+
+library(AUCRF)
+
+#Create Obese.num group
+metadata$obese.num[metadata$obese=="No"] <- 0
+metadata$obese.num[metadata$obese=="Yes"] <- 1
+obese <- factor(metadata$obese.num)
+
+#generate test set
+# get rid of those with 0 and only 4 other values
+testset <- Filter(function(x)(length(unique(x))>5), microb.norm)
+testset <- cbind(obese, testset)
+colnames(testset)[1] <- "obese"
+testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
+
+#Try AUCRF with default measures provided in readme
+#set.seed(3)
+#ArumugamAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+# list of 19 Measures, AUCopt = 0.7649212
+
+```
+
+```{r Turnbaugh_data, echo=FALSE, message=FALSE, warning=FALSE}
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+setwd("C:/users/marc/Desktop/obesity2/turnbaugh.twins")
+shared.data <- read.table("test.unique.good.filter.unique.precluster.pick.pick.an.shared", header=T)
+rownames(shared.data) <- shared.data[, 2]
+shared.data <- shared.data[, -c(1:3)]
+
+subsample.data <- read.table("test.unique.good.filter.unique.precluster.pick.pick.an.0.03.subsample.shared", header=T)
+rownames(subsample.data) <- subsample.data[, 2]
+subsample.data <- subsample.data[, -c(1:3)]
+
+metadata <- read.csv("turnbaugh.metadata.csv")
+rownames(metadata) <- metadata[, 4]
+metadata <- metadata[, -4]
+
+keep1 <- which(metadata$Sample == 1) # n =146
+
+#use the first sampling
+#subset data for only the first sampling
+s1.metadata <- metadata[keep1, ]
+s1.subsample.data <- subsample.data[keep1, ]
+
+#generate alpha diversity measures with vegan
+library(vegan)
+H <- diversity(s1.subsample.data)
+S <- specnumber(s1.subsample.data)
+J <- H/log(S)
+select.alpha.diversity <- as.data.frame(cbind(H, S, J))
+s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
+alpha.test <- s1.alpha.diversity
+
+#Generate phyla table data
+
+phyla.info <- read.csv("phyla.csv")
+phyla.table <- shared.data
+colnames(phyla.table) <- phyla.info[, 2]
+
+#add all the same columns up and then return the sum
+testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
+phyla.table <- as.data.frame(testing)
+rm(testing)
+
+#combine phyla that are not that abundant
+phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(4:5, 7:9)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
+s1.phyla.rel.abund <- phyla.table.rel.abund[rownames(shared.data), ]
+s1.phyla.rel.abund <- s1.phyla.rel.abund[keep1, ]
+
+#Create Obese Yes/No groups
+s1.metadata$obese[s1.metadata$BMI.category=="Lean" | s1.metadata$BMI.category=="Overweight"] <- "No"
+s1.metadata$obese[s1.metadata$BMI.category=="Obese"] <- "Yes"
+
+#Create Obese.num group
+s1.metadata$obese.num[s1.metadata$obese=="No"] <- 0
+s1.metadata$obese.num[s1.metadata$obese=="Yes"] <- 1
+
+#create groups to be used
+obese <- factor(s1.metadata$obese)
+bmi <- s1.metadata$BMI.category
+
+####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
+###########################################################################
+###########################################################################
+
+##Test BMI versus alpha diversity and phyla
+
+turnbaughH <- anova(lm(H ~ obese)) #P-value=0.06976
+turnbaughS <- anova(lm(S ~ obese)) #P-value=0.02435
+turnbaughJ <- anova(lm(J ~ obese)) #P-value=0.1253
+
+#B and F tests against obesity
+bacter <- s1.phyla.rel.abund$Bacteroidetes
+firm <- s1.phyla.rel.abund$Firmicutes
+BFratio <- bacter/firm
+
+turnbaughBacter <- anova(lm(bacter ~ obese)) #P-value=0.5771
+turnbaughFirm <- anova(lm(firm ~ obese)) #P-value=0.4073
+turnbaughBF <- anova(lm(BFratio ~ obese)) #P-value=0.2216
+
+####################################################################################### NMDS and PERMANOVA Analysis###################################
+###########################################################################
+
+set.seed(3)
+turnbaugh2 <- adonis(s1.subsample.data ~ obese, permutations=1000) 
+turnbaughPERM <- turnbaugh2$aov.tab
+#PERMANOVA=0.09491, pseudo-F=1.2114
+
+###########################################################################
+############ Relative Risk#################################################
+###########################################################################
+
+#Generate median values and put them into existing alpha.test dataframe
+#Shannon diversity
+
+alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
+
+##Shannon Diversity
+H.cat <- alpha.test$shannon.cat
+bmi.cat <- as.character(obese)
+test3 <- cbind(H.cat, bmi.cat)
+test3 <- test3[order(H.cat), ]
+orderedHCat <- test3[, 1]
+TurntotalN <- length(orderedHCat)
+TurnHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
+TurnHighShannonGroup <- as.data.frame(table(test3[c(1:TurnHHighTotal), 2]))
+TurnLowShannonGroup <- as.data.frame(table(test3[c((TurnHHighTotal + 1):TurntotalN), 2]))
+group1 <- c(TurnHighShannonGroup[2, 2], TurnHighShannonGroup[1, 2])
+group2 <- c(TurnLowShannonGroup[2, 2], TurnLowShannonGroup[1, 2])
+#Group1 (Higher than median), obese = 47 and non-obese = 26
+#Group2 (Lower than median), obese = 52 and non-obese = 21
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+turnbaughHEpi <- epi.2by2(r.test, method="cohort.count")
+turnbaughHMassoc <- turnbaughHEpi$massoc
+turnbaughHRR <- turnbaughHMassoc$RR.strata.score
+turnbaughHRRsig <- turnbaughHMassoc$chisq.strata
+## Risk Ratio = 1.11
+## CI = 0.88, 1.38
+## p-value = 0.376
+
+##Run the RR for B/F ratio
+Bacter = s1.phyla.rel.abund$Bacteroidetes
+Firm = s1.phyla.rel.abund$Firmicutes
+BFRatio = Bacter/Firm
+BFRatio <- as.data.frame(BFRatio)
+
+BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
+
+BFRatio.cat <- BFRatio$BFRatio.cat
+test4 <- cbind(BFRatio.cat, obese)
+test4 <- test4[order(BFRatio.cat), ]
+orderedBFCat <- test4[, 1]
+TurnBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
+TurnHighBFGroup <- as.data.frame(table(test4[c(1:TurnBFHighTotal), 2]))
+TurnLowBFGroup <- as.data.frame(table(test4[c((TurnBFHighTotal + 1):TurntotalN), 2]))
+group1 <- c(TurnHighBFGroup[2, 2], TurnHighBFGroup[1, 2])
+group2 <- c(TurnLowBFGroup[2, 2], TurnLowBFGroup[1, 2])
+#Group1 (Higher than median), obese = 54 and non-obese = 19
+#Group2 (Lower than median), obese = 45 and non-obese = 28
+
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+turnbaughBFEpi <- epi.2by2(r.test, method="cohort.count")
+turnbaughBFMassoc <- turnbaughBFEpi$massoc
+turnbaughBFRR <- turnbaughBFMassoc$RR.strata.score
+turnbaughBFRRsig <- turnbaughBFMassoc$chisq.strata
+## Risk Ratio = 0.83
+## CI = 0.66, 1.05
+## p-value = 0.111
+
+###########################################################################
+############ Classification using AUCRF ###################################
+###########################################################################
+
+#Create Obese.num group
+#s1.metadata$obese.num[s1.metadata$obese=="No"] <- 0
+#s1.metadata$obese.num[s1.metadata$obese=="Yes"] <- 1
+#obese <- factor(s1.metadata$obese.num)
+
+#generate test set
+# get rid of those with 0 and only 4 other values
+#testset <- Filter(function(x)(length(unique(x))>5), s1.subsample.data)
+#testset <- cbind(obese, testset)
+#colnames(testset)[1] <- "obese"
+#testset <- cbind(testset, H, S, J, s1.phyla.rel.abund)
+
+#Try AUCRF with default measures provided in readme
+#set.seed(3)
+#TurnbaughAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+# list of 11 Measures, AUCopt = 0.7764883
+
+```
+
 ```{r Table_3, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
-Bacteroidetes <- c(0.318, 0.209, 0.666, 0.093, 0.958, 0.352, 0.917, 0.191, 0.577)
-
-Firmicutes <- c(0.682, 0.257, 0.967, 0.360, 0.265, 0.636, 0.577, 0.293, 0.407)
-
-BFRatio <- c(0.631, 0.251, 0.871, 0.466, 0.990, 0.569, 0.674, 0.053, 0.222)
-
-Shannon <- c(0.016, 0.166, 0.667, 0.864, 0.203, 0.585, 0.749, 0.217, 0.070)
-
-OTURich <- c(0.016, 0.184, 0.907, 0.195, 0.035, 0.845, 0.751, 0.988, 0.024)
-
-Evenness <- c(0.032, 0.254, 0.632, 0.476, 0.537, 0.356, 0.556, 0.187, 0.125)
-
-BrayC <- c(0.064, 0.715, 0.005, 0.089, 0.573, 0.811, 0.892, 0.055, 0.095)
+Bacteroidetes <- signif(c(baxterBacter[1,5], rossBacter[1,5], goodrichBacter[1,5], escobarBacter[1,5], zupancicBacter[1,5], HMPBacter[1,5], WuBacter[1,5], ArumugamBacter[1,5], turnbaughBacter[1,5]), 3)
+
+Firmicutes <- signif(c(baxterFirm[1,5], rossFirm[1,5], goodrichFirm[1,5], escobarFirm[1,5], zupancicFirm[1,5], HMPFirm[1,5], WuFirm[1,5], ArumugamFirm[1,5], turnbaughFirm[1,5]), 3)
+
+BFRatio <- signif(c(baxterBF[1,5], rossBF[1,5], goodrichBF[1,5], escobarBF[1,5], zupancicBF[1,5], HMPBF[1,5], WuBF[1,5], ArumugamBF[1,5], turnbaughBF[1,5]), 3)
+
+Shannon <- signif(c(baxterH[1,5], rossH[1,5], goodrichH[1,5], escobarH[1,5], zupancicH[1,5], HMPH[1,5], WuH[1,5], ArumugamH[1,5], turnbaughH[1,5]), 3)
+
+OTURich <- signif(c(baxterS[1,5], rossS[1,5], goodrichS[1,5], escobarS[1,5], zupancicS[1,5], HMPS[1,5], WuS[1,5], ArumugamS[1,5], turnbaughS[1,5]),3)
+
+Evenness <- signif(c(baxterJ[1,5], rossJ[1,5], goodrichJ[1,5], escobarJ[1,5], zupancicJ[1,5], HMPJ[1,5], WuJ[1,5], ArumugamJ[1,5], turnbaughJ[1,5]), 3)
+
+BrayC <- signif(c(baxterPERM[1,6], rossPERM[1,6], goodrichPERM[1,6], escobarPERM[1,6], zupancicPERM[1,6], HMPPERM[1,6], WuPERM[1,6], ArumugamPERM[1,6], turnbaughPERM[1,6]), 3)
 
@@
 Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=207", "n=58", "n=507")
-tpos <- c(25, 16, 20, 6, 20, 52, 47, 2, 56) #obese in low diversity group
-tneg <- c(61, 112, 12, 9, 23, 21, 57, 27, 198) #nonobese in low diversity group
-cpos <- c(22, 10, 18, 4, 17, 47, 31, 3, 47) #obese in high diversity group
-cneg <- c(64, 118, 13, 11, 25, 26, 72, 26, 206) #nonobese in high diversity group
-RR <- c(1.14, 1.6, 1.08, 1.50, 1.15, 1.11, 1.50, 0.67, 1.19)
-low <- c(0.70, 0.75, 0.72, 0.53, 0.71, 0.88, 1.04, 0.12, 0.84)
-high <- c(1.85, 3.39, 1.61, 4.26, 1.87, 1.38, 2.16, 3.70, 1.68)
+tpos <- c(BaxLowShannonGroup[2, 2], HMPLowShannonGroup[2, 2], RossLowShannonGroup[2, 2], EscoLowShannonGroup[2, 2], AruLowShannonGroup[2, 2], TurnLowShannonGroup[2, 2], ZupaLowShannonGroup[2, 2], WuLowShannonGroup[2, 2], GoodLowShannonGroup[2, 2]) #obese in low diversity group
+tneg <- c(BaxLowShannonGroup[1, 2], HMPLowShannonGroup[1, 2], RossLowShannonGroup[1, 2], EscoLowShannonGroup[1, 2], AruLowShannonGroup[1, 2], TurnLowShannonGroup[1, 2], ZupaLowShannonGroup[1, 2], WuLowShannonGroup[1, 2], GoodLowShannonGroup[1, 2]) #nonobese in low diversity group
+cpos <- c(BaxHighShannonGroup[2, 2], HMPHighShannonGroup[2, 2], RossHighShannonGroup[2, 2], EscoHighShannonGroup[2, 2], AruHighShannonGroup[2, 2], TurnHighShannonGroup[2, 2], ZupaHighShannonGroup[2, 2], WuHighShannonGroup[2, 2], GoodHighShannonGroup[2, 2]) #obese in high diversity group
+cneg <- c(BaxHighShannonGroup[1, 2], HMPHighShannonGroup[1, 2], RossHighShannonGroup[1, 2], EscoHighShannonGroup[1, 2], AruHighShannonGroup[1, 2], TurnHighShannonGroup[1, 2], ZupaHighShannonGroup[1, 2], WuHighShannonGroup[1, 2], GoodHighShannonGroup[1, 2]) #nonobese in high diversity group
+RR <- signif(c(baxterHRR[1,1], HMPHRR[1,1], rossHRR[1,1], escobarHRR[1,1], ArumugamHRR[1,1], turnbaughHRR[1,1], zupancicHRR[1,1], WuHRR[1,1], goodrichHRR[1,1]), 2)
+low <- signif(c(baxterHRR[1,2], HMPHRR[1,2], rossHRR[1,2], escobarHRR[1,2], ArumugamHRR[1,2], turnbaughHRR[1,2], zupancicHRR[1,2], WuHRR[1,2], goodrichHRR[1,2]), 2)
+high <- signif(c(baxterHRR[1,3], HMPHRR[1,3], rossHRR[1,3], escobarHRR[1,3], ArumugamHRR[1,3], turnbaughHRR[1,3], zupancicHRR[1,3], WuHRR[1,3], goodrichHRR[1,3]), 2)
 
@@
 Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=201", "n=58", "n=507")
-tpos <- c(24, 13, 16, 3, 17, 45, 37, 2, 51) #obese in low BF group
-tneg <- c(62, 115, 16, 12, 26, 28, 67, 27, 203) #nonobese in low BF group
-cpos <- c(23, 13, 22, 7, 20, 54, 41, 3, 52) #obese in high BF group
-cneg <- c(63, 115, 9, 8, 22, 19, 62, 26, 201) #nonobese in high BF group
-RR <- c(1.04, 1.00, 0.70, 0.43, 0.83, 0.83, 0.89, 0.67, 0.98)
-low <- c(0.64, 0.48, 0.47, 0.14, 0.51, 0.66, 0.66, 0.12, 0.69)
-high <- c(1.70, 2.07, 1.07, 1.35, 1.35, 1.05, 1.27, 3.70, 1.38)
+tpos <- c(BaxLowBFGroup[2, 2], HMPLowBFGroup[2, 2], RossLowBFGroup[2, 2], EscoLowBFGroup[2, 2], AruLowBFGroup[2, 2], TurnLowBFGroup[2, 2], ZupaLowBFGroup[2, 2], WuLowBFGroup[2, 2], GoodLowBFGroup[2, 2]) #obese in low BF group
+tneg <- c(BaxLowBFGroup[1, 2], HMPLowBFGroup[1, 2], RossLowBFGroup[1, 2], EscoLowBFGroup[1, 2], AruLowBFGroup[1, 2], TurnLowBFGroup[1, 2], ZupaLowBFGroup[1, 2], WuLowBFGroup[1, 2], GoodLowBFGroup[1, 2]) #nonobese in low BF group
+cpos <- c(BaxHighBFGroup[2, 2], HMPHighBFGroup[2, 2], RossHighBFGroup[2, 2], EscoHighBFGroup[2, 2], AruHighBFGroup[2, 2], TurnHighBFGroup[2, 2], ZupaHighBFGroup[2, 2], WuHighBFGroup[2, 2], GoodHighBFGroup[2, 2]) #obese in high BF group
+cneg <- c(BaxHighBFGroup[1, 2], HMPHighBFGroup[1, 2], RossHighBFGroup[1, 2], EscoHighBFGroup[1, 2], AruHighBFGroup[1, 2], TurnHighBFGroup[1, 2], ZupaHighBFGroup[1, 2], WuHighBFGroup[1, 2], GoodHighBFGroup[1, 2]) #nonobese in high BF group
+RR <- signif(c(baxterBFRR[1,1], HMPBFRR[1,1], rossBFRR[1,1], escobarBFRR[1,1], ArumugamBFRR[1,1], turnbaughBFRR[1,1], zupancicBFRR[1,1], WuBFRR[1,1], goodrichBFRR[1,1]), 2)
+low <- signif(c(baxterBFRR[1,2], HMPBFRR[1,2], rossBFRR[1,2], escobarBFRR[1,2], ArumugamBFRR[1,2], turnbaughBFRR[1,2], zupancicBFRR[1,2], WuBFRR[1,2], goodrichBFRR[1,2]), 2)
+high <- signif(c(baxterBFRR[1,3], HMPBFRR[1,3], rossBFRR[1,3], escobarBFRR[1,3], ArumugamBFRR[1,3], turnbaughBFRR[1,3], zupancicBFRR[1,3], WuBFRR[1,3], goodrichBFRR[1,3]), 2)
 

### 2016-02-24 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
adding updated Amish data

@@
 We used the categorical variable of obese or non-obese to analyze each data set seperately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 3 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix.
+Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 2 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix.
 There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio.
@@
 Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.
-We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was 0.7553 +/- 0.0888.
-We found that the total range for the total number of variables used for the classification ranged from 3 (Ross and Escobar) to 32 (Zupancic).
+We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was 0.7553 +/- 0.0855.
+We found that the total range for the total number of variables used for the classification ranged from 3 (Ross and Escobar) to 40 (Zupancic).
 The total number of trees and node size was standardized to 1000 and 20 respectively for each data set.
@@
 For the Shannon diversity RR, all studies but Wu et al had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 2].
-This was confirmed with the pooled analysis showing an increased RR of 1.19 (CI 1.04, 1.36) for obesity in the lower then median Shannon diversity group (P-value = 0.0139).
+This was confirmed with the pooled analysis showing an increased RR of 1.18 (CI 1.03, 1.36) for obesity in the lower then median Shannon diversity group (P-value = 0.0187).
 When investigated the B/F ratio RR most studies showed that a low B/F ratio tended to have less obese individuals than those with a high B/F ratio [Figure 3].
 A notable exception to this was the HMP data set and the Baxter data set [Figure 3].
-The pooled RR was 0.86 (CI 0.75, 0.99) suggesting that a low B/F ratio does indeed have less obese individuals (P-value = 0.0296).
+The pooled RR was 0.86 (CI 0.75, 0.99) suggesting that a low B/F ratio does indeed have less obese individuals (P-value = 0.0376).
 
@@
 The final forest plot that we used for this analysis investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC in the the seperate study analysis looked when pooled together to assess the RR [Figure S4].
-There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.98 (CI 0.63, 1.53), P-value = 0.9225).
+There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.96 (CI 0.60, 1.52), P-value = 0.8575).
 Thus it would seem that although Ruminococcaceae is important for classification the directionality is study dependent [Figure S4].
@@
 This is also the case for the ZScore normalization analysis of the Shannon diversity.
-As an example, a lower than median Shannon diversity only has an overall 19% increase of obesity in it versus the higher than median Shannon diversity group.
+As an example, a lower than median Shannon diversity only has an overall 18% increase of obesity in it versus the higher than median Shannon diversity group.
 So even though there is a detectable signal it remains to be seen if this is a biologically significant phenomenon that could have a tangible effect on obesity.
@@
 metadata <- read.csv("amish_obesity_table2.csv")
+metadata2 <- read.csv("amish.metadata.csv")
 test <- metadata[!duplicated(metadata$submitted_sample_id_s), ]
 rownames(test) <- test[, 4]
+test2 <- metadata2[!duplicated(metadata2$SUBJID), ]
+rownames(test2) <- test2[, 1]
 shared.data <- read.table("combined.0.03.subsample.shared", header=T)
@@
 shared.data <- shared.data[, -c(1:3)]
+BactFamily <- read.csv("BacterialFamilyData.csv")
+rownames(BactFamily) <- BactFamily[, 1]
+BactFamily <- BactFamily[, -1]
 keep  <- rownames(shared.data)
 
-test2 <- test[keep, ]
-good.metadata <- test2
-test3 <- test2[!duplicated(test2$submitted_subject_id_s), ]
-keep <- rownames(test3)
+test3 <- test[keep, ]
+good.metadata <- test3
+test4 <- test2[keep, ]
+good.metadata2 <- test4
+
+family <- BactFamily[keep, ]
 microbiome <- shared.data[keep, ]
-metadata <- test3
-
-rm(good.metadata, shared.data, keep, test, test2, test3)
+metadata <- cbind(good.metadata, good.metadata2)
+
+rm(good.metadata, shared.data, keep, test, test2, test3, test4, BactFamily, good.metadata2, metadata2)
+
+test <- metadata[!duplicated(metadata$submitted_subject_id_s), ]
+keep  <- rownames(test)
+metadata <- test
+family <- family[keep, ]
+microbiome <- microbiome[keep, ]
+
+rm(test, keep)
+
+metadata <- metadata[complete.cases(metadata), ]
+keep <- rownames(metadata)
+family <- family[keep, ]
+microbiome <- microbiome[keep, ]
+
+rm(keep)
 
@@
 
+#Create BMI groups
+metadata$BMI.class[metadata$BMI<=24] <- "Normal"
+metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
+metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
+metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
+
 #Create Obese Yes/No groups
-# One problem is that we are assuming those with "not provided" are defaulting to the not affected group
-metadata$obese[metadata$subject_is_affected_s=="<not provided>" | metadata$subject_is_affected_s=="No"] <- "No"
-metadata$obese[metadata$subject_is_affected_s=="Yes"] <- "Yes"
-
-#Create Obese.num group
-metadata$obese.num[metadata$obese=="No"] <- 0
-metadata$obese.num[metadata$obese=="Yes"] <- 1
+metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
+metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
+
+#Create BMI class 2 groups
+metadata$BMI.class2[metadata$BMI<=24] <- "Normal"
+metadata$BMI.class2[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
+metadata$BMI.class2[metadata$BMI>=30] <- "Obese"
 
 #create groups to be used
+bmi <- metadata$BMI
 obese <- factor(metadata$obese)
@@
 
-zupancicH <- anova(lm(H ~ obese)) #P-value=0.2034
-zupancicS <- anova(lm(S ~ obese)) #P-value=0.03516
-zupancicJ <- anova(lm(J ~ obese)) #P-value=0.5369
+zupancicH <- anova(lm(H ~ obese)) #P-value=0.2671
+zupancicS <- anova(lm(S ~ obese)) #P-value=0.05423
+zupancicJ <- anova(lm(J ~ obese)) #P-value=0.6229
 
@@
 
-zupancicBacter <- anova(lm(bacter ~ obese)) #P-value=0.9582
-zupancicFirm <- anova(lm(firm ~ obese)) #P-value=0.2654
-zupancicBF <- anova(lm(BFratio ~ obese)) #P-value=0.9899
+zupancicBacter <- anova(lm(bacter ~ obese)) #P-value=0.8592
+zupancicFirm <- anova(lm(firm ~ obese)) #P-value=0.2564
+zupancicBF <- anova(lm(BFratio ~ obese)) #P-value=0.989
 
@@
 zupancicPERM <- zupancic2$aov.tab
-#PERMANOVA=0.5734, pseudo-F=0.95626
+#PERMANOVA=0.6863, pseudo-F=0.92888
 
@@
 zupancicHRRsig <- zupancicHMassoc$chisq.strata
-## Risk Ratio = 1.50
-## CI = 1.04, 2.16
-## p-value = 0.025
+## Risk Ratio = 1.49
+## CI = 1.01, 2.19
+## p-value = 0.043
 
@@
 zupancicBFRRsig <- zupancicBFMassoc$chisq.strata
-## Risk Ratio = 0.89
-## CI = 0.63, 1.27
-## p-value = 0.53
+## Risk Ratio = 0.94
+## CI = 0.64, 1.36
+## p-value = 0.727
 
@@
 #Create Obese.num group
+metadata$obese.num[metadata$obese=="No"] <- 0
+metadata$obese.num[metadata$obese=="Yes"] <- 1
 obese <- factor(metadata$obese.num)
@@
 #zupancicAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-# list of 42 Measures, AUCopt = 0.6365534
+#zupAUC <- fit$`OOB-AUCopt`
+# list of 40 Measures, AUCopt = 0.6596899
 
@@
 Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012", "2014")
-Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=207", "n=58", "n=507")
+Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=199", "n=58", "n=507")
 tpos <- c(BaxLowShannonGroup[2, 2], HMPLowShannonGroup[2, 2], RossLowShannonGroup[2, 2], EscoLowShannonGroup[2, 2], AruLowShannonGroup[2, 2], TurnLowShannonGroup[2, 2], ZupaLowShannonGroup[2, 2], WuLowShannonGroup[2, 2], GoodLowShannonGroup[2, 2]) #obese in low diversity group
@@
 **Figure 2: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity.**  Groups were divided for each study on high and low Shannon Diversity groups based on the median for that study.
-The overall pooled relative risk was 1.19 for the low diversity group (P-value = 0.0139).
+The overall pooled relative risk was 1.18 for the low diversity group (P-value = 0.0187).
 
@@
 Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012", "2014")
-Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=201", "n=58", "n=507")
+Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=199", "n=58", "n=507")
 tpos <- c(BaxLowBFGroup[2, 2], HMPLowBFGroup[2, 2], RossLowBFGroup[2, 2], EscoLowBFGroup[2, 2], AruLowBFGroup[2, 2], TurnLowBFGroup[2, 2], ZupaLowBFGroup[2, 2], WuLowBFGroup[2, 2], GoodLowBFGroup[2, 2]) #obese in low BF group
@@
 **Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low B/F ratio groups based on the median for that specific study.
-The overall pooled relative risk was 0.86 for the low diversity group (P-value = 0.0296).
+The overall pooled relative risk was 0.86 for the low diversity group (P-value = 0.0376).
 
@@
 turn.data <- read.csv("Turnbaugh.normalized.csv")
-#Load in Amish data
-amish.data <- read.csv("Amish.normalized.csv")
-
-#Amish STD = 0.5893173
-amish.ZScore.H <- amish.data$H.corr/0.5893173
+
 #Turnbaugh STD = 0.4818751
 turn.ZScore.H <- turn.data$H.corr/0.4818751
+
+#Amish STD = 0.5932651
+keepamish <- which(data1$dataset == "Amish")
+amishdata <- data1[keepamish, ]
+amish.ZScore.H <- amishdata$H.corr/0.5932651
+
 #MetaHit-Den STD = 0.4038558
@@
 metahit.ZScore.H <- metahitdata$H.corr/0.4038558
+
 #Korean STD = 0.5400615
@@
 Korean.ZScore.H <- Koreandata$H.corr/0.5400615
+
 #HMP STD = 0.6774514
@@
 HMP.ZScore.H <- hmpdata$H.corr/0.6774514
+
 #CCHC STD = 0.5784522
@@
 CCHC.ZScore.H <- CCHCdata$H.corr/0.5784522
+
 #GLNE07 STD = 0.4663871
@@
 GLNE.ZScore.H <- GLNEdata$H.corr/0.4663871
+
 #Escobar STD = 0.5378935
@@
 Escobar.ZScore.H <- escobardata$H.corr/0.5378935
+
 #COMBO STD = 0.6870763
@@
 COMBO.ZScore.H <- COMBOdata$H.corr/0.6870763
+
 #COMBO STD = 0.3489582
@@
 #data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit, COMBO, twinsUK
-HZScore <- c(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, COMBO.ZScore.H, twinsUK.ZScore.H)
+HZScore <- c(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, COMBO.ZScore.H, twinsUK.ZScore.H, amish.ZScore.H)
 data1 <- cbind(HZScore, data1)
 
-amish.data <- cbind(amish.ZScore.H, amish.data)
 turn.data <- cbind(turn.ZScore.H, turn.data)
@@
 t.obese <- as.character(turn.data$obese)
-am.obese <- as.character(amish.data$obese)
 
@@
 t.H.Zscore <- turn.data$turn.ZScore.H
-am.H.Zscore<- amish.data$amish.ZScore.H
 
@@
 comb1.obese <- c(d1.obese, t.obese)
-comb2.obese <- c(d1.obese, t.obese, am.obese)
 comb1.bmi.class <- c(d1.bmi.class, t.bmi.class)
 comb1.H.Zscore <- as.data.frame(c(d1.H.Zscore, t.H.Zscore))
-comb2.H.Zscore <- as.data.frame(c(d1.H.Zscore, t.H.Zscore, am.H.Zscore))
 
 group.data.test1 <- cbind(comb1.H.Zscore, comb1.obese, comb1.bmi.class)
-group.data.test2 <- cbind(comb2.H.Zscore, comb2.obese)
 colnames(group.data.test1) <- c("H.Zscore", "obese", "BMI.class")
-colnames(group.data.test2) <- c("H.Zscore", "obese")
 
@@
 t.zbf.corr <- turn.data$ZBF
-am.zbf.corr <- amish.data$ZBF
 
@@
 
-comb2.zbf.corr <- as.data.frame(c(d1.zbf.corr, t.zbf.corr, am.zbf.corr))
-group.data.test2 <- cbind(comb2.zbf.corr, group.data.test2)
-colnames(group.data.test2)[1] <- "zbf"
-
 ```
@@
 # Section A
-p1 <- ggplot(group.data.test2, aes(obese, H.Zscore)) 
+p1 <- ggplot(group.data.test, aes(obese, H.Zscore)) 
 p2 <- p1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
@@
 # Section B
-q1 <- ggplot(group.data.test1, aes(BMI.class, H.Zscore)) 
+q1 <- ggplot(group.data.test, aes(BMI.class, H.Zscore)) 
 q2 <- q1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
@@
 # Section C
-bf1 <- ggplot(group.data.test2, aes(obese, zbf)) 
+bf1 <- ggplot(group.data.test, aes(obese, zbf)) 
 bf2 <- bf1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
@@
 **Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** ZScore Normalized Shannon Diversity and Obesity.
-There was a significant difference between non-obese (n=1099) and obese groups (n=443) (P-value = 0.00493).
+There was a significant difference between non-obese (n=1099) and obese groups (n=435) (P-value = 0.005992).
 **B)** ZScore Normalized Shannon Diversity by BMI Group.
-Total n for lean, overweight, and obese groups are 509, 365, 461 respectively.
-Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = 0.03944) but no difference between lean and obese group (P-value = 0.107).
+Total n for lean, overweight, and obese groups are 575, 435, 524 respectively.
+Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = 0.018) but no difference between lean and obese group (P-value = 0.081).
 **C)** ZScore Normalized B/F Ratio and Obesity.
-There was no significant difference between the non-obese (n=1099) and obese (n=443) group based on B/F ratio (P-value = 0.4516).
+There was no significant difference between the non-obese (n=1099) and obese (n=435) group based on B/F ratio (P-value = 0.456).
 **D)** ZScore Normalized B/F Ratio by BMI Group.
-Total n for lean, overweight, and obese groups are 509, 365, 461 respectively.
-There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = 0.3317).
+Total n for lean, overweight, and obese groups are 575, 435, 524 respectively.
+There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = 0.398).
 

### 2016-02-24 microbiomeAndobesitySupplemental.Rmd
adding updated Amish data

@@
 Year <- c("2015", "2011", "2014", "2010", "2009", "2012", "2012")
-Total <- c("n=172", "n=256", "n=30", "n=85", "n=146", "n=201", "n=58")
-tpos <- c(36, 13, 7, 17, 22, 55, 2) #obese in low Rumi group
-tneg <- c(59, 140, 8, 26, 56, 103, 37) #nonobese in low Rumi group
-cpos <- c(11, 13, 3, 20, 25, 23, 3) #obese in high Rumi group
-cneg <- c(66, 90, 12, 22, 43, 26, 16) #nonobese in high Rumi group
-RR <- c(2.65, 0.67, 2.33, 0.83, 0.77, 0.74, 0.32)
-low <- c(1.45, 0.33, 0.74, 0.51, 0.48, 0.51, 0.06)
-high <- c(4.86, 1.39, 7.35, 1.35, 1.23, 1.07, 1.78)
+Total <- c("n=172", "n=256", "n=30", "n=85", "n=146", "n=199", "n=58")
+tpos <- c(36, 13, 7, 17, 22, 52, 2) #obese in low Rumi group
+tneg <- c(59, 140, 8, 26, 56, 110, 37) #nonobese in low Rumi group
+cpos <- c(11, 13, 3, 20, 25, 18, 3) #obese in high Rumi group
+cneg <- c(66, 90, 12, 22, 43, 19, 16) #nonobese in high Rumi group
+RR <- c(2.65, 0.67, 2.33, 0.83, 0.77, 0.66, 0.32)
+low <- c(1.45, 0.33, 0.74, 0.51, 0.48, 0.44, 0.06)
+high <- c(4.86, 1.39, 7.35, 1.35, 1.23, 0.98, 1.78)
 

### 2016-02-25 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
update table based on Amish data

@@
 
-Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic*", "HMP", "Wu", "Arumugam", "Turnbaugh**")
-
-Age <- c("54.3 +/- 9.9", "57.0 +/- 11.2", "61.5 +/- 8.9", "38.1 +/- 11.1", "48.2 +/- 13.2", "26.1 +/- 5.0", "26.4 +/- 9.4", "56.5 +/- 7.7", "21 - 32")
-
-Sex <- c("111|61", "48|15", "505|2", "14|16", "198|112", "127|129", "31|27", "45|40", "Not Available")
+Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "HMP", "Wu", "Arumugam", "Turnbaugh*")
+
+Age <- c("54.3 +/- 9.9", "57.0 +/- 11.2", "61.5 +/- 8.9", "38.1 +/- 11.1", "46.7 +/- 13.1", "26.1 +/- 5.0", "26.4 +/- 9.4", "56.5 +/- 7.7", "21 - 32")
+
+Sex <- c("111|61", "48|15", "505|2", "14|16", "112|87", "127|129", "31|27", "45|40", "Not Available")
 
@@
 
-BMI <- c("27.0 +/- 5.3", "31.6 +/- 5.3", "26.3 +/- 4.9", "27.4 +/- 4.5", "29.2 +/- 5.2", "24.1 +/- 3.4", "24.6 +/- 4.8", "27.8 +/- 5.8", "Not Available")
-
-Min <- c(17.5, 22.1, 16.2, 19.5, 16.7, 19, 14.0, 18.6, "Not Available")
+BMI <- c("27.0 +/- 5.3", "31.6 +/- 5.3", "26.3 +/- 4.9", "27.4 +/- 4.5", "27.7 +/- 5.3", "24.1 +/- 3.4", "24.6 +/- 4.8", "27.8 +/- 5.8", "Not Available")
+
+Min <- c(17.5, 22.1, 16.2, 19.5, 18.2, 19, 14.0, 18.6, "Not Available")
 
@@
 
-*If numerical BMI information was not available then ranges and demographic information was taken from the authors published manuscript for this table.
-
-**Only BMI group information was provided and only an age range was given.
+*Only BMI group information was provided and only an age range was given.
 

### 2016-02-25 code/Zupancic.Feb242016.dataAnalysis.R
Zupancic R code update

@@ -1,6 +1,6 @@
 #Zupancic et al.
 # Single study analysis
-#Feb 5th, 2016
+#Feb 24th, 2016
 
 ###########################################################################
 ############ Preparing Data Tables for Analysis ###########################
@@ -11,22 +11,44 @@ setwd("C:/users/marc/Desktop/obesity2/Amish/updatedGOOD")
 
 #Read in and match metadata to microbiome data
 metadata <- read.csv("amish_obesity_table2.csv")
+metadata2 <- read.csv("amish.metadata.csv")
 test <- metadata[!duplicated(metadata$submitted_sample_id_s), ]
 rownames(test) <- test[, 4]
+test2 <- metadata2[!duplicated(metadata2$SUBJID), ]
+rownames(test2) <- test2[, 1]
 shared.data <- read.table("combined.0.03.subsample.shared", header=T)
 rownames(shared.data) <- shared.data[, 2]
 shared.data <- shared.data[, -c(1:3)]
+BactFamily <- read.csv("BacterialFamilyData.csv")
+rownames(BactFamily) <- BactFamily[, 1]
+BactFamily <- BactFamily[, -1]
 keep  <- rownames(shared.data)
 
-test2 <- test[keep, ]
-good.metadata <- test2
-test3 <- test2[!duplicated(test2$submitted_subject_id_s), ]
-keep <- rownames(test3)
+test3 <- test[keep, ]
+good.metadata <- test3
+test4 <- test2[keep, ]
+good.metadata2 <- test4
+
+family <- BactFamily[keep, ]
 microbiome <- shared.data[keep, ]
-metadata <- test3
+metadata <- cbind(good.metadata, good.metadata2)
 
-rm(good.metadata, shared.data, keep, test, test2, test3)
+rm(good.metadata, shared.data, keep, test, test2, test3, test4, BactFamily, good.metadata2, metadata2)
 
+test <- metadata[!duplicated(metadata$submitted_subject_id_s), ]
+keep  <- rownames(test)
+metadata <- test
+family <- family[keep, ]
+microbiome <- microbiome[keep, ]
+
+rm(test, keep)
+
+metadata <- metadata[complete.cases(metadata), ]
+keep <- rownames(metadata)
+family <- family[keep, ]
+microbiome <- microbiome[keep, ]
+
+rm(keep)
 
[…]

### 2016-03-04 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
Work through Abstract and Introduction

@@
 ---
-title: "Looking for a Signal in the Noise: Revisiting Obesity and the Bacterial Microbiome"
+title: "Looking for a Signal in the Noise: Revisiting Obesity and the microbiome"
 output:
@@
 <br>
-**Running Title:** Bacterial Microbiome and Obesity
+**Running Title:** microbiome and Obesity
 <br>
@@
 *******
-##### **Abstract:**
-**Rationale:** The hypothesis that changes in the microbiome and obesity are associated has been widely accepted.
-However, two recent reports that re-examined existing data suggest that acceptance of this hypothesis is unwarranted.
-These studies, however, studied the original studies in isolation from each other.
-We hypothesized that by statistically pooling the individual results, the statistical power to identify a significant association would increase.
-In this meta analysis, we tested the significance of the association between the microbiome and obesity by increasing the number of data sets examined and investigating the effect of pooling the results across the individual studies.
-
-**Methods:** We identified 196 studies that tested the microbiome-obesity association hypothesis; however when we limited the studies to those that did microbiome sequencing, had publicly available sequences, and were either willing to share the BMI information for every individual or were already published, a total of 9 data sets were identified.
-We applied three analysis piplines to the data from each study.
-First, we tested the association between the community diversity, Bacteroides to Firmicutes ratio (B/F), and relative risk (RR) with the subjects' body mass index (BMI) and obesity category.
-We also used random forest to identify variables for classification of obese and non-obese indivdiuals.
-Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each studies.
-Finally, data normalization was used to investigate all data sets together for relationships between the microbiome and obesity.
-
-**Results:** There were no significant differences (P > 0.05) between the relative abundance of Bacteroidetes, Firmicutes, or thier ratio in non-obese and obese individuals.
-The Shannon diversity, OTU richness, relative risk, or oveall community dissimilarity was significant in at least one study (P<0.05).
-Overall, the random forest results for Area Under the Curve (AUC) was 0.7553 +/- 0.0888 and 7/9 studies contained an OTU that could be classified to the Ruminococcaceae family.
-When we pooled the results across studies, there was a signifcant difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04, 1.36, (P-value = 0.0139)) and B/F ratio (RR = 0.86, CI = 0.75, 0.99, (P-value = 0.0296)).
-
-**Conclusion:** We show that by pooling the results of many indivdiual studies we can test the generalizability of the bacterial microbiome.
-Our meta-analysis supports the  hypothesis that there is a difference between non-obese and obese individuals in Shannon diversity and B/F ratio, based on the currently avialable bacterial microbiome data included in this meta-analysis.
+##### Abstract
+**Rationale:** The hypothesis that changes in the microbiome are associated with obesity has been widely accepted.
+However, two recent studies re-analyzed published data and found that when the datasets were analyzed independently they could not support the hypothesis.
+In this meta analysis, we retested the hypothesis by increasing the number of data sets examined and investigating the effect of pooling the results across the individual datasets.
+
+**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 9 studies that fit our criteria.
+We applied three analysis pipelines to the data from each study.
+First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value.
+We also used the random forest machine learning algorithm to identify community features that would allow us to to predict whether a person was obese.
+Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F.
+Finally, we normalized the community diversity and B/F values within each dataset and pooled the data to test for an association with obesity.
+
+**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals.
+The Shannon diversity (N=1 study), OTU richness (N=2), and overall community dissimilarity (N=1) were significantly different between non-obese and obese individuals in at least one study (P<0.05).
+With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: 0.76; s.d.=0.09).
+In seven of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies.
+When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04-1.36, P-value = 0.01) and B/F ratio (RR = 0.86, CI = 0.75-0.99, P-value = 0.03).
+Finally, when we normalized the Shannon diversity data and pooled studies there was a significant difference between non-obese and obese individuals (difference in z-score=0.XX; P=0.0X) something about...
+B/F.
+
+**Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome.
+Our meta-analysis supports the  hypothesis that there is a difference between non-obese and obese individuals in Shannon diversity and B/F ratio, but that the effect sizes are small.
 
 *******
-##### **Introduction:**  
+##### Introduction
 Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014].
-This number increases to approximately 35% in adults (aged 20 or older) and these statisitics have seen little change since 2003 [@ogden_prevalence_2014].
+This number increases to approximately 35% in adults (aged 20 or older) and these statistics have seen little change since 2003 [@ogden_prevalence_2014].
 Traditionally the body mass index (BMI) has been used as the traditional method of classifying individuals as non-obese or obese [@lichtash_body_2013].
-However, other methods do exist, such as body adiposity index (BAI) but do not always correlate with important molecular measures traditionally associated with obesity [@lichtash_body_2013].
-Recently, there has been a lot of interest in the bacteiral microbiome and it's potential ability to modulate obesity [@brahe_can_2016; @dror_microbiota_2016].
-If this modulation potential is true then the bacterial microbiome could have an enormous impact and role to play in the future treatment of obesity and in helping to stem the rising obesity epidemic.
-
-There has been a continuous stream of studies that report that there is a link between the bacterial microbiome and obesity.
-The first research article to solidify the bacterial microbiomes' role in obesity was the landmark study by Turnbaugh et al.
-[-@turnbaugh_obesity-associated_2006] which provided direct evidence that by modulating the microbiome one could have an effect on the obesity phenotype of mice.
-This was quickly followed up by human studies that found observable differences in the bacterial microbiome between non-obese and obese individuals [@ley_microbial_2006; @turnbaugh_core_2009].
-Since the publication of these studies there have been a steady stream of publications on the bacterial microbiome with associated BMI data but without clear explicit mention of any significant correlations between the bacterial community and obesity [@ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; arumugam_enterotypes_2011; @goodrich_human_2014; @yatsunenko_human_2012; @wu_linking_2011; @escobar_gut_2014].
-However, two recent publications [@walters_meta-analyses_2014; @finucane_taxonomic_2014] re-analyzed some of the previously publishd data and came to the shocking conclusion that the human bacterial microbiome did not correlate with obesity.
-One common critique of these previous studies was that they never looked at the data pooled together and that this could result in a significant result where each individual study alone would not.
-
-Although there have been numerous RT-PCR and qPCR studies directed at examining the bacterial microbiome and obesity [@guo_development_2008; @schwiertz_microbiota_2010].
-The main method of choice to obtain the most complete picture of the overall microbial community is through 16S sequencing of a specific hypervariable region on the 16S rRNA gene.
-The sequencers used to complete this task have varied depending on research group with the most popular being the Roche 454 and Illumina MiSeq.
-A priori we decided that for this report we would limit ourselves to analyzing studies were 16S sequencing data is avaliable along with BMI information.
-We also analyzed every study with a uniform approach utilizing the mothur software package for sequence processing and clean up and R statistical software for analysis.
-
-The purpose of this study is to perform an extensive meta-analysis of the bacterial microbiome and obesity by analyzing and applying a more sytematic approach than has been used by previous studies in this area.
-Additionaly, we will include both seperate and pooled analysis to assess the bacterial microbiome and obesity.
-By including these two components we will build on the previous two re-analysis studies and provide a stronger assesment of the direction of effect or lack there of on the bacterial microbiome and obesity.
-In particular, our goal is to provide additional clarity on the role of both bacterial diversity and the Bacteroidetes/Firmicutes ratio impact on obesity.
+Recently, there has been increased interest in the role of the microbiome in modulating obesity [@brahe_can_2016; @dror_microbiota_2016].
+If the microbiome does affect obesity status, then manipulating the microbiome could have a significant role in the future treatment of obesity and in helping to stem the current epidemic.
+
+There has been a continuous stream of studies that report observing a link between the composition of microbiome and obesity.
+The first such study used genetically obese mice and observed that obese mice had a XXXXXer B/F ratio than lean mice [@Ley2005].
+Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009].
+Since the publication of these studies there have been a steady stream of publications on the microbiome with associated BMI data but without clear explicit mention of any significant correlations between the microbiome and obesity [@ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @arumugam_enterotypes_2011; @goodrich_human_2014; @yatsunenko_human_2012; @wu_linking_2011; @escobar_gut_2014].
+Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B/F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
+
+One critique of these re-analysis studies is that they did not aggregate the results across studies to increase the effective sample size.
+It is possible that there are small associations within each study that would not be significant, because the studies lacked sufficient statistical power.
+The purpose of this study was to perform a meta-analysis of the association between differences in the microbiome and obesity status by analyzing and applying a more systematic and synthetic approach than was used previously.
+Our goal was to provide additional clarity on the possible relationship between community diversity, B/F, and community composition and obesity status.
 
@@
 
+Although there have been numerous RT-PCR and qPCR studies directed at examining the microbiome and obesity [@guo_development_2008; @schwiertz_microbiota_2010].
+The main method of choice to obtain the most complete picture of the overall microbial community is through 16S sequencing of a specific hypervariable region on the 16S rRNA gene.
+The sequencers used to complete this task have varied depending on research group with the most popular being the Roche 454 and Illumina MiSeq.
+A priori we decided that for this report we would limit ourselves to analyzing studies were 16S sequencing data is avaliable along with BMI information.
+We also analyzed every study with a uniform approach utilizing the mothur software package for sequence processing and clean up and R statistical software for analysis.
+
 *Subject Demographics and Study Outcomes*
@@
 We found a large variation in reported outcomes with two studies finding bacterial community associations with obesity and three studies finding some association between either Firmicutes, Bacteroidetes, or the Bacteroidetes/Firmicutes ratio and obesity.
-However, a total of three manuscripts found no correlations between the bacterial microbiome and obesity.
+However, a total of three manuscripts found no correlations between the microbiome and obesity.
 This is in addition to the two re-analysis studies published in 2014.
@@
 
-Our meta-analysis helps to provide a little more clarity to the on going debate of whether or not there are specific bacterial microbiome variables that are influenced by obesity.
-Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a bacterial microbiome difference between non-obese and obese individuals.
+However, other methods do exist, such as body adiposity index (BAI) but do not always correlate with important molecular measures traditionally associated with obesity [@lichtash_body_2013].
+
+* hyper variation between humans may mute the actual signal.
+
+* just because we can't find a human-wide signal, doesn't mean microbiome isn't important for weight
+
+* @ley2006 watched humans lose weight and communities change
+* @turnbaugh20XX did fecal transplants (humans to mice) and transferred phenotype
+* @ley20XX did fecal transplants (pregnant women to mice) and transferred phenotype
+
+Our meta-analysis helps to provide a little more clarity to the on going debate of whether or not there are specific microbiome variables that are influenced by obesity.
+Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a microbiome difference between non-obese and obese individuals.
 However, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give.
-Here we perform an extensive literature review of the existing studies on the bacterial microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.
+Here we perform an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.
 We analyze the data using three different approaches.
@@
 
-Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the bacterial microbiome and obestiy could be observed.
-However, we found that the other two approaches, that used the pooled data, was able to identify a correlation between measures of the bacterial microbiome and obesity [Figure 2 & 3].
+Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the microbiome and obestiy could be observed.
+However, we found that the other two approaches, that used the pooled data, was able to identify a correlation between measures of the microbiome and obesity [Figure 2 & 3].
 For Shannon diversity both approaches using the pooled data found a correlation while only one of the approaches for the B/F ratio found a correlation.
@@
 Additionally, we are also limited in our BMI classifications by what the previous studies published and performed.
-In general, not enough studies have been done to investigate the bacterial microbiome in the context of other markers for obesity or on different ethnicities to be able to perform the analysis that we did.
+In general, not enough studies have been done to investigate the microbiome in the context of other markers for obesity or on different ethnicities to be able to perform the analysis that we did.
 Thus as more specific studies are published that address ethnicity and studies on the microbiome that use better markers of obesity are completed it is entirely possible that the noise associated with these measurements will decrease.
 
-Second, although we make attempts to normalize the data for the pooled analysis so that we can compare the bacterial microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.
+Second, although we make attempts to normalize the data for the pooled analysis so that we can compare the microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.
 However, what our analysis shows is that despite these differences, we are still able to identify an overall signal from the data.
@@
 Only a total of 1542 individuals were included in the pooled analysis which is relatively small compared to other meta-analysis in non-microbiome fields of study.
-It will be necessary to continuously update this analysis, as more studies become available, in order to obtain a more accurate estimate of the true effect that obesity has on the bacterial microbiome within humans.
+It will be necessary to continuously update this analysis, as more studies become available, in order to obtain a more accurate estimate of the true effect that obesity has on the microbiome within humans.
 
@@
 
-Despite the many problems that could add more noise to the measurements used, this study provides a reasonable first pass attempt to pool this data together in an effort to find a correlation between the bacterial microbiome and obesity.
-In doing so we believe that it provides a tanatalizing result that despite the large differences between studies it is still possible to detect a correlation between the bacterial microbiome and obesity.
+Despite the many problems that could add more noise to the measurements used, this study provides a reasonable first pass attempt to pool this data together in an effort to find a correlation between the microbiome and obesity.
+In doing so we believe that it provides a tanatalizing result that despite the large differences between studies it is still possible to detect a correlation between the microbiome and obesity.
 However, we emphaszie that it should still be viewed with caution since there are still many aspects of this analysis that could be improved upon.
 It truly is only a first attempt at trying to summerize and analyze the pooled data that is available.
-We hope that this study helps to add important information and clarity to the on-going dialogue of the bacterial microbiome's role in obesity.
+We hope that this study helps to add important information and clarity to the on-going dialogue of the microbiome's role in obesity.
 

### 2016-03-04 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
manuscript update: abstract and Figs

@@
 In seven of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies.
-When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04-1.36, P-value = 0.01) and B/F ratio (RR = 0.86, CI = 0.75-0.99, P-value = 0.03).
-Finally, when we normalized the Shannon diversity data and pooled studies there was a significant difference between non-obese and obese individuals (difference in z-score=0.XX; P=0.0X) something about...
+When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04-1.36, P-value = 0.02) and B/F ratio (RR = 0.86, CI = 0.75-0.99, P-value = 0.04).
+Finally, when we normalized the Shannon diversity data and pooled studies there was a significant difference between non-obese and obese individuals (difference in z-score=0.155; P=0.006) something about...
 B/F.
@@
 A total of 3 studies were from records identified by database searching [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014] and the other 6 were from records identified by other sources [@ross_16s_2015; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @conlan_species-level_2012].
-One of these studies (Baxter) has not been published at the time of this writing.
+One of these studies (Baxter) was used on a data set that was not originally designed to investigate obesity in order to help mitigate any possible publication bias associated with the bacterial microbiome and obesity.
 
@@
 **Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** ZScore Normalized Shannon Diversity and Obesity.
-There was a significant difference between non-obese (n=1099) and obese groups (n=435) (P-value = 0.005992).
+There was a significant difference between non-obese (n=1081) and obese groups (n=435) (P-value = 0.005992).
 **B)** ZScore Normalized Shannon Diversity by BMI Group.
-Total n for lean, overweight, and obese groups are 575, 435, 524 respectively.
+Total n for lean, overweight, and obese groups are 563, 435, 518 respectively.
 Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = 0.018) but no difference between lean and obese group (P-value = 0.081).
 **C)** ZScore Normalized B/F Ratio and Obesity.
-There was no significant difference between the non-obese (n=1099) and obese (n=435) group based on B/F ratio (P-value = 0.456).
+There was no significant difference between the non-obese (n=1081) and obese (n=435) group based on B/F ratio (P-value = 0.456).
 **D)** ZScore Normalized B/F Ratio by BMI Group.
-Total n for lean, overweight, and obese groups are 575, 435, 524 respectively.
+Total n for lean, overweight, and obese groups are 563, 435, 518 respectively.
 There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = 0.398).

### 2016-03-05 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
added ZScoreLogBF

@@
 
+###########################################################################
+############ ZScore Data Preperation ###################################
+###########################################################################
+
+BaxterZH <- scale(H)
+BaxterZLogBF <- scale(log(BFratio))
+BaxterBMI <- bmi
+
 ```
@@
 
+###########################################################################
+############ ZScore Data Preperation ###################################
+###########################################################################
+
+RossZH <- scale(H)
+RossZLogBF <- scale(log(BFratio))
+RossBMI <- bmi
+
 ```
@@
 
+###########################################################################
+############ ZScore Data Preperation ###################################
+###########################################################################
+
+GoodrichZH <- scale(H)
+GoodrichZLogBF <- scale(log(BFratio))
+GoodrichBMI <- bmi
+
 ```
@@
 
+###########################################################################
+############ ZScore Data Preperation ###################################
+###########################################################################
+
+EscobarZH <- scale(H)
+EscobarZLogBF <- scale(log(BFratio))
+EscobarBMI <- bmi
+
 ```
@@
 
+###########################################################################
+############ ZScore Data Preperation ###################################
+###########################################################################
+
+ZupancicZH <- scale(H)
+ZupancicZLogBF <- scale(log(BFratio))
+ZupancicBMI <- bmi
+
 ```
@@
 # list of 6 Measures, AUCopt = 0.7031773
+
+###########################################################################
+############ ZScore Data Preperation ###################################
+###########################################################################
+
+HMPZH <- scale(H)
+HMPZLogBF <- scale(log(BFratio))
+HMPBMI <- bmi
 
@@
 
+###########################################################################
+############ ZScore Data Preperation ###################################
+###########################################################################
+
+WuZH <- scale(H)
+WuZLogBF <- scale(log(BFratio))
+WuBMI <- bmi
+
 ```
@@
 
+###########################################################################
+############ ZScore Data Preperation ###################################
+###########################################################################
+
+ArumugamZH <- scale(H)
+ArumugamZLogBF <- scale(log(BFratio))
+ArumugamBMI <- bmi
+
 ```
@@
 
+###########################################################################
+############ ZScore Data Preperation ###################################
+###########################################################################
+
+TurnbaughZH <- scale(H)
+TurnbaughZLogBF <- scale(log(BFratio))
+TurnbaughObese <- as.character(obese)
+TurnbaughBMICat <- as.character(bmi)
+
 ```
@@
 
-#Load in BMI data
-data1 <- read.csv("combined.H.corr.data.csv")
-#Load in Turnbaugh data
-turn.data <- read.csv("Turnbaugh.normalized.csv")
-
-#Turnbaugh STD = 0.4818751
-turn.ZScore.H <- turn.data$H.corr/0.4818751
-
-#Amish STD = 0.5932651
-keepamish <- which(data1$dataset == "Amish")
-amishdata <- data1[keepamish, ]
-amish.ZScore.H <- amishdata$H.corr/0.5932651
-
-#MetaHit-Den STD = 0.4038558
-keepmetahit <- which(data1$dataset == "MetaHit")
-metahitdata <- data1[keepmetahit, ]
-metahit.ZScore.H <- metahitdata$H.corr/0.4038558
-
-#Korean STD = 0.5400615
-keepKorean <- which(data1$dataset == "Korean")
-Koreandata <- data1[keepKorean, ]
-Korean.ZScore.H <- Koreandata$H.corr/0.5400615
-
-#HMP STD = 0.6774514
-keepHMP <- which(data1$dataset == "HMP")
-hmpdata <- data1[keepHMP, ]
-HMP.ZScore.H <- hmpdata$H.corr/0.6774514
-
-#CCHC STD = 0.5784522
-keepCCHC <- which(data1$dataset == "CCHC")
-CCHCdata <- data1[keepCCHC, ]
-CCHC.ZScore.H <- CCHCdata$H.corr/0.5784522
-
-#GLNE07 STD = 0.4663871
-keepGLNE <- which(data1$dataset == "GLNE")
-GLNEdata <- data1[keepGLNE, ]
-GLNE.ZScore.H <- GLNEdata$H.corr/0.4663871
-
-#Escobar STD = 0.5378935
-keepescobar <- which(data1$dataset == "Escobar")
-escobardata <- data1[keepescobar, ]
-Escobar.ZScore.H <- escobardata$H.corr/0.5378935
-
-#COMBO STD = 0.6870763
-keepCOMBO <- which(data1$dataset == "COMBO")
-COMBOdata <- data1[keepCOMBO, ]
-COMBO.ZScore.H <- COMBOdata$H.corr/0.6870763
-
-#COMBO STD = 0.3489582
-keeptwinsUK <- which(data1$dataset == "twinsUK")
-twinsUKdata <- data1[keeptwinsUK, ]
-twinsUK.ZScore.H <- twinsUKdata$H.corr/0.3489582
-
-#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit, COMBO, twinsUK
-HZScore <- c(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, COMBO.ZScore.H, twinsUK.ZScore.H, amish.ZScore.H)
-data1 <- cbind(HZScore, data1)
-
-turn.data <- cbind(turn.ZScore.H, turn.data)
-rm(Escobar.ZScore.H, GLNE.ZScore.H, CCHC.ZScore.H, HMP.ZScore.H, Korean.ZScore.H, metahit.ZScore.H, amish.ZScore.H, turn.ZScore.H, COMBO.ZScore.H, twinsUK.ZScore.H)
-rm(keepmetahit, keepCCHC, keepescobar, keepGLNE, keepHMP, keepKorean)
+#data1 order Escobar, Baxter, Ross, HMP, Arumugam, Wu, Goodrich, Zupancic
+
+ZH <- c(EscobarZH, BaxterZH, RossZH, HMPZH, ArumugamZH, WuZH, GoodrichZH, ZupancicZH)
+bmi <- c(EscobarBMI, BaxterBMI, RossBMI, HMPBMI, ArumugamBMI, WuBMI, GoodrichBMI, ZupancicBMI)
+ZlogBF <- c(EscobarZLogBF, BaxterZLogBF, RossZLogBF, HMPZLogBF, ArumugamZLogBF, WuZLogBF, GoodrichZLogBF, ZupancicZLogBF)
+
+data1 <- as.data.frame(cbind(ZH, ZlogBF, bmi))
+
+turn.comb.meta <- as.data.frame(cbind(TurnbaughBMICat, TurnbaughObese))
+turn.comb.cont <- as.data.frame(cbind(TurnbaughZH, TurnbaughZLogBF))
+turn.data <- cbind(turn.comb.cont, turn.comb.meta)
+colnames(turn.data) <- c("ZH", "ZLogBF", "BMI.class", "obese")
 
@@
 #For ZScore normalization
-d1.obese <- data1$obese
+d1.obese <- as.character(data1$obese)
 t.obese <- as.character(turn.data$obese)
@@
 d1.bmi.class <- data1$BMI.class
-t.bmi.class <- as.character(turn.data$BMI.category)
-
-d1.H.Zscore <- data1$HZScore
-t.H.Zscore <- turn.data$turn.ZScore.H
+t.bmi.class <- as.character(turn.data$BMI.class)
+
+d1.H.Zscore <- data1$ZH
+t.H.Zscore <- turn.data$ZH
 
@@
 
-setwd("C:/Users/marc/Desktop/obesity2/BF.normalization")
-
 #Let's change the order so it makes sense
@@
 
-#Load in data
-data2 <- read.csv("combined.BF.corr.csv")
-
-#Create BMI groups
-data2$BMI.class[data2$bmi<=24] <- "Lean"
-data2$BMI.class[data2$bmi>24 & data2$bmi<30] <- "Overweight"
-data2$BMI.class[data2$bmi>=30] <- "Obese"
-
-#Create Obese Yes/No groups
-data2$obese[data2$BMI.class=="Lean" | data2$BMI.class=="Overweight"] <- "No"
-data2$obese[data2$BMI.class=="Obese"] <- "Yes"
-
 ##Run Zscore normalized Bacter and Firm overall analysis
 #data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
-d1.zbf.corr <- data2$ZBF
-t.zbf.corr <- turn.data$ZBF
+d1.lzbf.corr <- data1$ZlogBF
+t.lzbf.corr <- turn.data$ZLogBF
 
 #Bacteroidetes/Firmicutes Datasets
-comb.zbf.corr <- as.data.frame(c(d1.zbf.corr, t.zbf.corr))
-group.data.test <- cbind(comb.zbf.corr, group.data.test)
-colnames(group.data.test)[1] <- "zbf"
+comb.lzbf.corr <- as.data.frame(c(d1.lzbf.corr, t.lzbf.corr))
+group.data.test <- cbind(comb.lzbf.corr, group.data.test)
+colnames(group.data.test)[1] <- "ZLogBF"
+
+## Perform the statistical tests
+combZHTtest <- t.test(H.Zscore ~ obese, data=group.data.test) #P-value = 0.006061
+combZHANOVA <- anova(lm(H.Zscore ~ BMI.class, data=group.data.test)) #P-value = 0.0193
+a2 <- aov(H.Zscore ~ BMI.class, data=group.data.test)
+combZHTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
+combZHTukeyFinal <- combZHTukey$BMI.class
+
+combZLogBFTtest <- t.test(ZLogBF ~ obese, data=group.data.test) #P-value = 0.866
+combZLogBFANOVA <- anova(lm(ZLogBF ~ BMI.class, data=group.data.test)) #P-value = 0.106
+a2 <- aov(ZLogBF ~ BMI.class, data=group.data.test)
+combZLogBFTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
+combZLogBFTukeyFinal <- combZLogBFTukey$BMI.class
 
@@
 # Section C
-bf1 <- ggplot(group.data.test, aes(obese, zbf))
+bf1 <- ggplot(group.data.test, aes(obese, ZLogBF))
 bf2 <- bf1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
 bf3 <- bf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-bf4 <- bf3 + xlab("Obese Category") + ylab("ZScore Normalized B/F Ratio")
+bf4 <- bf3 + xlab("Obese Category") + ylab("ZScore Normalized Log B/F Ratio")
 bf5 <- bf4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
@@
 # Section D
-qbf1 <- ggplot(group.data.test, aes(BMI.class, zbf))
+qbf1 <- ggplot(group.data.test, aes(BMI.class, ZLogBF))
 qbf2 <- qbf1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
 qbf3 <- qbf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-qbf4 <- qbf3 + xlab("BMI Classification") + ylab("ZScore Normalized B/F Ratio")
+qbf4 <- qbf3 + xlab("BMI Classification") + ylab("ZScore Normalized Log B/F Ratio")
 qbf5 <- qbf4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
@@
 **Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** ZScore Normalized Shannon Diversity and Obesity.
-There was a significant difference between non-obese (n=1081) and obese groups (n=435) (P-value = 0.005992).
+There was a significant difference between non-obese (n=1081) and obese groups (n=435) (P-value = `r signif(combZHTtest$p.value, 3)`).
 **B)** ZScore Normalized Shannon Diversity by BMI Group.
 Total n for lean, overweight, and obese groups are 563, 435, 518 respectively.
-Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = 0.018) but no difference between lean and obese group (P-value = 0.081).
+Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = `r signif(combZHTukeyFinal[3, 4], 3)`) but no difference between lean and obese group (P-value = `r signif(combZHTukeyFinal[2, 4], 3)`).
 **C)** ZScore Normalized B/F Ratio and Obesity.
-There was no significant difference between the non-obese (n=1081) and obese (n=435) group based on B/F ratio (P-value = 0.456).
+There was no significant difference between the non-obese (n=1081) and obese (n=435) group based on B/F ratio (P-value = `r signif(combZLogBFTtest$p.value, 3)`).
 **D)** ZScore Normalized B/F Ratio by BMI Group.
 Total n for lean, overweight, and obese groups are 563, 435, 518 respectively.
-There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = 0.398).
+There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = `r signif(combZLogBFANOVA[1, 5], 3)`).
 

### 2016-03-05 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
updated references

@@
 There has been a continuous stream of studies that report observing a link between the composition of microbiome and obesity.
-The first such study used genetically obese mice and observed that obese mice had a XXXXXer B/F ratio than lean mice [@Ley2005].
+The first such study used genetically obese mice and observed that obese mice had a lower Bacteroidetes and higher Firmicutes than lean mice (lower B/F ratio associated with obesity) [@ley_obesity_2005].
 Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009].
-Since the publication of these studies there have been a steady stream of publications on the microbiome with associated BMI data but without clear explicit mention of any significant correlations between the microbiome and obesity [@ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @arumugam_enterotypes_2011; @goodrich_human_2014; @yatsunenko_human_2012; @wu_linking_2011; @escobar_gut_2014].
+Since the publication of these studies there have been a steady stream of publications on the microbiome with associated BMI data but without clear explicit mention of any significant correlations between the microbiome and obesity [@ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @escobar_gut_2014].
 Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B/F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
@@
 
-* @ley2006 watched humans lose weight and communities change
-* @turnbaugh20XX did fecal transplants (humans to mice) and transferred phenotype
-* @ley20XX did fecal transplants (pregnant women to mice) and transferred phenotype
+* [@ley_microbial_2006] watched humans lose weight and communities change
+* [@turnbaugh_effect_2009] did fecal transplants (humans to mice) and transferred phenotype
+* [@koren_host_2012] did fecal transplants (pregnant women to mice) and transferred phenotype
 

### 2016-03-08 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
added install packages into Rmd

@@
 
+```{r install_needed_packages, echo=FALSE, message=FALSE, warning=FALSE}
+deps = c("knitr","vegan","epiR","AUCRF","meta","metafor","ggplot2", "plyr", "reshape2", "RColorBrewer", "scales", "gridExtra");
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+
+```
+
 *******
@@
 ```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
-library(knitr)
 
@@
 #Generate Alpha Diversity Measures and alpha diversity table
-library(vegan)
 H <- diversity(microbiome)
@@
 ###########################################################################
-
-library(epiR)
 
@@
 ###########################################################################
-
-library(AUCRF)
 
@@
 
-library(epiR)
-
 #Generate median values and put them into existing alpha.test dataframe
@@
 #generate alpha diversity measures with vegan
-library(vegan)
 H <- diversity(microbiome)
@@
 #generate alpha diversity measures with vegan
-library(vegan)
 H <- diversity(test)
@@
 #Get alpha diversity of the samples
-library(vegan)
-
 H <- diversity(microbiome)
@@
 #Get alpha diversity of the samples
-library(vegan)
-
 H <- diversity(microb.norm)
@@
 
-library(AUCRF)
-
 #Create Obese.num group
@@
 #generate alpha diversity measures with vegan
-library(vegan)
 H <- diversity(s1.subsample.data)
@@
 
-library(meta)
-library(metafor)
-library(ggplot2)
-library(plyr)
-library(reshape2)
-library(RColorBrewer)
-library(scales)
-library(pheatmap)
-library(gridExtra)
-
 StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)

### 2016-03-08 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
fig modifications

@@
 res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
+RRSDPvalue <- anova.rma(res)
 
@@
 **Figure 2: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity.**  Groups were divided for each study on high and low Shannon Diversity groups based on the median for that study.
-The overall pooled relative risk was 1.18 for the low diversity group (P-value = 0.0187).
+The overall pooled relative risk was 1.18 for the low diversity group (P-value = `r signif(RRSDPvalue$QMp, 3)`).
 
@@
 res2 <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
+RRBFPvalue <- anova.rma(res2)
 
@@
 **Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low B/F ratio groups based on the median for that specific study.
-The overall pooled relative risk was 0.86 for the low diversity group (P-value = 0.0376).
+The overall pooled relative risk was 0.87 for the low diversity group (P-value = `r signif(RRBFPvalue$QMp, 3)`).
 

### 2016-03-08 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
Revise Methods section

--- 2016-03-08 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
@@
 *******
-##### Abstract
+## Abstract
 **Rationale:** The hypothesis that changes in the microbiome are associated with obesity has been widely accepted.
@@
 
-*******
-##### Introduction
+\newpage
+
+## Introduction
 Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014].
@@
 
-*******
-##### **Methods:**  
-*Literature Review and Study Inclusion*
-
-We attempted to stick to the original PRISMA guidelines for performing a meta-analysis as published by Moher et al [-@moher_preferred_2010].
-A detailed description of the literature seach and the exact formula we used for the initial PubMed search can be found in the online supplement.
-In brief, a total of 196 studies made it through the initial identification of studies.
-Screening removed a total of 184 studies leaving a total of 12 eligible studies for this meta-analysis.
-Upon a full reading of these specific manuscripts 2 were exculded and the reasons can be found in the online supplement.
-A total of 10 studies were ultimately deemed eligible for qualitative synthesis [Figure 1].
-One final study was removed from the quantitative analysis since no individual in the study was found to be able to be classified in the obese category based on the classical cutoffs.
-A total of 3 studies were from records identified by database searching [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014] and the other 6 were from records identified by other sources [@ross_16s_2015; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @conlan_species-level_2012].
-One of these studies (Baxter) was used on a data set that was not originally designed to investigate obesity in order to help mitigate any possible publication bias associated with the bacterial microbiome and obesity.
-
-Although there have been numerous RT-PCR and qPCR studies directed at examining the microbiome and obesity [@guo_development_2008; @schwiertz_microbiota_2010].
-The main method of choice to obtain the most complete picture of the overall microbial community is through 16S sequencing of a specific hypervariable region on the 16S rRNA gene.
-The sequencers used to complete this task have varied depending on research group with the most popular being the Roche 454 and Illumina MiSeq.
-A priori we decided that for this report we would limit ourselves to analyzing studies were 16S sequencing data is avaliable along with BMI information.
-We also analyzed every study with a uniform approach utilizing the mothur software package for sequence processing and clean up and R statistical software for analysis.
-
-*Subject Demographics and Study Outcomes*
-
-The overall demographics for individuals used in this meta-analysis split by their respective study can be found in table 1.
-A summary of the study type, study population, whether the data set is published, and the outcomes reported can be found in table 2.
-We found a large variation in reported outcomes with two studies finding bacterial community associations with obesity and three studies finding some association between either Firmicutes, Bacteroidetes, or the Bacteroidetes/Firmicutes ratio and obesity.
-However, a total of three manuscripts found no correlations between the microbiome and obesity.
-This is in addition to the two re-analysis studies published in 2014.
-The cutoffs we used in this study for the BMI groups were lean (BMI less than or equal to 24), overweight (BMI was greater than 24 but less than 30), and obese (BMI greater than or equal to 30).
-For analysis involving obese and non-obese groups the cutoffs we used were a BMI less than 30 (lean/non-obese) and a BMI of equal to or greater than 30 (obese).
-
-*Sequence Analysis Pipeline*
-
-All data was publically available and the sequence data was downloaded from the NCBI Sequence Read Archive, MG-RAST, or the European Nucleotide Archive.
-A total of 6/9 [@turnbaugh_core_2009; @ross_16s_2015; @zupancic_analysis_2012; @wu_linking_2011; @escobar_gut_2014; @conlan_species-level_2012] studies were processed by staying as close as possible to the mothur 454 standard operating procedure, which can be found at http://www.mothur.org/wiki/454_SOP.
-A total of 2/9 [@goodrich_human_2014] of the studies were processed by staying as close as possible to the mothur MiSeq standard operating procedure found at http://www.mothur.org/wiki/MiSeq_SOP.
-The final data set [@arumugam_enterotypes_2011] that was used took advantage of the Metaphlan2 package [@truong_metaphlan2_2015] to generate the necessary data for this analysis.
-A comparison of the similarity of Metaphlan2 to 16S sequencing using the Human Microbiome Project (HMP) data set can be found in the online supplement [Figure S1-S3].
-For the studies that utilized 16S rRNA gene sequencing three studies used primers targeting the V1-V3 region [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], two studies used the V4 region [@goodrich_human_2014], and one study used the V1-V2 region [@wu_linking_2011].
-Where there were two choices of data sets to use we used the V3-V5 data of the HMP [@conlan_species-level_2012] and the V2 data from the Turnbaugh study [-@turnbaugh_core_2009].
-A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/.
-
-*Data Analysis*
-
-We performed the analysis using R version 3.2.2 and R studio version 0.99.491.
-We split the overall analysis into three general categories.
-The first category involved following the same method employed by Finucane et al.
-and Walters et al.
-[-@finucane_taxonomic_2014; -@walters_meta-analyses_2014] in which each study was re-analyzed seperately for associations with BMI.
-The main associations that we explored for BMI in each seperate data set were the Bacteroidetes phyla, Firmicutes phyla, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, NMDS visualization of Bray-Curtis distance matrix PERMANOVA, and relative risk (RR).
-A two-way T-test was performed for comparison for non-obese versus obese.
-Finally, Random forest analysis to identify the best variables for classification of obese and non-obese was performed to look if there were similarities between data sets that we may have missed.
-The R package AUCRF (v1.1) was used for this.
-
-For the second category of analysis we pooled all the data from each seperate study together to asses whether the combined direction of all the studies increased an individual's RR for being obese.
-We investigated both Shannon diversity and the B/F ratio in detail for this part.
-In order to generate the RR we used the median value of each study and the number of those that were non-obese or obese in each group (above or below the median respectively) was recorded and a Fisher exact-test was performed.
-The relative risk from the the Fisher exact-test was obtained using the epiR (v0.9-6.9) package in R.
-We obtained the pooled RR and confidence intervals (CI) from the metafor (v1.9-8) package in R.
-Finally, we also assessed the potential bias in the two pooled analysis by using a funnel plot.
-
-The third and final category involved using  Zscore normalizations to pool all the data together to investigate the correlation between either BMI and Shannon diversity or BMI and the B/F ratio.
-We used a two-way t-test on the Zscore values to investigate if there was a difference between non-obese and obese individuals.
-In order to asses differences between the BMI categories (if data was available) we used ANOVA with a tukey post hoc test.
-Where appropriate we applied a Bonferroni correction for multiple comparisons.
-In all other cases a P-value under 0.05 was considered to be significant.
-
-*******
-##### Results:
-
-*Seperate Analysis:*
-
-We used the categorical variable of obese or non-obese to analyze each data set seperately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
+## Methods
+***Literature Review and Study Inclusion***
+We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010].
+A detailed description of our selection process and the exact search terms are provided in the Supplement [Figure 1].
+Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome.
+The initial search yielded 187 studies.
+We identified nine additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
+We then manually curated the 196 studies to select those studies that included BMI and sequence data.
+This yielded 10 eligible studies.
+An additional study was removed from our analysis because no individuals in the study had a BMI over 30.
+Among the final 9 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@ross_16s_2015; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @Huttenhower2012], and one dataset was used (Baxter) because at the time of our analysis it was unpublished and enabled us to mitigate against publication biases associated with the bacterial microbiome and obesity.
+The nine studies are summarized in Tables 1 and 2.
+
+***Sequence Analysis Pipeline***
+All sequence data were publicly available and were downloaded from the NCBI Sequence Read Archive, the European Nucleotide Archive, or the investigators' personal website.
+The nine studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011; @Huttenhower2012] or Illumina sequencing [@goodrich_human_2014; @arumugam_enterotypes_2011; Baxter].
+All of these studies used amplification-based 16S rRNA gene sequencing except for one, which used metagenomic shotgun sequencing [@arumugam_enterotypes_2011].
+Among the studies that sequenced the 16S rRNA gene, the researchers targeted the V1-V2  [@wu_linking_2011], V1-V3 [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], V3-V5 [@Huttenhower2012], and V4 [@goodrich_human_2014] regions.
+For those studies where multiple regions were sequenced,  we selected the region that corresponded to the largest number of subjects [@turnbaugh_core_2009; @Huttenhower2012].
+We processed the 16S rRNA gene sequence data as described in the original studies.
+The taxonomic data from the metagenomic shotgun sequence data were obtained using Metaphlan2 [@truong_metaphlan2_2015].
+A detailed and reproducible description of how the data were processed can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/.
+
+***Data Analysis***
+We performed the analysis using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`).
+We split the overall analysis into three general strategies.
+In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, and number of OTUs.
+A two-tailed Wilcoxon test was performed for comparison for non-obese and obese individuals.
+We also measured the Spearman correlation coefficient between each metric and the actual BMI value.
+Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices.
+This analysis was performed using the vegan R package.
+Finally, we used the AUCRF R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguish non-obese and obese individuals.
+
+In the second strategy, we pooled the relative risk (RR) values for each study.
+For each study we partitioned the subjects into a low or high group depending on whether their diversity or B/F was below or above the median value for the study.
+The RR was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group.
+We then performed a Fisher exact-test to test whether the RR was significantly different from 1.0 using the epiR (v0.9-6.9) and metafor (v1.9-8) packages.
+Finally, we assessed the potential publication bias in the diversity and B/F analyses using a funnel plot.
+
+In the third strategy, we normalized the diversity and B/F data within each study and pooled the normalized data across studies.
+We applied a Z-score normalization to the Shannon diversity data and log-transformed B/F data within each study to have a mean of zero and standard deviation of one.
+We then pooled Z-scores across studies and used a two-tailed T-test on the values to test for a significant difference in Z-scores among non-obese and obese individuals.
+To assess differences between the lean, overweight, and obese BMI categories we used ANOVA with a Tukey post hoc test.
+We also measured the Spearman correlation coefficient between the Z-scores for each metric and BMI values.
+
+## Results:
+
+***Separate Analysis***
+We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
 Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 2 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix.
@@
 
-*Classical Pooled Meta-analysis:*
+***Classical Pooled Meta-analysis***
 
@@
 
-The final forest plot that we used for this analysis investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC in the the seperate study analysis looked when pooled together to assess the RR [Figure S4].
+The final forest plot that we used for this analysis investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC in the the separate study analysis looked when pooled together to assess the RR [Figure S4].
 There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.96 (CI 0.60, 1.52), P-value = 0.8575).
@@
 
-*ZScore Normalization Pooled Analysis:*
-
-Using the third approach with Zscore normalization we found that there was a significant difference in the Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A].
+***Z-Score Normalization Pooled Analysis***
+
+Using the third approach with Z-score normalization we found that there was a significant difference in the Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A].
 Further there was also a significant difference in this normalized Shannon diversity value when the data was split between the three BMI groups (P-value = 0.03826) [Figure 5B].
 Using a Tukey post hoc test this difference was driven by the overweight versus the obese group (P-value = 0.03945).
-However, in contrast to this result we found that the B/F ratio, using the ZScore normalization, was not significant different between obese and non-obese indivdiuals (P-value = 0.4516) [Figure 5C].
-Additionally, we did not observe a difference when the indivdiuals were seperated by BMI groups (P-value = 0.3317) [Figure 5D].
+However, in contrast to this result we found that the B/F ratio, using the Z-score normalization, was not significant different between obese and non-obese individuals (P-value = 0.4516) [Figure 5C].
+Additionally, we did not observe a difference when the individuals were separated by BMI groups (P-value = 0.3317) [Figure 5D].
 
@@
 We analyze the data using three different approaches.
-The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses ZScore normalizations and does not use RR.
+The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses Z-score normalizations and does not use RR.
 
@@
 First,  it has been documented that in a number of groups BMI is not a good classification of obesity [@who_expert_consultation_appropriate_2004; @rahman_accuracy_2010].
-Of particular note, the asian population has a seperate set of guidelines for obesity [@who_expert_consultation_appropriate_2004].
+Of particular note, the asian population has a separate set of guidelines for obesity [@who_expert_consultation_appropriate_2004].
 Although the asain population in the studies analyzed are very small [Table 1] it is possible that this may contribute to some of the noise in the data.
@@
 Fourth, although a significant result exists for the RR for both Shannon diversity and the B/F ratio, it is small.
-This is also the case for the ZScore normalization analysis of the Shannon diversity.
+This is also the case for the Z-score normalization analysis of the Shannon diversity.
 As an example, a lower than median Shannon diversity only has an overall 18% increase of obesity in it versus the higher than median Shannon diversity group.
@@
 
-*******
-##### Tables:
-
-**Table 1:Summary Demographics of Individuals used in the Meta-analysis.**     
+\newpage
+
+## Tables
+
+**Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
 ```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
-*Only BMI group information was provided and only an age range was given.
-
-**Table 2:Summary of Important Study Characteristics**
+* Only BMI group information was provided and only an age range was given.
+
+\newpage
+
+**Table 2. Summary of Important Study Characteristics.**
 
@@
 
-Outcomes <- c("N/A (Unplublished)", "Mentions obesity, no specific correlations made though", "Christensenellaceae higher in low BMI individuals.
+Outcomes <- c("N/A (Unpublished)", "Mentions obesity, no specific correlations made though", "Christensenellaceae higher in low BMI individuals.
 No other correlations found.", "Firmicutes less abundant at higher BMI", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "No significant correlations", "Unweighted UniFrac Correlation with BMI", "No significant  correlations", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ")
@@
 
-**Table 3:Summary of P-values for Measurements of Interest for each Individual Study for Obese versus Normal**
+**Table 3. Summary of P-values for Measurements of Interest for each Individual Study for Obese versus Normal**
 
@@
 ###########################################################################
-############ ZScore Data Preperation ###################################
+############ Z-score Data Preparation ###################################
 ###########################################################################
@@
 #Create group analysis table to harness full data set
-#For ZScore normalization
+#For Z-score normalization
 d1.obese <- as.character(data1$obese)
@@
 
-d1.H.Zscore <- data1$ZH
-t.H.Zscore <- turn.data$ZH
+d1.H.Z-score <- data1$ZH
+t.H.Z-score <- turn.data$ZH
 
@@
 comb1.bmi.class <- c(d1.bmi.class, t.bmi.class)
-comb1.H.Zscore <- as.data.frame(c(d1.H.Zscore, t.H.Zscore))
-
-group.data.test1 <- cbind(comb1.H.Zscore, comb1.obese, comb1.bmi.class)
-colnames(group.data.test1) <- c("H.Zscore", "obese", "BMI.class")
+comb1.H.Z-score <- as.data.frame(c(d1.H.Z-score, t.H.Z-score))
+
+group.data.test1 <- cbind(comb1.H.Z-score, comb1.obese, comb1.bmi.class)
+colnames(group.data.test1) <- c("H.Z-score", "obese", "BMI.class")
 
@@
 
-##Run Zscore normalized Bacter and Firm overall analysis
+##Run Z-score normalized Bacter and Firm overall analysis
 #data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
@@
 ## Perform the statistical tests
-combZHTtest <- t.test(H.Zscore ~ obese, data=group.data.test) #P-value = 0.006061
-combZHANOVA <- anova(lm(H.Zscore ~ BMI.class, data=group.data.test)) #P-value = 0.0193
-a2 <- aov(H.Zscore ~ BMI.class, data=group.data.test)
+combZHTtest <- t.test(H.Z-score ~ obese, data=group.data.test) #P-value = 0.006061
+combZHANOVA <- anova(lm(H.Z-score ~ BMI.class, data=group.data.test)) #P-value = 0.0193
+a2 <- aov(H.Z-score ~ BMI.class, data=group.data.test)
 combZHTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
@@
 # Section A
-p1 <- ggplot(group.data.test, aes(obese, H.Zscore))
+p1 <- ggplot(group.data.test, aes(obese, H.Z-score))
 p2 <- p1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
 p3 <- p2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-p4 <- p3 + xlab("Obese Category") + ylab("ZScore Normalized H")
+p4 <- p3 + xlab("Obese Category") + ylab("Z-score Normalized H")
 p5 <- p4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
@@
 # Section B
-q1 <- ggplot(group.data.test, aes(BMI.class, H.Zscore))
+q1 <- ggplot(group.data.test, aes(BMI.class, H.Z-score))
 q2 <- q1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
 q3 <- q2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-q4 <- q3 + xlab("BMI Classification") + ylab("ZScore Normalized H")
+q4 <- q3 + xlab("BMI Classification") + ylab("Z-score Normalized H")
 q5 <- q4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
@@
 bf3 <- bf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-bf4 <- bf3 + xlab("Obese Category") + ylab("ZScore Normalized Log B/F Ratio")
+bf4 <- bf3 + xlab("Obese Category") + ylab("Z-score Normalized Log B/F Ratio")
 bf5 <- bf4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
@@
 qbf3 <- qbf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-qbf4 <- qbf3 + xlab("BMI Classification") + ylab("ZScore Normalized Log B/F Ratio")
+qbf4 <- qbf3 + xlab("BMI Classification") + ylab("Z-score Normalized Log B/F Ratio")
 qbf5 <- qbf4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
@@
 
-**Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** ZScore Normalized Shannon Diversity and Obesity.
+**Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** Z-score Normalized Shannon Diversity and Obesity.
 There was a significant difference between non-obese (n=1081) and obese groups (n=435) (P-value = `r signif(combZHTtest$p.value, 3)`).
-**B)** ZScore Normalized Shannon Diversity by BMI Group.
+**B)** Z-score Normalized Shannon Diversity by BMI Group.
 Total n for lean, overweight, and obese groups are 563, 435, 518 respectively.
 Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = `r signif(combZHTukeyFinal[3, 4], 3)`) but no difference between lean and obese group (P-value = `r signif(combZHTukeyFinal[2, 4], 3)`).
-**C)** ZScore Normalized B/F Ratio and Obesity.
+**C)** Z-score Normalized B/F Ratio and Obesity.
 There was no significant difference between the non-obese (n=1081) and obese (n=435) group based on B/F ratio (P-value = `r signif(combZLogBFTtest$p.value, 3)`).
-**D)** ZScore Normalized B/F Ratio by BMI Group.
+**D)** Z-score Normalized B/F Ratio by BMI Group.
 Total n for lean, overweight, and obese groups are 563, 435, 518 respectively.

### 2016-03-08 code/sequenceProcessing.Rmd
update sequence Processing Rmd

@@
 There are also a few reference files that you will need.
-The first is the [SILVA reference](http://www.mothur.org/wiki/Silva_reference_files) database and the second is the [RDP database](http://www.mothur.org/wiki/RDP_reference_files) of which we used release 102 and version 9 respectively.
+The first is the [SILVA reference](http://www.mothur.org/wiki/Silva_reference_files) database and the second is the [RDP database](http://www.mothur.org/wiki/RDP_reference_files) of which we used release 102 and version 9 respectively or the [greengenes database](http://www.mothur.org/wiki/Greengenes-formatted_databases) depending on what was used by each respective study.
 In general, we tried to keep close to the Schloss SOP for processsing sequence reads.
@@
 
-So now the biggest challenge that I have found so far is getting the data into a distance matrix not based on the closed-reference approach.
-The file is just too big for cluster.split so we have had to use the closed-reference approach for this data set.
+After this I seperated all the files and ran *chimera.uchime* on each of the samples in smaller groups.
+After this I recombined everything back together and then used the greengenes database to classify the sequences along with using the closed reference approach adopted by Goodrich et al.
+
+```{r, engine='bash', eval=FALSE}
+/share/scratch/schloss/mothur/mothur "#split.groups(fasta=combined.unique.good.filter.unique.precluster.fasta, count=combined.unique.good.filter.unique.precluster.count_table)"
+
+#!/bin/sh
+
+#Create new directories for chimera completed files
+mkdir chimera.completed
+mkdir mothurLogfiles
+
+#Make loop to run the chimera step and move completed samples to respective directory
+for chim in $(ls *.fasta)
+do
+	CFile=${chim//fasta/count_table}
+	/share/scratch/schloss/mothur/mothur "#chimera.uchime(fasta=$chim, count=$CFile, dereplicate=t, processors=8)"
+
+	mv $chim chimera.completed
+	mv $CFile chimera.completed
+	mv *logfile mothurLogfiles
+
+done
+
+exit 0
+
+#!/bin/sh
+
+/share/scratch/schloss/mothur/mothur "#merge.files(input=group1.fasta-group2.fasta-group3.fasta-group4.fasta-group5.fasta-group6.fasta-group7.fasta-group8.fasta-group9.fasta-group10.fasta-group11.fasta-group12.fasta-group13.fasta-group14.fasta-group15.fasta-group16.fasta-group17.fasta-group18.fasta-group19.fasta-group20.fasta-group21.fasta-group22.fasta-group23.fasta-group24.fasta-group25.fasta, output=combined.fasta)"
+/share/scratch/schloss/mothur/mothur "#merge.files(input=group1.groups-group2.groups-group3.groups-group4.groups-group5.groups-group6.groups-group7.groups-group8.groups-group9.groups-group10.groups-group11.groups-group12.groups-group13.groups-group14.groups-group15.groups-group16.groups-group17.groups-group18.groups-group19.groups-group20.groups-group21.groups-group22.groups-group23.groups-group24.groups-group25.groups, output=combined.groups)"
+/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.fasta)"
+/share/scratch/schloss/mothur/mothur "#count.seqs(name=combined.names, group=combined.groups)"
+
+mv *logfile mothurLogfiles
+
+exit 0
+```
+
+This here is the final line of code needed to generate the necessary shared files for downstream statistical analysis.
 

### 2016-03-11 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
moving heavy lifting to .R file

@@
 
+```{r Loading_data_tables, echo=FALSE, message=FALSE, warning=FALSE}
+
+setwd("C:/users/marc/Desktop")
+demographics <- read.csv("denovodemographicsTable.csv")
+OverallPValue <- read.csv("denovoOverallPTable.csv")
+ShannonRRTable <- read.csv("denovoShannonRRTable.csv")
+BFRatioRRTable <- read.csv("denovoBFRatioRRTable.csv")
+ZScoreCombinedData <- read.csv("denovoCombinedData.csv")
+ZScoreTurnbaughData <- read.csv("denovoTurnbaughData.csv")
+
+```
+
+```{r install_needed_packages, echo=FALSE, message=FALSE, warning=FALSE}
+deps = c("knitr","vegan","epiR","AUCRF","meta","metafor","ggplot2", "plyr", "reshape2", "RColorBrewer", "scales", "gridExtra");
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+
+```
+
 *******
@@
 ***Sequence Analysis Pipeline***
-All sequence data were publicly available and were downloaded from the NCBI Sequence Read Archive, the European Nucleotide Archive, or the investigators' personal website.
+All sequence data were publicly available and were downloaded from the NCBI Sequence Read Archive, the European Nucleotide Archive, or the investigators' personal website (https://gordonlab.wustl.edu/TurnbaughSE\_10\_09/STM\_2009.html).
 The nine studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011; @Huttenhower2012] or Illumina sequencing [@goodrich_human_2014; @arumugam_enterotypes_2011; Baxter].
@@
 
-```{r install_needed_packages, echo=FALSE, message=FALSE, warning=FALSE}
-deps = c("knitr","vegan","epiR","AUCRF","meta","metafor","ggplot2", "plyr", "reshape2", "RColorBrewer", "scales", "gridExtra");
-for (dep in deps){
-  if (dep %in% installed.packages()[,"Package"] == FALSE){
-    install.packages(as.character(dep), quiet=TRUE);
-  }
-  library(dep, verbose=FALSE, character.only=TRUE)
-}
-
-```
-
 \newpage
@@
 
-Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "HMP", "Wu", "Arumugam", "Turnbaugh*")
-
-Age <- c("54.3 +/- 9.9", "57.0 +/- 11.2", "61.5 +/- 8.9", "38.1 +/- 11.1", "46.7 +/- 13.1", "26.1 +/- 5.0", "26.4 +/- 9.4", "56.5 +/- 7.7", "21 - 32")
+Study <- as.character(demographics$Study)
+
+Age <- as.character(demographics$meanAge)
 
@@
 
-columbian.microb <- read.table("columbian.0.03.subsample.shared", header=T)
+columbian.microb <- read.table("EscobarPhylotypeSub.shared", header=T)
 metadata <- read.csv("columbian_dataset.csv")
@@
 #combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
+phylogenetic.info <- read.table("taxonomyKey.txt")
 rownames(phylogenetic.info) <- phylogenetic.info[,1]

### 2016-03-11 code/StrategydenovoAnalysis.R

@@ -208,17 +208,17 @@ BaxterBMI <- bmi
 
 #Get data for demographics table
 totalN <- length(rownames(demographics))
-meanAge <- mean(demographics$Age)
-SDAge <- sd(demographics$Age)
+meanAge <- format(round(mean(demographics$Age), 2), nsmall = 2)
+SDAge <- format(round(sd(demographics$Age), 2), nsmall = 2)
 temporary <- table(demographics$Gender)
 males <- temporary[names(temporary) == "m"]
 females <- temporary[names(temporary) == "f"]
 temporary <- table(demographics$White)
-ancestry <- temporary[names(temporary) == 1] / sum(temporary)
-meanBMI <- mean(demographics$BMI)
-SDBMI <- sd(demographics$BMI)
-minBMI <- min(demographics$BMI)
-maxBMI <- max(demographics$BMI)
+ancestry <- format(round(temporary[names(temporary) == 1] / sum(temporary), 2), nsmall = 2)
+meanBMI <- format(round(mean(demographics$BMI), 2), nsmall = 2)
+SDBMI <- format(round(sd(demographics$BMI), 2), nsmall = 2)
+minBMI <- format(round(min(demographics$BMI), 2), nsmall = 2)
+maxBMI <- format(round(max(demographics$BMI), 2), nsmall = 2)
 
 # Get data for P-value table
 overallPTable <- as.data.frame(t(c(baxterBacter$p.value, baxterFirm$p.value, 
@@ -467,17 +467,23 @@ RossBMI <- bmi
 
 #Get data for demographics table
 totalN <- c(totalN, length(rownames(edit.metadata2)))
-meanAge <- c(meanAge, mean(edit.metadata2$age_at_visit))
-SDAge <- c(SDAge, sd(edit.metadata2$age_at_visit))
+meanAge <- c(meanAge, 
+             format(round(
+               mean(edit.metadata2$age_at_visit), 2), nsmall = 2))
+             
+SDAge <- c(SDAge, 
+           format(round(
+             sd(edit.metadata2$age_at_visit), 2), nsmall = 2))
+
 temporary <- table(edit.metadata2$sex)
 males <- unname(c(males, temporary[names(temporary) == "M"]))
 females <- unname(c(females, temporary[names(temporary) == "F"]))
[…]

### 2016-03-13 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
Manuscript Code Update

@@
 Age <- as.character(demographics$meanAge)
-
-Sex <- c("111|61", "48|15", "505|2", "14|16", "112|87", "127|129", "31|27", "45|40", "Not Available")
-
-Ancestry <- c(87.8, 0, "Not Available", 0, 100, 83, "Not Available", "Not Available", 51.4)
-
-BMI <- c("27.0 +/- 5.3", "31.6 +/- 5.3", "26.3 +/- 4.9", "27.4 +/- 4.5", "27.7 +/- 5.3", "24.1 +/- 3.4", "24.6 +/- 4.8", "27.8 +/- 5.8", "Not Available")
-
-Min <- c(17.5, 22.1, 16.2, 19.5, 18.2, 19, 14.0, 18.6, "Not Available")
-
-Max <- c(46.9, 47.9, 44.8, 37.6, 51.1, 34, 41.3, 40.2, "Not Available")
-
-table1 <- cbind(Study, Age, Sex, Ancestry, BMI, Min, Max)
-
-colnames(table1) <- c("Study", "Age (Mean +/- SD)", "Sex (F|M)", "European Ancestry (%)", "BMI (Mean +/- SD)", "Min BMI","Max BMI")
+AgeSD <- as.character(format(round(
+  demographics$SDAge, 2), nsmall = 2))
+AgeTable <- paste(Age," (",AgeSD,")", sep="")
+
+Sex <- paste(demographics$females, "|", demographics$males)
+
+Ancestry <- as.character(format(round(
+  demographics$ancestry*100, 2), nsmall = 2))
+
+BMI <- paste(demographics$meanBMI, " (", demographics$SDBMI, ")", sep="")
+
+Min <- as.character(demographics$minBMI)
+
+Max <- as.character(demographics$maxBMI)
+
+table1 <- cbind(Study, AgeTable, Sex, Ancestry, BMI, Min, Max)
+table1[table1=="NA" | table1=="NA | NA" | table1=="NA (NA)" | 
+         table1=="      NA"] <- "Not Available"
+
+colnames(table1) <- c("Study", "Age (SD)", "Sex (F|M)", "European Ancestry (%)", "BMI (SD)", "Min BMI","Max BMI")
 
@@
 
-* Only BMI group information was provided and only an age range was given.
+* For Turnbaugh only BMI group information was provided and only an age range was given.
 
@@
 
-```{r Data_Baxter, echo=FALSE, message=FALSE, warning=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-# Reading in the necessary Data
-setwd("C:/Users/marc/Desktop/obesity2/data1")
-demographics <- read.csv("demographics.v2.csv")
-microbiome <- read.csv("data1.subsample.otus.csv")
-phylogenetic.info <- read.csv("data1.summary.taxonomy.csv")
-taxonomy <- read.csv("data1.taxonomy.csv")
-
-# Minor modifications and Rownames adustments of data tables
-rownames(demographics) <- demographics[,1]
-rownames(microbiome) <- microbiome[,2]
-demographics <- demographics[,-1]
-microbiome <- microbiome[,-2]
-test.samples <- rownames(demographics)
-microbiome <- microbiome[,-c(1:2)]
-
-# Create Table for phyla information
-phylogenetic.info <- phylogenetic.info[-c(1:12),]
-phyla <- which(phylogenetic.info$taxlevel == 2)
-phyla.table <- phylogenetic.info[phyla, ]
-phyla.table <- phyla.table[,-c(1:2,4:5)]
-phyla.names <- as.character(phyla.table[,1])
-phyla.table <- phyla.table[,-1]
-phyla.table <- as.data.frame(t(phyla.table))
-colnames(phyla.table) <- phyla.names
-rownames(phyla.table) <- rownames(demographics)
-
-#Add phyla together that are not very abundant and delete them from the table
-phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deferribacteres", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "Cyanobacteria_Chloroplast", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(1, 4:7, 9:11, 13, 15)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
-
-#Generate Alpha Diversity Measures and alpha diversity table
-H <- diversity(microbiome)
-S <- specnumber(microbiome)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-
-#Create age quartiles
-
-demographics$age.quartile <- with(demographics, cut(Age, breaks=quantile(Age, probs=seq(0,1, by=0.25)), include.lowest=TRUE))
-
-#Create Obese Yes/No groups
-demographics$obese[demographics$BMI.classification=="Normal" | demographics$BMI.classification=="Overweight"] <- "No"
-demographics$obese[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Yes"
-
-#Create Obese.num groups
-demographics$obese.num[demographics$obese=="No"] <- 0
-demographics$obese.num[demographics$obese=="Yes"] <- 1
-obese <- factor(demographics$obese.num)
-
-######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-
-#Create a column with obese and extreme obese as single entity
-demographics$BMIclass2[demographics$BMI.classification=="Normal"] <- "Normal"
-demographics$BMIclass2[demographics$BMI.classification=="Overweight"] <- "Overweight"                     
-demographics$BMIclass2[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Obese"
-
-##Test BMI versus alpha diversity and phyla
-
-bmi <- demographics$BMI
-
-baxterH <- anova(lm(H ~ obese)) #P-value=0.01643
-baxterS <- anova(lm(S ~ obese)) #P-value=0.01571
-baxterJ <- anova(lm(J ~ obese)) #P-value=0.03241
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-baxterBacter <- anova(lm(bacter ~ obese)) #P-value=0.3175
-baxterFirm <- anova(lm(firm ~ obese)) #P-value=0.6817
-baxterBF <- anova(lm(BFratio ~ obese)) #P-value=0.6305
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-baxter2 <- adonis(microbiome ~ obese, permutations=1000)
-baxterPERM <- baxter2$aov.tab
-#Not significant PERMANOVA = 0.06394, pseudo-F = 1.4246
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-#Generate median values and put them into existing alpha.test dataframe
-
-#Shannon diversity
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-##Run the RR for Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-orderedHCat <- test3[, 1]
-BaxtotalN <- length(orderedHCat)
-BaxHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
-BaxHighShannonGroup <- as.data.frame(table(test3[c(1:BaxHHighTotal), 2]))
-BaxLowShannonGroup <- as.data.frame(table(test3[c((BaxHHighTotal + 1):BaxtotalN), 2]))
-#Group1 (Higher than median), obese = 22 and non-obese = 64
-#Group2 (Lower than median), obese = 25 and non-obese = 61
-
-group1 <- c(BaxHighShannonGroup[2, 2], BaxHighShannonGroup[1, 2])
-group2 <- c(BaxLowShannonGroup[2, 2], BaxLowShannonGroup[1, 2])
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-baxterHEpi <- epi.2by2(r.test, method="cohort.count")
-baxterHMassoc <- baxterHEpi$massoc
-baxterHRR <- baxterHMassoc$RR.strata.score
-baxterHRRsig <- baxterHMassoc$chisq.strata
-## Risk Ratio = 1.14
-## CI = 0.70, 1.85
-## p-value = 0.608
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-orderedBFCat <- test4[, 1]
-BaxBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
-BaxHighBFGroup <- as.data.frame(table(test4[c(1:BaxBFHighTotal), 2]))
-BaxLowBFGroup <- as.data.frame(table(test4[c((BaxBFHighTotal + 1):BaxtotalN), 2]))
-#Group1 (Higher than median), obese = 23 and non-obese = 63
-#Group2 (Lower than median), obese = 24 and non-obese = 62
-
-group1 <- c(BaxHighBFGroup[2, 2], BaxHighBFGroup[1, 2])
-group2 <- c(BaxLowBFGroup[2, 2], BaxLowBFGroup[1, 2])
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-baxterBFEpi <- epi.2by2(r.test, method="cohort.count")
-baxterBFMassoc <- baxterBFEpi$massoc
-baxterBFRR <- baxterBFMassoc$RR.strata.score
-baxterBFRRsig <- baxterBFMassoc$chisq.strata
-## Risk Ratio = 1.04
-## CI = 0.64, 1.70
-## p-value = 0.864
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-#generate test set
-  # get rid of values that only have 0 and something else
-testset <- Filter(function(x)(length(unique(x))>2), microbiome)
-  # get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), microbiome)
-
-#Need to add phyla and alpha diversity measures to the dataset
-
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-testset <- cbind(obese, testset)
-
-#Try AUCRF with default measures provided in readme
-#set.seed(3)
-#baxterAUCfit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-
-###########################################################################
-############ Z-score Data Preparation ###################################
-###########################################################################
-
-BaxterZH <- scale(H)
-BaxterZLogBF <- scale(log(BFratio))
-BaxterBMI <- bmi
-
-```
-
-```{r Data_Ross, echo=FALSE, message=FALSE, warning=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/hispanic/")
-
-hispanic.microb <- read.table("hispanic.subsample.shared", header=T)
-
-metadata <- read.csv("s40168-015-0072-y-s1.csv")
-sample.match <- read.csv("Hispanic_dataset.csv")
-
-#Create a microbiome data table with sample names from metadata
-test <- cbind(as.character(sample.match$Run_s), as.character(sample.match$Library_Name_s))
-test <- test[order(test[, 1]), ]
-
-test2 <- cbind(test, hispanic.microb)
-rownames(test2) <- test2[, 2]
-his.microb.edit <- test2[, -c(1:5)]
-edit.metadata <- metadata[, -c(2:5)]
-rownames(edit.metadata) <- edit.metadata[, 1]
-edit.metadata <- edit.metadata[, -1]
-
-#Create a metadata file in the order of the microbiome data
-order1 <- rownames(his.microb.edit)
-edit.metadata2 <- edit.metadata[order1, ]
-
-#Get alpha diversity of the samples
-H <- diversity(his.microb.edit)
-S <- specnumber(his.microb.edit)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(his.microb.edit)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- his.microb.edit
-colnames(phyla.table) <- phyla.names
-  #add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-  #combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Deinococcus-Thermus", "Elusimicrobia", "Fusobacteria", "Lentisphaerae", "Synergistetes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(3:4, 6:7, 9:10)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-phyla.table.rel.abund <- as.data.frame(phyla.table.rel.abund[, -8])
-
-#Create BMI groups
-edit.metadata2$BMI.class[edit.metadata2$BMI<=24] <- "Normal"
-edit.metadata2$BMI.class[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
-edit.metadata2$BMI.class[edit.metadata2$BMI>=30 & edit.metadata2$BMI<40] <- "Obese"
-edit.metadata2$BMI.class[edit.metadata2$BMI>=40] <- "Extreme Obesity"
-
-#Create Obese Yes/No groups
-edit.metadata2$obese[edit.metadata2$BMI.class=="Normal" | edit.metadata2$BMI.class=="Overweight"] <- "No"
-edit.metadata2$obese[edit.metadata2$BMI.class=="Obese" | edit.metadata2$BMI.class=="Extreme Obesity"] <- "Yes"
-
-#Create BMI class 2 groups
-edit.metadata2$BMI.class2[edit.metadata2$BMI<=24] <- "Normal"
-edit.metadata2$BMI.class2[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
-edit.metadata2$BMI.class2[edit.metadata2$BMI>=30] <- "Obese"
-
-#Get paitent demographics to be tested
-
-bmi <- edit.metadata2$BMI
-obese <- factor(edit.metadata2$obese)
-
-######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-
-rossH <- anova(lm(H ~ obese)) #P-value=0.1657
-rossS <- anova(lm(S ~ obese)) #P-value=0.1841
-rossJ <- anova(lm(J ~ obese)) #P-value=0.2538
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-rossBacter <- anova(lm(bacter ~ obese)) #P-value=0.2092
-rossFirm <- anova(lm(firm ~ obese)) #P-value=0.2569
-rossBF <- anova(lm(BFratio ~ obese)) #P-value=0.2508
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-ross2 <- adonis(his.microb.edit ~ obese, permutations=1000)
-rossPERM <- ross2$aov.tab
-#PERMANOVA=0.7153, pseudo-F=0.73249
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-orderedHCat <- test3[, 1]
-RosstotalN <- length(orderedHCat)
-RossHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
-RossHighShannonGroup <- as.data.frame(table(test3[c(1:RossHHighTotal), 2]))
-RossLowShannonGroup <- as.data.frame(table(test3[c((RossHHighTotal + 1):RosstotalN), 2]))
-group1 <- c(RossHighShannonGroup[2, 2], RossHighShannonGroup[1, 2])
-group2 <- c(RossLowShannonGroup[2, 2], RossLowShannonGroup[1, 2])
-#Group1 (Higher than median), obese = 18 and non-obese = 13
-#Group2 (Lower than median), obese = 20 and non-obese = 12
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group1", "group2")
-
-rossHEpi <- epi.2by2(r.test, method="cohort.count")
-rossHMassoc <- rossHEpi$massoc
-rossHRR <- rossHMassoc$RR.strata.score
-rossHRRsig <- rossHMassoc$chisq.strata
-## Risk Ratio = 1.08
-## CI = 0.72, 1.61
-## p-value = 0.719
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-orderedBFCat <- test4[, 1]
-RossBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
-RossHighBFGroup <- as.data.frame(table(test4[c(1:RossBFHighTotal), 2]))
-RossLowBFGroup <- as.data.frame(table(test4[c((RossBFHighTotal + 1):RosstotalN), 2]))
-group1 <- c(RossHighBFGroup[2, 2], RossHighBFGroup[1, 2])
-group2 <- c(RossLowBFGroup[2, 2], RossLowBFGroup[1, 2])
-#Group1 (Higher than median), obese = 22 and non-obese = 9
-#Group2 (Lower than median), obese = 16 and non-obese = 16
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-rossBFEpi <- epi.2by2(r.test, method="cohort.count")
-rossBFMassoc <- rossBFEpi$massoc
-rossBFRR <- rossBFMassoc$RR.strata.score
-rossBFRRsig <- rossBFMassoc$chisq.strata
-## Risk Ratio = 0.70
-## CI = 0.47, 1.07
-## p-value = 0.089
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-#Create Obese.num group
-edit.metadata2$obese.num[edit.metadata2$obese=="No"] <- 0
-edit.metadata2$obese.num[edit.metadata2$obese=="Yes"] <- 1
-obese <- factor(edit.metadata2$obese.num)
-
-#generate test set
-  # get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), his.microb.edit)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-#set.seed(3)
-#rossAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-# list of 3 Measures, AUCopt = 0.7526
-
-###########################################################################
-############ Z-score Data Preparation ###################################
-###########################################################################
-
-RossZH <- scale(H)
-RossZLogBF <- scale(log(BFratio))
-RossBMI <- bmi
-
-```
-
-```{r Goodrich_data, echo=FALSE, message=FALSE, warning=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/twinsUK")
-
-#Read in and match metadata to microbiome data
-metadata <- read.csv("TwinsUKStudy_metadata.csv")
-rownames(metadata) <- metadata[, 1]
-shared.data <- read.table("combined.tx.1.subsample.shared", header=T)
-rownames(shared.data) <- shared.data[, 2]
-microbiome <- shared.data[, -c(1:3)]
-keep  <- rownames(metadata)
-
-microbiome<- microbiome[keep, ]
-
-rm(good.metadata, shared.data, keep)
-
-#generate alpha diversity measures with vegan
-H <- diversity(microbiome)
-S <- specnumber(microbiome)
-J <- H/log(S)
-alpha.diversity <- as.data.frame(cbind(H, S, J))
-alpha.test <- alpha.diversity
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.table("phyla.txt", header=T)
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(microbiome)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- microbiome
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Chloroflexi", "Deferribacteres", "Elusimicrobia", "Fusobacteria", "Gemmatimonadetes", "Lentisphaerae", "OD1", "Planctomycetes", "Spirochaetes", "SR1", "Synergistetes", "Tenericutes")], 1, sum)
-phyla.table <- phyla.table[, -c(1, 4:6, 8:12, 14:17)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-
-#Create BMI groups
-metadata$BMI.class[metadata$body_mass_index_s<=24] <- "Normal"
-metadata$BMI.class[metadata$body_mass_index_s>24 & metadata$body_mass_index_s<30] <- "Overweight"
-metadata$BMI.class[metadata$body_mass_index_s>=30 & metadata$body_mass_index_s<40] <- "Obese"
-metadata$BMI.class[metadata$body_mass_index_s>=40] <- "Extreme Obesity"
-
-#Create Obese Yes/No groups
-metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
-metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
-
-#Create a column with obese and extreme obese as single entity
-metadata$BMIclass2[metadata$BMI.class=="Normal"] <- "Normal"
-metadata$BMIclass2[metadata$BMI.class=="Overweight"] <- "Overweight"                     
-metadata$BMIclass2[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Obese"
-
-#Get paitent demographics to be tested
-bmi <- metadata$body_mass_index_s
-obese <- factor(metadata$obese)
-
-###########################################################################
-######### First Level Analysis & Alpha Diversity with BMI #################
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-goodrichH <- anova(lm(H ~ obese)) #P-value=0.6671
-goodrichS <- anova(lm(S ~ obese)) #P-value=0.9073
-goodrichJ <- anova(lm(J ~ obese)) #P-value=0.6319
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-goodrichBacter <- anova(lm(bacter ~ obese)) #P-value=0.6661
-goodrichFirm <- anova(lm(firm ~ obese)) #P-value=0.9688
-goodrichBF <- anova(lm(BFratio ~ obese)) #P-value=0.8709
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-goodrich2 <- adonis(microbiome ~ obese, permutations=1000)
-goodrichPERM <- goodrich2$aov.tab
-#PERMANOVA=0.004995, pseudo-F=2.9308
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-orderedHCat <- test3[, 1]
-GoodtotalN <- length(orderedHCat)
-GoodHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
-GoodHighShannonGroup <- as.data.frame(table(test3[c(1:GoodHHighTotal), 2]))
-GoodLowShannonGroup <- as.data.frame(table(test3[c((GoodHHighTotal + 1):GoodtotalN), 2]))
-group1 <- c(GoodHighShannonGroup[2, 2], GoodHighShannonGroup[1, 2])
-group2 <- c(GoodLowShannonGroup[2, 2], GoodLowShannonGroup[1, 2])
-#Group1 (Higher than median), obese = 47 and non-obese = 206
-#Group2 (Lower than median), obese = 56 and non-obese = 198
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-goodrichHEpi <- epi.2by2(r.test, method="cohort.count")
-goodrichHMassoc <- goodrichHEpi$massoc
-goodrichHRR <- goodrichHMassoc$RR.strata.score
-goodrichHRRsig <- goodrichHMassoc$chisq.strata
-## Risk Ratio = 1.19
-## CI = 0.84, 1.68
-## p-value = 0.332
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-orderedBFCat <- test4[, 1]
-GoodBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
-GoodHighBFGroup <- as.data.frame(table(test4[c(1:GoodBFHighTotal), 2]))
-GoodLowBFGroup <- as.data.frame(table(test4[c((GoodBFHighTotal + 1):GoodtotalN), 2]))
-group1 <- c(GoodHighBFGroup[2, 2], GoodHighBFGroup[1, 2])
-group2 <- c(GoodLowBFGroup[2, 2], GoodLowBFGroup[1, 2])
-#Group1 (Higher than median), obese = 52 and non-obese = 201
-#Group2 (Lower than median), obese = 51 and non-obese = 203
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-goodrichBFEpi <- epi.2by2(r.test, method="cohort.count")
-goodrichBFMassoc <- goodrichBFEpi$massoc
-goodrichBFRR <- goodrichBFMassoc$RR.strata.score
-goodrichBFRRsig <- goodrichBFMassoc$chisq.strata
-## Risk Ratio = 0.98
-## CI = 0.69, 1.38
-## p-value = 0.894
-
-###########################################################################
-############ Classification using AUCRF####################################
-###########################################################################
-
-#Create Obese.num group
-metadata$obese.num[metadata$obese=="No"] <- 0
-metadata$obese.num[metadata$obese=="Yes"] <- 1
-obese <- factor(metadata$obese.num)
-
-#generate test set
-# get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), microbiome)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-#set.seed(3)
-#goodrichAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-# list of 8 Measures, AUCopt = 0.6767159
-
-###########################################################################
-############ Z-score Data Preparation ###################################
-###########################################################################
-
-GoodrichZH <- scale(H)
-GoodrichZLogBF <- scale(log(BFratio))
-GoodrichBMI <- bmi
-
-```
-
-```{r escobar_data, echo=FALSE, message=FALSE, warning=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/columbian/")
-
-columbian.microb <- read.table("EscobarPhylotypeSub.shared", header=T)
-metadata <- read.csv("columbian_dataset.csv")
-
-#Organize the microbiome shared data
-rownames(columbian.microb) <- columbian.microb[, 2]
-columbian.microb <- columbian.microb[, -c(1:3)]
-
-#Organize the metadata
-rownames(metadata) <- metadata[, 5]
-
-#Get only data that we are interested in
-edit.metadata <- metadata[, -c(1:7, 9, 13:15, 16:52)]
-
-#Sort metadata into the same order as the microbiome data
-order1 <- rownames(columbian.microb)
-edit.metadata2 <- edit.metadata[order1, ]
-
-#Get alpha diversity of the samples
-H <- diversity(columbian.microb)
-S <- specnumber(columbian.microb)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.table("taxonomyKey.txt")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(columbian.microb)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- columbian.microb
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(4:5, 7:9)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
-
-#Create Obese Yes/No groups
-edit.metadata2$obese[edit.metadata2$description_s=="Adequate weight male stool sample" | edit.metadata2$description_s=="Overweight male stool sample" | edit.metadata2$description_s=="Adequate weight female stool sample" | edit.metadata2$description_s=="Overweight female stool sample"] <- "No"
-edit.metadata2$obese[edit.metadata2$description_s=="Obese male stool sample" | edit.metadata2$description_s=="Obese female stool sample"] <- "Yes"
-
-#Create BMI groups
-edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s<=24] <- "Normal"
-edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
-edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=30 & edit.metadata2$body_mass_index_s<40] <- "Obese"
-edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=40] <- "Extreme Obesity"
-
-#Create BMI groups2
-edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s<=24] <- "Normal"
-edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
-edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>=30] <- "Obese"
-
-#Get paitent demographics data to be tested
-bmi <- edit.metadata2$body_mass_index_s
-obese <- factor(edit.metadata2$obese)
-
-######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-escobarH <- anova(lm(H ~ obese)) #P-value=0.8636
-escobarS <- anova(lm(S ~ obese)) #P-value=0.1947
-escobarJ <- anova(lm(J ~ obese)) #P-value=0.4762
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-escobarBacter <- anova(lm(bacter ~ obese)) #P-value=0.09253
-escobarFirm <- anova(lm(firm ~ obese)) #P-value=0.3602
-escobarBF <- anova(lm(BFratio ~ obese)) #P-value=0.4663
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-escobar2 <- adonis(columbian.microb ~ obese, permutations=1000)
-escobarPERM <- escobar2$aov.tab
-#PERMANOVA=0.08891, pseudo-F=1.3797
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-orderedHCat <- test3[, 1]
-EscototalN <- length(orderedHCat)
-EscoHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
-EscoHighShannonGroup <- as.data.frame(table(test3[c(1:EscoHHighTotal), 2]))
-EscoLowShannonGroup <- as.data.frame(table(test3[c((EscoHHighTotal + 1):EscototalN), 2]))
-group1 <- c(EscoHighShannonGroup[2, 2], EscoHighShannonGroup[1, 2])
-group2 <- c(EscoLowShannonGroup[2, 2], EscoLowShannonGroup[1, 2])
-#Group1 (Higher than median), obese = 4 and non-obese = 11
-#Group2 (Lower than median), obese = 6 and non-obese = 9
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group1", "group2")
-
-escobarHEpi <- epi.2by2(r.test, method="cohort.count")
-escobarHMassoc <- escobarHEpi$massoc
-escobarHRR <- escobarHMassoc$RR.strata.score
-escobarHRRsig <- escobarHMassoc$chisq.strata
-## Risk Ratio = 1.50
-## CI = 0.53, 4.26
-## p-value = 0.439
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-orderedBFCat <- test4[, 1]
-EscoBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
-EscoHighBFGroup <- as.data.frame(table(test4[c(1:EscoBFHighTotal), 2]))
-EscoLowBFGroup <- as.data.frame(table(test4[c((EscoBFHighTotal + 1):EscototalN), 2]))
-group1 <- c(EscoHighBFGroup[2, 2], EscoHighBFGroup[1, 2])
-group2 <- c(EscoLowBFGroup[2, 2], EscoLowBFGroup[1, 2])
-#Group1 (Higher than median), obese = 7 and non-obese = 8
-#Group2 (Lower than median), obese = 3 and non-obese = 12
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-escobarBFEpi <- epi.2by2(r.test, method="cohort.count")
-escobarBFMassoc <- escobarBFEpi$massoc
-escobarBFRR <- escobarBFMassoc$RR.strata.score
-escobarBFRRsig <- escobarBFMassoc$chisq.strata
-## Risk Ratio = 0.43
-## CI = 0.14, 1.35
-## p-value = 0.121
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-#Create Obese.num group
-edit.metadata2$obese.num[edit.metadata2$obese=="No"] <- 0
-edit.metadata2$obese.num[edit.metadata2$obese=="Yes"] <- 1
-obese <- factor(edit.metadata2$obese.num)
-
-#generate test set
-  # get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), columbian.microb)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-#set.seed(3)
-#escobarAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-# list of 3 Measures, AUCopt = 0.95
-
-###########################################################################
-############ Z-score Data Preparation ###################################
-###########################################################################
-
-EscobarZH <- scale(H)
-EscobarZLogBF <- scale(log(BFratio))
-EscobarBMI <- bmi
-
-```
-
-```{r Zupancic_data, echo=FALSE, message=FALSE, warning=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/Amish/updatedGOOD")
-
-#Read in and match metadata to microbiome data
-metadata <- read.csv("amish_obesity_table2.csv")
-metadata2 <- read.csv("amish.metadata.csv")
-test <- metadata[!duplicated(metadata$submitted_sample_id_s), ]
-rownames(test) <- test[, 4]
-test2 <- metadata2[!duplicated(metadata2$SUBJID), ]
-rownames(test2) <- test2[, 1]
-shared.data <- read.table("combined.0.03.subsample.shared", header=T)
-rownames(shared.data) <- shared.data[, 2]
-shared.data <- shared.data[, -c(1:3)]
-BactFamily <- read.csv("BacterialFamilyData.csv")
-rownames(BactFamily) <- BactFamily[, 1]
-BactFamily <- BactFamily[, -1]
-keep  <- rownames(shared.data)
-
-test3 <- test[keep, ]
-good.metadata <- test3
-test4 <- test2[keep, ]
-good.metadata2 <- test4
-
-family <- BactFamily[keep, ]
-microbiome <- shared.data[keep, ]
-metadata <- cbind(good.metadata, good.metadata2)
-
-rm(good.metadata, shared.data, keep, test, test2, test3, test4, BactFamily, good.metadata2, metadata2)
-
-test <- metadata[!duplicated(metadata$submitted_subject_id_s), ]
-keep  <- rownames(test)
-metadata <- test
-family <- family[keep, ]
-microbiome <- microbiome[keep, ]
-
-rm(test, keep)
-
-metadata <- metadata[complete.cases(metadata), ]
-keep <- rownames(metadata)
-family <- family[keep, ]
-microbiome <- microbiome[keep, ]
-
-rm(keep)
-
-#generate alpha diversity measures with vegan
-H <- diversity(microbiome)
-S <- specnumber(microbiome)
-J <- H/log(S)
-select.alpha.diversity <- as.data.frame(cbind(H, S, J))
-s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
-alpha.test <- s1.alpha.diversity
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(microbiome)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- microbiome
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Spirochaetes")], 1, sum)
-phyla.table <- phyla.table[, -c(4, 6)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:6)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-
-#Create BMI groups
-metadata$BMI.class[metadata$BMI<=24] <- "Normal"
-metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
-metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
-metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
-
-#Create Obese Yes/No groups
-metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
-metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
-
-#Create BMI class 2 groups
-metadata$BMI.class2[metadata$BMI<=24] <- "Normal"
-metadata$BMI.class2[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
-metadata$BMI.class2[metadata$BMI>=30] <- "Obese"
-
-#create groups to be used
-bmi <- metadata$BMI
-obese <- factor(metadata$obese)
-
-####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-zupancicH <- anova(lm(H ~ obese)) #P-value=0.2671
-zupancicS <- anova(lm(S ~ obese)) #P-value=0.05423
-zupancicJ <- anova(lm(J ~ obese)) #P-value=0.6229
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-zupancicBacter <- anova(lm(bacter ~ obese)) #P-value=0.8592
-zupancicFirm <- anova(lm(firm ~ obese)) #P-value=0.2564
-zupancicBF <- anova(lm(BFratio ~ obese)) #P-value=0.989
-
-####################################################################################### NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-zupancic2 <- adonis(microbiome ~ obese, permutations=1000)
-zupancicPERM <- zupancic2$aov.tab
-#PERMANOVA=0.6863, pseudo-F=0.92888
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-orderedHCat <- test3[, 1]
-ZupatotalN <- length(orderedHCat)
-ZupaHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
-ZupaHighShannonGroup <- as.data.frame(table(test3[c(1:ZupaHHighTotal), 2]))
-ZupaLowShannonGroup <- as.data.frame(table(test3[c((ZupaHHighTotal + 1):ZupatotalN), 2]))
-group1 <- c(ZupaHighShannonGroup[2, 2], ZupaHighShannonGroup[1, 2])
-group2 <- c(ZupaLowShannonGroup[2, 2], ZupaLowShannonGroup[1, 2])
-#Group1 (Higher than median), obese = 31 and non-obese = 72
-#Group2 (Lower than median), obese = 47 and non-obese = 57
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-zupancicHEpi <- epi.2by2(r.test, method="cohort.count")
-zupancicHMassoc <- zupancicHEpi$massoc
-zupancicHRR <- zupancicHMassoc$RR.strata.score
-zupancicHRRsig <- zupancicHMassoc$chisq.strata
-## Risk Ratio = 1.49
-## CI = 1.01, 2.19
-## p-value = 0.043
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-orderedBFCat <- test4[, 1]
-ZupaBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
-ZupaHighBFGroup <- as.data.frame(table(test4[c(1:ZupaBFHighTotal), 2]))
-ZupaLowBFGroup <- as.data.frame(table(test4[c((ZupaBFHighTotal + 1):ZupatotalN), 2]))
-group1 <- c(ZupaHighBFGroup[2, 2], ZupaHighBFGroup[1, 2])
-group2 <- c(ZupaLowBFGroup[2, 2], ZupaLowBFGroup[1, 2])
-#Group1 (Higher than median), obese = 41 and non-obese = 62
-#Group2 (Lower than median), obese = 37 and non-obese = 67
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-zupancicBFEpi <- epi.2by2(r.test, method="cohort.count")
-zupancicBFMassoc <- zupancicBFEpi$massoc
-zupancicBFRR <- zupancicBFMassoc$RR.strata.score
-zupancicBFRRsig <- zupancicBFMassoc$chisq.strata
-## Risk Ratio = 0.94
-## CI = 0.64, 1.36
-## p-value = 0.727
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-#Create Obese.num group
-metadata$obese.num[metadata$obese=="No"] <- 0
-metadata$obese.num[metadata$obese=="Yes"] <- 1
-obese <- factor(metadata$obese.num)
-
-#generate test set
-# get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), microbiome)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-#set.seed(3)
-#zupancicAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-#zupAUC <- fit$`OOB-AUCopt`
-# list of 40 Measures, AUCopt = 0.6596899
-
-###########################################################################
-############ Z-score Data Preparation ###################################
-###########################################################################
-
-ZupancicZH <- scale(H)
-ZupancicZLogBF <- scale(log(BFratio))
-ZupancicBMI <- bmi
-
-```
-
-```{r HMP_data, echo=FALSE, message=FALSE, warning=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-##Read in relevant data
-
-setwd("C:/users/marc/Desktop/obesity2/HMP.analysis")
-microbiome <- read.table("Stool.an.0.03.subsample.shared", header=T)
-meta.cat <- read.table("categorical.metadata", header=T)
-meta.cont <- read.table("continuous.metadata", header=T)
-
-#Only interested in first visit microbiome data
-#need to create microbiome data set for only that
-
-test <- microbiome[grep("\\.01\\.", microbiome$Group), ] #selecting by .01.
-
-microb.rownames <- gsub("([0-9]+).*", "\\1", test$Group) #extracting only numeric before first "."
-rownames(test) <- microb.rownames
-
-#Get rid of information not used in downstram analysis
-test <- test[, -c(1:3)]
-
-#Subset data to the microbiome total n
-select.meta.cat <- meta.cat[microb.rownames, ]
-select.meta.cont <- meta.cont[microb.rownames, ]
-
-#generate alpha diversity measures with vegan
-H <- diversity(test)
-S <- specnumber(test)
-J <- H/log(S)
-select.alpha.diversity <- as.data.frame(cbind(H, S, J))
-alpha.test <- select.alpha.diversity
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(test)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- test
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(1, 4, 6, 7, 9, 10:12)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
-
-#Create Obese Yes/No groups
-select.meta.cat$obese[select.meta.cat$BMI_C=="normal" | select.meta.cat$BMI_C=="overweight"] <- "No"
-select.meta.cat$obese[select.meta.cat$BMI_C=="obese" | select.meta.cat$BMI_C=="extreme obesity"] <- "Yes"
-
-#Create BMI groups2
-select.meta.cat$BMI.class2[select.meta.cont$DTPBMI<=24] <- "Normal"
-select.meta.cat$BMI.class2[select.meta.cont$DTPBMI>24 & select.meta.cont$DTPBMI<30] <- "Overweight"
-select.meta.cat$BMI.class2[select.meta.cont$DTPBMI>=30] <- "Obese"
-
-#Get paitent demographics data to be tested
-bmi <- select.meta.cont$DTPBMI
-obese <- factor(select.meta.cat$obese)
-
-######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-HMPH <- anova(lm(H ~ obese)) #P-value=0.585
-HMPS <- anova(lm(S ~ obese)) #P-value=0.845
-HMPJ <- anova(lm(J ~ obese)) #P-value=0.3559
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-HMPBacter <- anova(lm(bacter ~ obese)) #P-value=0.3518
-HMPFirm <- anova(lm(firm ~ obese)) #P-value=0.636
-HMPBF <- anova(lm(BFratio ~ obese)) #P-value=0.5688
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-HMP2 <- adonis(test ~ obese, permutations=1000)
-HMPPERM <- HMP2$aov.tab
-#PERMANOVA=0.8112, pseudo-F=0.7024
-
-######################################################################################## Relative Risk ###############################################
-###########################################################################
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-orderedHCat <- test3[, 1]
-HMPtotalN <- length(orderedHCat)
-HMPHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
-HMPHighShannonGroup <- as.data.frame(table(test3[c(1:HMPHHighTotal), 2]))
-HMPLowShannonGroup <- as.data.frame(table(test3[c((HMPHHighTotal + 1):HMPtotalN), 2]))
-group1 <- c(HMPHighShannonGroup[2, 2], HMPHighShannonGroup[1, 2])
-group2 <- c(HMPLowShannonGroup[2, 2], HMPLowShannonGroup[1, 2])
-#Group1 (Higher than median), obese = 10 and non-obese = 118
-#Group2 (Lower than median), obese = 16 and non-obese = 112
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group1", "group2")
-
-HMPHEpi <- epi.2by2(r.test, method="cohort.count")
-HMPHMassoc <- HMPHEpi$massoc
-HMPHRR <- HMPHMassoc$RR.strata.score
-HMPHRRsig <- HMPHMassoc$chisq.strata
-## Risk Ratio = 1.60
-## CI = 0.75, 3.39
-## p-value = 0.214
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-orderedBFCat <- test4[, 1]
-HMPBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
-HMPHighBFGroup <- as.data.frame(table(test4[c(1:HMPBFHighTotal), 2]))
-HMPLowBFGroup <- as.data.frame(table(test4[c((HMPBFHighTotal + 1):HMPtotalN), 2]))
-group1 <- c(HMPHighBFGroup[2, 2], HMPHighBFGroup[1, 2])
-group2 <- c(HMPLowBFGroup[2, 2], HMPLowBFGroup[1, 2])
-#Group1 (Higher than median), obese = 13 and non-obese = 115
-#Group2 (Lower than median), obese = 13 and non-obese = 115
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-HMPBFEpi <- epi.2by2(r.test, method="cohort.count")
-HMPBFMassoc <- HMPBFEpi$massoc
-HMPBFRR <- HMPBFMassoc$RR.strata.score
-HMPBFRRsig <- HMPBFMassoc$chisq.strata
-## Risk Ratio = 1.00
-## CI = 0.48, 2.07
-## p-value = 1.00
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-#Create Obese.num group
-select.meta.cat$obese.num[select.meta.cat$obese=="No"] <- 0
-select.meta.cat$obese.num[select.meta.cat$obese=="Yes"] <- 1
-obese <- factor(select.meta.cat$obese.num)
-
-#generate test set
-  # get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), test)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-#set.seed(3)
-#HMPAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-# list of 6 Measures, AUCopt = 0.7031773
-
-###########################################################################
-############ Z-score Data Preparation ###################################
-###########################################################################
-
-HMPZH <- scale(H)
-HMPZLogBF <- scale(log(BFratio))
-HMPBMI <- bmi
-
-```
-
-```{r Wu_data, echo=FALSE, message=FALSE, warning=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/COMBO/")
-
-microbiome <- read.table("combined.good.subsample.shared", header=T)
-rownames(microbiome) <- microbiome$Group
-microbiome <- microbiome[, -c(1:3)]
-seqData <- read.csv("COMBO_data_table.csv", header=T)
-seqData <- seqData[-1, ]
-rownames(seqData) <- seqData$Run_s
-metadata <- read.table("BMI_and_Counts.txt", header=T)
-metadata <- metadata[, c(1, 2)]
-
-#get seqData set in line with microbiome data
-namesToKeep <- rownames(microbiome)
-test <- seqData[namesToKeep, ]
-seqData <- test
-rm(test)
-
-#change all rownames to match sample IDs
-rownames(seqData) <- seqData$submitted_subject_id_s
-rownames(microbiome) <- rownames(seqData)
-rownames(metadata) <- metadata[, 1]
-
-#Match the metadata now with the microbiome data
-namesToKeep <- rownames(microbiome)
-test <- metadata[namesToKeep, ]
-metadata <- test
-rm(test)
-rm(seqData, keep, namesToKeep)
-
-#Get alpha diversity of the samples
-H <- diversity(microbiome)
-S <- specnumber(microbiome)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.table("combined.good.phyla.txt", header=T)
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(microbiome)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- microbiome
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Synergistetes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(4, 6:7)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:6)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-rm(keep, phyla.names, phyla.total, phylogenetic.info, phyla.table, phyla.good)
-
-#Create BMI groups
-metadata$BMI.class[metadata$BMI<=24] <- "Normal"
-metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
-metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
-metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
-
-#Create Obese Yes/No groups
-metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
-metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
-
-#Create a column with obese and extreme obese as single entity
-metadata$BMIclass2[metadata$BMI.class=="Normal"] <- "Normal"
-metadata$BMIclass2[metadata$BMI.class=="Overweight"] <- "Overweight"                     
-metadata$BMIclass2[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Obese"
-
-#Get paitent demographics to be tested
-bmi <- metadata$BMI
-obese <- factor(metadata$obese)
-
-###########################################################################
-######### First Level Analysis & Alpha Diversity with BMI #################
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-WuH <- anova(lm(H ~ obese)) #P-value=0.7493
-WuS <- anova(lm(S ~ obese)) #P-value=0.7511
-WuJ <- anova(lm(J ~ obese)) #P-value=0.5557
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-WuBacter <- anova(lm(bacter ~ obese)) #P-value=0.9173
-WuFirm <- anova(lm(firm ~ obese)) #P-value=0.5765
-WuBF <- anova(lm(BFratio ~ obese)) #P-value=0.6735
-
-###########################################################################
-############ NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-Wu2 <- adonis(microbiome ~ obese, permutations=1000)
-WuPERM <- Wu2$aov.tab
-#PERMANOVA=0.8921, pseudo-F=0.66065
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-orderedHCat <- test3[, 1]
-WutotalN <- length(orderedHCat)
-WuHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
-WuHighShannonGroup <- as.data.frame(table(test3[c(1:WuHHighTotal), 2]))
-WuLowShannonGroup <- as.data.frame(table(test3[c((WuHHighTotal + 1):WutotalN), 2]))
-group1 <- c(WuHighShannonGroup[2, 2], WuHighShannonGroup[1, 2])
-group2 <- c(WuLowShannonGroup[2, 2], WuLowShannonGroup[1, 2])
-#Group1 (Higher than median), obese = 3 and non-obese = 26
-#Group2 (Lower than median), obese = 2 and non-obese = 27
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-WuHEpi <- epi.2by2(r.test, method="cohort.count")
-WuHMassoc <- WuHEpi$massoc
-WuHRR <- WuHMassoc$RR.strata.score
-WuHRRsig <- WuHMassoc$chisq.strata
-## Risk Ratio = 0.67
-## CI = 0.12, 3.70
-## p-value = 0.64
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-orderedBFCat <- test4[, 1]
-WuBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
-WuHighBFGroup <- as.data.frame(table(test4[c(1:WuBFHighTotal), 2]))
-WuLowBFGroup <- as.data.frame(table(test4[c((WuBFHighTotal + 1):WutotalN), 2]))
-group1 <- c(WuHighBFGroup[2, 2], WuHighBFGroup[1, 2])
-group2 <- c(WuLowBFGroup[2, 2], WuLowBFGroup[1, 2])
-#Group1 (Higher than median), obese = 3 and non-obese = 26
-#Group2 (Lower than median), obese = 2 and non-obese = 27
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-WuBFEpi <- epi.2by2(r.test, method="cohort.count")
-WuBFMassoc <- WuBFEpi$massoc
-WuBFRR <- WuBFMassoc$RR.strata.score
-WuBFRRsig <- WuBFMassoc$chisq.strata
-## Risk Ratio = 0.67
-## CI = 0.12, 3.70
-## p-value = 0.64
-
-###########################################################################
-############ Classification using AUCRF####################################
-###########################################################################
-
-#Create Obese.num group
-metadata$obese.num[metadata$obese=="No"] <- 0
-metadata$obese.num[metadata$obese=="Yes"] <- 1
-obese <- factor(metadata$obese.num)
-
-#generate test set
-# get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), microbiome)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-#set.seed(3)
-#WuAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-# list of 15 Measures, AUCopt = 0.7075472
-
-###########################################################################
-############ Z-score Data Preparation ###################################
-###########################################################################
-
-WuZH <- scale(H)
-WuZLogBF <- scale(log(BFratio))
-WuBMI <- bmi
-
-```
-
-```{r Arumugam_data, echo=FALSE, message=FALSE, warning=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/MetaHit/")
-
-microb <- read.csv("finalized_metaHit_shared.csv")
-rownames(microb) <- microb[, 1]
-microb <- microb[,-1]
-
-metadata <- read.csv("MetaHit_metadata.csv")
-rownames(metadata) <- metadata[, 1]
-metadata <- metadata[, -1]
-
-phyla <- read.csv("phyla.csv")
-rownames(phyla) <- phyla[, 1]
-phyla <- phyla[, -1]
-
-# Generate Overall Relative Abundance - normalize to a value of 100
-# Needs to be done since not all measures have 100 bacteria from metaphlan2
-overall <- rowSums(microb)
-microb.norm <- (microb / overall) * 100
-
-phyla.table <- read.csv("phyla.csv", header = T)
-rownames(phyla.table) <- phyla.table[, 1]
-phyla.table <- phyla.table[, -1]
-phyla.table <- as.data.frame(t(phyla.table))
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:10)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-rm(phyla.table, phyla.info)
-
-#Get alpha diversity of the samples
-H <- diversity(microb.norm)
-S <- specnumber(microb.norm)
-J <- H/log(S)
-alpha.diversity.shannon <- cbind(H,S,J)
-alpha.test <- as.data.frame(alpha.diversity.shannon)
-# Seems to look okay....
-
-#Create BMI groups
-metadata$BMI.class[metadata$BMI<=24] <- "Normal"
-metadata$BMI.class[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"
-metadata$BMI.class[metadata$BMI>=30 & metadata$BMI<40] <- "Obese"
-metadata$BMI.class[metadata$BMI>=40] <- "Extreme Obesity"
-
-#Create Obese Yes/No groups
-metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
-metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"
-
-#Create Obese.num group
-metadata$obese.num[metadata$obese=="No"] <- 0
-metadata$obese.num[metadata$obese=="Yes"] <- 1
-
-#Create a column with obese and extreme obese as single entity
-metadata$BMIclass2[metadata$BMI<=24] <- "Normal"
-metadata$BMIclass2[metadata$BMI>24 & metadata$BMI<30] <- "Overweight"      
-metadata$BMIclass2[metadata$BMI>=30] <- "Obese"
-
-#Get paitent demographics to be tested
-bmi <- metadata$BMI
-obese <- factor(metadata$obese)
-
-####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-ArumugamH <- anova(lm(H ~ obese)) #P-value=0.2174
-ArumugamS <- anova(lm(S ~ obese)) #P-value=0.9878
-ArumugamJ <- anova(lm(J ~ obese)) #P-value=0.1874
-
-#B and F tests against obesity
-bacter <- phyla.table.rel.abund$Bacteroidetes
-firm <- phyla.table.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-ArumugamBacter <- anova(lm(bacter ~ obese)) #P-value=0.1911
-ArumugamFirm <- anova(lm(firm ~ obese)) #P-value=0.2925
-ArumugamBF <- anova(lm(BFratio ~ obese)) #P-value=0.05254
-
-####################################################################################### NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-Arumugam2 <- adonis(microb.norm ~ obese, permutations=1000)
-ArumugamPERM <- Arumugam2$aov.tab
-#PERMANOVA=0.05495, pseudo-F=1.6047
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-orderedHCat <- test3[, 1]
-ArutotalN <- length(orderedHCat)
-AruHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
-AruHighShannonGroup <- as.data.frame(table(test3[c(1:AruHHighTotal), 2]))
-AruLowShannonGroup <- as.data.frame(table(test3[c((AruHHighTotal + 1):ArutotalN), 2]))
-group1 <- c(AruHighShannonGroup[2, 2], AruHighShannonGroup[1, 2])
-group2 <- c(AruLowShannonGroup[2, 2], AruLowShannonGroup[1, 2])
-#Group1 (Higher than median), obese = 17 and non-obese = 25
-#Group2 (Lower than median), obese = 20 and non-obese = 23
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-ArumugamHEpi <- epi.2by2(r.test, method="cohort.count")
-ArumugamHMassoc <- ArumugamHEpi$massoc
-ArumugamHRR <- ArumugamHMassoc$RR.strata.score
-ArumugamHRRsig <- ArumugamHMassoc$chisq.strata
-## Risk Ratio = 1.15
-## CI = 0.71, 1.87
-## p-value = 0.575
-
-##Run the RR for B/F ratio
-Bacter = phyla.table.rel.abund$Bacteroidetes
-Firm = phyla.table.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-orderedBFCat <- test4[, 1]
-AruBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
-AruHighBFGroup <- as.data.frame(table(test4[c(1:AruBFHighTotal), 2]))
-AruLowBFGroup <- as.data.frame(table(test4[c((AruBFHighTotal + 1):ArutotalN), 2]))
-group1 <- c(AruHighBFGroup[2, 2], AruHighBFGroup[1, 2])
-group2 <- c(AruLowBFGroup[2, 2], AruLowBFGroup[1, 2])
-#Group1 (Higher than median), obese = 20 and non-obese = 22
-#Group2 (Lower than median), obese = 17 and non-obese = 26
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-ArumugamBFEpi <- epi.2by2(r.test, method="cohort.count")
-ArumugamBFMassoc <- ArumugamBFEpi$massoc
-ArumugamBFRR <- ArumugamBFMassoc$RR.strata.score
-ArumugamBFRRsig <- ArumugamBFMassoc$chisq.strata
-## Risk Ratio = 0.83
-## CI = 0.51, 1.35
-## p-value = 0.452
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-#Create Obese.num group
-metadata$obese.num[metadata$obese=="No"] <- 0
-metadata$obese.num[metadata$obese=="Yes"] <- 1
-obese <- factor(metadata$obese.num)
-
-#generate test set
-# get rid of those with 0 and only 4 other values
-testset <- Filter(function(x)(length(unique(x))>5), microb.norm)
-testset <- cbind(obese, testset)
-colnames(testset)[1] <- "obese"
-testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-#set.seed(3)
-#ArumugamAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-# list of 19 Measures, AUCopt = 0.7649212
-
-###########################################################################
-############ Z-score Data Preparation ###################################
-###########################################################################
-
-ArumugamZH <- scale(H)
-ArumugamZLogBF <- scale(log(BFratio))
-ArumugamBMI <- bmi
-
-```
-
-```{r Turnbaugh_data, echo=FALSE, message=FALSE, warning=FALSE}
-###########################################################################
-############ Preparing Data Tables for Analysis ###########################
-###########################################################################
-
-setwd("C:/users/marc/Desktop/obesity2/turnbaugh.twins")
-shared.data <- read.table("test.unique.good.filter.unique.precluster.pick.pick.an.shared", header=T)
-rownames(shared.data) <- shared.data[, 2]
-shared.data <- shared.data[, -c(1:3)]
-
-subsample.data <- read.table("test.unique.good.filter.unique.precluster.pick.pick.an.0.03.subsample.shared", header=T)
-rownames(subsample.data) <- subsample.data[, 2]
-subsample.data <- subsample.data[, -c(1:3)]
-
-metadata <- read.csv("turnbaugh.metadata.csv")
-rownames(metadata) <- metadata[, 4]
-metadata <- metadata[, -4]
-
-keep1 <- which(metadata$Sample == 1) # n =146
-
-#use the first sampling
-#subset data for only the first sampling
-s1.metadata <- metadata[keep1, ]
-s1.subsample.data <- subsample.data[keep1, ]
-
-#generate alpha diversity measures with vegan
-H <- diversity(s1.subsample.data)
-S <- specnumber(s1.subsample.data)
-J <- H/log(S)
-select.alpha.diversity <- as.data.frame(cbind(H, S, J))
-s1.alpha.diversity <- as.data.frame(select.alpha.diversity)
-alpha.test <- s1.alpha.diversity
-
-#Generate phyla table data
-
-phyla.info <- read.csv("phyla.csv")
-phyla.table <- shared.data
-colnames(phyla.table) <- phyla.info[, 2]
-
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing)
-
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(4:5, 7:9)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-s1.phyla.rel.abund <- phyla.table.rel.abund[rownames(shared.data), ]
-s1.phyla.rel.abund <- s1.phyla.rel.abund[keep1, ]
-
-#Create Obese Yes/No groups
-s1.metadata$obese[s1.metadata$BMI.category=="Lean" | s1.metadata$BMI.category=="Overweight"] <- "No"
-s1.metadata$obese[s1.metadata$BMI.category=="Obese"] <- "Yes"
-
-#Create Obese.num group
-s1.metadata$obese.num[s1.metadata$obese=="No"] <- 0
-s1.metadata$obese.num[s1.metadata$obese=="Yes"] <- 1
-
-#create groups to be used
-obese <- factor(s1.metadata$obese)
-bmi <- s1.metadata$BMI.category
-
-####################################################################################### First Level Analysis & Alpha Diversity with BMI #############
-###########################################################################
-###########################################################################
-
-##Test BMI versus alpha diversity and phyla
-
-turnbaughH <- anova(lm(H ~ obese)) #P-value=0.06976
-turnbaughS <- anova(lm(S ~ obese)) #P-value=0.02435
-turnbaughJ <- anova(lm(J ~ obese)) #P-value=0.1253
-
-#B and F tests against obesity
-bacter <- s1.phyla.rel.abund$Bacteroidetes
-firm <- s1.phyla.rel.abund$Firmicutes
-BFratio <- bacter/firm
-
-turnbaughBacter <- anova(lm(bacter ~ obese)) #P-value=0.5771
-turnbaughFirm <- anova(lm(firm ~ obese)) #P-value=0.4073
-turnbaughBF <- anova(lm(BFratio ~ obese)) #P-value=0.2216
-
-####################################################################################### NMDS and PERMANOVA Analysis###################################
-###########################################################################
-
-set.seed(3)
-turnbaugh2 <- adonis(s1.subsample.data ~ obese, permutations=1000)
-turnbaughPERM <- turnbaugh2$aov.tab
-#PERMANOVA=0.09491, pseudo-F=1.2114
-
-###########################################################################
-############ Relative Risk#################################################
-###########################################################################
-
-#Generate median values and put them into existing alpha.test dataframe
-#Shannon diversity
-
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-##Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-orderedHCat <- test3[, 1]
-TurntotalN <- length(orderedHCat)
-TurnHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
-TurnHighShannonGroup <- as.data.frame(table(test3[c(1:TurnHHighTotal), 2]))
-TurnLowShannonGroup <- as.data.frame(table(test3[c((TurnHHighTotal + 1):TurntotalN), 2]))
-group1 <- c(TurnHighShannonGroup[2, 2], TurnHighShannonGroup[1, 2])
-group2 <- c(TurnLowShannonGroup[2, 2], TurnLowShannonGroup[1, 2])
-#Group1 (Higher than median), obese = 47 and non-obese = 26
-#Group2 (Lower than median), obese = 52 and non-obese = 21
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-turnbaughHEpi <- epi.2by2(r.test, method="cohort.count")
-turnbaughHMassoc <- turnbaughHEpi$massoc
-turnbaughHRR <- turnbaughHMassoc$RR.strata.score
-turnbaughHRRsig <- turnbaughHMassoc$chisq.strata
-## Risk Ratio = 1.11
-## CI = 0.88, 1.38
-## p-value = 0.376
-
-##Run the RR for B/F ratio
-Bacter = s1.phyla.rel.abund$Bacteroidetes
-Firm = s1.phyla.rel.abund$Firmicutes
-BFRatio = Bacter/Firm
-BFRatio <- as.data.frame(BFRatio)
-
-BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})
-
-BFRatio.cat <- BFRatio$BFRatio.cat
-test4 <- cbind(BFRatio.cat, obese)
-test4 <- test4[order(BFRatio.cat), ]
-orderedBFCat <- test4[, 1]
-TurnBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
-TurnHighBFGroup <- as.data.frame(table(test4[c(1:TurnBFHighTotal), 2]))
-TurnLowBFGroup <- as.data.frame(table(test4[c((TurnBFHighTotal + 1):TurntotalN), 2]))
-group1 <- c(TurnHighBFGroup[2, 2], TurnHighBFGroup[1, 2])
-group2 <- c(TurnLowBFGroup[2, 2], TurnLowBFGroup[1, 2])
-#Group1 (Higher than median), obese = 54 and non-obese = 19
-#Group2 (Lower than median), obese = 45 and non-obese = 28
-
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-turnbaughBFEpi <- epi.2by2(r.test, method="cohort.count")
-turnbaughBFMassoc <- turnbaughBFEpi$massoc
-turnbaughBFRR <- turnbaughBFMassoc$RR.strata.score
-turnbaughBFRRsig <- turnbaughBFMassoc$chisq.strata
-## Risk Ratio = 0.83
-## CI = 0.66, 1.05
-## p-value = 0.111
-
-###########################################################################
-############ Classification using AUCRF ###################################
-###########################################################################
-
-#Create Obese.num group
-#s1.metadata$obese.num[s1.metadata$obese=="No"] <- 0
-#s1.metadata$obese.num[s1.metadata$obese=="Yes"] <- 1
-#obese <- factor(s1.metadata$obese.num)
-
-#generate test set
-# get rid of those with 0 and only 4 other values
-#testset <- Filter(function(x)(length(unique(x))>5), s1.subsample.data)
-#testset <- cbind(obese, testset)
-#colnames(testset)[1] <- "obese"
-#testset <- cbind(testset, H, S, J, s1.phyla.rel.abund)
-
-#Try AUCRF with default measures provided in readme
-#set.seed(3)
-#TurnbaughAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
-# list of 11 Measures, AUCopt = 0.7764883
-
-###########################################################################
-############ Z-score Data Preparation ###################################
-###########################################################################
-
-TurnbaughZH <- scale(H)
-TurnbaughZLogBF <- scale(log(BFratio))
-TurnbaughObese <- as.character(obese)
-TurnbaughBMICat <- as.character(bmi)
-
-```
-
 ```{r Table_3, echo=FALSE, message=FALSE, warning=FALSE}
 
-Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "HMP", "Wu", "Arumugam", "Turnbaugh")
-
-Bacteroidetes <- signif(c(baxterBacter[1,5], rossBacter[1,5], goodrichBacter[1,5], escobarBacter[1,5], zupancicBacter[1,5], HMPBacter[1,5], WuBacter[1,5], ArumugamBacter[1,5], turnbaughBacter[1,5]), 3)
-
-Firmicutes <- signif(c(baxterFirm[1,5], rossFirm[1,5], goodrichFirm[1,5], escobarFirm[1,5], zupancicFirm[1,5], HMPFirm[1,5], WuFirm[1,5], ArumugamFirm[1,5], turnbaughFirm[1,5]), 3)
-
-BFRatio <- signif(c(baxterBF[1,5], rossBF[1,5], goodrichBF[1,5], escobarBF[1,5], zupancicBF[1,5], HMPBF[1,5], WuBF[1,5], ArumugamBF[1,5], turnbaughBF[1,5]), 3)
-
-Shannon <- signif(c(baxterH[1,5], rossH[1,5], goodrichH[1,5], escobarH[1,5], zupancicH[1,5], HMPH[1,5], WuH[1,5], ArumugamH[1,5], turnbaughH[1,5]), 3)
-
-OTURich <- signif(c(baxterS[1,5], rossS[1,5], goodrichS[1,5], escobarS[1,5], zupancicS[1,5], HMPS[1,5], WuS[1,5], ArumugamS[1,5], turnbaughS[1,5]),3)
-
-Evenness <- signif(c(baxterJ[1,5], rossJ[1,5], goodrichJ[1,5], escobarJ[1,5], zupancicJ[1,5], HMPJ[1,5], WuJ[1,5], ArumugamJ[1,5], turnbaughJ[1,5]), 3)
-
-BrayC <- signif(c(baxterPERM[1,6], rossPERM[1,6], goodrichPERM[1,6], escobarPERM[1,6], zupancicPERM[1,6], HMPPERM[1,6], WuPERM[1,6], ArumugamPERM[1,6], turnbaughPERM[1,6]), 3)
+Study <- as.character(OverallPValue$X)
+Bacteroidetes <- format(round(OverallPValue$Bacteroidetes, 3), nsmall=3)
+Firmicutes <- format(round(OverallPValue$Firmicutes, 3), nsmall=3)
+BFRatio <- format(round(OverallPValue$BFRatio, 3), nsmall=3)
+Shannon <- format(round(OverallPValue$Shannon, 3), nsmall=3)
+OTURich <- format(round(OverallPValue$OTURich, 3), nsmall=3)
+Evenness <- format(round(OverallPValue$Evenness, 3), nsmall=3)
+BrayC <- format(round(OverallPValue$BrayC, 3), nsmall=3)
 
 table3 <- cbind(Study, Bacteroidetes, Firmicutes, BFRatio, Shannon, OTURich, Evenness, BrayC)
-
 colnames(table3) <- c("Study", "Bacteroidetes", "Firmicutes", "B/F Ratio", "Shannon Diversity", "OTU Richness", "Evenness", "Bray Curtis")
@@
 
-StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
-Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu", "Goodrich")
-Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012", "2014")
-Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=199", "n=58", "n=507")
-tpos <- c(BaxLowShannonGroup[2, 2], HMPLowShannonGroup[2, 2], RossLowShannonGroup[2, 2], EscoLowShannonGroup[2, 2], AruLowShannonGroup[2, 2], TurnLowShannonGroup[2, 2], ZupaLowShannonGroup[2, 2], WuLowShannonGroup[2, 2], GoodLowShannonGroup[2, 2]) #obese in low diversity group
-tneg <- c(BaxLowShannonGroup[1, 2], HMPLowShannonGroup[1, 2], RossLowShannonGroup[1, 2], EscoLowShannonGroup[1, 2], AruLowShannonGroup[1, 2], TurnLowShannonGroup[1, 2], ZupaLowShannonGroup[1, 2], WuLowShannonGroup[1, 2], GoodLowShannonGroup[1, 2]) #nonobese in low diversity group
-cpos <- c(BaxHighShannonGroup[2, 2], HMPHighShannonGroup[2, 2], RossHighShannonGroup[2, 2], EscoHighShannonGroup[2, 2], AruHighShannonGroup[2, 2], TurnHighShannonGroup[2, 2], ZupaHighShannonGroup[2, 2], WuHighShannonGroup[2, 2], GoodHighShannonGroup[2, 2]) #obese in high diversity group
-cneg <- c(BaxHighShannonGroup[1, 2], HMPHighShannonGroup[1, 2], RossHighShannonGroup[1, 2], EscoHighShannonGroup[1, 2], AruHighShannonGroup[1, 2], TurnHighShannonGroup[1, 2], ZupaHighShannonGroup[1, 2], WuHighShannonGroup[1, 2], GoodHighShannonGroup[1, 2]) #nonobese in high diversity group
-RR <- signif(c(baxterHRR[1,1], HMPHRR[1,1], rossHRR[1,1], escobarHRR[1,1], ArumugamHRR[1,1], turnbaughHRR[1,1], zupancicHRR[1,1], WuHRR[1,1], goodrichHRR[1,1]), 2)
-low <- signif(c(baxterHRR[1,2], HMPHRR[1,2], rossHRR[1,2], escobarHRR[1,2], ArumugamHRR[1,2], turnbaughHRR[1,2], zupancicHRR[1,2], WuHRR[1,2], goodrichHRR[1,2]), 2)
-high <- signif(c(baxterHRR[1,3], HMPHRR[1,3], rossHRR[1,3], escobarHRR[1,3], ArumugamHRR[1,3], turnbaughHRR[1,3], zupancicHRR[1,3], WuHRR[1,3], goodrichHRR[1,3]), 2)
+StudyNo <- as.numeric(ShannonRRTable$X)
+Study <- as.character(ShannonRRTable$Study)
+Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2010", "2009")
+Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63","n=85", "n=146")
+tpos <- as.numeric(ShannonRRTable$tposH) #obese in low diversity group
+tneg <- as.numeric(ShannonRRTable$tnegH) #nonobese in low diversity group
+cpos <- as.numeric(ShannonRRTable$cposH) #obese in high diversity group
+cneg <- as.numeric(ShannonRRTable$cnegH) #nonobese in high diversity group
+RR <- as.numeric(ShannonRRTable$RRH)
+low <- as.numeric(ShannonRRTable$lowH)
+high <- as.numeric(ShannonRRTable$highH)
 
@@
 **Figure 2: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity.**  Groups were divided for each study on high and low Shannon Diversity groups based on the median for that study.
-The overall pooled relative risk was 1.18 for the low diversity group (P-value = `r signif(RRSDPvalue$QMp, 3)`).
+The overall pooled relative risk was 1.17 for the low diversity group (P-value = `r signif(RRSDPvalue$QMp, 3)`).
 
@@
 
-StudyNo <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
-Study <- c("Baxter", "HMP", "Ross", "Escobar", "Arumugam", "Turnbaugh", "Zupancic", "Wu", "Goodrich")
-Year <- c("2015", "2011", "2015", "2014", "2010", "2009", "2012", "2012", "2014")
-Total <- c("n=172", "n=256", "n=63", "n=30", "n=85", "n=146", "n=199", "n=58", "n=507")
-tpos <- c(BaxLowBFGroup[2, 2], HMPLowBFGroup[2, 2], RossLowBFGroup[2, 2], EscoLowBFGroup[2, 2], AruLowBFGroup[2, 2], TurnLowBFGroup[2, 2], ZupaLowBFGroup[2, 2], WuLowBFGroup[2, 2], GoodLowBFGroup[2, 2]) #obese in low BF group
-tneg <- c(BaxLowBFGroup[1, 2], HMPLowBFGroup[1, 2], RossLowBFGroup[1, 2], EscoLowBFGroup[1, 2], AruLowBFGroup[1, 2], TurnLowBFGroup[1, 2], ZupaLowBFGroup[1, 2], WuLowBFGroup[1, 2], GoodLowBFGroup[1, 2]) #nonobese in low BF group
-cpos <- c(BaxHighBFGroup[2, 2], HMPHighBFGroup[2, 2], RossHighBFGroup[2, 2], EscoHighBFGroup[2, 2], AruHighBFGroup[2, 2], TurnHighBFGroup[2, 2], ZupaHighBFGroup[2, 2], WuHighBFGroup[2, 2], GoodHighBFGroup[2, 2]) #obese in high BF group
-cneg <- c(BaxHighBFGroup[1, 2], HMPHighBFGroup[1, 2], RossHighBFGroup[1, 2], EscoHighBFGroup[1, 2], AruHighBFGroup[1, 2], TurnHighBFGroup[1, 2], ZupaHighBFGroup[1, 2], WuHighBFGroup[1, 2], GoodHighBFGroup[1, 2]) #nonobese in high BF group
-RR <- signif(c(baxterBFRR[1,1], HMPBFRR[1,1], rossBFRR[1,1], escobarBFRR[1,1], ArumugamBFRR[1,1], turnbaughBFRR[1,1], zupancicBFRR[1,1], WuBFRR[1,1], goodrichBFRR[1,1]), 2)
-low <- signif(c(baxterBFRR[1,2], HMPBFRR[1,2], rossBFRR[1,2], escobarBFRR[1,2], ArumugamBFRR[1,2], turnbaughBFRR[1,2], zupancicBFRR[1,2], WuBFRR[1,2], goodrichBFRR[1,2]), 2)
-high <- signif(c(baxterBFRR[1,3], HMPBFRR[1,3], rossBFRR[1,3], escobarBFRR[1,3], ArumugamBFRR[1,3], turnbaughBFRR[1,3], zupancicBFRR[1,3], WuBFRR[1,3], goodrichBFRR[1,3]), 2)
+StudyNo <- as.numeric(BFRatioRRTable$X)
+Study <- as.character(BFRatioRRTable$Study)
+Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2010", "2009")
+Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63","n=85", "n=146")
+tpos <- as.numeric(BFRatioRRTable$tposBF) #obese in low diversity group
+tneg <- as.numeric(BFRatioRRTable$tnegBF) #nonobese in low diversity group
+cpos <- as.numeric(BFRatioRRTable$cposBF) #obese in high diversity group
+cneg <- as.numeric(BFRatioRRTable$cnegBF) #nonobese in high diversity group
+RR <- as.numeric(BFRatioRRTable$RRBF)
+low <- as.numeric(BFRatioRRTable$lowBF)
+high <- as.numeric(BFRatioRRTable$highBF)
 
@@
 **Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low B/F ratio groups based on the median for that specific study.
-The overall pooled relative risk was 0.87 for the low diversity group (P-value = `r signif(RRBFPvalue$QMp, 3)`).
+The overall pooled relative risk was 0.90 for the low diversity group (P-value = `r signif(RRBFPvalue$QMp, 3)`).
 
@@
 
-setwd("C:/Users/marc/Desktop/obesity2/BMI.normalization")
-
 #data1 order Escobar, Baxter, Ross, HMP, Arumugam, Wu, Goodrich, Zupancic
 
-ZH <- c(EscobarZH, BaxterZH, RossZH, HMPZH, ArumugamZH, WuZH, GoodrichZH, ZupancicZH)
-bmi <- c(EscobarBMI, BaxterBMI, RossBMI, HMPBMI, ArumugamBMI, WuBMI, GoodrichBMI, ZupancicBMI)
-ZlogBF <- c(EscobarZLogBF, BaxterZLogBF, RossZLogBF, HMPZLogBF, ArumugamZLogBF, WuZLogBF, GoodrichZLogBF, ZupancicZLogBF)
+ZH <- as.numeric(ZScoreCombinedData$ZH)
+bmi <- as.numeric(ZScoreCombinedData$BMI)
+BMI.class <- as.character(ZScoreCombinedData$BMICat)
+BMI.class[BMI.class == "Normal"] <- "Lean"
+obese <- as.character(ZScoreCombinedData$Obese)
+ZlogBF <- as.numeric(ZScoreCombinedData$ZLogBF)
 
 data1 <- as.data.frame(cbind(ZH, ZlogBF, bmi))
-
-turn.comb.meta <- as.data.frame(cbind(TurnbaughBMICat, TurnbaughObese))
-turn.comb.cont <- as.data.frame(cbind(TurnbaughZH, TurnbaughZLogBF))
-turn.data <- cbind(turn.comb.cont, turn.comb.meta)
-colnames(turn.data) <- c("ZH", "ZLogBF", "BMI.class", "obese")
-
-#Create BMI groups
-data1$BMI.class[data1$bmi<=24] <- "Lean"
-data1$BMI.class[data1$bmi>24 & data1$bmi<30] <- "Overweight"
-data1$BMI.class[data1$bmi>=30] <- "Obese"
-
-#Create Obese Yes/No groups
-data1$obese[data1$BMI.class=="Lean" | data1$BMI.class=="Overweight"] <- "No"
-data1$obese[data1$BMI.class=="Obese"] <- "Yes"
 
@@
 #For Z-score normalization
-d1.obese <- as.character(data1$obese)
-t.obese <- as.character(turn.data$obese)
-
-d1.bmi.class <- data1$BMI.class
-t.bmi.class <- as.character(turn.data$BMI.class)
-
-d1.H.Z-score <- data1$ZH
-t.H.Z-score <- turn.data$ZH
+t.obese <- as.character(ZScoreTurnbaughData$Obese)
+t.bmi.class <- as.character(ZScoreTurnbaughData$BMICat)
+t.ZH <- ZScoreTurnbaughData$ZH
+t.ZlogBF <- ZScoreTurnbaughData$ZLogBF
 
 #Combine everything together
-comb1.obese <- c(d1.obese, t.obese)
-comb1.bmi.class <- c(d1.bmi.class, t.bmi.class)
-comb1.H.Z-score <- as.data.frame(c(d1.H.Z-score, t.H.Z-score))
-
-group.data.test1 <- cbind(comb1.H.Z-score, comb1.obese, comb1.bmi.class)
-colnames(group.data.test1) <- c("H.Z-score", "obese", "BMI.class")
+comb1.obese <- c(obese, t.obese)
+comb1.bmi.class <- c(BMI.class, t.bmi.class)
+comb1.ZH <- as.data.frame(c(ZH, t.ZH))
+group.data.test1 <- cbind(comb1.ZH, comb1.obese, comb1.bmi.class)
+colnames(group.data.test1) <- c("ZH", "obese", "BMI.class")
 
@@
 
-##Run Z-score normalized Bacter and Firm overall analysis
-#data1 order Escobar, GLNE, CCHC, HMP, Korean, MetaHit
-d1.lzbf.corr <- data1$ZlogBF
-t.lzbf.corr <- turn.data$ZLogBF
-
 #Bacteroidetes/Firmicutes Datasets
-comb.lzbf.corr <- as.data.frame(c(d1.lzbf.corr, t.lzbf.corr))
-group.data.test <- cbind(comb.lzbf.corr, group.data.test)
+comb.ZlogBF <- as.data.frame(c(ZlogBF, t.ZlogBF))
+group.data.test <- cbind(comb.ZlogBF, group.data.test)
 colnames(group.data.test)[1] <- "ZLogBF"
@@
 ## Perform the statistical tests
-combZHTtest <- t.test(H.Z-score ~ obese, data=group.data.test) #P-value = 0.006061
-combZHANOVA <- anova(lm(H.Z-score ~ BMI.class, data=group.data.test)) #P-value = 0.0193
-a2 <- aov(H.Z-score ~ BMI.class, data=group.data.test)
+combZHTtest <- t.test(ZH ~ obese, data=group.data.test) #P-value = 0.006492
+combZHANOVA <- anova(lm(ZH ~ BMI.class, data=group.data.test)) #P-value = 0.01588
+a2 <- aov(ZH ~ BMI.class, data=group.data.test)
 combZHTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
@@
 
-combZLogBFTtest <- t.test(ZLogBF ~ obese, data=group.data.test) #P-value = 0.866
-combZLogBFANOVA <- anova(lm(ZLogBF ~ BMI.class, data=group.data.test)) #P-value = 0.106
+combZLogBFTtest <- t.test(ZLogBF ~ obese, data=group.data.test) #P-value = 0.6337
+combZLogBFANOVA <- anova(lm(ZLogBF ~ BMI.class, data=group.data.test)) #P-value = 0.7125
 a2 <- aov(ZLogBF ~ BMI.class, data=group.data.test)
@@
 # Section A
-p1 <- ggplot(group.data.test, aes(obese, H.Z-score))
+p1 <- ggplot(group.data.test, aes(obese, ZH))
 p2 <- p1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
@@
 # Section B
-q1 <- ggplot(group.data.test, aes(BMI.class, H.Z-score))
+q1 <- ggplot(group.data.test, aes(BMI.class, ZH))
 q2 <- q1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")

### 2016-03-13 code/StrategydenovoAnalysis.R

@@ -1298,7 +1298,7 @@ temporary <- table(metadata$sex_s)
 males <- unname(c(males, temporary[names(temporary) == "male"]))
 females <- unname(c(females, temporary[names(temporary) == "female"]))
 #temporary <- table(demographics$White)
-ancestry <- unname(c(ancestry, format(round(100, 2), nsmall = 2)))
+ancestry <- unname(c(ancestry, format(round(1, 2), nsmall = 2)))
 meanBMI <- c(meanBMI, 
              format(round(mean(metadata$BMI), 2), nsmall = 2))
 SDBMI <- c(SDBMI, 
@@ -2368,7 +2368,7 @@ highBF <- c(highBF, turnbaughBFRR[1,3])
 TurnbaughData <- as.data.frame(cbind(TurnbaughZH, TurnbaughZLogBF))
 TurnbaughData$Study <- "Turnbaugh"
 TurnbaughData <- as.data.frame(cbind(TurnbaughData, TurnbaughObese, TurnbaughBMICat))
-colnames(TurnbaughData) <- c("ZH", "ZLogBF", "Study", "BMICat", "Obese")
+colnames(TurnbaughData) <- c("ZH", "ZLogBF", "Study", "Obese", "BMICat")
 
 
 rm(TurnLowShannonGroup, TurnHighShannonGroup, turnbaughHRR, turnbaughBacter, 

### 2016-03-14 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
removed MetaHit Data

--- 2016-03-13 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
@@
 
-setwd("C:/users/marc/Desktop")
-demographics <- read.csv("denovodemographicsTable.csv")
-OverallPValue <- read.csv("denovoOverallPTable.csv")
-ShannonRRTable <- read.csv("denovoShannonRRTable.csv")
-BFRatioRRTable <- read.csv("denovoBFRatioRRTable.csv")
-ZScoreCombinedData <- read.csv("denovoCombinedData.csv")
-ZScoreTurnbaughData <- read.csv("denovoTurnbaughData.csv")
+demographics <- read.csv("results/tables/denovodemographicsTable.csv")
+OverallPValue <- read.csv("results/tables/denovoOverallPTable.csv")
+ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
+BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
+ZScoreCombinedData <- read.csv("results/tables/denovoCombinedData.csv")
+ZScoreTurnbaughData <- read.csv("results/tables/denovoTurnbaughData.csv")
 
@@
 **Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals.
-The Shannon diversity (N=1 study), OTU richness (N=2), and overall community dissimilarity (N=1) were significantly different between non-obese and obese individuals in at least one study (P<0.05).
+Only one diversity measure (OTU richness (N=1)) was significantly different between non-obese and obese individuals (P<0.05).
 With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: 0.76; s.d.=0.09).
 In seven of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies.
-When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04-1.36, P-value = 0.02) and B/F ratio (RR = 0.86, CI = 0.75-0.99, P-value = 0.04).
-Finally, when we normalized the Shannon diversity data and pooled studies there was a significant difference between non-obese and obese individuals (difference in z-score=0.155; P=0.006) something about...
+When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = 1.16, CI = 1.01-1.34, P-value = 0.04) but not B/F ratio (RR = 0.98, CI = 0.85-1.13, P-value = 0.81).
+Finally, when we normalized the Shannon diversity data and pooled studies there was a significant difference between non-obese and obese individuals (difference in z-score=0.155; P=0.02) something about...
 B/F.
@@
 For the Shannon diversity RR, all studies but Wu et al had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 2].
-This was confirmed with the pooled analysis showing an increased RR of 1.18 (CI 1.03, 1.36) for obesity in the lower then median Shannon diversity group (P-value = 0.0187).
-When investigated the B/F ratio RR most studies showed that a low B/F ratio tended to have less obese individuals than those with a high B/F ratio [Figure 3].
-A notable exception to this was the HMP data set and the Baxter data set [Figure 3].
-The pooled RR was 0.86 (CI 0.75, 0.99) suggesting that a low B/F ratio does indeed have less obese individuals (P-value = 0.0376).
+This was confirmed with the pooled analysis showing an increased RR of 1.16 (CI 1.01, 1.34) for obesity in the lower then median Shannon diversity group (P-value = 0.0412).
+When investigated the B/F ratio RR most studies showed that there was no significant difference between obese and non-obese individuals [Figure 3].
+The pooled RR was 0.98 (CI 0.85, 1.13) suggesting that a low B/F ratio does indeed have less obese individuals (P-value = 0.811).
 
@@
 
-Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "HMP", "Wu", "Arumugam", "Turnbaugh")
+Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "HMP", "Wu", "Turnbaugh")
 
-Type <- c("Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational")
+Type <- c("Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational")
 
 Population <- c("Adults without Cancer from Texas, Ontario, Michigan, Massachusetts", "Hispanic adults with and without type 2 diabetes from Texas
-", "Adult twins and mothers from the United Kingdom", "Healthy adults from Columbia", "Amish Adults with and without metabolic syndrome from Pennsylvania", "Healthy Adults from Texas and Missouri", "Healthy Adults from Pennsylvania", "Healthy Adults from Denmark", "Adult twins and mothers from Missouri")
+", "Adult twins and mothers from the United Kingdom", "Healthy adults from Columbia", "Amish Adults with and without metabolic syndrome from Pennsylvania", "Healthy Adults from Texas and Missouri", "Healthy Adults from Pennsylvania", "Adult twins and mothers from Missouri")
 
-Published <- c("No", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes","Yes")
+Published <- c("No", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes","Yes")
 
 Outcomes <- c("N/A (Unpublished)", "Mentions obesity, no specific correlations made though", "Christensenellaceae higher in low BMI individuals.
-No other correlations found.", "Firmicutes less abundant at higher BMI", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "No significant correlations", "Unweighted UniFrac Correlation with BMI", "No significant  correlations", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ")
+No other correlations found.", "Firmicutes less abundant at higher BMI", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "No significant correlations", "Unweighted UniFrac Correlation with BMI", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ")
 
@@
 
-![Figure 1](Sze_ObesityMicrobiomeMeta_PeerJ_2016_files/figure-html/PrismaFlowGraph.png)\
+![Figure 1](results/figures/PrismaFlowGraph.png)\
 
@@
 
-```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
 
@@
 Study <- as.character(ShannonRRTable$Study)
-Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2010", "2009")
-Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63","n=85", "n=146")
+Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
+Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
 tpos <- as.numeric(ShannonRRTable$tposH) #obese in low diversity group
@@
 **Figure 2: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity.**  Groups were divided for each study on high and low Shannon Diversity groups based on the median for that study.
-The overall pooled relative risk was 1.17 for the low diversity group (P-value = `r signif(RRSDPvalue$QMp, 3)`).
+The overall pooled relative risk was 1.16 for the low diversity group (P-value = `r signif(RRSDPvalue$QMp, 3)`).
 
-```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
 
@@
 Study <- as.character(BFRatioRRTable$Study)
-Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2010", "2009")
-Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63","n=85", "n=146")
+Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
+Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
 tpos <- as.numeric(BFRatioRRTable$tposBF) #obese in low diversity group
@@
 **Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low B/F ratio groups based on the median for that specific study.
-The overall pooled relative risk was 0.90 for the low diversity group (P-value = `r signif(RRBFPvalue$QMp, 3)`).
+The overall pooled relative risk was 0.98 for the low diversity group (P-value = `r signif(RRBFPvalue$QMp, 3)`).
 
-```{r Figure_4, echo=FALSE}
+```{r Figure_4, echo=FALSE, fig.path="results/figures/"}
 par(mfrow=c(1,2))
@@
 
-```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
 # Section A

### 2016-03-14 code/StrategydenovoAnalysisMetaHitDel.R

@@ -1,0 +1,2162 @@
+## Strategy 1
+## Obesesity and the bacterial microbiome
+## Marc Sze
+## March 9, 2016
+
+library(vegan)
+library(epiR)
+library(AUCRF)
+
+############# BAXTER ######################################################
+
+
+###########################################################################
+############ Preparing Data Tables for Analysis ###########################
+###########################################################################
+
+# Reading in the necessary Data
+demographics <- read.csv("data/process/Baxter/demographics.v2.csv")
+microbiome <- read.csv("data/process/Baxter/data1.subsample.otus.csv")
+phylogenetic.info <- read.csv("data/process/Baxter/data1.summary.taxonomy.csv")
+
+# Minor modifications and Rownames adustments of data tables
+rownames(demographics) <- demographics[,1]
+rownames(microbiome) <- microbiome[,2]
+demographics <- demographics[,-1]
+microbiome <- microbiome[,-2]
+test.samples <- rownames(demographics)
+microbiome <- microbiome[,-c(1:2)]
+
+# Create Table for phyla information
+phylogenetic.info <- phylogenetic.info[-c(1:12),]
+phyla <- which(phylogenetic.info$taxlevel == 2)
+phyla.table <- phylogenetic.info[phyla, ]
+phyla.table <- phyla.table[,-c(1:2,4:5)]
+phyla.names <- as.character(phyla.table[,1])
+phyla.table <- phyla.table[,-1]
+phyla.table <- as.data.frame(t(phyla.table))
+colnames(phyla.table) <- phyla.names
+rownames(phyla.table) <- rownames(demographics)
+rm(phyla, phyla.names, phylogenetic.info)
+
+#Add phyla together that are not very abundant and delete them from the table
+phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deferribacteres", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "Cyanobacteria_Chloroplast", "TM7")], 1, sum)
+phyla.table <- phyla.table[, -c(1, 4:7, 9:11, 13, 15)]
+
+#Create a relative abundance table for phyla
+phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
+phyla.table.rel.abund <- (phyla.table/phyla.total)*100
[…]

### 2016-03-14 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
manuscript code update

@@
 ZScoreTurnbaughData <- read.csv("results/tables/denovoTurnbaughData.csv")
+AUCRFData <- read.csv("results/tables/denovoAUCRFDataTable.csv")
 
@@
 
+```{r Data_Load_Fig4, echo=FALSE, message=FALSE, warning=FALSE}
+
+#data1 order Escobar, Baxter, Ross, HMP, Arumugam, Wu, Goodrich, Zupancic
+
+ZH <- as.numeric(ZScoreCombinedData$ZH)
+bmi <- as.numeric(ZScoreCombinedData$BMI)
+BMI.class <- as.character(ZScoreCombinedData$BMICat)
+BMI.class[BMI.class == "Normal"] <- "Lean"
+obese <- as.character(ZScoreCombinedData$Obese)
+ZlogBF <- as.numeric(ZScoreCombinedData$ZLogBF)
+
+data1 <- as.data.frame(cbind(ZH, ZlogBF, bmi))
+
+#Create group analysis table to harness full data set
+#For Z-score normalization
+t.obese <- as.character(ZScoreTurnbaughData$Obese)
+t.bmi.class <- as.character(ZScoreTurnbaughData$BMICat)
+t.ZH <- ZScoreTurnbaughData$ZH
+t.ZlogBF <- ZScoreTurnbaughData$ZLogBF
+
+#Combine everything together
+comb1.obese <- c(obese, t.obese)
+comb1.bmi.class <- c(BMI.class, t.bmi.class)
+comb1.ZH <- as.data.frame(c(ZH, t.ZH))
+group.data.test1 <- cbind(comb1.ZH, comb1.obese, comb1.bmi.class)
+colnames(group.data.test1) <- c("ZH", "obese", "BMI.class")
+
+#Let's change the order so it makes sense
+group.data.test1$BMI.class <- factor(group.data.test1$BMI.class, levels = c("Lean", "Overweight", "Obese"))
+
+group.data.test <- group.data.test1
+
+#Bacteroidetes/Firmicutes Datasets
+comb.ZlogBF <- as.data.frame(c(ZlogBF, t.ZlogBF))
+group.data.test <- cbind(comb.ZlogBF, group.data.test)
+colnames(group.data.test)[1] <- "ZLogBF"
+
+## Perform the statistical tests
+combZHTtest <- t.test(ZH ~ obese, data=group.data.test) #P-value = 0.006492
+combZHANOVA <- anova(lm(ZH ~ BMI.class, data=group.data.test)) #P-value = 0.01588
+a2 <- aov(ZH ~ BMI.class, data=group.data.test)
+combZHTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
+combZHTukeyFinal <- combZHTukey$BMI.class
+
+combZLogBFTtest <- t.test(ZLogBF ~ obese, data=group.data.test) #P-value = 0.6337
+combZLogBFANOVA <- anova(lm(ZLogBF ~ BMI.class, data=group.data.test)) #P-value = 0.7125
+a2 <- aov(ZLogBF ~ BMI.class, data=group.data.test)
+combZLogBFTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
+combZLogBFTukeyFinal <- combZLogBFTukey$BMI.class
+
+```
+
 *******
@@
 
-**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 9 studies that fit our criteria.
+**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 8 studies that fit our criteria.
 We applied three analysis pipelines to the data from each study.
@@
 Only one diversity measure (OTU richness (N=1)) was significantly different between non-obese and obese individuals (P<0.05).
-With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: 0.76; s.d.=0.09).
+With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`).
 In seven of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies.
 When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = 1.16, CI = 1.01-1.34, P-value = 0.04) but not B/F ratio (RR = 0.98, CI = 0.85-1.13, P-value = 0.81).
-Finally, when we normalized the Shannon diversity data and pooled studies there was a significant difference between non-obese and obese individuals (difference in z-score=0.155; P=0.02) something about...
-B/F.
+Finally, when we normalized the Shannon diversity data and pooled studies there was a significant difference between non-obese and obese individuals (difference in mean z-score = `r format(round(combZHTtest$estimate[1]-combZHTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZHTtest$p.value, 2), nsmall=2)`).
+Similar to RR there was no significant difference in the normalized B/F ratio between non-obese and obese individuals (difference in mean z-score = `r format(round(combZLogBFTtest$estimate[1]-combZLogBFTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`).
 
@@
 
-```{r Data_Load_Fig4, echo=FALSE, message=FALSE, warning=FALSE}
-
-#data1 order Escobar, Baxter, Ross, HMP, Arumugam, Wu, Goodrich, Zupancic
-
-ZH <- as.numeric(ZScoreCombinedData$ZH)
-bmi <- as.numeric(ZScoreCombinedData$BMI)
-BMI.class <- as.character(ZScoreCombinedData$BMICat)
-BMI.class[BMI.class == "Normal"] <- "Lean"
-obese <- as.character(ZScoreCombinedData$Obese)
-ZlogBF <- as.numeric(ZScoreCombinedData$ZLogBF)
-
-data1 <- as.data.frame(cbind(ZH, ZlogBF, bmi))
-
-#Create group analysis table to harness full data set
-#For Z-score normalization
-t.obese <- as.character(ZScoreTurnbaughData$Obese)
-t.bmi.class <- as.character(ZScoreTurnbaughData$BMICat)
-t.ZH <- ZScoreTurnbaughData$ZH
-t.ZlogBF <- ZScoreTurnbaughData$ZLogBF
-
-#Combine everything together
-comb1.obese <- c(obese, t.obese)
-comb1.bmi.class <- c(BMI.class, t.bmi.class)
-comb1.ZH <- as.data.frame(c(ZH, t.ZH))
-group.data.test1 <- cbind(comb1.ZH, comb1.obese, comb1.bmi.class)
-colnames(group.data.test1) <- c("ZH", "obese", "BMI.class")
-
-#Let's change the order so it makes sense
-group.data.test1$BMI.class <- factor(group.data.test1$BMI.class, levels = c("Lean", "Overweight", "Obese"))
-
-group.data.test <- group.data.test1
-
-#Bacteroidetes/Firmicutes Datasets
-comb.ZlogBF <- as.data.frame(c(ZlogBF, t.ZlogBF))
-group.data.test <- cbind(comb.ZlogBF, group.data.test)
-colnames(group.data.test)[1] <- "ZLogBF"
-
-## Perform the statistical tests
-combZHTtest <- t.test(ZH ~ obese, data=group.data.test) #P-value = 0.006492
-combZHANOVA <- anova(lm(ZH ~ BMI.class, data=group.data.test)) #P-value = 0.01588
-a2 <- aov(ZH ~ BMI.class, data=group.data.test)
-combZHTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-combZHTukeyFinal <- combZHTukey$BMI.class
-
-combZLogBFTtest <- t.test(ZLogBF ~ obese, data=group.data.test) #P-value = 0.6337
-combZLogBFANOVA <- anova(lm(ZLogBF ~ BMI.class, data=group.data.test)) #P-value = 0.7125
-a2 <- aov(ZLogBF ~ BMI.class, data=group.data.test)
-combZLogBFTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-combZLogBFTukeyFinal <- combZLogBFTukey$BMI.class
-
-```
-
 ```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}

### 2016-03-15 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
TwinsUK denovo added

@@
 AUCRFData <- read.csv("results/tables/denovoAUCRFDataTable.csv")
+PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
 
@@
 With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`).
-In seven of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies.
+In 7/8 of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies.
 When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = 1.16, CI = 1.01-1.34, P-value = 0.04) but not B/F ratio (RR = 0.98, CI = 0.85-1.13, P-value = 0.81).
@@
 
+**Table 4. Summary of Power Calculations for each Individual for Shannon Diversity and B/F ratio for Non-obese versus Obese**
+
+```{r Table_4, echo=FALSE, message=FALSE, warning=FALSE}
+
+Study <- as.character(PowerTable$Study)
+PowerH <- PowerTable$StudyPowerH
+PowerBF <- PowerTable$StudyPowerBF
+
+table4 <- cbind(Study, PowerH, PowerBF)
+colnames(table4) <- c("Study", "Shannon Diversity Power (%)", "B/F Ratio Power (%)")
+
+kable(table4, format = "markdown", align = 'c', padding = 2)
+```
+
 *******
@@
 **Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** Z-score Normalized Shannon Diversity and Obesity.
-There was a significant difference between non-obese (n=1081) and obese groups (n=435) (P-value = `r signif(combZHTtest$p.value, 3)`).
+There was a significant difference between non-obese (n = `r length(which(group.data.test$obese == "No"))`) and obese groups (n = `r length(which(group.data.test$obese == "Yes"))`) (P-value = `r signif(combZHTtest$p.value, 3)`).
 **B)** Z-score Normalized Shannon Diversity by BMI Group.
-Total n for lean, overweight, and obese groups are 563, 435, 518 respectively.
+Total n for lean, overweight, and obese groups are `r length(which(group.data.test$BMI.class == "Lean"))`, `r length(which(group.data.test$BMI.class == "Overweight"))`, `r length(which(group.data.test$BMI.class == "Obese"))` respectively.
 Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = `r signif(combZHTukeyFinal[3, 4], 3)`) but no difference between lean and obese group (P-value = `r signif(combZHTukeyFinal[2, 4], 3)`).
 **C)** Z-score Normalized B/F Ratio and Obesity.
-There was no significant difference between the non-obese (n=1081) and obese (n=435) group based on B/F ratio (P-value = `r signif(combZLogBFTtest$p.value, 3)`).
+There was no significant difference between the non-obese (n = `r length(which(group.data.test$obese == "No"))`) and obese (n = `r length(which(group.data.test$obese == "Yes"))`) group based on B/F ratio (P-value = `r signif(combZLogBFTtest$p.value, 3)`).
 **D)** Z-score Normalized B/F Ratio by BMI Group.
-Total n for lean, overweight, and obese groups are 563, 435, 518 respectively.
+Total n for lean, overweight, and obese groups are `r length(which(group.data.test$BMI.class == "Lean"))`, `r length(which(group.data.test$BMI.class == "Overweight"))`, `r length(which(group.data.test$BMI.class == "Obese"))` respectively.
 There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = `r signif(combZLogBFANOVA[1, 5], 3)`).

### 2016-03-15 code/StrategydenovoAnalysisMetaHitDel.R
add TwinsUK denovo and class RR

@@ -187,15 +187,104 @@ testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
 testset <- cbind(obese, testset)
 
 #Try AUCRF with default measures provided in readme
-#set.seed(3)
-#baxterAUCfit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+set.seed(3)
+baxterAUCfit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
+baxAUC <- baxterAUCfit$`OOB-AUCopt`
+baxKopt <- baxterAUCfit$Kopt
+
+
+###########################################################################
+########## Similar Classification OTUs Data Table Prep ####################
+###########################################################################
+
+Clostridiales <- microbiome$Otu00479
+Lachnospiraceae <- microbiome$Otu00036
+Ruminococcaceae <- microbiome$Otu00320
+
+##Run the RR for Clostridiales
+test <- as.data.frame(Clostridiales)
+test <- within(test, {Clostridiales.cat = ifelse(Clostridiales <= median(Clostridiales), "less", "higher")})
+
+test4 <- cbind(test, obese)
+test4 <- test4[order(Clostridiales.cat), ]
+orderedtestCat <- test4[, 2]
+BaxtestHighTotal <- length(orderedtestCat[
+  orderedClostridialesCat=="higher"])
+BaxHightestGroup <- as.data.frame(
+  table(test4[c(1:BaxtestHighTotal), 3]))
+BaxLowtestGroup <- as.data.frame(
+  table(test4[c((BaxtestHighTotal + 1):BaxtotalN), 3]))
+
+group1 <- c(BaxHightestGroup[2, 2], BaxHightestGroup[1, 2])
+group2 <- c(BaxLowtestGroup[2, 2], BaxLowtestGroup[1, 2])
+r.test <- rbind(group2, group1)
+colnames(r.test) <- c("Obese", "Not.Obese")
+rownames(r.test) <- c("group2", "group1")
+
+baxterClostEpi <- epi.2by2(r.test, method="cohort.count")
+baxterClostMassoc <- baxterBFEpi$massoc
+baxterClostsRR <- baxterBFMassoc$RR.strata.score
+baxterClostRRsig <- baxterBFMassoc$chisq.strata
+
+
+##Run the RR for Lachnospiraceae
+test <- as.data.frame(Lachnospiraceae)
[…]

### 2016-03-17 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
Readability and Common OTU Class

@@
 Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.
-We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was 0.7553 +/- 0.0855.
+We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)` +/- `r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`.
 We found that the total range for the total number of variables used for the classification ranged from 3 (Ross and Escobar) to 40 (Zupancic).
@@
 One aspect we found surprising was that there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study.
-The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/9 studies), an OTU that could be classified to the family Lachnospiraceae (5/9 studies), and the Shannon diversity metric (4/9 studies).
+The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/8 studies), an OTU that could be classified to the family Lachnospiraceae (6/8 studies), Clostridiales (6/8) and Shannon diversity metric (5/8 studies).
 

### 2016-03-17 code/CommonClassOTUsMetaHitDel.R

@@ -7,7 +7,7 @@
 library(vegan)
 library(epiR)
 library(AUCRF)
-
+source("code/UsedFunctions.R")
 
 ###########################################################################
 ############ Preparing Data Tables for Analysis ###########################
@@ -122,7 +122,8 @@ GoodrichDemo$BMIclass2[GoodrichDemo$BMI.class=="Normal"] <- "Normal"
 GoodrichDemo$BMIclass2[GoodrichDemo$BMI.class=="Overweight"] <- "Overweight"                     
 GoodrichDemo$BMIclass2[GoodrichDemo$BMI.class=="Obese" | 
                          GoodrichDemo$BMI.class=="Extreme Obesity"] <- "Obese"
-
+colnames(GoodrichDemo)[12] <- "BMI"
+GoodrichDemo$BMI <- as.numeric(as.character(GoodrichDemo$BMI))
 rm(keep)
 
 
@@ -315,91 +316,113 @@ rm(keep1)
 
 
 ###########################################################################
-########## Similar Classification OTUs Data Table Prep ####################
+########## Visualizing How the Different OTUs look ########################
+################### Clostridiales OTUs ####################################
 ###########################################################################
 
-Clostridiales <- microbiome$Otu00479
-Lachnospiraceae <- microbiome$Otu00036
-Ruminococcaceae <- microbiome$Otu00320
+library(gridExtra)
 
-##Run the RR for Clostridiales
-test <- as.data.frame(Clostridiales)
-test <- within(test, {Clostridiales.cat = ifelse(Clostridiales <= median(Clostridiales), "less", "higher")})
+a1 <- ggBaseBox(BaxterMicrobiome, BaxterDemo, "Otu00479", "Baxter", log=TRUE)
+b1 <- ggBaseBox(RossMicrobiome, RossDemo, "Otu00102", "Ross", log=TRUE)
+c1 <- ggBaseBox(GoodrichMicrobiome, GoodrichDemo, 
+                "Otu000287", "Goodrich", log=TRUE)
+d1 <- ggBaseBox(GoodrichMicrobiome, GoodrichDemo, 
[…]

### 2016-03-17 code/UsedFunctions.R

@@ -1,0 +1,60 @@
+## Functions For Meta Analysis Paper on
+## Obesesity and the bacterial microbiome
+## Marc Sze
+## March 16, 2016
+
+
+## Function to graph OTU versus BMI linear regression with log scale option
+## Default for log is FALSE. 
+## Example:
+## q1 <- ggBaseLinear(RossMicrobiome, RossDemo, "Otu00102", log=TRUE)
+ggBaseLinear <- function(microbiome, demo, otuNum, log=FALSE){
+  deps = c("ggplot2");
+  for (dep in deps){
+    if (dep %in% installed.packages()[,"Package"] == FALSE){
+      install.packages(as.character(dep), quiet=TRUE);
+    }
+    library(dep, verbose=FALSE, character.only=TRUE)
+  }
+  
+  if(log==FALSE){
+    OTUData <- as.data.frame(microbiome[otuNum])
+  } else{
+    OTUData <- as.data.frame(log(microbiome[otuNum] + 1))
+  }
+  bmi <- demo$BMI
+  combined <- cbind(OTUData, bmi)
+  ggplot(combined, aes(x=combined[, 1], y=combined[, 2])) + 
+    geom_point(shape=1) + 
+    geom_smooth(method=lm) + xlab(otuNum) + ylab(colnames(combined)[2])
+}
+
+
+## Function to graph OTU versus BMI  boxplot with log scale option
+## Default for log is FALSE. 
+## Example:
+## q1 <- ggBaseLinear(RossMicrobiome, RossDemo, "Otu00102", "Ross", log=TRUE)
+ggBaseBox <- function(microbiome, demo, otuNum, title, log=FALSE){
+  deps = c("ggplot2", "RColorBrewer");
+  for (dep in deps){
+    if (dep %in% installed.packages()[,"Package"] == FALSE){
+      install.packages(as.character(dep), quiet=TRUE);
+    }
+    library(dep, verbose=FALSE, character.only=TRUE)
+  }
+  
+  if(log==FALSE){
+    OTUData <- as.data.frame(microbiome[otuNum])
+  } else{
+    OTUData <- as.data.frame(log(microbiome[otuNum] + 1))
+  }
+  obese <- as.factor(demo$obese)
+  combined <- cbind(OTUData, obese)
+  ggplot(combined, aes(obese, combined[, 1])) + 
+    geom_boxplot(aes(fill = obese)) + 
[…]

### 2016-03-17 code/UsedFunctions.R
R functions update

@@ -3,12 +3,92 @@
 ## Marc Sze
 ## March 16, 2016
 
+## Function to make table for RR Analysis and run RR Test.
+## Needed Libraries include epiR.  
+## An Example of how to use:
+  ## RunRR(alpha.test, demographics, "obese", "H")
+RunRR <- function(testData, demo, testVar1, testVar2){
+  
+  # Check for necessary Libraries and load in not already attached
+  deps = c("epiR");
+  for (dep in deps){
+    if (dep %in% installed.packages()[,"Package"] == FALSE){
+      install.packages(as.character(dep), quiet=TRUE);
+    }
+    library(dep, verbose=FALSE, character.only=TRUE)
+  }
+  
+  # Generate categorical Data
+  Var1 <- demo[testVar1]
+  
+  # Generate Median Categorical Data
+  VarOfInt <- testData[testVar2]
+  VarOfInt <- as.data.frame(within(VarOfInt, 
+                     {Var.cat = ifelse(
+                       VarOfInt[,1] <= median(VarOfInt[,1]), 
+                       "less", "higher")}))
+  VarOfInt <- cbind(VarOfInt, Var1)
+  VarOfInt <- VarOfInt[order(VarOfInt[, 2]), ]
+  
+  # Generate Test table
+  totalN <- length(VarOfInt[, 2])
+  HighTotal <- length(which(VarOfInt[, 2]=="higher"))
+  HighMedianGroup <- as.data.frame(table(
+    VarOfInt[c(1:HighTotal), 3]))
+  LowMedianGroup <- as.data.frame(table(
+    VarOfInt[c((HighTotal + 1):totalN), 3]))
+  ConGroup <- c(HighMedianGroup[2, 2], HighMedianGroup[1, 2])
+  TestGroup <- c(LowMedianGroup[2, 2], LowMedianGroup[1, 2])
+  r.test <- rbind(TestGroup, ConGroup)
+  colnames(r.test) <- c("Yes", "No")
+  rownames(r.test) <- c("TestGroup", "ConGroup")
+  
+  # Run test and store important variables
+  EpiTest <- epi.2by2(r.test, method="cohort.count")
+  MassocTest <- EpiTest$massoc
+  RRValues <- MassocTest$RR.strata.score
+  RRsignif <- MassocTest$chisq.strata
+  
+  # Create Other variables to store for future use if needed
+  tpos <- LowMedianGroup[2, 2]
[…]

### 2016-03-17 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
manuscript updates

@@
 An additional study was removed from our analysis because no individuals in the study had a BMI over 30.
-Among the final 9 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@ross_16s_2015; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @Huttenhower2012], and one dataset was used (Baxter) because at the time of our analysis it was unpublished and enabled us to mitigate against publication biases associated with the bacterial microbiome and obesity.
+Among the final 9 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and one dataset was used (Baxter) because at the time of our analysis it was unpublished and enabled us to mitigate against publication biases associated with the bacterial microbiome and obesity.
 The nine studies are summarized in Tables 1 and 2.
@@
 All sequence data were publicly available and were downloaded from the NCBI Sequence Read Archive, the European Nucleotide Archive, or the investigators' personal website (https://gordonlab.wustl.edu/TurnbaughSE\_10\_09/STM\_2009.html).
-The nine studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011; @Huttenhower2012] or Illumina sequencing [@goodrich_human_2014; @arumugam_enterotypes_2011; Baxter].
-All of these studies used amplification-based 16S rRNA gene sequencing except for one, which used metagenomic shotgun sequencing [@arumugam_enterotypes_2011].
-Among the studies that sequenced the 16S rRNA gene, the researchers targeted the V1-V2  [@wu_linking_2011], V1-V3 [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], V3-V5 [@Huttenhower2012], and V4 [@goodrich_human_2014] regions.
-For those studies where multiple regions were sequenced,  we selected the region that corresponded to the largest number of subjects [@turnbaugh_core_2009; @Huttenhower2012].
+The nine studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011;
+@human_microbiome_project_consortium_structure_2012] or Illumina sequencing [@goodrich_human_2014; Baxter].
+All of these studies used amplification-based 16S rRNA gene sequencing.
+Among the studies that sequenced the 16S rRNA gene, the researchers targeted the V1-V2  [@wu_linking_2011], V1-V3 [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], V3-V5 [@human_microbiome_project_consortium_structure_2012], and V4 [@goodrich_human_2014] regions.
+For those studies where multiple regions were sequenced,  we selected the region that corresponded to the largest number of subjects [@turnbaugh_core_2009;
+@human_microbiome_project_consortium_structure_2012].
 We processed the 16S rRNA gene sequence data as described in the original studies.

### 2016-03-17 code/StrategydenovoAnalysisMetaHitDel.R
main analysis R file update

@@ -6,6 +6,7 @@
 library(vegan)
 library(epiR)
 library(AUCRF)
+source("code/UsedFunctions.R")
 
 ############# BAXTER ######################################################
 
@@ -107,34 +108,8 @@ baxterPERM <- baxter2$aov.tab
 ############ Relative Risk#################################################
 ###########################################################################
 
-#Generate median values and put them into existing alpha.test dataframe
-
-#Shannon diversity
-alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})
-
-##Run the RR for Shannon Diversity
-H.cat <- alpha.test$shannon.cat
-bmi.cat <- as.character(obese)
-test3 <- cbind(H.cat, bmi.cat)
-test3 <- test3[order(H.cat), ]
-orderedHCat <- test3[, 1]
-BaxtotalN <- length(orderedHCat)
-BaxHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
-BaxHighShannonGroup <- as.data.frame(table(test3[c(1:BaxHHighTotal), 2]))
-BaxLowShannonGroup <- as.data.frame(table(test3[c((BaxHHighTotal + 1):BaxtotalN), 2]))
-#Group1 (Higher than median), obese = 22 and non-obese = 64
-#Group2 (Lower than median), obese = 25 and non-obese = 61
-
-group1 <- c(BaxHighShannonGroup[2, 2], BaxHighShannonGroup[1, 2])
-group2 <- c(BaxLowShannonGroup[2, 2], BaxLowShannonGroup[1, 2])
-r.test <- rbind(group2, group1)
-colnames(r.test) <- c("Obese", "Not.Obese")
-rownames(r.test) <- c("group2", "group1")
-
-baxterHEpi <- epi.2by2(r.test, method="cohort.count")
-baxterHMassoc <- baxterHEpi$massoc
-baxterHRR <- baxterHMassoc$RR.strata.score
-baxterHRRsig <- baxterHMassoc$chisq.strata
+# Run Shannon Diversity RR Test
+BaxterHRR <- RunRR(alpha.test, demographics, "obese", "H")
 ## Risk Ratio = 1.14
 ## CI = 0.70, 1.85
[…]

### 2016-03-18 code/StrategydenovoAnalysisMetaHitDel.R
finish adding power calc to code

@@ -3,6 +3,7 @@
 ## Marc Sze
 ## March 9, 2016
 
+library(statmod)
 library(vegan)
 library(epiR)
 library(AUCRF)
@@ -164,8 +165,22 @@ combinedData$Study <- "Baxter"
 StudyPowerH <- NonParaPowerSim(
   alpha.test, demographics, "H", "obese", n=1000)
 
+set.seed(3)
+StudyPowerHRR <- power.fisher.test(
+  BaxterHRR$tpos/(BaxterHRR$tpos+BaxterHRR$tneg), 
+  BaxterHRR$cpos/(BaxterHRR$cpos+BaxterHRR$cneg), 
+  BaxterHRR$tpos+BaxterHRR$tneg, BaxterHRR$cpos+BaxterHRR$cneg, 
+  alpha=0.05, nsim=1000, alternative="two.sided")*100
+
 StudyPowerBF <- NonParaPowerSim(
   BFRatio, demographics, "BFRatio", "obese", n=1000)
+
+set.seed(3)
+StudyPowerBFRR <- power.fisher.test(
+  BaxterBFRR$tpos/(BaxterBFRR$tpos+BaxterBFRR$tneg), 
+  BaxterBFRR$cpos/(BaxterBFRR$cpos+BaxterBFRR$cneg), 
+  BaxterBFRR$tpos+BaxterBFRR$tneg, BaxterBFRR$cpos+BaxterBFRR$cneg, 
+  alpha=0.05, nsim=1000, alternative="two.sided")*100
 
 rm(baxterBacter, baxterFirm, baxterBF, baxterH, baxterS, baxterJ, baxterPERM, 
    BaxterZH, BaxterZLogBF, BaxterBMI, baxter2)
@@ -346,10 +361,23 @@ combinedData <- rbind(combinedData, RossData)
 StudyPowerH <- c(StudyPowerH, 
                  NonParaPowerSim(
                    alpha.test, metadata, "H", "obese", n=1000))
+set.seed(3)
+StudyPowerHRR <- c(StudyPowerHRR, power.fisher.test(
+  RossHRR$tpos/(RossHRR$tpos+RossHRR$tneg), 
+  RossHRR$cpos/(RossHRR$cpos+RossHRR$cneg), 
+  RossHRR$tpos+RossHRR$tneg, RossHRR$cpos+RossHRR$cneg, 
+  alpha=0.05, nsim=1000, alternative="two.sided")*100)
 
[…]

### 2016-03-18 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
power calc add to paper

@@
 ```{r install_needed_packages, echo=FALSE, message=FALSE, warning=FALSE}
-deps = c("knitr","vegan","epiR","AUCRF","meta","metafor","ggplot2", "plyr", "reshape2", "RColorBrewer", "scales", "gridExtra");
+deps = c("knitr","vegan","epiR","AUCRF","meta","metafor","ggplot2", "plyr", "reshape2", "RColorBrewer", "scales", "gridExtra", "pwr", "statmod");
 for (dep in deps){
@@
 
-**Table 4. Summary of Power Calculations for each Individual for Shannon Diversity and B/F ratio for Non-obese versus Obese**
+**Table 4. Summary of Power Calculations for Shannon Diversity and B/F ratio for Non-obese versus Obese**
 
@@
 
-Study <- as.character(PowerTable$Study)
-PowerH <- PowerTable$StudyPowerH
-PowerBF <- PowerTable$StudyPowerBF
+Study <- c(as.character(PowerTable$Study), "COMBINED")
+keep <- which(group.data.test$obese=="Yes")
+CombinedH <- pwr.t2n.test(d=mean(group.data.test$ZH[-keep])-mean(group.data.test$ZH[keep]), n1=length(group.data.test$ZH[-keep]), n2=length(group.data.test$ZH[keep]))
+CombinedBF <- pwr.t2n.test(d=mean(group.data.test$ZLogBF[-keep])-mean(group.data.test$ZLogBF[keep]), n1=length(group.data.test$ZLogBF[-keep]), n2=length(group.data.test$ZLogBF[keep]))
 
-table4 <- cbind(Study, PowerH, PowerBF)
-colnames(table4) <- c("Study", "Shannon Diversity Power (%)", "B/F Ratio Power (%)")
+set.seed(3)
+CombinedHRR <- power.fisher.test(
+  sum(ShannonRRTable$tposH)/
+        (sum(ShannonRRTable$tposH)+sum(ShannonRRTable$tnegH)), 
+  sum(ShannonRRTable$cposH)/
+    (sum(ShannonRRTable$cposH)+sum(ShannonRRTable$cnegH)), 
+  sum(ShannonRRTable$tposH)+sum(ShannonRRTable$tnegH), sum(ShannonRRTable$cposH)+sum(ShannonRRTable$cnegH), 
+  alpha=0.05, nsim=1000, alternative="two.sided")*100
+
+set.seed(3)
+CombinedBFRR <- power.fisher.test(
+  sum(BFRatioRRTable$tposBF)/
+        (sum(BFRatioRRTable$tposBF)+sum(BFRatioRRTable$tnegBF)), 
+  sum(BFRatioRRTable$cposBF)/
+    (sum(BFRatioRRTable$cposBF)+sum(BFRatioRRTable$cnegBF)), 
+  sum(BFRatioRRTable$tposBF)+sum(BFRatioRRTable$tnegBF), sum(BFRatioRRTable$cposBF)+sum(BFRatioRRTable$cnegBF), 
+  alpha=0.05, nsim=1000, alternative="two.sided")*100
+
+PowerH <- c(PowerTable$StudyPowerH, format(round(CombinedH$power*100, 1), nsmall=1))
+PowerHRR <- c(PowerTable$StudyPowerHRR, CombinedHRR)
+PowerBF <- c(PowerTable$StudyPowerBF, format(round(CombinedBF$power*100, 1), nsmall=1))
+PowerBFRR <-c(PowerTable$StudyPowerBFRR, CombinedBFRR)
+
+table4 <- cbind(Study, PowerH, PowerHRR, PowerBF, PowerBFRR)
+colnames(table4) <- c("Study", "Shannon Diversity Power (%)", "Shannon RR Power (%)", "B/F Ratio Power (%)", "B/F ratio RR (%)")
 

### 2016-03-18 microbiomeAndobesitySupplemental.Rmd
large supplemental updates

--- 2016-02-19 microbiomeAndobesitySupplemental.Rmd
@@
 
-```{r data_table_loading, include=FALSE}
-setwd("C:/Users/marc/Desktop/obesity2/HMP.analysis")
-
-metaphlanData <- read.table("HMP_metaphlan2_merged_abundance_table.txt", header=T)
-IDconversion <- read.csv("HMP.stool.IDs.csv")
-microbiome <- read.table("Stool.an.0.03.subsample.shared", header=T)
-
-#Only interested in first visit microbiome data
-#need to create microbiome data set for only that
-
-Data16S <- microbiome[grep("\\.01\\.", microbiome$Group), ] #selecting by .01.
-
-microb.rownames <- gsub("([0-9]+).*", "\\1", Data16S$Group) #extracting only numeric before first "."
-rownames(Data16S) <- microb.rownames
-
-#Get rid of information not used in downstram analysis
-Data16S <- Data16S[, -c(1:3)]
-
-#Get phyla information
-#Edited out non phyla information first with sed in linux
-#combined new labels with previous taxonomy file with excel
-phylogenetic.info <- read.csv("phyla.data.csv")
-rownames(phylogenetic.info) <- phylogenetic.info[,1]
-phylogenetic.info <- phylogenetic.info[,-c(1)]
-phyla.names <- as.character(phylogenetic.info$Taxonomy)
-keep <- colnames(Data16S)
-phyla.good <- phylogenetic.info[keep, ]
-phyla.names <- as.character(phyla.good[,2])
-phyla.table <- Data16S
-colnames(phyla.table) <- phyla.names
-#add all the same columns up and then return the sum
-testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
-phyla.table <- as.data.frame(testing)
-rm(testing, keep, microb.rownames, phyla.names)
-
-#combine phyla that are not that abundant
-phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "TM7")], 1, sum)
-phyla.table <- phyla.table[, -c(1, 4, 6, 7, 9, 10:12)]
-
-#Create a relative abundance table for phyla
-phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
-phyla.table.rel.abund <- (phyla.table/phyla.total)*100
-rm(phyla.total)
-
-# Create comparable data set with 
-metaphlanPhlyaData <- read.csv("HMP.metaphlan.phyla.data.csv", header=T)
-rownames(metaphlanPhlyaData) <- metaphlanPhlyaData$ID
-metaphlanPhlyaData <- metaphlanPhlyaData[,-1]
-metaphlanPhlyaData <- as.data.frame(t(metaphlanPhlyaData))
-
-#combine phyla that are not that abundant
-metaphlanPhlyaData$other <- apply(metaphlanPhlyaData[, c("Acidobacteria", "Candidatus_Saccharibacteria", "Deinococcus_Thermus", "Fusobacteria", "Spirochaetes", "Synergistetes", "Tenericutes")], 1, sum)
-metaphlanPhlyaData <- metaphlanPhlyaData[, -c(1, 4:5, 7, 9:11)]
-
-#Create a relative abundance table for phyla
-metaphlanPhlyaData.total <- apply(metaphlanPhlyaData[, c(1:6)], 1, sum)
-metaphlanPhlyaData.rel.abund <- (metaphlanPhlyaData/metaphlanPhlyaData.total)*100
-
-#Need to get sampleIDs to match 16S data
-metaphlanNames <- rownames(metaphlanPhlyaData)
-rownames(IDconversion) <- IDconversion[,1]
-matchedData <- IDconversion[metaphlanNames, c(1,3)]
-IndIDmatched <- matchedData[,2]
-
-#Need to find and remove duplicated values (No idea which sample goes with which)
-test1 <- as.data.frame(table(matchedData$Individual.ID))
-good <- which(test1$Freq == 1)
-unique <- test1[good, ]
-uniqueNames <- as.character(unique$Var1)
-
-test <- phyla.table.rel.abund[rownames(phyla.table.rel.abund) %in% uniqueNames, ]
-finalUniqueNames <- rownames(test)
-UpdatedMatchedData <- matchedData[matchedData$Individual.ID %in% finalUniqueNames, ]
-
-goodMetaphlanPhylaData.rel.abund <- metaphlanPhlyaData.rel.abund[rownames(metaphlanPhlyaData.rel.abund) %in% rownames(UpdatedMatchedData), ]
-rownames(goodMetaphlanPhylaData.rel.abund) <- UpdatedMatchedData[, 2]
-testing <- goodMetaphlanPhylaData.rel.abund[order(rownames(goodMetaphlanPhylaData.rel.abund)), ]
-
-phyla16SData <- test
-phylaMetaphlanData <- testing
-rm(finalUniqueNames, good, IndIDmatched, metaphlanNames, metaphlanPhlyaData.total, uniqueNames, goodMetaphlanPhylaData.rel.abund, matchedData, metaphlanPhlyaData, metaphlanPhlyaData.rel.abund, phyla.good, phyla.table, phyla.table.rel.abund, phylogenetic.info, test, test1, testing, unique, UpdatedMatchedData)
-
-# Can start off with a general bar plot of the two side by side 
-library(qdap)
-library(plyr)
-library(reshape2)
-library(ggplot2)
-library(RColorBrewer)
-library(scales)
-library(gridExtra) #using grid.arrange() can add different plots together
-library(plotrix)
-library(ggExtra)
-library(BlandAltmanLeh)
-
-#phyla16SData <- phyla16SData[,-5] only if want to ignore unclassified
-phylaMetaphlanData$unclassified <- 0
-phylaMetaphlanData <- subset(phylaMetaphlanData, select=c(Actinobacteria:Proteobacteria, unclassified, Verrucomicrobia, other))
-phyla16SData$dataset <- 1 
-phylaMetaphlanData$dataset <- 2
-
-```
-
-**In-Depth Overview of Search Strategy:**
+###In-Depth Overview of Search Strategy:
 
@@
 The second study was excluded because they did not have BMI information available and when contacted the authors never returned any correspondance [@zeevi_personalized_2015].
+A thrid study was excluded since it did not use 16S rRNA gene sequencing for their bacterial microbiome analysis [@arumugam_enterotypes_2011]
 
-This included a total of 10 studies in the qualitative synthesis of the analysis.
+Once these 3 studies were excluded there was a total of 9 studies in the qualitative synthesis of the analysis.
 Because we decided a prioi to use the standard definition for BMI group classification one study from this ten did not have any individuals who were obese by this criteria [@nam_comparative_2011] and was excluded from the final quantitative synthesis and analysis.
@@
 
-* PCR, qPCR, or RT-PCR used as main analysis Tool
+* PCR, qPCR, metagenomic sequencing, or RT-PCR used as main analysis Tool
 * TRFLP or clone sequencing used to asses the bacterial community
@@
 
-**A Comparison of 16S Amplicon Sequencing and Metaphlan2:**
+###Supplemental Results:
 
-```{r Figure_S1, echo=FALSE}
-testset <- rbind(phyla16SData, phylaMetaphlanData)
-testset$dataset <- factor(testset$dataset)
-melt.testset <- melt(testset, id.vars=c("dataset"))
-colnames(melt.testset)[2] <- "phyla"
-barplotAnalysisData <- ddply(melt.testset, c("dataset", "phyla"), summarise, mean = mean(value), sd=sd(value), sem=sd(value/sqrt(length(value))))
+![Figure S1](results/figures/suppFigClostridiales.png)\
 
-bpg1 <- ggplot(barplotAnalysisData, aes(x=dataset, y=mean, fill=phyla)) + geom_bar(position = "fill", stat="identity")
-bpg2 <- bpg1 + scale_fill_brewer(palette="Set1") + scale_y_continuous(label=percent)
-bpg3 <- bpg2 + theme(axis.text=element_text(size=15), axis.title=element_text(size=20, face="bold"))
-bpg4 <- bpg3 + xlab("Analysis Used") + ylab("Mean Relative Abundance (%)")
-bpg5 <- bpg4 + guides(fill=guide_legend(title="Phyla")) + theme(legend.title = element_text(size=16, face="bold"))
-bpg6 <- bpg5 + theme(axis.line=element_line(colour="black"), 
-                 panel.grid.major = element_blank(), 
-                 panel.grid.minor = element_blank(), 
-                 panel.border = element_blank(), 
-                 panel.background = element_blank())
-bpg7 <- bpg6 + scale_x_discrete(breaks = 1:2, labels=c("16S", "Metaphlan2"))
+**Figure S1: Boxplots of All Clostridiales OTUs Important for Non-Obese and Obese Classification.** A total of 6/8 studies had at least 1 OTU that was classified to Clostridiales.
+However, overall there was a wide range in direction both between studies and within studies.
 
-bpg7
+![Figure S2](results/figures/suppFigLachno.png)\
+
+**Figure S2: Boxplots of All Lachnospiraceae OTUs Important for Non-Obese and Obese Classification.** A total of 6/8 studies had at least 1 OTU that was classified to Lachnospiraceae.
+However, overall there was a wide range in direction both between studies and within studies.
+
+![Figure S3](results/figures/suppFigRumi.png)\
+
+**Figure S3: Boxplots of All Ruminococcaceae OTUs Important for Non-Obese and Obese Classification.** A total of 7/8 studies had at least 1 OTU that was classified to Ruminococcaceae  However, overall there was a wide range in direction both between studies and within studies.
+
+**Table 1. Summary of BLAST alignment of Representative OTU for Clostridiales, Lachnospiraceae, and Ruminococcaceae classification OTUs**
+
+```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
+
+library(knitr)
+
+Study <- c("Wu", "Turnbaugh", "Ross", "Escobar", "Zupancic", "Baxter", "Goodrich", "HMP")
+VariableRegion <- c("V1-V2", "V2", "V1-V3", "V1-V3", "V1-V3", "V4", "V4", "V3-V5")
+ClosMostPrev <- c("OTU191:3 Clostridium", "None", "OTU102:8 Ruminococcus", "None", "OTU322:3 Parvimonas | OTU62:3 Parvimonas | OTU256:4 Clostridium", "OTU479:4 Anaerostipes", "OTU287:5 Clostridium | OTU662:8 Clostridium | OTU145:6 Clostridium | OTU244:3 Clostridium | OTU442:4 Clostridium", "OTU293:7 Clostridium")
+ClostHighMatch <- c("Anaerovorax odorimutans(81|95)", "None", "Ruminococcus champanellensis(100|100)", "None", "OTU322:Eubacterium hallii(92|100) | OTU62:Eubacterium hallii(92|99) | OTU256:Romboutsia lituseburensis(92|99)", "Anaerosporbacter mobilis(100|94)", "OTU287:Clostridium fimetarium(100|94) |  OTU662:Ruminiclostridium thermocellum(100|91) | OTU145:Ruminococcus lactaris(100|100) | OTU244:Anaerosporobacter mobilis(100|94) | OTU442:Clostridium propionicum(100|92)", "Anaerobacterium chartisolvens (100|91)")
+LachMostPrev <- c("OTU75:6 Clostridium | OTU229:5 Peptostreptococcus | OTU5: 4 Blautia", "OTU318:", "OTU177:3 Roseburia", "OTU67:6 Eubacterium | OTU74:5 Blautia | OTU80:4 Blautia", "None", "OTU36:8 Clostridium", "OTU180:4 Clostridium | OTU75: 5 Clostridium | OTU122:5 Roseburia | OTU92:8 Clostridium | OTU673:7 Clostridium", "None")
+LachHighMatch <- c("Clostridium bolteae(100|96)", "", "Clostridium Spp(100|98)", "OTU67:Dorea formicigenerans(100|96) | OTU74:Blautia luti(100|96) | OTU80:Blautia producta(100|93)", "None", "Lachnobacterium bovis(100|96)", "OTU180:Hungatella effluvii(100|95) | OTU75:Ruminococcus gnavus(100|100) | OTU122:Eubacterium ramulus(100|96) | OTU92:Fusicatenibacter saccharivorans(100|94) | OTU673:Bacteroides xylanolyticus(92|97) | OTU523:Eubacterium rectale(100|97)", "None")
+RumiMostPrev <- c("OTU123:4 Clostridium", "OTU30:", "OTU22:3 Oscillibacter", "None", "OTU398:3 Eubacterium | OTU850:3 Ethanoligenens | OTU817:8 Ruminococcus", "OTU320:4 Clostridium", "OTU68:5 Ruminococcus | OTU255:3 Blautia | OTU1145:3 Ethanoligenens | OTU558:8 Ruminococcus | OTU297:3 Oscillibacter | OTU6:4 Clostridium | OTU132:2 Ruminoclostridium", "OTU252:10 Clostridium")
+RumiHighMatch <- c("Anaerotruncus colihominis(85|95)", "", "Oscillibacter Spp(100|97)", "None", "OTU398:Oscillibacter valericigene(77|96) | OTU850:Papillibacter cinnamivorans(90|91) | OTU817:Ruminococcus callidus(91|95)", "Sporobacter termitidis(100|94)", "OTU68:Acetanaerobacterium elongatum(100|89) | OTU255:Pseudoflaconifractor capillosus(100|91) | OTU1145:Subdoligranulum variable(100|95) | OTU558:Ruminococcus champanellensis(92|98) | OTU297:Flavonifractor plautii(100|95) | OTU6:Oscillibacter valericigenes(100|93) | OTU132:Intestinimonas butyriciproducens(100|92)", "Clostridium Spp(100|97)")
+
+table1 <- cbind(Study, VariableRegion, ClosMostPrev, ClostHighMatch, LachMostPrev, LachHighMatch, RumiMostPrev, RumiHighMatch)
+colnames(table1) <- c("Study", "Hypervariable Region", "Clostridiales OTU Most Prevalent Match", "Clostridiales OTU Highest Match (Query Cover|Identity)", "Lachnospiraceae OTU Most Prevalent Match", "Lachnospiraceae OTU Highest Match (Query Cover|Identity)", "Ruminococcaceae OTU Most Prevalent Match", "Ruminococcaceae OTU Highest Match (Query Cover|Identity)")
+
+kable(table1, format = "markdown", align = 'c', padding = 2)
 ```
-
-**Figure S1:Comparison of Phyla between 16S Sequencing and Metaphlan2.**
-
-```{r Figure_S2, echo=FALSE, warning=FALSE, message=FALSE}
-graphData <- as.data.frame(cbind(phyla16SData$Bacteroidetes, phylaMetaphlanData$Bacteroidetes))
-colnames(graphData) <- c("rRNA_16S", "Metaphlan2")
-
-p1 <- ggplot(graphData, aes(x=rRNA_16S, y=Metaphlan2)) + geom_point(colour="blue", size=4)
-p2 <- p1 + geom_smooth(method="lm", colour="black", se=TRUE)
-p3 <- p2 + theme(axis.text=element_text(size=15), axis.title=element_text(size=20, face="bold"))
-p4 <- p3 + xlim(0, 100) + xlab("16S") + ylim(0, 100)
-p5 <- p4 + theme(axis.line=element_line(colour="black"), 
-                 panel.grid.major = element_blank(), 
-                 panel.grid.minor = element_blank(), 
-                 panel.border = element_blank(), 
-                 panel.background = element_blank())
-
-ggExtra::ggMarginal(p5, type = "histogram")
-
-```
-
-**Figure S2: Bacteroidetes Linear Correlation between 16S Sequencing and Metaphlan2.**  Bacteroidetes has the best correlation of all the phyla (P-value = 0.0002277, R^2^ = 0.2358).
-The histogram for the 16S samples are on the x-axis while the histogram for Metaphlan2 is on the y-axis.
-
-```{r Figure_S3, echo=FALSE}
-
-ba <- bland.altman.plot(phyla16SData$Bacteroidetes, phylaMetaphlanData$Bacteroidetes, graph.sys="ggplot2", main="Bland Altman Plot of Bacteroidetes", xlab="Means", ylab="Differences")
-
-ba1 <- ba + theme(axis.text=element_text(size=15), axis.title=element_text(size=25, face="bold"))
-ba2 <- ba1 + ylab("Difference") + xlab("Mean of Measurements")
-ba3 <- ba2 + geom_point(size=4)
-ba4 <- ba3 + theme(axis.line=element_line(colour="black"), 
-                     panel.grid.major = element_blank(), 
-                     panel.grid.minor = element_blank(), 
-                     panel.border = element_blank(), 
-                     panel.background = element_blank())
-
-ba4
-```
-
-**Figure S3: Bland-Altman plot of Bacteroidetes measurements for Metaphlan2 and 16S sequencing.**
 

### 2016-03-20 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
Manuscript power and sequence sim update

--- 2016-03-17 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
@@
 PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
-
+ModelledPowerH <- read.csv("results/tables/denovoModelledPowerH.csv")
+ModelledPowerBF <- read.csv("results/tables/denovoModelledPowerBF.csv")
 ```
@@
 
-Study <- as.character(PowerTable$Study)
-PowerH <- PowerTable$StudyPowerH
-PowerBF <- PowerTable$StudyPowerBF
+Study <- c(as.character(PowerTable$Study))
+PowerH <- c(PowerTable$StudyPowerH)
+GroupsOposH <- c(ShannonRRTable$tposH + ShannonRRTable$cposH)
+GroupsNonOposH <- c(ShannonRRTable$tnegH + ShannonRRTable$cnegH)
+GroupsH <- paste(GroupsOposH, "|", GroupsNonOposH, sep="")
 
-table4 <- cbind(Study, PowerH, PowerBF)
-colnames(table4) <- c("Study", "Shannon Diversity Power (%)", "B/F Ratio Power (%)")
+PowerHRR <- c(PowerTable$StudyPowerHRR)
+GroupsLowH <- c(format(round(ShannonRRTable$tposH / (ShannonRRTable$tposH + ShannonRRTable$tnegH), 1), nsmall=1))
+GroupsHighH <- c(format(round(ShannonRRTable$cposH / (ShannonRRTable$cposH + ShannonRRTable$cnegH), 1), nsmall=1))
+GroupsHRR <- paste(GroupsLowH, "|", GroupsHighH, sep="")
+
+PowerBF <- c(PowerTable$StudyPowerBF)
+GroupsOposBF <- c(BFRatioRRTable$tposBF + BFRatioRRTable$cposBF)
+GroupsNonOposBF <- c(BFRatioRRTable$tnegBF + BFRatioRRTable$cnegBF)
+GroupsBF <- paste(GroupsOposBF, "|", GroupsNonOposBF, sep="")
+
+PowerBFRR <-c(PowerTable$StudyPowerBFRR)
+GroupsLowBF <- c(format(round(BFRatioRRTable$tposBF / (BFRatioRRTable$tposBF + BFRatioRRTable$tnegBF), 1), nsmall=1))
+GroupsHighBF <- c(format(round(BFRatioRRTable$cposBF / (BFRatioRRTable$cposBF + BFRatioRRTable$cnegBF), 1), nsmall=1))
+GroupsBFRR <- paste(GroupsLowBF, "|", GroupsHighBF, sep="")
+
+table4 <- cbind(Study, PowerH, GroupsH, PowerHRR, GroupsHRR, PowerBF, GroupsBF, PowerBFRR, GroupsBFRR)
+colnames(table4) <- c("Study", "Shannon Diversity Power (%)", "Group n (Obese|Non-Obese)", "Shannon RR Power (%)", "Group Proportion (Low|High)", "B/F Ratio Power (%)", "Group n (Obese|Non-Obese)", "B/F ratio RR (%)", "Group Proportion (Low|High)")
 
 kable(table4, format = "markdown", align = 'c', padding = 2)
+```
+
+**Table 5. Summary of Power Calculations Simulations for Shannon Diversity for Non-obese versus Obese**
+
+```{r Table_5, echo=FALSE, message=FALSE, warning=FALSE}
+
+table5 <- ModelledPowerH[, -1]
+
+colnames(table5) <- c("Estimated Power for 1% Difference", "Estimated Power for 5% Difference", "Estimated Power for 10% Difference", "Estimated Power for 15% Difference", "Actual % Difference", "Study")
+
+kable(table5, format = "markdown", align = 'c', padding = 2)
+
+```
+
+**Table 6. Summary of Power Calculations Simulations for B/F ratio for Non-obese versus Obese**
+
+```{r Table_6, echo=FALSE, message=FALSE, warning=FALSE}
+
+table6 <- ModelledPowerBF[, -1]
+
+colnames(table6) <- c("Estimated Power for 1% Difference", "Estimated Power for 5% Difference", "Estimated Power for 10% Difference", "Estimated Power for 15% Difference", "Actual % Difference", "Study")
+
+kable(table6, format = "markdown", align = 'c', padding = 2)
 ```

### 2016-03-20 microbiomeAndobesitySupplemental.Rmd
Manuscript power and sequence sim update

@@
 
-```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Table_S1, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
+**Table S2. Top Sequence Similarity by Variable Region Based on the Representative OTU for Clostridiales, Lachnospiraceae, and Ruminococcaceae**
+
+```{r Table_S2, echo=FALSE, message=FALSE, warning=FALSE}
+
+VariableRegion <- c("V1-V2 vs V2", "V1-V3", "V4")
+
+TopHitClos <- c("None", "45.2 | 45.2 | 46.5", "60.9 | 60.9 | 32.8")
+AverageClos <- c("None", "34.6 | 34.6 | 59.4", "58.9 | 58.9 | 33.2")
+TopHitLachno <- c("TBC", "40.7 | 40.7 | 50", "61.3 | 61.3 | 32.8")
+AverageLachno <- c("TBC", "37.9 | 37.9 | 52.1", "59.4 | 59.4 | 33.1")
+TopHitRumi <- c("TBC", "15.5 | 15.5 | 84.1", "60.5 | 60.5 | 32.8")
+AverageRumi <- c("TBC", "15.2 | 15.2 | 84.1", "57.5 | 57.5 | 33.2")
+
+table2 <- cbind(VariableRegion, TopHitClos, AverageClos, TopHitLachno, AverageLachno, TopHitRumi, AverageRumi)
+colnames(table2) <- c("Variable Region", "Top Similarity Colstridiales (Id | Sim | Gaps)", "Average Similarity Colstridiales (Id | Sim | Gaps)", "Top Similarity Lachnospiraceae (Id | Sim | Gaps)", "Average Similarity Lachnospiraceae (Id | Sim | Gaps)", "Top Similarity Ruminococcaceae (Id | Sim | Gaps)", "Average Similarity Ruminococcaceae (Id | Sim | Gaps)")
+
+kable(table2, format = "markdown", align = 'c', padding = 2)
+
+```
+
 *******

### 2016-03-20 code/PowerVariations.R
Power code update

@@ -280,7 +280,7 @@ ModelledPowerH <- as.data.frame(
     ActualDiff, 1), nsmall=1)))
 colnames(ModelledPowerH) <- c("Pow1Diff", "Pow5Diff", "Pow10Diff", "Pow15Diff", "ActualPerDiff")
 ModelledPowerH$Study <- Study
-
+write.csv(ModelledPowerH, "results/tables/denovoModelledPowerH.csv")
 
 ## Generate Estimates for BF Ratio
 
@@ -553,7 +553,7 @@ ModelledPowerBF <- as.data.frame(
     ActualDiff, 1), nsmall=1)))
 colnames(ModelledPowerBF) <- c("Pow1Diff", "Pow5Diff", "Pow10Diff", "Pow15Diff", "ActualPerDiff")
 ModelledPowerBF$Study <- Study
-
+write.csv(ModelledPowerBF, "results/tables/denovoModelledPowerBF.csv")
 
 #######################################################################################
 

### 2016-03-20 code/StrategydenovoAnalysisMetaHitDel.R
main code update add pow

@@ -60,6 +60,8 @@ obese <- factor(demographics$obese.num)
 baxterH <- wilcox.test(alpha.test$H ~ obese) #P-value=0.065
 MeanNonObeseH <- mean(alpha.test$H[which(demographics$obese == "No")])
 MeanObeseH <- mean(alpha.test$H[which(demographics$obese == "Yes")])
+SDNonObeseH <- sd(alpha.test$H[which(demographics$obese == "No")])
+SDObeseH <- sd(alpha.test$H[which(demographics$obese == "Yes")])
 baxterS <- wilcox.test(alpha.test$S ~ obese) #P-value=0.03923
 baxterJ <- wilcox.test(alpha.test$J ~ obese) #P-value=0.1254
 
@@ -71,7 +73,8 @@ averageStudyBF <- mean(BFratio)
 sdBF <- sd(BFratio)
 MeanNonObeseBF <- mean(BFratio[which(demographics$obese == "No")])
 MeanObeseBF <- mean(BFratio[which(demographics$obese == "Yes")])
-
+SDNonObeseBF <- sd(BFratio[which(demographics$obese == "No")])
+SDObeseBF <- sd(BFratio[which(demographics$obese == "Yes")])
 baxterBacter <- wilcox.test(bacter ~ obese) #P-value=0.3175
 baxterFirm <- wilcox.test(firm ~ obese) #P-value=0.6817
 baxterBF <- wilcox.test(BFratio ~ obese) #P-value=0.6305
@@ -257,6 +260,11 @@ MeanNonObeseH <- c(MeanNonObeseH,
                    mean(alpha.test$H[which(metadata$obese == "No")]))
 MeanObeseH <- c(MeanObeseH, 
                 mean(alpha.test$H[which(metadata$obese == "Yes")]))
+SDNonObeseH <- c(SDNonObeseH, 
+                 sd(alpha.test$H[which(metadata$obese == "No")]))
+SDObeseH <- c(SDObeseH, 
+              sd(alpha.test$H[which(metadata$obese == "Yes")]))
+
 rossS <- wilcox.test(alpha.test$S ~ obese) #P-value=0.2492
 rossJ <- wilcox.test(alpha.test$J ~ obese) #P-value=0.3826
 
@@ -265,6 +273,12 @@ bacter <- phyla.table.rel.abund$Bacteroidetes
 firm <- phyla.table.rel.abund$Firmicutes
 BFratio <- bacter/firm
 averageStudyBF <- c(averageStudyBF, mean(BFratio))
[…]

### 2016-03-20 code/UsedFunctions.R

@@ -191,6 +191,20 @@ DefNonParaPowerSim <- function(mean1, mean2, n1, n2, sdselect, iters=1000){
   
 }
 
+
+## Re write of Function to get estimated Power based on user inputed means and sd
+## to better simulate the actual data (add ability for two differet
+## SDs to be entered.
+## Example on how to use:
+  ## DefNonParaPowerSim(3.0, 3.35, 20, 50, 0.2, 0.2, n=1000)
+DefNonParaPowerSim2 <- function(mean1, mean2, n1, n2, sd1, sd2, iters=1000){
+  set.seed(3)
+  PvalSim <- replicate(iters, wilcox.test(
+    rnorm(n1, mean1, sd1), rnorm(n2, mean2, sd2))$p.value)
+  return((sum(PvalSim < 0.05)) / 10)
+  
+}
+
 ## Function to graph OTU versus BMI linear regression with log scale option
 ## Default for log is FALSE. Necessary library is ggplot2. 
 ## Example on how to use:

### 2016-03-21 code/sequenceProcessing.Rmd
Update sequence processing Rmd

--- 2016-02-12 code/sequenceProcessing.Rmd
@@
     keep_md: yes
-bibliography: metaAnalysis.obesity.microbiome.bib
+bibliography: ../metaAnalysis.obesity.microbiome.bib
 ---
@@
 This recount starts from time of download all the way to generation of the final shared file used for the meta-analysis in the study.
-You will also need the latest version of mothur, SRA Toolkit, Metaphlan2, Python 2.7, and Python 3+ installed on your system.
+You will also need the latest version of mothur, SRA Toolkit, and Python 3+ installed on your system.
 There are also a few reference files that you will need.
 The first is the [SILVA reference](http://www.mothur.org/wiki/Silva_reference_files) database and the second is the [RDP database](http://www.mothur.org/wiki/RDP_reference_files) of which we used release 102 and version 9 respectively.
-In general, we tried to keep close to the Schloss SOP for processsing sequence reads.
-However, where this was not possible (mostly due to a high loss of reads resulting in a high loss of samples) we defaulted to the processing used by the published authors.
-Where it was not possible to use a *de novo* approach (mostly due to high numbers of unique sequences) and the authors used a closed-reference approach we used the RDP database instead of greengenes.
-
+In general, we tried to keep as close as possible to the Schloss SOP for processsing sequence reads for either 454 or the MiSeq.
+Most of the differences between the SOP as written and what was performed involved the use of flowgrams.
+Most of the data in the SRA were stored in the Fastq format and this step was not possible.
 <br>
@@
 ```{r, engine='bash', eval=FALSE}
-/share/scratch/schloss/mothur/mothur data1.batch
+mnt/EXT/Schloss-data/bin/mothur Baxter.batch
 ```
-
-If you thought this was difficult to get data just wait until later on in this pipeline to see just how crazy it can get trying to get the data to a useable point.
-<br>
-<br>
 
@@
 We can download the entire data set again with the code mentioned above but with one small substitution at the ending.
-Here we go......
 
@@
 do  
-  /share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"  
+  mnt/EXT/Schloss-data/bin/mothur "#fastq.info(fastq=$sample2)"  
   rm *logfile  
@@
   QFILE=${TRIM//fasta/qual}  
-  /share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=5, maxhomop=8, qwindowaverage=25, qwindowsize=50, processors=8)"  
+  mnt/EXT/Schloss-data/bin/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=5, maxhomop=8, qwindowaverage=25, qwindowsize=50, processors=8)"  
 done
@@
 cat \*trim.fasta > combined.trim.fasta  
-/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"
+mnt/EXT/Schloss-data/bin/mothur "#unique.seqs(fasta=combined.trim.fasta)"
 
@@
 
-After this little program is run we follow the mothur 454 SOP as written and can use the following command to complete the processing using the put together *Ross.batch* file and running mothur.
-One final difference from the standardized mothur approach that was made was during the screen.seqs command after alignment a hard cutoff *minlength=200* and *maxlength=250* was used.
+After this little program is run we follow the mothur 454 SOP pretty much as  written and can use the following command to complete the processing using the put together *Ross.batch* file and running mothur.
+Moving forward for the screen.seqs command we use a minlength cutoff of 200 in addition to what is written in the 454 SOP for all data sets.
 
 ```{r, engine='bash', eval=FALSE}
-/share/scratch/schloss/mothur/mothur Ross.batch
+mnt/EXT/Schloss-data/bin/mothur Ross.batch
 ```
@@
 ##### **Goodrich et al. 2014:**  
-This is a second larger twin data set that builds on the Turnbaugh et al.
-findings.
-The paper can be found [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4255478/).
+This is a second larger twin data set and the paper can be found [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4255478/).
 The individuals that participated in the study were part of the TwinsUK registry and is somewhat heavily skewed toward women.
@@
 # mkdir sra.complete  
-for start in \$(ls \*sra)  
+for start in $(ls *sra)  
 do  
@@
 
-# Unpack fastq files  
+#Unpack fastq files  
 mkdir fastq.complete  
@@
 do  
-  /share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$info)"  
+	mnt/EXT/Schloss-data/bin/mothur "#fastq.info(fastq=$info)"  
 	mv $info fastq.complete  
@@
 
-# Screen the fasta file that came from the fastq  
-mkdir fasta.complete  
+#Trim the fasta file that came from the fastq  
+mkdir fasta.trimming.complete  
 for next in $(ls *fasta)  
 do  
-  /share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$next, maxambig=0, maxlength=275, processors=8)"  
-  mv $next fasta.complete # should store silva file in here and then move it to working directory  
+	QualFile=${next//fasta/qual}
+	/mnt/EXT/Schloss-data/bin/mothur "#trim.seqs(fasta=$next, qfile=$QualFile, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"  
+	mv $next fasta.trimming.complete  
+	mv $QualFile fasta.trimming.complete
 	mv *.logfile logfile.complete  
@@
 
-mv ~/twinsUK/fasta.complete/silva.v4.fasta ~/twinsUK  
-# Align the screened fasta file to V4 region of SILVA  
-mkdir good.fasta.complete  
-for sample2 in $(ls *good.fasta)  
+#Create a groups file for each sample
+for sample2 in $(ls *trim.fasta)
 do
-  /share/scratch/schloss/mothur/mothur "#align.seqs(fasta=$sample2, reference=silva.v4.fasta, processors=8)"  
-  mv $sample2 good.fasta.complete  
-  mv *logfile logfile.complete  
+	groupfile=${sample2//fasta/groups}
+	grep -e '^>' $sample2 > test.fasta  
+	sed 's/>//' test.fasta > test2.fasta  
+	sed 's/\..*$//' test2.fasta > test3.fasta  
+	paste -d"\t" test2.fasta test3.fasta > ${groupfile}
+
 done
 
-#Screen the aligned sequences to specific location  
-mkdir align.complete  
-for sample3 in $(ls *.align)  
-do  
-  /share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$sample3, start=1968, end=11546, maxhomop=8, processors=8)"  
-  mv $sample3 align.complete  
-done
-
-# Filter the screened sequences to get a fasta file back  
-mkdir align.good.complete  
-for sample4 in $(ls *good.align)  
-do  
-  /share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=$sample4, vertical=T, trump=., processors=8)"  
-  mv $sample4 align.good.complete  
-  mv *logfile logfile.complete  
-done
-
-# Make the combined fasta file  
-mkdir filter.fasta.complete  
-cat *filter.fasta > combined.fasta  
-mv *filter.fasta filter.fasta.complete
-
-# Make the combined group file  
-grep -e '^>' combined.fasta > test.fasta  
-sed 's/>//' test.fasta > test2.fasta  
-sed 's/\..*$//' test2.fasta > test3.fasta  
-paste -d"\t" test2.fasta test3.fasta > combined.groups
-
+#Make the combined fasta file  
+mkdir combined.complete  
+cat *trim.fasta > combined.fasta  
+mv *trim.fasta combined.complete
+cat *trim.groups > combined.groups
+mv *trim.groups combined.complete
 ```
 
-After this program was run I thought perhaps naively that I could simply run a set a batch file and be done with this data set.
-Well not quite.
-The sheer size and the way our server was configured meant that if I called much more than 130GB of RAM my job would run the risk of starting over regularly.
-So I did use the batch method but only upto *pre.cluster*.
-After this the *chimera.uchime* step was just taking too long so I seperated everything again and did the *chimera.uchime* separately.
-This also meant having to redo the unique seqs step.
-The code for this is below:
+From here you can run a batch file with mothur to get the final output files.
 
@@
 /share/scratch/schloss/mothur/mothur Goodrich.batch
-```
-
-So now the biggest challenge that I have found so far is getting the data into a distance matrix not based on the closed-reference approach.
-The file is just too big for cluster.split so we have had to use the closed-reference approach for this data set.
-
-```{r, engine='bash', eval=FALSE}
-/share/scratch/schloss/mothur/mothur Goodrich2.batch
 ```
@@
 We can download the entire data set again with the code mentioned above but with a few substitutions versus the previous two studies.
-Here we go......
 
@@
 Unlike the Ross, et al study we used the mothur specified defaults of *qwindowaverage=35* and *qwindowsize=50*.
-It should also be noted that we had *maxambig=0* and *maxhomop=8*.
 Since the data come seperated already we do not need an oligo file and all we have to do is loop the mothur *trim.seqs* command.
@@
   QFILE=\${TRIM//fasta/qual}  
-/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50, processors=8)"  
+mnt/EXT/Schloss-data/bin/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50, processors=8)"  
 done
@@
 
-/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"
+mnt/EXT/Schloss-data/bin/mothur "#unique.seqs(fasta=combined.trim.fasta)"
 
@@
 ```{r, engine='bash', eval=FALSE}
-/share/scratch/schloss/mothur/mothur Escobar.batch
+mnt/EXT/Schloss-data/bin/mothur Escobar.batch
 ```
@@
 Now this data set included the sff files so we can use sffinfo function and start from the very start of the mothur 454 pipleine.
-Plus one for this data set so far.
+However, to keep things consistent we will keep with the fastq approach that we have been using from the previous data sets.
 The below program was used and you can use either *#!/bin/sh* or *#!/bin/bash* and text editor to create it and then run it.
-Also you should be aware that this part of the processing will take a LONG time to get through.
 
@@
 # Make a directory to store the reference files
+# Remember to move these back into the necessary directory later
+# When you get to the batch file point,
 mkdir reference.files
@@
 do
-	/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"	
+	mnt/EXT/Schloss-data/bin/mothur "#fastq.info(fastq=$sample2)"	
 	mv $sample2 completed.fastq
@@
 	Quality=${Trimming//fasta/qual}
-	/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
+	mnt/EXT/Schloss-data/bin/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
 	mv $Trimming first.trim.completed
@@
 ```{r, engine='bash', eval=FALSE}
-/share/scratch/schloss/mothur/mothur Zupancic.batch
+mnt/EXT/Schloss-data/bin/mothur Zupancic.batch
 ```
@@
 With that you should now be able to generate the subsample shared file that was used for the downstream analysis.
-This was a headache of a data set to work with.
-I think the next one will be better.
-At least I hope so....
-<br><br>
-
-##### **Yatsunenko et al. 2012:**  
-This data set was another famous one looking at the microbiome of indivdiuals at different geographical locations and pretty different diets.
-The link to the paper is [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3376388/).
-A couple problems from a data processing and sequence clean up point of view is that they used a HiSeq, single end, 100bp total length, sequencing run.
-This wreaked absolute havoc with a lot of the downstream steps mostly with trying to figure out a way to reduce the artificial diversity within the files.
-Oh...
-and the data was not on a public database but stored on MG-RAST.
-I had a lot trouble downloading the data on command line all at once (maybe I'm just not smart enough to do it).
-So I had to manually download the sequence files.Thats right I am just a glutton for punishment.
-After you have downloaded the files the first part of the code is:
-
-```{r, engine='bash', eval=FALSE}
-#!/bin/sh  
-mkdir fna.done  
-mkdir logfiles.done  
-mkdir bad.accnos.done
-
-# Algin the sequences
-mkdir good.fna.done  
-for sample in $(ls *.fna)  
-do  
-  /share/scratch/schloss/mothur/mothur "#align.seqs(fasta=\$sample, reference=silva.bacteria.fasta, processors=8)"  
-  mv $sample good.fna.done  
-  mv *logfile logfiles.done  
-done
-
-```
-
-For the next code it might be best to just double check that the start and end are what is written here.
-To make it go faster I made a custom pcr.seqs based on the start and end of the sequences on the aligned file.
-It shouldn't make a difference but.....you never know.
-
-```{r, engine='bash', eval=FALSE}
-#!/bin/sh
-
-# Screen the sequences to the same start and end
-mkdir align.done  
-for sample2 in $(ls *.align)  
-do  
-  /share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$sample2, start=8, end=2448, maxhomop=8, processors=8)"  
-  mv $sample2 align.done  
-  mv *logfile logfiles.done  
-done
-
-# Filter the sequences
-mkdir good.align.done  
-for sample3 in \$(ls \*good.align)  
-do  
-  /share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=\$sample3, vertical=T, trump=., processors=8)"  
-  mv \$sample3 good.align.done  
-  mv \*logfile logfiles.done  
-done
-
-# Create a groups file and combine the fasta and groups together
-mkdir completed.good.groups  
-for sample in \$(ls \*.good.filter.fasta)  
-do  
-  grep -e '^>' \$sample > test.fasta  
-  sed 's/>//' test.fasta > test2.fasta  
-  sed 's/_.*//' test2.fasta > test3.fasta  
-  paste -d"\t" test2.fasta test3.fasta > \${sample}.groups
-
-  rm test.fasta  
-  rm test2.fasta  
-  rm test3.fasta  
-  mv ~/Yatsunenko.Nature/${sample}.groups  ~/Yatsunenko.Nature/completed.good.groups
-
-done
-```
-
-From here I thought I might be able to get away with combining the data set and running everything together.
-Well that was not the case since it is just too large for the memory I have readily available without having the job restarting.
-The plan is to seperate everything back out again and run the *chimera.uchime* function.
-Then combine them and run the *pre.cluster* function.
-The code below shows what was used for this part:
-
-```{r, engine='bash', eval=FALSE}
-#!/bin/sh  
-cat *fasta > combined.fasta  
-cat *groups > combined.groups
-
-/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.fasta)"    
-/share/scratch/schloss/mothur/mothur Yatsunenko.batch
-
-done
-```
-
 <br><br>
@@
 The one exception to the workflow provided in the notebook is that you will need to make sure that you do not use a closed reference data base approach for OTU clustering and instead use an open reference approach.
-<br><br>
-
-##### **Nam, et al. 2011:**  
-This study recruited 20 individuals that were from the area surronding Kyung Hee University in South Korea and can be found [here](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0022109).
-The data from this project can be found under the accession number PRJDA6057 and we downloaded the data from DRR000776.
-There is some small difference to the wget, which is mostly associated with the directory where the data is stored.
-From here you can run the SRA tool kit command to get the sff file.
-Once this is done you can then run mothur with the batch file and get the sub sample shared file that was used in this analysis.
-We have also provided the oligo file in this repository *korean6.oligos* that was used so that you can easily use it on this data set.
-
-```{r, engine='bash', eval=FALSE}
-wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByRun/sra/DRR/DRR000/DRR000776
-
-sff-dump DRR000776.sra
-
-/share/scratch/schloss/mothur/mothur Nam.batch''
-
-```
-
-Compared to the previous data sets this was one of the most straight forward to actually process.
+We used the default approach listed rather than repeat the whole analysis again.
 <br><br>
@@
 
-# Trim the fastas according to quality 
-mkdir first.trim.completed
-for Trimming in $(ls *fasta)
-do
-	Quality=${Trimming//fasta/qual}
-	/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
-	mv $Trimming first.trim.completed
-	mv $Quality first.trim.completed
-	mv *logfile logfile.completed
-done
-
-mkdir scrap.fasta.completed
-mkdir scrap.qual.completed
-mkdir trim.qual.completed
-mv *scrap.qual scrap.qual.completed
-mv *scrap.fasta scrap.fasta.completed
-mv *trim.qual trim.qual.completed
+# Combine the qaulity and fasta files together
+cat *fasta > combined.fasta
+cat *qual > combined.qual
 
@@
 
-So for the next part I needed to make a group file so I renamed the fasta file to represent the sampleID label so that I could with more ease create the labels I needed within the group file.
-The code for it is....
+After you have done this you can run a batch file with mothur and that should generate the subsample shared file that we used for this anlaysis.
+It should be noted that this data set came with the barcodes and they needed to be removed so this differed form the previous data sets in this regard.
+However, this is all taken care of in the batch file.
 
@@
 
-groupN="SRR327637 SRR327638 SRR327639 SRR327640 SRR327641 SRR327642 SRR327643 SRR327644 SRR327645 SRR327646 SRR327647 SRR327648 SRR327649 SRR327650 SRR327651 SRR327652 SRR327653 SRR327654 SRR327655 SRR327656 SRR327657 SRR327658 SRR327659 SRR327660 SRR327661 SRR327662 SRR327663 SRR327664 SRR327665 SRR327666 SRR327667 SRR327668 SRR327669 SRR327670 SRR327671 SRR327672 SRR327673 SRR327674 SRR327675 SRR327676 SRR327677 SRR327678 SRR327679 SRR327680 SRR327681 SRR327682 SRR327683 SRR327684 SRR327685 SRR327686 SRR327687 SRR327688 SRR327689 SRR327690 SRR327691 SRR327692 SRR327693 SRR327694 SRR327695 SRR327696 SRR327697 SRR327698 SRR327699 SRR327700 SRR327701"
-
-for V in ${groupN}
-do
-	grep -e '^>' ${V}.trim.fasta > test.fasta
-	sed 's/>//' test.fasta > test2.fasta
-	sed "s/$/\t$V/" test2.fasta > ${V}.groups
-	rm test.fasta
-	rm test2.fasta
-done
-
-cat *fasta > combined.fasta
-cat *groups > combined.groups
-
+mnt/EXT/Schloss-data/bin/mothur Wu.batch
 ```
-
-After you have done this you can run a batch file with mothur and that should generate the subsample shared file that we used for this anlaysis.
-
-```{r, engine='bash', eval=FALSE}
-/share/scratch/schloss/mothur/mothur Wu.batch
-```
-<br><br>
-
-##### **Arumugam, et al. 2010 (MetaHit):**  
-We only used the Danish individuals for this study and left the others alone since many had IBD or other disease conditions.
-For this data set Metaphlan2 was used to generate the abundance data.
-Metaphlan2 and how well it associates with 16S data is also discussed in the supplemental of the draft manuscript we are writing.
-To test this we downloaded ~40 samples from the HMP data set which had both metagenome and 16S data.
-We ran the same code that we used with the MetaHit data to generate the abundance data using Metaphlan2 and compared that result with the obtained 16S values.
-So here is code that was used below:
-
-```{r, engine='bash', eval=FALSE}
-#!/bin/sh
-
-PATH=\$PATH:~/metaphlan2  
-export mpa_dir=~/metaphlan2  
-mkdir completed.combined.fq
-
-test_samples="MH0001 MH0002 MH0003 MH0004 MH0005 MH0006 MH0007 MH0008 
-MH0009 MH0010 MH0011 MH0012 MH0013 MH0014 MH0015 MH0016 MH0017 MH0018 
-MH0019 MH0020 MH0021 MH0022 MH0023 MH0024 MH0025 MH0026 MH0027 MH0028 
-MH0030 MH0031 MH0032 MH0033 MH0034 MH0035 MH0036 MH0037 MH0038 MH0039 
-MH0040 MH0041 MH0042 MH0043 MH0044 MH0045 MH0046 MH0047 MH0048 MH0049 
-MH0050 MH0051 MH0052 MH0053 MH0054 MH0055 MH0056 MH0057 MH0058 MH0059 
-MH0060 MH0061 MH0062 MH0063 MH0064 MH0065 MH0066 MH0067 MH0068 MH0069 
-MH0070 MH0071 MH0072 MH0073 MH0074 MH0075 MH0076 MH0077 MH0078 MH0079 
-MH0080 MH0081 MH0082 MH0083 MH0084 MH0085 MH0086"
-
-for s in ${test_samples}  
-do  
-  cat ${s}*.1.fq > ${s}.combined.1.fq  
-  cat ${s}*.2.fq > ${s}.combined.2.fq  
-  mv ${s}*combined.1.fq completed.combined.fq  
-  mv ${s}*combined.2.fq completed.combined.fq  
-done
-
-```
-
-This code was used to combine the fastq files since there are multiple files per sample.  To actually implement the metaphlan2 program for these files see below:
-
-```{r, engine='bash', eval=FALSE}
-for s in ${test_samples}  
-do  
-  metaphlan2.py /mnt/EXT/Schloss-data/msze/test/danish.obesity.metagenome/raw.data.sep/completed.combined.fq/${s}.combined.1.fq,/mnt/EXT/Schloss-data/msze/test/danish.obesity.metagenome/raw.data.sep/completed.combined.fq/${s}.combined.2.fq --mpa_pkl db_v20/mpa_v20_m200.pkl --bowtie2db db_v20/mpa_v20_m200 --input_type fastq --bowtie2out profiled_samples/${s}.bowtie2.bz2 -o profiled_samples/${s}.txt
-
-done
-```
-
-In order to combine all the files into a single data table the following code was run:
-```{r, engine='bash', eval=FALSE}
-python ~/metaphlan2/utils/merge_metaphlan_tables.py profiled_samples*.txt > profiled_samples/output/merged_abundance_table.txt
-```
-
-For the 16S and metaphlan2 comparison the following code was run to obtain the necessary data tables:
-
-```{r, engine='bash', eval=FALSE}
-#!/bin/sh  
-PATH=\$PATH:~/metaphlan2  
-export mpa_dir=~/metaphlan2
-
-test_samples="SRS012849 SRS012969 SRS013098 SRS013216 SRS013638 SRS013639 SRS016437 SRS016438 SRS016990 SRS017916 SRS020622 SRS021153 SRS021219 SRS022093 SRS024663 SRS043667 SRS045528 SRS045739 SRS049402 SRS049896 SRS050026 SRS053356 SRS053573 SRS053649 SRS054352 SRS055533 SRS056273 SRS058070 SRS063489 SRS063518 SRS064973 SRS074670 SRS074964 SRS075078 SRS075341 SRS075963 SRS076929 SRS077086 SRS077194 SRS077294 SRS077335 SRS077502 SRS077552 SRS077753 SRS077849 SRS078242 SRS078419 SRS078665 SRS097889 SRS098514 SRS098571 SRS098644 SRS098717 SRS098827 SRS0100021 SRS101376 SRS101433 SRS103987 SRS0104197 SRS104311 SRS104400 SRS104485 SRS105153 SRS140492 SRS140513 SRS140645 SRS142503 SRS142505 SRS142599 SRS142712 SRS142890 SRS143070 SRS143181 SRS143342 SRS143417 SRS143598 SRS143780 SRS143876 SRS143991 SRS144362 SRS144506 SRS144537 SRS145497 SRS146764 SRS146812 SRS146813 SRS147022 SRS147088 SRS147139 SRS147346 SRS147445 SRS147652 SRS147766 SRS147919 SRS148196 SRS148424 SRS148721"
-
-for s in ${test_samples}  
-do  
-#Download and unpack necessary files  
-  wget -r -np -nd -k ftp://public-ftp.hmpdacc.org/Illumina/PHASEII/stool/${s}.tar.bz2  
-tar xvjf ${s}.tar.bz2  
-  mv ${s}/${s}.denovo_duplicates_marked.trimmed.1.fastq ~/metaphlan2  
-  mv ${s}/${s}.denovo_duplicates_marked.trimmed.2.fastq ~/metaphlan2  
-  metaphlan2.py ${s}.denovo_duplicates_marked.trimmed.1.fastq,${s}.denovo_duplicates_marked.trimmed.2.fastq --mpa_pkl db_v20/mpa_v20_m200.pkl --bowtie2db db_v20/mpa_v20_m200 --input_type fastq --bowtie2out profiled_samples/${s}.bowtie2.bz2 -o profiled_samples/${s}.txt
-  rm -r ${s}  
-  rm ${s}.denovo_duplicates_marked.trimmed.1.fastq  
-  rm ${s}.denovo_duplicates_marked.trimmed.2.fastq  
-done
-```
-
-You can then go into the directory with all the seperate data tables and combined them into one with the following command:
-
-```{r, engine='bash', eval=FALSE}
-python ~/metaphlan2/utils/merge_metaphlan_tables.py profiled_samples*.txt > profiled_samples/output/merged_abundance_table.txt
-```
-
-Now I decided, perhaps foolishly, to manually curate the combined file.
-What I mean by this is that I manually deleted viruses and eukaryotic cells that were identified instead of using a computer to do this.
-I also then manually created data tables with phyla and the lowest taxonomic identification.
-The data table with the lowest taxonomic identification information was used to generate alpha diversity measures and to complete and PERMANOVA analysis that was used.
-These manually curated files are provided in this repository for you to use as you want.
-Finally, I used relative abundance instead of subsampling in these samples.
-So I had to redo the relative abundance after removing the viruses and eukaryotic cells.
-This part was done in R.
-<br><br>
 
@@
 We used the V2 region.
-For the most part it was relatively straight forward, just be aware that this took a lot of RAM to create a distance matrix.
-The code I used is below:
+For the most part it was relatively straight forward.
+The code used and instruction are below:
 
 ```{r, engine='bash', eval=FALSE}
-grep -e '^>' combined.trim.fasta > test.fasta  
+grep -e '^>' V2.fasta > test.fasta  
 sed 's/>//' test.fasta > test2.fasta    
@@
 ```{r, engine='bash', eval=FALSE}
-/share/scratch/schloss/mothur/mothur Turnbaugh.batch
+mnt/EXT/Schloss-data/bin/mothur Turnbaugh.batch
 ```

### 2016-03-23 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
manuscript and Suppl update

--- 2016-03-17 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
@@
 PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
+ModelledPower1PerDiff <- read.csv("results/tables/denovoModelled1PerDiff.csv")
+ModelledPower5PerDiff <- read.csv("results/tables/denovoModelled5PerDiff.csv")
+ModelledPower10PerDiff <- read.csv("results/tables/denovoModelled10PerDiff.csv")
+ModelledPower15PerDiff <- read.csv("results/tables/denovoModelled15PerDiff.csv")
+ActualDiff <- read.csv("results/tables/denovoActualDiff.csv")
+PowerModelHRR <- read.csv("results/tables/denovoPowerModelHRR.csv")
+PowerModelBFRR <- read.csv("results/tables/denovoPowerModelBFRR.csv")
+SimulatedN_H <- read.csv("results/tables/denovoSimulatedN_H.csv")
+SimulatedN_BF <- read.csv("results/tables/denovoSimulatedN_BF.csv")
+RRSimNNeeded <- read.csv("results/tables/denovoRRSimNNeeded.csv")
 
@@
 ```{r install_needed_packages, echo=FALSE, message=FALSE, warning=FALSE}
-deps = c("knitr","vegan","epiR","AUCRF","meta","metafor","ggplot2", "plyr", "reshape2", "RColorBrewer", "scales", "gridExtra");
+deps = c("knitr","vegan","epiR","AUCRF","metafor","ggplot2", "plyr", "reshape2", "RColorBrewer", "scales", "gridExtra", "pwr", "statmod");
 for (dep in deps){
@@
 
-**Table 3. Summary of P-values for Measurements of Interest for each Individual Study for Obese versus Normal**
+**Table 3. Summary of Power Calculation Simulations for Shannon Diversity for Non-obese versus Obese**
 
 ```{r Table_3, echo=FALSE, message=FALSE, warning=FALSE}
+ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
+ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
 
-Study <- as.character(OverallPValue$X)
-Bacteroidetes <- format(round(OverallPValue$Bacteroidetes, 3), nsmall=3)
-Firmicutes <- format(round(OverallPValue$Firmicutes, 3), nsmall=3)
-BFRatio <- format(round(OverallPValue$BFRatio, 3), nsmall=3)
-Shannon <- format(round(OverallPValue$Shannon, 3), nsmall=3)
-OTURich <- format(round(OverallPValue$OTURich, 3), nsmall=3)
-Evenness <- format(round(OverallPValue$Evenness, 3), nsmall=3)
-BrayC <- format(round(OverallPValue$BrayC, 3), nsmall=3)
+Study <- paste(as.character(PowerTable$Study), " (", ContTotal, "|", ObeseTotal, ")", sep="")
+ActualDiffH <- format(round(ActualDiff$H, 3), nsmall=3)
+Shannon <- format(round(OverallPValue$Shannon, 2), nsmall=2)
+OneP <- format(round(ModelledPower1PerDiff$H, 3), nsmall=3)
+FiveP <- format(round(ModelledPower5PerDiff$H, 3), nsmall=3)
+TenP <- format(round(ModelledPower10PerDiff$H, 3), nsmall=3)
+FifteenP <- format(round(ModelledPower15PerDiff$H, 3), nsmall=3)
 
-table3 <- cbind(Study, Bacteroidetes, Firmicutes, BFRatio, Shannon, OTURich, Evenness, BrayC)
-colnames(table3) <- c("Study", "Bacteroidetes", "Firmicutes", "B/F Ratio", "Shannon Diversity", "OTU Richness", "Evenness", "Bray Curtis")
+table3 <- cbind(Study, ActualDiffH, Shannon, OneP, FiveP, TenP, FifteenP)
+colnames(table3) <- c("Study (Non-Obese|Obese)", "Actual Percent Difference", "P-value", "Estimated Power for 1% Difference", "Estimated Power for 5% Difference", "Estimated Power for 10% Difference", "Estimated Power for 15% Difference")
 
@@
 
-**Table 4. Summary of Power Calculations for each Individual for Shannon Diversity and B/F ratio for Non-obese versus Obese**
+**Table 4. Summary of Power Calculation Simulations for B/F Ratio for Non-obese versus Obese**
 
@@
 
-Study <- as.character(PowerTable$Study)
-PowerH <- PowerTable$StudyPowerH
-PowerBF <- PowerTable$StudyPowerBF
+ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
+ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
 
-table4 <- cbind(Study, PowerH, PowerBF)
-colnames(table4) <- c("Study", "Shannon Diversity Power (%)", "B/F Ratio Power (%)")
+Study <- paste(as.character(PowerTable$Study), " (", ContTotal, "|", ObeseTotal, ")", sep="") 
+ActualDiffBF <- format(round(ActualDiff$BFratio, 3), nsmall=3)
+BF <- format(round(OverallPValue$BFRatio, 2), nsmall=2)
+OneP <- format(round(ModelledPower1PerDiff$BFratio, 3), nsmall=3)
+FiveP <- format(round(ModelledPower5PerDiff$BFratio, 3), nsmall=3)
+TenP <- format(round(ModelledPower10PerDiff$BFratio, 3), nsmall=3)
+FifteenP <- format(round(ModelledPower15PerDiff$BFratio, 3), nsmall=3)
+
+table4 <- cbind(Study, ActualDiffBF, BF, OneP, FiveP, TenP, FifteenP)
+colnames(table4) <- c("Study (Non-Obese|Obese)", "Actual Percent Difference", "P-value", "Estimated Power for 1% Difference", "Estimated Power for 5% Difference", "Estimated Power for 10% Difference", "Estimated Power for 15% Difference")
 
 kable(table4, format = "markdown", align = 'c', padding = 2)
+
+```
+
+**Table 5. Summary of Sample Size Needed Simulations for Shannon Diversity for Non-obese versus Obese**
+
+```{r Table_5, echo=FALSE, message=FALSE, warning=FALSE}
+
+ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
+ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
+
+Study <- paste(as.character(PowerTable$Study), " (", ContTotal, "|", ObeseTotal, ")", sep="")
+
+NNeeded <- format(round(SimulatedN_H[, -1],0), nsmall=0)
+table5 <- cbind(Study, NNeeded)
+colnames(table5) <- c("Study (Non-Obese|Obese)", "Actual n/Group Needed", "Estimated n/Group for 1% Difference", "Estimated n/Group for 5% Difference", "Estimated n/Group for 10% Difference", "Estimated n/Group for 15% Difference")
+
+kable(table5, format = "markdown", align = 'c', padding = 2)
+```
+
+**Table 6. Summary of Sample Size Needed Simulations for B/F Ratio for Non-obese versus Obese**
+
+```{r Table_6, echo=FALSE, message=FALSE, warning=FALSE}
+
+ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
+ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
+
+Study <- paste(as.character(PowerTable$Study), " (", ContTotal, "|", ObeseTotal, ")", sep="")
+
+NNeeded <- format(round(SimulatedN_BF[, -1],0), nsmall=0)
+table6 <- cbind(Study, NNeeded)
+colnames(table6) <- c("Study (Non-Obese|Obese)", "Actual n/Group Needed", "Estimated n/Group for 1% Difference", "Estimated n/Group for 5% Difference", "Estimated n/Group for 10% Difference", "Estimated n/Group for 15% Difference")
+
+kable(table6, format = "markdown", align = 'c', padding = 2)
 ```
@@
 #To create a graph visualize the data
-forest(res, at=log(c(0.3, 1, 5)), atransf=exp, xlab="RR of Obesity in high versus low Shannon diversity based on Median", cex=0.8)
+forest(res, at=log(c(0.3, 1, 5)), atransf=exp, mlab= "Pooled Relative Risk", xlab="RR of Obesity in high versus low Shannon diversity based on Median", cex=0.8)
 par(font=2) #Switch to bold
@@
 #To create a graph visualize the data
-forest(res2, at=log(c(0.1, 1, 3)), atransf=exp, xlab="RR of Obesity in high versus low B/F ratio based on Median", cex=0.8)
+forest(res2, at=log(c(0.1, 1, 3)), atransf=exp, mlab= "Pooled Relative Risk", xlab="RR of Obesity in high versus low B/F ratio based on Median", cex=0.8)
 par(font=2) #Switch to bold
@@
 
-```{r Figure_4, echo=FALSE, fig.path="results/figures/"}
-par(mfrow=c(1,2))
-funnel(res, main="A")
-funnel(res2, main="B")
-```
-
-**Figure 4: Funnel Plot of the Shannon Diversity and Bacteroidetes/Firmicutes Ratio Relative Risk.**  **A)** Overall there does not seem to be any bias associated with the studies selected for the Shannon Diversity analysis with studies falling on either side of the predicted value and those with smaller total n falling further away from this.
-**B)** For the Bacteroidetes/Firmicutes ratio there does not seem to be any associated bias.
-The overall pattern is similar to that observed for the Shannon Diversity analysis.
-
-```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
 # Section A
@@
 p3 <- p2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-p4 <- p3 + xlab("Obese Category") + ylab("Z-score Normalized H")
+p4 <- p3 + xlab("") + ylab("Z-score Normalized H") + scale_x_discrete(breaks=c("No", "Yes"), labels=c("Non-Obese", "Obese"))
 p5 <- p4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
@@
 q3 <- q2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-q4 <- q3 + xlab("BMI Classification") + ylab("Z-score Normalized H")
+q4 <- q3 + xlab("") + ylab("Z-score Normalized H")
 q5 <- q4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
@@
 bf3 <- bf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-bf4 <- bf3 + xlab("Obese Category") + ylab("Z-score Normalized Log B/F Ratio")
+bf4 <- bf3 + xlab("") + ylab("Z-score Normalized Log B/F Ratio") + 
+  scale_x_discrete(breaks=c("No", "Yes"), labels=c("Non-Obese", "Obese"))
 bf5 <- bf4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
@@
 qbf3 <- qbf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-qbf4 <- qbf3 + xlab("BMI Classification") + ylab("Z-score Normalized Log B/F Ratio")
+qbf4 <- qbf3 + xlab("") + ylab("Z-score Normalized Log B/F Ratio")
 qbf5 <- qbf4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
@@
 
-**Figure 5: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** Z-score Normalized Shannon Diversity and Obesity.
+**Figure 4: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** Z-score Normalized Shannon Diversity and Obesity.
 There was a significant difference between non-obese (n = `r length(which(group.data.test$obese == "No"))`) and obese groups (n = `r length(which(group.data.test$obese == "Yes"))`) (P-value = `r signif(combZHTtest$p.value, 3)`).

### 2016-03-23 microbiomeAndobesitySupplemental.Rmd
manuscript and Suppl update

--- 2016-03-18 microbiomeAndobesitySupplemental.Rmd
@@
 <br><br>
+
+```{r Loading_data_tables, echo=FALSE, message=FALSE, warning=FALSE}
+
+OverallPValue <- read.csv("results/tables/denovoOverallPTable.csv")
+ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
+BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
+
+```
 
@@
 
+**Table S1. Summary of P-values for Measurements of Interest for each Individual Study for Obese versus Normal**
+
+```{r Table_S1, echo=FALSE, message=FALSE, warning=FALSE}
+
+library(knitr)
+
+Study <- as.character(OverallPValue$X)
+Bacteroidetes <- format(round(OverallPValue$Bacteroidetes, 2), nsmall=2)
+Firmicutes <- format(round(OverallPValue$Firmicutes, 2), nsmall=2)
+BFRatio <- format(round(OverallPValue$BFRatio, 2), nsmall=2)
+Shannon <- format(round(OverallPValue$Shannon, 2), nsmall=2)
+OTURich <- format(round(OverallPValue$OTURich, 2), nsmall=2)
+Evenness <- format(round(OverallPValue$Evenness, 2), nsmall=2)
+BrayC <- format(round(OverallPValue$BrayC, 2), nsmall=2)
+
+tableS1 <- cbind(Study, Bacteroidetes, Firmicutes, BFRatio, Shannon, OTURich, Evenness, BrayC)
+colnames(tableS1) <- c("Study", "Bacteroidetes", "Firmicutes", "B/F Ratio", "Shannon Diversity", "OTU Richness", "Evenness", "Bray Curtis")
+
+kable(tableS1, format = "markdown", align = 'c', padding = 2)
+
+```
+
 ![Figure S1](results/figures/suppFigClostridiales.png)\
@@
 
-**Table 1. Summary of BLAST alignment of Representative OTU for Clostridiales, Lachnospiraceae, and Ruminococcaceae classification OTUs**
+**Table 2. Summary of BLAST alignment of Representative OTU for Clostridiales, Lachnospiraceae, and Ruminococcaceae classification OTUs**
 
-```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
-
-library(knitr)
+```{r Table_S2, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-table1 <- cbind(Study, VariableRegion, ClosMostPrev, ClostHighMatch, LachMostPrev, LachHighMatch, RumiMostPrev, RumiHighMatch)
-colnames(table1) <- c("Study", "Hypervariable Region", "Clostridiales OTU Most Prevalent Match", "Clostridiales OTU Highest Match (Query Cover|Identity)", "Lachnospiraceae OTU Most Prevalent Match", "Lachnospiraceae OTU Highest Match (Query Cover|Identity)", "Ruminococcaceae OTU Most Prevalent Match", "Ruminococcaceae OTU Highest Match (Query Cover|Identity)")
+tableS2 <- cbind(Study, VariableRegion, ClosMostPrev, ClostHighMatch, LachMostPrev, LachHighMatch, RumiMostPrev, RumiHighMatch)
+colnames(tableS2) <- c("Study", "Hypervariable Region", "Clostridiales OTU Most Prevalent Match", "Clostridiales OTU Highest Match (Query Cover|Identity)", "Lachnospiraceae OTU Most Prevalent Match", "Lachnospiraceae OTU Highest Match (Query Cover|Identity)", "Ruminococcaceae OTU Most Prevalent Match", "Ruminococcaceae OTU Highest Match (Query Cover|Identity)")
 
-kable(table1, format = "markdown", align = 'c', padding = 2)
+kable(tableS2, format = "markdown", align = 'c', padding = 2)
 ```
+
+**Table S3. Top Sequence Similarity by Variable Region Based on the Representative OTU for Clostridiales, Lachnospiraceae, and Ruminococcaceae**
+
+```{r Table_S3, echo=FALSE, message=FALSE, warning=FALSE}
+
+VariableRegion <- c("V1-V2 vs V2", "V1-V3", "V4")
+
+TopHitClos <- c("None", "50.3|39.1", "89.4|2.8")
+AverageClos <- c("None", "48.0|42.4", "87.7|2.4")
+TopHitLachno <- c("TBC", "40.1|47.9", "92.9|0.8")
+AverageLachno <- c("TBC", "35.9|52.5", "85.8|5.6")
+TopHitRumi <- c("TBC", "30.4|68.4", "86.6|0.8")
+AverageRumi <- c("TBC", "29.2|69.1", "81.6|6.6")
+
+tableS3 <- cbind(VariableRegion, TopHitClos, AverageClos, TopHitLachno, AverageLachno, TopHitRumi, AverageRumi)
+colnames(tableS3) <- c("Variable Region", "Top Similarity Colstridiales (Id|Gaps)", "Average Similarity Colstridiales (Id|Gaps)", "Top Similarity Lachnospiraceae (Id|Gaps)", "Average Similarity Lachnospiraceae (Id|Gaps)", "Top Similarity Ruminococcaceae (Id|Gaps)", "Average Similarity Ruminococcaceae (Id|Gaps)")
+
+kable(tableS3, format = "markdown", align = 'c', padding = 2)
+
+```
+
+```{r Figure_S4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+
+library(metafor)
+
+# Shannon Diversity Data for Funnel Plot
+StudyNo <- as.numeric(ShannonRRTable$X)
+Study <- as.character(ShannonRRTable$Study)
+Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
+Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
+tpos <- as.numeric(ShannonRRTable$tposH) #obese in low diversity group
+tneg <- as.numeric(ShannonRRTable$tnegH) #nonobese in low diversity group
+cpos <- as.numeric(ShannonRRTable$cposH) #obese in high diversity group
+cneg <- as.numeric(ShannonRRTable$cnegH) #nonobese in high diversity group
+RR <- as.numeric(ShannonRRTable$RRH)
+low <- as.numeric(ShannonRRTable$lowH)
+high <- as.numeric(ShannonRRTable$highH)
+
+numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
+raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
+
+dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
+k <- length(dat$StudyNo)
+dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
+dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
+dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
+dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
+
+#Used to generate the RE Model estimate based on all the studies
+res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
+
+# B/F ratio data used to generate Funnel Plot
+StudyNo <- as.numeric(BFRatioRRTable$X)
+Study <- as.character(BFRatioRRTable$Study)
+Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
+Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
+tpos <- as.numeric(BFRatioRRTable$tposBF) #obese in low diversity group
+tneg <- as.numeric(BFRatioRRTable$tnegBF) #nonobese in low diversity group
+cpos <- as.numeric(BFRatioRRTable$cposBF) #obese in high diversity group
+cneg <- as.numeric(BFRatioRRTable$cnegBF) #nonobese in high diversity group
+RR <- as.numeric(BFRatioRRTable$RRBF)
+low <- as.numeric(BFRatioRRTable$lowBF)
+high <- as.numeric(BFRatioRRTable$highBF)
+
+numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
+raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
+
+dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
+k <- length(dat$StudyNo)
+dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
+dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
+dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
+dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
+
+#Used to generate the RE Model estimate based on all the studies
+res2 <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
+
+# Generate Funnel Plots for supplemental
+par(mfrow=c(1,2))
+funnel(res, main="A")
+funnel(res2, main="B")
+```
+
+**Figure 4: Funnel Plot of the Shannon Diversity and Bacteroidetes/Firmicutes Ratio Relative Risk.**  **A)** Overall there does not seem to be any bias associated with the studies selected for the Shannon Diversity analysis with studies falling on either side of the predicted value and those with smaller total n falling further away from this.
+**B)** For the Bacteroidetes/Firmicutes ratio there does not seem to be any associated bias.
+The overall pattern is similar to that observed for the Shannon Diversity analysis.
 

### 2016-03-23 code/numberSimulationsGood.R
minor changes to Nsim code

@@ -35,8 +35,8 @@ for(i in 1:8){
 BFActualNeededN <- rep(0, 8)
 
 for(i in 1:8){
-  cohensD <- (SummaryStatsByObeseGroup$MeanNonObeseBF[i] -  
-                SummaryStatsByObeseGroup$MeanObeseBF[i]) / 
+  cohensD <- abs((SummaryStatsByObeseGroup$MeanNonObeseBF[i] -  
+                SummaryStatsByObeseGroup$MeanObeseBF[i])) / 
     SummaryStatsByStudy$sdBF[i]  
   # This needs to change based on study
   
@@ -131,7 +131,7 @@ for(i in 1:8){
 
 ########## FIFTEEN PERCENT #####################################################
 
-FifteenPercentDiff <- SummaryStatsByStudy$averageStudyH * 0.10
+FifteenPercentDiff <- SummaryStatsByStudy$averageStudyH * 0.15
 HFifteenPer <- rep(0, 8)
 
 for(i in 1:8){
@@ -144,7 +144,7 @@ for(i in 1:8){
 }
 
 
-FifteenPercentDiff <- SummaryStatsByStudy$averageStudyBF * 0.10
+FifteenPercentDiff <- SummaryStatsByStudy$averageStudyBF * 0.15
 BFratioFifteenPer <- rep(0, 8)
 
 for(i in 1:8){

### 2016-03-23 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
combined n power sim add

@@
 Additionally, we did not observe a difference when the individuals were separated by BMI groups (P-value = 0.3317) [Figure 5D].
+
+Using a small (0.2), medium (0.5), and large (0.8) cohen value we estimated the power of the combined data set as well as the average cohen value from across studies for Shannon Diversity (0.06) and BF ratio (0.01).
+
+small effect for combined data power calculation:
+`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.2, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`  
+medium effect for combined data power calculation:
+`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.5, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`  
+large effect for combined data power calculation:
+`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.8, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`  
+average effect for all Shannon Diversity:
+`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.06, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`
+average effect for all BF ratio:
+`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.01, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`
 

### 2016-03-23 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
Methods revisions to bring in power analysis

@@
 The RR was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group.
-We then performed a Fisher exact-test to test whether the RR was significantly different from 1.0 using the epiR (v0.9-6.9) and metafor (v1.9-8) packages.
+We then performed a Fisher exact-test to test whether the RR was significantly different from 1.0 using the epiR (v0.9-6.9)[@citation] and metafor (v1.9-8)[@citation] packages.
 Finally, we assessed the potential publication bias in the diversity and B/F analyses using a funnel plot.
@@
 We also measured the Spearman correlation coefficient between the Z-scores for each metric and BMI values.
+
+We performed power analyses to assess the ability to detect varying effect sizes for the observed study designs.
+We also predicted the number of individuals required to observe a power of 80% when non-obese and obese individuals were equally represented in the study.
+Power calculations assumed the data were normally distributed.
+Since the observed data were not always normally distributed, the observed power is probably overestimated and the number of required subjects is probably underestimated.
+We used Fisher's exact test to calculate power for relative risk data.
+**To make comparisons of power calculations across studies meaningful, we assumed the proportion of obese individuals was consistent and used the proportion of obese individuals across all studies.** Power analyses were performed using the statmod [@citation] and pwr [@citation] R packages.
 

### 2016-03-24 code/MakeFiguresCode.R
Code update for new figs

@@ -1,0 +1,270 @@
+## Make Figures for Manuscript
+## Obesesity and the bacterial microbiome
+## Marc Sze
+## March 23, 2016
+
+library(ggplot2)
+
+
+OverallPValue <- read.csv("results/tables/denovoOverallPTable.csv")
+PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
+ModelledPower1PerDiff <- read.csv("results/tables/denovoModelled1PerDiff.csv")
+ModelledPower5PerDiff <- read.csv("results/tables/denovoModelled5PerDiff.csv")
+ModelledPower10PerDiff <- read.csv("results/tables/denovoModelled10PerDiff.csv")
+ModelledPower15PerDiff <- read.csv("results/tables/denovoModelled15PerDiff.csv")
+ActualDiff <- read.csv("results/tables/denovoActualDiff.csv")
+SimulatedN_H <- read.csv("results/tables/denovoSimulatedN_H.csv")
+SimulatedN_BF <- read.csv("results/tables/denovoSimulatedN_BF.csv")
+
+##################### SHANNON DIVERSITY ###################################
+
+# A data frame that has all the necessary data for power
+Study <- as.character(PowerTable$Study)
+ActualDiffH <- as.numeric(format(round(ActualDiff$H, 3), nsmall=3))
+Shannon <- as.numeric(format(round(OverallPValue$Shannon, 2), nsmall=2))
+OneP <- as.numeric(format(round(ModelledPower1PerDiff$H, 3), nsmall=3))
+FiveP <- as.numeric(format(round(ModelledPower5PerDiff$H, 3), nsmall=3))
+TenP <- as.numeric(format(round(ModelledPower10PerDiff$H, 3), nsmall=3))
+FifteenP <- as.numeric(format(round(ModelledPower15PerDiff$H, 3), nsmall=3))
+
+DataTable1 <- as.data.frame(
+  cbind(Shannon, ActualDiffH, OneP, FiveP, TenP, FifteenP))
+DataTable1$Study <- Study
+TableLegend <- paste(as.character(Study), ", P-value=", Shannon, sep="")
+DataTable1$Legend <- TableLegend
+
+library(reshape)
+test <- melt(DataTable1, id=c("ActualDiffH", "Shannon", "Study", "Legend"))
+
+test$EstDiff[test$variable=="OneP"] <- 1
[…]

### 2016-03-24 code/MakeRRFiguresCode.R

@@ -1,0 +1,15 @@
+## Make Custom RR Figures for Manuscript
+## Obesesity and the bacterial microbiome
+## Marc Sze
+## March 24, 2016
+
+
+
+
+
+
+
+
+
+
+

### 2016-03-24 code/PowerVariations.R

@@ -428,22 +428,33 @@ write.csv(ActualDiff, "results/tables/denovoActualDiff.csv")
 ## used for the control group and the treatment group would vary by:
 ## 1%, 5%, 10%, 15%
 
+# Create base proportion based on data for control group
+# Vary the obese proportion by 1, 5, 10, 15 percent
+
+BasePropotionsH <- (ShannonRRTable$cposH)/(ShannonRRTable$cposH + 
+                                            ShannonRRTable$cnegH)
+
+BasePropotionsBF <- (BFRatioRRTable$cposBF)/(BFRatioRRTable$cposBF + 
+                                              BFRatioRRTable$cnegBF)
 #One percent
 
 OnePercentHRR <- rep(0, 8)
+
 for(i in 1:8){
-  OnePercentHRR[i] <- power.fisher.test(0.38, 0.39, 
-                              ShannonRRTable$tnegH[i] + ShannonRRTable$cnegH[i], 
-                              ShannonRRTable$tposH[i] + ShannonRRTable$cposH[i], 
-                              alpha=0.05, nsim=1000, alternative="two.sided")
+  OnePercentHRR[i] <- power.fisher.test(BasePropotionsH[i], 
+                                        BasePropotionsH[i] + 0.01, 
+                            ShannonRRTable$cposH[i] + ShannonRRTable$cnegH[i], 
+                            ShannonRRTable$tposH[i] + ShannonRRTable$tnegH[i], 
+                            alpha=0.05, nsim=1000, alternative="two.sided")
 }
 
 OnePercentBFRR <- rep(0, 8)
 for(i in 1:8){
-  OnePercentBFRR[i] <- power.fisher.test(0.38, 0.39, 
-                              BFRatioRRTable$tnegBF[i] + BFRatioRRTable$cnegBF[i], 
-                              BFRatioRRTable$tposBF[i] + BFRatioRRTable$cposBF[i], 
-                              alpha=0.05, nsim=1000, alternative="two.sided")
+  OnePercentBFRR[i] <- power.fisher.test(BasePropotionsBF[i], 
+                                        BasePropotionsBF[i] + 0.01, 
+                            BFRatioRRTable$cposBF[i] + BFRatioRRTable$cnegBF[i], 
+                            BFRatioRRTable$tposBF[i] + BFRatioRRTable$tnegBF[i], 
[…]

### 2016-03-24 code/StrategydenovoAnalysisMetaHitDel.R

@@ -1778,7 +1778,8 @@ rm(WuBacter, WuFirm, WuBF, WuH, WuS, WuJ, WuPERM, WuZH, WuZLogBF,
 ############ Preparing Data Tables for Analysis ###########################
 ###########################################################################
 
-microbiome <- read.table("data/process/Turnbaugh/TurnbaughSub.shared", header=T)
+microbiome <- read.table("data/process/Turnbaugh/TurnbaughGoodSub.shared", 
+                         header=T)
 microbiome <- EditTable(microbiome, 2, delRange=c(1:3))
 metadata <- read.csv("data/process/Turnbaugh/turnbaugh.metadata.csv")
 metadata <- EditTable(metadata, 4, delRange=4)
@@ -1795,8 +1796,8 @@ rm(keep1)
 alpha.test <- makeAlphaTable(microbiome)
 
 #Get phyla information
-phyla.table <- MakePhylaTable(microbiome, "data/process/Turnbaugh/phyla.csv", 
-                              csv = T)
+phyla.table <- MakePhylaTable(microbiome, 
+                              "data/process/Turnbaugh/taxonomyKey.txt")
 
 #combine phyla that are not that abundant
 phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Synergistetes", "TM7", 
@@ -1821,7 +1822,7 @@ bmi <- metadata$BMI.category
 
 ##Test BMI versus alpha diversity and phyla
 
-turnbaughH <- wilcox.test(alpha.test$H ~ obese) #P-value=0.1162
+turnbaughH <- wilcox.test(alpha.test$H ~ obese) #P-value=0.9699
 MeanNonObeseH <- c(MeanNonObeseH, 
                    mean(alpha.test$H[which(metadata$obese == "No")]))
 MeanObeseH <- c(MeanObeseH, 
@@ -1833,7 +1834,7 @@ SDObeseH <- c(SDObeseH,
 averageStudyH <- c(averageStudyH, mean(alpha.test$H))
 sdH <- c(sdH, sd(alpha.test$H))
 
-turnbaughS <- wilcox.test(alpha.test$S ~ obese) #P-value=0.05479
+turnbaughS <- wilcox.test(alpha.test$S ~ obese) #P-value=0.5436
 MeanNonObeseS <- c(MeanNonObeseS, 
[…]

### 2016-03-24 code/numberSimulationsGood.R

@@ -172,21 +172,29 @@ write.csv(SimulatedN_BF, "results/tables/denovoSimulatedN_BF.csv")
 ####################### Estimation of n for percent differences
 ####################### Between two proportions (ACTUAL)
 
+# First need to calculate the base proportions
+TreatmentProportionH <- (ShannonRRTable$tposH)/(ShannonRRTable$tposH + 
+                                                 ShannonRRTable$tnegH)
+
+ControlProportionH <- (ShannonRRTable$cposH)/(ShannonRRTable$cposH + 
+                                              ShannonRRTable$cnegH)
+
+TreatmentProportionBF <- (BFRatioRRTable$tposBF)/(BFRatioRRTable$tposBF + 
+                                                  BFRatioRRTable$tnegBF)
+
+ControlProportionBF <- (BFRatioRRTable$cposBF)/(BFRatioRRTable$cposBF + 
+                                               BFRatioRRTable$cnegBF)
+
+
 # Shannon Diversity
 # Calculate the Effect size (h)
+# Calculate Simulated N for each data set
 
 effectSize <- rep(0, 8)
-for(i in 1:8){
-  TotalTreatment <- ShannonRRTable$tposH[i] + ShannonRRTable$tnegH[i]
-  ControlTreatment <- ShannonRRTable$cposH[i] + ShannonRRTable$cnegH[i]
-  
-  effectSize[i] <- ES.h(ShannonRRTable$tposH[i]/TotalTreatment, 
-                        ShannonRRTable$cposH[i]/ControlTreatment)
-}
-
-# Calculate Simulated N for each data set 
 RRActSimNH <- rep(0, 8)
 for(i in 1:8){
+  effectSize[i] <- ES.h(ControlProportionH[i], 
+                        TreatmentProportionH[i])
   try(
   RRActSimNH[i] <- pwr.2p.test(h=abs(effectSize[i]), sig.level=0.05, 
                                power=0.8, alternative = "two.sided")$n)
@@ -194,24 +202,158 @@ for(i in 1:8){
 
 # BF Ratio
 # Calculate the Effect size (h)
+# Calculate Simulated N for each data set 
 
 effectSize <- rep(0, 8)
-for(i in 1:8){
-  TotalTreatment <- BFRatioRRTable$tposBF[i] + BFRatioRRTable$tnegBF[i]
[…]

### 2016-03-24 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
another manuscript update

--- 2016-03-23 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
@@
 SimulatedN_BF <- read.csv("results/tables/denovoSimulatedN_BF.csv")
-RRSimNNeeded <- read.csv("results/tables/denovoRRSimNNeeded.csv")
+RRSimNNeededH <- read.csv("results/tables/denovoRRSimNNeededH.csv")
+RRSimNNeededBF <- read.csv("results/tables/denovoRRSimNNeededBF.csv")
 
@@
 
+ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
+ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
+
+ObBreakdown <- paste(ContTotal, " | ", ObeseTotal, sep="")
+
 BMI <- paste(demographics$meanBMI, " (", demographics$SDBMI, ")", sep="")
@@
 
-table1 <- cbind(Study, AgeTable, Sex, Ancestry, BMI, Min, Max)
+table1 <- cbind(Study, AgeTable, Sex, Ancestry, ObBreakdown, BMI, Min, Max)
 table1[table1=="NA" | table1=="NA | NA" | table1=="NA (NA)" | 
@@
 
-colnames(table1) <- c("Study", "Age (SD)", "Sex (F|M)", "European Ancestry (%)", "BMI (SD)", "Min BMI","Max BMI")
+colnames(table1) <- c("Study", "Age (SD)", "Sex (F|M)", "European Ancestry (%)", "Obese (No|Yes)", "BMI (SD)", "Min BMI","Max BMI")
 
@@
 
-**Table 3. Summary of Power Calculation Simulations for Shannon Diversity for Non-obese versus Obese**
+```{r Table_5, echo=FALSE, message=FALSE, warning=FALSE}
 
-```{r Table_3, echo=FALSE, message=FALSE, warning=FALSE}
-ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
-ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
-
-Study <- paste(as.character(PowerTable$Study), " (", ContTotal, "|", ObeseTotal, ")", sep="")
-ActualDiffH <- format(round(ActualDiff$H, 3), nsmall=3)
-Shannon <- format(round(OverallPValue$Shannon, 2), nsmall=2)
-OneP <- format(round(ModelledPower1PerDiff$H, 3), nsmall=3)
-FiveP <- format(round(ModelledPower5PerDiff$H, 3), nsmall=3)
-TenP <- format(round(ModelledPower10PerDiff$H, 3), nsmall=3)
-FifteenP <- format(round(ModelledPower15PerDiff$H, 3), nsmall=3)
-
-table3 <- cbind(Study, ActualDiffH, Shannon, OneP, FiveP, TenP, FifteenP)
-colnames(table3) <- c("Study (Non-Obese|Obese)", "Actual Percent Difference", "P-value", "Estimated Power for 1% Difference", "Estimated Power for 5% Difference", "Estimated Power for 10% Difference", "Estimated Power for 15% Difference")
-
-kable(table3, format = "markdown", align = 'c', padding = 2)
-```
-
-**Table 4. Summary of Power Calculation Simulations for B/F Ratio for Non-obese versus Obese**
-
-```{r Table_4, echo=FALSE, message=FALSE, warning=FALSE}
-
-ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
-ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
-
-Study <- paste(as.character(PowerTable$Study), " (", ContTotal, "|", ObeseTotal, ")", sep="") 
-ActualDiffBF <- format(round(ActualDiff$BFratio, 3), nsmall=3)
-BF <- format(round(OverallPValue$BFRatio, 2), nsmall=2)
-OneP <- format(round(ModelledPower1PerDiff$BFratio, 3), nsmall=3)
-FiveP <- format(round(ModelledPower5PerDiff$BFratio, 3), nsmall=3)
-TenP <- format(round(ModelledPower10PerDiff$BFratio, 3), nsmall=3)
-FifteenP <- format(round(ModelledPower15PerDiff$BFratio, 3), nsmall=3)
-
-table4 <- cbind(Study, ActualDiffBF, BF, OneP, FiveP, TenP, FifteenP)
-colnames(table4) <- c("Study (Non-Obese|Obese)", "Actual Percent Difference", "P-value", "Estimated Power for 1% Difference", "Estimated Power for 5% Difference", "Estimated Power for 10% Difference", "Estimated Power for 15% Difference")
-
-kable(table4, format = "markdown", align = 'c', padding = 2)
+source("code/MakeFiguresCode.R")
+grid.arrange(a2, b2, nrow=2)
 
 ```
-
-**Table 5. Summary of Sample Size Needed Simulations for Shannon Diversity for Non-obese versus Obese**
-
-```{r Table_5, echo=FALSE, message=FALSE, warning=FALSE}
-
-ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
-ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
-
-Study <- paste(as.character(PowerTable$Study), " (", ContTotal, "|", ObeseTotal, ")", sep="")
-
-NNeeded <- format(round(SimulatedN_H[, -1],0), nsmall=0)
-table5 <- cbind(Study, NNeeded)
-colnames(table5) <- c("Study (Non-Obese|Obese)", "Actual n/Group Needed", "Estimated n/Group for 1% Difference", "Estimated n/Group for 5% Difference", "Estimated n/Group for 10% Difference", "Estimated n/Group for 15% Difference")
-
-kable(table5, format = "markdown", align = 'c', padding = 2)
-```
-
-**Table 6. Summary of Sample Size Needed Simulations for B/F Ratio for Non-obese versus Obese**
+**Figure 5. Summary of Power and Sample Size Simulations for Shannon Diversity for Non-obese versus Obese**
 
@@
 
-ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
-ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
+grid.arrange(c2, d2, nrow=2)
 
-Study <- paste(as.character(PowerTable$Study), " (", ContTotal, "|", ObeseTotal, ")", sep="")
-
-NNeeded <- format(round(SimulatedN_BF[, -1],0), nsmall=0)
-table6 <- cbind(Study, NNeeded)
-colnames(table6) <- c("Study (Non-Obese|Obese)", "Actual n/Group Needed", "Estimated n/Group for 1% Difference", "Estimated n/Group for 5% Difference", "Estimated n/Group for 10% Difference", "Estimated n/Group for 15% Difference")
-
-kable(table6, format = "markdown", align = 'c', padding = 2)
 ```
+**Figure 6. Summary of Power and Sample Size Simulations for Bacteroidetes/Firmicutes Ratio for Non-obese versus Obese**
 
@@
 p4 <- p3 + xlab("") + ylab("Z-score Normalized H") + scale_x_discrete(breaks=c("No", "Yes"), labels=c("Non-Obese", "Obese"))
-p5 <- p4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
+p5 <- p4 + guides(fill=FALSE)
 p6 <- p5 + theme(axis.line=element_line(colour="black"),
@@
 # Section B
-q1 <- ggplot(group.data.test, aes(BMI.class, ZH))
-q2 <- q1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
-q3 <- q2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-q4 <- q3 + xlab("") + ylab("Z-score Normalized H")
-q5 <- q4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
-q6 <- q5 + theme(axis.line=element_line(colour="black"),
-                 panel.grid.major = element_blank(),
-                 panel.grid.minor = element_blank(),
-                 panel.border = element_blank(),
-                 panel.background = element_blank())
-q7 <- q6 + ggtitle('B') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
-
-# Section C
 bf1 <- ggplot(group.data.test, aes(obese, ZLogBF))
@@
   scale_x_discrete(breaks=c("No", "Yes"), labels=c("Non-Obese", "Obese"))
-bf5 <- bf4 + guides(fill=guide_legend(title="Obese")) + theme(legend.title = element_text(size=8, face="bold"))
+bf5 <- bf4 + guides(fill=FALSE)
 bf6 <- bf5 + theme(axis.line=element_line(colour="black"),
@@
                    panel.background = element_blank())
-bf7 <- bf6 + ggtitle('C') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
+bf7 <- bf6 + ggtitle('B') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
 
-# Section D
-qbf1 <- ggplot(group.data.test, aes(BMI.class, ZLogBF))
-qbf2 <- qbf1 + geom_boxplot(aes(fill = BMI.class)) + scale_fill_brewer(palette="Set1")
-qbf3 <- qbf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-qbf4 <- qbf3 + xlab("") + ylab("Z-score Normalized Log B/F Ratio")
-qbf5 <- qbf4 + guides(fill=guide_legend(title="BMI Class")) + theme(legend.title = element_text(size=8, face="bold"))
-qbf6 <- qbf5 + theme(axis.line=element_line(colour="black"),
-                     panel.grid.major = element_blank(),
-                     panel.grid.minor = element_blank(),
-                     panel.border = element_blank(),
-                     panel.background = element_blank())
-qbf7 <- qbf6 + ggtitle('D') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
-
-grid.arrange(p7, q7, bf7, qbf7)
+grid.arrange(p7, bf7, ncol=2)
 ```

### 2016-03-25 code/MakeFiguresCode.R
code update nice Figs

@@ -55,7 +55,8 @@ a2 <- a1 + geom_jitter(aes(colour = Legend), size=5, alpha=7/10,
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.border = element_blank(),
-           panel.background = element_blank()) + 
+           panel.background = element_blank(), 
+           legend.key = element_rect(fill = "white")) + 
   geom_vline(aes(xintercept=ActualDiffH[1], colour=Legend[1]), 
                 linetype="dashed") + 
   geom_vline(aes(xintercept=ActualDiffH[2], colour=Legend[2]), 
@@ -114,8 +115,9 @@ b2 <- b1 + geom_jitter(aes(colour = Study), size=5, alpha=7/10,
                               panel.grid.major = element_blank(),
                               panel.grid.minor = element_blank(),
                               panel.border = element_blank(),
-                              panel.background = element_blank()) + 
-  ylim(0, 5) + 
+                              panel.background = element_blank(), 
+                              legend.position="none") + 
+  ylim(0, 5) +  
   geom_hline(aes(yintercept=LogActualNNeededH[1], colour=Study[1]), 
              linetype="dashed") + 
   geom_hline(aes(yintercept=LogActualNNeededH[2], colour=Study[2]), 
@@ -182,7 +184,8 @@ c2 <- c1 + geom_jitter(aes(colour = Legend), size=5, alpha=7/10,
                               panel.grid.major = element_blank(),
                               panel.grid.minor = element_blank(),
                               panel.border = element_blank(),
-                              panel.background = element_blank()) + 
+                              panel.background = element_blank(), 
+                              legend.key = element_rect(fill = "white")) + 
   geom_vline(aes(xintercept=ActualDiffBF[1], colour=Legend[1]), 
              linetype="dashed") + 
   geom_vline(aes(xintercept=ActualDiffBF[2], colour=Legend[2]), 
@@ -245,7 +248,8 @@ d2 <- d1 + geom_jitter(aes(colour = Study), size=5, alpha=7/10,
                               panel.grid.major = element_blank(),
                               panel.grid.minor = element_blank(),
                               panel.border = element_blank(),
-                              panel.background = element_blank()) + 
+                              panel.background = element_blank(), 
+                              legend.position="none") + 
   ylim(0, 7.5) + 
   geom_hline(aes(yintercept=LogActualNNeededBF[1], colour=Study[1]), 
              linetype="dashed") + 

### 2016-03-25 code/MakeROCFigure.R

@@ -1,0 +1,229 @@
+## Make Figures for Manuscript
+## Obesesity and the bacterial microbiome
+## Marc Sze
+## March 23, 2016
+
+library(ggplot2)
+library(AUCRF)
+library(pROC)
+source("code/StrategydenovoAnalysisMetaHitDel.R")
+
+rm(alpha.test, AUCRFDataTable, BFRatio, BFRatioRRTable, data, combinedData, 
+   demographicsTable, metadata, microbiome, overallPTable, phyla.table.rel.abund, 
+   PowerTable, ShannonRRTable, SummaryStatsByStudy, SummaryStatsByObeseGroup, 
+   testset, ancestry, averageStudyB, averageStudyBF, averageStudyF, 
+   averageStudyH, averageStudyJ, averageStudyS, Bacter, bacter, baxAUC, 
+   baxKopt, BaxterBFRR, BaxterHRR, BFratio, bmi, cnegBF, cnegH, cposBF, cposH, 
+   escoAUC, EscobarBFRR, EscobarHRR, escoKopt, females, Firm, firm, goodAUC, 
+   goodKopt, H, highBF, highH, HMPAUC, HMPBFRR, HMPHRR, HMPKopt, J, KoptAll, 
+   lowBF, lowH, males, maxBMI, meanAge, meanBMI, MeanNonObeseB, MeanNonObeseBF, 
+   MeanNonObeseF, MeanNonObeseH, MeanNonObeseJ, MeanNonObeseS, MeanObeseB, 
+   MeanObeseBF, MeanObeseF, MeanObeseH, MeanObeseJ, MeanObeseS, microb.rownames, 
+   minBMI, OOBAUCAll, rossAUC, RossBFRR, RossHRR, rossKopt, RRBF, RRH, S, SDAge, 
+   sdB, sdBF, SDBMI, sdF, sdH, sdJ, SDNonObeseB, SDNonObeseBF, SDNonObeseF, 
+   SDNonObeseH, SDNonObeseJ, SDNonObeseS, SDObeseB, SDObeseBF, SDObeseF, 
+   SDObeseH, SDObeseJ, SDObeseS, sdS, StudyPowerBF, StudyPowerBFRR, 
+   StudyPowerH, StudyPowerHRR, temporary, tnegBF, tnegH, totalN, tposBF, tposH, 
[…]

### 2016-03-25 code/MakeRRFiguresCode.R

@@ -3,13 +3,181 @@
 ## Marc Sze
 ## March 24, 2016
 
+library(ggplot2)
+library(metafor)
+library(wesanderson)
+
+# Adaption of code from Joey ggforest wrapper
+
+# First need to load and create the necessary rma class file
+# For Shannon Diversity
+
+ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
+
+StudyNo <- as.numeric(ShannonRRTable$X)
+Study <- as.character(ShannonRRTable$Study)
+Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
+Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
+tpos <- as.numeric(ShannonRRTable$tposH) #obese in low diversity group
+tneg <- as.numeric(ShannonRRTable$tnegH) #nonobese in low diversity group
+cpos <- as.numeric(ShannonRRTable$cposH) #obese in high diversity group
+cneg <- as.numeric(ShannonRRTable$cnegH) #nonobese in high diversity group
+RR <- as.numeric(ShannonRRTable$RRH)
+low <- as.numeric(ShannonRRTable$lowH)
+high <- as.numeric(ShannonRRTable$highH)
+
+numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
+raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
+
+dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
+k <- length(dat$StudyNo)
+dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
+dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
+dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
+dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
+
+#Used to generate the RE Model estimate based on all the studies
+ShannonRes <- rma(ai=tpos, bi=tneg, ci=cpos, 
+                  di=cneg, data=raw.overall, measure="RR", 
+                  slab=paste(Study, Year, Total, sep=", "), method="REML")
+RRSDPvalue <- anova.rma(ShannonRes)
+
[…]

### 2016-03-25 code/StrategydenovoAnalysisMetaHitDel.R

@@ -7,6 +7,7 @@ library(statmod)
 library(vegan)
 library(epiR)
 library(AUCRF)
+library(pROC)
 source("code/UsedFunctions.R")
 
 ############# BAXTER ######################################################
@@ -158,7 +159,6 @@ set.seed(3)
 baxterAUCfit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
 baxAUC <- baxterAUCfit$`OOB-AUCopt`
 baxKopt <- baxterAUCfit$Kopt
-
 
 ###########################################################################
 ############ Z-score Data Preparation ###################################

### 2016-03-25 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
manuscript update writing

@@
 }
+
+```
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+# Load specific R files to source data from for graphs
+
+source("code/MakeRRFiguresCode.R")
+source("code/MakeFiguresCode.R")
 
@@
 Only one diversity measure (OTU richness (N=1)) was significantly different between non-obese and obese individuals (P<0.05).
-With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`).
+With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`) and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05).
 In 7/8 of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies.
-When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = 1.16, CI = 1.01-1.34, P-value = 0.04) but not B/F ratio (RR = 0.98, CI = 0.85-1.13, P-value = 0.81).
-Finally, when we normalized the Shannon diversity data and pooled studies there was a significant difference between non-obese and obese individuals (difference in mean z-score = `r format(round(combZHTtest$estimate[1]-combZHTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZHTtest$p.value, 2), nsmall=2)`).
-Similar to RR there was no significant difference in the normalized B/F ratio between non-obese and obese individuals (difference in mean z-score = `r format(round(combZLogBFTtest$estimate[1]-combZLogBFTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`).
+When we pooled the results across studies, there was no significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(10^test3R$LogFC[1],2), nsmall=2)`, CI = `r format(round(10^ShannonRes$ci.lb, 2), nsmall=2)`-`r format(round(10^ShannonRes$ci.ub, 2), nsmall=2)`, P-value = `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`) and B/F ratio (RR = `r format(round(10^test3BF$LogFC[1],2), nsmall=2)`, CI = `r format(round(10^BFRes$ci.lb, 2), nsmall=2)`-`r format(round(10^BFRes$ci.ub, 2), nsmall=2)`, P-value = `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`).
+When we normalized the Shannon diversity data and pooled studies there was also no significant difference between non-obese and obese individuals (difference in mean z-score = `r format(round(combZHTtest$estimate[1]-combZHTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZHTtest$p.value, 2), nsmall=2)`).
+There was also no significant difference in the normalized B/F ratio between non-obese and obese individuals (difference in mean z-score = `r format(round(combZLogBFTtest$estimate[1]-combZLogBFTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`).
+Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.
 
 **Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome.
-Our meta-analysis supports the  hypothesis that there is a difference between non-obese and obese individuals in Shannon diversity and B/F ratio, but that the effect sizes are small.
+Our meta-analysis provides support for the null hypothesis that there is no difference between non-obese and obese individuals in Shannon diversity and B/F ratio, that the effect sizes are small, and that a very large number of indivdiuals would have to be tested to reach sufficient study power.
 
@@
 This analysis was performed using the vegan R package.
-Finally, we used the AUCRF R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguish non-obese and obese individuals.
+Finally, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguish non-obese and obese individuals.
+To generate Receiver Operater Characteristic (ROC) curves we used the pROC (v1.8) R package.
+A bonferroni adujusted p-value was used to assess differences between each study.
 
@@
 We then pooled Z-scores across studies and used a two-tailed T-test on the values to test for a significant difference in Z-scores among non-obese and obese individuals.
-To assess differences between the lean, overweight, and obese BMI categories we used ANOVA with a Tukey post hoc test.
-We also measured the Spearman correlation coefficient between the Z-scores for each metric and BMI values.
+
+In the final component of the analysis we performed power and sample number simulations for different effect sizes for each study.
+This was completed for all the different components measured as part of first strategy.
+We also calculated the actual sample size needed based on the effect size of each indivdiual study.
+To complete this analysis we used a combination of the pwr (v1.1-3) R package and base R functions.
 
@@
 We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 2 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix.
+Using a P-value cutoff for significance as less than 0.05 we found a only 1 significant result for OTU richness.
 There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio.
-The full summary of the exact P-values for each respective study and variable tested can be found in table 3.
+The full summary of the exact P-values for each respective study and variable tested can be found in table S1 of the online supplement.
 
@@
 We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)` +/- `r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`.
-We found that the total range for the total number of variables used for the classification ranged from 3 (Ross and Escobar) to 40 (Zupancic).
+We found that the range for the total number of variables used for the classification ranged from 6 (HMP study) to 42 (Zupancic).
 The total number of trees and node size was standardized to 1000 and 20 respectively for each data set.
+Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05, Figure 2).
 One aspect we found surprising was that there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study.
-The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/8 studies), an OTU that could be classified to the family Lachnospiraceae (6/8 studies), Clostridiales (6/8) and Shannon diversity metric (5/8 studies).
+The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/8 studies), an OTU that could be classified to the family Lachnospiraceae (6/8 studies), Clostridiales (6/8), Bacteroides (5/8), and Shannon diversity metric (4/8 studies).
+
+Using the top three common taxonomic classifications identified across studies we investigated this further.
+Although these taxonomic level classifications showed up commonly across studies, the overall direction and magnitude varied quite a bit depending on the study [Figure S1-3].
+Further, when the sequences were BLAST the identifications within each OTU within and between groups varied quite widely with many having relatively poor query coverage and identity [Table S2].
+When the sequence identity for each specific variable region was compared using the EMBOSS Needle pairwise sequence alignment (http://www.ebi.ac.uk/Tools/psa/emboss_needle/nucleotide.html) the V4 region had the highest similarity [Table S3] but all had greater than 4% dissimilarity.
 
@@
 We next moved on to the pooled meta-analysis.
-For the Shannon diversity RR, all studies but Wu et al had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 2].
-This was confirmed with the pooled analysis showing an increased RR of 1.16 (CI 1.01, 1.34) for obesity in the lower then median Shannon diversity group (P-value = 0.0412).
-When investigated the B/F ratio RR most studies showed that there was no significant difference between obese and non-obese individuals [Figure 3].
-The pooled RR was 0.98 (CI 0.85, 1.13) suggesting that a low B/F ratio does indeed have less obese individuals (P-value = 0.811).
+For the Shannon diversity RR, all studies but Wu, Goodrich, and Turnbaugh had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 3A].
+The was  pooled analysis showed an increased RR of `r format(round(10^test3R$LogFC[1],2), nsmall=2)` (CI `r format(round(10^ShannonRes$ci.lb, 2), nsmall=2)`, `r format(round(10^ShannonRes$ci.ub, 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was not significant (P-value = `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`).
+Similarily, when we investigated the B/F ratio RR  most studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`)[Figure 3B].
+The pooled RR was `r format(round(10^test3BF$LogFC[1],2), nsmall=2)` (CI `r format(round(10^BFRes$ci.lb, 2), nsmall=2)`, `r format(round(10^BFRes$ci.ub, 2), nsmall=2)`).
+The differences in RR that was observed also do not seem to be related to the variable region chosen [Figure 3].
 
 We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis.
-From the graph we can observe that there does not seem to be any significant  bias [Figure 4].
+From the graph we can observe that there does not seem to be any significant  bias [Figure S4].
 Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.
-
-The final forest plot that we used for this analysis investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC in the the separate study analysis looked when pooled together to assess the RR [Figure S4].
-There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.96 (CI 0.60, 1.52), P-value = 0.8575).
-Thus it would seem that although Ruminococcaceae is important for classification the directionality is study dependent [Figure S4].
 
@@
 
-Using the third approach with Z-score normalization we found that there was a significant difference in the Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A].
-Further there was also a significant difference in this normalized Shannon diversity value when the data was split between the three BMI groups (P-value = 0.03826) [Figure 5B].
-Using a Tukey post hoc test this difference was driven by the overweight versus the obese group (P-value = 0.03945).
-However, in contrast to this result we found that the B/F ratio, using the Z-score normalization, was not significant different between obese and non-obese individuals (P-value = 0.4516) [Figure 5C].
-Additionally, we did not observe a difference when the individuals were separated by BMI groups (P-value = 0.3317) [Figure 5D].
+Using the third approach with Z-score normalization we found that there was no significant difference in the Shannon diversity between the obese and non-obese group (P-value = `r format(round(combZHTtest$p.value, 2), nsmall=2)`) [Figure 5A].
+Additionally,  we found that the there was also no significant difference in B/F ratio, using the Z-score normalization, between obese and non-obese individuals (P-value = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`) [Figure 5B].
 
-Using a small (0.2), medium (0.5), and large (0.8) cohen value we estimated the power of the combined data set as well as the average cohen value from across studies for Shannon Diversity (0.06) and BF ratio (0.01).
+**Power and Sample Size Estimates and Simulations**
 
-small effect for combined data power calculation:
-`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.2, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`  
-medium effect for combined data power calculation:
-`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.5, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`  
-large effect for combined data power calculation:
-`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.8, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`  
-average effect for all Shannon Diversity:
-`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.06, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`
-average effect for all BF ratio:
-`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.01, sig.level = 0.05, alternative="two.sided")$power,2), nsmall=2)`
+For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 5A] compared to the B/F ratio [Figure 6A] differences.
+When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity [Figure 5A].
+However, this was not the case for the B/F ratio where even a 15% difference between non-obese and obese groups only resulted in an average power between studies of `r format(round(mean(ModelledPower15PerDiff$BFratio)*100, 2), nsmall=2)`% [Figure 6A].
+Looking at the sample sizes showed that on average to detect a 5% difference at 80% power between non-obese and obese individuals a total n of `r format(round(mean(RRSimNNeededH$RRFivePercentNH), 0), nsmall=0)` would be needed for each group [Figure 5B].
+In contrast in order to detect a 5% difference with 80% power in the B/F ratio a sample size of `r format(round(mean(RRSimNNeededBF$RRFivePercentNBF), 0), nsmall=0)` per group [Figure 6B].
+A breakdown of the power simulations and sample size estimates of the other variables and RR can be found in the online supplement.
+
+We also investigated if the combined total n for the third strategy was sufficiently powered.
+Using a small (0.2), medium (0.5), and large (0.8) cohen value we estimated the power of the combined data sets as well as the average cohen value from across studies for Shannon Diversity (0.06) and BF ratio (0.01).
+For the small effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.2, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
+For the medium effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.5, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
+For the large effect for combined data power calculation a value of
+`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.8, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
+The average actual cohens d value for Shannon Diversity for all 8 studies of 0.06 gave a power of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.06, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
+Finally, the average actual cohens d value for B/F ratio for all 8 studies of 0.01 yieled a power of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.01, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
 
@@
 ##### Discussion:
-
-However, other methods do exist, such as body adiposity index (BAI) but do not always correlate with important molecular measures traditionally associated with obesity [@lichtash_body_2013].
-
-* hyper variation between humans may mute the actual signal.
-
-* just because we can't find a human-wide signal, doesn't mean microbiome isn't important for weight
-
-* [@ley_microbial_2006] watched humans lose weight and communities change
-* [@turnbaugh_effect_2009] did fecal transplants (humans to mice) and transferred phenotype
-* [@koren_host_2012] did fecal transplants (pregnant women to mice) and transferred phenotype
 
@@
 Here we perform an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.
-We analyze the data using three different approaches.
+We analyzed the data using three different approaches and also performed post-hoc power and sample size calculations and simulations.
 The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses Z-score normalizations and does not use RR.
 
-Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the microbiome and obestiy could be observed.
-However, we found that the other two approaches, that used the pooled data, was able to identify a correlation between measures of the microbiome and obesity [Figure 2 & 3].
-For Shannon diversity both approaches using the pooled data found a correlation while only one of the approaches for the B/F ratio found a correlation.
-Taken together we believe that this analysis provides a strong preliminary finding that there is a small but detectable difference between non-obese and obese individuals with respect to both Shannon diversity and the B/F ratio.
-However, it should be noted that our analysis would indicate that the signal for Shannon diversity may be more robust than that of the B/F ratio since the two different pooled analysis approaches found Shannon diversity significant while only one did for the B/F ratio.
+Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the microbiome and obesityy could be observed.
+Further, even though random forest could identify specific variables with similar taxonomic classifications that resulted in good ROC curves [Figure 1] these variables varied widely by both their actual sequence and directionality.
+Similarily, we found that the other two approaches, that used the pooled data approach, was also not able to identify a correlation between measures of the microbiome and obesity [Figure 3 & 4].
+For these approaches both Shannon diversity and B/F ratio failed to have a significant correlation to obesity.
+In order to also assess the amount of our results that could be due to insufficient power we performed power and sample size analysis.
+Overall we found that the most studies for Shannon Diversity would be powered to find a 15% difference, however, the vast majority of reported differences between studies ranged from `r format(round(min(ActualDiff$H), 2), nsmall=2)`-`r format(round(max(ActualDiff$H), 2), nsmall=2)` percent.
+Although the effect sizes for the BF ratio were larger than that for Shannon Diversity [Figure 5A & 6A] the estimated power remained low.
+Additionaly, we would also need a relatively large sample size to be able to detect the actual differences in each study for both Shannon Diversity and B/F ratio [Figure 5B & 6B].
+Taken together we believe that this analysis builds on the previous two re-anlaysis and provides additional information as to whether there is a difference between non-obese and obese human individuals with respect to both Shannon diversity and the B/F ratio.
 
@@
 Second, although we make attempts to normalize the data for the pooled analysis so that we can compare the microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.
-However, what our analysis shows is that despite these differences, we are still able to identify an overall signal from the data.
-Third, this meta-analysis and review of the data to date is still rather preliminary.
-Only a total of 1542 individuals were included in the pooled analysis which is relatively small compared to other meta-analysis in non-microbiome fields of study.
-It will be necessary to continuously update this analysis, as more studies become available, in order to obtain a more accurate estimate of the true effect that obesity has on the microbiome within humans.
+To help mitigate some of these problems we analyzed each data set separately and processed all samples through a uniform pipeline in an attempt to mitigate some of the variation that might occur due to sequence processing.
+It should also be noted that we also do not see any specific bias in the forest plots [Figure 3] based on hypervariable region.
+Further many of the studies show very similar actual effect sizes for Shannon Diversity difference between non-obese and obese individuals.
+However, even with all these precautions and safety checks we do realize that it may not be possible to compare these studies together and that further studies need to be completed to verify the observations made in this report.
 
-Fourth, although a significant result exists for the RR for both Shannon diversity and the B/F ratio, it is small.
-This is also the case for the Z-score normalization analysis of the Shannon diversity.
-As an example, a lower than median Shannon diversity only has an overall 18% increase of obesity in it versus the higher than median Shannon diversity group.
-So even though there is a detectable signal it remains to be seen if this is a biologically significant phenomenon that could have a tangible effect on obesity.
+Third, although we may not have detected a signal between non-obese and obese indivdiuals our results indicate that we may not have been sufficiently powered to do so.
+However, taking both the sample size calculations and power calculations together would indicate that if there is a true difference to be found between non-obese and obese individuals this difference would be very small.
+Additionally, due to the very small signal another important question would be whether or not this difference is truly biologically meaningful.
 
-Despite the many problems that could add more noise to the measurements used, this study provides a reasonable first pass attempt to pool this data together in an effort to find a correlation between the microbiome and obesity.
-In doing so we believe that it provides a tanatalizing result that despite the large differences between studies it is still possible to detect a correlation between the microbiome and obesity.
-However, we emphaszie that it should still be viewed with caution since there are still many aspects of this analysis that could be improved upon.
-It truly is only a first attempt at trying to summerize and analyze the pooled data that is available.
-We hope that this study helps to add important information and clarity to the on-going dialogue of the microbiome's role in obesity.
+Fourth, although the ROC curves for each study were relatively good there was very little actual overlap between studies based on sequence similarity and directionality.
+This would suggest that within each study classification of obese and non-obese is relatively strong but not generalizable across studies.
+This might indicate an ancestry, ethnicity, or a specific environmental effect such as diet difference between study populations that might cause this dissimilarity.
+It is also possible that the random forest model overfit the data making it less generalizable across studies.
+
+Although we failed to find an effect it is not realistic to necessarily state that there is no microbiome impact on obesity.
+There is a large body of very strong evidence in mice on the effect the microbiome has on obesity [@turnbaugh_effect_2009; @koren_host_2012].
+It is not possible to discount these findings and it may be the case that in humans there is just too many other variables and factors that effect the microbiome which makes the signal to hard to detect in a cross sectional manner.
+Some well designed studies in humans and diet, in the pasat, have shown that the noise inherent to microbiome studies can be overcome with very careful attention made to controlling the study conditions [@ley_microbial_2006].
+
+Overall, this study provides a reasonable follow up to the two previous re-analysis studies to assess the correlation between the microbiome and obesity.
+In doing so we believe that we have provided evidence that if there is an effect size present it is small and will take many more studies and samples to be analyzed to confirm a difference between non-obese and obese individuals.
 
@@
 We would also like to thank Shawn Whitefield for her help with making sure the manuscript adhered to the guidelines set out for a meta-analysis.
-Finally, we would like to thank Rick Bushman, Hongzhe Li, and Pixu Shi in agreeing to release and helping to get a hold of the subject specific metadata from the Wu et al study.
+We would also like to thank Rick Bushman, Hongzhe Li, and Pixu Shi in agreeing to release and helping to get a hold of the subject specific metadata from the Wu et al study.
+Finally, we would like to thank all the authors who have made there data publicly available.
+Without this initial forethought studies such as these would not be possible.
 
@@
 
-```{r Table_5, echo=FALSE, message=FALSE, warning=FALSE}
-
-source("code/MakeFiguresCode.R")
-grid.arrange(a2, b2, nrow=2)
-
-```
-**Figure 5. Summary of Power and Sample Size Simulations for Shannon Diversity for Non-obese versus Obese**
-
-```{r Table_6, echo=FALSE, message=FALSE, warning=FALSE}
-
-grid.arrange(c2, d2, nrow=2)
-
-```
-**Figure 6. Summary of Power and Sample Size Simulations for Bacteroidetes/Firmicutes Ratio for Non-obese versus Obese**
-
 *******
@@
 ```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+combinedData <- read.csv("results/tables/ROCcombinedData.csv", stringsAsFactors = FALSE)
 
-StudyNo <- as.numeric(ShannonRRTable$X)
-Study <- as.character(ShannonRRTable$Study)
-Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
-Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
-tpos <- as.numeric(ShannonRRTable$tposH) #obese in low diversity group
-tneg <- as.numeric(ShannonRRTable$tnegH) #nonobese in low diversity group
-cpos <- as.numeric(ShannonRRTable$cposH) #obese in high diversity group
-cneg <- as.numeric(ShannonRRTable$cnegH) #nonobese in high diversity group
-RR <- as.numeric(ShannonRRTable$RRH)
-low <- as.numeric(ShannonRRTable$lowH)
-high <- as.numeric(ShannonRRTable$highH)
+diag <- data.frame(x=seq(1, 0), y = seq(0, 1))
+diag$Study <- "Control"
 
-numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
-raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
-
-dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
-k <- length(dat$StudyNo)
-dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
-dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
-dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
-dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
-
-#Used to generate the RE Model estimate based on all the studies
-res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
-RRSDPvalue <- anova.rma(res)
-
-#To create a graph visualize the data
-forest(res, at=log(c(0.3, 1, 5)), atransf=exp, mlab= "Pooled Relative Risk", xlab="RR of Obesity in high versus low Shannon diversity based on Median", cex=0.8)
-par(font=2) #Switch to bold
-
-text(5.6, 10.3, "Relative Risk [95% CI]", pos=2, cex=0.8)
-text(-6.5, 10.3, "Study, Year, and n", pos=4, cex=0.8)
+rocPlot <- ggplot(data=combinedData, 
+                  aes(x = Sensitivity, y = Specificity))
+rocPlot + geom_line(aes(group=Study, color=Study)) + 
+  scale_colour_brewer(palette="Set1") + 
+  scale_x_continuous(trans = "reverse") + 
+  geom_line(data=diag, aes(x=x, y=y), linetype="dashed") + 
+  theme(axis.line=element_line(colour="black"),
+       panel.grid.major = element_blank(),
+       panel.grid.minor = element_blank(),
+       panel.border = element_blank(),
+       panel.background = element_blank(), 
+       legend.key = element_rect(fill = "white"))
 
@@
 
-**Figure 2: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity.**  Groups were divided for each study on high and low Shannon Diversity groups based on the median for that study.
-The overall pooled relative risk was 1.16 for the low diversity group (P-value = `r signif(RRSDPvalue$QMp, 3)`).
+**Figure 2: Receiver Operating Characteristic Curve of the Random Forest Analysis of Each Study.** After adjusting for multiple comparisons there was no significant difference between any of the 8 studies examined (P-value > 0.05).
 
@@
 
-StudyNo <- as.numeric(BFRatioRRTable$X)
-Study <- as.character(BFRatioRRTable$Study)
-Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
-Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
-tpos <- as.numeric(BFRatioRRTable$tposBF) #obese in low diversity group
-tneg <- as.numeric(BFRatioRRTable$tnegBF) #nonobese in low diversity group
-cpos <- as.numeric(BFRatioRRTable$cposBF) #obese in high diversity group
-cneg <- as.numeric(BFRatioRRTable$cnegBF) #nonobese in high diversity group
-RR <- as.numeric(BFRatioRRTable$RRBF)
-low <- as.numeric(BFRatioRRTable$lowBF)
-high <- as.numeric(BFRatioRRTable$highBF)
+grid.arrange(HRRG1, BFRRG1, ncol=2)
 
-numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
-raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
-
-dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
-k <- length(dat$StudyNo)
-dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
-dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
-dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
-dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
-
-#Used to generate the RE Model estimate based on all the studies
-res2 <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
-RRBFPvalue <- anova.rma(res2)
-
-#To create a graph visualize the data
-forest(res2, at=log(c(0.1, 1, 3)), atransf=exp, mlab= "Pooled Relative Risk", xlab="RR of Obesity in high versus low B/F ratio based on Median", cex=0.8)
-par(font=2) #Switch to bold
-
-text(5.6, 10.3, "Relative Risk [95% CI]", pos=2, cex=0.8)
-text(-6.0, 10.3, "Study, Year, and n", pos=4, cex=0.8)
 ```
 
-**Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low B/F ratio groups based on the median for that specific study.
-The overall pooled relative risk was 0.98 for the low diversity group (P-value = `r signif(RRBFPvalue$QMp, 3)`).
+**Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity or Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low Shannon Diversity or Bacteroidetes/Firmicutes (B/F) ratio groups based on the median for that study.
+Red represents studies that used the 16S rRNA V4 region, yellow represents studies that used a V1-V3 approach, blue represents studies that used either a V1-V2 or just a V2 approach, and orange represents studies that used the V3-V5 region.
+**A)** The overall pooled relative risk for Shannon Diversity was `r format(round(10^test3R$LogFC[1],2), nsmall=2)` with a P-value of `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`.
+**B)** The overall pooled relative risk for the B/F ratio was  `r format(round(10^test3BF$LogFC[1],2), nsmall=2)` with a P-value of `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`.
 
@@
 There was a significant difference between non-obese (n = `r length(which(group.data.test$obese == "No"))`) and obese groups (n = `r length(which(group.data.test$obese == "Yes"))`) (P-value = `r signif(combZHTtest$p.value, 3)`).
-**B)** Z-score Normalized Shannon Diversity by BMI Group.
-Total n for lean, overweight, and obese groups are `r length(which(group.data.test$BMI.class == "Lean"))`, `r length(which(group.data.test$BMI.class == "Overweight"))`, `r length(which(group.data.test$BMI.class == "Obese"))` respectively.
-Using an ANOVA with tukey post-hoc testing there was a significant diference between the overweight and obese group (P-value = `r signif(combZHTukeyFinal[3, 4], 3)`) but no difference between lean and obese group (P-value = `r signif(combZHTukeyFinal[2, 4], 3)`).
-**C)** Z-score Normalized B/F Ratio and Obesity.
+**B)** Z-score Normalized B/F Ratio and Obesity.
 There was no significant difference between the non-obese (n = `r length(which(group.data.test$obese == "No"))`) and obese (n = `r length(which(group.data.test$obese == "Yes"))`) group based on B/F ratio (P-value = `r signif(combZLogBFTtest$p.value, 3)`).
-**D)** Z-score Normalized B/F Ratio by BMI Group.
-Total n for lean, overweight, and obese groups are `r length(which(group.data.test$BMI.class == "Lean"))`, `r length(which(group.data.test$BMI.class == "Overweight"))`, `r length(which(group.data.test$BMI.class == "Obese"))` respectively.
-There was no significant difference between any of the three groups based on ANOVA with Tukey post-hoc testing (P-value = `r signif(combZLogBFANOVA[1, 5], 3)`).
+
+```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+
+source("code/MakeFiguresCode.R")
+
+grid.arrange(a2, b2, nrow=2)
+
+```
+
+**Figure 5.
+Summary of Power and Sample Size Simulations for Shannon Diversity for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
+
+```{r Figure_6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+
+grid.arrange(c2, d2, nrow=2)
+
+```
+
+**Figure 6.
+Summary of Power and Sample Size Simulations for Bacteroidetes/Firmicutes Ratio for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
 

### 2016-03-25 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
manuscript update writing

@@
 
-```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
 
@@
 
-```{r Figure_4, echo=FALSE}
+```{r Figure_4, echo=FALSE, fig.path="results/figures/"}
 par(mfrow=c(1,2))

### 2016-03-28 code/sequenceProcessing.Rmd
add where to find batch files
final mods to seq processing rmd

@@
 Most of the data in the SRA were stored in the Fastq format and this step was not possible.
+The batch files can be found in the data/process directory under the respective study directory.
 <br>
@@
 sed 's/\s.*$//' test2.fasta > test3.fasta
-paste -d" " test2.fasta test3.fasta > V2.groups''
+sed 's/.*_//' test3.fasta > test4.fasta
+paste -d"\t" test3.fasta test4.fasta > V2.groups''
 ```

### 2016-03-28 code/numberSimulationsSupplement.R
code for n sim suppl

@@ -1,0 +1,320 @@
+## Power Calculations Variations: Estimating needed n
+## Obesesity and the bacterial microbiome: Supplemental variables
+## Bacteroidetes, Firmicutes, OTU Richness, Evenness
+## Marc Sze
+## March 28, 2016
+
+
+library(statmod)
+library(pwr)
+source("code/UsedFunctions.R")
+
+ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
+BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
+PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
+SummaryStatsByObeseGroup <- read.csv("results/tables/denovoSSbyObeseGroup.csv")
+SummaryStatsByStudy <- read.csv("results/tables/denovoSSbyStudy.csv")
+ActualDiff <- read.csv("results/tables/denovoActualDiff.csv")
+
+Study <- as.character(PowerTable$Study)
+
+######### ACTUAL ###########################################################
+
+
+BacterActualNeededN <- rep(0, 8)
+
+for(i in 1:8){
+  cohensD <- (SummaryStatsByObeseGroup$MeanNonObeseB[i] -  
+                SummaryStatsByObeseGroup$MeanObeseB[i]) / 
+    SummaryStatsByStudy$sdB[i] 
+  # This needs to change based on study
+  
+  BacterActualNeededN[i] <- pwr.t.test(d=cohensD, 
+                                  power=0.8, sig.level=0.05, type="two.sample", 
+                                  alternative="two.sided")$n
+}
+
+FirmActualNeededN <- rep(0, 8)
+
+for(i in 1:8){
+  cohensD <- abs((SummaryStatsByObeseGroup$MeanNonObeseF[i] -  
+                    SummaryStatsByObeseGroup$MeanObeseF[i])) / 
+    SummaryStatsByStudy$sdF[i]  
+  # This needs to change based on study
+  
+  FirmActualNeededN[i] <- pwr.t.test(d=cohensD, 
+                                   power=0.8, sig.level=0.05, type="two.sample", 
+                                   alternative="two.sided")$n
+}
+
+
+SActualNeededN <- rep(0, 8)
+
+for(i in 1:8){
+  cohensD <- abs((SummaryStatsByObeseGroup$MeanNonObeseS[i] -  
+                    SummaryStatsByObeseGroup$MeanObeseS[i])) / 
[…]

### 2016-03-29 code/MakeFiguresCodeSupplemental.R
update code for suppl

@@ -18,6 +18,8 @@ SimulatedN_B <- read.csv("results/tables/denovoSimulatedN_B.csv")
 SimulatedN_F <- read.csv("results/tables/denovoSimulatedN_F.csv")
 SimulatedN_S <- read.csv("results/tables/denovoSimulatedN_S.csv")
 SimulatedN_J <- read.csv("results/tables/denovoSimulatedN_J.csv")
+ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
+BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
 
 ##################### OTU RICHNESS ###########################################
 
@@ -266,8 +268,501 @@ jn2 <- jn1 + geom_jitter(aes(colour = Study), size=5, alpha=7/10,
   ggtitle('B') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
 
 
+##################### FIRMICUTES ###########################################
+
+# A data frame that has all the necessary data for power
+Study <- as.character(PowerTable$Study)
+ActualDiffF <- as.numeric(format(round(ActualDiff$firm, 3), nsmall=3))
+FPvalue <- as.numeric(format(round(OverallPValue$Firmicutes, 2), nsmall=2))
+OneP <- as.numeric(format(round(ModelledPower1PerDiff$firm, 3), nsmall=3))
+FiveP <- as.numeric(format(round(ModelledPower5PerDiff$firm, 3), nsmall=3))
+TenP <- as.numeric(format(round(ModelledPower10PerDiff$firm, 3), nsmall=3))
+FifteenP <- as.numeric(format(round(ModelledPower15PerDiff$firm, 3), nsmall=3))
+
+DataTable1 <- as.data.frame(
+  cbind(FPvalue, ActualDiffF, OneP, FiveP, TenP, FifteenP))
+DataTable1$Study <- Study
+TableLegend <- paste(as.character(Study), ", P-value=", FPvalue, sep="")
+DataTable1$Legend <- TableLegend
+
+test <- melt(DataTable1, id=c("ActualDiffF", "FPvalue", "Study", "Legend"))
+
+test$EstDiff[test$variable=="OneP"] <- 1
+test$EstDiff[test$variable=="FiveP"] <- 5
+test$EstDiff[test$variable=="TenP"] <- 10
+test$EstDiff[test$variable=="FifteenP"] <- 15
+
[…]

### 2016-03-29 code/StrategydenovoAnalysisMetaHitDel.R

@@ -2058,9 +2058,12 @@ lowH <- c(BaxterHRR$lowCI, RossHRR$lowCI, GoodrichHRR$lowCI, EscobarHRR$lowCI,
 highH <- c(BaxterHRR$highCI, RossHRR$highCI, GoodrichHRR$highCI, 
            EscobarHRR$highCI, ZupancicHRR$highCI, HMPHRR$highCI, WuHRR$highCI, 
            TurnbaughHRR$highCI)
+pValueH <- c(BaxterHRR$pValue, RossHRR$pValue, GoodrichHRR$pValue, 
+           EscobarHRR$pValue, ZupancicHRR$pValue, HMPHRR$pValue, WuHRR$pValue, 
+           TurnbaughHRR$pValue)
 
 ShannonRRTable <- as.data.frame(cbind(tposH, tnegH, cposH, cnegH, 
-                                      RRH, lowH, highH))
+                                      RRH, lowH, highH, pValueH))
 ShannonRRTable$Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", 
                           "HMP", "Wu", "Turnbaugh")
 write.csv(ShannonRRTable, "results/tables/denovoShannonRRTable.csv")
@@ -2086,8 +2089,12 @@ highBF <- c(BaxterBFRR$highCI, RossBFRR$highCI, GoodrichBFRR$highCI,
             EscobarBFRR$highCI, ZupancicBFRR$highCI, HMPBFRR$highCI, 
             WuBFRR$highCI, TurnbaughBFRR$highCI)
 
+pValueBF <- c(BaxterBFRR$pValue, RossBFRR$pValue, GoodrichBFRR$pValue, 
+           EscobarBFRR$pValue, ZupancicBFRR$pValue, HMPBFRR$pValue, 
+           WuBFRR$pValue, TurnbaughBFRR$pValue)
+
 BFRatioRRTable <- as.data.frame(cbind(tposBF, tnegBF, cposBF, cnegBF, RRBF, 
-                                      lowBF, highBF))
+                                      lowBF, highBF, pValueBF))
 BFRatioRRTable$Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", 
                           "HMP", "Wu", "Turnbaugh")
 write.csv(BFRatioRRTable, "results/tables/denovoBFRatioRRTable.csv")

### 2016-03-29 microbiomeAndobesitySupplemental.Rmd
supp add all proposed figs

@@
 
-**Figure 4: Funnel Plot of the Shannon Diversity and Bacteroidetes/Firmicutes Ratio Relative Risk.**  **A)** Overall there does not seem to be any bias associated with the studies selected for the Shannon Diversity analysis with studies falling on either side of the predicted value and those with smaller total n falling further away from this.
+**Figure S4: Funnel Plot of the Shannon Diversity and Bacteroidetes/Firmicutes Ratio Relative Risk.**  **A)** Overall there does not seem to be any bias associated with the studies selected for the Shannon Diversity analysis with studies falling on either side of the predicted value and those with smaller total n falling further away from this.
 **B)** For the Bacteroidetes/Firmicutes ratio there does not seem to be any associated bias.
@@
 
+```{r Figure_S5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+
+source("code/MakeFiguresCodeSupplemental.R")
+library(gridExtra)
+
+grid.arrange(s2, sn2, nrow=2)
+
+```
+
+**Figure S5.
+Summary of Power and Sample Size Simulations for OTU Richness (S) for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
+
+```{r Figure_S6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+
+grid.arrange(j2, jn2, nrow=2)
+
+```
+
+**Figure S6.
+Summary of Power and Sample Size Simulations for Evenness (J) for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
+
+```{r Figure_S7, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+
+grid.arrange(b2, bn2, nrow=2)
+```
+
+**Figure S7.
+Summary of Power and Sample Size Simulations for Bacteroidetes for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
+
+```{r Figure_S8, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+
+grid.arrange(f2, fn2, nrow=2)
+```
+
+**Figure S8.
+Summary of Power and Sample Size Simulations for Firmicutes for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
+
+```{r Figure_S9, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+
+grid.arrange(RRH2, HRRn2, nrow=2)
+```
+
+**Figure S9.
+Summary of Power and Sample Size Simulations for Relative Risk  for Obesity based on Shannon Diversity**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
+
+```{r Figure_S10, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+
+grid.arrange(RRBF2, BFRRn2, nrow=2)
+```
+
+**Figure S10.
+Summary of Power and Sample Size Simulations for Relative Risk  for Obesity based on B/F Ratio**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
+
 *******

### 2016-03-30 code/sequenceProcessing.Rmd
minor sequence processing update

@@
 The batch files can be found in the data/process directory under the respective study directory.
+Our specific server used a PBS script for job submission and an example of it can be found below.
+If you want to copy it simply save it as a .pbs file.
+
+```{r, engine='bash', eval=FALSE}
+#!/bin/sh
+#PBS -l nodes=1:ppn=8,mem=44gb
+#PBS -l walltime=500:00:00
+#PBS -j oe
+#PBS -m abe
+#PBS -V
+#PBS -M email@location.com
+#PBS -q first
+
+echo "ncpus-2.pbs"
+cat $PBS_NODEFILE
+qstat -f $PBS_JOBID
+
+cd $PBS_O_WORKDIR
+
+NCPUS=`wc -l $PBS_NODEFILE | awk '{print $1}'`
+
+/mnt/EXT/Schloss-data/bin/mothur Turnbaugh1.batch
+
+echo "qsub working directory absolute is"
+echo $PBS_O_WORKDIR
+exit 0
+
+```
+
 <br>

### 2016-03-30 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
Fix typos

@@
 *Current Address:*  
-1526 MSRB I  
+1520A Medical Science Research Building I
+1500 W. Medical Center Dr.
 Ann Arbor, MI  
@@
 Email: pschloss@umich.edu  
-Phone: 734-936-2951    
+Phone: 734-647-5801    
 <br>

### 2016-03-31 microbiomeAndobesitySupplemental.Rmd
update code for pdf

@@
 
-```{r Figure_S4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_S4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 funnel(res2, main="B")
+
 ```
@@
 
-```{r Figure_S5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_S5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_S6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S7, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_S7, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S8, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_S8, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S9, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_S9, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S10, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_S10, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 

### 2016-03-31 code/CommonClassOTUsMetaHitDel.R

@@ -345,6 +345,7 @@ j1 <- ggBaseBox(ZupancicMicrobiome, ZupancicDemo,
 k1 <- ggBaseBox(HMPmicrobiome, HMPDemo, "Otu000293", "HMP", log=TRUE)
 l1 <- ggBaseBox(WuMicrobiome, WuDemo, "Otu00191", "Wu", log=TRUE)
 
+pdf("results/figures/suppFigClostridiales.pdf")
 png("results/figures/suppFigClostridiales.png")
 grid.arrange(a1, b1, c1, d1, e1, f1, g1, h1, i1, h1, j1, k1, l1)
 dev.off()
@@ -382,6 +383,7 @@ l1 <- ggBaseBox(WuMicrobiome, WuDemo, "Otu00229", "Wu", log=TRUE)
 m1 <- ggBaseBox(WuMicrobiome, WuDemo, "Otu00005", "Wu", log=TRUE)
 n1 <- ggBaseBox(WuMicrobiome, WuDemo, "Otu00067", "Wu", log=TRUE)
 
+pdf("results/figures/suppFigLachno.pdf")
 png("results/figures/suppFigLachno.png")
 grid.arrange(a1, b1, c1, d1, e1, f1, g1, h1, i1, h1, j1, k1, l1, m1, n1)
 dev.off()
@@ -420,6 +422,7 @@ k1 <- ggBaseBox(ZupancicMicrobiome, ZupancicDemo,
 l1 <- ggBaseBox(HMPmicrobiome, HMPDemo, "Otu000252", "HMP", log=TRUE)
 m1 <- ggBaseBox(WuMicrobiome, WuDemo, "Otu00123", "Wu", log=TRUE)
 
+pdf("results/figures/suppFigRumi.pdf")
 png("results/figures/suppFigRumi.png")
 grid.arrange(a1, b1, c1, d1, e1, f1, g1, h1, i1, h1, j1, k1, l1, m1)
 dev.off()

### 2016-04-04 submission/Sze_Obesity_PeerJ_2016.Rmd
Light editing of text

--- 2016-03-30 Sze_ObesityMicrobiomeMetaNoMetaHit_PeerJ_2016.Rmd
@@
 ## Introduction
+
 Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014].
@@
 One critique of these re-analysis studies is that they did not aggregate the results across studies to increase the effective sample size.
-It is possible that there are small associations within each study that would not be significant, because the studies lacked sufficient statistical power.
+It is possible that there are small associations within each study that would not be significant because the studies lacked sufficient statistical power.
 The purpose of this study was to perform a meta-analysis of the association between differences in the microbiome and obesity status by analyzing and applying a more systematic and synthetic approach than was used previously.
-Our goal was to provide additional clarity on the possible relationship between community diversity, B/F, and community composition and obesity status.
+Furthermore, we sought to provide additional clarity on the possible relationship between community diversity, B/F, and community composition and obesity status and the ability to detect significant differences in these variables if they were actually present.
 
 ## Methods
+
 ***Literature Review and Study Inclusion***
@@
 All sequence data were publicly available and were downloaded from the NCBI Sequence Read Archive, the European Nucleotide Archive, or the investigators' personal website (https://gordonlab.wustl.edu/TurnbaughSE\_10\_09/STM\_2009.html).
-The nine studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011;
-@human_microbiome_project_consortium_structure_2012] or Illumina sequencing [@goodrich_human_2014; Baxter].
+The nine studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012] or Illumina sequencing [@goodrich_human_2014; Baxter].
 All of these studies used amplification-based 16S rRNA gene sequencing.
 Among the studies that sequenced the 16S rRNA gene, the researchers targeted the V1-V2  [@wu_linking_2011], V1-V3 [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], V3-V5 [@human_microbiome_project_consortium_structure_2012], and V4 [@goodrich_human_2014] regions.
-For those studies where multiple regions were sequenced,  we selected the region that corresponded to the largest number of subjects [@turnbaugh_core_2009;
-@human_microbiome_project_consortium_structure_2012].
+For those studies where multiple regions were sequenced,  we selected the region that corresponded to the largest number of subjects [@turnbaugh_core_2009; @human_microbiome_project_consortium_structure_2012].
 We processed the 16S rRNA gene sequence data as described in the original studies.
@@
 Finally, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguish non-obese and obese individuals.
-To generate Receiver Operater Characteristic (ROC) curves we used the pROC (v1.8) R package.
-A bonferroni adujusted p-value was used to assess differences between each study.
+To generate Receiver Operator Characteristic (ROC) curves we used the pROC (v1.8) R package.
+A Bonferroni-adjusted p-value was used to assess differences between each study.
 
@@
 This was completed for all the different components measured as part of first strategy.
-We also calculated the actual sample size needed based on the effect size of each indivdiual study.
+We also calculated the actual sample size needed based on the effect size of each individual study.
 To complete this analysis we used a combination of the pwr (v1.1-3) R package and base R functions.
 
-## Results:
+## Results
 
 ***Separate Analysis***
-We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
+We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
 Using a P-value cutoff for significance as less than 0.05 we found a only 1 significant result for OTU richness.
@@
 
-**Power and Sample Size Estimates and Simulations**
+***Power and Sample Size Estimates and Simulations***
 
@@
 
-*******
-##### Discussion:
+## Discussion:
 
@@
 
-Third, although we may not have detected a signal between non-obese and obese indivdiuals our results indicate that we may not have been sufficiently powered to do so.
+Third, although we may not have detected a signal between non-obese and obese individuals our results indicate that we may not have been sufficiently powered to do so.
 However, taking both the sample size calculations and power calculations together would indicate that if there is a true difference to be found between non-obese and obese individuals this difference would be very small.
@@
 
-*******
-##### Acknowledgements:
+## Acknowledgements
 
-The authors would like to thank Nielson Baxter for his help and suggestions during the sequence analysis portion.
-We would also like to thank Shawn Whitefield for her help with making sure the manuscript adhered to the guidelines set out for a meta-analysis.
-We would also like to thank Rick Bushman, Hongzhe Li, and Pixu Shi in agreeing to release and helping to get a hold of the subject specific metadata from the Wu et al study.
+The authors would like to thank Nielson Baxter and Shawn Whitefield for their suggestions on the manuscript.
 Finally, we would like to thank all the authors who have made there data publicly available.

### 2016-04-04 submission/Sze_Obesity_PeerJ_2016_Supp.Rmd
Light editing of text

--- 2016-03-29 microbiomeAndobesitySupplemental.Rmd
@@
 
-The inital search strategy included looking for all papers that initially fit under the below NCBI PubMed advanced search criteria.
+The initial search strategy included looking for all papers that initially fit under the below NCBI PubMed advanced search criteria.
 The terms included in this criteria were that the manuscript had to have "Bacterial Microbiome" and "Obesity, BMI, bmi, obesity" in their manuscript criteria, it was not published more than 10 years ago, they were not review articles, and it contained research on humans only.
@@
 
-From this total we browsed abstracts for mention of stool or feces examination, that did not involve children, was not a clinical trial for probiotics or other diet related treatments, did not only have participants with inflammatory bowel disease, the articles were in English, did not only use PCR, qPCR, or RT-PCR only for their analysis, and sequecning that used only clone libraries.
+From this total we browsed abstracts for mention of stool or feces examination, that did not involve children, was not a clinical trial for probiotics or other diet related treatments, did not only have participants with inflammatory bowel disease, the articles were in English, did not only use PCR, qPCR, or RT-PCR only for their analysis, and sequencing that used only clone libraries.
 This ultimately excluded all but a total of 11 studies.
@@
 They also did not have obesity as part of their results in the actual published manuscript.
-The second study was excluded because they did not have BMI information available and when contacted the authors never returned any correspondance [@zeevi_personalized_2015].
-A thrid study was excluded since it did not use 16S rRNA gene sequencing for their bacterial microbiome analysis [@arumugam_enterotypes_2011]
+The second study was excluded because they did not have BMI information available and when contacted the authors never returned any correspondence [@zeevi_personalized_2015].
+A third study was excluded since it did not use 16S rRNA gene sequencing for their bacterial microbiome analysis [@arumugam_enterotypes_2011]
 
@@
 * Sequencing Data not publicly available for download
-* BMI not available and authors do not return correspondance
+* BMI not available and authors do not return correspondence
 * Samples were not stool or feces

### 2016-04-04 submission/Sze_Obesity_PeerJ_2016.Rmd
Work with YAML header material

@@
 ---
-title: "Looking for a Signal in the Noise: Revisiting Obesity and the microbiome"
+title: "Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome"
 output:
@@
     keep_md: yes
-  word_document:
-    reference_docx: manuscript_template.docx
-bibliography: metaAnalysis.obesity.microbiome.bib
+bibliography: submission/references.bib
 ---

### 2016-04-04 submission/Sze_Obesity_PeerJ_2016_Supp.Rmd
Work with YAML header material

@@
 ---
-title: "Bacterial Microbiome and Obesity Meta-analysis Supplemental"
-author: "Marc Sze"
-date: "January 26, 2016"
+title: "Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome"
 output:
@@
     keep_md: yes
-bibliography: metaAnalysis.obesity.microbiome.bib
+bibliography: submission/references.bib
 ---
-<br><br>
+
+Marc A Sze<sup>1</sup> and Patrick D Schloss<sup>1</sup>
 

### 2016-04-04 submission/Sze_Obesity_PeerJ_2016.Rmd

@@
 
-```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 combinedData <- read.csv("results/tables/ROCcombinedData.csv", stringsAsFactors = FALSE)
@@
 
-```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 # Section A
@@
 
-```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
+```{r Figure_6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 

### 2016-05-03 code/old/StrategydenovoAnalysisMetaHitDel.R
fixed obese non-obese class problem in OLD

@@ -494,9 +494,11 @@ metadata <- EditTable(metadata, 7, delRange=0)
 microbiome <- read.table("data/process/Goodrich/GoodrichGoodSub.shared", header=T)
 microbiome <- EditTable(microbiome, 2, delRange=c(1:3))
 keep  <- which(metadata$body_mass_index_s != "<not provided>")
-microbiome<- microbiome[keep, ]
+keep2 <- rownames(metadata[which(
+  metadata$body_mass_index_s != "<not provided>"), ])
+microbiome<- microbiome[keep2, ]
 metadata <- metadata[keep, ]
-rm(keep)
+rm(keep, keep2)
 
 #generate alpha diversity measures with vegan
 alpha.test <- makeAlphaTable(microbiome)
@@ -531,7 +533,7 @@ obese <- factor(metadata$obese)
 
 ##Test BMI versus alpha diversity and phyla
 
-goodrichH <- wilcox.test(alpha.test$H ~ obese) #P-value=0.614
+goodrichH <- wilcox.test(alpha.test$H ~ obese) #P-value=0.001384
 MeanNonObeseH <- c(MeanNonObeseH, 
                    mean(alpha.test$H[which(metadata$obese == "No")]))
 MeanObeseH <- c(MeanObeseH, 
@@ -1785,12 +1787,14 @@ metadata <- read.csv("data/process/Turnbaugh/turnbaugh.metadata.csv")
 metadata <- EditTable(metadata, 4, delRange=4)
 
 keep1 <- which(metadata$Sample == 1) # n =146
+keep2 <- rownames(metadata[which(metadata$Sample == 1),])
 
 #use the first sampling
 #subset data for only the first sampling
 metadata <- metadata[keep1, ]
-microbiome <- microbiome[keep1, ]
-rm(keep1)
+microbiome <- microbiome[keep2, ]
+rm(keep1, keep2)
+
 
 #generate alpha diversity measures with vegan
 alpha.test <- makeAlphaTable(microbiome)
@@ -1822,7 +1826,7 @@ bmi <- metadata$BMI.category
 
 ##Test BMI versus alpha diversity and phyla
 
-turnbaughH <- wilcox.test(alpha.test$H ~ obese) #P-value=0.9699
+turnbaughH <- wilcox.test(alpha.test$H ~ obese) #P-value=0.1077
[…]

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
modifying data tables used pt1

@@
 
+alpha_tests.summary <- read.table("../data/process/alpha_tests.summary", header=T)
+alpha_composite.summary <- read.table("../data/process/alpha_composite.summary")
+alpha_power.obs <- read.table("../data/process/alpha_power.observed", 
+                              header = T, row.names = 1)
+alpha_power.predicted <- read.table("../data/process/alpha_power.predicted", 
+                                    header = T)
+
+beta_tests.summary <- read.table("../data/process/beta_tests.summary", 
+                                 header=T, row.names = 1)
+rand_forest.genus.roc_data <- read.table("../data/process/random_forest.genus.roc_data", header=T)
+
+rand_forest.genus.summary <- read.table("../data/process/random_forest.genus.summary", 
+                                        header = T, row.names = 1)
+
+rand_forest.genus.train_test <- 
+  read.table("../data/process/random_forest.genus.train_test", header = T)
+
+rand_forest.otu.roc_data <- 
+  read.table("../data/process/random_forest.otu.roc_data", header = T)
+
+rand_forest.otu.summary <- 
+  read.table("../data/process/random_forest.otu.summary", 
+             header = T, row.names = 1)
+
+RR_composite <- read.table("../data/process/relative_risk.composite", header = T)
+RR_summary <- read.table("../data/process/relative_risk.summary", header = T)
+RR_pow.obs <- read.table("../data/process/rr_power.observed", 
+                         header = T, row.names = 1)
+RR_pow.pred <- read.table("../data/process/rr_power.predicted", header = T)
+
+# Hold on to these for now as I update the rest of the rmd.
 demographics <- read.csv("results/tables/denovodemographicsTable.csv")

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
incorporating Pat values into Rmd

@@
 # Hold on to these for now as I update the rest of the rmd.
-demographics <- read.csv("results/tables/denovodemographicsTable.csv")
-OverallPValue <- read.csv("results/tables/denovoOverallPTable.csv")
-ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
-BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
-ZScoreCombinedData <- read.csv("results/tables/denovoCombinedData.csv")
-ZScoreTurnbaughData <- read.csv("results/tables/denovoTurnbaughData.csv")
-AUCRFData <- read.csv("results/tables/denovoAUCRFDataTable.csv")
-PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
-ModelledPower1PerDiff <- read.csv("results/tables/denovoModelled1PerDiff.csv")
-ModelledPower5PerDiff <- read.csv("results/tables/denovoModelled5PerDiff.csv")
-ModelledPower10PerDiff <- read.csv("results/tables/denovoModelled10PerDiff.csv")
-ModelledPower15PerDiff <- read.csv("results/tables/denovoModelled15PerDiff.csv")
-ActualDiff <- read.csv("results/tables/denovoActualDiff.csv")
-PowerModelHRR <- read.csv("results/tables/denovoPowerModelHRR.csv")
-PowerModelBFRR <- read.csv("results/tables/denovoPowerModelBFRR.csv")
-SimulatedN_H <- read.csv("results/tables/denovoSimulatedN_H.csv")
-SimulatedN_BF <- read.csv("results/tables/denovoSimulatedN_BF.csv")
-RRSimNNeededH <- read.csv("results/tables/denovoRRSimNNeededH.csv")
-RRSimNNeededBF <- read.csv("results/tables/denovoRRSimNNeededBF.csv")
+#demographics <- read.csv("results/tables/denovodemographicsTable.csv")
+#OverallPValue <- read.csv("results/tables/denovoOverallPTable.csv")
+#ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
+#BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
+#ZScoreCombinedData <- read.csv("results/tables/denovoCombinedData.csv")
+#ZScoreTurnbaughData <- read.csv("results/tables/denovoTurnbaughData.csv")
+#AUCRFData <- read.csv("results/tables/denovoAUCRFDataTable.csv")
+#PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
+#ModelledPower1PerDiff <- read.csv("results/tables/denovoModelled1PerDiff.csv")
+#ModelledPower5PerDiff <- read.csv("results/tables/denovoModelled5PerDiff.csv")
+#ModelledPower10PerDiff <- read.csv("results/tables/denovoModelled10PerDiff.csv")
+#ModelledPower15PerDiff <- read.csv("results/tables/denovoModelled15PerDiff.csv")
+#ActualDiff <- read.csv("results/tables/denovoActualDiff.csv")
+#PowerModelHRR <- read.csv("results/tables/denovoPowerModelHRR.csv")
+#PowerModelBFRR <- read.csv("results/tables/denovoPowerModelBFRR.csv")
+#SimulatedN_H <- read.csv("results/tables/denovoSimulatedN_H.csv")
+#SimulatedN_BF <- read.csv("results/tables/denovoSimulatedN_BF.csv")
+#RRSimNNeededH <- read.csv("results/tables/denovoRRSimNNeededH.csv")
+#RRSimNNeededBF <- read.csv("results/tables/denovoRRSimNNeededBF.csv")
 
@@
 
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
+```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
 # Load specific R files to source data from for graphs
@@
 
-```{r Data_Load_Fig4, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Data_Load_Fig4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
 
@@
 First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value.
-We also used the random forest machine learning algorithm to identify community features that would allow us to to predict whether a person was obese.
+We also used the random forest machine learning algorithm to identify community features that would allow us to to predict whether a person was obese based on either genus or OTU level.
 Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F.
@@
 **Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals.
-Only one diversity measure (OTU richness (N=1)) was significantly different between non-obese and obese individuals (P<0.05).
-With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`) and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05).
+A total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` diversity measurements  were found to be significantly different between non-obese and obese indivdiuals (P < 0.05) (Shannon Diversity (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannon"), ]))`), observed richness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "sobs"), ]))`), and evenness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannoneven"), ]))`)).
+With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`) and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05).
 In 7/8 of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies.
@@
 **Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
-```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
 
@@
 
-```{r Table_2, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Table_2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
 
@@
 
-```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
 combinedData <- read.csv("results/tables/ROCcombinedData.csv", stringsAsFactors = FALSE)
@@
 
-```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 # Section A
@@
 
-```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_6, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
add Pat data pt2 more readable code

@@
 A total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` diversity measurements  were found to be significantly different between non-obese and obese indivdiuals (P < 0.05) (Shannon Diversity (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannon"), ]))`), observed richness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "sobs"), ]))`), and evenness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannoneven"), ]))`)).
-With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`) and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05).
-In 7/8 of the studies an OTU related to the Ruminococcaceae family was important in classification; however, the direction of the association varied between studies.
-When we pooled the results across studies, there was no significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(10^test3R$LogFC[1],2), nsmall=2)`, CI = `r format(round(10^ShannonRes$ci.lb, 2), nsmall=2)`-`r format(round(10^ShannonRes$ci.ub, 2), nsmall=2)`, P-value = `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`) and B/F ratio (RR = `r format(round(10^test3BF$LogFC[1],2), nsmall=2)`, CI = `r format(round(10^BFRes$ci.lb, 2), nsmall=2)`-`r format(round(10^BFRes$ci.ub, 2), nsmall=2)`, P-value = `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`).
+With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals for both the genus (mean AUC: `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`) and OTU (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)`) level, and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05).
+When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`).
 When we normalized the Shannon diversity data and pooled studies there was also no significant difference between non-obese and obese individuals (difference in mean z-score = `r format(round(combZHTtest$estimate[1]-combZHTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZHTtest$p.value, 2), nsmall=2)`).

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
complete Pat data abstract

@@
                                     header = T)
+alpha.data.ALL <- read.table("../data/process/alpha.data", header = T)
 
@@
 When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`).
-When we normalized the Shannon diversity data and pooled studies there was also no significant difference between non-obese and obese individuals (difference in mean z-score = `r format(round(combZHTtest$estimate[1]-combZHTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZHTtest$p.value, 2), nsmall=2)`).
-There was also no significant difference in the normalized B/F ratio between non-obese and obese individuals (difference in mean z-score = `r format(round(combZLogBFTtest$estimate[1]-combZLogBFTtest$estimate[2], 2), nsmall=2)`; P = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`).
+When we used a linear mixed-effect model correcting for study there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 4), nsmall=4)`) between non-obese and obese individuals.
 Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
add Pat data to results pt1

@@
 Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.
-We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(AUCRFData$OOBAUCAll), 2), nsmall=2)` +/- `r format(round(sd(AUCRFData$OOBAUCAll), 2), nsmall=2)`.
-We found that the range for the total number of variables used for the classification ranged from 6 (HMP study) to 42 (Zupancic).
-The total number of trees and node size was standardized to 1000 and 20 respectively for each data set.
+We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.
+We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al.
+study) to `r max(rand_forest.genus.summary$k_opt` (Goodrich, et al.
+Study).
+For the OTU level the number of variables ranged from `r min(rand_forest.otu.summary$k_opt)` (Wu, et al.
+study) to `r max(rand_forest.otu.summary$k_opt)` (Goodrich, et al.
+study).
+The total number of trees and node size was standardized to 500 and 20 respectively for each data set.
 Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05, Figure 2).

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
add Pat data RR results pt2

@@
 For the Shannon diversity RR, all studies but Wu, Goodrich, and Turnbaugh had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 3A].
-The was  pooled analysis showed an increased RR of `r format(round(10^test3R$LogFC[1],2), nsmall=2)` (CI `r format(round(10^ShannonRes$ci.lb, 2), nsmall=2)`, `r format(round(10^ShannonRes$ci.ub, 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was not significant (P-value = `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`).
-Similarily, when we investigated the B/F ratio RR  most studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`)[Figure 3B].
-The pooled RR was `r format(round(10^test3BF$LogFC[1],2), nsmall=2)` (CI `r format(round(10^BFRes$ci.lb, 2), nsmall=2)`, `r format(round(10^BFRes$ci.ub, 2), nsmall=2)`).
+The pooled analysis showed an increased RR of `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['shannon', 'ci_lb'],2), nsmall=2)` - `r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was significant (P-value = `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`).
+In contrast, when we investigated the B/F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`)[Figure 3B].
+The pooled RR was `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)` - `r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`).
 The differences in RR that was observed also do not seem to be related to the variable region chosen [Figure 3].

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
add Pat data LME results pt3

@@
 
-***Z-Score Normalization Pooled Analysis***
+***Linear Mixed-Effect models for Pooled Analysis***
 
-Using the third approach with Z-score normalization we found that there was no significant difference in the Shannon diversity between the obese and non-obese group (P-value = `r format(round(combZHTtest$p.value, 2), nsmall=2)`) [Figure 5A].
-Additionally,  we found that the there was also no significant difference in B/F ratio, using the Z-score normalization, between obese and non-obese individuals (P-value = `r format(round(combZLogBFTtest$p.value, 2), nsmall=2)`) [Figure 5B].
+Using the third approach which utilizes a linear mixed-effect model to control for the effect of study on the Shannon diversity  we found that there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 5A].
+Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`.
+However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2)), nsmall=2)`, Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2)), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2)), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2)), nsmall=2)`using this approach, between obese and non-obese individuals [Figure 5B].
 

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
add Pat data power results pt4

@@
 RR_pow.pred <- read.table("../data/process/rr_power.predicted", header = T)
+
+shannond <- (mean(alpha.data.ALL[which(alpha.data.ALL$obese == FALSE), ]$shannon)-mean(alpha.data.ALL[which(alpha.data.ALL$obese == TRUE), ]$shannon))/sd(alpha.data.ALL$shannon)
+
+bf_ratiod <- (mean(alpha.data.ALL[which(alpha.data.ALL$obese == FALSE), ]$bf_ratio)-mean(alpha.data.ALL[which(alpha.data.ALL$obese == TRUE), ]$bf_ratio))/sd(alpha.data.ALL$bf_ratio)
 
@@
 For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 5A] compared to the B/F ratio [Figure 6A] differences.
-When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity [Figure 5A].
-However, this was not the case for the B/F ratio where even a 15% difference between non-obese and obese groups only resulted in an average power between studies of `r format(round(mean(ModelledPower15PerDiff$BFratio)*100, 2), nsmall=2)`% [Figure 6A].
-Looking at the sample sizes showed that on average to detect a 5% difference at 80% power between non-obese and obese individuals a total n of `r format(round(mean(RRSimNNeededH$RRFivePercentNH), 0), nsmall=0)` would be needed for each group [Figure 5B].
-In contrast in order to detect a 5% difference with 80% power in the B/F ratio a sample size of `r format(round(mean(RRSimNNeededBF$RRFivePercentNBF), 0), nsmall=0)` per group [Figure 6B].
-A breakdown of the power simulations and sample size estimates of the other variables and RR can be found in the online supplement.
+When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity (predicted percent power = `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`) [Figure 5A].
+This predicted power was also similar for evenness (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`).
+However, this was not the case for the B/F ratio (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)` +/- `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`%) [Figure 6A] or the other measures that were tested (observed richness, Bacteroidetes, Firmicutes) where even a 15% difference between non-obese and obese groups only resulted in a relatively poor average power between studies.
+Looking at the sample sizes showed that on average to detect a 5% difference in Shannon diversity at 80% power between non-obese and obese individuals a total n of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) would be needed for each group [Figure 5B].
+In contrast in order to detect a 5% difference with 80% power in the B/F ratio a sample size of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) per group [Figure 6B].
+A complete breakdown of the power simulations and sample size estimates for the other variables investigated and RR can be found in the online supplement.
 
-We also investigated if the combined total n for the third strategy was sufficiently powered.
+We also investigated if the combined total n for the linear mixed effect model was sufficiently powered.
 Using a small (0.2), medium (0.5), and large (0.8) cohen value we estimated the power of the combined data sets as well as the average cohen value from across studies for Shannon Diversity (0.06) and BF ratio (0.01).
-For the small effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.2, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
-For the medium effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.5, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
+For the small effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=0.2, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
+For the medium effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=0.5, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
 For the large effect for combined data power calculation a value of
-`r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.8, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
-The average actual cohens d value for Shannon Diversity for all 8 studies of 0.06 gave a power of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.06, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
-Finally, the average actual cohens d value for B/F ratio for all 8 studies of 0.01 yieled a power of `r format(round(pwr.t2n.test(n1=length(which(group.data.test$obese == "No")), n2=length(which(group.data.test$obese == "Yes")), d=0.01, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
+`r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=0.8, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
+The average actual cohens d value for Shannon Diversity for all `r length(rownames(beta_tests.summary))` studies of `r format(round(shannond, 2), nsmall=2)` gave a power of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=shannond, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
+Finally, the average actual cohens d value for B/F ratio for all `r length(rownames(beta_tests.summary))` studies of `r format(round(bf_ratiod, 2), nsmall=2)` yieled a power of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=bf_ratiod, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
 

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
add Pat data discussion ES pt1

@@
 
-# Hold on to these for now as I update the rest of the rmd.
-#demographics <- read.csv("results/tables/denovodemographicsTable.csv")
-#OverallPValue <- read.csv("results/tables/denovoOverallPTable.csv")
-#ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
-#BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
-#ZScoreCombinedData <- read.csv("results/tables/denovoCombinedData.csv")
-#ZScoreTurnbaughData <- read.csv("results/tables/denovoTurnbaughData.csv")
-#AUCRFData <- read.csv("results/tables/denovoAUCRFDataTable.csv")
-#PowerTable <- read.csv("results/tables/denovoPowerTable.csv")
-#ModelledPower1PerDiff <- read.csv("results/tables/denovoModelled1PerDiff.csv")
-#ModelledPower5PerDiff <- read.csv("results/tables/denovoModelled5PerDiff.csv")
-#ModelledPower10PerDiff <- read.csv("results/tables/denovoModelled10PerDiff.csv")
-#ModelledPower15PerDiff <- read.csv("results/tables/denovoModelled15PerDiff.csv")
-#ActualDiff <- read.csv("results/tables/denovoActualDiff.csv")
-#PowerModelHRR <- read.csv("results/tables/denovoPowerModelHRR.csv")
-#PowerModelBFRR <- read.csv("results/tables/denovoPowerModelBFRR.csv")
-#SimulatedN_H <- read.csv("results/tables/denovoSimulatedN_H.csv")
-#SimulatedN_BF <- read.csv("results/tables/denovoSimulatedN_BF.csv")
-#RRSimNNeededH <- read.csv("results/tables/denovoRRSimNNeededH.csv")
-#RRSimNNeededBF <- read.csv("results/tables/denovoRRSimNNeededBF.csv")
+#Using a quick for loop to generate the necessary effect size information
+variableTested <- seq(1:6)
+variables <- colnames(alpha_power.obs)[1:6]
+alpha_effectSize <- matrix(nrow=length(rownames(beta_tests.summary)), ncol=length(variableTested))
+studies <- rownames(beta_tests.summary)
+
+for(i in variableTested){
+  tempdata <- c()
+  for(j in studies){
+  tempdata <- rbind(tempdata, abs((mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == FALSE), variables[i]]) - mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == TRUE), variables[i]]))/mean(alpha.data.ALL[, variables[i]]) * 100))
+  }
+  alpha_effectSize[, i] <- tempdata
+}
+
+colnames(alpha_effectSize) <- variables
+rownames(alpha_effectSize) <- studies
 
@@
 source("code/MakeFiguresCode.R")
-
-```
-
-```{r Data_Load_Fig4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
-
-#data1 order Escobar, Baxter, Ross, HMP, Arumugam, Wu, Goodrich, Zupancic
-
-ZH <- as.numeric(ZScoreCombinedData$ZH)
-bmi <- as.numeric(ZScoreCombinedData$BMI)
-BMI.class <- as.character(ZScoreCombinedData$BMICat)
-BMI.class[BMI.class == "Normal"] <- "Lean"
-obese <- as.character(ZScoreCombinedData$Obese)
-ZlogBF <- as.numeric(ZScoreCombinedData$ZLogBF)
-
-data1 <- as.data.frame(cbind(ZH, ZlogBF, bmi))
-
-#Create group analysis table to harness full data set
-#For Z-score normalization
-t.obese <- as.character(ZScoreTurnbaughData$Obese)
-t.bmi.class <- as.character(ZScoreTurnbaughData$BMICat)
-t.ZH <- ZScoreTurnbaughData$ZH
-t.ZlogBF <- ZScoreTurnbaughData$ZLogBF
-
-#Combine everything together
-comb1.obese <- c(obese, t.obese)
-comb1.bmi.class <- c(BMI.class, t.bmi.class)
-comb1.ZH <- as.data.frame(c(ZH, t.ZH))
-group.data.test1 <- cbind(comb1.ZH, comb1.obese, comb1.bmi.class)
-colnames(group.data.test1) <- c("ZH", "obese", "BMI.class")
-
-#Let's change the order so it makes sense
-group.data.test1$BMI.class <- factor(group.data.test1$BMI.class, levels = c("Lean", "Overweight", "Obese"))
-
-group.data.test <- group.data.test1
-
-#Bacteroidetes/Firmicutes Datasets
-comb.ZlogBF <- as.data.frame(c(ZlogBF, t.ZlogBF))
-group.data.test <- cbind(comb.ZlogBF, group.data.test)
-colnames(group.data.test)[1] <- "ZLogBF"
-
-## Perform the statistical tests
-combZHTtest <- t.test(ZH ~ obese, data=group.data.test) #P-value = 0.006492
-combZHANOVA <- anova(lm(ZH ~ BMI.class, data=group.data.test)) #P-value = 0.01588
-a2 <- aov(ZH ~ BMI.class, data=group.data.test)
-combZHTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-combZHTukeyFinal <- combZHTukey$BMI.class
-
-combZLogBFTtest <- t.test(ZLogBF ~ obese, data=group.data.test) #P-value = 0.6337
-combZLogBFANOVA <- anova(lm(ZLogBF ~ BMI.class, data=group.data.test)) #P-value = 0.7125
-a2 <- aov(ZLogBF ~ BMI.class, data=group.data.test)
-combZLogBFTukey <- TukeyHSD(x=a2, 'BMI.class', conf.level=0.95)
-combZLogBFTukeyFinal <- combZLogBFTukey$BMI.class
 
@@
 In order to also assess the amount of our results that could be due to insufficient power we performed power and sample size analysis.
-Overall we found that the most studies for Shannon Diversity would be powered to find a 15% difference, however, the vast majority of reported differences between studies ranged from `r format(round(min(ActualDiff$H), 2), nsmall=2)`-`r format(round(max(ActualDiff$H), 2), nsmall=2)` percent.
+Overall we found that most studies for Shannon Diversity would be powered to find a 15% difference, however, the vast majority of reported effect sizes between studies ranged from `r format(round(min(alpha_effectSize[, 'shannon']), 2), nsmall=2)`-`r format(round(max(alpha_effectSize[, 'shannon']), 2), nsmall=2)` percent.
 Although the effect sizes for the BF ratio were larger than that for Shannon Diversity [Figure 5A & 6A] the estimated power remained low.

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
fixing minor typos of Pat data changes so far

@@
 }
-
-```
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
-# Load specific R files to source data from for graphs
-
-source("code/MakeRRFiguresCode.R")
-source("code/MakeFiguresCode.R")
 
@@
 Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.
-We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.
+We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.
 We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al.
-study) to `r max(rand_forest.genus.summary$k_opt` (Goodrich, et al.
+study) to `r max(rand_forest.genus.summary$k_opt)` (Goodrich, et al.
 Study).
@@
 Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`.
-However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2)), nsmall=2)`, Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2)), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2)), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2)), nsmall=2)`using this approach, between obese and non-obese individuals [Figure 5B].
+However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`, Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`using this approach, between obese and non-obese individuals [Figure 5B].
 
@@
 
-Study <- as.character(demographics$Study)
+Study <- as.character(rownames(beta_tests.summary))
 

### 2016-05-23 submission/Sze_Obesity_PeerJ_2016.Rmd
got a working Rmd to render

@@
     keep_md: yes
-bibliography: submission/references.bib
+bibliography: reference.bib
 ---
@@
 
-![Figure 1](results/figures/PrismaFlowGraph.png)\
+![Figure 1](../results/figures/Figure_1.png)\
 
@@
 Red represents studies that used the 16S rRNA V4 region, yellow represents studies that used a V1-V3 approach, blue represents studies that used either a V1-V2 or just a V2 approach, and orange represents studies that used the V3-V5 region.
-**A)** The overall pooled relative risk for Shannon Diversity was `r format(round(10^test3R$LogFC[1],2), nsmall=2)` with a P-value of `r format(round(RRSDPvalue$QMp, 2), nsmall=2)`.
-**B)** The overall pooled relative risk for the B/F ratio was  `r format(round(10^test3BF$LogFC[1],2), nsmall=2)` with a P-value of `r format(round(RRBFPvalue$QMp, 2), nsmall=2)`.
+**A)** The overall pooled relative risk for Shannon Diversity was `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` with a P-value of `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`.
+**B)** The overall pooled relative risk for the B/F ratio was  `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` with a P-value of `r format(round(RR_composite['bf_ratio', 'p_value'], 2), nsmall=2)`.
 
@@
 **Figure 4: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** Z-score Normalized Shannon Diversity and Obesity.
-There was a significant difference between non-obese (n = `r length(which(group.data.test$obese == "No"))`) and obese groups (n = `r length(which(group.data.test$obese == "Yes"))`) (P-value = `r signif(combZHTtest$p.value, 3)`).
+There was a significant difference between non-obese (n = `r length(which(alpha.data.ALL$obese == FALSE))`) and obese groups (n = `r length(which(alpha.data.ALL$obese == TRUE))`) (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`).
 **B)** Z-score Normalized B/F Ratio and Obesity.
-There was no significant difference between the non-obese (n = `r length(which(group.data.test$obese == "No"))`) and obese (n = `r length(which(group.data.test$obese == "Yes"))`) group based on B/F ratio (P-value = `r signif(combZLogBFTtest$p.value, 3)`).
+There was no significant difference between the non-obese (n = `r length(which(alpha.data.ALL$obese == FALSE))`) and obese (n = `r length(which(alpha.data.ALL$obese == TRUE))`) group based on B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`).
 

### 2016-05-24 code/run_alpha_diversity.R
Small edits

@@ -1,6 +1,7 @@
-library("lme4")
-
 source('code/utilities.R')
+get_dependencies("lme4")
+
+
 
 composite_analysis <- function(alpha, dataset, is_obese, pow){
 
@@ -27,8 +28,8 @@ single_analysis <- function(alpha, pow, is_obese){
 	n_non <- length(non_obese)
 	n_obese <- length(obese)
 
-	ci_obese <- unname(quantile(obese, c(0.025, 0.975)))
-	ci_non <- unname(quantile(non_obese, c(0.025, 0.975)))
+	ci_obese <- unname(quantile(obese, c(0.025, 0.975), na.rm=T))
+	ci_non <- unname(quantile(non_obese, c(0.025, 0.975), na.rm=T))
 
 
 	return(

### 2016-05-24 code/run_beta_diversity.R

@@ -38,6 +38,7 @@ run <- function(datasets){
 	summary_data <- NULL
 
 	for(d in datasets){
+		print(d)
 		set.seed(1976)
 		beta_file <- paste0('data/', d, '/', d, '.braycurtis.0.03.lt.ave.dist')
 		beta <- read_lt_matrix(beta_file)

### 2016-05-24 code/finish_clustersplit.R
Remove samples that have NA for obesity value

@@ -1,0 +1,41 @@
+level_regexp <- "^0\\.03"
+
+extra_names_file <- "data/zeevi/zeevi.renamed.unique.good.filter.unique.precluster.denovo.uchime.pick.pick.count_table.extra.temp"
+extras <- read.table(file=extra_names_file, header=T, stringsAsFactors=F)[,1]
+
+list_files <- list.files('data/zeevi', pattern='*.list', full.names=T)
+file_number <- as.numeric(gsub(".*fasta.(\\d*)\\..*", "\\1", list_files))
+file_order <- order(file_number)
+list_files <- list_files[file_order]
+
+combined_data <- rep("", length(list_files))
+
+for(i in 0:(length(list_files)-1)){
+
+	list_line <- ""
+	list_file <- scan(list_files[i+1], what=character(), sep="\n", quiet=T)
+	hits <- grepl(level_regexp, list_file)
+
+	if(sum(hits) == 1){
+		list_line <- list_file[hits]
+	} else {
+		list_file[1] <- gsub("^unique", "0.00", list_file[1])
+		cutoffs <- as.numeric(gsub("^(0\\.\\d\\d)\\t.*", "\\1", list_file))
+
+		best_cutoff <- which.max(cutoffs[cutoffs <= 0.03])
+		list_line <- list_file[best_cutoff]
+	}
+
+	combined_data[i+1] <- gsub("^0\\.\\d\\d\\t\\d*\\t", "", list_line)
+}
+
+clustered <- paste(combined_data, collapse='\t')
+extra_merged <- paste(extras, collapse='\t')
+
+combined_data <- paste(clustered, extra_merged, sep='\t')
+
+n_otus <- sum(nchar(combined_data) - nchar(gsub("\t", "", combined_data)) + 1)
+
+new_list_line <- paste("0.03", n_otus, combined_data, sep='\t')
+
+write(new_list_line, file="data/zeevi/zeevi.renamed.unique.good.filter.unique.precluster.pick.pick.an.unique_list.list")

### 2016-05-24 code/plot_roc_data.R

@@ -1,0 +1,58 @@
+capwords <- function(s, strict = FALSE) {
+		cap <- function(s) paste(toupper(substring(s, 1, 1)),
+									{s <- substring(s, 2); if(strict) tolower(s) else s},
+														 sep = "", collapse = " " )
+		sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
+}
+
+plot_roc_data <- function(data_type){
+
+	roc_curve_file <- paste0("data/process/random_forest.", data_type, ".roc_data")
+	roc_curve <- read.table(file=roc_curve_file, header=T, stringsAsFactors=F)
+
+	rf_summary_file <- paste0("data/process/random_forest.", data_type, ".summary")
+	rf_summary <- read.table(file=rf_summary_file, header=T, stringsAsFactors=F)
+
+	N <- ifelse(data_type == "otu", "OTUs", "Genera")
+
+	rf_summary <- rf_summary[order(rf_summary$auc_cv, decreasing=T),]
+	datasets <- rf_summary$dataset
+
+	study_name <- capwords(datasets)
+	study_name[study_name == "Hmp"] <- "HMP"
+
+	basic_colors <- c("red", "lightseagreen", "dodgerblue", "black")
+	colors <- rep(basic_colors, length.out=length(datasets))
+	names(colors) <- datasets
+
+	basic_lty <- 1:3
+	lty <- rep(basic_lty, each=length(basic_colors), length.out=length(datasets))
+	names(lty) <- datasets
+
+	plot_roc <- function(study){
+		study_data <- roc_curve[roc_curve$dataset == study,]
+		points(study_data$specificity~study_data$sensitivity, type="s", col=colors[study], lty=lty[study], lwd=2)
+	}
+
+
+	pdf(file=paste0("results/figures/roc_curve.", data_type, ".pdf"))
+	par(mar=c(4,4,0.5,0.5))
+	plot(NA, xlim=c(1,0), ylim=c(0,1), axes=F, ylab="Specificity", xlab="Sensitivity")
+
+	abline(a=1, b=-1, col="gray")
+
+	sapply(datasets, plot_roc)
+
+	axis(1)
+	axis(2, las=2)
+	box()
+
+	legend_name <- paste(
+		study_name,
+		" (AUC-CV=",
+		format(rf_summary$auc_cv, digits=2),
[…]

### 2016-05-24 code/run_alpha_diversity.R

@@ -63,6 +63,13 @@ run <- function(datasets){
 
 		stopifnot(alpha$group == metadata$sample)
 
+		na_obesity <- is.na(metadata$obese)
+		metadata <- metadata[!na_obesity,]
+		alpha <- alpha[!na_obesity,]
+
+		stopifnot(rownames(alpha) == metadata$sample)
+
+
 		bf_relabund <- get_bacteroides_firmicutes(d)
 
 		#transformations were worked out with a qqplot to check for normality

### 2016-05-24 code/run_aucrf_genus.R

@@ -79,6 +79,12 @@ run <- function(datasets){
 
 		stopifnot(rownames(shared) == metadata$sample)
 
+		na_obesity <- is.na(metadata$obese)
+		metadata <- metadata[!na_obesity,]
+		shared <- shared[!na_obesity,]
+
+		stopifnot(rownames(shared) == metadata$sample)
+
 		keep_otus <- apply(shared > 0, 2, sum) > (0.1 * nrow(shared))
 
 		rel_abund_keep <- shared[,keep_otus] / n_seqs

### 2016-05-24 code/run_aucrf_otus.R

@@ -25,6 +25,13 @@ run <- function(datasets){
 
 		stopifnot(rownames(shared) == metadata$sample)
 
+		na_obesity <- is.na(metadata$obese)
+		metadata <- metadata[!na_obesity,]
+		shared <- shared[!na_obesity,]
+
+		stopifnot(rownames(shared) == metadata$sample)
+
+
 		keep_otus <- apply(shared > 0, 2, sum) > (0.1 * nrow(shared))
 
 		rel_abund_keep <- shared[,keep_otus] / n_seqs

### 2016-05-24 code/run_beta_diversity.R

@@ -43,8 +43,14 @@ run <- function(datasets){
 		beta <- read_lt_matrix(beta_file)
 
 		metadata_file <- paste0('data/', d, '/', d, '.metadata')
-		metadata <- read.table(file=metadata_file, header=T)
+		metadata <- read.table(file=metadata_file, header=T, stringsAsFactors=F)
 		metadata <- metadata[metadata$sample %in% rownames(beta),]
+
+		stopifnot(rownames(beta) == metadata$sample)
+
+		na_obesity <- is.na(metadata$obese)
+		metadata <- metadata[!na_obesity,]
+		beta <- beta[!na_obesity,!na_obesity]
 
 		stopifnot(rownames(beta) == metadata$sample)
 

### 2016-05-24 code/run_relative_risk.R

@@ -52,6 +52,12 @@ run <- function(datasets){
 
 		stopifnot(alpha$group == metadata$sample)
 
+		na_obesity <- is.na(metadata$obese)
+		metadata <- metadata[!na_obesity,]
+		alpha <- alpha[!na_obesity,]
+
+		stopifnot(rownames(alpha) == metadata$sample)
+
 		bf_relabund <- get_bacteroides_firmicutes(d)
 
 		shannon <- analyze_indiv_study(alpha$shannon, metadata$obese)

### 2016-05-24 code/run_alpha_diversity.R
Fix access to sample names in alpha file

@@ -68,7 +68,7 @@ run <- function(datasets){
 		metadata <- metadata[!na_obesity,]
 		alpha <- alpha[!na_obesity,]
 
-		stopifnot(rownames(alpha) == metadata$sample)
+		stopifnot(alpha$group == metadata$sample)
 
 
 		bf_relabund <- get_bacteroides_firmicutes(d)

### 2016-05-24 code/run_relative_risk.R

@@ -56,7 +56,7 @@ run <- function(datasets){
 		metadata <- metadata[!na_obesity,]
 		alpha <- alpha[!na_obesity,]
 
-		stopifnot(rownames(alpha) == metadata$sample)
+		stopifnot(alpha$group == metadata$sample)
 
 		bf_relabund <- get_bacteroides_firmicutes(d)
 

### 2016-05-24 code/run_alpha_diversity.R
Fix row handling with bf calculations

@@ -42,7 +42,7 @@ single_analysis <- function(alpha, pow, is_obese){
 	)
 }
 
-datasets <- c('baxter', 'goodrich', 'escobar', 'hmp', 'ross', 'turnbaugh', 'wu', 'zupancic')
+datasets <- c('baxter', 'goodrich', 'escobar', 'hmp', 'ross', 'turnbaugh', 'wu', 'zupancic', 'zeevi')
 
 run <- function(datasets){
 
@@ -67,11 +67,13 @@ run <- function(datasets){
 		na_obesity <- is.na(metadata$obese)
 		metadata <- metadata[!na_obesity,]
 		alpha <- alpha[!na_obesity,]
-
 		stopifnot(alpha$group == metadata$sample)
 
-
 		bf_relabund <- get_bacteroides_firmicutes(d)
+		bf_relabund <- bf_relabund[alpha$group,]
+		stopifnot(rownames(bf_relabund) == metadata$sample)
+
+
 
 		#transformations were worked out with a qqplot to check for normality
 		shannon <- single_analysis(alpha$shannon, pow=2, metadata$obese)

### 2016-05-24 code/run_relative_risk.R

@@ -55,10 +55,11 @@ run <- function(datasets){
 		na_obesity <- is.na(metadata$obese)
 		metadata <- metadata[!na_obesity,]
 		alpha <- alpha[!na_obesity,]
-
 		stopifnot(alpha$group == metadata$sample)
 
 		bf_relabund <- get_bacteroides_firmicutes(d)
+		bf_relabund <- bf_relabund[alpha$group,]
+		stopifnot(rownames(bf_relabund) == metadata$sample)
 
 		shannon <- analyze_indiv_study(alpha$shannon, metadata$obese)
 		sobs <- analyze_indiv_study(alpha$sobs, metadata$obese)

### 2016-05-24 submission/Sze_Obesity_PeerJ_2016.Rmd
fixing spacing in text

@@
 When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`).
-When we used a linear mixed-effect model correcting for study there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 4), nsmall=4)`) between non-obese and obese individuals.
+When we used a linear mixed-effect model correcting for study there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) between non-obese and obese individuals.
 Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.
@@
 Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.
-We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.
+We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)` for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.
 We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al.
@@
 The pooled analysis showed an increased RR of `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['shannon', 'ci_lb'],2), nsmall=2)` - `r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was significant (P-value = `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`).
-In contrast, when we investigated the B/F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`)[Figure 3B].
+In contrast, when we investigated the B/F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`) [Figure 3B].
 The pooled RR was `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)` - `r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`).
@@
 Using the third approach which utilizes a linear mixed-effect model to control for the effect of study on the Shannon diversity  we found that there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 5A].
-Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`.
-However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`, Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`using this approach, between obese and non-obese individuals [Figure 5B].
+Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`).
+However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) using this approach, between obese and non-obese individuals [Figure 5B].
 

### 2016-05-24 code/run_alpha_EffectSize.R
add R script for alpha effect size

@@ -1,0 +1,34 @@
+# Using a quick for loop to generate the necessary effect size information
+# This would be to generate actual effect sizes observed.
+
+# Read in required data
+
+alpha_power.obs <- read.table("../data/process/alpha_power.observed", 
+                              header = T, row.names = 1)
+beta_tests.summary <- read.table("../data/process/beta_tests.summary", 
+                                 header=T, row.names = 1)
+alpha.data.ALL <- read.table("../data/process/alpha.data", header = T)
+
+# Create variables needed to run the for loop
+variableTested <- seq(1:6)
+variables <- colnames(alpha_power.obs)[1:6]
+alpha_effectSize <- matrix(nrow=length(rownames(beta_tests.summary)), ncol=length(variableTested))
+studies <- rownames(beta_tests.summary)
+
+# For loops to create an effect size table
+for(i in variableTested){
+  tempdata <- c()
+  for(j in studies){
+    tempdata <- rbind(tempdata, abs((mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == FALSE), variables[i]]) - mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == TRUE), variables[i]]))/mean(alpha.data.ALL[, variables[i]]) * 100))
+  }
+  alpha_effectSize[, i] <- tempdata
+}
+
+colnames(alpha_effectSize) <- variables
+rownames(alpha_effectSize) <- studies
+
+
+write.table(alpha_effectSize, file="../data/process/alpha_EffectSize.summary", 
+            quote = F, sep='\t', row.names=T)
+
+

### 2016-05-24 submission/Sze_Obesity_PeerJ_2016.Rmd
added Pat data table 1 demographics

@@
 alpha.data.ALL <- read.table("../data/process/alpha.data", header = T)
+alpha_effectSize <- read.table("../data/process/alpha_EffectSize.summary")
 
@@
 bf_ratiod <- (mean(alpha.data.ALL[which(alpha.data.ALL$obese == FALSE), ]$bf_ratio)-mean(alpha.data.ALL[which(alpha.data.ALL$obese == TRUE), ]$bf_ratio))/sd(alpha.data.ALL$bf_ratio)
-
-#Using a quick for loop to generate the necessary effect size information
-variableTested <- seq(1:6)
-variables <- colnames(alpha_power.obs)[1:6]
-alpha_effectSize <- matrix(nrow=length(rownames(beta_tests.summary)), ncol=length(variableTested))
-studies <- rownames(beta_tests.summary)
-
-for(i in variableTested){
-  tempdata <- c()
-  for(j in studies){
-  tempdata <- rbind(tempdata, abs((mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == FALSE), variables[i]]) - mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == TRUE), variables[i]]))/mean(alpha.data.ALL[, variables[i]]) * 100))
-  }
-  alpha_effectSize[, i] <- tempdata
-}
-
-colnames(alpha_effectSize) <- variables
-rownames(alpha_effectSize) <- studies
 
@@
 **Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
-```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
+```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
 
-Study <- as.character(rownames(beta_tests.summary))
+table1 <- read.table("../results/tables/table1.demographics", header=T)
 
-Age <- as.character(demographics$meanAge)
-AgeSD <- as.character(format(round(
-  demographics$SDAge, 2), nsmall = 2))
-AgeTable <- paste(Age," (",AgeSD,")", sep="")
-
-Sex <- paste(demographics$females, "|", demographics$males)
-
-Ancestry <- as.character(format(round(
-  demographics$ancestry*100, 2), nsmall = 2))
-
-ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
-ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
-
-ObBreakdown <- paste(ContTotal, " | ", ObeseTotal, sep="")
-
-BMI <- paste(demographics$meanBMI, " (", demographics$SDBMI, ")", sep="")
-
-Min <- as.character(demographics$minBMI)
-
-Max <- as.character(demographics$maxBMI)
-
-table1 <- cbind(Study, AgeTable, Sex, Ancestry, ObBreakdown, BMI, Min, Max)
-table1[table1=="NA" | table1=="NA | NA" | table1=="NA (NA)" |
-         table1=="      NA"] <- "Not Available"
-
-colnames(table1) <- c("Study", "Age (SD)", "Sex (F|M)", "European Ancestry (%)", "Obese (No|Yes)", "BMI (SD)", "Min BMI","Max BMI")
+colnames(table1) <- c("Study", "Age (SD)", "Sex (F|M)", "European Ancestry (%)", 
+                      "Obese (No|Yes)", "BMI (SD)", "Min BMI","Max BMI")
 

### 2016-05-24 submission/Sze_Obesity_PeerJ_2016.Rmd
added Pat data table 2 study info

@@
 
-```{r Table_2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
+```{r Table_2, echo=FALSE, message=FALSE, warning=FALSE}
+# Old order
+#Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "HMP", "Wu", #"Turnbaugh")
 
-Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "HMP", "Wu", "Turnbaugh")
+Study <- as.character(table1$Study)
+  #New Order
+  #baxter, escobar, hmp, ross, schubert, turnbaugh, wu, zupancic, goodrich, zeevi
 
@@
 
-Population <- c("Adults without Cancer from Texas, Ontario, Michigan, Massachusetts", "Hispanic adults with and without type 2 diabetes from Texas
-", "Adult twins and mothers from the United Kingdom", "Healthy adults from Columbia", "Amish Adults with and without metabolic syndrome from Pennsylvania", "Healthy Adults from Texas and Missouri", "Healthy Adults from Pennsylvania", "Adult twins and mothers from Missouri")
+Population <- c("Adults without Cancer from Texas, Ontario, Michigan, Massachusetts", "Healthy adults from Columbia", "Healthy Adults from Pennsylvania", "Hispanic adults with and without type 2 diabetes from Texas
+", "Nondiarrheal indivdiuals without infection history", "Adult twins and mothers from Missouri", "Healthy Adults from Texas and Missouri", "Amish Adults with and without metabolic syndrome from Pennsylvania", "Adult twins and mothers from the United Kingdom")
 
-Published <- c("No", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes","Yes")
+Published <- c("Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes","Yes", "Yes")
 
-Outcomes <- c("N/A (Unpublished)", "Mentions obesity, no specific correlations made though", "Christensenellaceae higher in low BMI individuals.
-No other correlations found.", "Firmicutes less abundant at higher BMI", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "No significant correlations", "Unweighted UniFrac Correlation with BMI", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ")
+Outcomes <- c("Did not investigate control populations for relationship with BMI", "Firmicutes less abundant at higher BMI", "No significant correlations", "Mentions obesity, no specific correlations made though", "Did not investigate control populations for relationship with BMI", "Unweighted UniFrac Correlation with BMI", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "Christensenellaceae higher in low BMI individuals.
+No other correlations found.")
 

### 2016-05-25 code/run_alpha_diversity.R
Fix removal of NAs from bf_ratio data

@@ -70,7 +70,7 @@ run <- function(datasets){
 		stopifnot(alpha$group == metadata$sample)
 
 		bf_relabund <- get_bacteroides_firmicutes(d)
-		bf_relabund <- bf_relabund[alpha$group,]
+		bf_relabund <- bf_relabund[!na_obesity,]
 		stopifnot(rownames(bf_relabund) == metadata$sample)
 
 

### 2016-05-25 code/run_alpha_EffectSize.R
Revert "fixing spacing in text"
This reverts commit 377bf96a6dcbc68889505ce512452c503e29df88.

@@ -1,34 +1,0 @@
-# Using a quick for loop to generate the necessary effect size information
-# This would be to generate actual effect sizes observed.
-
-# Read in required data
-
-alpha_power.obs <- read.table("../data/process/alpha_power.observed", 
-                              header = T, row.names = 1)
-beta_tests.summary <- read.table("../data/process/beta_tests.summary", 
-                                 header=T, row.names = 1)
-alpha.data.ALL <- read.table("../data/process/alpha.data", header = T)
-
-# Create variables needed to run the for loop
-variableTested <- seq(1:6)
-variables <- colnames(alpha_power.obs)[1:6]
-alpha_effectSize <- matrix(nrow=length(rownames(beta_tests.summary)), ncol=length(variableTested))
-studies <- rownames(beta_tests.summary)
-
-# For loops to create an effect size table
-for(i in variableTested){
-  tempdata <- c()
-  for(j in studies){
-    tempdata <- rbind(tempdata, abs((mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == FALSE), variables[i]]) - mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == TRUE), variables[i]]))/mean(alpha.data.ALL[, variables[i]]) * 100))
-  }
-  alpha_effectSize[, i] <- tempdata
-}
-
-colnames(alpha_effectSize) <- variables
-rownames(alpha_effectSize) <- studies
-
-
-write.table(alpha_effectSize, file="../data/process/alpha_EffectSize.summary", 
-            quote = F, sep='\t', row.names=T)
-
-

### 2016-05-25 submission/Sze_Obesity_PeerJ_2016.Rmd

--- 2016-05-24 submission/Sze_Obesity_PeerJ_2016.Rmd
@@
 When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`).
-When we used a linear mixed-effect model correcting for study there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) between non-obese and obese individuals.
+When we used a linear mixed-effect model correcting for study there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 4), nsmall=4)`) between non-obese and obese individuals.
 Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.
@@
 Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.
-We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)` for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.
+We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.
 We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al.
@@
 The pooled analysis showed an increased RR of `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['shannon', 'ci_lb'],2), nsmall=2)` - `r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was significant (P-value = `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`).
-In contrast, when we investigated the B/F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`) [Figure 3B].
+In contrast, when we investigated the B/F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`)[Figure 3B].
 The pooled RR was `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)` - `r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`).
@@
 Using the third approach which utilizes a linear mixed-effect model to control for the effect of study on the Shannon diversity  we found that there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 5A].
-Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`).
-However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) using this approach, between obese and non-obese individuals [Figure 5B].
+Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`.
+However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`, Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`using this approach, between obese and non-obese individuals [Figure 5B].
 

### 2016-05-25 code/old/run_make_table1.R
Revert to b57c36013cf624

@@ -1,136 +1,0 @@
-# Table 1 Creation 
-# Make all the measurements for the table here so it just needs to be read into
-# the Rmd.
-
-
-# Read in necessary tables
-meta.bax <- read.table("../data/baxter/baxter.metadata", header = T)
-meta.esc <- read.table("../data/escobar/escobar.metadata", header = T)
-meta.goo <- read.table("../data/goodrich/goodrich.metadata", header = T)
-meta.hmp <- read.table("../data/hmp/hmp.metadata", header = T)
-meta.ros <- read.table("../data/ross/ross.metadata", header = T)
-meta.sch <- read.table("../data/schubert/schubert.metadata", header = T)
-meta.tur <- read.table("../data/turnbaugh/turnbaugh.metadata", header = T)
-meta.wu <- read.table("../data/wu/wu.metadata", header = T)
-meta.zee <- read.table("../data/zeevi/zeevi.metadata", header = T)
-meta.zup <- read.table("../data/zupancic/zupancic.metadata", header = T)
-beta_tests.summary <- read.table("../data/process/beta_tests.summary", 
-                                 header=T, row.names = 1)
-
-# Create data table entries
-Study <- as.character(rownames(beta_tests.summary))
-
-Age <- round(c(mean(meta.bax$age), mean(meta.esc$age), mean(meta.hmp$age), 
-               mean(meta.ros$age), mean(meta.sch$age), mean(meta.tur$age), 
-               mean(meta.wu$age), mean(as.numeric(as.character(meta.zup$age)), 
-                                       na.rm=TRUE), mean(meta.goo$age)), 2)
-
-AgeSD <- round(c(sd(meta.bax$age), sd(meta.esc$age), sd(meta.hmp$age), 
-                 sd(meta.ros$age), sd(meta.sch$age), sd(meta.tur$age), 
-                 sd(meta.wu$age), sd(as.numeric(as.character(meta.zup$age)), 
-                                     na.rm=TRUE), sd(meta.goo$age)), 2)
-
-AgeTable <- paste(as.character(Age)," (",as.character(AgeSD),")", sep="")
-
-# No female/male information from Turnbaugh study so have to use NA 
-females <- c(
-  length(meta.bax$sex[meta.bax$sex == 'f']), 
-  length(meta.esc$sex[meta.esc$sex == 'f']), 
-  length(meta.hmp$sex[meta.hmp$sex == 'f']), 
[…]

### 2016-05-25 code/run_alpha_diversity.R

@@ -70,7 +70,7 @@ run <- function(datasets){
 		stopifnot(alpha$group == metadata$sample)
 
 		bf_relabund <- get_bacteroides_firmicutes(d)
-		bf_relabund <- bf_relabund[!na_obesity,]
+		bf_relabund <- bf_relabund[alpha$group,]
 		stopifnot(rownames(bf_relabund) == metadata$sample)
 
 

### 2016-05-25 submission/Sze_Obesity_PeerJ_2016.Rmd
abstract text modifications

--- 2016-05-24 submission/Sze_Obesity_PeerJ_2016.Rmd
@@
 alpha.data.ALL <- read.table("../data/process/alpha.data", header = T)
-alpha_effectSize <- read.table("../data/process/alpha_EffectSize.summary")
 
@@
 
+#Using a quick for loop to generate the necessary effect size information
+variableTested <- seq(1:6)
+variables <- colnames(alpha_power.obs)[1:6]
+alpha_effectSize <- matrix(nrow=length(rownames(beta_tests.summary)), ncol=length(variableTested))
+studies <- rownames(beta_tests.summary)
+
+for(i in variableTested){
+  tempdata <- c()
+  for(j in studies){
+  tempdata <- rbind(tempdata, abs((mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == FALSE), variables[i]]) - mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == TRUE), variables[i]]))/mean(alpha.data.ALL[, variables[i]]) * 100))
+  }
+  alpha_effectSize[, i] <- tempdata
+}
+
+colnames(alpha_effectSize) <- variables
+rownames(alpha_effectSize) <- studies
+
 ```
@@
 
-**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 8 studies that fit our criteria.
+**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 10 studies that fit our criteria.
 We applied three analysis pipelines to the data from each study.
 First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value.
-We also used the random forest machine learning algorithm to identify community features that would allow us to to predict whether a person was obese based on either genus or OTU level.
+We also used the random forest machine learning algorithm to identify community features that would allow us to to classify whether a person was obese based on either genus or OTU level.
 Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F.
-Finally, we normalized the community diversity and B/F values within each dataset and pooled the data to test for an association with obesity.
+Finally, we utilized linear mixed-effect models to correct for variation by study and analyzed the pooled community diversity and B/F values within each dataset to test for an association with obesity.
 
@@
 
-**Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome.
-Our meta-analysis provides support for the null hypothesis that there is no difference between non-obese and obese individuals in Shannon diversity and B/F ratio, that the effect sizes are small, and that a very large number of indivdiuals would have to be tested to reach sufficient study power.
+**Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome and its effect on obesity.
+Our meta-analysis provides support for the observation that there is a difference between non-obese and obese individuals in Shannon diversity but that the effect sizes are small, and that a very large number of indivdiuals would have to be tested to reach sufficient study power.
 
@@
 **Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
-```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
 
-table1 <- read.table("../results/tables/table1.demographics", header=T)
+Study <- as.character(rownames(beta_tests.summary))
 
-colnames(table1) <- c("Study", "Age (SD)", "Sex (F|M)", "European Ancestry (%)", 
-                      "Obese (No|Yes)", "BMI (SD)", "Min BMI","Max BMI")
+Age <- as.character(demographics$meanAge)
+AgeSD <- as.character(format(round(
+  demographics$SDAge, 2), nsmall = 2))
+AgeTable <- paste(Age," (",AgeSD,")", sep="")
+
+Sex <- paste(demographics$females, "|", demographics$males)
+
+Ancestry <- as.character(format(round(
+  demographics$ancestry*100, 2), nsmall = 2))
+
+ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
+ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
+
+ObBreakdown <- paste(ContTotal, " | ", ObeseTotal, sep="")
+
+BMI <- paste(demographics$meanBMI, " (", demographics$SDBMI, ")", sep="")
+
+Min <- as.character(demographics$minBMI)
+
+Max <- as.character(demographics$maxBMI)
+
+table1 <- cbind(Study, AgeTable, Sex, Ancestry, ObBreakdown, BMI, Min, Max)
+table1[table1=="NA" | table1=="NA | NA" | table1=="NA (NA)" |
+         table1=="      NA"] <- "Not Available"
+
+colnames(table1) <- c("Study", "Age (SD)", "Sex (F|M)", "European Ancestry (%)", "Obese (No|Yes)", "BMI (SD)", "Min BMI","Max BMI")
 

### 2016-05-25 submission/Sze_Obesity_PeerJ_2016.Rmd
update methods text

@@
 The first such study used genetically obese mice and observed that obese mice had a lower Bacteroidetes and higher Firmicutes than lean mice (lower B/F ratio associated with obesity) [@ley_obesity_2005].
-Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009].
+Translation of this result to humans by the same researchers did not observe this effect, but did found that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009].
 Since the publication of these studies there have been a steady stream of publications on the microbiome with associated BMI data but without clear explicit mention of any significant correlations between the microbiome and obesity [@ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @escobar_gut_2014].
@@
 The initial search yielded 187 studies.
-We identified nine additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
+We identified ten additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
 We then manually curated the 196 studies to select those studies that included BMI and sequence data.
@@
 An additional study was removed from our analysis because no individuals in the study had a BMI over 30.
-Among the final 9 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and one dataset was used (Baxter) because at the time of our analysis it was unpublished and enabled us to mitigate against publication biases associated with the bacterial microbiome and obesity.
-The nine studies are summarized in Tables 1 and 2.
+Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity.
+The ten studies are summarized in Tables 1 and 2.
 
@@
 All sequence data were publicly available and were downloaded from the NCBI Sequence Read Archive, the European Nucleotide Archive, or the investigators' personal website (https://gordonlab.wustl.edu/TurnbaughSE\_10\_09/STM\_2009.html).
-The nine studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012] or Illumina sequencing [@goodrich_human_2014; Baxter].
+In total seven studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012; @schubert_microbiome_2014] and three studies used Illumina sequencing [@goodrich_human_2014; @zeevi_personalized_2015; @baxter_microbiota-based_2016].
 All of these studies used amplification-based 16S rRNA gene sequencing.
-Among the studies that sequenced the 16S rRNA gene, the researchers targeted the V1-V2  [@wu_linking_2011], V1-V3 [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], V3-V5 [@human_microbiome_project_consortium_structure_2012], and V4 [@goodrich_human_2014] regions.
+Among the studies that sequenced the 16S rRNA gene, the researchers targeted the V1-V2  [@wu_linking_2011], V1-V3 [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], V3-V5 [@human_microbiome_project_consortium_structure_2012; @schubert_microbiome_2014], V4 [@goodrich_human_2014; @baxter_microbiota-based_2016; ], and V3-4 [@zeevi_personalized_2015] regions.
 For those studies where multiple regions were sequenced,  we selected the region that corresponded to the largest number of subjects [@turnbaugh_core_2009; @human_microbiome_project_consortium_structure_2012].
-We processed the 16S rRNA gene sequence data as described in the original studies.
-The taxonomic data from the metagenomic shotgun sequence data were obtained using Metaphlan2 [@truong_metaphlan2_2015].
+We processed the 16S rRNA gene sequence data using a standardized mothur pipeline.
+For a few studies author defined quality scores were used to maximize the number of samples for each study.
 A detailed and reproducible description of how the data were processed can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/.
@@
 We split the overall analysis into three general strategies.
-In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, and number of OTUs.
+In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, observed richness, and evenness.
 A two-tailed Wilcoxon test was performed for comparison for non-obese and obese individuals.
-We also measured the Spearman correlation coefficient between each metric and the actual BMI value.
 Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices.
 This analysis was performed using the vegan R package.
-Finally, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguish non-obese and obese individuals.
+Finally, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguishing non-obese and obese individuals.
 To generate Receiver Operator Characteristic (ROC) curves we used the pROC (v1.8) R package.
@@
 
-In the second strategy, we pooled the relative risk (RR) values for each study.
+In the second strategy, we pooled the relative risk (RR) values to assess whether there is an increased risk of obesity that is detectable based on specific microbiome metrics.
 For each study we partitioned the subjects into a low or high group depending on whether their diversity or B/F was below or above the median value for the study.
 The RR was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group.
-We then performed a Fisher exact-test to test whether the RR was significantly different from 1.0 using the epiR (v0.9-6.9) and metafor (v1.9-8) packages.
+We then performed a Fisher exact-test to investigate whether the RR was significantly different from 1.0 using the epiR (v0.9-6.9) and metafor (v1.9-8) packages.
 Finally, we assessed the potential publication bias in the diversity and B/F analyses using a funnel plot.
 
-In the third strategy, we normalized the diversity and B/F data within each study and pooled the normalized data across studies.
-We applied a Z-score normalization to the Shannon diversity data and log-transformed B/F data within each study to have a mean of zero and standard deviation of one.
-We then pooled Z-scores across studies and used a two-tailed T-test on the values to test for a significant difference in Z-scores among non-obese and obese individuals.
-
-In the final component of the analysis we performed power and sample number simulations for different effect sizes for each study.
+In the third strategy, we corrected for study effects and used linear mixed-effect models to test the pooled diversity and B/F ratio data for a difference between obese and non-obese groups.
+Additionally, we also performed power and sample number simulations for different effect sizes for each study.
 This was completed for all the different components measured as part of first strategy.

### 2016-05-25 submission/Sze_Obesity_PeerJ_2016.Rmd
paragraph1 results edits

@@
 We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Using a P-value cutoff for significance as less than 0.05 we found a only 1 significant result for OTU richness.
+Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))`significant results; two for Shannon diversity, three for observed richness, and one for evenness.
 There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio.

### 2016-05-25 submission/Sze_Obesity_PeerJ_2016_Supp.Rmd
Table S1 creation and removing old data links

@@
 
-OverallPValue <- read.csv("results/tables/denovoOverallPTable.csv")
-ShannonRRTable <- read.csv("results/tables/denovoShannonRRTable.csv")
-BFRatioRRTable <- read.csv("results/tables/denovoBFRatioRRTable.csv")
-
+alpha_tests.summary <- read.table("../data/process/alpha_tests.summary", header=T)
+beta_tests.summary <- read.table("../data/process/beta_tests.summary", header=T)
 ```
@@
 
-Study <- as.character(OverallPValue$X)
-Bacteroidetes <- format(round(OverallPValue$Bacteroidetes, 2), nsmall=2)
-Firmicutes <- format(round(OverallPValue$Firmicutes, 2), nsmall=2)
-BFRatio <- format(round(OverallPValue$BFRatio, 2), nsmall=2)
-Shannon <- format(round(OverallPValue$Shannon, 2), nsmall=2)
-OTURich <- format(round(OverallPValue$OTURich, 2), nsmall=2)
-Evenness <- format(round(OverallPValue$Evenness, 2), nsmall=2)
-BrayC <- format(round(OverallPValue$BrayC, 2), nsmall=2)
+Study <- unique(as.character(alpha_tests.summary$dataset))
+Bacteroidetes <- format(round(alpha_tests.summary[
+  which(alpha_tests.summary$metric == 'bacteroidetes'), 'p.value'], 3), nsmall=3)
+Firmicutes <- format(round(alpha_tests.summary[
+  which(alpha_tests.summary$metric == 'firmicutes'), 'p.value'], 3), nsmall=3)
+BFRatio <- format(round(alpha_tests.summary[
+  which(alpha_tests.summary$metric == 'bf_ratio'), 'p.value'], 3), nsmall=3)
+Shannon <- format(round(alpha_tests.summary[
+  which(alpha_tests.summary$metric == 'shannon'), 'p.value'], 3), nsmall=3)
+OTURich <- format(round(alpha_tests.summary[
+  which(alpha_tests.summary$metric == 'sobs'), 'p.value'], 3), nsmall=3)
+Evenness <- format(round(alpha_tests.summary[
+  which(alpha_tests.summary$metric == 'shannoneven'), 'p.value'], 3), nsmall=3)
+BrayC <- format(round(beta_tests.summary$p_value, 3), nsmall=3)
 
@@
 
-![Figure S1](results/figures/suppFigClostridiales.png)\
-
 **Figure S1: Boxplots of All Clostridiales OTUs Important for Non-Obese and Obese Classification.** A total of 6/8 studies had at least 1 OTU that was classified to Clostridiales.
@@
 
-![Figure S2](results/figures/suppFigLachno.png)\
-
 **Figure S2: Boxplots of All Lachnospiraceae OTUs Important for Non-Obese and Obese Classification.** A total of 6/8 studies had at least 1 OTU that was classified to Lachnospiraceae.
 However, overall there was a wide range in direction both between studies and within studies.
-
-![Figure S3](results/figures/suppFigRumi.png)\
 
@@
 
-```{r Table_S2, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Table_S2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
 
@@
 
-```{r Table_S3, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Table_S3, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
 
@@
 
-```{r Figure_S4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_S4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_S5, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_S6, echo=FALSE, message=FALSE, warning=FALSE, eval = FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S7, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_S7, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S8, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_S8, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S9, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_S9, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-```{r Figure_S10, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_S10, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 

### 2016-05-25 submission/Sze_Obesity_PeerJ_2016.Rmd
remove parts on Ruminococcace and such

@@
 Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05, Figure 2).
-One aspect we found surprising was that there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study.
-The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/8 studies), an OTU that could be classified to the family Lachnospiraceae (6/8 studies), Clostridiales (6/8), Bacteroides (5/8), and Shannon diversity metric (4/8 studies).
-
-Using the top three common taxonomic classifications identified across studies we investigated this further.
-Although these taxonomic level classifications showed up commonly across studies, the overall direction and magnitude varied quite a bit depending on the study [Figure S1-3].
-Further, when the sequences were BLAST the identifications within each OTU within and between groups varied quite widely with many having relatively poor query coverage and identity [Table S2].
-When the sequence identity for each specific variable region was compared using the EMBOSS Needle pairwise sequence alignment (http://www.ebi.ac.uk/Tools/psa/emboss_needle/nucleotide.html) the V4 region had the highest similarity [Table S3] but all had greater than 4% dissimilarity.
 

### 2016-05-25 submission/Sze_Obesity_PeerJ_2016_Supp.Rmd
remove BLAST and Ruminococcace stuff

@@
 
-**Figure S1: Boxplots of All Clostridiales OTUs Important for Non-Obese and Obese Classification.** A total of 6/8 studies had at least 1 OTU that was classified to Clostridiales.
-However, overall there was a wide range in direction both between studies and within studies.
-
-**Figure S2: Boxplots of All Lachnospiraceae OTUs Important for Non-Obese and Obese Classification.** A total of 6/8 studies had at least 1 OTU that was classified to Lachnospiraceae.
-However, overall there was a wide range in direction both between studies and within studies.
-
-**Figure S3: Boxplots of All Ruminococcaceae OTUs Important for Non-Obese and Obese Classification.** A total of 7/8 studies had at least 1 OTU that was classified to Ruminococcaceae  However, overall there was a wide range in direction both between studies and within studies.
-
-**Table 2. Summary of BLAST alignment of Representative OTU for Clostridiales, Lachnospiraceae, and Ruminococcaceae classification OTUs**
-
-```{r Table_S2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
-
-Study <- c("Wu", "Turnbaugh", "Ross", "Escobar", "Zupancic", "Baxter", "Goodrich", "HMP")
-VariableRegion <- c("V1-V2", "V2", "V1-V3", "V1-V3", "V1-V3", "V4", "V4", "V3-V5")
-ClosMostPrev <- c("OTU191:3 Clostridium", "None", "OTU102:8 Ruminococcus", "None", "OTU322:3 Parvimonas | OTU62:3 Parvimonas | OTU256:4 Clostridium", "OTU479:4 Anaerostipes", "OTU287:5 Clostridium | OTU662:8 Clostridium | OTU145:6 Clostridium | OTU244:3 Clostridium | OTU442:4 Clostridium", "OTU293:7 Clostridium")
-ClostHighMatch <- c("Anaerovorax odorimutans(81|95)", "None", "Ruminococcus champanellensis(100|100)", "None", "OTU322:Eubacterium hallii(92|100) | OTU62:Eubacterium hallii(92|99) | OTU256:Romboutsia lituseburensis(92|99)", "Anaerosporbacter mobilis(100|94)", "OTU287:Clostridium fimetarium(100|94) |  OTU662:Ruminiclostridium thermocellum(100|91) | OTU145:Ruminococcus lactaris(100|100) | OTU244:Anaerosporobacter mobilis(100|94) | OTU442:Clostridium propionicum(100|92)", "Anaerobacterium chartisolvens (100|91)")
-LachMostPrev <- c("OTU75:6 Clostridium | OTU229:5 Peptostreptococcus | OTU5: 4 Blautia", "OTU318:", "OTU177:3 Roseburia", "OTU67:6 Eubacterium | OTU74:5 Blautia | OTU80:4 Blautia", "None", "OTU36:8 Clostridium", "OTU180:4 Clostridium | OTU75: 5 Clostridium | OTU122:5 Roseburia | OTU92:8 Clostridium | OTU673:7 Clostridium", "None")
-LachHighMatch <- c("Clostridium bolteae(100|96)", "", "Clostridium Spp(100|98)", "OTU67:Dorea formicigenerans(100|96) | OTU74:Blautia luti(100|96) | OTU80:Blautia producta(100|93)", "None", "Lachnobacterium bovis(100|96)", "OTU180:Hungatella effluvii(100|95) | OTU75:Ruminococcus gnavus(100|100) | OTU122:Eubacterium ramulus(100|96) | OTU92:Fusicatenibacter saccharivorans(100|94) | OTU673:Bacteroides xylanolyticus(92|97) | OTU523:Eubacterium rectale(100|97)", "None")
-RumiMostPrev <- c("OTU123:4 Clostridium", "OTU30:", "OTU22:3 Oscillibacter", "None", "OTU398:3 Eubacterium | OTU850:3 Ethanoligenens | OTU817:8 Ruminococcus", "OTU320:4 Clostridium", "OTU68:5 Ruminococcus | OTU255:3 Blautia | OTU1145:3 Ethanoligenens | OTU558:8 Ruminococcus | OTU297:3 Oscillibacter | OTU6:4 Clostridium | OTU132:2 Ruminoclostridium", "OTU252:10 Clostridium")
-RumiHighMatch <- c("Anaerotruncus colihominis(85|95)", "", "Oscillibacter Spp(100|97)", "None", "OTU398:Oscillibacter valericigene(77|96) | OTU850:Papillibacter cinnamivorans(90|91) | OTU817:Ruminococcus callidus(91|95)", "Sporobacter termitidis(100|94)", "OTU68:Acetanaerobacterium elongatum(100|89) | OTU255:Pseudoflaconifractor capillosus(100|91) | OTU1145:Subdoligranulum variable(100|95) | OTU558:Ruminococcus champanellensis(92|98) | OTU297:Flavonifractor plautii(100|95) | OTU6:Oscillibacter valericigenes(100|93) | OTU132:Intestinimonas butyriciproducens(100|92)", "Clostridium Spp(100|97)")
-
-tableS2 <- cbind(Study, VariableRegion, ClosMostPrev, ClostHighMatch, LachMostPrev, LachHighMatch, RumiMostPrev, RumiHighMatch)
-colnames(tableS2) <- c("Study", "Hypervariable Region", "Clostridiales OTU Most Prevalent Match", "Clostridiales OTU Highest Match (Query Cover|Identity)", "Lachnospiraceae OTU Most Prevalent Match", "Lachnospiraceae OTU Highest Match (Query Cover|Identity)", "Ruminococcaceae OTU Most Prevalent Match", "Ruminococcaceae OTU Highest Match (Query Cover|Identity)")
-
-kable(tableS2, format = "markdown", align = 'c', padding = 2)
-```
-
-**Table S3. Top Sequence Similarity by Variable Region Based on the Representative OTU for Clostridiales, Lachnospiraceae, and Ruminococcaceae**
-
-```{r Table_S3, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
-
-VariableRegion <- c("V1-V2 vs V2", "V1-V3", "V4")
-
-TopHitClos <- c("None", "50.3|39.1", "89.4|2.8")
-AverageClos <- c("None", "48.0|42.4", "87.7|2.4")
-TopHitLachno <- c("TBC", "40.1|47.9", "92.9|0.8")
-AverageLachno <- c("TBC", "35.9|52.5", "85.8|5.6")
-TopHitRumi <- c("TBC", "30.4|68.4", "86.6|0.8")
-AverageRumi <- c("TBC", "29.2|69.1", "81.6|6.6")
-
-tableS3 <- cbind(VariableRegion, TopHitClos, AverageClos, TopHitLachno, AverageLachno, TopHitRumi, AverageRumi)
-colnames(tableS3) <- c("Variable Region", "Top Similarity Colstridiales (Id|Gaps)", "Average Similarity Colstridiales (Id|Gaps)", "Top Similarity Lachnospiraceae (Id|Gaps)", "Average Similarity Lachnospiraceae (Id|Gaps)", "Top Similarity Ruminococcaceae (Id|Gaps)", "Average Similarity Ruminococcaceae (Id|Gaps)")
-
-kable(tableS3, format = "markdown", align = 'c', padding = 2)
-
-```
-
-```{r Figure_S4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_S1, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-**Figure S4: Funnel Plot of the Shannon Diversity and Bacteroidetes/Firmicutes Ratio Relative Risk.**  **A)** Overall there does not seem to be any bias associated with the studies selected for the Shannon Diversity analysis with studies falling on either side of the predicted value and those with smaller total n falling further away from this.
+**Figure S1: Funnel Plot of the Shannon Diversity and Bacteroidetes/Firmicutes Ratio Relative Risk.**  **A)** Overall there does not seem to be any bias associated with the studies selected for the Shannon Diversity analysis with studies falling on either side of the predicted value and those with smaller total n falling further away from this.
 **B)** For the Bacteroidetes/Firmicutes ratio there does not seem to be any associated bias.
@@
 
-```{r Figure_S5, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_S2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
-**Figure S5.
+**Figure S2.
 Summary of Power and Sample Size Simulations for OTU Richness (S) for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
@@
 
-```{r Figure_S6, echo=FALSE, message=FALSE, warning=FALSE, eval = FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
+```{r Figure_S3, echo=FALSE, message=FALSE, warning=FALSE, eval = FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
 
@@
 
+**Figure S3.
+Summary of Power and Sample Size Simulations for Evenness (J) for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
+
+```{r Figure_S4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
+
+grid.arrange(b2, bn2, nrow=2)
+```
+
+**Figure S4.
+Summary of Power and Sample Size Simulations for Bacteroidetes for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
+
+```{r Figure_S5, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
+
+grid.arrange(f2, fn2, nrow=2)
+```
+
+**Figure S5.
+Summary of Power and Sample Size Simulations for Firmicutes for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**A)** The dotted lines represent the actual effect size for each specific study.
+**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
+
+```{r Figure_S6, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
+
+grid.arrange(RRH2, HRRn2, nrow=2)
+```
+
 **Figure S6.
-Summary of Power and Sample Size Simulations for Evenness (J) for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+Summary of Power and Sample Size Simulations for Relative Risk  for Obesity based on Shannon Diversity**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
 **A)** The dotted lines represent the actual effect size for each specific study.
@@
 
-grid.arrange(b2, bn2, nrow=2)
+grid.arrange(RRBF2, BFRRn2, nrow=2)
 ```
@@
 **Figure S7.
-Summary of Power and Sample Size Simulations for Bacteroidetes for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-```{r Figure_S8, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-grid.arrange(f2, fn2, nrow=2)
-```
-
-**Figure S8.
-Summary of Power and Sample Size Simulations for Firmicutes for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-```{r Figure_S9, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-grid.arrange(RRH2, HRRn2, nrow=2)
-```
-
-**Figure S9.
-Summary of Power and Sample Size Simulations for Relative Risk  for Obesity based on Shannon Diversity**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-```{r Figure_S10, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-grid.arrange(RRBF2, BFRRn2, nrow=2)
-```
-
-**Figure S10.
 Summary of Power and Sample Size Simulations for Relative Risk  for Obesity based on B/F Ratio**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.

### 2016-05-25 submission/Sze_Obesity_PeerJ_2016.Rmd
text reflects new Figure nums

@@
 We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))`significant results; two for Shannon diversity, three for observed richness, and one for evenness.
+Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` significant results; two for Shannon diversity, three for observed richness, and one for evenness.
 There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio.
@@
 We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis.
-From the graph we can observe that there does not seem to be any significant  bias [Figure S4].
+From the graph we can observe that there does not seem to be any significant  bias [Figure S1].
 Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.
@@
 In contrast in order to detect a 5% difference with 80% power in the B/F ratio a sample size of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) per group [Figure 6B].
-A complete breakdown of the power simulations and sample size estimates for the other variables investigated and RR can be found in the online supplement.
+A complete breakdown of the power simulations and sample size estimates for the other variables investigated and RR can be found in the online supplement [Figure S2-S7].
 

### 2016-05-26 submission/Sze_Obesity_PeerJ_2016.Rmd

@@
 We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` significant results; two for Shannon diversity, three for observed richness, and one for evenness.
-There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio.
+Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05), ]))` significant results; two for Shannon diversity, three for observed richness, one for evenness, and one for Bacteroidetes [Table S1].
+There were no significant results for Firmicutes or the B/F ratio.
 The full summary of the exact P-values for each respective study and variable tested can be found in table S1 of the online supplement.
@@
 We analyzed the data using three different approaches and also performed post-hoc power and sample size calculations and simulations.
-The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses Z-score normalizations and does not use RR.
+The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach uses linear mixed effect models to adjust for study effects and invetigates both predicted power and sample number needed for the observed effect sizes.
 
@@
 Further, even though random forest could identify specific variables with similar taxonomic classifications that resulted in good ROC curves [Figure 1] these variables varied widely by both their actual sequence and directionality.
-Similarily, we found that the other two approaches, that used the pooled data approach, was also not able to identify a correlation between measures of the microbiome and obesity [Figure 3 & 4].
-For these approaches both Shannon diversity and B/F ratio failed to have a significant correlation to obesity.
+Similarily, for the B/F ratio we found that the other two approaches, that used the pooled data approach, were also not able to identify a correlation between measures of the microbiome and obesity [Figure 3 & 4].
+For these approaches both the B/F ratio failed to have a significant correlation to obesity.
+However, a small but significant differences was observed between non-obese and obese individauls  for Shannon diversity [Figure 3 & 4].
 In order to also assess the amount of our results that could be due to insufficient power we performed power and sample size analysis.
-Overall we found that most studies for Shannon Diversity would be powered to find a 15% difference, however, the vast majority of reported effect sizes between studies ranged from `r format(round(min(alpha_effectSize[, 'shannon']), 2), nsmall=2)`-`r format(round(max(alpha_effectSize[, 'shannon']), 2), nsmall=2)` percent.
+Overall, we found that most studies for Shannon Diversity would be powered to find a 15% difference, however, the vast majority of reported effect sizes between studies ranged from `r format(round(min(alpha_effectSize[, 'shannon']), 2), nsmall=2)`-`r format(round(max(alpha_effectSize[, 'shannon']), 2), nsmall=2)` percent.
 Although the effect sizes for the BF ratio were larger than that for Shannon Diversity [Figure 5A & 6A] the estimated power remained low.
@@
 
-Second, although we make attempts to normalize the data for the pooled analysis so that we can compare the microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.
-To help mitigate some of these problems we analyzed each data set separately and processed all samples through a uniform pipeline in an attempt to mitigate some of the variation that might occur due to sequence processing.
+Second, although we make attempts to correct for study effects in our pooled analysis  so that we can compare the microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.
+To help mitigate some of these problems we analyzed each data set separately and processed all samples through a uniform pipeline in an attempt to mitigate a portion of the variation that might occur due to sequence processing.
 It should also be noted that we also do not see any specific bias in the forest plots [Figure 3] based on hypervariable region.
@@
 However, taking both the sample size calculations and power calculations together would indicate that if there is a true difference to be found between non-obese and obese individuals this difference would be very small.
-Additionally, due to the very small signal another important question would be whether or not this difference is truly biologically meaningful.
+Additionally, due to the very small effect size potentially the more important question that does need to be raised is whether or not this difference is truly biologically meaningful.
 
 Fourth, although the ROC curves for each study were relatively good there was very little actual overlap between studies based on sequence similarity and directionality.
-This would suggest that within each study classification of obese and non-obese is relatively strong but not generalizable across studies.
+Further none of the selected variables from each indivdiual study did a respectable job of classification when applied to the other data sets.
+This would suggest that within each indivdiual study classification of obese and non-obese is relatively strong but not generalizable across studies.
 This might indicate an ancestry, ethnicity, or a specific environmental effect such as diet difference between study populations that might cause this dissimilarity.
-It is also possible that the random forest model overfit the data making it less generalizable across studies.
+Additionally, an alternative explanation that may be more reasonable is that the random forest model is overfit for some of the studies examined which explains why the variables selected are less generalizable across studies.
+In actuality it may be a combination of all of the factors mentioned as to why generalizability is problematic.
 
@@
 There is a large body of very strong evidence in mice on the effect the microbiome has on obesity [@turnbaugh_effect_2009; @koren_host_2012].
-It is not possible to discount these findings and it may be the case that in humans there is just too many other variables and factors that effect the microbiome which makes the signal to hard to detect in a cross sectional manner.
-Some well designed studies in humans and diet, in the pasat, have shown that the noise inherent to microbiome studies can be overcome with very careful attention made to controlling the study conditions [@ley_microbial_2006].
+It is not possible to discount these findings and it may be the case that in humans there is just too many other variables and factors that effect the microbiome which makes the signal too hard to detect in a cross sectional manner.
+Some well designed studies in humans and diet, in the past, have shown that the noise inherent to microbiome studies can be overcome with very careful attention made to controlling the study conditions [@ley_microbial_2006].
 
-Overall, this study provides a reasonable follow up to the two previous re-analysis studies to assess the correlation between the microbiome and obesity.
+Overall, this study provides a reasonable follow up to the two previous re-analysis studies to assess the association between the microbiome and obesity.
 In doing so we believe that we have provided evidence that if there is an effect size present it is small and will take many more studies and samples to be analyzed to confirm a difference between non-obese and obese individuals.
@@
 The authors would like to thank Nielson Baxter and Shawn Whitefield for their suggestions on the manuscript.
-Finally, we would like to thank all the authors who have made there data publicly available.
+Finally, we would like to thank all the authors who have made their data publicly available.
 Without this initial forethought studies such as these would not be possible.

### 2016-05-26 code/make_table1.R
R code to make table 1

@@ -1,0 +1,147 @@
+# Table 1 Creation 
+# Make all the measurements for the table here so it just needs to be read into
+# the Rmd.
+
+
+# Read in necessary tables
+meta.bax <- read.table("../data/baxter/baxter.metadata", header = T)
+meta.esc <- read.table("../data/escobar/escobar.metadata", header = T)
+meta.goo <- read.table("../data/goodrich/goodrich.metadata", header = T)
+meta.hmp <- read.table("../data/hmp/hmp.metadata", header = T)
+meta.ros <- read.table("../data/ross/ross.metadata", header = T)
+meta.sch <- read.table("../data/schubert/schubert.metadata", header = T)
+meta.tur <- read.table("../data/turnbaugh/turnbaugh.metadata", header = T)
+meta.wu <- read.table("../data/wu/wu.metadata", header = T)
+meta.zee <- read.table("../data/zeevi/zeevi.metadata", header = T)
+meta.zup <- read.table("../data/zupancic/zupancic.metadata", header = T)
+beta_tests.summary <- read.table("../data/process/beta_tests.summary", 
+                                 header=T, row.names = 1)
+
+# Create data table entries
+Study <- as.character(rownames(beta_tests.summary))
+
+Age <- round(c(mean(meta.bax$age), mean(meta.esc$age), mean(meta.hmp$age), 
+               mean(meta.ros$age), mean(meta.sch$age), mean(meta.tur$age), 
+               mean(meta.wu$age), mean(as.numeric(as.character(meta.zup$age)), 
+                                       na.rm=TRUE), mean(meta.goo$age), 
+               mean(meta.zee$age, na.rm=TRUE)), 2)
+
+AgeSD <- round(c(sd(meta.bax$age), sd(meta.esc$age), sd(meta.hmp$age), 
+                 sd(meta.ros$age), sd(meta.sch$age), sd(meta.tur$age), 
+                 sd(meta.wu$age), sd(as.numeric(as.character(meta.zup$age)), 
+                                     na.rm=TRUE), sd(meta.goo$age), 
+               sd(meta.zee$age, na.rm = TRUE)), 2)
+
+AgeTable <- paste(as.character(Age)," (",as.character(AgeSD),")", sep="")
+
+# No female/male information from Turnbaugh study so have to use NA 
+females <- c(
+  length(meta.bax$sex[meta.bax$sex == 'f']), 
[…]

### 2016-05-26 code/run_aucrf_genus.R
Calculate confidence interval for accuracy

@@ -38,26 +38,11 @@ get_genus_shared <- function(dataset){
 	return(shared_genus)
 }
 
-get_metrics <- function(predicted, reference){
-
-	contingency_table <- table(predicted = predicted, reference=reference)
-
-	sensitivity <- contingency_table[2,2] / sum(contingency_table[,2])
-	specificity <- contingency_table[1,1] / sum(contingency_table[,1])
-	accuracy <- (contingency_table[1,1] + contingency_table[2,2]) / sum(contingency_table)
-	posPredValue <- contingency_table[2,2] / sum(contingency_table[2,])
-	negPredValue <- contingency_table[1,1] / sum(contingency_table[1,])
-
-	return(c(sensitivity = sensitivity, specificity = specificity,
-		 			accuracy = accuracy, posPredValue = posPredValue,
-					negPredValue = negPredValue))
-}
-
 run <- function(datasets){
 
 	datasets <- unlist(strsplit(datasets, split=" "))
 
-#	datasets <- c('baxter', 'escobar', 'turnbaugh', 'wu', 'ross', 'hmp', 'zupancic')
+#	datasets <- c('baxter', 'escobar', 'turnbaugh', 'wu', 'ross', 'hmp', 'zupancic', 'zeevi', 'schubert', 'goodrich')
 
 	test_data <- list()
 	model <- list()
@@ -65,6 +50,7 @@ run <- function(datasets){
 	model_summary <- NULL
 	roc_summary <- NULL
 
+	datasets <- c("baxter", "zupancic")
 
 	for(d in datasets){
 		print(d)
@@ -108,17 +94,17 @@ run <- function(datasets){
 		model[[d]]$opt_threshold <- roc[[d]]$thresholds[opt_index]
 
 		predicted <- factor(probabilities >= model[[d]]$opt_threshold, levels=c("FALSE", "TRUE"))
-		metrics <- get_metrics(predicted, metadata$obese)
 
-		model[[d]]$opt_accuracy <- metrics["accuracy"]
-		model[[d]]$opt_posPredValue <- metrics["posPredValue"]
-		model[[d]]$opt_negPredValue <- metrics["negPredValue"]
+		accuracy <- ci.coords(roc[[d]], x=model[[d]]$opt_threshold, input = "threshold", ret="acc", progress="none")
 
[…]

### 2016-05-26 code/run_aucrf_otus.R
Output observed ROC data

@@ -48,7 +48,7 @@ run <- function(datasets){
 
 		auc_cv <- cross_validate(model)
 
-		roc_summary <- rbind(roc_summary, data.frame(dataset=d, sensitivity = auc_cv$sens_spec["sens"], specificity = auc_cv$sens_spec["spec"]))
+		roc_summary <- rbind(roc_summary, data.frame(dataset=d, sensitivity = roc$sensitivities, specificity = roc$specificities))
 
 		model_summary <- rbind(model_summary, c(d, auc, auc_cv$cv_est, k_opt, otus))
 
@@ -57,6 +57,7 @@ run <- function(datasets){
 	colnames(model_summary) <- c("dataset", "auc_lci", "auc", "auc_hci", "auc_cv", "k_opt",  "otus")
 	write.table(model_summary, file="data/process/random_forest.otu.summary", quote=F, sep='\t', row.names=F)
 
+	colnames(roc_summary) <- c("dataset", "sensitivity", "specificity")
 	write.table(roc_summary, file="data/process/random_forest.otu.roc_data", quote=F, sep='\t', row.names=F)
 
 }

### 2016-05-26 code/plot_alpha_diversity.R
Move Bacteroides legend

@@ -81,7 +81,7 @@ pdf(file="alpha.pdf")
 diversity_plot('shannon', leg=c(0,9))
 diversity_plot('shannoneven', leg=c(0,9))
 diversity_plot('sobs', leg=c(0,9))
-diversity_plot('bacteroidetes', leg=c(0,9))
+diversity_plot('bacteroidetes', leg=c(0.5,10))
 diversity_plot('firmicutes', leg=c(0,9))
 diversity_plot('bf_ratio', leg=c(4,9))
 dev.off()

### 2016-05-26 submission/Sze_Obesity_PeerJ_2016.Rmd

@@
 **Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
-```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
+```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
 
-Study <- as.character(rownames(beta_tests.summary))
-
-Age <- as.character(demographics$meanAge)
-AgeSD <- as.character(format(round(
-  demographics$SDAge, 2), nsmall = 2))
-AgeTable <- paste(Age," (",AgeSD,")", sep="")
-
-Sex <- paste(demographics$females, "|", demographics$males)
-
-Ancestry <- as.character(format(round(
-  demographics$ancestry*100, 2), nsmall = 2))
-
-ObeseTotal <- ShannonRRTable$tposH + ShannonRRTable$cposH
-ContTotal <- ShannonRRTable$tnegH + ShannonRRTable$cnegH
-
-ObBreakdown <- paste(ContTotal, " | ", ObeseTotal, sep="")
-
-BMI <- paste(demographics$meanBMI, " (", demographics$SDBMI, ")", sep="")
-
-Min <- as.character(demographics$minBMI)
-
-Max <- as.character(demographics$maxBMI)
-
-table1 <- cbind(Study, AgeTable, Sex, Ancestry, ObBreakdown, BMI, Min, Max)
-table1[table1=="NA" | table1=="NA | NA" | table1=="NA (NA)" |
-         table1=="      NA"] <- "Not Available"
+table1 <- read.table("../results/tables/table1.demographics", header = T)
 

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016.Rmd
add Table 2 to manuscript

@@
 
-```{r Table_2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
+```{r Table_2, echo=FALSE, message=FALSE, warning=FALSE}
 
-Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "HMP", "Wu", "Turnbaugh")
+Study <- as.character(table1$Study)
+#New Order
+  #baxter, escobar, hmp, ross, schubert, turnbaugh, wu, zupancic, goodrich, zeevi
 
-Type <- c("Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational")
+Type <- c("Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational")
 
-Population <- c("Adults without Cancer from Texas, Ontario, Michigan, Massachusetts", "Hispanic adults with and without type 2 diabetes from Texas
-", "Adult twins and mothers from the United Kingdom", "Healthy adults from Columbia", "Amish Adults with and without metabolic syndrome from Pennsylvania", "Healthy Adults from Texas and Missouri", "Healthy Adults from Pennsylvania", "Adult twins and mothers from Missouri")
+Population <- c("Adults without Cancer from Texas, Ontario, Michigan, Massachusetts", "Healthy adults from Columbia", "Healthy Adults from Pennsylvania", "Hispanic adults with and without type 2 diabetes from Texas
+", "Nondiarrheal indivdiuals without infection history", "Adult twins and mothers from Missouri", "Healthy Adults from Texas and Missouri", "Amish Adults with and without metabolic syndrome from Pennsylvania", "Adult twins and mothers from the United Kingdom", "Healthy Adults from Tel Aviv")
 
-Published <- c("No", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes","Yes")
+Published <- c("Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes","Yes", "Yes", "Yes")
 
-Outcomes <- c("N/A (Unpublished)", "Mentions obesity, no specific correlations made though", "Christensenellaceae higher in low BMI individuals.
-No other correlations found.", "Firmicutes less abundant at higher BMI", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "No significant correlations", "Unweighted UniFrac Correlation with BMI", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ")
+Outcomes <- c("Did not investigate control populations for relationship with BMI", "Firmicutes less abundant at higher BMI", "No significant correlations", "Mentions obesity, no specific correlations made though", "Did not investigate control populations for relationship with BMI", "Unweighted UniFrac Correlation with BMI", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "Christensenellaceae higher in low BMI individuals.
+No other correlations found.", "Found associations with postprandial (post meal) glycemic responses, which correlates with BMI, and Proteobacteria, Actinobacteria, and Enterobacteriaceae")
 

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016.Rmd
change manuscript Fig 2 legend

@@
 
-**Figure 2: Receiver Operating Characteristic Curve of the Random Forest Analysis of Each Study.** After adjusting for multiple comparisons there was no significant difference between any of the 8 studies examined (P-value > 0.05).
+**Figure 2: Individual and Combined comparison of Obese and Non-Obese groups for Shannon Diversity and B/F Ratio.** A) For Shannon diversity only Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 2), nsmall = 2)` and Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 5), nsmall = 5)`) studies indivdiually were signficantly different for obese and non-obese groups.
+When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 5), nsmall = 5)`).
+B) For the B/F ratio there was no individual study that was significant nor was their a difference when the data was combined and study effect corrected for between Non-Obese and Obese groups (P > 0.05).
 

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016.Rmd

@@
 **Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity or Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low Shannon Diversity or Bacteroidetes/Firmicutes (B/F) ratio groups based on the median for that study.
-Red represents studies that used the 16S rRNA V4 region, yellow represents studies that used a V1-V3 approach, blue represents studies that used either a V1-V2 or just a V2 approach, and orange represents studies that used the V3-V5 region.
+Red represents studies that used the 16S rRNA V4 region, yellow represents studies that used a V1-V3 approach, blue represents studies that used either a V1-V2 or just a V2 approach, green represents studies that used a V3-V4 region, and orange represents studies that used the V3-V5 region.
 **A)** The overall pooled relative risk for Shannon Diversity was `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` with a P-value of `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`.

### 2016-05-27 code/run_aucrf_genus.R
Remove testing dataaset vector

@@ -50,7 +50,6 @@ run <- function(datasets){
 	model_summary <- NULL
 	roc_summary <- NULL
 
-	datasets <- c("baxter", "zupancic")
 
 	for(d in datasets){
 		print(d)

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016.Rmd
edit Figure 4 legend

@@
 
-![Figure 1](../results/figures/Figure_1.png)\
-
 **Figure 1: PRISMA Flow Diagram of Total Records Searched.**  This was adapted from:  Moher D, Liberati A, Tetzlaff J, Altman DG, The PRISMA Group (2009).
@@
 
-```{r Figure_2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}
-combinedData <- read.csv("results/tables/ROCcombinedData.csv", stringsAsFactors = FALSE)
-
-diag <- data.frame(x=seq(1, 0), y = seq(0, 1))
-diag$Study <- "Control"
-
-rocPlot <- ggplot(data=combinedData,
-                  aes(x = Sensitivity, y = Specificity))
-rocPlot + geom_line(aes(group=Study, color=Study)) +
-  scale_colour_brewer(palette="Set1") +
-  scale_x_continuous(trans = "reverse") +
-  geom_line(data=diag, aes(x=x, y=y), linetype="dashed") +
-  theme(axis.line=element_line(colour="black"),
-       panel.grid.major = element_blank(),
-       panel.grid.minor = element_blank(),
-       panel.border = element_blank(),
-       panel.background = element_blank(),
-       legend.key = element_rect(fill = "white"))
-
-```
-
-**Figure 2: Individual and Combined comparison of Obese and Non-Obese groups for Shannon Diversity and B/F Ratio.** A) For Shannon diversity only Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 2), nsmall = 2)` and Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 5), nsmall = 5)`) studies indivdiually were signficantly different for obese and non-obese groups.
+**Figure 2: Individual and Combined comparison of Obese and Non-Obese groups for Shannon Diversity and B/F Ratio.** **A)** For Shannon diversity only Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 2), nsmall = 2)` and Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 5), nsmall = 5)`) studies indivdiually were signficantly different for obese and non-obese groups.
 When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 5), nsmall = 5)`).
-B) For the B/F ratio there was no individual study that was significant nor was their a difference when the data was combined and study effect corrected for between Non-Obese and Obese groups (P > 0.05).
-
-```{r Figure_3, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-grid.arrange(HRRG1, BFRRG1, ncol=2)
-
-```
+**B)** For the B/F ratio there was no individual study that was significant nor was their a difference when the data was combined and study effect corrected for between Non-Obese and Obese groups (P > 0.05).
 
@@
 
-```{r Figure_4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-# Section A
-p1 <- ggplot(group.data.test, aes(obese, ZH))
-p2 <- p1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
-p3 <- p2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-p4 <- p3 + xlab("") + ylab("Z-score Normalized H") + scale_x_discrete(breaks=c("No", "Yes"), labels=c("Non-Obese", "Obese"))
-p5 <- p4 + guides(fill=FALSE)
-p6 <- p5 + theme(axis.line=element_line(colour="black"),
-                 panel.grid.major = element_blank(),
-                 panel.grid.minor = element_blank(),
-                 panel.border = element_blank(),
-                 panel.background = element_blank())
-p7 <- p6 + ggtitle('A') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
-
-# Section B
-bf1 <- ggplot(group.data.test, aes(obese, ZLogBF))
-bf2 <- bf1 + geom_boxplot(aes(fill = obese)) + scale_fill_brewer(palette="Set1")
-bf3 <- bf2 + theme(axis.text=element_text(size=8), axis.title=element_text(size=10, face="bold"))
-bf4 <- bf3 + xlab("") + ylab("Z-score Normalized Log B/F Ratio") +
-  scale_x_discrete(breaks=c("No", "Yes"), labels=c("Non-Obese", "Obese"))
-bf5 <- bf4 + guides(fill=FALSE)
-bf6 <- bf5 + theme(axis.line=element_line(colour="black"),
-                   panel.grid.major = element_blank(),
-                   panel.grid.minor = element_blank(),
-                   panel.border = element_blank(),
-                   panel.background = element_blank())
-bf7 <- bf6 + ggtitle('B') + theme(plot.title=element_text(hjust=0, face="bold", size=20))
-
-grid.arrange(p7, bf7, ncol=2)
-```
-
-**Figure 4: Pooled Analysis of Shannon Diversity and B/F ratio by either Obesity Status or BMI Groups.**  **A)** Z-score Normalized Shannon Diversity and Obesity.
-There was a significant difference between non-obese (n = `r length(which(alpha.data.ALL$obese == FALSE))`) and obese groups (n = `r length(which(alpha.data.ALL$obese == TRUE))`) (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`).
-**B)** Z-score Normalized B/F Ratio and Obesity.
-There was no significant difference between the non-obese (n = `r length(which(alpha.data.ALL$obese == FALSE))`) and obese (n = `r length(which(alpha.data.ALL$obese == TRUE))`) group based on B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`).
+**Figure 4: ROC Curves for Each Study based on Classification of Non-Obese or Obese Groups at the Genus or OTU level.**  **A)** No significant difference was found between the different ROC curves based on Study for the genus level.
+The overall AUCs ranged from `r round(min(rand_forest.genus.summary$auc), 2)` to `r round(max(rand_forest.genus.summary$auc), 2)` with a mean of `r round(mean(rand_forest.genus.summary$auc), 2)`.
+**B)** Similarly no difference was found between the ROC curves based on Study at the OTU level.
+The overall mean of the AUCs were slightly higher overall than the genus level `r round(mean(rand_forest.otu.summary$auc), 2)` but the range was greater going from `r round(min(rand_forest.otu.summary$auc), 2)` to `r round(max(rand_forest.otu.summary$auc), 2)`.
 
@@
 
-**Figure 5.
-Summary of Power and Sample Size Simulations for Shannon Diversity for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**Figure 5: Summary of Power and Sample Size Simulations for Shannon Diversity for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
 **A)** The dotted lines represent the actual effect size for each specific study.

### 2016-05-27 code/run_aucrf_genus.R
Add sensitivity and specificity to roc_summary

@@ -104,6 +104,8 @@ run <- function(datasets){
 
 		model_summary <- rbind(model_summary, c(d, model[[d]]$auc,
 			 									model[[d]]$auc_cv["cv_est"], k_opt, otus,
+												model[[d]]$opt_sensitivity,
+												model[[d]]$opt_specificity,
 												model[[d]]$opt_threshold))
 	}
 

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016.Rmd
edit Figure 5 legend

@@
 
+**Figure 5: Overall Accuracy of Each Study to Predict Non-Obese and Obese Individuals based on that Study's Classification Variables** Overall there is no significant difference between the studes in mean accuracy.
+Additionally, the accuracy for each study group to predict Non-Obese and Obese individuals was relatively low.
+
 ```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016.Rmd
change figure numbers to correct order

@@
 
-```{r Figure_5, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-source("code/MakeFiguresCode.R")
-
-grid.arrange(a2, b2, nrow=2)
-
-```
-
-**Figure 5: Summary of Power and Sample Size Simulations for Shannon Diversity for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
+**Figure 6: Summary of Power and Sample Size Simulations for Shannon Diversity for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
 **A)** The dotted lines represent the actual effect size for each specific study.
@@
 
-```{r Figure_6, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-grid.arrange(c2, d2, nrow=2)
-
-```
-
-**Figure 6.
+**Figure 7.
 Summary of Power and Sample Size Simulations for Bacteroidetes/Firmicutes Ratio for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016_Supp.Rmd
Fix supplement figure legends

@@
 
-**Table S1. Summary of P-values for Measurements of Interest for each Individual Study for Obese versus Normal**
+```{r load_needed_data_tables, echo=FALSE, message=FALSE, warning=FALSE}
 
-```{r Table_S1, echo=FALSE, message=FALSE, warning=FALSE}
-
-library(knitr)
-
-Study <- unique(as.character(alpha_tests.summary$dataset))
-Bacteroidetes <- format(round(alpha_tests.summary[
-  which(alpha_tests.summary$metric == 'bacteroidetes'), 'p.value'], 3), nsmall=3)
-Firmicutes <- format(round(alpha_tests.summary[
-  which(alpha_tests.summary$metric == 'firmicutes'), 'p.value'], 3), nsmall=3)
-BFRatio <- format(round(alpha_tests.summary[
-  which(alpha_tests.summary$metric == 'bf_ratio'), 'p.value'], 3), nsmall=3)
-Shannon <- format(round(alpha_tests.summary[
-  which(alpha_tests.summary$metric == 'shannon'), 'p.value'], 3), nsmall=3)
-OTURich <- format(round(alpha_tests.summary[
-  which(alpha_tests.summary$metric == 'sobs'), 'p.value'], 3), nsmall=3)
-Evenness <- format(round(alpha_tests.summary[
-  which(alpha_tests.summary$metric == 'shannoneven'), 'p.value'], 3), nsmall=3)
-BrayC <- format(round(beta_tests.summary$p_value, 3), nsmall=3)
-
-tableS1 <- cbind(Study, Bacteroidetes, Firmicutes, BFRatio, Shannon, OTURich, Evenness, BrayC)
-colnames(tableS1) <- c("Study", "Bacteroidetes", "Firmicutes", "B/F Ratio", "Shannon Diversity", "OTU Richness", "Evenness", "Bray Curtis")
-
-kable(tableS1, format = "markdown", align = 'c', padding = 2)
+alpha_tests.summary <- read.table("../data/process/alpha_tests.summary", header=T)
+alpha_composite.summary <- read.table("../data/process/alpha_composite.summary")
 
@@
 
-```{r Figure_S1, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
+**Figure S1: Individual and Combined comparison of Obese and Non-Obese groups for Observed OTUs.**  For observed OTUs Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 2), nsmall = 2)`, Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 5), nsmall = 5)`), and Turnbaugh (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'turnbaugh' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 2), nsmall = 2)`) studies individually were signficantly different for obese and non-obese groups.
+When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 5), nsmall = 5)`)
 
-library(metafor)
+**Figure S2: Individual and Combined comparison of Obese and Non-Obese groups for Evenness.**  For evenness only the Goodrich study (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannoneven'), 'p.value'], 3), nsmall = 3)`)  individually was signficantly different for obese and non-obese groups.
+When the data sets were combined and study effect was corrected for,  there was an overall significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 2), nsmall = 2)`)
 
-# Shannon Diversity Data for Funnel Plot
-StudyNo <- as.numeric(ShannonRRTable$X)
-Study <- as.character(ShannonRRTable$Study)
-Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
-Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
-tpos <- as.numeric(ShannonRRTable$tposH) #obese in low diversity group
-tneg <- as.numeric(ShannonRRTable$tnegH) #nonobese in low diversity group
-cpos <- as.numeric(ShannonRRTable$cposH) #obese in high diversity group
-cneg <- as.numeric(ShannonRRTable$cnegH) #nonobese in high diversity group
-RR <- as.numeric(ShannonRRTable$RRH)
-low <- as.numeric(ShannonRRTable$lowH)
-high <- as.numeric(ShannonRRTable$highH)
+**Figure S3: Individual and Combined comparison of Obese and Non-Obese groups for Bacteroidetes Relative Abundance.**  Only the Escobar study (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'escobar' & alpha_tests.summary$metric == 'bacteroidetes'), 'p.value'], 3), nsmall = 3)`)  individually was signficantly different for obese and non-obese groups.
+When the data sets were combined and study effect was corrected for,  there was no overall significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall = 2)`)
 
-numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
-raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
+**Figure S4: Individual and Combined comparison of Obese and Non-Obese groups for Firmicutes Relative Abundance.**  No study was individually signficantly different for obese and non-obese groups.
+When the data sets were combined and study effect was corrected for,  there was also no overall significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall = 2)`)
 
-dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
-k <- length(dat$StudyNo)
-dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
-dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
-dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
-dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
-
-#Used to generate the RE Model estimate based on all the studies
-res <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
-
-# B/F ratio data used to generate Funnel Plot
-StudyNo <- as.numeric(BFRatioRRTable$X)
-Study <- as.character(BFRatioRRTable$Study)
-Year <- c("2015", "2015", "2014", "2014", "2012", "2011", "2012", "2009")
-Total <- c("n=172", "n=63", "n=507", "n=30", "n=200", "n=256", "n=63", "n=146")
-tpos <- as.numeric(BFRatioRRTable$tposBF) #obese in low diversity group
-tneg <- as.numeric(BFRatioRRTable$tnegBF) #nonobese in low diversity group
-cpos <- as.numeric(BFRatioRRTable$cposBF) #obese in high diversity group
-cneg <- as.numeric(BFRatioRRTable$cnegBF) #nonobese in high diversity group
-RR <- as.numeric(BFRatioRRTable$RRBF)
-low <- as.numeric(BFRatioRRTable$lowBF)
-high <- as.numeric(BFRatioRRTable$highBF)
-
-numeric.data <-as.data.frame(cbind(tpos, tneg, cpos, cneg))
-raw.overall <- as.data.frame(cbind(StudyNo, Study,Year, Total, numeric.data))
-
-dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data=raw.overall, append=TRUE)
-k <- length(dat$StudyNo)
-dat.fm <- data.frame(study=factor(rep(1:k, each = 4)))
-dat.fm$grp <- factor(rep(c("T", "T", "C", "C"), k), levels = c("T", "C"))
-dat.fm$out <- factor(rep(c("+", "-", "+", "-"), k), levels = c("+", "-"))
-dat.fm$freq <- with(raw.overall, c(rbind(tpos, tneg, cpos, cneg)))
-
-#Used to generate the RE Model estimate based on all the studies
-res2 <- rma(ai=tpos, bi=tneg, ci=cpos, di=cneg, data=raw.overall, measure="RR", slab=paste(Study, Year, Total, sep=", "), method="REML")
-
-# Generate Funnel Plots for supplemental
-par(mfrow=c(1,2))
-funnel(res, main="A")
-funnel(res2, main="B")
-
-```
-
-**Figure S1: Funnel Plot of the Shannon Diversity and Bacteroidetes/Firmicutes Ratio Relative Risk.**  **A)** Overall there does not seem to be any bias associated with the studies selected for the Shannon Diversity analysis with studies falling on either side of the predicted value and those with smaller total n falling further away from this.
+**Figure S5: Funnel Plot of the Shannon Diversity and Bacteroidetes/Firmicutes Ratio Relative Risk.**  **A)** Overall there does not seem to be any bias associated with the studies selected for the Shannon Diversity analysis with studies falling on either side of the predicted value and those with smaller total n falling further away from this.
 **B)** For the Bacteroidetes/Firmicutes ratio there does not seem to be any associated bias.
@@
 
-```{r Figure_S2, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-source("code/MakeFiguresCodeSupplemental.R")
-library(gridExtra)
-
-grid.arrange(s2, sn2, nrow=2)
-
-```
-
-**Figure S2.
+**Figure S6.
 Summary of Power and Sample Size Simulations for OTU Richness (S) for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
@@
 
-```{r Figure_S3, echo=FALSE, message=FALSE, warning=FALSE, eval = FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-grid.arrange(j2, jn2, nrow=2)
-
-```
-
-**Figure S3.
+**Figure S7.
 Summary of Power and Sample Size Simulations for Evenness (J) for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
@@
 
-```{r Figure_S4, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-grid.arrange(b2, bn2, nrow=2)
-```
-
-**Figure S4.
+**Figure S8.
 Summary of Power and Sample Size Simulations for Bacteroidetes for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
@@
 
-```{r Figure_S5, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-grid.arrange(f2, fn2, nrow=2)
-```
-
-**Figure S5.
+**Figure S9.
 Summary of Power and Sample Size Simulations for Firmicutes for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
@@
 
-```{r Figure_S6, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-grid.arrange(RRH2, HRRn2, nrow=2)
-```
-
-**Figure S6.
+**Figure S10.
 Summary of Power and Sample Size Simulations for Relative Risk  for Obesity based on Shannon Diversity**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
@@
 
-```{r Figure_S7, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE,  fig.path="results/figures/", dev = c("pdf", "png")}
-
-grid.arrange(RRBF2, BFRRn2, nrow=2)
-```
-
-**Figure S7.
+**Figure S11.
 Summary of Power and Sample Size Simulations for Relative Risk  for Obesity based on B/F Ratio**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016_Supp.Rmd
update supplement search criteria

@@
 From this total we browsed abstracts for mention of stool or feces examination, that did not involve children, was not a clinical trial for probiotics or other diet related treatments, did not only have participants with inflammatory bowel disease, the articles were in English, did not only use PCR, qPCR, or RT-PCR only for their analysis, and sequencing that used only clone libraries.
-This ultimately excluded all but a total of 11 studies.
+This ultimately excluded all but a total of 13 studies.
 
-From this total of 12 studies the full text was reviewed for whether or not sequencing data was publicly available, BMI information (either categorical or continuous) was available in a supplement or, if it was not available, whether authors upon contact were willing to share this information or direct us to repositories that stored this specific information.
+From this total of 11 studies with the full text was reviewed for whether or not sequencing data was publicly available, BMI information (either categorical or continuous) was available in a supplement or, if it was not available, whether authors upon contact were willing to share this information or direct us to repositories that stored this specific information.
 One study was excluded [@yatsunenko_human_2012] because it contained children and their sequencing of the 16S rRNA gene involved amplicons of only 100bp in length.
 They also did not have obesity as part of their results in the actual published manuscript.
-The second study was excluded because they did not have BMI information available and when contacted the authors never returned any correspondence [@zeevi_personalized_2015].
-A third study was excluded since it did not use 16S rRNA gene sequencing for their bacterial microbiome analysis [@arumugam_enterotypes_2011]
+The second study was excluded because it did not use 16S rRNA gene sequencing for their bacterial microbiome analysis [@arumugam_enterotypes_2011].
 
-Once these 3 studies were excluded there was a total of 9 studies in the qualitative synthesis of the analysis.
-Because we decided a prioi to use the standard definition for BMI group classification one study from this ten did not have any individuals who were obese by this criteria [@nam_comparative_2011] and was excluded from the final quantitative synthesis and analysis.
+Once these 2 studies were excluded there was a total of 11 studies in the qualitative synthesis of the analysis.
+Because we decided a prioi to use the standard definition for BMI group classification one study from this 11 did not have any individuals who were obese by this criteria [@nam_comparative_2011] and was excluded from the final quantitative synthesis and analysis.
 

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016_Supp.Rmd
adjust so p-value show up

@@
 
-**Figure S1: Individual and Combined comparison of Obese and Non-Obese groups for Observed OTUs.**  For observed OTUs Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 2), nsmall = 2)`, Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 5), nsmall = 5)`), and Turnbaugh (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'turnbaugh' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 2), nsmall = 2)`) studies individually were signficantly different for obese and non-obese groups.
-When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 5), nsmall = 5)`)
+**Figure S1: Individual and Combined comparison of Obese and Non-Obese groups for Observed OTUs.**  For observed OTUs Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 2), nsmall = 2)`, Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 8), nsmall = 8)`), and Turnbaugh (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'turnbaugh' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 2), nsmall = 2)`) studies individually were signficantly different for obese and non-obese groups.
+When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 8), nsmall = 8)`)
 
 **Figure S2: Individual and Combined comparison of Obese and Non-Obese groups for Evenness.**  For evenness only the Goodrich study (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannoneven'), 'p.value'], 3), nsmall = 3)`)  individually was signficantly different for obese and non-obese groups.
-When the data sets were combined and study effect was corrected for,  there was an overall significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 2), nsmall = 2)`)
+When the data sets were combined and study effect was corrected for,  there was an overall significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall = 2)`)
 

### 2016-05-27 code/run_aucrf_genus.R
Fix testing output header

@@ -144,15 +144,12 @@ run <- function(datasets){
 
 				result <- c(train, test,
 									accuracy[2], accuracy[1],
-									accuracy[3], ci(roc_test), train_test_p)
+									accuracy[3], train_test_p)
 				testing_summary <- rbind(testing_summary, result)
 			} else {
-				result <- c(train, test, model[[train]]$opt_sensitivity,
-																	model[[train]]$opt_specificity,
-																	model[[train]]$opt_accuracy,
+				result <- c(train, test, model[[train]]$opt_accuracy,
 																	model[[train]]$opt_accuracy_lci,
-																	model[[train]]$opt_accuracy_uci,
-																	model[[train]]$auc, NA)
+																	model[[train]]$opt_accuracy_uci, NA)
 				testing_summary <- rbind(testing_summary, result)
 
 			}
@@ -161,7 +158,7 @@ run <- function(datasets){
 
 	colnames(testing_summary) <- c("train", "test",
 																	"accuracy", "accuracy_lci", "accuracy_uci",
-																	"auc_lci", "auc", "auc_hci", "p_value")
+																	"p_value")
 	rownames(testing_summary) <- NULL
 	write.table(testing_summary, file="data/process/random_forest.genus.train_test", quote=F, sep='\t', row.names=F)
 

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016.Rmd

@@
 First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value.
-We also used the random forest machine learning algorithm to identify community features that would allow us to to classify whether a person was obese based on either genus or OTU level.
+We investigated each study seperately as well as together while correcting for study effects.
 Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F.
-Finally, we utilized linear mixed-effect models to correct for variation by study and analyzed the pooled community diversity and B/F values within each dataset to test for an association with obesity.
+Finally, We also used the random forest machine learning algorithm to identify community features that would allow us to to classify whether a person was obese based on either genus or OTU level and assessed how well these variables did across study.
+Additionally, we also investigated power and esimated sample sizes needed to adequately test our observer effect sizes amongst the different study populations.
 
-**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals.
+**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese groups.
 A total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` diversity measurements  were found to be significantly different between non-obese and obese indivdiuals (P < 0.05) (Shannon Diversity (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannon"), ]))`), observed richness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "sobs"), ]))`), and evenness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannoneven"), ]))`)).
-With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals for both the genus (mean AUC: `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`) and OTU (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)`) level, and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05).
-When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`).
-When we used a linear mixed-effect model correcting for study there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) between non-obese and obese individuals.
+When we used a linear mixed-effect model correcting for study effects there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 8), nsmall=8)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) between non-obese and obese individuals.
+When we pooled the results across studies and looked at whether there was an increased RR of obesity we found that there was for Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`).
+Overall random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals for both the genus (mean AUC: `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`) and OTU (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)`) level, and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05).
 Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.
@@
 **Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome and its effect on obesity.
-Our meta-analysis provides support for the observation that there is a difference between non-obese and obese individuals in Shannon diversity but that the effect sizes are small, and that a very large number of indivdiuals would have to be tested to reach sufficient study power.
+Our meta-analysis provides support for the observation that there is a difference between non-obese and obese individuals in Shannon diversity but that the effect sizes is small, and that a very large number of indivdiuals would have to be tested to reach a sufficient study power.
 
@@
 We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010].
-A detailed description of our selection process and the exact search terms are provided in the Supplement [Figure 1].
+A detailed description of our selection process and the exact search terms are provided in the Supplement and in figure 1.
 Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome.
@@
 In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, observed richness, and evenness.
-A two-tailed Wilcoxon test was performed for comparison for non-obese and obese individuals.
+After each variable was transformed to fit a normal distribution A two-tailed t-test was performed for comparison for non-obese and obese individuals.
+Additionaly, for this strategy, we performed a pooled analysis on these measured variables using linear mixed effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups.
 Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices.
 This analysis was performed using the vegan R package.
-Finally, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguishing non-obese and obese individuals.
-To generate Receiver Operator Characteristic (ROC) curves we used the pROC (v1.8) R package.
-A Bonferroni-adjusted p-value was used to assess differences between each study.
 
@@
 
-In the third strategy, we corrected for study effects and used linear mixed-effect models to test the pooled diversity and B/F ratio data for a difference between obese and non-obese groups.
+In the third strategy, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguishing non-obese and obese individuals.
+To generate Receiver Operator Characteristic (ROC) curves we used the pROC (v1.8) R package.
+A Bonferroni-adjusted p-value was used to assess differences between each study.
+
 Additionally, we also performed power and sample number simulations for different effect sizes for each study.
@@
 
-***Separate Analysis***
+***Separate and Pooled Analysis***
 We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05), ]))` significant results; two for Shannon diversity, three for observed richness, one for evenness, and one for Bacteroidetes [Table S1].
-There were no significant results for Firmicutes or the B/F ratio.
-The full summary of the exact P-values for each respective study and variable tested can be found in table S1 of the online supplement.
-
-Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.
-We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)` for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.
-We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al.
-study) to `r max(rand_forest.genus.summary$k_opt)` (Goodrich, et al.
-Study).
-For the OTU level the number of variables ranged from `r min(rand_forest.otu.summary$k_opt)` (Wu, et al.
-study) to `r max(rand_forest.otu.summary$k_opt)` (Goodrich, et al.
-study).
-The total number of trees and node size was standardized to 500 and 20 respectively for each data set.
-Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05, Figure 2).
+Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05), ]))` significant results; two for Shannon diversity [Figure 2A], three for observed richness [Figure S1], one for evenness [Figure S2], and one for Bacteroidetes [Figure S3].
+There were no significant results for Firmicutes [Figure S4] or the B/F ratio [Figure 2B].
+Using the linear mixed-effect model to analyze the data pooled together while also controling study effects we found that for Shannon diversity  there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 2A].
+Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`) [Figure S1].
+However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`)  [Figure 2B & S2-S4] using this approach, between obese and non-obese individuals.
 
@@
 We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis.
-From the graph we can observe that there does not seem to be any significant  bias [Figure S1].
+From the graph we can observe that there does not seem to be any significant  bias [Figure S5].
 Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.
 
-***Linear Mixed-Effect models for Pooled Analysis***
+***Random Forest Classification and Predicition Accuracy***
 
-Using the third approach which utilizes a linear mixed-effect model to control for the effect of study on the Shannon diversity  we found that there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 5A].
-Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`).
-However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) using this approach, between obese and non-obese individuals [Figure 5B].
+Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.
+We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)` for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.
+We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al.
+study) to `r max(rand_forest.genus.summary$k_opt)` (Goodrich, et al.
+Study).
+For the OTU level the number of variables ranged from `r min(rand_forest.otu.summary$k_opt)` (Wu, et al.
+study) to `r max(rand_forest.otu.summary$k_opt)` (Goodrich, et al.
+study).
+The total number of trees and node size was standardized to 500 and 20 respectively for each data set.
+Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05) regardless if it was using genus or OTU level classification [Figure 4].
+Overall most studies performed poorly when used on other data sets for classification [Figure 5] with no significant difference between studies for their mean accuracy for prediction of non-obese and obese indivdiuals (P-value > 0.05).
 
@@
 
-For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 5A] compared to the B/F ratio [Figure 6A] differences.
-When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity (predicted percent power = `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`) [Figure 5A].
+For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 6A] compared to the B/F ratio [Figure 7A] differences.
+When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity (predicted percent power = `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`) [Figure 6A].
 This predicted power was also similar for evenness (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`).
-However, this was not the case for the B/F ratio (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)` +/- `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`%) [Figure 6A] or the other measures that were tested (observed richness, Bacteroidetes, Firmicutes) where even a 15% difference between non-obese and obese groups only resulted in a relatively poor average power between studies.
-Looking at the sample sizes showed that on average to detect a 5% difference in Shannon diversity at 80% power between non-obese and obese individuals a total n of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) would be needed for each group [Figure 5B].
-In contrast in order to detect a 5% difference with 80% power in the B/F ratio a sample size of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) per group [Figure 6B].
-A complete breakdown of the power simulations and sample size estimates for the other variables investigated and RR can be found in the online supplement [Figure S2-S7].
+However, this was not the case for the B/F ratio (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)` +/- `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`%) [Figure 7A] or the other measures that were tested (observed richness, Bacteroidetes, Firmicutes) where even a 15% difference between non-obese and obese groups only resulted in a relatively poor average power between studies.
+Looking at the sample sizes showed that on average to detect a 5% difference in Shannon diversity at 80% power between non-obese and obese individuals a total n of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) would be needed for each group [Figure 6B].
+In contrast in order to detect a 5% difference with 80% power in the B/F ratio a sample size of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) per group [Figure 7B].
+A complete breakdown of the power simulations and sample size estimates for the other variables investigated and RR can be found in the online supplement [Figure S6-S11].
 

### 2016-05-31 submission/Sze_Obesity_PeerJ_2016.Rmd

@@
 title: "Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome"
+bibliography: references.bib
 output:
-  html_document:
-    keep_md: yes
-bibliography: reference.bib
+  pdf_document:
+    keep_tex: true
+    includes:
+      in_header: header.tex
+csl: msystems.csl
+fontsize: 11pt
+geometry: margin=1.0in
 ---
 
-Marc A Sze<sup>1</sup> and Patrick D Schloss<sup>1</sup>
+```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
+opts_chunk$set("tidy" = TRUE)
+opts_chunk$set("echo" = FALSE)
+opts_chunk$set("eval" = TRUE)
+opts_chunk$set("warning" = FALSE)
+opts_chunk$set("cache" = FALSE)
 
-<sup>1</sup> Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI, USA  
-<br>
-<br>
-<br>
-<br>
-<br>
-<br>
-Corresponding author: Patrick D Schloss  
-*Current Address:*  
-1520A Medical Science Research Building I
-1500 W. Medical Center Dr.
-Ann Arbor, MI  
-USA 48109-5620  
-Email: pschloss@umich.edu  
-Phone: 734-647-5801    
-<br>
-<br>
-**Contributions:** MAS (planned, designed, and performed experiments, completed data analysis, wrote first draft), PDS (planned and designed the study, data interpretation, and edited subsequent drafts).
-<br>
-**Running Title:** microbiome and Obesity
-<br>
-
-```{r Loading_data_tables, echo=FALSE, message=FALSE, warning=FALSE}
-
-alpha_tests.summary <- read.table("../data/process/alpha_tests.summary", header=T)
-alpha_composite.summary <- read.table("../data/process/alpha_composite.summary")
-alpha_power.obs <- read.table("../data/process/alpha_power.observed", 
-                              header = T, row.names = 1)
-alpha_power.predicted <- read.table("../data/process/alpha_power.predicted", 
-                                    header = T)
-alpha.data.ALL <- read.table("../data/process/alpha.data", header = T)
-
-beta_tests.summary <- read.table("../data/process/beta_tests.summary", 
-                                 header=T, row.names = 1)
-rand_forest.genus.roc_data <- read.table("../data/process/random_forest.genus.roc_data", header=T)
-
-rand_forest.genus.summary <- read.table("../data/process/random_forest.genus.summary", 
-                                        header = T, row.names = 1)
-
-rand_forest.genus.train_test <- 
-  read.table("../data/process/random_forest.genus.train_test", header = T)
-
-rand_forest.otu.roc_data <- 
-  read.table("../data/process/random_forest.otu.roc_data", header = T)
-
-rand_forest.otu.summary <- 
-  read.table("../data/process/random_forest.otu.summary", 
-             header = T, row.names = 1)
-
-RR_composite <- read.table("../data/process/relative_risk.composite", header = T)
-RR_summary <- read.table("../data/process/relative_risk.summary", header = T)
-RR_pow.obs <- read.table("../data/process/rr_power.observed", 
-                         header = T, row.names = 1)
-RR_pow.pred <- read.table("../data/process/rr_power.predicted", header = T)
-
-shannond <- (mean(alpha.data.ALL[which(alpha.data.ALL$obese == FALSE), ]$shannon)-mean(alpha.data.ALL[which(alpha.data.ALL$obese == TRUE), ]$shannon))/sd(alpha.data.ALL$shannon)
-
-bf_ratiod <- (mean(alpha.data.ALL[which(alpha.data.ALL$obese == FALSE), ]$bf_ratio)-mean(alpha.data.ALL[which(alpha.data.ALL$obese == TRUE), ]$bf_ratio))/sd(alpha.data.ALL$bf_ratio)
-
-#Using a quick for loop to generate the necessary effect size information
-variableTested <- seq(1:6)
-variables <- colnames(alpha_power.obs)[1:6]
-alpha_effectSize <- matrix(nrow=length(rownames(beta_tests.summary)), ncol=length(variableTested))
-studies <- rownames(beta_tests.summary)
-
-for(i in variableTested){
-  tempdata <- c()
-  for(j in studies){
-  tempdata <- rbind(tempdata, abs((mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == FALSE), variables[i]]) - mean(alpha.data.ALL[which(alpha.data.ALL$dataset == j & alpha.data.ALL$obese == TRUE), variables[i]]))/mean(alpha.data.ALL[, variables[i]]) * 100))
-  }
-  alpha_effectSize[, i] <- tempdata
+inline_hook <- function(x){
+	print(x)
+	if(is.numeric(x)){
+		if(abs(x - round(x)) < .Machine$double.eps^0.5){
+			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
+		} else {
+			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
 		}
 
-colnames(alpha_effectSize) <- variables
-rownames(alpha_effectSize) <- studies
+	} else {
+    	paste(x)      
+	}
+}
+knitr::knit_hooks$set(inline=inline_hook)
+```
+\begin{center}
+\vspace{25mm}
+
+**Running Title:** The Human Microbiome and Obesity
+<br>
+Marc A Sze and Patrick D Schloss${^\dagger}$
+<br>
+**Contributions:** Both authors contributed to the planning, design, execution, interpretation, and writing of the analyses.
+
+\vspace{30mm}
+
+$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu
+
+Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+\end{center}
+
+\newpage
+\linenumbers
+
+```{r datasets}
+beta <- read.table(file="data/process/beta_tests.summary", header=T)
+genus <- read.table(file="data/process/random_forest.genus.summary", header=T)
+otu <- read.table(file="data/process/random_forest.otu.summary", header=T)
+test <- read.table(file="data/process/random_forest.genus.train_test", header=T)
+test_summary <- aggregate(accuracy~train, data=test, median)
 
@@
 
-```{r install_needed_packages, echo=FALSE, message=FALSE, warning=FALSE}
-deps = c("knitr","vegan","epiR","AUCRF","metafor","ggplot2", "plyr", "reshape2", "RColorBrewer", "scales", "gridExtra", "pwr", "statmod");
-for (dep in deps){
-  if (dep %in% installed.packages()[,"Package"] == FALSE){
-    install.packages(as.character(dep), quiet=TRUE);
-  }
-  library(dep, verbose=FALSE, character.only=TRUE)
-}
+## Abstract - this could be significantly shorter if we don't do the structured format.
 
-```
-
-*******
-## Abstract
-**Rationale:** The hypothesis that changes in the microbiome are associated with obesity has been widely accepted.
-However, two recent studies re-analyzed published data and found that when the datasets were analyzed independently they could not support the hypothesis.
+**Rationale:** The hypothesis that changes in the microbiome are associated with obesity has been widely accepted; however, two recent studies re-analyzed published data and found that when the datasets were analyzed independently they could not support the hypothesis.
 In this meta analysis, we retested the hypothesis by increasing the number of data sets examined and investigating the effect of pooling the results across the individual datasets.
 
-**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 10 studies that fit our criteria.
-We applied three analysis pipelines to the data from each study.
-First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value.
-We investigated each study seperately as well as together while correcting for study effects.
-Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F.
-Finally, We also used the random forest machine learning algorithm to identify community features that would allow us to to classify whether a person was obese based on either genus or OTU level and assessed how well these variables did across study.
-Additionally, we also investigated power and esimated sample sizes needed to adequately test our observer effect sizes amongst the different study populations.
+**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis or had a large number of samples to 10 studies that fit our criteria.
+We applied four analysis pipelines to the data from each study to determine whether there was a significant association between variation in the gut microbiome with obesity status.
+First, we tested for an association with alpha-diversity metrics.
+Second, we calculated the relative risk (RR) of being obese based on alpha-diversity metrics.
+Third, we tested for a community-wide association with obesity based on Bray-Curtis distances between samples.
+Finally, we used the random forest machine learning algorithm to identify community features that would allow us to to classify whether a person was obese based on either genus or OTU level and assessed how well these variables did across other studies.
+Additionally, we also investigated power and estimated sample sizes needed to adequately test our observer effect sizes amongst the different study populations.
 
-**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese groups.
-A total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` diversity measurements  were found to be significantly different between non-obese and obese indivdiuals (P < 0.05) (Shannon Diversity (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannon"), ]))`), observed richness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "sobs"), ]))`), and evenness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannoneven"), ]))`)).
-When we used a linear mixed-effect model correcting for study effects there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 8), nsmall=8)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) between non-obese and obese individuals.
-When we pooled the results across studies and looked at whether there was an increased RR of obesity we found that there was for Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`).
-Overall random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals for both the genus (mean AUC: `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`) and OTU (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)`) level, and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05).
-Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.
+**Results:** When considered in isolation, there were few significant associations between alpha diversity metrics and obesity when we used the observed diversity values or relative risk values; however, when we combined the data into random effects models we observed significant differences in the Shannon diversity, number of observed OTUs, and Shannon evenness.
+Although these tests yielded small P-values, the general effect sizes were quite small.
+At the community level, we observed significant differences in the community structure of obese and non-obese individuals in `sum(beta$p_value<0.05)` of the 10 studies.
+The ability of Random Forest machine learning models to classify obese and non-obese was limited.
+When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]` (median=`r median(test_summary$accuracy)`).
+Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect modest-sized differences in any of the measurements used.
 
-**Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome and its effect on obesity.
-Our meta-analysis provides support for the observation that there is a difference between non-obese and obese individuals in Shannon diversity but that the effect sizes is small, and that a very large number of indivdiuals would have to be tested to reach a sufficient study power.
+**Conclusion:** Synthesizing data from 10 microbiome studies, we find statistical support for a relationship between the microbial communities found in human feces and their obesity status; however, this association is relatively weak.
+Furthermore, this analysis demonstrates that most studies that analyze human feces are underpowered for detecting modest differences in diversity.
 
@@
 
-Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014].
+Obesity is a growing health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014].
 This number increases to approximately 35% in adults (aged 20 or older) and these statistics have seen little change since 2003 [@ogden_prevalence_2014].
@@
 
-There has been a continuous stream of studies that report observing a link between the composition of microbiome and obesity.
-The first such study used genetically obese mice and observed that obese mice had a lower Bacteroidetes and higher Firmicutes than lean mice (lower B/F ratio associated with obesity) [@ley_obesity_2005].
-Translation of this result to humans by the same researchers did not observe this effect, but did found that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009].
-Since the publication of these studies there have been a steady stream of publications on the microbiome with associated BMI data but without clear explicit mention of any significant correlations between the microbiome and obesity [@ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @escobar_gut_2014].
-Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B/F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
+There have been several studies that report observing a link between the composition of microbiome and obesity in animal models and in humans.
+The first such study used genetically obese mice and observed the ratio of the relative abundances of Bacteroidetes to Firmicutes (B:F) was lower in obese mice than lean mice [@ley_obesity_2005].
+Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009].
+Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
+Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since different communities yield variation in animal weight [REFERENCES].
+The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.
 
-One critique of these re-analysis studies is that they did not aggregate the results across studies to increase the effective sample size.
-It is possible that there are small associations within each study that would not be significant because the studies lacked sufficient statistical power.
+The recent publication of additional studies that collected BMI data for each subject as well as other studies that were not included in the earlier re-analysis studies offered the opportunity to revisit the question relating the structure of the human microbiome to obesity [@ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @escobar_gut_2014; @baxter_microbiota-based_2016; @schubert_microbiome_2014].
+One critique of the prior re-analysis studies is that the authors did not aggregate the results across studies to increase the effective sample size.
+It is possible that there were small associations within each study that were not statistically significant because the studies lacked sufficient power.
+Alternatively, diversity metrics may mask the appropriate signal and it is necessary to measure the association at the level of microbial populations.
+Walters et al.
+[-@walters_meta-analyses_2014] demonstrated that Random Forest machine learning models were capable of predicting obesity status within a single cohort, but did not attempt to test the models on other cohorts.
 The purpose of this study was to perform a meta-analysis of the association between differences in the microbiome and obesity status by analyzing and applying a more systematic and synthetic approach than was used previously.
-Furthermore, we sought to provide additional clarity on the possible relationship between community diversity, B/F, and community composition and obesity status and the ability to detect significant differences in these variables if they were actually present.
 
@@
 We processed the 16S rRNA gene sequence data using a standardized mothur pipeline.
-For a few studies author defined quality scores were used to maximize the number of samples for each study.
+Briefly, our pipelines attempted to follow previously recommended approaches for 454 and Illumina sequencing data [@Kozich_2013; @Schloss2011].
+All sequences were screened for chimeras using UCHIME and assigned to operational taxonomic units using the average neighbor algorithm using a 3% distance threshold [@Edgar2011; @Westcott2015].
+All sequence processing was performed using mothur (v.1.37.0) [@mothur].
 A detailed and reproducible description of how the data were processed can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/.
@@
 ***Data Analysis***
-We performed the analysis using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`).
-We split the overall analysis into three general strategies.
-In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, observed richness, and evenness.
-After each variable was transformed to fit a normal distribution A two-tailed t-test was performed for comparison for non-obese and obese individuals.
-Additionaly, for this strategy, we performed a pooled analysis on these measured variables using linear mixed effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups.
+We split the overall meta-analysis into three general strategies.
+First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B:F), Shannon diversity, observed richness, and evenness.
+After each variable was transformed to fit a normal distribution a two-tailed t-test was performed for comparison for non-obese and obese individuals.
+We performed a pooled analysis on these measured variables using linear random effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups using the lme4 (v.`r packageDescription("lme4")$Version`) R package.
 Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices.
-This analysis was performed using the vegan R package.
-
-In the second strategy, we pooled the relative risk (RR) values to assess whether there is an increased risk of obesity that is detectable based on specific microbiome metrics.
-For each study we partitioned the subjects into a low or high group depending on whether their diversity or B/F was below or above the median value for the study.
+This analysis was performed using the vegan (v.`r packageDescription("vegan")$Version`) R package.
+Second, we pooled the relative risk (RR) values to assess whether there is an increased risk of obesity that is detectable based on specific microbiome metrics.
+For each study we partitioned the subjects into a low or high group depending on whether their diversity or B:F was below or above the median value for the study.
 The RR was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group.
-We then performed a Fisher exact-test to investigate whether the RR was significantly different from 1.0 using the epiR (v0.9-6.9) and metafor (v1.9-8) packages.
-Finally, we assessed the potential publication bias in the diversity and B/F analyses using a funnel plot.
-
-In the third strategy, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguishing non-obese and obese individuals.
-To generate Receiver Operator Characteristic (ROC) curves we used the pROC (v1.8) R package.
-A Bonferroni-adjusted p-value was used to assess differences between each study.
-
-Additionally, we also performed power and sample number simulations for different effect sizes for each study.
+We then performed a Fisher exact-test to investigate whether the RR was significantly different from 1.0 using the epiR (`r packageDescription("epiR")$Version`) and metafor (`r packageDescription("metafor")$Version`) packages.
+Third, we used the AUCRF (`r packageDescription("AUCRF")$Version`) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguishing non-obese and obese individuals.
+To generate Receiver Operator Characteristic (ROC) curves we used the pROC (`r packageDescription("pROC")$Version`) R package.
+Finally, we performed power and sample number simulations for different effect sizes for each study.
 This was completed for all the different components measured as part of first strategy.
 We also calculated the actual sample size needed based on the effect size of each individual study.
-To complete this analysis we used a combination of the pwr (v1.1-3) R package and base R functions.
+To complete this analysis we used the pwr (`r packageDescription("pwr")$Version`) R package and base R functions.
+The analysis was performed using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`).
 
@@
 ***Separate and Pooled Analysis***
-We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
+We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B:F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
 Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05), ]))` significant results; two for Shannon diversity [Figure 2A], three for observed richness [Figure S1], one for evenness [Figure S2], and one for Bacteroidetes [Figure S3].
-There were no significant results for Firmicutes [Figure S4] or the B/F ratio [Figure 2B].
+There were no significant results for Firmicutes [Figure S4] or the B:F ratio [Figure 2B].
 Using the linear mixed-effect model to analyze the data pooled together while also controling study effects we found that for Shannon diversity  there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 2A].
 Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`) [Figure S1].
-However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`)  [Figure 2B & S2-S4] using this approach, between obese and non-obese individuals.
+However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B:F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`)  [Figure 2B & S2-S4] using this approach, between obese and non-obese individuals.
 
@@
 The pooled analysis showed an increased RR of `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['shannon', 'ci_lb'],2), nsmall=2)` - `r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was significant (P-value = `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`).
-In contrast, when we investigated the B/F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`) [Figure 3B].
+In contrast, when we investigated the B:F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`) [Figure 3B].
 The pooled RR was `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)` - `r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`).
@@
 
-For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 6A] compared to the B/F ratio [Figure 7A] differences.
+For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 6A] compared to the B:F ratio [Figure 7A] differences.
 When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity (predicted percent power = `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`) [Figure 6A].
 This predicted power was also similar for evenness (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`).
-However, this was not the case for the B/F ratio (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)` +/- `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`%) [Figure 7A] or the other measures that were tested (observed richness, Bacteroidetes, Firmicutes) where even a 15% difference between non-obese and obese groups only resulted in a relatively poor average power between studies.
+However, this was not the case for the B:F ratio (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)` +/- `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`%) [Figure 7A] or the other measures that were tested (observed richness, Bacteroidetes, Firmicutes) where even a 15% difference between non-obese and obese groups only resulted in a relatively poor average power between studies.
 Looking at the sample sizes showed that on average to detect a 5% difference in Shannon diversity at 80% power between non-obese and obese individuals a total n of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) would be needed for each group [Figure 6B].
-In contrast in order to detect a 5% difference with 80% power in the B/F ratio a sample size of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) per group [Figure 7B].
+In contrast in order to detect a 5% difference with 80% power in the B:F ratio a sample size of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) per group [Figure 7B].
 A complete breakdown of the power simulations and sample size estimates for the other variables investigated and RR can be found in the online supplement [Figure S6-S11].
@@
 The average actual cohens d value for Shannon Diversity for all `r length(rownames(beta_tests.summary))` studies of `r format(round(shannond, 2), nsmall=2)` gave a power of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=shannond, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
-Finally, the average actual cohens d value for B/F ratio for all `r length(rownames(beta_tests.summary))` studies of `r format(round(bf_ratiod, 2), nsmall=2)` yieled a power of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=bf_ratiod, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
+Finally, the average actual cohens d value for B:F ratio for all `r length(rownames(beta_tests.summary))` studies of `r format(round(bf_ratiod, 2), nsmall=2)` yieled a power of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=bf_ratiod, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
 
@@
 Further, even though random forest could identify specific variables with similar taxonomic classifications that resulted in good ROC curves [Figure 1] these variables varied widely by both their actual sequence and directionality.
-Similarily, for the B/F ratio we found that the other two approaches, that used the pooled data approach, were also not able to identify a correlation between measures of the microbiome and obesity [Figure 3 & 4].
-For these approaches both the B/F ratio failed to have a significant correlation to obesity.
+Similarily, for the B:F ratio we found that the other two approaches, that used the pooled data approach, were also not able to identify a correlation between measures of the microbiome and obesity [Figure 3 & 4].
+For these approaches both the B:F ratio failed to have a significant correlation to obesity.
 However, a small but significant differences was observed between non-obese and obese individauls  for Shannon diversity [Figure 3 & 4].
@@
 Although the effect sizes for the BF ratio were larger than that for Shannon Diversity [Figure 5A & 6A] the estimated power remained low.
-Additionaly, we would also need a relatively large sample size to be able to detect the actual differences in each study for both Shannon Diversity and B/F ratio [Figure 5B & 6B].
-Taken together we believe that this analysis builds on the previous two re-anlaysis and provides additional information as to whether there is a difference between non-obese and obese human individuals with respect to both Shannon diversity and the B/F ratio.
+Additionaly, we would also need a relatively large sample size to be able to detect the actual differences in each study for both Shannon Diversity and B:F ratio [Figure 5B & 6B].
+Taken together we believe that this analysis builds on the previous two re-anlaysis and provides additional information as to whether there is a difference between non-obese and obese human individuals with respect to both Shannon diversity and the B:F ratio.
 
@@
 
-**Figure 2: Individual and Combined comparison of Obese and Non-Obese groups for Shannon Diversity and B/F Ratio.** **A)** For Shannon diversity only Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 2), nsmall = 2)` and Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 5), nsmall = 5)`) studies indivdiually were signficantly different for obese and non-obese groups.
+**Figure 2: Individual and Combined comparison of Obese and Non-Obese groups for Shannon Diversity and B:F Ratio.** **A)** For Shannon diversity only Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 2), nsmall = 2)` and Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 5), nsmall = 5)`) studies indivdiually were signficantly different for obese and non-obese groups.
 When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 5), nsmall = 5)`).
-**B)** For the B/F ratio there was no individual study that was significant nor was their a difference when the data was combined and study effect corrected for between Non-Obese and Obese groups (P > 0.05).
+**B)** For the B:F ratio there was no individual study that was significant nor was their a difference when the data was combined and study effect corrected for between Non-Obese and Obese groups (P > 0.05).
 
-**Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity or Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low Shannon Diversity or Bacteroidetes/Firmicutes (B/F) ratio groups based on the median for that study.
+**Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity or Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low Shannon Diversity or Bacteroidetes/Firmicutes (B:F) ratio groups based on the median for that study.
 Red represents studies that used the 16S rRNA V4 region, yellow represents studies that used a V1-V3 approach, blue represents studies that used either a V1-V2 or just a V2 approach, green represents studies that used a V3-V4 region, and orange represents studies that used the V3-V5 region.
 **A)** The overall pooled relative risk for Shannon Diversity was `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` with a P-value of `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`.
-**B)** The overall pooled relative risk for the B/F ratio was  `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` with a P-value of `r format(round(RR_composite['bf_ratio', 'p_value'], 2), nsmall=2)`.
+**B)** The overall pooled relative risk for the B:F ratio was  `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` with a P-value of `r format(round(RR_composite['bf_ratio', 'p_value'], 2), nsmall=2)`.
 

### 2016-05-31 code/run_alpha_diversity.R
Update alpha composite statistics

@@ -101,8 +101,12 @@ run <- function(datasets){
 
 	write.table(summary_data, file="data/process/alpha_tests.summary", quote=F, sep='\t', row.names=F)
 
-#	composite_analysis <- function(alpha, dataset, is_obese, pow){
-	composite_p <- NULL
+	non_mean <- aggregate(mean_non~metric, data=summary_data, mean)
+	obese_mean <- aggregate(mean_obese~metric, data=summary_data, mean)
+
+	composite_p <- rep(NA, nrow(non_mean))
+	names(composite_p) <- non_mean$metric
+
 	composite_p["shannon"] <- composite_analysis(composite_data$shannon, composite_data$dataset, composite_data$obese, 2)
 	composite_p["sobs"] <- composite_analysis(composite_data$sobs, composite_data$dataset, composite_data$obese, 0.5)
 	composite_p["shannoneven"] <- composite_analysis(composite_data$shannoneven, composite_data$dataset, composite_data$obese, 4)
@@ -110,5 +114,5 @@ run <- function(datasets){
 	composite_p["firmicutes"] <- composite_analysis(composite_data$firmicutes, composite_data$dataset, composite_data$obese, 1)
 	composite_p["bf_ratio"] <- composite_analysis(composite_data$bf_ratio, composite_data$dataset, composite_data$obese, 0.5)
 
-	write.table(file="data/process/alpha_composite.summary", data.frame(composite_p), quote=F, sep='\t')
+	write.table(file="data/process/alpha_composite.summary", data.frame(non_mean$metric, non_mean$mean_non, obese_mean$mean_obese, composite_p), quote=F, sep='\t', row.names=F)
 }

### 2016-05-31 code/run_alpha_diversity.R
Update alpha composite statistics

@@ -114,5 +114,5 @@ run <- function(datasets){
 	composite_p["firmicutes"] <- composite_analysis(composite_data$firmicutes, composite_data$dataset, composite_data$obese, 1)
 	composite_p["bf_ratio"] <- composite_analysis(composite_data$bf_ratio, composite_data$dataset, composite_data$obese, 0.5)
 
-	write.table(file="data/process/alpha_composite.summary", data.frame(non_mean$metric, non_mean$mean_non, obese_mean$mean_obese, composite_p), quote=F, sep='\t', row.names=F)
+	write.table(file="data/process/alpha_composite.summary", data.frame(metric=non_mean$metric, non_obese=non_mean$mean_non, obese=obese_mean$mean_obese, p = composite_p), quote=F, sep='\t', row.names=F)
 }

### 2016-05-31 submission/Sze_Obesity_PeerJ_2016.Rmd
Refactor Abstract and Importance sections

@@
 ---
+
+# Lack of power and small effect size confounds the ability to differentiate obese and non-obese individuals using gut microbiome data
 
@@
 ```{r datasets}
+alpha <- read.table(file="data/process/alpha_composite.summary", header=T)
+shannon_effect_size <- 100*abs(1-alpha[alpha$metric=="shannon","obese"]/alpha[alpha$metric=="shannon","non_obese"])
 beta <- read.table(file="data/process/beta_tests.summary", header=T)
+
 genus <- read.table(file="data/process/random_forest.genus.summary", header=T)
 otu <- read.table(file="data/process/random_forest.otu.summary", header=T)
+
 test <- read.table(file="data/process/random_forest.genus.train_test", header=T)
@@
 
+power <- read.table(file="data/process/alpha_power.predicted", header=T)
+well_powered <- sum(power[power$metric=="shannon" &
+ 												power$effect_size==0.05, "power"] > 0.8)
+
+number_string <- c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine")
 ```
 
-## Abstract - this could be significantly shorter if we don't do the structured format.
+## Abstract
+Two recent studies have re-analyzed published data and found that when datasets are analyzed independently there was limited support the widely accepted hypothesis that changes in the microbiome are associated with obesity.
+This hypothesis was reconsidered by increasing the number of data sets and pooling the results across the individual datasets.
+The Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines were applied to identify 10 studies for an updated and more synthetic analysis.
+Alpha-diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances.
+Although these tests yielded small P-values, the difference between the Shannon diversity index of obese and non-obese individuals was `r shannon_effect_size`%.
+A power analysis demonstrated that `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity.
+When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]` (median=`r median(test_summary$accuracy)`).
+Although there is statistical support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.
 
-**Rationale:** The hypothesis that changes in the microbiome are associated with obesity has been widely accepted; however, two recent studies re-analyzed published data and found that when the datasets were analyzed independently they could not support the hypothesis.
-In this meta analysis, we retested the hypothesis by increasing the number of data sets examined and investigating the effect of pooling the results across the individual datasets.
-
-**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis or had a large number of samples to 10 studies that fit our criteria.
-We applied four analysis pipelines to the data from each study to determine whether there was a significant association between variation in the gut microbiome with obesity status.
-First, we tested for an association with alpha-diversity metrics.
-Second, we calculated the relative risk (RR) of being obese based on alpha-diversity metrics.
-Third, we tested for a community-wide association with obesity based on Bray-Curtis distances between samples.
-Finally, we used the random forest machine learning algorithm to identify community features that would allow us to to classify whether a person was obese based on either genus or OTU level and assessed how well these variables did across other studies.
-Additionally, we also investigated power and estimated sample sizes needed to adequately test our observer effect sizes amongst the different study populations.
-
-**Results:** When considered in isolation, there were few significant associations between alpha diversity metrics and obesity when we used the observed diversity values or relative risk values; however, when we combined the data into random effects models we observed significant differences in the Shannon diversity, number of observed OTUs, and Shannon evenness.
-Although these tests yielded small P-values, the general effect sizes were quite small.
-At the community level, we observed significant differences in the community structure of obese and non-obese individuals in `sum(beta$p_value<0.05)` of the 10 studies.
-The ability of Random Forest machine learning models to classify obese and non-obese was limited.
-When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]` (median=`r median(test_summary$accuracy)`).
-Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect modest-sized differences in any of the measurements used.
-
-**Conclusion:** Synthesizing data from 10 microbiome studies, we find statistical support for a relationship between the microbial communities found in human feces and their obesity status; however, this association is relatively weak.
-Furthermore, this analysis demonstrates that most studies that analyze human feces are underpowered for detecting modest differences in diversity.
+## Importance
+As interest in the human microbiome grows there is an increasing number of studies that can be used to test numerous hypotheses across human populations.
+The hypothesis that variation in the gut microbiota can explain or be used to predict obesity status has received considerable attention and is frequently mentioned as an example for the role of the microbiome in human health.
+Here we assess this hypothesis using ten independent studies and find that although there is an association, it is smaller than can be detected by most microbiome studies.
+Furthermore, we directly tested the ability to predict obesity status based on the composition of an individual's microbiome and find that the median classification accuracy is between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]`%.
+This type of analysis can be used to design future studies and expanded to explore other hypotheses.
 

### 2016-05-31 submission/Sze_Obesity_PeerJ_2016.Rmd
Refactor Introduction and Methods sections

@@
 Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009].
+They also showed that the relative abundance of Bacteroidetes and Firmicutes increased and decreased, respectively, as obese individuals lost weight while on a fat or carbohydrate restricted diet [@ley_microbial_2006].
 Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
@@
 One critique of the prior re-analysis studies is that the authors did not aggregate the results across studies to increase the effective sample size.
-It is possible that there were small associations within each study that were not statistically significant because the studies lacked sufficient power.
+It is possible that there were small associations within each study that were not statistically significant because the individual studies lacked sufficient power.
 Alternatively, diversity metrics may mask the appropriate signal and it is necessary to measure the association at the level of microbial populations.
@@
 Briefly, our pipelines attempted to follow previously recommended approaches for 454 and Illumina sequencing data [@Kozich_2013; @Schloss2011].
-All sequences were screened for chimeras using UCHIME and assigned to operational taxonomic units using the average neighbor algorithm using a 3% distance threshold [@Edgar2011; @Westcott2015].
+All sequences were screened for chimeras using UCHIME and assigned to operational taxonomic units (OTUs) using the average neighbor algorithm using a 3% distance threshold [@Edgar2011; @Westcott2015].
 All sequence processing was performed using mothur (v.1.37.0) [@mothur].
-A detailed and reproducible description of how the data were processed can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/.
 
 ***Data Analysis***
-We split the overall meta-analysis into three general strategies.
-First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B:F), Shannon diversity, observed richness, and evenness.
-After each variable was transformed to fit a normal distribution a two-tailed t-test was performed for comparison for non-obese and obese individuals.
+We split the overall meta-analysis into three general strategies using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`).
+First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B:F), Shannon diversity, observed richness, and Shannon evenness.
+After each variable was transformed to fit a normal distribution a two-tailed t-test was performed for comparison of non-obese and obese individuals (i.e.
+BMI > 35.0).
 We performed a pooled analysis on these measured variables using linear random effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups using the lme4 (v.`r packageDescription("lme4")$Version`) R package.
@@
 This analysis was performed using the vegan (v.`r packageDescription("vegan")$Version`) R package.
-Second, we pooled the relative risk (RR) values to assess whether there is an increased risk of obesity that is detectable based on specific microbiome metrics.
-For each study we partitioned the subjects into a low or high group depending on whether their diversity or B:F was below or above the median value for the study.
-The RR was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group.
-We then performed a Fisher exact-test to investigate whether the RR was significantly different from 1.0 using the epiR (`r packageDescription("epiR")$Version`) and metafor (`r packageDescription("metafor")$Version`) packages.
-Third, we used the AUCRF (`r packageDescription("AUCRF")$Version`) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguishing non-obese and obese individuals.
-To generate Receiver Operator Characteristic (ROC) curves we used the pROC (`r packageDescription("pROC")$Version`) R package.
-Finally, we performed power and sample number simulations for different effect sizes for each study.
-This was completed for all the different components measured as part of first strategy.
+For both analyses, the datasets were rarefied (N=1000) so that each study within a study had the same number of sequences.
+Second, for each study we partitioned the subjects into a low or high group depending on whether their alpha-diversity metrics were below or above the median value for the study.
+The relative risk (RR) was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group.
+We then performed a Fisher exact-test to investigate whether the RR was significantly different from 1.0 within each study and across all of the studies using the epiR (`r packageDescription("epiR")$Version`) and metafor (`r packageDescription("metafor")$Version`) packages.
+Third, we used the AUCRF (`r packageDescription("AUCRF")$Version`) R package to generate Random Forest models.
+For each study we developed models using either OTUs or genus-level phylotypes.
+The quality of each model was assessed by measuring the area under the curve (AUC) of the Receive Operating Characteristic (ROC) using ten-fold cross validation.
+Because the genus-level phylotype models were developed using a common reference, it was possible to use one study's model (i.e.
+the training set) to classify the samples from the other studies (i.e.
+the testing sets).
+The optimum threshold for the training set was set as the probability threshold that had the highest combined sensitivity and specificity.
+This threshold was then used to calculate the accuracy of the model applied to the test studies.
+To generate Receiver Operator Characteristic (ROC) curves and calculate the accuracy of the models we used the pROC (`r packageDescription("pROC")$Version`) R package.
+Finally, we performed power and sample number simulations for different effect sizes for each study using the pwr (`r packageDescription("pwr")$Version`) R package and base R functions.
 We also calculated the actual sample size needed based on the effect size of each individual study.
-To complete this analysis we used the pwr (`r packageDescription("pwr")$Version`) R package and base R functions.
-The analysis was performed using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`).
+
+***Reproducible methods***
+A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_mBio_2016/.
 

### 2016-06-01 submission/Sze_Obesity_PeerJ_2016.Rmd
Add references to mouse studies

@@
 Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since different communities yield variation in animal weight [REFERENCES].
+Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since direct manipulation of the communities yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Cox2014; @Mahana2016].
 The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.

### 2016-06-01 submission/Sze_Obesity_PeerJ_2016.Rmd
Work through Results section

--- 2016-05-31 submission/Sze_Obesity_PeerJ_2016.Rmd
@@
 
-# Lack of power and small effect size confounds the ability to differentiate obese and non-obese individuals using gut microbiome data
+# Lack of power and small effect size confounds the ability to differentiate non-obese and obese individuals using gut microbiome data
 
@@
 ```{r datasets}
-alpha <- read.table(file="data/process/alpha_composite.summary", header=T)
-shannon_effect_size <- 100*abs(1-alpha[alpha$metric=="shannon","obese"]/alpha[alpha$metric=="shannon","non_obese"])
+alpha <- read.table(file="data/process/alpha_tests.summary", header=T)
+
+alpha_composite <- read.table(file="data/process/alpha_composite.summary", header=T)
+alpha_effect_size <- 100*abs(1-alpha_composite[,"obese"]/alpha_composite[,"non_obese"])
+names(alpha_effect_size) <- alpha_composite$metric
+
+relative <- read.table(file="data/process/relative_risk.summary", header=T)
+
+relative_composite <- read.table(file="data/process/relative_risk.composite", header=T)
+
 beta <- read.table(file="data/process/beta_tests.summary", header=T)
 
+otu <- read.table(file="data/process/random_forest.otu.summary", header=T)
+
 genus <- read.table(file="data/process/random_forest.genus.summary", header=T)
-otu <- read.table(file="data/process/random_forest.otu.summary", header=T)
 
@@
  												power$effect_size==0.05, "power"] > 0.8)
+
+rr_power <- read.table(file="data/process/rr_power.predicted", header=T)
 
@@
 Alpha-diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances.
-Although these tests yielded small P-values, the difference between the Shannon diversity index of obese and non-obese individuals was `r shannon_effect_size`%.
+Although these tests yielded small P-values, the difference between the Shannon diversity index of non-obese and obese individuals was `r alpha_effect_size["shannon"]`%.
 A power analysis demonstrated that `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity.
@@
 Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since different communities yield variation in animal weight [REFERENCES].
+Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since direct manipulation of the communities yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016].
 The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.
@@
 
-***Separate and Pooled Analysis***
-We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B:F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05), ]))` significant results; two for Shannon diversity [Figure 2A], three for observed richness [Figure S1], one for evenness [Figure S2], and one for Bacteroidetes [Figure S3].
-There were no significant results for Firmicutes [Figure S4] or the B:F ratio [Figure 2B].
-Using the linear mixed-effect model to analyze the data pooled together while also controling study effects we found that for Shannon diversity  there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 2A].
-Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`) [Figure S1].
-However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B:F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`)  [Figure 2B & S2-S4] using this approach, between obese and non-obese individuals.
+```{r}
+alpha_sig <- alpha[alpha$p.value < 0.05,]
+sig_counts <- table(alpha_sig$metric)
 
-***Classical Pooled Meta-analysis***
+stopifnot(sig_counts['bf_ratio']==0)
+stopifnot(sig_counts['firmicutes']==0)
+```
 
-We next moved on to the pooled meta-analysis.
-For the Shannon diversity RR, all studies but Wu, Goodrich, and Turnbaugh had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 3A].
-The pooled analysis showed an increased RR of `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['shannon', 'ci_lb'],2), nsmall=2)` - `r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was significant (P-value = `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`).
-In contrast, when we investigated the B:F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`) [Figure 3B].
-The pooled RR was `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)` - `r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`).
-The differences in RR that was observed also do not seem to be related to the variable region chosen [Figure 3].
+***Alpha diversity analysis***
+We calculated the Shannon diversity index, richness, and Shannon evenness,  the relative abundance of *Bacteroidetes* and *Firmicutes*, and the ratio of their relative abundance (B:F) for each sample.
+Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the ten studies.
+The B:F and the relative abundance of *Firmicutes* were not significantly associated with obesity in any study.
+We identified `r sum(alpha$p.value < 0.05)` P-values less than 0.05: `r number_string[sig_counts["sobs"]]` studies indicated obese individuals had a lower richness, `r number_string[sig_counts["shannon"]]` studies indicated a significantly lower diversity, `r number_string[sig_counts["shannoneven"]]` study indicated a significantly lower evenness, and `r number_string[sig_counts["bacteroidetes"]]` study indicated a significantly higher relative abundance of *Bacteroidetes* (Figures 2 and S1).
+These results largely match those of the Walters and Finucane re-analysis studies.
+Interestingly, although only `r number_string[sig_counts["shannon"]]` of the 10 studies observed the previously reported association between lower diversity and obesity, the other studies appeared to have the same trend, albeit the differences were not statistically significant.
+We used a random effects linear model to combine the studies using the study as the random effect and found statistical support for decreased richness, evenness, and diversity among obese individuals (all P<`r max(alpha_composite[alpha_composite$p <0.05, "p"])`).
+Although there was a significant relationship between these metrics and diversity and obesity status, the effect size was quite small.
+The obese individuals averaged `r alpha_effect_size["sobs"]`% lower richness, `r alpha_effect_size["shannoneven"]`% lower evenness, and `r alpha_effect_size["shannon"]`% lower diversity.
+There were no significant associations when we pooled the phylum-level metrics across studies.
+These results indicate that obese individuals do have a statistically significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant.
 
-We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis.
-From the graph we can observe that there does not seem to be any significant  bias [Figure S5].
-Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.
+```{r}
+rr_sig <- relative[relative$p.value < 0.05,]
+sig_counts <- table(rr_sig$metric)
 
-***Random Forest Classification and Predicition Accuracy***
+stopifnot(nrow(rr_sig) == 3)
+stopifnot(sig_counts['bf_ratio']==0)
+stopifnot(sig_counts['firmicutes']==0)
+stopifnot(sig_counts['bacteroidetes']==0)
 
-Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.
-We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)` for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.
-We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al.
-study) to `r max(rand_forest.genus.summary$k_opt)` (Goodrich, et al.
-Study).
-For the OTU level the number of variables ranged from `r min(rand_forest.otu.summary$k_opt)` (Wu, et al.
-study) to `r max(rand_forest.otu.summary$k_opt)` (Goodrich, et al.
-study).
-The total number of trees and node size was standardized to 500 and 20 respectively for each data set.
-Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05) regardless if it was using genus or OTU level classification [Figure 4].
-Overall most studies performed poorly when used on other data sets for classification [Figure 5] with no significant difference between studies for their mean accuracy for prediction of non-obese and obese indivdiuals (P-value > 0.05).
+concatenate_rr <- function(data){
 
-***Power and Sample Size Estimates and Simulations***
+	paste0(format(data$rr, digits=2, nsmall=2L), " (", format(data$ci_lb, digits=2, nsmall=2L), "-", format(data$ci_ub, digits=2, nsmall=2L), ")")
 
-For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 6A] compared to the B:F ratio [Figure 7A] differences.
-When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity (predicted percent power = `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`) [Figure 6A].
-This predicted power was also similar for evenness (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`).
-However, this was not the case for the B:F ratio (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)` +/- `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`%) [Figure 7A] or the other measures that were tested (observed richness, Bacteroidetes, Firmicutes) where even a 15% difference between non-obese and obese groups only resulted in a relatively poor average power between studies.
-Looking at the sample sizes showed that on average to detect a 5% difference in Shannon diversity at 80% power between non-obese and obese individuals a total n of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) would be needed for each group [Figure 6B].
-In contrast in order to detect a 5% difference with 80% power in the B:F ratio a sample size of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) per group [Figure 7B].
-A complete breakdown of the power simulations and sample size estimates for the other variables investigated and RR can be found in the online supplement [Figure S6-S11].
+}
+```
 
-We also investigated if the combined total n for the linear mixed effect model was sufficiently powered.
-Using a small (0.2), medium (0.5), and large (0.8) cohen value we estimated the power of the combined data sets as well as the average cohen value from across studies for Shannon Diversity (0.06) and BF ratio (0.01).
-For the small effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=0.2, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
-For the medium effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=0.5, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
-For the large effect for combined data power calculation a value of
-`r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=0.8, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.
-The average actual cohens d value for Shannon Diversity for all `r length(rownames(beta_tests.summary))` studies of `r format(round(shannond, 2), nsmall=2)` gave a power of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=shannond, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
-Finally, the average actual cohens d value for B:F ratio for all `r length(rownames(beta_tests.summary))` studies of `r format(round(bf_ratiod, 2), nsmall=2)` yieled a power of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=bf_ratiod, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.
+***Relative risk***
+Building upon the alpha diversity analysis we calculated the relative risk of being obese based on whether an individual's alpha diversity metrics were below or above the median metric for that study.
+The results using relative risk largely matched those of using the untransformed alpha diversity data.
+Across the ten studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2).
+Again, although the relative risk values were not significant for other studies, the values tended to be above one.
+When we pooled the data using a random effects model, the relative risk associated with having a richness, evenness, or diversity below the median for the population was significantly associated with obesity (all P<`r max(relative_composite[relative_composite$p_value <0.05, "p_value"])`).
+The relative risks associated with alpha diversity were small.
+The relative risk of having a low richness was `r concatenate_rr(relative_composite["sobs",])`, low evenness was `r concatenate_rr(relative_composite["shannoneven",])`, and low diversity was `r concatenate_rr(relative_composite["shannon",])`.
+There were no significant difference in the phylum-level metrics.
+Again, the relative risk results indicate that individuals with a lower richness, evenness, or diversity are at statistically significant increased risk of being obese, it is questionable whether that risk is biologically or clinically relevant.
 
-## Discussion:
+```{r}
+beta_sig <- beta[beta$p_value < 0.05,]
+stopifnot(beta_sig$dataset == c("escobar", "turnbaugh", "goodrich"))
+```
+
+***Beta diversity analysis***
+Following the approach used by the Walters and Finucane re-analysis studies, for each dataset we calculated a Bray-Curtis distance matrix to measure the difference in the membership and structure of the individuals from each study.
+We then used PERMANOVA to test for a significant differences between the structure of non-obese and obese individuals.
+The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05).
+Because it was not possible to ascertain the directionality of the difference in community structure nor perform a pooled analysis using studies that had non-overlapping 16S rRNA gene sequence regions it is unclear whether these differences reflect a broader, but perhaps small, shift in community structure between non-obese and obese individuals.
+
+```{r}
+stopifnot(test_summary[which.min(test_summary$accuracy), "train"]=="turnbaugh")
+stopifnot(test_summary[which.max(test_summary$accuracy), "train"]=="hmp")
+```
+
+***Development of a microbiome-based classifier of obesity***
+The Walters re-analysis study suggested that it was possible to classify individuals as being obese or non-obese based on the composition of their microbiota.
+We repeated this analysis with additional datasets using OTU and genus-level phylotype data.
+For each study we developed a Random Forest machine learning model to classify individuals.
+Using ten-fold cross validation, the observed AUC values varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A).
+So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset.
+The the observed AUC values for the models applied to the training datastes varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B).
+For each model we identified the probability where the sum of the sensitivity and specificity was the highest.
+We then used this to identify a threshold to calculate the accuracy of the models when applied to the other nine datasets (Figure 5).
+Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`).
+When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance.
+The ability to predict obesity status using the relative abundance of OTUs and genera in the communities is only marginally better than random.
+These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.
+
+```{r}
+which_div_at_05 <- which(power[power$metric=="shannon" & power$effect_size==0.05,"power"] > 0.80)
+
+stopifnot(power[which_div_at_05, "study"] == "goodrich")
+
+which_div_at_10 <- which(power[power$metric=="shannon" & power$effect_size==0.10,"power"] > 0.80)
+
+which_bf_at_15 <- which(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"] > 0.80)
+
+stopifnot(length(which_bf_at_15) == 0)
+
+stopifnot(sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.10,"power"] > 0.80) == 0)
+
+range_effect_size <- range(abs(100*(1-alpha[alpha$metric == "shannon","mean_obese"]/alpha[alpha$metric=="shannon", "mean_non"])))
+
+```
+
+***Power and Sample Size Estimate Simulations***
+The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals.
+To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S3-S8).
+Although there is no biological rationale for these effect sizes, they represent a range that is plausible.
+Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6).
+None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S5).
+In fact, the maximum power among any of the studies to detect a 15% difference in B:F values was `r max(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"])`.
+Among the tests for relative risk, none of the studies had sufficient power to detect a Cohen's d of 0.10 and only `r number_string[sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.15,"power"] > 0.80)]` studies had sufficient power to detect a Cohen's d of 0.15.
+We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups.
+To detect a 1, 5, 10, and 15% difference in Shannon index, the median sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively.
+To detect a 1, 5, 10, and 15% difference in B:F values, the median sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)`  individuals, respectively.
+To detect a 1, 5, 10, and 15% difference in relative risk values using Shannon diversity, the median sampling effort per group was approximately `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively.
+These estimates indicate that most microbiome studies are underpowered to detect modest effect sizes using either metric.
+In the case of obesity, the studies were underpowered to detect the `signif(range_effect_size, 1)[1]` to `r signif(range_effect_size, 1)[2]`% difference in diversity that was observed across the studies.
+
+## Discussion
 
@@
 Further none of the selected variables from each indivdiual study did a respectable job of classification when applied to the other data sets.
-This would suggest that within each indivdiual study classification of obese and non-obese is relatively strong but not generalizable across studies.
+This would suggest that within each indivdiual study classification of non-obese and obese is relatively strong but not generalizable across studies.
 This might indicate an ancestry, ethnicity, or a specific environmental effect such as diet difference between study populations that might cause this dissimilarity.
@@
 
-**Figure 2: Individual and Combined comparison of Obese and Non-Obese groups for Shannon Diversity and B:F Ratio.** **A)** For Shannon diversity only Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 2), nsmall = 2)` and Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 5), nsmall = 5)`) studies indivdiually were signficantly different for obese and non-obese groups.
+**Figure 2: Individual and Combined comparison of Obese and Non-Obese groups for Shannon Diversity and B:F Ratio.** **A)** For Shannon diversity only Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 2), nsmall = 2)` and Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 5), nsmall = 5)`) studies indivdiually were signficantly different for non-obese and obese groups.
 When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 5), nsmall = 5)`).

### 2016-06-01 code/finish_clustersplit.R
Reorganize code folder

@@ -1,41 +1,0 @@
-level_regexp <- "^0\\.03"
-
-extra_names_file <- "data/zeevi/zeevi.renamed.unique.good.filter.unique.precluster.denovo.uchime.pick.pick.count_table.extra.temp"
-extras <- read.table(file=extra_names_file, header=T, stringsAsFactors=F)[,1]
-
-list_files <- list.files('data/zeevi', pattern='*.list', full.names=T)
-file_number <- as.numeric(gsub(".*fasta.(\\d*)\\..*", "\\1", list_files))
-file_order <- order(file_number)
-list_files <- list_files[file_order]
-
-combined_data <- rep("", length(list_files))
-
-for(i in 0:(length(list_files)-1)){
-
-	list_line <- ""
-	list_file <- scan(list_files[i+1], what=character(), sep="\n", quiet=T)
-	hits <- grepl(level_regexp, list_file)
-
-	if(sum(hits) == 1){
-		list_line <- list_file[hits]
-	} else {
-		list_file[1] <- gsub("^unique", "0.00", list_file[1])
-		cutoffs <- as.numeric(gsub("^(0\\.\\d\\d)\\t.*", "\\1", list_file))
-
-		best_cutoff <- which.max(cutoffs[cutoffs <= 0.03])
-		list_line <- list_file[best_cutoff]
-	}
-
-	combined_data[i+1] <- gsub("^0\\.\\d\\d\\t\\d*\\t", "", list_line)
-}
-
-clustered <- paste(combined_data, collapse='\t')
-extra_merged <- paste(extras, collapse='\t')
-
-combined_data <- paste(clustered, extra_merged, sep='\t')
-
-n_otus <- sum(nchar(combined_data) - nchar(gsub("\t", "", combined_data)) + 1)
-
-new_list_line <- paste("0.03", n_otus, combined_data, sep='\t')
-
-write(new_list_line, file="data/zeevi/zeevi.renamed.unique.good.filter.unique.precluster.pick.pick.an.unique_list.list")

### 2016-06-01 submission/Sze_Obesity_PeerJ_2016.Rmd
Code in the number of studies

@@
 
+n <- length(unique(alpha$dataset))
+
 alpha_composite <- read.table(file="data/process/alpha_composite.summary", header=T)
@@
 
-number_string <- c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine")
+number_string <- c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine", "ten")
 ```
@@
 The hypothesis that variation in the gut microbiota can explain or be used to predict obesity status has received considerable attention and is frequently mentioned as an example for the role of the microbiome in human health.
-Here we assess this hypothesis using ten independent studies and find that although there is an association, it is smaller than can be detected by most microbiome studies.
+Here we assess this hypothesis using `r number_string[n]` independent studies and find that although there is an association, it is smaller than can be detected by most microbiome studies.
 Furthermore, we directly tested the ability to predict obesity status based on the composition of an individual's microbiome and find that the median classification accuracy is between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]`%.
@@
 The initial search yielded 187 studies.
-We identified ten additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
+We identified number_string[n] additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
 We then manually curated the 196 studies to select those studies that included BMI and sequence data.
@@
 Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity.
-The ten studies are summarized in Tables 1 and 2.
+The number_string[n] studies are summarized in Tables 1 and 2.
 
@@
 We calculated the Shannon diversity index, richness, and Shannon evenness,  the relative abundance of *Bacteroidetes* and *Firmicutes*, and the ratio of their relative abundance (B:F) for each sample.
-Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the ten studies.
+Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the number_string[n] studies.
 The B:F and the relative abundance of *Firmicutes* were not significantly associated with obesity in any study.
@@
 The results using relative risk largely matched those of using the untransformed alpha diversity data.
-Across the ten studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2).
+Across the number_string[n] studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2).
 Again, although the relative risk values were not significant for other studies, the values tended to be above one.
@@
 For each model we identified the probability where the sum of the sensitivity and specificity was the highest.
-We then used this to identify a threshold to calculate the accuracy of the models when applied to the other nine datasets (Figure 5).
+We then used this to identify a threshold to calculate the accuracy of the models when applied to the other number_string[n-1] datasets (Figure 5).
 Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`).

### 2016-06-01 code/run_alpha_diversity.R
Add Schubert dataset

@@ -42,7 +42,7 @@ single_analysis <- function(alpha, pow, is_obese){
 	)
 }
 
-datasets <- c('baxter', 'goodrich', 'escobar', 'hmp', 'ross', 'turnbaugh', 'wu', 'zupancic', 'zeevi')
+datasets <- c('baxter', 'goodrich', 'escobar', 'hmp', 'ross', 'schubert', 'turnbaugh', 'wu', 'zupancic', 'zeevi')
 
 run <- function(datasets){
 

### 2016-06-02 submission/Sze_Obesity_PeerJ_2016.Rmd
Edit discussion section

@@
 csl: msystems.csl
-fontsize: 11pt
+fontsize: 12pt
 geometry: margin=1.0in
@@
 The Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines were applied to identify 10 studies for an updated and more synthetic analysis.
-Alpha-diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances.
+Alpha diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances.
 Although these tests yielded small P-values, the difference between the Shannon diversity index of non-obese and obese individuals was `r alpha_effect_size["shannon"]`%.
@@
 For both analyses, the datasets were rarefied (N=1000) so that each study within a study had the same number of sequences.
-Second, for each study we partitioned the subjects into a low or high group depending on whether their alpha-diversity metrics were below or above the median value for the study.
+Second, for each study we partitioned the subjects into a low or high group depending on whether their alpha diversity metrics were below or above the median value for the study.
 The relative risk (RR) was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group.
@@
 So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset.
-The the observed AUC values for the models applied to the training datastes varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B).
+The the observed AUC values for the models applied to the training datasets varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B).
 For each model we identified the probability where the sum of the sensitivity and specificity was the highest.
@@
 
-Our meta-analysis helps to provide a little more clarity to the on going debate of whether or not there are specific microbiome variables that are influenced by obesity.
-Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a microbiome difference between non-obese and obese individuals.
-However, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give.
-Here we perform an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.
-We analyzed the data using three different approaches and also performed post-hoc power and sample size calculations and simulations.
-The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach uses linear mixed effect models to adjust for study effects and invetigates both predicted power and sample number needed for the observed effect sizes.
+Our meta-analysis helps to provide a clarity to the ongoing debate of whether or not there are specific microbiome-based markers that can be associated with obesity.
+We performed an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.
+By statistically pooling the data from ten studies, we observed significant, but small, relationships between richness, evenness, and diversity and obesity status as well as the relative risk of being obese based on these metrics.
+We also generated Random Forest machine learning models trained on each dataset and tested on the remaining datasets.
+This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome is limited.
+Finally, we assessed the ability of each study to detect defined differences in alpha diversity and observed that most studies were underpowered to detect modest effect sizes.
+Considering these datasets are among the largest published, it appears that most microbiome studies are underpowered to detect differences in alpha diversity.
 
-Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the microbiome and obesityy could be observed.
-Further, even though random forest could identify specific variables with similar taxonomic classifications that resulted in good ROC curves [Figure 1] these variables varied widely by both their actual sequence and directionality.
-Similarily, for the B:F ratio we found that the other two approaches, that used the pooled data approach, were also not able to identify a correlation between measures of the microbiome and obesity [Figure 3 & 4].
-For these approaches both the B:F ratio failed to have a significant correlation to obesity.
-However, a small but significant differences was observed between non-obese and obese individauls  for Shannon diversity [Figure 3 & 4].
-In order to also assess the amount of our results that could be due to insufficient power we performed power and sample size analysis.
-Overall, we found that most studies for Shannon Diversity would be powered to find a 15% difference, however, the vast majority of reported effect sizes between studies ranged from `r format(round(min(alpha_effectSize[, 'shannon']), 2), nsmall=2)`-`r format(round(max(alpha_effectSize[, 'shannon']), 2), nsmall=2)` percent.
-Although the effect sizes for the BF ratio were larger than that for Shannon Diversity [Figure 5A & 6A] the estimated power remained low.
-Additionaly, we would also need a relatively large sample size to be able to detect the actual differences in each study for both Shannon Diversity and B:F ratio [Figure 5B & 6B].
-Taken together we believe that this analysis builds on the previous two re-anlaysis and provides additional information as to whether there is a difference between non-obese and obese human individuals with respect to both Shannon diversity and the B:F ratio.
+Alpha diversity metrics are attractive because they distill a complex dataset to a single value.
+For example, diversity is a measure of the entropy in a community and integrates richness and evenness information.
+Two communities with little taxonomic similarity can have the same diversity.
+Among ecologists the relevance of these metrics is questioned because it is difficult to ascribe a mechanistic interpretation to their relationship with stability or disease.
+Regardless, the concept of a biologically significant effect size needs to be developed among microbiome researchers.
+Alternative metrics could include the ability to detect a defined difference in the relative abundance of an OTU representing a defined relative abundance.
+What makes for a biologically significant difference or relative abundance is an important point that has yet to be discussed in the microbiome field.
+The use of operationally defined effect sizes should be adequate until it is possible to decide upon an accepted practice.
 
-We do acknowledge that this study does have a few limitations that are worth mentioning.
-First,  it has been documented that in a number of groups BMI is not a good classification of obesity [@who_expert_consultation_appropriate_2004; @rahman_accuracy_2010].
-Of particular note, the asian population has a separate set of guidelines for obesity [@who_expert_consultation_appropriate_2004].
-Although the asain population in the studies analyzed are very small [Table 1] it is possible that this may contribute to some of the noise in the data.
-However, it is important to keep in mind that for these measures, specifically the RR, there does not seem to be any study bias due to studies having more ethnic diversity [Figure 4].
-Additionally, we are also limited in our BMI classifications by what the previous studies published and performed.
-In general, not enough studies have been done to investigate the microbiome in the context of other markers for obesity or on different ethnicities to be able to perform the analysis that we did.
-Thus as more specific studies are published that address ethnicity and studies on the microbiome that use better markers of obesity are completed it is entirely possible that the noise associated with these measurements will decrease.
+By selecting a range of possible effect sizes, we were able to demonstrate that most studies are underpowered to detect modest differences in alpha diversity metrics and phylum-level relative abundances.
+Several factors interact to limit the power of microbiome studies.
+There is wide interpersonal variation in the diversity and structure of the human microbiome.
+In addition, the common experimental designs limit their power.
+As we observed, most of the studies included in our analysis were unbalanced for the variable that we were interested in.
+This was also true of those studies that originally sought to identify associations with obesity.
+Even with a balanced design, we showed that it was necessary to obtain approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)` and `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)` sequences per sample to detect a 5% difference in Shannon diversity or B:F, respectively.
+It was interesting that these sample sizes agreed across studies regardless of their sequencing method, region with in the 16S rRNA gene, or subject population (Figure 6).
+This suggests that regardless of the treatment or category, these sample sizes represent a good starting point for subject recruitment when using stool samples.
+Unfortunately, few studies have been published with this level of subject recruitment.
+This is troubling since the positive predictive rate of a significant finding in an underpowered study is small leading to results that cannot be reproduced [ref].
+Future microbiome studies should articulate the basis for their experimental design.
 
-Second, although we make attempts to correct for study effects in our pooled analysis  so that we can compare the microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.
-To help mitigate some of these problems we analyzed each data set separately and processed all samples through a uniform pipeline in an attempt to mitigate a portion of the variation that might occur due to sequence processing.
-It should also be noted that we also do not see any specific bias in the forest plots [Figure 3] based on hypervariable region.
-Further many of the studies show very similar actual effect sizes for Shannon Diversity difference between non-obese and obese individuals.
-However, even with all these precautions and safety checks we do realize that it may not be possible to compare these studies together and that further studies need to be completed to verify the observations made in this report.
+Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that there was not a consistent association between alpha diversity and obesity; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference.
+Our analysis also used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from 3 studies [refs] and a metagenomic study and the Walters study used 16S rRNA gene sequence data from 5 studies [refs]; two studies were included in both studies.
+Our analysis included 4 of these studies [refs] and excluded 3 of the studies because they were too small [ref], only utilized metagenomic data [ref], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering.
+The additional six datasets were published after the two reviews were performed and include datasets with more samples than were found in the original studies.
+Our collection of `r number_string[n]` studies allowed us to largely use the same sequence analysis pipeline for all datasets and relied heavily on the availability of public data and access to metadata that included variables beyond the needs of the original study.
+To execute this analysis, we created an automated data analysis pipeline, which can be easily updated to add additional studies as they become available.
+Similarly, it would be possible to adapt this pipeline to other body sites and treatment or variables (e.g.
+subject's sex or age).
 
-Third, although we may not have detected a signal between non-obese and obese individuals our results indicate that we may not have been sufficiently powered to do so.
-However, taking both the sample size calculations and power calculations together would indicate that if there is a true difference to be found between non-obese and obese individuals this difference would be very small.
-Additionally, due to the very small effect size potentially the more important question that does need to be raised is whether or not this difference is truly biologically meaningful.
-
-Fourth, although the ROC curves for each study were relatively good there was very little actual overlap between studies based on sequence similarity and directionality.
-Further none of the selected variables from each indivdiual study did a respectable job of classification when applied to the other data sets.
-This would suggest that within each indivdiual study classification of non-obese and obese is relatively strong but not generalizable across studies.
-This might indicate an ancestry, ethnicity, or a specific environmental effect such as diet difference between study populations that might cause this dissimilarity.
-Additionally, an alternative explanation that may be more reasonable is that the random forest model is overfit for some of the studies examined which explains why the variables selected are less generalizable across studies.
-In actuality it may be a combination of all of the factors mentioned as to why generalizability is problematic.
+Similar to our study, the Walters analysis [@walters_meta-analyses_2014], the authors generated Random Forest machine learning models to differentiate between non-obese and obese individuals.
+They obtained similar AUC values to our analysis; however, they did not attempt to test these models on the other studies in their analysis.
+When we performed the inter-dataset cross validation the median accuracy across datasets was only `r 100*median(test_summary$accuracy)`% indicating that the models did a poor job when applied to other datasets.
+This could be due to differences in subject populations and methods.
+Considering the median AUC for models trained and tested on the same data with ten-fold cross validation only varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)` and there was not a strong signal in the alpha diversity data, we suspect that there is insufficient signal to reliably classify individuals.
 
 Although we failed to find an effect it is not realistic to necessarily state that there is no microbiome impact on obesity.
-There is a large body of very strong evidence in mice on the effect the microbiome has on obesity [@turnbaugh_effect_2009; @koren_host_2012].
-It is not possible to discount these findings and it may be the case that in humans there is just too many other variables and factors that effect the microbiome which makes the signal too hard to detect in a cross sectional manner.
-Some well designed studies in humans and diet, in the past, have shown that the noise inherent to microbiome studies can be overcome with very careful attention made to controlling the study conditions [@ley_microbial_2006].
-
-Overall, this study provides a reasonable follow up to the two previous re-analysis studies to assess the association between the microbiome and obesity.
-In doing so we believe that we have provided evidence that if there is an effect size present it is small and will take many more studies and samples to be analyzed to confirm a difference between non-obese and obese individuals.
+There is strong evidence in murine models of obesity that the microbiome and level of adiposity can be manipulated via genetic manipulation of the animal and manipulation of the community through antibiotics or colonizing germ free mice with diverse fecal material from human donors [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016; @ley_obesity_2005].
+These studies appear to conflict with the observations using human subjects.
+Recalling the large interpersonal variation in the structure of the microbiome, it is possible that each individual has their own signatures of obesity.
+Alternatively, it could be that the involvement of the microbiome in obesity is at the level of a common set of metabolites that can be produced from different structures of the microbiome.
 
 ## Acknowledgements
-
-The authors would like to thank Nielson Baxter and Shawn Whitefield for their suggestions on the manuscript.
-Finally, we would like to thank all the authors who have made their data publicly available.
-Without this initial forethought studies such as these would not be possible.
+The authors would like to thank Nielson Baxter and Shawn Whitefield for their suggestions on the development of the manuscript.
+We are grateful to the authors of the studies used in our meta-analysis who have made their data publicly available or available to us directly.
+Without their forethought studies such as this would not be possible.
+This work was supported in part by funding from the National Institutes of Health to PDS (U01AI2425501 and P30DK034933).
 

### 2016-06-02 submission/Sze_Obesity_PeerJ_2016.Rmd
Muck with figure legends

@@
 
-## Tables
-
 **Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
-```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
-
-table1 <- read.table("../results/tables/table1.demographics", header = T)
-
-colnames(table1) <- c("Study", "Age (SD)", "Sex (F|M)", "European Ancestry (%)", "Obese (No|Yes)", "BMI (SD)", "Min BMI","Max BMI")
-
-kable(table1, format = "markdown", align = 'c', padding = 2)
-
-```
-
-* For Turnbaugh only BMI group information was provided and only an age range was given.
 
@@
 
-**Table 2. Summary of Important Study Characteristics.**
+**Figure 1: PRISMA flow diagram of total records searched [@Moher2009].**
 
-```{r Table_2, echo=FALSE, message=FALSE, warning=FALSE}
+**Figure 2: Individual and combined comparison of obese and non-obese groups for Shannon diversity (A) and B:F (B).**
 
-Study <- as.character(table1$Study)
-#New Order
-  #baxter, escobar, hmp, ross, schubert, turnbaugh, wu, zupancic, goodrich, zeevi
+**Figure 3: Meta analysis of the relative risk of obesity based on Shannon diversity (A) or B:F (B).**
 
-Type <- c("Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational", "Observational")
+**Figure 4: ROC curves for each study based on classification of non-obese or obese groups using OTUs (A) or genus-level classification (B).**
 
-Population <- c("Adults without Cancer from Texas, Ontario, Michigan, Massachusetts", "Healthy adults from Columbia", "Healthy Adults from Pennsylvania", "Hispanic adults with and without type 2 diabetes from Texas
-", "Nondiarrheal indivdiuals without infection history", "Adult twins and mothers from Missouri", "Healthy Adults from Texas and Missouri", "Amish Adults with and without metabolic syndrome from Pennsylvania", "Adult twins and mothers from the United Kingdom", "Healthy Adults from Tel Aviv")
+**Figure 5: Overall accuracy of each study to predict non-obese and obese individuals based on that study's Random Forest machine learning model applied to each of the other studies.**
 
-Published <- c("Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes","Yes", "Yes", "Yes")
+**Figure 6: Power (A) and sample size simulations (B) for Shannon diversity for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-Outcomes <- c("Did not investigate control populations for relationship with BMI", "Firmicutes less abundant at higher BMI", "No significant correlations", "Mentions obesity, no specific correlations made though", "Did not investigate control populations for relationship with BMI", "Unweighted UniFrac Correlation with BMI", "Decrease diversity, Bacteroidetes, and Actinobacteria correlated with increased BMI ", "Bacteroidetes/Firmicutes ratio adjusted BMI correlation", "Christensenellaceae higher in low BMI individuals.
-No other correlations found.", "Found associations with postprandial (post meal) glycemic responses, which correlates with BMI, and Proteobacteria, Actinobacteria, and Enterobacteriaceae")
+\newpage
 
-table2 <- cbind(Study, Type, Population, Published, Outcomes)
+**Figure S1: Individual and Combined comparison of Obese and Non-Obese groups Based on Evenness (A), Richness (B), or the Relative Abundance of Bacteroidetes (C) and Firmictues (D).**
 
-colnames(table2) <- c("Study", "Study Type", "Population", "Published", "Outcomes Reported")
+**Figure S2: Meta Analysis of the Relative Risk of Obesity Based on Evenness (A), Richness (B), or the Relative Abundance of Bacteroidetes (C) and Firmictues (D).**
 
-kable(table2, format = "markdown", align = 'c', padding = 2)
-```
+**Figure S3: Power (A) and sample size simulations (B) for B:F for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-*******
-##### Figures:
+**Figure S4: Power (A) and sample size simulations (B) for richness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure 1: PRISMA Flow Diagram of Total Records Searched.**  This was adapted from:  Moher D, Liberati A, Tetzlaff J, Altman DG, The PRISMA Group (2009).
-Preferred Reporting Items for Systematic Reviews and Meta-Analyses: The PRISMA Statement.
-PLoS Med 6(7): e1000097.
-doi:10.1371/journal.pmed1000097.
+**Figure S5: Power (A) and sample size simulations (B) for evenness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure 2: Individual and Combined comparison of Obese and Non-Obese groups for Shannon Diversity and B:F Ratio.** **A)** For Shannon diversity only Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 2), nsmall = 2)` and Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 5), nsmall = 5)`) studies indivdiually were signficantly different for non-obese and obese groups.
-When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 5), nsmall = 5)`).
-**B)** For the B:F ratio there was no individual study that was significant nor was their a difference when the data was combined and study effect corrected for between Non-Obese and Obese groups (P > 0.05).
+**Figure S6: Power (A) and sample size simulations (B) for the relative abundance of Bacteroidetes for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure 3: Meta Analysis of the Relative Risk of Obesity Based on Shannon Diversity or Bacteroidetes/Firmicutes Ratio.**  Groups were divided for each study on high and low Shannon Diversity or Bacteroidetes/Firmicutes (B:F) ratio groups based on the median for that study.
-Red represents studies that used the 16S rRNA V4 region, yellow represents studies that used a V1-V3 approach, blue represents studies that used either a V1-V2 or just a V2 approach, green represents studies that used a V3-V4 region, and orange represents studies that used the V3-V5 region.
-**A)** The overall pooled relative risk for Shannon Diversity was `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` with a P-value of `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`.
-**B)** The overall pooled relative risk for the B:F ratio was  `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` with a P-value of `r format(round(RR_composite['bf_ratio', 'p_value'], 2), nsmall=2)`.
+**Figure S7: Power (A) and sample size simulations (B) for the relative abundance of Firmicutes for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure 4: ROC Curves for Each Study based on Classification of Non-Obese or Obese Groups at the Genus or OTU level.**  **A)** No significant difference was found between the different ROC curves based on Study for the genus level.
-The overall AUCs ranged from `r round(min(rand_forest.genus.summary$auc), 2)` to `r round(max(rand_forest.genus.summary$auc), 2)` with a mean of `r round(mean(rand_forest.genus.summary$auc), 2)`.
-**B)** Similarly no difference was found between the ROC curves based on Study at the OTU level.
-The overall mean of the AUCs were slightly higher overall than the genus level `r round(mean(rand_forest.otu.summary$auc), 2)` but the range was greater going from `r round(min(rand_forest.otu.summary$auc), 2)` to `r round(max(rand_forest.otu.summary$auc), 2)`.
+**Figure S8: Power (A) and sample size simulations (B) for relative risk of obesity based on Shannon diversity.**  Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure 5: Overall Accuracy of Each Study to Predict Non-Obese and Obese Individuals based on that Study's Classification Variables** Overall there is no significant difference between the studes in mean accuracy.
-Additionally, the accuracy for each study group to predict Non-Obese and Obese individuals was relatively low.
+\newpage
 
-**Figure 6: Summary of Power and Sample Size Simulations for Shannon Diversity for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-**Figure 7.
-Summary of Power and Sample Size Simulations for Bacteroidetes/Firmicutes Ratio for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-*******
-##### References:
+## References

### 2016-06-02 submission/Sze_Obesity_PeerJ_2016.Rmd
Add references

@@
 Unfortunately, few studies have been published with this level of subject recruitment.
-This is troubling since the positive predictive rate of a significant finding in an underpowered study is small leading to results that cannot be reproduced [ref].
+This is troubling since the positive predictive rate of a significant finding in an underpowered study is small leading to results that cannot be reproduced [@Ioannidis2005].
 Future microbiome studies should articulate the basis for their experimental design.
@@
 Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that there was not a consistent association between alpha diversity and obesity; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference.
-Our analysis also used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from 3 studies [refs] and a metagenomic study and the Walters study used 16S rRNA gene sequence data from 5 studies [refs]; two studies were included in both studies.
-Our analysis included 4 of these studies [refs] and excluded 3 of the studies because they were too small [ref], only utilized metagenomic data [ref], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering.
-The additional six datasets were published after the two reviews were performed and include datasets with more samples than were found in the original studies.
+Our analysis also used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from 3 studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from 5 studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012].
+Our analysis included 4 of these studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011] and excluded 3 of the studies because they were too small [@ley_microbial_2006], only utilized metagenomic data [@Qin2010], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering [@yatsunenko_human_2012].
+The additional `r number_string[n-3]` datasets were published after the two reviews were performed and include datasets with more samples than were found in the original studies.
 Our collection of `r number_string[n]` studies allowed us to largely use the same sequence analysis pipeline for all datasets and relied heavily on the availability of public data and access to metadata that included variables beyond the needs of the original study.

### 2016-06-02 submission/Sze_Obesity_mBio_2016.Rmd
Get write.paper to work

--- 2016-06-02 submission/Sze_Obesity_PeerJ_2016.Rmd
@@
 title: "Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome"
-bibliography: references.bib
+bibliography: reference.bib
 output:
@@
       in_header: header.tex
-csl: msystems.csl
+csl: mbio.csl
 fontsize: 12pt
@@
 
-```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
+```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, message=FALSE}
+source('code/utilities.R')
+get_dependencies(c("lme4", "vegan", "epiR", "metafor", "AUCRF", "pROC", "pwr"))
+
 opts_chunk$set("tidy" = TRUE)
@@
 
-**Running Title:** The Human Microbiome and Obesity
-<br>
+Running Title: The Human Microbiome and Obesity
+
+\vspace{10mm}
+
 Marc A Sze and Patrick D Schloss${^\dagger}$
-<br>
-**Contributions:** Both authors contributed to the planning, design, execution, interpretation, and writing of the analyses.
 
-\vspace{30mm}
+\vspace{10mm}
+
+Contributions: Both authors contributed to the planning, design, execution, interpretation, and writing of the analyses.
+
+\vspace{20mm}
 
@@
 ## Abstract
+
 Two recent studies have re-analyzed published data and found that when datasets are analyzed independently there was limited support the widely accepted hypothesis that changes in the microbiome are associated with obesity.
@@
 A power analysis demonstrated that `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity.
-When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]` (median=`r median(test_summary$accuracy)`).
+When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r 100*range(test_summary$accuracy)[1]` and `r 100*range(test_summary$accuracy)[2]`% (median=`r 100*median(test_summary$accuracy)`%).
 Although there is statistical support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.
 
-## Importance
+### Importance
+
 As interest in the human microbiome grows there is an increasing number of studies that can be used to test numerous hypotheses across human populations.
@@
 Here we assess this hypothesis using `r number_string[n]` independent studies and find that although there is an association, it is smaller than can be detected by most microbiome studies.
-Furthermore, we directly tested the ability to predict obesity status based on the composition of an individual's microbiome and find that the median classification accuracy is between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]`%.
+Furthermore, we directly tested the ability to predict obesity status based on the composition of an individual's microbiome and find that the median classification accuracy is between `r 100*range(test_summary$accuracy)[1]` and `r 100*range(test_summary$accuracy)[2]`%.
 This type of analysis can be used to design future studies and expanded to explore other hypotheses.
@@
 
-***Literature Review and Study Inclusion***
+***Literature Review and Study Inclusion.***
 We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010].
@@
 
-***Sequence Analysis Pipeline***
+***Sequence Analysis Pipeline.***
 All sequence data were publicly available and were downloaded from the NCBI Sequence Read Archive, the European Nucleotide Archive, or the investigators' personal website (https://gordonlab.wustl.edu/TurnbaughSE\_10\_09/STM\_2009.html).
@@
 
-***Data Analysis***
+***Data Analysis.***
 We split the overall meta-analysis into three general strategies using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`).
@@
 
-***Reproducible methods***
-A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_mBio_2016/.
+***Reproducible methods.***
+A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_Obesity_mBio_2016/.
 
@@
 
-***Alpha diversity analysis***
+***Alpha diversity analysis.***
 We calculated the Shannon diversity index, richness, and Shannon evenness,  the relative abundance of *Bacteroidetes* and *Firmicutes*, and the ratio of their relative abundance (B:F) for each sample.
-Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the number_string[n] studies.
+Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the `r number_string[n]` studies.
 The B:F and the relative abundance of *Firmicutes* were not significantly associated with obesity in any study.
@@
 
-***Relative risk***
+***Relative risk.***
 Building upon the alpha diversity analysis we calculated the relative risk of being obese based on whether an individual's alpha diversity metrics were below or above the median metric for that study.
@@
 
-***Beta diversity analysis***
+***Beta diversity analysis.***
 Following the approach used by the Walters and Finucane re-analysis studies, for each dataset we calculated a Bray-Curtis distance matrix to measure the difference in the membership and structure of the individuals from each study.
@@
 
-***Development of a microbiome-based classifier of obesity***
+***Development of a microbiome-based classifier of obesity.***
 The Walters re-analysis study suggested that it was possible to classify individuals as being obese or non-obese based on the composition of their microbiota.
@@
 
-***Power and Sample Size Estimate Simulations***
+***Power and Sample Size Estimate Simulations.***
 The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals.
@@
 ## Acknowledgements
+
 The authors would like to thank Nielson Baxter and Shawn Whitefield for their suggestions on the development of the manuscript.

### 2016-06-02 submission/Sze_Obesity_mBio_2016.tex
Get write.paper to work

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}
\usepackage{comment}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\begin{document}
\maketitle

\section{Lack of power and small effect size confounds the ability to
differentiate non-obese and obese individuals using gut microbiome
data}\label{lack-of-power-and-small-effect-size-confounds-the-ability-to-differentiate-non-obese-and-obese-individuals-using-gut-microbiome-data}

\begin{center}
\vspace{25mm}

Running Title: The Human Microbiome and Obesity

\vspace{10mm}

Marc A Sze and Patrick D Schloss${^\dagger}$

\vspace{10mm}

Contributions: Both authors contributed to the planning, design, execution, interpretation, and writing of the analyses.

\vspace{20mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
\end{center}

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

Two recent studies have re-analyzed published data and found that when
datasets are analyzed independently there was limited support the widely
accepted hypothesis that changes in the microbiome are associated with
obesity. This hypothesis was reconsidered by increasing the number of
data sets and pooling the results across the individual datasets. The
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines were applied to identify 10 studies for an updated
and more synthetic analysis. Alpha diversity metrics and the relative
risk of obesity based on those metrics were used to identify a limited
number of significant associations with obesity; however, when the
results of the studies were pooled using a random effects model
significant associations were observed between Shannon diversity, number
of observed OTUs, and Shannon evenness and obesity status They were not
observed for the ratio of \emph{Bacteroidetes} and \emph{Firmicutes} or
their individual relative abundances. Although these tests yielded small
P-values, the difference between the Shannon diversity index of
non-obese and obese individuals was 2.07\%. A power analysis
demonstrated that one of the studies had sufficient power to detect a
5\% difference in diversity. When models trained on one dataset were
then tested using the other 9 datasets, the median accuracy varied
between 33.01 and 64.77\% (median=56.67\%). Although there is
statistical support for a relationship between the microbial communities
found in human feces and obesity status, this association is relatively
weak and its detection is confounded by large interpersonal variation
and insufficient sample sizes.

\subsubsection{Importance}\label{importance}

As interest in the human microbiome grows there is an increasing number
of studies that can be used to test numerous hypotheses across human
populations. The hypothesis that variation in the gut microbiota can
explain or be used to predict obesity status has received considerable
attention and is frequently mentioned as an example for the role of the
microbiome in human health. Here we assess this hypothesis using ten
independent studies and find that although there is an association, it
is smaller than can be detected by most microbiome studies. Furthermore,
we directly tested the ability to predict obesity status based on the
composition of an individual's microbiome and find that the median
classification accuracy is between 33.01 and 64.77\%. This type of
analysis can be used to design future studies and expanded to explore
other hypotheses.

\newpage

\subsection{Introduction}\label{introduction}

Obesity is a growing health concern with approximately 20\% of the youth
(aged 2-19) in the United States classified as either overweight or
obese (1). This number increases to approximately 35\% in adults (aged
20 or older) and these statistics have seen little change since 2003
(1). Traditionally the body mass index (BMI) has been used as the
traditional method of classifying individuals as non-obese or obese (2).
Recently, there has been increased interest in the role of the
microbiome in modulating obesity (3, 4). If the microbiome does affect
obesity status, then manipulating the microbiome could have a
significant role in the future treatment of obesity and in helping to
stem the current epidemic.

There have been several studies that report observing a link between the
composition of microbiome and obesity in animal models and in humans.
The first such study used genetically obese mice and observed the ratio
of the relative abundances of Bacteroidetes to Firmicutes (B:F) was
lower in obese mice than lean mice (5). Translation of this result to
humans by the same researchers did not observe this effect, but did find
that obese individuals had a lower diversity than lean individuals (6).
They also showed that the relative abundance of Bacteroidetes and
Firmicutes increased and decreased, respectively, as obese individuals
lost weight while on a fat or carbohydrate restricted diet (7). Two
re-analysis studies interrogated previously published microbiome and
obesity data and concluded that the previously reported differences in
community diversity and B:F among non-obese and obese individuals could
not be generalized (8, 9). Regardless of the results using human
populations, mechanistic studies using animal models that were
manipulated with antibiotics or colonization with varied communities
appears to support the association since direct manipulation of the
communities yielded variation in animal weight (10--13). The purported
association between the differences in the microbiome and obesity have
been widely repeated with little attention given to the lack of a clear
signal in human cohort studies.

The recent publication of additional studies that collected BMI data for
each subject as well as other studies that were not included in the
earlier re-analysis studies offered the opportunity to revisit the
question relating the structure of the human microbiome to obesity
(14--22). One critique of the prior re-analysis studies is that the
authors did not aggregate the results across studies to increase the
effective sample size. It is possible that there were small associations
within each study that were not statistically significant because the
individual studies lacked sufficient power. Alternatively, diversity
metrics may mask the appropriate signal and it is necessary to measure
the association at the level of microbial populations. Walters et al.
(8) demonstrated that Random Forest machine learning models were capable
of predicting obesity status within a single cohort, but did not attempt
to test the models on other cohorts. The purpose of this study was to
perform a meta-analysis of the association between differences in the
microbiome and obesity status by analyzing and applying a more
systematic and synthetic approach than was used previously.

\subsection{Methods}\label{methods}

\textbf{\emph{Literature Review and Study Inclusion.}} We followed the
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines to identify studies to include in our meta-analysis
(23). A detailed description of our selection process and the exact
search terms are provided in the Supplement and in figure 1. Briefly, we
searched PubMed for original research studies that involved studying
obesity and the human microbiome. The initial search yielded 187
studies. We identified number\_string{[}n{]} additional studies that
were not designed to explicitly test for an association between the
microbiome and obesity. We then manually curated the 196 studies to
select those studies that included BMI and sequence data. This yielded
10 eligible studies. An additional study was removed from our analysis
because no individuals in the study had a BMI over 30. Among the final
10 studies, 3 were from identified from our PubMed search (6, 15, 20), 5
were originally identified from the 9 studies that did not explicitly
investigate obesity but included BMI data (14, 18, 19, 24, 25), and two
datasets were used (21, 22) because these publications did not
specifically look for any metabolic or obesity conditions but had
control populations and enabled us to help mitigate against publication
biases associated with the bacterial microbiome and obesity. The
number\_string{[}n{]} studies are summarized in Tables 1 and 2.

\textbf{\emph{Sequence Analysis Pipeline.}} All sequence data were
publicly available and were downloaded from the NCBI Sequence Read
Archive, the European Nucleotide Archive, or the investigators' personal
website
(\url{https://gordonlab.wustl.edu/TurnbaughSE/_10/_09/STM/_2009.html}).
In total seven studies used 454 (6, 14, 15, 19, 20, 22, 25) and three
studies used Illumina sequencing (18, 21, 24). All of these studies used
amplification-based 16S rRNA gene sequencing. Among the studies that
sequenced the 16S rRNA gene, the researchers targeted the V1-V2 (19),
V1-V3 (14, 15, 20), V3-V5 (22, 25), V4 {[}(18); (21); {]}, and V3-4 (24)
regions. For those studies where multiple regions were sequenced, we
selected the region that corresponded to the largest number of subjects
(6, 25). We processed the 16S rRNA gene sequence data using a
standardized mothur pipeline. Briefly, our pipelines attempted to follow
previously recommended approaches for 454 and Illumina sequencing data
(26, 27). All sequences were screened for chimeras using UCHIME and
assigned to operational taxonomic units (OTUs) using the average
neighbor algorithm using a 3\% distance threshold (28, 29). All sequence
processing was performed using mothur (v.1.37.0) (30).

\textbf{\emph{Data Analysis.}} We split the overall meta-analysis into
three general strategies using R (3.3.0). First, we followed the
approach employed by Finucane et al (9) and Walters et al (8) where each
study was re-analyzed separately to identify associations between BMI
and the relative abundance of Bacteroidetes and Firmicutes, the ratio of
Bacteroidetes and Firmicutes relative abundances (B:F), Shannon
diversity, observed richness, and Shannon evenness. After each variable
was transformed to fit a normal distribution a two-tailed t-test was
performed for comparison of non-obese and obese individuals (i.e.~BMI
\textgreater{} 35.0). We performed a pooled analysis on these measured
variables using linear random effect models to correct for study effect
to asses differences on the combined dataset between non-obese and obese
groups using the lme4 (v.1.1-12) R package. Next, we compared the
community structure from non-obese and obese individuals using PERMANOVA
analysis of Bray-Curtis distance matrices. This analysis was performed
using the vegan (v.2.3-5) R package. For both analyses, the datasets
were rarefied (N=1000) so that each study within a study had the same
number of sequences. Second, for each study we partitioned the subjects
into a low or high group depending on whether their alpha diversity
metrics were below or above the median value for the study. The relative
risk (RR) was then calculated as the ratio of the number of obese
individuals in the low group to the number of obese individuals in the
high group. We then performed a Fisher exact-test to investigate whether
the RR was significantly different from 1.0 within each study and across
all of the studies using the epiR (0.9-77) and metafor (1.9-8) packages.
Third, we used the AUCRF (1.1) R package to generate Random Forest
models. For each study we developed models using either OTUs or
genus-level phylotypes. The quality of each model was assessed by
measuring the area under the curve (AUC) of the Receive Operating
Characteristic (ROC) using ten-fold cross validation. Because the
genus-level phylotype models were developed using a common reference, it
was possible to use one study's model (i.e.~the training set) to
classify the samples from the other studies (i.e.~the testing sets). The
optimum threshold for the training set was set as the probability
threshold that had the highest combined sensitivity and specificity.
This threshold was then used to calculate the accuracy of the model
applied to the test studies. To generate Receiver Operator
Characteristic (ROC) curves and calculate the accuracy of the models we
used the pROC (1.8) R package. Finally, we performed power and sample
number simulations for different effect sizes for each study using the
pwr (1.1-3) R package and base R functions. We also calculated the
actual sample size needed based on the effect size of each individual
study.

\textbf{\emph{Reproducible methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_Obesity_mBio_2016/}.

\subsection{Results}\label{results}

\textbf{\emph{Alpha diversity analysis.}} We calculated the Shannon
diversity index, richness, and Shannon evenness, the relative abundance
of \emph{Bacteroidetes} and \emph{Firmicutes}, and the ratio of their
relative abundance (B:F) for each sample. Once we transformed each of
the six alpha diversity metrics to make them normally distributed, we
used a t-test to identify significant associations between the alpha
diversity metric and whether an individual was obese for each of the ten
studies. The B:F and the relative abundance of \emph{Firmicutes} were
not significantly associated with obesity in any study. We identified 7
P-values less than 0.05: three studies indicated obese individuals had a
lower richness, two studies indicated a significantly lower diversity,
one study indicated a significantly lower evenness, and one study
indicated a significantly higher relative abundance of
\emph{Bacteroidetes} (Figures 2 and S1). These results largely match
those of the Walters and Finucane re-analysis studies. Interestingly,
although only two of the 10 studies observed the previously reported
association between lower diversity and obesity, the other studies
appeared to have the same trend, albeit the differences were not
statistically significant. We used a random effects linear model to
combine the studies using the study as the random effect and found
statistical support for decreased richness, evenness, and diversity
among obese individuals (all P\textless{}0.011). Although there was a
significant relationship between these metrics and diversity and obesity
status, the effect size was quite small. The obese individuals averaged
7.47\% lower richness, 0.88\% lower evenness, and 2.07\% lower
diversity. There were no significant associations when we pooled the
phylum-level metrics across studies. These results indicate that obese
individuals do have a statistically significant lower diversity than
non-obese individuals; however, it is questionable whether the
difference is biologically significant.

\textbf{\emph{Relative risk.}} Building upon the alpha diversity
analysis we calculated the relative risk of being obese based on whether
an individual's alpha diversity metrics were below or above the median
metric for that study. The results using relative risk largely matched
those of using the untransformed alpha diversity data. Across the
number\_string{[}n{]} studies and six metrics, the only significant
relative risk values were the richness, evenness, and diversity values
from the Goodrich study (Figures 3 and S2). Again, although the relative
risk values were not significant for other studies, the values tended to
be above one. When we pooled the data using a random effects model, the
relative risk associated with having a richness, evenness, or diversity
below the median for the population was significantly associated with
obesity (all P\textless{}0.0044). The relative risks associated with
alpha diversity were small. The relative risk of having a low richness
was 1.30 (1.13-1.49), low evenness was 1.20 (1.06-1.37), and low
diversity was 1.27 (1.09-1.48). There were no significant difference in
the phylum-level metrics. Again, the relative risk results indicate that
individuals with a lower richness, evenness, or diversity are at
statistically significant increased risk of being obese, it is
questionable whether that risk is biologically or clinically relevant.

\textbf{\emph{Beta diversity analysis.}} Following the approach used by
the Walters and Finucane re-analysis studies, for each dataset we
calculated a Bray-Curtis distance matrix to measure the difference in
the membership and structure of the individuals from each study. We then
used PERMANOVA to test for a significant differences between the
structure of non-obese and obese individuals. The Escobar, Goodrich, and
Turnbaugh datasets indicated a significant difference in community
structure (all P\textless{}0.05). Because it was not possible to
ascertain the directionality of the difference in community structure
nor perform a pooled analysis using studies that had non-overlapping 16S
rRNA gene sequence regions it is unclear whether these differences
reflect a broader, but perhaps small, shift in community structure
between non-obese and obese individuals.

\textbf{\emph{Development of a microbiome-based classifier of obesity.}}
The Walters re-analysis study suggested that it was possible to classify
individuals as being obese or non-obese based on the composition of
their microbiota. We repeated this analysis with additional datasets
using OTU and genus-level phylotype data. For each study we developed a
Random Forest machine learning model to classify individuals. Using
ten-fold cross validation, the observed AUC values varied between 0.52
and 0.69 indicating a relatively poor ability to classify individuals
(Figure 4A). So that we could test models on other datasets, we trained
models using genus-level phylotype data for each dataset. The the
observed AUC values for the models applied to the training datasets
varied between 0.51 and 0.65, again indicating a relatively poor ability
to classify individuals from the original dataset (Figure 4B). For each
model we identified the probability where the sum of the sensitivity and
specificity was the highest. We then used this to identify a threshold
to calculate the accuracy of the models when applied to the other
number\_string{[}n-1{]} datasets (Figure 5). Although there considerable
variation in accuracy values for each model, the median accuracy for
each model varied between 0.33 (Turnbaugh) and 0.65 (HMP) (median=0.57).
When we considered the number of samples, balance of non-obese and obese
individuals, and region within the 16S rRNA gene it was not possible to
identify factors that predictably affected model performance. The
ability to predict obesity status using the relative abundance of OTUs
and genera in the communities is only marginally better than random.
These results suggest that given the large diversity of microbiome
compositions it is difficult to identify a taxonomic signal that can be
associated with obesity.

\textbf{\emph{Power and Sample Size Estimate Simulations.}} The
inability to detect a difference between non-obese and obese individuals
could be due to the lack of a true effect or because the study had
insufficient statistical power to detect a difference because of
insufficient sampling, large interpersonal variation, and unbalanced
sampling of non-obese and obese individuals. To assess this, we
calculated the power to detect differences of 1, 5, 10, and 15\% in each
of the alpha diversity metrics using the sample sizes used in each of
the studies (Figures 6, S3-S8). Although there is no biological
rationale for these effect sizes, they represent a range that is
plausible. Only the Goodrich study had power greater than 0.80 to detect
a 5\% difference in Shannon diversity and six of the studies had enough
power to detect a 10\% difference (Figure 6). None of the studies had
sufficient power to detect a 15\% difference between B:F values (Figure
S5). In fact, the maximum power among any of the studies to detect a
15\% difference in B:F values was 0.25. Among the tests for relative
risk, none of the studies had sufficient power to detect a Cohen's d of
0.10 and only two studies had sufficient power to detect a Cohen's d of
0.15. We next estimated how many individuals would need to have been
sampled to have sufficient power to detect the four effect sizes
assuming the observed interpersonal variation from each study and
balanced sampling between the two groups. To detect a 1, 5, 10, and 15\%
difference in Shannon index, the median sampling effort per group was
approximately 3,400, 140, 35, and 16 individuals, respectively. To
detect a 1, 5, 10, and 15\% difference in B:F values, the median
sampling effort per group was approximately 160,000, 6,300, 1,600, and
700 individuals, respectively. To detect a 1, 5, 10, and 15\% difference
in relative risk values using Shannon diversity, the median sampling
effort per group was approximately 39,000, 1,500, 380, and 170
individuals, respectively. These estimates indicate that most microbiome
studies are underpowered to detect modest effect sizes using either
metric. In the case of obesity, the studies were underpowered to detect
the \texttt{signif(range\_effect\_size,\ 1){[}1{]}} to 6\% difference in
diversity that was observed across the studies.

\subsection{Discussion}\label{discussion}

Our meta-analysis helps to provide a clarity to the ongoing debate of
whether or not there are specific microbiome-based markers that can be
associated with obesity. We performed an extensive literature review of
the existing studies on the microbiome and obesity and perform a
meta-analysis on the studies that remained based on our inclusion and
exclusion criteria. By statistically pooling the data from ten studies,
we observed significant, but small, relationships between richness,
evenness, and diversity and obesity status as well as the relative risk
of being obese based on these metrics. We also generated Random Forest
machine learning models trained on each dataset and tested on the
remaining datasets. This analysis demonstrated that the ability to
reliably classify individuals as being obese based on the composition of
their microbiome is limited. Finally, we assessed the ability of each
study to detect defined differences in alpha diversity and observed that
most studies were underpowered to detect modest effect sizes.
Considering these datasets are among the largest published, it appears
that most microbiome studies are underpowered to detect differences in
alpha diversity.

Alpha diversity metrics are attractive because they distill a complex
dataset to a single value. For example, diversity is a measure of the
entropy in a community and integrates richness and evenness information.
Two communities with little taxonomic similarity can have the same
diversity. Among ecologists the relevance of these metrics is questioned
because it is difficult to ascribe a mechanistic interpretation to their
relationship with stability or disease. Regardless, the concept of a
biologically significant effect size needs to be developed among
microbiome researchers. Alternative metrics could include the ability to
detect a defined difference in the relative abundance of an OTU
representing a defined relative abundance. What makes for a biologically
significant difference or relative abundance is an important point that
has yet to be discussed in the microbiome field. The use of
operationally defined effect sizes should be adequate until it is
possible to decide upon an accepted practice.

By selecting a range of possible effect sizes, we were able to
demonstrate that most studies are underpowered to detect modest
differences in alpha diversity metrics and phylum-level relative
abundances. Several factors interact to limit the power of microbiome
studies. There is wide interpersonal variation in the diversity and
structure of the human microbiome. In addition, the common experimental
designs limit their power. As we observed, most of the studies included
in our analysis were unbalanced for the variable that we were interested
in. This was also true of those studies that originally sought to
identify associations with obesity. Even with a balanced design, we
showed that it was necessary to obtain approximately 140 and 6,300
sequences per sample to detect a 5\% difference in Shannon diversity or
B:F, respectively. It was interesting that these sample sizes agreed
across studies regardless of their sequencing method, region with in the
16S rRNA gene, or subject population (Figure 6). This suggests that
regardless of the treatment or category, these sample sizes represent a
good starting point for subject recruitment when using stool samples.
Unfortunately, few studies have been published with this level of
subject recruitment. This is troubling since the positive predictive
rate of a significant finding in an underpowered study is small leading
to results that cannot be reproduced (31). Future microbiome studies
should articulate the basis for their experimental design.

Two previous reviews (8, 9) have stated that there was not a consistent
association between alpha diversity and obesity; however, neither of
these studies made an attempt to pool the existing data together to try
and harness the additional power that this would give and they did not
assess whether the studies were sufficiently powered to detect a
difference. Our analysis also used 16S rRNA gene sequence data from ten
studies whereas the Finucane study used 16S rRNA gene sequence data from
3 studies (7, 10, 25) and a metagenomic study (32) and the Walters study
used 16S rRNA gene sequence data from 5 studies (10, 15, 19, 25, 33);
two studies were included in both analyses (10, 25). Our analysis
included 4 of these studies (10, 15, 19, 25) and excluded 3 of the
studies because they were too small (7), only utilized metagenomic data
(32), or used short single read Illumina HiSeq data that has a high
error rate making it untractable for \emph{de novo} OTU clustering (33).
The additional seven datasets were published after the two reviews were
performed and include datasets with more samples than were found in the
original studies. Our collection of ten studies allowed us to largely
use the same sequence analysis pipeline for all datasets and relied
heavily on the availability of public data and access to metadata that
included variables beyond the needs of the original study. To execute
this analysis, we created an automated data analysis pipeline, which can
be easily updated to add additional studies as they become available.
Similarly, it would be possible to adapt this pipeline to other body
sites and treatment or variables (e.g.~subject's sex or age).

Similar to our study, the Walters analysis (8), the authors generated
Random Forest machine learning models to differentiate between non-obese
and obese individuals. They obtained similar AUC values to our analysis;
however, they did not attempt to test these models on the other studies
in their analysis. When we performed the inter-dataset cross validation
the median accuracy across datasets was only 56.67\% indicating that the
models did a poor job when applied to other datasets. This could be due
to differences in subject populations and methods. Considering the
median AUC for models trained and tested on the same data with ten-fold
cross validation only varied between 0.51 and 0.65 and there was not a
strong signal in the alpha diversity data, we suspect that there is
insufficient signal to reliably classify individuals.

Although we failed to find an effect it is not realistic to necessarily
state that there is no microbiome impact on obesity. There is strong
evidence in murine models of obesity that the microbiome and level of
adiposity can be manipulated via genetic manipulation of the animal and
manipulation of the community through antibiotics or colonizing germ
free mice with diverse fecal material from human donors (5, 10--13).
These studies appear to conflict with the observations using human
subjects. Recalling the large interpersonal variation in the structure
of the microbiome, it is possible that each individual has their own
signatures of obesity. Alternatively, it could be that the involvement
of the microbiome in obesity is at the level of a common set of
metabolites that can be produced from different structures of the
microbiome.

\subsection{Acknowledgements}\label{acknowledgements}

The authors would like to thank Nielson Baxter and Shawn Whitefield for
their suggestions on the development of the manuscript. We are grateful
to the authors of the studies used in our meta-analysis who have made
their data publicly available or available to us directly. Without their
forethought studies such as this would not be possible. This work was
supported in part by funding from the National Institutes of Health to
PDS (U01AI2425501 and P30DK034933).

\newpage

\textbf{Table 1. Summary Demographics of Individuals used in the
Meta-analysis.}

\newpage

\textbf{Figure 1: PRISMA flow diagram of total records searched (34).}

\textbf{Figure 2: Individual and combined comparison of obese and
non-obese groups for Shannon diversity (A) and B:F (B).}

\textbf{Figure 3: Meta analysis of the relative risk of obesity based on
Shannon diversity (A) or B:F (B).}

\textbf{Figure 4: ROC curves for each study based on classification of
non-obese or obese groups using OTUs (A) or genus-level classification
(B).}

\textbf{Figure 5: Overall accuracy of each study to predict non-obese
and obese individuals based on that study's Random Forest machine
learning model applied to each of the other studies.}

\textbf{Figure 6: Power (A) and sample size simulations (B) for Shannon
diversity for differentiating between non-obese versus obese for effect
sizes of 1, 5, 10, and 15\%.} Power calculations use the sampling
distribution from the original studies and the sample size estimations
assume an equal amount of sampling from each treatment group.

\newpage

\textbf{Figure S1: Individual and Combined comparison of Obese and
Non-Obese groups Based on Evenness (A), Richness (B), or the Relative
Abundance of Bacteroidetes (C) and Firmictues (D).}

\textbf{Figure S2: Meta Analysis of the Relative Risk of Obesity Based
on Evenness (A), Richness (B), or the Relative Abundance of
Bacteroidetes (C) and Firmictues (D).}

\textbf{Figure S3: Power (A) and sample size simulations (B) for B:F for
differentiating between non-obese versus obese for effect sizes of 1, 5,
10, and 15\%.} Power calculations use the sampling distribution from the
original studies and the sample size estimations assume an equal amount
of sampling from each treatment group.

\textbf{Figure S4: Power (A) and sample size simulations (B) for
richness for differentiating between non-obese versus obese for effect
sizes of 1, 5, 10, and 15\%.} Power calculations use the sampling
distribution from the original studies and the sample size estimations
assume an equal amount of sampling from each treatment group.

\textbf{Figure S5: Power (A) and sample size simulations (B) for
evenness for differentiating between non-obese versus obese for effect
sizes of 1, 5, 10, and 15\%.} Power calculations use the sampling
distribution from the original studies and the sample size estimations
assume an equal amount of sampling from each treatment group.

\textbf{Figure S6: Power (A) and sample size simulations (B) for the
relative abundance of Bacteroidetes for differentiating between
non-obese versus obese for effect sizes of 1, 5, 10, and 15\%.} Power
calculations use the sampling distribution from the original studies and
the sample size estimations assume an equal amount of sampling from each
treatment group.

\textbf{Figure S7: Power (A) and sample size simulations (B) for the
relative abundance of Firmicutes for differentiating between non-obese
versus obese for effect sizes of 1, 5, 10, and 15\%.} Power calculations
use the sampling distribution from the original studies and the sample
size estimations assume an equal amount of sampling from each treatment
group.

\textbf{Figure S8: Power (A) and sample size simulations (B) for
relative risk of obesity based on Shannon diversity.} Power calculations
use the sampling distribution from the original studies and the sample
size estimations assume an equal amount of sampling from each treatment
group.

\newpage

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-ogdenux5fprevalenceux5f2014}{}
1. \textbf{Ogden CL}, \textbf{Carroll MD}, \textbf{Kit BK},
\textbf{Flegal KM}. 2014. Prevalence of childhood and adult obesity in
the United States, 2011-2012. JAMA \textbf{311}:806--814.
doi:\href{https://doi.org/10.1001/jama.2014.732}{10.1001/jama.2014.732}.

\hypertarget{ref-lichtashux5fbodyux5f2013}{}
2. \textbf{Lichtash CT}, \textbf{Cui J}, \textbf{Guo X}, \textbf{Chen
Y-DI}, \textbf{Hsueh WA}, \textbf{Rotter JI}, \textbf{Goodarzi MO}.
2013. Body adiposity index versus body mass index and other
anthropometric traits as correlates of cardiometabolic risk factors.
PloS One \textbf{8}:e65954.
doi:\href{https://doi.org/10.1371/journal.pone.0065954}{10.1371/journal.pone.0065954}.

\hypertarget{ref-braheux5fcanux5f2016}{}
3. \textbf{Brahe LK}, \textbf{Astrup A}, \textbf{Larsen LH}. 2016. Can
We Prevent Obesity-Related Metabolic Diseases by Dietary Modulation of
the Gut Microbiota? Advances in Nutrition (Bethesda, Md)
\textbf{7}:90--101.
doi:\href{https://doi.org/10.3945/an.115.010587}{10.3945/an.115.010587}.

\hypertarget{ref-drorux5fmicrobiotaux5f2016}{}
4. \textbf{Dror T}, \textbf{Dickstein Y}, \textbf{Dubourg G},
\textbf{Paul M}. 2016. Microbiota manipulation for weight change.
Microbial Pathogenesis.
doi:\href{https://doi.org/10.1016/j.micpath.2016.01.002}{10.1016/j.micpath.2016.01.002}.

\hypertarget{ref-leyux5fobesityux5f2005}{}
5. \textbf{Ley RE}, \textbf{Bäckhed F}, \textbf{Turnbaugh P},
\textbf{Lozupone CA}, \textbf{Knight RD}, \textbf{Gordon JI}. 2005.
Obesity alters gut microbial ecology. Proceedings of the National
Academy of Sciences of the United States of America
\textbf{102}:11070--11075.
doi:\href{https://doi.org/10.1073/pnas.0504978102}{10.1073/pnas.0504978102}.

\hypertarget{ref-turnbaughux5fcoreux5f2009}{}
6. \textbf{Turnbaugh PJ}, \textbf{Hamady M}, \textbf{Yatsunenko T},
\textbf{Cantarel BL}, \textbf{Duncan A}, \textbf{Ley RE}, \textbf{Sogin
ML}, \textbf{Jones WJ}, \textbf{Roe BA}, \textbf{Affourtit JP},
\textbf{Egholm M}, \textbf{Henrissat B}, \textbf{Heath AC},
\textbf{Knight R}, \textbf{Gordon JI}. 2009. A core gut microbiome in
obese and lean twins. Nature \textbf{457}:480--484.
doi:\href{https://doi.org/10.1038/nature07540}{10.1038/nature07540}.

\hypertarget{ref-leyux5fmicrobialux5f2006}{}
7. \textbf{Ley RE}, \textbf{Turnbaugh PJ}, \textbf{Klein S},
\textbf{Gordon JI}. 2006. Microbial ecology: Human gut microbes
associated with obesity. Nature \textbf{444}:1022--1023.
doi:\href{https://doi.org/10.1038/4441022a}{10.1038/4441022a}.

\hypertarget{ref-waltersux5fmeta-analysesux5f2014}{}
8. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
Meta-analyses of human gut microbes associated with obesity and IBD.
FEBS letters \textbf{588}:4223--4233.
doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.

\hypertarget{ref-finucaneux5ftaxonomicux5f2014}{}
9. \textbf{Finucane MM}, \textbf{Sharpton TJ}, \textbf{Laurent TJ},
\textbf{Pollard KS}. 2014. A taxonomic signature of obesity in the
microbiome? Getting to the guts of the matter. PloS One
\textbf{9}:e84689.
doi:\href{https://doi.org/10.1371/journal.pone.0084689}{10.1371/journal.pone.0084689}.

\hypertarget{ref-turnbaughux5fobesity-associatedux5f2006}{}
10. \textbf{Turnbaugh PJ}, \textbf{Ley RE}, \textbf{Mahowald MA},
\textbf{Magrini V}, \textbf{Mardis ER}, \textbf{Gordon JI}. 2006. An
obesity-associated gut microbiome with increased capacity for energy
harvest. Nature \textbf{444}:1027--31.
doi:\href{https://doi.org/10.1038/nature05414}{10.1038/nature05414}.

\hypertarget{ref-Koren2012}{}
11. \textbf{Koren O}, \textbf{Goodrich JK}, \textbf{Cullender TC},
\textbf{Spor A}, \textbf{Laitinen K}, \textbf{Bäckhed HK},
\textbf{Gonzalez A}, \textbf{Werner JJ}, \textbf{Angenent LT},
\textbf{Knight R}, \textbf{Bäckhed F}, \textbf{Isolauri E},
\textbf{Salminen S}, \textbf{Ley RE}. 2012. Host remodeling of the gut
microbiome and metabolic changes during pregnancy. Cell
\textbf{150}:470--480.
doi:\href{https://doi.org/10.1016/j.cell.2012.07.008}{10.1016/j.cell.2012.07.008}.

\hypertarget{ref-Cox2014}{}
12. \textbf{Cox LM}, \textbf{Yamanishi S}, \textbf{Sohn J},
\textbf{Alekseyenko AV}, \textbf{Leung JM}, \textbf{Cho I}, \textbf{Kim
SG}, \textbf{Li H}, \textbf{Gao Z}, \textbf{Mahana D}, \textbf{Rodriguez
JGZ}, \textbf{Rogers AB}, \textbf{Robine N}, \textbf{Loke P},
\textbf{Blaser MJ}. 2014. Altering the intestinal microbiota during a
critical developmental window has lasting metabolic consequences. Cell
\textbf{158}:705--721.
doi:\href{https://doi.org/10.1016/j.cell.2014.05.052}{10.1016/j.cell.2014.05.052}.

\hypertarget{ref-Mahana2016}{}
13. \textbf{Mahana D}, \textbf{Trent CM}, \textbf{Kurtz ZD},
\textbf{Bokulich NA}, \textbf{Battaglia T}, \textbf{Chung J},
\textbf{Müller CL}, \textbf{Li H}, \textbf{Bonneau RA}, \textbf{Blaser
MJ}. 2016. Antibiotic perturbation of the murine gut microbiome enhances
the adiposity, insulin resistance, and liver disease associated with
high-fat diet. Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0297-9}{10.1186/s13073-016-0297-9}.

\hypertarget{ref-rossux5f16sux5f2015}{}
14. \textbf{Ross MC}, \textbf{Muzny DM}, \textbf{McCormick JB},
\textbf{Gibbs RA}, \textbf{Fisher-Hoch SP}, \textbf{Petrosino JF}. 2015.
16S gut community of the Cameron County Hispanic Cohort. Microbiome
\textbf{3}:7.
doi:\href{https://doi.org/10.1186/s40168-015-0072-y}{10.1186/s40168-015-0072-y}.

\hypertarget{ref-zupancicux5fanalysisux5f2012}{}
15. \textbf{Zupancic ML}, \textbf{Cantarel BL}, \textbf{Liu Z},
\textbf{Drabek EF}, \textbf{Ryan KA}, \textbf{Cirimotich S},
\textbf{Jones C}, \textbf{Knight R}, \textbf{Walters WA},
\textbf{Knights D}, \textbf{Mongodin EF}, \textbf{Horenstein RB},
\textbf{Mitchell BD}, \textbf{Steinle N}, \textbf{Snitker S},
\textbf{Shuldiner AR}, \textbf{Fraser CM}. 2012. Analysis of the gut
microbiota in the old order Amish and its relation to the metabolic
syndrome. PloS One \textbf{7}:e43052.
doi:\href{https://doi.org/10.1371/journal.pone.0043052}{10.1371/journal.pone.0043052}.

\hypertarget{ref-namux5fcomparativeux5f2011}{}
16. \textbf{Nam Y-D}, \textbf{Jung M-J}, \textbf{Roh SW}, \textbf{Kim
M-S}, \textbf{Bae J-W}. 2011. Comparative analysis of Korean human gut
microbiota by barcoded pyrosequencing. PloS One \textbf{6}:e22109.
doi:\href{https://doi.org/10.1371/journal.pone.0022109}{10.1371/journal.pone.0022109}.

\hypertarget{ref-arumugamux5fenterotypesux5f2011}{}
17. \textbf{Arumugam M}, \textbf{Raes J}, \textbf{Pelletier E},
\textbf{Le Paslier D}, \textbf{Yamada T}, \textbf{Mende DR},
\textbf{Fernandes GR}, \textbf{Tap J}, \textbf{Bruls T}, \textbf{Batto
J-M}, \textbf{Bertalan M}, \textbf{Borruel N}, \textbf{Casellas F},
\textbf{Fernandez L}, \textbf{Gautier L}, \textbf{Hansen T},
\textbf{Hattori M}, \textbf{Hayashi T}, \textbf{Kleerebezem M},
\textbf{Kurokawa K}, \textbf{Leclerc M}, \textbf{Levenez F},
\textbf{Manichanh C}, \textbf{Nielsen HB}, \textbf{Nielsen T},
\textbf{Pons N}, \textbf{Poulain J}, \textbf{Qin J},
\textbf{Sicheritz-Ponten T}, \textbf{Tims S}, \textbf{Torrents D},
\textbf{Ugarte E}, \textbf{Zoetendal EG}, \textbf{Wang J},
\textbf{Guarner F}, \textbf{Pedersen O}, \textbf{Vos WM de},
\textbf{Brunak S}, \textbf{Doré J}, \textbf{MetaHIT Consortium},
\textbf{Antolín M}, \textbf{Artiguenave F}, \textbf{Blottiere HM},
\textbf{Almeida M}, \textbf{Brechot C}, \textbf{Cara C},
\textbf{Chervaux C}, \textbf{Cultrone A}, \textbf{Delorme C},
\textbf{Denariaz G}, \textbf{Dervyn R}, \textbf{Foerstner KU},
\textbf{Friss C}, \textbf{Guchte M van de}, \textbf{Guedon E},
\textbf{Haimet F}, \textbf{Huber W}, \textbf{Hylckama-Vlieg J van},
\textbf{Jamet A}, \textbf{Juste C}, \textbf{Kaci G}, \textbf{Knol J},
\textbf{Lakhdari O}, \textbf{Layec S}, \textbf{Le Roux K},
\textbf{Maguin E}, \textbf{Mérieux A}, \textbf{Melo Minardi R},
\textbf{M'rini C}, \textbf{Muller J}, \textbf{Oozeer R},
\textbf{Parkhill J}, \textbf{Renault P}, \textbf{Rescigno M},
\textbf{Sanchez N}, \textbf{Sunagawa S}, \textbf{Torrejon A},
\textbf{Turner K}, \textbf{Vandemeulebrouck G}, \textbf{Varela E},
\textbf{Winogradsky Y}, \textbf{Zeller G}, \textbf{Weissenbach J},
\textbf{Ehrlich SD}, \textbf{Bork P}. 2011. Enterotypes of the human gut
microbiome. Nature \textbf{473}:174--180.
doi:\href{https://doi.org/10.1038/nature09944}{10.1038/nature09944}.

\hypertarget{ref-goodrichux5fhumanux5f2014}{}
18. \textbf{Goodrich JK}, \textbf{Waters JL}, \textbf{Poole AC},
\textbf{Sutter JL}, \textbf{Koren O}, \textbf{Blekhman R},
\textbf{Beaumont M}, \textbf{Van Treuren W}, \textbf{Knight R},
\textbf{Bell JT}, \textbf{Spector TD}, \textbf{Clark AG}, \textbf{Ley
RE}. 2014. Human genetics shape the gut microbiome. Cell
\textbf{159}:789--799.
doi:\href{https://doi.org/10.1016/j.cell.2014.09.053}{10.1016/j.cell.2014.09.053}.

\hypertarget{ref-wuux5flinkingux5f2011}{}
19. \textbf{Wu GD}, \textbf{Chen J}, \textbf{Hoffmann C},
\textbf{Bittinger K}, \textbf{Chen Y-Y}, \textbf{Keilbaugh SA},
\textbf{Bewtra M}, \textbf{Knights D}, \textbf{Walters WA},
\textbf{Knight R}, \textbf{Sinha R}, \textbf{Gilroy E}, \textbf{Gupta
K}, \textbf{Baldassano R}, \textbf{Nessel L}, \textbf{Li H},
\textbf{Bushman FD}, \textbf{Lewis JD}. 2011. Linking long-term dietary
patterns with gut microbial enterotypes. Science (New York, NY)
\textbf{334}:105--108.
doi:\href{https://doi.org/10.1126/science.1208344}{10.1126/science.1208344}.

\hypertarget{ref-escobarux5fgutux5f2014}{}
20. \textbf{Escobar JS}, \textbf{Klotz B}, \textbf{Valdes BE},
\textbf{Agudelo GM}. 2014. The gut microbiota of Colombians differs from
that of Americans, Europeans and Asians. BMC microbiology
\textbf{14}:311.
doi:\href{https://doi.org/10.1186/s12866-014-0311-6}{10.1186/s12866-014-0311-6}.

\hypertarget{ref-baxterux5fmicrobiota-basedux5f2016}{}
21. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-schubertux5fmicrobiomeux5f2014}{}
22. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
distinguish patients with Clostridium difficile infection and non-C.
difficile-associated diarrhea from healthy controls. mBio
\textbf{5}:e01021--01014.
doi:\href{https://doi.org/10.1128/mBio.01021-14}{10.1128/mBio.01021-14}.

\hypertarget{ref-moherux5fpreferredux5f2010}{}
23. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
\textbf{Altman DG}, \textbf{PRISMA Group}. 2010. Preferred reporting
items for systematic reviews and meta-analyses: The PRISMA statement.
International Journal of Surgery (London, England) \textbf{8}:336--341.
doi:\href{https://doi.org/10.1016/j.ijsu.2010.02.007}{10.1016/j.ijsu.2010.02.007}.

\hypertarget{ref-zeeviux5fpersonalizedux5f2015}{}
24. \textbf{Zeevi D}, \textbf{Korem T}, \textbf{Zmora N},
\textbf{Israeli D}, \textbf{Rothschild D}, \textbf{Weinberger A},
\textbf{Ben-Yacov O}, \textbf{Lador D}, \textbf{Avnit-Sagi T},
\textbf{Lotan-Pompan M}, \textbf{Suez J}, \textbf{Mahdi JA},
\textbf{Matot E}, \textbf{Malka G}, \textbf{Kosower N}, \textbf{Rein M},
\textbf{Zilberman-Schapira G}, \textbf{Dohnalová L},
\textbf{Pevsner-Fischer M}, \textbf{Bikovsky R}, \textbf{Halpern Z},
\textbf{Elinav E}, \textbf{Segal E}. 2015. Personalized Nutrition by
Prediction of Glycemic Responses. Cell \textbf{163}:1079--1094.
doi:\href{https://doi.org/10.1016/j.cell.2015.11.001}{10.1016/j.cell.2015.11.001}.

\hypertarget{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{}
25. \textbf{Human Microbiome Project Consortium}. 2012. Structure,
function and diversity of the healthy human microbiome. Nature
\textbf{486}:207--214.
doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.

\hypertarget{ref-Kozichux5f2013}{}
26. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq Illumina sequencing platform.
Applied and environmental microbiology \textbf{79}:5112--5120.

\hypertarget{ref-Schloss2011}{}
27. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
Reducing the effects of PCR amplification and sequencing artifacts on
16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.

\hypertarget{ref-Edgar2011}{}
28. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.

\hypertarget{ref-Westcott2015}{}
29. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
methods outperform reference-based methods for assigning 16S rRNA gene
sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.

\hypertarget{ref-mothur}{}
30. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{others}. 2009. Introducing mothur:
open-source, platform-independent, community-supported software for
describing and comparing microbial communities. Applied and
environmental microbiology \textbf{75}:7537--7541.

\hypertarget{ref-Ioannidis2005}{}
31. \textbf{Ioannidis JPA}. 2005. Why most published research findings
are false. PLoS Med \textbf{2}:e124.
doi:\href{https://doi.org/10.1371/journal.pmed.0020124}{10.1371/journal.pmed.0020124}.

\hypertarget{ref-Qin2010}{}
32. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
\textbf{Burgdorf KS}, \textbf{Manichanh C}, \textbf{Nielsen T},
\textbf{Pons N}, \textbf{Levenez F}, \textbf{Yamada T}, \textbf{Mende
DR}, \textbf{Li J}, \textbf{Xu J}, \textbf{Li S}, \textbf{Li D},
\textbf{Cao J}, \textbf{Wang B}, \textbf{Liang H}, \textbf{Zheng H},
\textbf{Xie Y}, \textbf{Tap J}, \textbf{Lepage P}, \textbf{Bertalan M},
\textbf{Batto J-M}, \textbf{Hansen T}, \textbf{Paslier DL},
\textbf{Linneberg A}, \textbf{Nielsen HB}, \textbf{Pelletier E},
\textbf{Renault P}, \textbf{Sicheritz-Ponten T}, \textbf{Turner K},
\textbf{Zhu H}, \textbf{Yu C}, \textbf{Li S}, \textbf{Jian M},
\textbf{Zhou Y}, \textbf{Li Y}, \textbf{Zhang X}, \textbf{Li S},
\textbf{Qin N}, \textbf{Yang H}, \textbf{Wang J}, \textbf{Brunak S},
\textbf{Doré J}, \textbf{Guarner F}, \textbf{Kristiansen K},
\textbf{Pedersen O}, \textbf{Parkhill J}, \textbf{Weissenbach J},
\textbf{Antolin M}, \textbf{Artiguenave F}, \textbf{Blottiere H},
\textbf{Borruel N}, \textbf{Bruls T}, \textbf{Casellas F},
\textbf{Chervaux C}, \textbf{Cultrone A}, \textbf{Delorme C},
\textbf{Denariaz G}, \textbf{Dervyn R}, \textbf{Forte M}, \textbf{Friss
C}, \textbf{Guchte M van de}, \textbf{Guedon E}, \textbf{Haimet F},
\textbf{Jamet A}, \textbf{Juste C}, \textbf{Kaci G}, \textbf{Kleerebezem
M}, \textbf{Knol J}, \textbf{Kristensen M}, \textbf{Layec S},
\textbf{Roux KL}, \textbf{Leclerc M}, \textbf{Maguin E}, \textbf{Minardi
RM}, \textbf{Oozeer R}, \textbf{Rescigno M}, \textbf{Sanchez N},
\textbf{Tims S}, \textbf{Torrejon T}, \textbf{Varela E}, \textbf{Vos W
de}, \textbf{Winogradsky Y}, \textbf{Zoetendal E}, \textbf{Bork P},
\textbf{Ehrlich SD}, \textbf{Wang J}. 2010. A human gut microbial gene
catalogue established by metagenomic sequencing. Nature
\textbf{464}:59--65.
doi:\href{https://doi.org/10.1038/nature08821}{10.1038/nature08821}.

\hypertarget{ref-yatsunenkoux5fhumanux5f2012}{}
33. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
\textbf{Trehan I}, \textbf{Dominguez-Bello MG}, \textbf{Contreras M},
\textbf{Magris M}, \textbf{Hidalgo G}, \textbf{Baldassano RN},
\textbf{Anokhin AP}, \textbf{Heath AC}, \textbf{Warner B},
\textbf{Reeder J}, \textbf{Kuczynski J}, \textbf{Caporaso JG},
\textbf{Lozupone CA}, \textbf{Lauber C}, \textbf{Clemente JC},
\textbf{Knights D}, \textbf{Knight R}, \textbf{Gordon JI}. 2012. Human
gut microbiome viewed across age and geography. Nature
\textbf{486}:222--227.
doi:\href{https://doi.org/10.1038/nature11053}{10.1038/nature11053}.

\hypertarget{ref-Moher2009}{}
34. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
\textbf{Altman DG}. 2009. Preferred reporting items for systematic
reviews and meta-analyses: The PRISMA statement. PLoS Med
\textbf{6}:e1000097.
doi:\href{https://doi.org/10.1371/journal.pmed.1000097}{10.1371/journal.pmed.1000097}.

\end{document}

### 2016-06-03 code/make_table1.R
Bring in Table 1

@@ -1,147 +1,67 @@
-# Table 1 Creation 
-# Make all the measurements for the table here so it just needs to be read into
-# the Rmd.
+capwords <- function(s, strict = FALSE) {
+		s <- as.character(s)
+		cap <- function(s) paste(toupper(substring(s, 1, 1)),
+									{s <- substring(s, 2); if(strict) tolower(s) else s},
+														 sep = "", collapse = " " )
+		sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
+}
 
+make_study_label <- function(dataset){
+	study <- ifelse(dataset=="hmp", "HMP", capwords(dataset))
+}
 
-# Read in necessary tables
-meta.bax <- read.table("../data/baxter/baxter.metadata", header = T)
-meta.esc <- read.table("../data/escobar/escobar.metadata", header = T)
-meta.goo <- read.table("../data/goodrich/goodrich.metadata", header = T)
-meta.hmp <- read.table("../data/hmp/hmp.metadata", header = T)
-meta.ros <- read.table("../data/ross/ross.metadata", header = T)
-meta.sch <- read.table("../data/schubert/schubert.metadata", header = T)
-meta.tur <- read.table("../data/turnbaugh/turnbaugh.metadata", header = T)
-meta.wu <- read.table("../data/wu/wu.metadata", header = T)
-meta.zee <- read.table("../data/zeevi/zeevi.metadata", header = T)
-meta.zup <- read.table("../data/zupancic/zupancic.metadata", header = T)
-beta_tests.summary <- read.table("../data/process/beta_tests.summary", 
-                                 header=T, row.names = 1)
+format_p <- function(p_value){
+	char_p <- format(round(p_value, 3), nsmall=3L)
 
-# Create data table entries
-Study <- as.character(rownames(beta_tests.summary))
+	if(p_value < 0.001){
+		char_p <- "<0.001"
+	}
 
-Age <- round(c(mean(meta.bax$age), mean(meta.esc$age), mean(meta.hmp$age), 
-               mean(meta.ros$age), mean(meta.sch$age), mean(meta.tur$age), 
-               mean(meta.wu$age), mean(as.numeric(as.character(meta.zup$age)), 
-                                       na.rm=TRUE), mean(meta.goo$age), 
[…]

### 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
Render manuscript

@@
     \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
   \newcommand{\euro}{€}
@@
 \usepackage[margin=1.0in]{geometry}
-\usepackage{hyperref}
-\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
-\hypersetup{unicode=true,
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
             pdftitle={Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome},
-            pdfborder={0 0 0},
-            breaklinks=true}
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
 \urlstyle{same}  % don't use monospace font for urls
@@
 
+\hyperdef{}{references}{\label{references}}
 \subsection*{References}\label{references}
@@
 
-\hypertarget{refs}{}
-\hypertarget{ref-ogdenux5fprevalenceux5f2014}{}
+\hyperdef{}{ref-ogdenux5fprevalenceux5f2014}{\label{ref-ogdenux5fprevalenceux5f2014}}
 1. \textbf{Ogden CL}, \textbf{Carroll MD}, \textbf{Kit BK},
@@
 the United States, 2011-2012. JAMA \textbf{311}:806--814.
-doi:\href{https://doi.org/10.1001/jama.2014.732}{10.1001/jama.2014.732}.
+doi:\url{http://doi.org/10.1001/jama.2014.732}.
 
-\hypertarget{ref-lichtashux5fbodyux5f2013}{}
+\hyperdef{}{ref-lichtashux5fbodyux5f2013}{\label{ref-lichtashux5fbodyux5f2013}}
 2. \textbf{Lichtash CT}, \textbf{Cui J}, \textbf{Guo X}, \textbf{Chen
@@
 PloS One \textbf{8}:e65954.
-doi:\href{https://doi.org/10.1371/journal.pone.0065954}{10.1371/journal.pone.0065954}.
+doi:\url{http://doi.org/10.1371/journal.pone.0065954}.
 
-\hypertarget{ref-braheux5fcanux5f2016}{}
+\hyperdef{}{ref-braheux5fcanux5f2016}{\label{ref-braheux5fcanux5f2016}}
 3. \textbf{Brahe LK}, \textbf{Astrup A}, \textbf{Larsen LH}. 2016. Can
@@
 the Gut Microbiota? Advances in Nutrition (Bethesda, Md)
-\textbf{7}:90--101.
-doi:\href{https://doi.org/10.3945/an.115.010587}{10.3945/an.115.010587}.
+\textbf{7}:90--101. doi:\url{http://doi.org/10.3945/an.115.010587}.
 
-\hypertarget{ref-drorux5fmicrobiotaux5f2016}{}
+\hyperdef{}{ref-drorux5fmicrobiotaux5f2016}{\label{ref-drorux5fmicrobiotaux5f2016}}
 4. \textbf{Dror T}, \textbf{Dickstein Y}, \textbf{Dubourg G},
@@
 Microbial Pathogenesis.
-doi:\href{https://doi.org/10.1016/j.micpath.2016.01.002}{10.1016/j.micpath.2016.01.002}.
+doi:\url{http://doi.org/10.1016/j.micpath.2016.01.002}.
 
-\hypertarget{ref-leyux5fobesityux5f2005}{}
+\hyperdef{}{ref-leyux5fobesityux5f2005}{\label{ref-leyux5fobesityux5f2005}}
 5. \textbf{Ley RE}, \textbf{Bäckhed F}, \textbf{Turnbaugh P},
@@
 \textbf{102}:11070--11075.
-doi:\href{https://doi.org/10.1073/pnas.0504978102}{10.1073/pnas.0504978102}.
+doi:\url{http://doi.org/10.1073/pnas.0504978102}.
 
-\hypertarget{ref-turnbaughux5fcoreux5f2009}{}
+\hyperdef{}{ref-turnbaughux5fcoreux5f2009}{\label{ref-turnbaughux5fcoreux5f2009}}
 6. \textbf{Turnbaugh PJ}, \textbf{Hamady M}, \textbf{Yatsunenko T},
@@
 obese and lean twins. Nature \textbf{457}:480--484.
-doi:\href{https://doi.org/10.1038/nature07540}{10.1038/nature07540}.
+doi:\url{http://doi.org/10.1038/nature07540}.
 
-\hypertarget{ref-leyux5fmicrobialux5f2006}{}
+\hyperdef{}{ref-leyux5fmicrobialux5f2006}{\label{ref-leyux5fmicrobialux5f2006}}
 7. \textbf{Ley RE}, \textbf{Turnbaugh PJ}, \textbf{Klein S},
@@
 associated with obesity. Nature \textbf{444}:1022--1023.
-doi:\href{https://doi.org/10.1038/4441022a}{10.1038/4441022a}.
+doi:\url{http://doi.org/10.1038/4441022a}.
 
-\hypertarget{ref-waltersux5fmeta-analysesux5f2014}{}
+\hyperdef{}{ref-waltersux5fmeta-analysesux5f2014}{\label{ref-waltersux5fmeta-analysesux5f2014}}
 8. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
@@
 FEBS letters \textbf{588}:4223--4233.
-doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.
+doi:\url{http://doi.org/10.1016/j.febslet.2014.09.039}.
 
-\hypertarget{ref-finucaneux5ftaxonomicux5f2014}{}
+\hyperdef{}{ref-finucaneux5ftaxonomicux5f2014}{\label{ref-finucaneux5ftaxonomicux5f2014}}
 9. \textbf{Finucane MM}, \textbf{Sharpton TJ}, \textbf{Laurent TJ},
@@
 \textbf{9}:e84689.
-doi:\href{https://doi.org/10.1371/journal.pone.0084689}{10.1371/journal.pone.0084689}.
+doi:\url{http://doi.org/10.1371/journal.pone.0084689}.
 
-\hypertarget{ref-turnbaughux5fobesity-associatedux5f2006}{}
+\hyperdef{}{ref-turnbaughux5fobesity-associatedux5f2006}{\label{ref-turnbaughux5fobesity-associatedux5f2006}}
 10. \textbf{Turnbaugh PJ}, \textbf{Ley RE}, \textbf{Mahowald MA},
@@
 harvest. Nature \textbf{444}:1027--31.
-doi:\href{https://doi.org/10.1038/nature05414}{10.1038/nature05414}.
+doi:\url{http://doi.org/10.1038/nature05414}.
 
-\hypertarget{ref-Koren2012}{}
+\hyperdef{}{ref-Koren2012}{\label{ref-Koren2012}}
 11. \textbf{Koren O}, \textbf{Goodrich JK}, \textbf{Cullender TC},
@@
 \textbf{150}:470--480.
-doi:\href{https://doi.org/10.1016/j.cell.2012.07.008}{10.1016/j.cell.2012.07.008}.
+doi:\url{http://doi.org/10.1016/j.cell.2012.07.008}.
 
-\hypertarget{ref-Cox2014}{}
+\hyperdef{}{ref-Cox2014}{\label{ref-Cox2014}}
 12. \textbf{Cox LM}, \textbf{Yamanishi S}, \textbf{Sohn J},
@@
 \textbf{158}:705--721.
-doi:\href{https://doi.org/10.1016/j.cell.2014.05.052}{10.1016/j.cell.2014.05.052}.
+doi:\url{http://doi.org/10.1016/j.cell.2014.05.052}.
 
-\hypertarget{ref-Mahana2016}{}
+\hyperdef{}{ref-Mahana2016}{\label{ref-Mahana2016}}
 13. \textbf{Mahana D}, \textbf{Trent CM}, \textbf{Kurtz ZD},
@@
 high-fat diet. Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0297-9}{10.1186/s13073-016-0297-9}.
+doi:\url{http://doi.org/10.1186/s13073-016-0297-9}.
 
-\hypertarget{ref-rossux5f16sux5f2015}{}
+\hyperdef{}{ref-rossux5f16sux5f2015}{\label{ref-rossux5f16sux5f2015}}
 14. \textbf{Ross MC}, \textbf{Muzny DM}, \textbf{McCormick JB},
@@
 16S gut community of the Cameron County Hispanic Cohort. Microbiome
-\textbf{3}:7.
-doi:\href{https://doi.org/10.1186/s40168-015-0072-y}{10.1186/s40168-015-0072-y}.
+\textbf{3}:7. doi:\url{http://doi.org/10.1186/s40168-015-0072-y}.
 
-\hypertarget{ref-zupancicux5fanalysisux5f2012}{}
+\hyperdef{}{ref-zupancicux5fanalysisux5f2012}{\label{ref-zupancicux5fanalysisux5f2012}}
 15. \textbf{Zupancic ML}, \textbf{Cantarel BL}, \textbf{Liu Z},
@@
 syndrome. PloS One \textbf{7}:e43052.
-doi:\href{https://doi.org/10.1371/journal.pone.0043052}{10.1371/journal.pone.0043052}.
+doi:\url{http://doi.org/10.1371/journal.pone.0043052}.
 
-\hypertarget{ref-namux5fcomparativeux5f2011}{}
+\hyperdef{}{ref-namux5fcomparativeux5f2011}{\label{ref-namux5fcomparativeux5f2011}}
 16. \textbf{Nam Y-D}, \textbf{Jung M-J}, \textbf{Roh SW}, \textbf{Kim
@@
 microbiota by barcoded pyrosequencing. PloS One \textbf{6}:e22109.
-doi:\href{https://doi.org/10.1371/journal.pone.0022109}{10.1371/journal.pone.0022109}.
+doi:\url{http://doi.org/10.1371/journal.pone.0022109}.
 
-\hypertarget{ref-arumugamux5fenterotypesux5f2011}{}
+\hyperdef{}{ref-arumugamux5fenterotypesux5f2011}{\label{ref-arumugamux5fenterotypesux5f2011}}
 17. \textbf{Arumugam M}, \textbf{Raes J}, \textbf{Pelletier E},
@@
 microbiome. Nature \textbf{473}:174--180.
-doi:\href{https://doi.org/10.1038/nature09944}{10.1038/nature09944}.
+doi:\url{http://doi.org/10.1038/nature09944}.
 
-\hypertarget{ref-goodrichux5fhumanux5f2014}{}
+\hyperdef{}{ref-goodrichux5fhumanux5f2014}{\label{ref-goodrichux5fhumanux5f2014}}
 18. \textbf{Goodrich JK}, \textbf{Waters JL}, \textbf{Poole AC},
@@
 \textbf{159}:789--799.
-doi:\href{https://doi.org/10.1016/j.cell.2014.09.053}{10.1016/j.cell.2014.09.053}.
+doi:\url{http://doi.org/10.1016/j.cell.2014.09.053}.
 
-\hypertarget{ref-wuux5flinkingux5f2011}{}
+\hyperdef{}{ref-wuux5flinkingux5f2011}{\label{ref-wuux5flinkingux5f2011}}
 19. \textbf{Wu GD}, \textbf{Chen J}, \textbf{Hoffmann C},
@@
 patterns with gut microbial enterotypes. Science (New York, NY)
-\textbf{334}:105--108.
-doi:\href{https://doi.org/10.1126/science.1208344}{10.1126/science.1208344}.
+\textbf{334}:105--108. doi:\url{http://doi.org/10.1126/science.1208344}.
 
-\hypertarget{ref-escobarux5fgutux5f2014}{}
+\hyperdef{}{ref-escobarux5fgutux5f2014}{\label{ref-escobarux5fgutux5f2014}}
 20. \textbf{Escobar JS}, \textbf{Klotz B}, \textbf{Valdes BE},
@@
 that of Americans, Europeans and Asians. BMC microbiology
-\textbf{14}:311.
-doi:\href{https://doi.org/10.1186/s12866-014-0311-6}{10.1186/s12866-014-0311-6}.
+\textbf{14}:311. doi:\url{http://doi.org/10.1186/s12866-014-0311-6}.
 
-\hypertarget{ref-baxterux5fmicrobiota-basedux5f2016}{}
+\hyperdef{}{ref-baxterux5fmicrobiota-basedux5f2016}{\label{ref-baxterux5fmicrobiota-basedux5f2016}}
 21. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
@@
 Genome Medicine \textbf{8}:37.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+doi:\url{http://doi.org/10.1186/s13073-016-0290-3}.
 
-\hypertarget{ref-schubertux5fmicrobiomeux5f2014}{}
+\hyperdef{}{ref-schubertux5fmicrobiomeux5f2014}{\label{ref-schubertux5fmicrobiomeux5f2014}}
 22. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
@@
 \textbf{5}:e01021--01014.
-doi:\href{https://doi.org/10.1128/mBio.01021-14}{10.1128/mBio.01021-14}.
+doi:\url{http://doi.org/10.1128/mBio.01021-14}.
 
-\hypertarget{ref-moherux5fpreferredux5f2010}{}
+\hyperdef{}{ref-moherux5fpreferredux5f2010}{\label{ref-moherux5fpreferredux5f2010}}
 23. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
@@
 International Journal of Surgery (London, England) \textbf{8}:336--341.
-doi:\href{https://doi.org/10.1016/j.ijsu.2010.02.007}{10.1016/j.ijsu.2010.02.007}.
+doi:\url{http://doi.org/10.1016/j.ijsu.2010.02.007}.
 
-\hypertarget{ref-zeeviux5fpersonalizedux5f2015}{}
+\hyperdef{}{ref-zeeviux5fpersonalizedux5f2015}{\label{ref-zeeviux5fpersonalizedux5f2015}}
 24. \textbf{Zeevi D}, \textbf{Korem T}, \textbf{Zmora N},
@@
 Prediction of Glycemic Responses. Cell \textbf{163}:1079--1094.
-doi:\href{https://doi.org/10.1016/j.cell.2015.11.001}{10.1016/j.cell.2015.11.001}.
+doi:\url{http://doi.org/10.1016/j.cell.2015.11.001}.
 
-\hypertarget{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{}
+\hyperdef{}{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{\label{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}}
 25. \textbf{Human Microbiome Project Consortium}. 2012. Structure,
 function and diversity of the healthy human microbiome. Nature
-\textbf{486}:207--214.
-doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
+\textbf{486}:207--214. doi:\url{http://doi.org/10.1038/nature11234}.
 
-\hypertarget{ref-Kozichux5f2013}{}
+\hyperdef{}{ref-Kozichux5f2013}{\label{ref-Kozichux5f2013}}
 26. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
@@
 
-\hypertarget{ref-Schloss2011}{}
+\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
 27. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
@@
 16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
-doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
+doi:\url{http://doi.org/10.1371/journal.pone.0027310}.
 
-\hypertarget{ref-Edgar2011}{}
+\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
 28. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
@@
 and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
-doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
+doi:\url{http://doi.org/10.1093/bioinformatics/btr381}.
 
-\hypertarget{ref-Westcott2015}{}
+\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
 29. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
@@
 sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
+doi:\url{http://doi.org/10.7717/peerj.1487}.
 
-\hypertarget{ref-mothur}{}
+\hyperdef{}{ref-mothur}{\label{ref-mothur}}
 30. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
@@
 
-\hypertarget{ref-Ioannidis2005}{}
+\hyperdef{}{ref-Ioannidis2005}{\label{ref-Ioannidis2005}}
 31. \textbf{Ioannidis JPA}. 2005. Why most published research findings
 are false. PLoS Med \textbf{2}:e124.
-doi:\href{https://doi.org/10.1371/journal.pmed.0020124}{10.1371/journal.pmed.0020124}.
+doi:\url{http://doi.org/10.1371/journal.pmed.0020124}.
 
-\hypertarget{ref-Qin2010}{}
+\hyperdef{}{ref-Qin2010}{\label{ref-Qin2010}}
 32. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
@@
 catalogue established by metagenomic sequencing. Nature
-\textbf{464}:59--65.
-doi:\href{https://doi.org/10.1038/nature08821}{10.1038/nature08821}.
+\textbf{464}:59--65. doi:\url{http://doi.org/10.1038/nature08821}.
 
-\hypertarget{ref-yatsunenkoux5fhumanux5f2012}{}
+\hyperdef{}{ref-yatsunenkoux5fhumanux5f2012}{\label{ref-yatsunenkoux5fhumanux5f2012}}
 33. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
@@
 gut microbiome viewed across age and geography. Nature
-\textbf{486}:222--227.
-doi:\href{https://doi.org/10.1038/nature11053}{10.1038/nature11053}.
+\textbf{486}:222--227. doi:\url{http://doi.org/10.1038/nature11053}.
 
-\hypertarget{ref-Moher2009}{}
+\hyperdef{}{ref-Moher2009}{\label{ref-Moher2009}}
 34. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
@@
 \textbf{6}:e1000097.
-doi:\href{https://doi.org/10.1371/journal.pmed.1000097}{10.1371/journal.pmed.1000097}.
+doi:\url{http://doi.org/10.1371/journal.pmed.1000097}.
 

### 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
Echo output from building table

@@
 
+\begin{verbatim}
+## 
+## > capwords <- function(s, strict = FALSE) {
+## +     s <- as.character(s)
+## +     cap <- function(s) paste(toupper(substring(s, 1, 1)), {
+## +         s <- su .... [TRUNCATED] 
+## 
+## > make_study_label <- function(dataset) {
+## +     study <- ifelse(dataset == "hmp", "HMP", capwords(dataset))
+## + }
+## 
+## > format_p <- function(p_value) {
+## +     char_p <- format(round(p_value, 3), nsmall = 3)
+## +     if (p_value < 0.001) {
+## +         char_p <- "<0.001"
+## +    .... [TRUNCATED] 
+## 
+## > get_study_summary <- function(study, beta = beta_summary) {
+## +     metadata_file <- paste0("data/", study, "/", study, ".metadata")
+## +     metadata <- .... [TRUNCATED] 
+## 
+## > beta_summary <- read.table("data/process/beta_tests.summary", 
+## +     header = T, row.names = 1)
+## 
+## > datasets <- sort(rownames(beta_summary))
+## 
+## > study_summary <- t(sapply(datasets, get_study_summary))
+## 
+## > kable(study_summary, row.names = FALSE, col.names = c("Study", 
+## +     "Subjects (N)", "Obese (%)", "Average BMI (Min-Max)", "Female (%)", 
+## +     "Av ..." ... [TRUNCATED] 
+## 
+## 
+## Study        Subjects (N)    Obese (%)    Average BMI (Min-Max)    Female (%)    Average Age (Min-Max)    Non-Hispanic White (%)    AMOVA (P-value) 
+## ----------  --------------  -----------  -----------------------  ------------  -----------------------  ------------------------  -----------------
+## Baxter           172           27.3         27.0 (17.5-46.9)          64.5         54.3 (29.0-80.0)                87.8                  0.078      
+## Escobar           30           33.3         27.4 (19.5-37.6)          46.7         38.1 (21.0-60.0)                 NA                   0.047      
+## Goodrich         982           19.7         26.3 (16.2-52.4)          98.9         61.0 (23.0-86.0)                 NA                  <0.001      
+## Hmp              287           10.8         24.3 (19.0-34.0)          49.1         26.3 (18.0-40.0)                81.5                  0.322      
+## Ross              63           60.3         31.6 (22.1-47.9)          76.2         57.0 (33.0-81.0)                0.0                   0.845      
+## Schubert         104           32.7         28.2 (18.5-62.5)          66.3         52.8 (19.0-88.0)                82.7                  0.180      
+## Turnbaugh        146           67.8                NA                  NA                 NA                       51.4                  0.040      
+## Wu                64            7.8         24.3 (14.0-41.3)          53.1         26.3 (2.16-50.0)                 NA                   0.577      
+## Zeevi            731            NA          26.4 (16.4-47.0)           NA          43.4 (18.0-70.0)                 NA                   0.135      
+## Zupancic         207           36.2         28.2 (18.2-127.0)         57.0         46.7 (20.0-79.0)               100.0                  0.206
+\end{verbatim}
+
 \newpage

### 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
Trying to get table to render in pdf

--- 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{longtable,booktabs}
 \usepackage{graphicx,grffile}
@@
 
+\begin{verbatim}
+## 
+## > capwords <- function(s, strict = FALSE) {
+## +     s <- as.character(s)
+## +     cap <- function(s) paste(toupper(substring(s, 1, 1)), {
+## +         s <- su .... [TRUNCATED] 
+## 
+## > make_study_label <- function(dataset) {
+## +     study <- ifelse(dataset == "hmp", "HMP", capwords(dataset))
+## + }
+## 
+## > format_p <- function(p_value) {
+## +     char_p <- format(round(p_value, 3), nsmall = 3)
+## +     if (p_value < 0.001) {
+## +         char_p <- "<0.001"
+## +    .... [TRUNCATED] 
+## 
+## > get_study_summary <- function(study, beta = beta_summary) {
+## +     metadata_file <- paste0("data/", study, "/", study, ".metadata")
+## +     metadata <- .... [TRUNCATED] 
+## 
+## > beta_summary <- read.table("data/process/beta_tests.summary", 
+## +     header = T, row.names = 1)
+## 
+## > datasets <- sort(rownames(beta_summary))
+## 
+## > study_summary <- t(sapply(datasets, get_study_summary))
+## 
+## > table1 <- kable(study_summary, row.names = FALSE, 
+## +     col.names = c("Study", "Subjects (N)", "Obese (%)", "Average BMI (Min-Max)", 
+## +         "Fe ..." ... [TRUNCATED]
+\end{verbatim}
+
+\begin{longtable}[c]{@{}lccccccc@{}}
+\toprule
+Study & Subjects (N) & Obese (\%) & Average BMI (Min-Max) & Female (\%)
+& Average Age (Min-Max) & Non-Hispanic White (\%) & AMOVA
+(P-value)\tabularnewline
+\midrule
+\endhead
+Baxter & 172 & 27.3 & 27.0 (17.5-46.9) & 64.5 & 54.3 (29.0-80.0) & 87.8
+& 0.078\tabularnewline
+Escobar & 30 & 33.3 & 27.4 (19.5-37.6) & 46.7 & 38.1 (21.0-60.0) & NA &
+0.047\tabularnewline
+Goodrich & 982 & 19.7 & 26.3 (16.2-52.4) & 98.9 & 61.0 (23.0-86.0) & NA
+& \textless{}0.001\tabularnewline
+Hmp & 287 & 10.8 & 24.3 (19.0-34.0) & 49.1 & 26.3 (18.0-40.0) & 81.5 &
+0.322\tabularnewline
+Ross & 63 & 60.3 & 31.6 (22.1-47.9) & 76.2 & 57.0 (33.0-81.0) & 0.0 &
+0.845\tabularnewline
+Schubert & 104 & 32.7 & 28.2 (18.5-62.5) & 66.3 & 52.8 (19.0-88.0) &
+82.7 & 0.180\tabularnewline
+Turnbaugh & 146 & 67.8 & NA & NA & NA & 51.4 & 0.040\tabularnewline
+Wu & 64 & 7.8 & 24.3 (14.0-41.3) & 53.1 & 26.3 (2.16-50.0) & NA &
+0.577\tabularnewline
+Zeevi & 731 & NA & 26.4 (16.4-47.0) & NA & 43.4 (18.0-70.0) & NA &
+0.135\tabularnewline
+Zupancic & 207 & 36.2 & 28.2 (18.2-127.0) & 57.0 & 46.7 (20.0-79.0) &
+100.0 & 0.206\tabularnewline
+\bottomrule
+\end{longtable}
+
 \newpage

### 2016-06-03 code/make_table1.R

@@ -64,4 +64,4 @@ datasets <- sort(rownames(beta_summary))
 
 study_summary <- t(sapply(datasets, get_study_summary))
 
-kable(study_summary, row.names=FALSE, col.names=c("Study", "Subjects (N)", "Obese (%)", "Average BMI (Min-Max)", "Female (%)", "Average Age (Min-Max)", "Non-Hispanic White (%)", "AMOVA (P-value)"), align=c('l',rep('c', 7)), digits=3)
+table1 <- kable(study_summary, row.names=FALSE, col.names=c("Study", "Subjects (N)", "Obese (%)", "Average BMI (Min-Max)", "Female (%)", "Average Age (Min-Max)", "Non-Hispanic White (%)", "AMOVA (P-value)"), align=c('l',rep('c', 7)), digits=3)

### 2016-06-03 results/tables/table_1.Rmd
Clean up results/tables folder

capwords <- function(s, strict = FALSE) {
		s <- as.character(s)
		cap <- function(s) paste(toupper(substring(s, 1, 1)),
									{s <- substring(s, 2); if(strict) tolower(s) else s},
														 sep = "", collapse = " " )
		sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
}

make_study_label <- function(dataset){
	study <- ifelse(dataset=="hmp", "HMP", capwords(dataset))
}

format_p <- function(p_value){
	char_p <- format(round(p_value, 3), nsmall=3L)

	if(p_value < 0.001){
		char_p <- "<0.001"
	}

	return(char_p)
}

get_study_summary <- function(study, beta = beta_summary){
	metadata_file <- paste0('data/', study, '/', study, '.metadata')
	metadata <- read.table(file=metadata_file, header=T, stringsAsFactors=FALSE, na.strings=c("NA", "NULL"))

	n_subjects <- nrow(metadata)

	per_obese <- format(round(100 * mean(metadata$obese), 1), nsmall=1L)

	bmi <- "NA"
	if(sum(is.na(metadata$bmi)) != nrow(metadata)){
		mean_bmi <- format(signif(mean(metadata$bmi, na.rm=T), 3), nsmall=1L)
		min_bmi <- format(signif(min(metadata$bmi, na.rm=T), 3), nsmall=1L)
		max_bmi <- format(signif(max(metadata$bmi, na.rm=T), 3), nsmall=1L)
		bmi <- paste0(mean_bmi, " (", min_bmi, "-", max_bmi, ")")
	}

	per_female <- format(round(100 * mean(tolower(metadata$sex) == 'f'), 1), nsmall=1L)

	age <- "NA"
	if(sum(is.na(metadata$age)) != nrow(metadata)){
		mean_age <- format(signif(mean(metadata$age, na.rm=T), 3), nsmall=1L)
		min_age <- format(signif(min(metadata$age, na.rm=T), 3), nsmall=1L)
		max_age <- format(signif(max(metadata$age, na.rm=T), 3), nsmall=1L)
		age <- paste0(mean_age, " (", min_age, "-", max_age, ")")
	}

	per_white <- format(round(100*mean(metadata$white), 1), nsmall=1L)

	list(study = capwords(study),
				n_subjects = n_subjects,
				per_obese = per_obese,
				bmi_summary = bmi,
				per_female = per_female,
				age_summary = age,
				per_white = per_white,
				beta_p = format_p(beta_summary[study,'p_value']))
}

beta_summary <- read.table("data/process/beta_tests.summary",
															header=T, row.names = 1)
datasets <- sort(rownames(beta_summary))

study_summary <- t(sapply(datasets, get_study_summary))

table1 <- kable(study_summary, row.names=FALSE, col.names=c("Study", "Subjects (N)", "Obese (%)", "Average BMI (Min-Max)", "Female (%)", "Average Age (Min-Max)", "Non-Hispanic White (%)", "AMOVA (P-value)"), align=c('l',rep('c', 7)), digits=3)

### 2016-06-03 results/tables/table_1.Rmd
Create xtable version of Table 1

@@
+---
+output:
+  pdf_document:
+    keep_tex: true
+    includes:
+      in_header: table_header.tex
+tables: true
+geometry: margin=0.75in
+---
+
+**Table 1. Summary of obesity, demographic, and beta-diversity analysis data for the studies used in the meta-analysis.** NA indicates that those metadata were not available for that study.
+
+```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
+source('code/utilities.R')
+get_dependencies('xtable')
+
+options(xtable.sanitize.text.function=identity)
+
 capwords <- function(s, strict = FALSE) {
@@
 	list(study = capwords(study),
-				n_subjects = n_subjects,
+				n_subjects = as.character(n_subjects),
 				per_obese = per_obese,
@@
 
-table1 <- kable(study_summary, row.names=FALSE, col.names=c("Study", "Subjects (N)", "Obese (%)", "Average BMI (Min-Max)", "Female (%)", "Average Age (Min-Max)", "Non-Hispanic White (%)", "AMOVA (P-value)"), align=c('l',rep('c', 7)), digits=3)
+#colnames(study_summary) <- LETTERS[1:8]
+colnames(study_summary) <- c("Study", "\\bigcell{c}{Subjects \\\\ (N)}",
+													"\\bigcell{c}{Obese \\\\ (\\%)}",
+													"\\bigcell{c}{Average BMI \\\\ (Min-Max)}",
+													"\\bigcell{c}{Female \\\\ (\\%)}",
+													"\\bigcell{c}{Average Age \\\\ (Min-Max)}",
+													"\\bigcell{c}{Non-Hispanic \\\\ White (\\%)}",
+													"\\bigcell{c}{AMOVA \\\\ (P-value)}")
+
+colnames(study_summary) <- paste0("\\textbf{", colnames(study_summary), "}")
+
+composite <- xtable(study_summary, label="")
+
+align(composite) <- "llccccccc"
+
+print(composite,
+		include.rownames=FALSE,
+		include.colnames=TRUE,
+		hline.after = c(-1,0, nrow(composite)),
+		comment=FALSE,
+		size="normalsize",
+		table.placement="!ht"
+		)
+```

### 2016-06-03 submission/supp_text.Rmd
Incorporate supplementary text file

--- 2016-05-27 submission/Sze_Obesity_PeerJ_2016_Supp.Rmd
@@
 ---
-title: "Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome"
+title: ""
+bibliography: reference.bib
 output:
-  html_document:
-    keep_md: yes
-bibliography: reference.bib
+  pdf_document:
+    keep_tex: true
+    includes:
+      in_header: header.tex
+csl: mbio.csl
+fontsize: 12pt
+geometry: margin=1.0in
 ---
 
-Marc A Sze<sup>1</sup> and Patrick D Schloss<sup>1</sup>
+## Supplementary Text 1: Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome
 
-```{r Loading_data_tables, echo=FALSE, message=FALSE, warning=FALSE}
+Marc A Sze and Patrick D Schloss
 
-alpha_tests.summary <- read.table("../data/process/alpha_tests.summary", header=T)
-beta_tests.summary <- read.table("../data/process/beta_tests.summary", header=T)
-```
-
-###In-Depth Overview of Search Strategy:
+### In-Depth Overview of Search Strategy
 
@@
 
-*Inclusion Criteria:*
+***Inclusion Criteria:***
 
@@
 
-*Exclusion Criteria:*
+***Exclusion Criteria:***
 
@@
 
-###Supplemental Results:
-
-```{r load_needed_data_tables, echo=FALSE, message=FALSE, warning=FALSE}
-
-alpha_tests.summary <- read.table("../data/process/alpha_tests.summary", header=T)
-alpha_composite.summary <- read.table("../data/process/alpha_composite.summary")
-
-```
-
-**Figure S1: Individual and Combined comparison of Obese and Non-Obese groups for Observed OTUs.**  For observed OTUs Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 2), nsmall = 2)`, Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 5), nsmall = 5)`), and Turnbaugh (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'turnbaugh' & alpha_tests.summary$metric == 'sobs'), 'p.value'], 2), nsmall = 2)`) studies individually were signficantly different for obese and non-obese groups.
-When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 5), nsmall = 5)`)
-
-**Figure S2: Individual and Combined comparison of Obese and Non-Obese groups for Evenness.**  For evenness only the Goodrich study (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannoneven'), 'p.value'], 3), nsmall = 3)`)  individually was signficantly different for obese and non-obese groups.
-When the data sets were combined and study effect was corrected for,  there was an overall significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 2), nsmall = 2)`)
-
-**Figure S3: Individual and Combined comparison of Obese and Non-Obese groups for Bacteroidetes Relative Abundance.**  Only the Escobar study (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'escobar' & alpha_tests.summary$metric == 'bacteroidetes'), 'p.value'], 3), nsmall = 3)`)  individually was signficantly different for obese and non-obese groups.
-When the data sets were combined and study effect was corrected for,  there was no overall significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall = 2)`)
-
-**Figure S4: Individual and Combined comparison of Obese and Non-Obese groups for Firmicutes Relative Abundance.**  No study was individually signficantly different for obese and non-obese groups.
-When the data sets were combined and study effect was corrected for,  there was also no overall significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall = 2)`)
-
-**Figure S5: Funnel Plot of the Shannon Diversity and Bacteroidetes/Firmicutes Ratio Relative Risk.**  **A)** Overall there does not seem to be any bias associated with the studies selected for the Shannon Diversity analysis with studies falling on either side of the predicted value and those with smaller total n falling further away from this.
-**B)** For the Bacteroidetes/Firmicutes ratio there does not seem to be any associated bias.
-The overall pattern is similar to that observed for the Shannon Diversity analysis.
-
-**Figure S6.
-Summary of Power and Sample Size Simulations for OTU Richness (S) for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-**Figure S7.
-Summary of Power and Sample Size Simulations for Evenness (J) for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-**Figure S8.
-Summary of Power and Sample Size Simulations for Bacteroidetes for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-**Figure S9.
-Summary of Power and Sample Size Simulations for Firmicutes for Non-obese versus Obese**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-**Figure S10.
-Summary of Power and Sample Size Simulations for Relative Risk  for Obesity based on Shannon Diversity**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-**Figure S11.
-Summary of Power and Sample Size Simulations for Relative Risk  for Obesity based on B/F Ratio**  P-values listed in the legend represent the outcome of a wilcoxson rank sum test between non-obese and obese for each specific data set.
-**A)** The dotted lines represent the actual effect size for each specific study.
-**B)** The dotted lines represent the needed n to achieve an 80% power with the actual study effect size.
-
-*******
-##### References:
+### References

### 2016-06-03 submission/supp_text.tex
Incorporate supplementary text file

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\ImportTok}[1]{{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
\newcommand{\BuiltInTok}[1]{{#1}}
\newcommand{\ExtensionTok}[1]{{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}
\usepackage{comment}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\begin{document}

\subsection{Supplementary Text 1: Looking for a Signal in the Noise:
Revisiting Obesity and the
Microbiome}\label{supplementary-text-1-looking-for-a-signal-in-the-noise-revisiting-obesity-and-the-microbiome}

Marc A Sze and Patrick D Schloss

\subsubsection{In-Depth Overview of Search
Strategy}\label{in-depth-overview-of-search-strategy}

The initial search strategy included looking for all papers that
initially fit under the below NCBI PubMed advanced search criteria. The
terms included in this criteria were that the manuscript had to have
``Bacterial Microbiome'' and ``Obesity, BMI, bmi, obesity'' in their
manuscript criteria, it was not published more than 10 years ago, they
were not review articles, and it contained research on humans only. The
below formula when put into PubMed should recapitulate our initial
search on the website.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{((}\NormalTok{(((((((Bacterial Microbiome) AND (Obesity or bmi or body mass index or BMI or obesity) AND }\StringTok{"last 10 years"}\NormalTok{[PDat] AND Humans[Mesh])) NOT review[ptyp]) AND }\StringTok{"last 10 years"}\NormalTok{[PDat] AND Humans[Mesh])) AND }\StringTok{"last 10 years"}\NormalTok{[PDat] AND Humans[Mesh])) AND }\StringTok{"last 10 years"}\NormalTok{[PDat] AND Humans[Mesh])}
\end{Highlighting}
\end{Shaded}

This search yielded a total of 187 manuscripts. From two previous other
reviews of obesity and the bacterial microbiome along with knowledge of
two other published papers that investigated obesity but were missed by
the database search we obtained a total of 7 more articles. We also had
access to normal healthy individuals from an unpublished dataset. This
brought our total number of records to 196.

From this total we browsed abstracts for mention of stool or feces
examination, that did not involve children, was not a clinical trial for
probiotics or other diet related treatments, did not only have
participants with inflammatory bowel disease, the articles were in
English, did not only use PCR, qPCR, or RT-PCR only for their analysis,
and sequencing that used only clone libraries. This ultimately excluded
all but a total of 13 studies.

From this total of 11 studies with the full text was reviewed for
whether or not sequencing data was publicly available, BMI information
(either categorical or continuous) was available in a supplement or, if
it was not available, whether authors upon contact were willing to share
this information or direct us to repositories that stored this specific
information. One study was excluded (1) because it contained children
and their sequencing of the 16S rRNA gene involved amplicons of only
100bp in length. They also did not have obesity as part of their results
in the actual published manuscript. The second study was excluded
because it did not use 16S rRNA gene sequencing for their bacterial
microbiome analysis (2).

Once these 2 studies were excluded there was a total of 11 studies in
the qualitative synthesis of the analysis. Because we decided a prioi to
use the standard definition for BMI group classification one study from
this 11 did not have any individuals who were obese by this criteria (3)
and was excluded from the final quantitative synthesis and analysis.

\textbf{\emph{Inclusion Criteria:}}

\begin{itemize}
\tightlist
\item
  Contains mention of Bacterial Microbiome and Obesity
\item
  BMI, bmi, or obesity could be referenced instead of Obesity
\item
  Not published more than 10 years ago
\item
  Research on Humans only
\item
  At least one specific result examining obesity and a bacterial
  microbiome measure
\item
  Participants did not have Inflammatory Bowel Disease or Cancer
\item
  Greater than 100bp single or dual end reads for 16S Sequencing
\item
  DNA obtained from stool or feces
\end{itemize}

\textbf{\emph{Exclusion Criteria:}}

\begin{itemize}
\tightlist
\item
  PCR, qPCR, metagenomic sequencing, or RT-PCR used as main analysis
  Tool
\item
  TRFLP or clone sequencing used to asses the bacterial community
\item
  Utilization of 100bp or less single end reads for sequencing
\item
  Sequencing Data not publicly available for download
\item
  BMI not available and authors do not return correspondence
\item
  Samples were not stool or feces
\item
  Study contained children
\item
  Study was a review
\end{itemize}

\subsubsection*{References}\label{references}
\addcontentsline{toc}{subsubsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-yatsunenkoux5fhumanux5f2012}{}
1. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
\textbf{Trehan I}, \textbf{Dominguez-Bello MG}, \textbf{Contreras M},
\textbf{Magris M}, \textbf{Hidalgo G}, \textbf{Baldassano RN},
\textbf{Anokhin AP}, \textbf{Heath AC}, \textbf{Warner B},
\textbf{Reeder J}, \textbf{Kuczynski J}, \textbf{Caporaso JG},
\textbf{Lozupone CA}, \textbf{Lauber C}, \textbf{Clemente JC},
\textbf{Knights D}, \textbf{Knight R}, \textbf{Gordon JI}. 2012. Human
gut microbiome viewed across age and geography. Nature
\textbf{486}:222--227.
doi:\href{https://doi.org/10.1038/nature11053}{10.1038/nature11053}.

\hypertarget{ref-arumugamux5fenterotypesux5f2011}{}
2. \textbf{Arumugam M}, \textbf{Raes J}, \textbf{Pelletier E},
\textbf{Le Paslier D}, \textbf{Yamada T}, \textbf{Mende DR},
\textbf{Fernandes GR}, \textbf{Tap J}, \textbf{Bruls T}, \textbf{Batto
J-M}, \textbf{Bertalan M}, \textbf{Borruel N}, \textbf{Casellas F},
\textbf{Fernandez L}, \textbf{Gautier L}, \textbf{Hansen T},
\textbf{Hattori M}, \textbf{Hayashi T}, \textbf{Kleerebezem M},
\textbf{Kurokawa K}, \textbf{Leclerc M}, \textbf{Levenez F},
\textbf{Manichanh C}, \textbf{Nielsen HB}, \textbf{Nielsen T},
\textbf{Pons N}, \textbf{Poulain J}, \textbf{Qin J},
\textbf{Sicheritz-Ponten T}, \textbf{Tims S}, \textbf{Torrents D},
\textbf{Ugarte E}, \textbf{Zoetendal EG}, \textbf{Wang J},
\textbf{Guarner F}, \textbf{Pedersen O}, \textbf{Vos WM de},
\textbf{Brunak S}, \textbf{Doré J}, \textbf{MetaHIT Consortium},
\textbf{Antolín M}, \textbf{Artiguenave F}, \textbf{Blottiere HM},
\textbf{Almeida M}, \textbf{Brechot C}, \textbf{Cara C},
\textbf{Chervaux C}, \textbf{Cultrone A}, \textbf{Delorme C},
\textbf{Denariaz G}, \textbf{Dervyn R}, \textbf{Foerstner KU},
\textbf{Friss C}, \textbf{Guchte M van de}, \textbf{Guedon E},
\textbf{Haimet F}, \textbf{Huber W}, \textbf{Hylckama-Vlieg J van},
\textbf{Jamet A}, \textbf{Juste C}, \textbf{Kaci G}, \textbf{Knol J},
\textbf{Lakhdari O}, \textbf{Layec S}, \textbf{Le Roux K},
\textbf{Maguin E}, \textbf{Mérieux A}, \textbf{Melo Minardi R},
\textbf{M'rini C}, \textbf{Muller J}, \textbf{Oozeer R},
\textbf{Parkhill J}, \textbf{Renault P}, \textbf{Rescigno M},
\textbf{Sanchez N}, \textbf{Sunagawa S}, \textbf{Torrejon A},
\textbf{Turner K}, \textbf{Vandemeulebrouck G}, \textbf{Varela E},
\textbf{Winogradsky Y}, \textbf{Zeller G}, \textbf{Weissenbach J},
\textbf{Ehrlich SD}, \textbf{Bork P}. 2011. Enterotypes of the human gut
microbiome. Nature \textbf{473}:174--180.
doi:\href{https://doi.org/10.1038/nature09944}{10.1038/nature09944}.

\hypertarget{ref-namux5fcomparativeux5f2011}{}
3. \textbf{Nam Y-D}, \textbf{Jung M-J}, \textbf{Roh SW}, \textbf{Kim
M-S}, \textbf{Bae J-W}. 2011. Comparative analysis of Korean human gut
microbiota by barcoded pyrosequencing. PloS One \textbf{6}:e22109.
doi:\href{https://doi.org/10.1371/journal.pone.0022109}{10.1371/journal.pone.0022109}.

\end{document}

### 2016-06-03 submission/Sze_Obesity_mBio_2016.Rmd
Bring in Table 1
Echo output from building table
Trying to get table to render in pdf
Remove Taable 1 from pdf - wasn't rendering correctly
Add AMOVA reference

@@
 
-The recent publication of additional studies that collected BMI data for each subject as well as other studies that were not included in the earlier re-analysis studies offered the opportunity to revisit the question relating the structure of the human microbiome to obesity [@ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @escobar_gut_2014; @baxter_microbiota-based_2016; @schubert_microbiome_2014].
+The recent publication of additional studies that collected BMI data for each subject as well as other studies that were not included in the earlier re-analysis studies offered the opportunity to revisit the question relating the structure of the human microbiome to obesity.
 One critique of the prior re-analysis studies is that the authors did not aggregate the results across studies to increase the effective sample size.
@@
 
-## Methods
-
-***Literature Review and Study Inclusion.***
-We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010].
-A detailed description of our selection process and the exact search terms are provided in the Supplement and in figure 1.
-Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome.
-The initial search yielded 187 studies.
-We identified number_string[n] additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
-We then manually curated the 196 studies to select those studies that included BMI and sequence data.
-This yielded 10 eligible studies.
-An additional study was removed from our analysis because no individuals in the study had a BMI over 30.
-Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity.
-The number_string[n] studies are summarized in Tables 1 and 2.
-
-***Sequence Analysis Pipeline.***
-All sequence data were publicly available and were downloaded from the NCBI Sequence Read Archive, the European Nucleotide Archive, or the investigators' personal website (https://gordonlab.wustl.edu/TurnbaughSE\_10\_09/STM\_2009.html).
-In total seven studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012; @schubert_microbiome_2014] and three studies used Illumina sequencing [@goodrich_human_2014; @zeevi_personalized_2015; @baxter_microbiota-based_2016].
-All of these studies used amplification-based 16S rRNA gene sequencing.
-Among the studies that sequenced the 16S rRNA gene, the researchers targeted the V1-V2  [@wu_linking_2011], V1-V3 [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], V3-V5 [@human_microbiome_project_consortium_structure_2012; @schubert_microbiome_2014], V4 [@goodrich_human_2014; @baxter_microbiota-based_2016; ], and V3-4 [@zeevi_personalized_2015] regions.
-For those studies where multiple regions were sequenced, we selected the region that corresponded to the largest number of subjects [@turnbaugh_core_2009; @human_microbiome_project_consortium_structure_2012].
-We processed the 16S rRNA gene sequence data using a standardized mothur pipeline.
-Briefly, our pipelines attempted to follow previously recommended approaches for 454 and Illumina sequencing data [@Kozich_2013; @Schloss2011].
-All sequences were screened for chimeras using UCHIME and assigned to operational taxonomic units (OTUs) using the average neighbor algorithm using a 3% distance threshold [@Edgar2011; @Westcott2015].
-All sequence processing was performed using mothur (v.1.37.0) [@mothur].
-
-***Data Analysis.***
-We split the overall meta-analysis into three general strategies using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`).
-First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B:F), Shannon diversity, observed richness, and Shannon evenness.
-After each variable was transformed to fit a normal distribution a two-tailed t-test was performed for comparison of non-obese and obese individuals (i.e.
-BMI > 35.0).
-We performed a pooled analysis on these measured variables using linear random effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups using the lme4 (v.`r packageDescription("lme4")$Version`) R package.
-Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices.
-This analysis was performed using the vegan (v.`r packageDescription("vegan")$Version`) R package.
-For both analyses, the datasets were rarefied (N=1000) so that each study within a study had the same number of sequences.
-Second, for each study we partitioned the subjects into a low or high group depending on whether their alpha diversity metrics were below or above the median value for the study.
-The relative risk (RR) was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group.
-We then performed a Fisher exact-test to investigate whether the RR was significantly different from 1.0 within each study and across all of the studies using the epiR (`r packageDescription("epiR")$Version`) and metafor (`r packageDescription("metafor")$Version`) packages.
-Third, we used the AUCRF (`r packageDescription("AUCRF")$Version`) R package to generate Random Forest models.
-For each study we developed models using either OTUs or genus-level phylotypes.
-The quality of each model was assessed by measuring the area under the curve (AUC) of the Receive Operating Characteristic (ROC) using ten-fold cross validation.
-Because the genus-level phylotype models were developed using a common reference, it was possible to use one study's model (i.e.
-the training set) to classify the samples from the other studies (i.e.
-the testing sets).
-The optimum threshold for the training set was set as the probability threshold that had the highest combined sensitivity and specificity.
-This threshold was then used to calculate the accuracy of the model applied to the test studies.
-To generate Receiver Operator Characteristic (ROC) curves and calculate the accuracy of the models we used the pROC (`r packageDescription("pROC")$Version`) R package.
-Finally, we performed power and sample number simulations for different effect sizes for each study using the pwr (`r packageDescription("pwr")$Version`) R package and base R functions.
-We also calculated the actual sample size needed based on the effect size of each individual study.
-
-***Reproducible methods.***
-A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_Obesity_mBio_2016/.
-
 ## Results
 
+***Meta-analysis***
+
+[@ross_16s_2015; @zupancic_analysis_2012; @goodrich_human_2014; @wu_linking_2011; @escobar_gut_2014; @baxter_microbiota-based_2016; @schubert_microbiome_2014; @zeevi_personalized_2015; @human_microbiome_project_consortium_structure_2012; @turnbaugh_obesity-associated_2006]
 ```{r}
@@
 Following the approach used by the Walters and Finucane re-analysis studies, for each dataset we calculated a Bray-Curtis distance matrix to measure the difference in the membership and structure of the individuals from each study.
-We then used PERMANOVA to test for a significant differences between the structure of non-obese and obese individuals.
+We then used AMOVA to test for a significant differences between the structure of non-obese and obese individuals.
 The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05).
@@
 
+## Methods
+
+***Literature Review and Study Inclusion.***
+We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010].
+A detailed description of our selection process and the exact search terms are provided in the Supplement and in figure 1.
+Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome.
+The initial search yielded 187 studies.
+We identified number_string[n] additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
+We then manually curated the 196 studies to select those studies that included BMI and sequence data.
+This yielded 10 eligible studies.
+An additional study was removed from our analysis because no individuals in the study had a BMI over 30.
+Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity.
+The number_string[n] studies are summarized in Tables 1 and 2.
+
+***Sequence Analysis Pipeline.***
+All sequence data were publicly available and were downloaded from the NCBI Sequence Read Archive, the European Nucleotide Archive, or the investigators' personal website (https://gordonlab.wustl.edu/TurnbaughSE\_10\_09/STM\_2009.html).
+In total seven studies used 454 [@turnbaugh_core_2009; @zupancic_analysis_2012; @ross_16s_2015; @escobar_gut_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012; @schubert_microbiome_2014] and three studies used Illumina sequencing [@goodrich_human_2014; @zeevi_personalized_2015; @baxter_microbiota-based_2016].
+All of these studies used amplification-based 16S rRNA gene sequencing.
+Among the studies that sequenced the 16S rRNA gene, the researchers targeted the V1-V2  [@wu_linking_2011], V1-V3 [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], V3-V5 [@human_microbiome_project_consortium_structure_2012; @schubert_microbiome_2014], V4 [@goodrich_human_2014; @baxter_microbiota-based_2016; ], and V3-4 [@zeevi_personalized_2015] regions.
+For those studies where multiple regions were sequenced, we selected the region that corresponded to the largest number of subjects [@turnbaugh_core_2009; @human_microbiome_project_consortium_structure_2012].
+We processed the 16S rRNA gene sequence data using a standardized mothur pipeline.
+Briefly, our pipelines attempted to follow previously recommended approaches for 454 and Illumina sequencing data [@Kozich_2013; @Schloss2011].
+All sequences were screened for chimeras using UCHIME and assigned to operational taxonomic units (OTUs) using the average neighbor algorithm using a 3% distance threshold [@Edgar2011; @Westcott2015].
+All sequence processing was performed using mothur (v.1.37.0) [@mothur].
+
+***Data Analysis.***
+We split the overall meta-analysis into three general strategies using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`).
+First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B:F), Shannon diversity, observed richness, and Shannon evenness.
+After each variable was transformed to fit a normal distribution a two-tailed t-test was performed for comparison of non-obese and obese individuals (i.e.
+BMI > 35.0).
+We performed a pooled analysis on these measured variables using linear random effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups using the lme4 (v.`r packageDescription("lme4")$Version`) R package.
+Next, we compared the community structure from non-obese and obese individuals using analysis of molecular variance (AMOVA) with Bray-Curtis distance matrices [@Anderson2001].
+This analysis was performed using the vegan (v.`r packageDescription("vegan")$Version`) R package.
+For both analyses, the datasets were rarefied (N=1000) so that each study within a study had the same number of sequences.
+Second, for each study we partitioned the subjects into a low or high group depending on whether their alpha diversity metrics were below or above the median value for the study.
+The relative risk (RR) was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group.
+We then performed a Fisher exact-test to investigate whether the RR was significantly different from 1.0 within each study and across all of the studies using the epiR (`r packageDescription("epiR")$Version`) and metafor (`r packageDescription("metafor")$Version`) packages.
+Third, we used the AUCRF (`r packageDescription("AUCRF")$Version`) R package to generate Random Forest models.
+For each study we developed models using either OTUs or genus-level phylotypes.
+The quality of each model was assessed by measuring the area under the curve (AUC) of the Receive Operating Characteristic (ROC) using ten-fold cross validation.
+Because the genus-level phylotype models were developed using a common reference, it was possible to use one study's model (i.e.
+the training set) to classify the samples from the other studies (i.e.
+the testing sets).
+The optimum threshold for the training set was set as the probability threshold that had the highest combined sensitivity and specificity.
+This threshold was then used to calculate the accuracy of the model applied to the test studies.
+To generate Receiver Operator Characteristic (ROC) curves and calculate the accuracy of the models we used the pROC (`r packageDescription("pROC")$Version`) R package.
+Finally, we performed power and sample number simulations for different effect sizes for each study using the pwr (`r packageDescription("pwr")$Version`) R package and base R functions.
+We also calculated the actual sample size needed based on the effect size of each individual study.
+
+***Reproducible methods.***
+A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_Obesity_mBio_2016/.
+
 ## Acknowledgements
@@
 This work was supported in part by funding from the National Institutes of Health to PDS (U01AI2425501 and P30DK034933).
-
-\newpage
-
-**Table 1. Summary Demographics of Individuals used in the Meta-analysis.**
 

### 2016-06-03 results/tables/table_1.Rmd
Add sequencing information

--- 2016-06-03 results/tables/table_1.Rmd
@@
+---
+output:
+  pdf_document:
+    keep_tex: true
+    includes:
+      in_header: table_header.tex
+tables: true
+geometry: margin=0.75in
+---
+
+**Table 1.
+Summary of obesity, demographic, sequencing, and beta-diversity analysis data for the studies used in the meta-analysis.** NA indicates that those metadata were not available for that study.
+
+```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
+source('code/utilities.R')
+get_dependencies('xtable')
+
+options(xtable.sanitize.text.function=identity)
+
 capwords <- function(s, strict = FALSE) {
@@
 }
+
+region <- c(
+	baxter = "V4",
+	escobar = "V2",
+	hmp = "V3-V5",
+	ross = "V1-V3",
+	schubert = "V4",
+	turnbaugh = "V2",
+	wu = "V1-V2",
+	zupancic = "V3-V5",
+	goodrich = "V4",
+	zeevi = "V3-V4"
+)
+
+technology <-  c(
+	baxter = "MiSeq",
+	escobar = "454",
+	hmp = "454",
+	ross = "454",
+	schubert = "MiSeq",
+	turnbaugh = "454",
+	wu = "454",
+	zupancic = "454",
+	goodrich = "MiSeq",
+	zeevi = "MiSeq"
+)
 
@@
 				per_white = per_white,
+				method = technology[study],
+				region = region[study],
 				beta_p = format_p(beta_summary[study,'p_value']))
@@
 
-table1 <- kable(study_summary, row.names=FALSE, col.names=c("Study", "Subjects (N)", "Obese (%)", "Average BMI (Min-Max)", "Female (%)", "Average Age (Min-Max)", "Non-Hispanic White (%)", "AMOVA (P-value)"), align=c('l',rep('c', 7)), digits=3)
+#colnames(study_summary) <- LETTERS[1:8]
+colnames(study_summary) <- c("Study", "\\bigcell{c}{Subjects \\\\ (N)}",
+													"\\bigcell{c}{Obese \\\\ (\\%)}",
+													"\\bigcell{c}{Average BMI \\\\ (Min-Max)}",
+													"\\bigcell{c}{Female \\\\ (\\%)}",
+													"\\bigcell{c}{Average Age \\\\ (Min-Max)}",
+													"\\bigcell{c}{Non-Hispanic \\\\ White (\\%)}",
+													"\\bigcell{c}{Sequencing \\\\ Method}",
+													"\\bigcell{c}{16S rRNA \\\\ Gene Region}",
+													"\\bigcell{c}{AMOVA \\\\ (P-value)}")
+
+colnames(study_summary) <- paste0("\\textbf{", colnames(study_summary), "}")
+
+composite <- xtable(study_summary, label="")
+
+align(composite) <- "llccccccccc"
+
+print(composite,
+		include.rownames=FALSE,
+		include.colnames=TRUE,
+		hline.after = c(-1,0, nrow(composite)),
+		comment=FALSE,
+		size="scriptsize",
+		table.placement="!ht"
+		)
+```

### 2016-06-03 submission/Sze_Obesity_mBio_2016.Rmd
Editing manuscript

@@
 The Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines were applied to identify 10 studies for an updated and more synthetic analysis.
-Alpha diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances.
+Alpha diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status.
+They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances.
 Although these tests yielded small P-values, the difference between the Shannon diversity index of non-obese and obese individuals was `r alpha_effect_size["shannon"]`%.
-A power analysis demonstrated that `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity.
-When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r 100*range(test_summary$accuracy)[1]` and `r 100*range(test_summary$accuracy)[2]`% (median=`r 100*median(test_summary$accuracy)`%).
-Although there is statistical support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.
+A power analysis demonstrated that only `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity.
+When Random Forest machine learning models were trained on one dataset and then tested using the other 9 datasets, the median accuracy varied between `r 100*range(test_summary$accuracy)[1]` and `r 100*range(test_summary$accuracy)[2]`% (median=`r 100*median(test_summary$accuracy)`%).
+Although there was support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.
 
@@
 There have been several studies that report observing a link between the composition of microbiome and obesity in animal models and in humans.
-The first such study used genetically obese mice and observed the ratio of the relative abundances of Bacteroidetes to Firmicutes (B:F) was lower in obese mice than lean mice [@ley_obesity_2005].
+The first such study used genetically obese mice and observed the ratio of the relative abundances of *Bacteroidetes* to *Firmicutes* (B:F) was lower in obese mice than lean mice [@ley_obesity_2005].
 Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009].
-They also showed that the relative abundance of Bacteroidetes and Firmicutes increased and decreased, respectively, as obese individuals lost weight while on a fat or carbohydrate restricted diet [@ley_microbial_2006].
+They also showed that the relative abundance of *Bacteroidetes* and *Firmicutes* increased and decreased, respectively, as obese individuals lost weight while on a fat or carbohydrate restricted diet [@ley_microbial_2006].
 Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since direct manipulation of the communities yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016].
+Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since these manipulation yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016].
 The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.
@@
 
-***Meta-analysis***
+***Literature Review and Study Inclusion.***
+We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010].
+A detailed description of our selection process and the exact search terms are provided in the Supplemental Text and in Figure 1.
+Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome.
+The initial search yielded 187 studies.
+We identified number_string[n] additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
+We then manually curated the 196 studies to select those studies that included BMI and sequence data.
+This yielded 10 eligible studies.
+An additional study was removed from our analysis because no individuals in the study had a BMI over 30.
+Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_obesity-associated_2006; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity.
+The `r number_string[n]` studies are summarized in Table 1.
 
-[@ross_16s_2015; @zupancic_analysis_2012; @goodrich_human_2014; @wu_linking_2011; @escobar_gut_2014; @baxter_microbiota-based_2016; @schubert_microbiome_2014; @zeevi_personalized_2015; @human_microbiome_project_consortium_structure_2012; @turnbaugh_obesity-associated_2006]
 ```{r}
@@
 The B:F and the relative abundance of *Firmicutes* were not significantly associated with obesity in any study.
-We identified `r sum(alpha$p.value < 0.05)` P-values less than 0.05: `r number_string[sig_counts["sobs"]]` studies indicated obese individuals had a lower richness, `r number_string[sig_counts["shannon"]]` studies indicated a significantly lower diversity, `r number_string[sig_counts["shannoneven"]]` study indicated a significantly lower evenness, and `r number_string[sig_counts["bacteroidetes"]]` study indicated a significantly higher relative abundance of *Bacteroidetes* (Figures 2 and S1).
+We identified `r number_string[sum(alpha$p.value < 0.05)]` P-values that were less than 0.05: `r number_string[sig_counts["sobs"]]` studies indicated obese individuals had a lower richness, `r number_string[sig_counts["shannon"]]` studies indicated a significantly lower diversity, `r number_string[sig_counts["shannoneven"]]` study indicated a significantly lower evenness, and `r number_string[sig_counts["bacteroidetes"]]` study indicated a significantly higher relative abundance of *Bacteroidetes* (Figures 2 and S1).
 These results largely match those of the Walters and Finucane re-analysis studies.
-Interestingly, although only `r number_string[sig_counts["shannon"]]` of the 10 studies observed the previously reported association between lower diversity and obesity, the other studies appeared to have the same trend, albeit the differences were not statistically significant.
+Interestingly, although only `r number_string[sig_counts["shannon"]]` of the `r number_string[n]` studies observed the previously reported association between lower diversity and obesity, the other studies appeared to have the same trend, albeit the differences were not statistically significant.
 We used a random effects linear model to combine the studies using the study as the random effect and found statistical support for decreased richness, evenness, and diversity among obese individuals (all P<`r max(alpha_composite[alpha_composite$p <0.05, "p"])`).
-Although there was a significant relationship between these metrics and diversity and obesity status, the effect size was quite small.
+Although there was a significant relationship between these metrics and obesity status, the effect size was quite small.
 The obese individuals averaged `r alpha_effect_size["sobs"]`% lower richness, `r alpha_effect_size["shannoneven"]`% lower evenness, and `r alpha_effect_size["shannon"]`% lower diversity.
@@
 ***Relative risk.***
-Building upon the alpha diversity analysis we calculated the relative risk of being obese based on whether an individual's alpha diversity metrics were below or above the median metric for that study.
+Building upon the alpha diversity analysis we calculated the relative risk of being obese based on an individual's alpha diversity metrics relative to the median metric for that study.
 The results using relative risk largely matched those of using the untransformed alpha diversity data.
-Across the number_string[n] studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2).
+Across the `r number_string[n]` studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2).
 Again, although the relative risk values were not significant for other studies, the values tended to be above one.
@@
 Following the approach used by the Walters and Finucane re-analysis studies, for each dataset we calculated a Bray-Curtis distance matrix to measure the difference in the membership and structure of the individuals from each study.
-We then used AMOVA to test for a significant differences between the structure of non-obese and obese individuals.
+We then used AMOVA to test for a significant differences between the structure of non-obese and obese individuals (Table 1).
 The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05).
@@
 ***Development of a microbiome-based classifier of obesity.***
-The Walters re-analysis study suggested that it was possible to classify individuals as being obese or non-obese based on the composition of their microbiota.
+The Walters re-analysis study suggested that it was possible to classify individuals as being non-obese or obese  based on the composition of their microbiota.
 We repeated this analysis with additional datasets using OTU and genus-level phylotype data.
@@
 So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset.
-The the observed AUC values for the models applied to the training datasets varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B).
+The cross-validated AUC values for the models applied to the training datasets varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B).
 For each model we identified the probability where the sum of the sensitivity and specificity was the highest.
-We then used this to identify a threshold to calculate the accuracy of the models when applied to the other number_string[n-1] datasets (Figure 5).
+We then used this probability to define a threshold for calculating the accuracy of the models when applied to the other `r number_string[n-1]` datasets (Figure 5).
 Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`).
@@
 We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups.
-To detect a 1, 5, 10, and 15% difference in Shannon index, the median sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively.
-To detect a 1, 5, 10, and 15% difference in B:F values, the median sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)`  individuals, respectively.
-To detect a 1, 5, 10, and 15% difference in relative risk values using Shannon diversity, the median sampling effort per group was approximately `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively.
+To detect a 1, 5, 10, or 15% difference in Shannon index, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively.
+To detect a 1, 5, 10, and 15% difference in B:F values, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively.
+To detect a 1, 5, 10, and 15% difference in relative risk values using Shannon diversity, the median required sampling effort per group was approximately `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively.
 These estimates indicate that most microbiome studies are underpowered to detect modest effect sizes using either metric.
@@
 Our meta-analysis helps to provide a clarity to the ongoing debate of whether or not there are specific microbiome-based markers that can be associated with obesity.
-We performed an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.
+We performed an extensive literature review of the existing studies on the microbiome and obesity and performed a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.
 By statistically pooling the data from ten studies, we observed significant, but small, relationships between richness, evenness, and diversity and obesity status as well as the relative risk of being obese based on these metrics.
 We also generated Random Forest machine learning models trained on each dataset and tested on the remaining datasets.
-This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome is limited.
+This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome was limited.
 Finally, we assessed the ability of each study to detect defined differences in alpha diversity and observed that most studies were underpowered to detect modest effect sizes.
@@
 Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that there was not a consistent association between alpha diversity and obesity; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference.
-Our analysis also used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from 3 studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from 5 studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012].
-Our analysis included 4 of these studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011] and excluded 3 of the studies because they were too small [@ley_microbial_2006], only utilized metagenomic data [@Qin2010], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering [@yatsunenko_human_2012].
+Our analysis also used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from three studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from five studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012].
+Our analysis included four of these studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011] and excluded three of the studies because they were too small [@ley_microbial_2006], only utilized metagenomic data [@Qin2010], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering [@yatsunenko_human_2012].
 The additional `r number_string[n-3]` datasets were published after the two reviews were performed and include datasets with more samples than were found in the original studies.
@@
 
-Similar to our study, the Walters analysis [@walters_meta-analyses_2014], the authors generated Random Forest machine learning models to differentiate between non-obese and obese individuals.
+Similar to our study, the Walters et al [@walters_meta-analyses_2014] generated Random Forest machine learning models to differentiate between non-obese and obese individuals.
 They obtained similar AUC values to our analysis; however, they did not attempt to test these models on the other studies in their analysis.
@@
 
-Although we failed to find an effect it is not realistic to necessarily state that there is no microbiome impact on obesity.
+Although we failed to find an effect this does not necessarily mean that there is no role for the microbiome in obesity.
 There is strong evidence in murine models of obesity that the microbiome and level of adiposity can be manipulated via genetic manipulation of the animal and manipulation of the community through antibiotics or colonizing germ free mice with diverse fecal material from human donors [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016; @ley_obesity_2005].
@@
 ## Methods
-
-***Literature Review and Study Inclusion.***
-We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010].
-A detailed description of our selection process and the exact search terms are provided in the Supplement and in figure 1.
-Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome.
-The initial search yielded 187 studies.
-We identified number_string[n] additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
-We then manually curated the 196 studies to select those studies that included BMI and sequence data.
-This yielded 10 eligible studies.
-An additional study was removed from our analysis because no individuals in the study had a BMI over 30.
-Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity.
-The number_string[n] studies are summarized in Tables 1 and 2.
 
@@
 We split the overall meta-analysis into three general strategies using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`).
-First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B:F), Shannon diversity, observed richness, and Shannon evenness.
+First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of *Bacteroidetes* and Firmicutes, the ratio of *Bacteroidetes* and *Firmicutes* relative abundances (B:F), Shannon diversity, observed richness, and Shannon evenness.
 After each variable was transformed to fit a normal distribution a two-tailed t-test was performed for comparison of non-obese and obese individuals (i.e.
@@
 
-**Figure S1: Individual and Combined comparison of Obese and Non-Obese groups Based on Evenness (A), Richness (B), or the Relative Abundance of Bacteroidetes (C) and Firmictues (D).**
+**Figure S1: Individual and Combined comparison of Obese and Non-Obese groups Based on Evenness (A), Richness (B), or the Relative Abundance of *Bacteroidetes* (C) and Firmictues (D).**
 
-**Figure S2: Meta Analysis of the Relative Risk of Obesity Based on Evenness (A), Richness (B), or the Relative Abundance of Bacteroidetes (C) and Firmictues (D).**
+**Figure S2: Meta Analysis of the Relative Risk of Obesity Based on Evenness (A), Richness (B), or the Relative Abundance of *Bacteroidetes* (C) and Firmictues (D).**
 
@@
 
-**Figure S6: Power (A) and sample size simulations (B) for the relative abundance of Bacteroidetes for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S6: Power (A) and sample size simulations (B) for the relative abundance of *Bacteroidetes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure S7: Power (A) and sample size simulations (B) for the relative abundance of Firmicutes for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S7: Power (A) and sample size simulations (B) for the relative abundance of *Firmicutes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 

### 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
Remove Taable 1 from pdf - wasn't rendering correctly
Editing manuscript

--- 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
@@
     \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
   \newcommand{\euro}{€}
@@
 \usepackage[margin=1.0in]{geometry}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
+\usepackage{hyperref}
+\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
+\hypersetup{unicode=true,
             pdftitle={Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
+            pdfborder={0 0 0},
+            breaklinks=true}
 \urlstyle{same}  % don't use monospace font for urls
@@
 earlier re-analysis studies offered the opportunity to revisit the
-question relating the structure of the human microbiome to obesity
-(14--22). One critique of the prior re-analysis studies is that the
-authors did not aggregate the results across studies to increase the
-effective sample size. It is possible that there were small associations
-within each study that were not statistically significant because the
-individual studies lacked sufficient power. Alternatively, diversity
-metrics may mask the appropriate signal and it is necessary to measure
-the association at the level of microbial populations. Walters et al.
-(8) demonstrated that Random Forest machine learning models were capable
-of predicting obesity status within a single cohort, but did not attempt
-to test the models on other cohorts. The purpose of this study was to
+question relating the structure of the human microbiome to obesity. One
+critique of the prior re-analysis studies is that the authors did not
+aggregate the results across studies to increase the effective sample
+size. It is possible that there were small associations within each
+study that were not statistically significant because the individual
+studies lacked sufficient power. Alternatively, diversity metrics may
+mask the appropriate signal and it is necessary to measure the
+association at the level of microbial populations. Walters et al. (8)
+demonstrated that Random Forest machine learning models were capable of
+predicting obesity status within a single cohort, but did not attempt to
+test the models on other cohorts. The purpose of this study was to
 perform a meta-analysis of the association between differences in the
@@
 
-\subsection{Methods}\label{methods}
+\subsection{Results}\label{results}
 
-\textbf{\emph{Literature Review and Study Inclusion.}} We followed the
-Preferred Reporting Items for Systematic Reviews and Meta-Analyses
-(PRISMA) guidelines to identify studies to include in our meta-analysis
-(23). A detailed description of our selection process and the exact
-search terms are provided in the Supplement and in figure 1. Briefly, we
-searched PubMed for original research studies that involved studying
-obesity and the human microbiome. The initial search yielded 187
-studies. We identified number\_string{[}n{]} additional studies that
-were not designed to explicitly test for an association between the
-microbiome and obesity. We then manually curated the 196 studies to
-select those studies that included BMI and sequence data. This yielded
-10 eligible studies. An additional study was removed from our analysis
-because no individuals in the study had a BMI over 30. Among the final
-10 studies, 3 were from identified from our PubMed search (6, 15, 20), 5
-were originally identified from the 9 studies that did not explicitly
-investigate obesity but included BMI data (14, 18, 19, 24, 25), and two
-datasets were used (21, 22) because these publications did not
-specifically look for any metabolic or obesity conditions but had
-control populations and enabled us to help mitigate against publication
-biases associated with the bacterial microbiome and obesity. The
-number\_string{[}n{]} studies are summarized in Tables 1 and 2.
+\textbf{\emph{Meta-analysis}}
 
-\textbf{\emph{Sequence Analysis Pipeline.}} All sequence data were
-publicly available and were downloaded from the NCBI Sequence Read
-Archive, the European Nucleotide Archive, or the investigators' personal
-website
-(\url{https://gordonlab.wustl.edu/TurnbaughSE/_10/_09/STM/_2009.html}).
-In total seven studies used 454 (6, 14, 15, 19, 20, 22, 25) and three
-studies used Illumina sequencing (18, 21, 24). All of these studies used
-amplification-based 16S rRNA gene sequencing. Among the studies that
-sequenced the 16S rRNA gene, the researchers targeted the V1-V2 (19),
-V1-V3 (14, 15, 20), V3-V5 (22, 25), V4 {[}(18); (21); {]}, and V3-4 (24)
-regions. For those studies where multiple regions were sequenced, we
-selected the region that corresponded to the largest number of subjects
-(6, 25). We processed the 16S rRNA gene sequence data using a
-standardized mothur pipeline. Briefly, our pipelines attempted to follow
-previously recommended approaches for 454 and Illumina sequencing data
-(26, 27). All sequences were screened for chimeras using UCHIME and
-assigned to operational taxonomic units (OTUs) using the average
-neighbor algorithm using a 3\% distance threshold (28, 29). All sequence
-processing was performed using mothur (v.1.37.0) (30).
-
-\textbf{\emph{Data Analysis.}} We split the overall meta-analysis into
-three general strategies using R (3.3.0). First, we followed the
-approach employed by Finucane et al (9) and Walters et al (8) where each
-study was re-analyzed separately to identify associations between BMI
-and the relative abundance of Bacteroidetes and Firmicutes, the ratio of
-Bacteroidetes and Firmicutes relative abundances (B:F), Shannon
-diversity, observed richness, and Shannon evenness. After each variable
-was transformed to fit a normal distribution a two-tailed t-test was
-performed for comparison of non-obese and obese individuals (i.e.~BMI
-\textgreater{} 35.0). We performed a pooled analysis on these measured
-variables using linear random effect models to correct for study effect
-to asses differences on the combined dataset between non-obese and obese
-groups using the lme4 (v.1.1-12) R package. Next, we compared the
-community structure from non-obese and obese individuals using PERMANOVA
-analysis of Bray-Curtis distance matrices. This analysis was performed
-using the vegan (v.2.3-5) R package. For both analyses, the datasets
-were rarefied (N=1000) so that each study within a study had the same
-number of sequences. Second, for each study we partitioned the subjects
-into a low or high group depending on whether their alpha diversity
-metrics were below or above the median value for the study. The relative
-risk (RR) was then calculated as the ratio of the number of obese
-individuals in the low group to the number of obese individuals in the
-high group. We then performed a Fisher exact-test to investigate whether
-the RR was significantly different from 1.0 within each study and across
-all of the studies using the epiR (0.9-77) and metafor (1.9-8) packages.
-Third, we used the AUCRF (1.1) R package to generate Random Forest
-models. For each study we developed models using either OTUs or
-genus-level phylotypes. The quality of each model was assessed by
-measuring the area under the curve (AUC) of the Receive Operating
-Characteristic (ROC) using ten-fold cross validation. Because the
-genus-level phylotype models were developed using a common reference, it
-was possible to use one study's model (i.e.~the training set) to
-classify the samples from the other studies (i.e.~the testing sets). The
-optimum threshold for the training set was set as the probability
-threshold that had the highest combined sensitivity and specificity.
-This threshold was then used to calculate the accuracy of the model
-applied to the test studies. To generate Receiver Operator
-Characteristic (ROC) curves and calculate the accuracy of the models we
-used the pROC (1.8) R package. Finally, we performed power and sample
-number simulations for different effect sizes for each study using the
-pwr (1.1-3) R package and base R functions. We also calculated the
-actual sample size needed based on the effect size of each individual
-study.
-
-\textbf{\emph{Reproducible methods.}} A detailed and reproducible
-description of how the data were processed and analyzed can be found at
-\url{https://github.com/SchlossLab/Sze_Obesity_mBio_2016/}.
-
-\subsection{Results}\label{results}
+(10, 14--22)
 
@@
 the membership and structure of the individuals from each study. We then
-used PERMANOVA to test for a significant differences between the
-structure of non-obese and obese individuals. The Escobar, Goodrich, and
-Turnbaugh datasets indicated a significant difference in community
-structure (all P\textless{}0.05). Because it was not possible to
-ascertain the directionality of the difference in community structure
-nor perform a pooled analysis using studies that had non-overlapping 16S
-rRNA gene sequence regions it is unclear whether these differences
-reflect a broader, but perhaps small, shift in community structure
-between non-obese and obese individuals.
+used AMOVA to test for a significant differences between the structure
+of non-obese and obese individuals. The Escobar, Goodrich, and Turnbaugh
+datasets indicated a significant difference in community structure (all
+P\textless{}0.05). Because it was not possible to ascertain the
+directionality of the difference in community structure nor perform a
+pooled analysis using studies that had non-overlapping 16S rRNA gene
+sequence regions it is unclear whether these differences reflect a
+broader, but perhaps small, shift in community structure between
+non-obese and obese individuals.
 
@@
 rate of a significant finding in an underpowered study is small leading
-to results that cannot be reproduced (31). Future microbiome studies
+to results that cannot be reproduced (23). Future microbiome studies
 should articulate the basis for their experimental design.
@@
 studies whereas the Finucane study used 16S rRNA gene sequence data from
-3 studies (7, 10, 25) and a metagenomic study (32) and the Walters study
-used 16S rRNA gene sequence data from 5 studies (10, 15, 19, 25, 33);
-two studies were included in both analyses (10, 25). Our analysis
-included 4 of these studies (10, 15, 19, 25) and excluded 3 of the
+3 studies (7, 10, 22) and a metagenomic study (24) and the Walters study
+used 16S rRNA gene sequence data from 5 studies (10, 15, 17, 22, 25);
+two studies were included in both analyses (10, 22). Our analysis
+included 4 of these studies (10, 15, 17, 22) and excluded 3 of the
 studies because they were too small (7), only utilized metagenomic data
-(32), or used short single read Illumina HiSeq data that has a high
-error rate making it untractable for \emph{de novo} OTU clustering (33).
+(24), or used short single read Illumina HiSeq data that has a high
+error rate making it untractable for \emph{de novo} OTU clustering (25).
 The additional seven datasets were published after the two reviews were
@@
 
+\subsection{Methods}\label{methods}
+
+\textbf{\emph{Literature Review and Study Inclusion.}} We followed the
+Preferred Reporting Items for Systematic Reviews and Meta-Analyses
+(PRISMA) guidelines to identify studies to include in our meta-analysis
+(26). A detailed description of our selection process and the exact
+search terms are provided in the Supplement and in figure 1. Briefly, we
+searched PubMed for original research studies that involved studying
+obesity and the human microbiome. The initial search yielded 187
+studies. We identified number\_string{[}n{]} additional studies that
+were not designed to explicitly test for an association between the
+microbiome and obesity. We then manually curated the 196 studies to
+select those studies that included BMI and sequence data. This yielded
+10 eligible studies. An additional study was removed from our analysis
+because no individuals in the study had a BMI over 30. Among the final
+10 studies, 3 were from identified from our PubMed search (6, 15, 18), 5
+were originally identified from the 9 studies that did not explicitly
+investigate obesity but included BMI data (14, 16, 17, 21, 22), and two
+datasets were used (19, 20) because these publications did not
+specifically look for any metabolic or obesity conditions but had
+control populations and enabled us to help mitigate against publication
+biases associated with the bacterial microbiome and obesity. The
+number\_string{[}n{]} studies are summarized in Tables 1 and 2.
+
+\textbf{\emph{Sequence Analysis Pipeline.}} All sequence data were
+publicly available and were downloaded from the NCBI Sequence Read
+Archive, the European Nucleotide Archive, or the investigators' personal
+website
+(\url{https://gordonlab.wustl.edu/TurnbaughSE/_10/_09/STM/_2009.html}).
+In total seven studies used 454 (6, 14, 15, 17, 18, 20, 22) and three
+studies used Illumina sequencing (16, 19, 21). All of these studies used
+amplification-based 16S rRNA gene sequencing. Among the studies that
+sequenced the 16S rRNA gene, the researchers targeted the V1-V2 (17),
+V1-V3 (14, 15, 18), V3-V5 (20, 22), V4 {[}(16); (19); {]}, and V3-4 (21)
+regions. For those studies where multiple regions were sequenced, we
+selected the region that corresponded to the largest number of subjects
+(6, 22). We processed the 16S rRNA gene sequence data using a
+standardized mothur pipeline. Briefly, our pipelines attempted to follow
+previously recommended approaches for 454 and Illumina sequencing data
+(27, 28). All sequences were screened for chimeras using UCHIME and
+assigned to operational taxonomic units (OTUs) using the average
+neighbor algorithm using a 3\% distance threshold (29, 30). All sequence
+processing was performed using mothur (v.1.37.0) (31).
+
+\textbf{\emph{Data Analysis.}} We split the overall meta-analysis into
+three general strategies using R (3.3.0). First, we followed the
+approach employed by Finucane et al (9) and Walters et al (8) where each
+study was re-analyzed separately to identify associations between BMI
+and the relative abundance of Bacteroidetes and Firmicutes, the ratio of
+Bacteroidetes and Firmicutes relative abundances (B:F), Shannon
+diversity, observed richness, and Shannon evenness. After each variable
+was transformed to fit a normal distribution a two-tailed t-test was
+performed for comparison of non-obese and obese individuals (i.e.~BMI
+\textgreater{} 35.0). We performed a pooled analysis on these measured
+variables using linear random effect models to correct for study effect
+to asses differences on the combined dataset between non-obese and obese
+groups using the lme4 (v.1.1-12) R package. Next, we compared the
+community structure from non-obese and obese individuals using analysis
+of molecular variance (AMOVA) with Bray-Curtis distance matrices (32).
+This analysis was performed using the vegan (v.2.3-5) R package. For
+both analyses, the datasets were rarefied (N=1000) so that each study
+within a study had the same number of sequences. Second, for each study
+we partitioned the subjects into a low or high group depending on
+whether their alpha diversity metrics were below or above the median
+value for the study. The relative risk (RR) was then calculated as the
+ratio of the number of obese individuals in the low group to the number
+of obese individuals in the high group. We then performed a Fisher
+exact-test to investigate whether the RR was significantly different
+from 1.0 within each study and across all of the studies using the epiR
+(0.9-77) and metafor (1.9-8) packages. Third, we used the AUCRF (1.1) R
+package to generate Random Forest models. For each study we developed
+models using either OTUs or genus-level phylotypes. The quality of each
+model was assessed by measuring the area under the curve (AUC) of the
+Receive Operating Characteristic (ROC) using ten-fold cross validation.
+Because the genus-level phylotype models were developed using a common
+reference, it was possible to use one study's model (i.e.~the training
+set) to classify the samples from the other studies (i.e.~the testing
+sets). The optimum threshold for the training set was set as the
+probability threshold that had the highest combined sensitivity and
+specificity. This threshold was then used to calculate the accuracy of
+the model applied to the test studies. To generate Receiver Operator
+Characteristic (ROC) curves and calculate the accuracy of the models we
+used the pROC (1.8) R package. Finally, we performed power and sample
+number simulations for different effect sizes for each study using the
+pwr (1.1-3) R package and base R functions. We also calculated the
+actual sample size needed based on the effect size of each individual
+study.
+
+\textbf{\emph{Reproducible methods.}} A detailed and reproducible
+description of how the data were processed and analyzed can be found at
+\url{https://github.com/SchlossLab/Sze_Obesity_mBio_2016/}.
+
 \subsection{Acknowledgements}\label{acknowledgements}
@@
 
-\textbf{Table 1. Summary Demographics of Individuals used in the
-Meta-analysis.}
-
-\newpage
-
-\textbf{Figure 1: PRISMA flow diagram of total records searched (34).}
+\textbf{Figure 1: PRISMA flow diagram of total records searched (33).}
 
@@
 
-\hyperdef{}{references}{\label{references}}
 \subsection*{References}\label{references}
@@
 
-\hyperdef{}{ref-ogdenux5fprevalenceux5f2014}{\label{ref-ogdenux5fprevalenceux5f2014}}
+\hypertarget{refs}{}
+\hypertarget{ref-ogdenux5fprevalenceux5f2014}{}
 1. \textbf{Ogden CL}, \textbf{Carroll MD}, \textbf{Kit BK},
@@
 the United States, 2011-2012. JAMA \textbf{311}:806--814.
-doi:\url{http://doi.org/10.1001/jama.2014.732}.
+doi:\href{https://doi.org/10.1001/jama.2014.732}{10.1001/jama.2014.732}.
 
-\hyperdef{}{ref-lichtashux5fbodyux5f2013}{\label{ref-lichtashux5fbodyux5f2013}}
+\hypertarget{ref-lichtashux5fbodyux5f2013}{}
 2. \textbf{Lichtash CT}, \textbf{Cui J}, \textbf{Guo X}, \textbf{Chen
@@
 PloS One \textbf{8}:e65954.
-doi:\url{http://doi.org/10.1371/journal.pone.0065954}.
+doi:\href{https://doi.org/10.1371/journal.pone.0065954}{10.1371/journal.pone.0065954}.
 
-\hyperdef{}{ref-braheux5fcanux5f2016}{\label{ref-braheux5fcanux5f2016}}
+\hypertarget{ref-braheux5fcanux5f2016}{}
 3. \textbf{Brahe LK}, \textbf{Astrup A}, \textbf{Larsen LH}. 2016. Can
@@
 the Gut Microbiota? Advances in Nutrition (Bethesda, Md)
-\textbf{7}:90--101. doi:\url{http://doi.org/10.3945/an.115.010587}.
+\textbf{7}:90--101.
+doi:\href{https://doi.org/10.3945/an.115.010587}{10.3945/an.115.010587}.
 
-\hyperdef{}{ref-drorux5fmicrobiotaux5f2016}{\label{ref-drorux5fmicrobiotaux5f2016}}
+\hypertarget{ref-drorux5fmicrobiotaux5f2016}{}
 4. \textbf{Dror T}, \textbf{Dickstein Y}, \textbf{Dubourg G},
@@
 Microbial Pathogenesis.
-doi:\url{http://doi.org/10.1016/j.micpath.2016.01.002}.
+doi:\href{https://doi.org/10.1016/j.micpath.2016.01.002}{10.1016/j.micpath.2016.01.002}.
 
-\hyperdef{}{ref-leyux5fobesityux5f2005}{\label{ref-leyux5fobesityux5f2005}}
+\hypertarget{ref-leyux5fobesityux5f2005}{}
 5. \textbf{Ley RE}, \textbf{Bäckhed F}, \textbf{Turnbaugh P},
@@
 \textbf{102}:11070--11075.
-doi:\url{http://doi.org/10.1073/pnas.0504978102}.
+doi:\href{https://doi.org/10.1073/pnas.0504978102}{10.1073/pnas.0504978102}.
 
-\hyperdef{}{ref-turnbaughux5fcoreux5f2009}{\label{ref-turnbaughux5fcoreux5f2009}}
+\hypertarget{ref-turnbaughux5fcoreux5f2009}{}
 6. \textbf{Turnbaugh PJ}, \textbf{Hamady M}, \textbf{Yatsunenko T},
@@
 obese and lean twins. Nature \textbf{457}:480--484.
-doi:\url{http://doi.org/10.1038/nature07540}.
+doi:\href{https://doi.org/10.1038/nature07540}{10.1038/nature07540}.
 
-\hyperdef{}{ref-leyux5fmicrobialux5f2006}{\label{ref-leyux5fmicrobialux5f2006}}
+\hypertarget{ref-leyux5fmicrobialux5f2006}{}
 7. \textbf{Ley RE}, \textbf{Turnbaugh PJ}, \textbf{Klein S},
@@
 associated with obesity. Nature \textbf{444}:1022--1023.
-doi:\url{http://doi.org/10.1038/4441022a}.
+doi:\href{https://doi.org/10.1038/4441022a}{10.1038/4441022a}.
 
-\hyperdef{}{ref-waltersux5fmeta-analysesux5f2014}{\label{ref-waltersux5fmeta-analysesux5f2014}}
+\hypertarget{ref-waltersux5fmeta-analysesux5f2014}{}
 8. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
@@
 FEBS letters \textbf{588}:4223--4233.
-doi:\url{http://doi.org/10.1016/j.febslet.2014.09.039}.
+doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.
 
-\hyperdef{}{ref-finucaneux5ftaxonomicux5f2014}{\label{ref-finucaneux5ftaxonomicux5f2014}}
+\hypertarget{ref-finucaneux5ftaxonomicux5f2014}{}
 9. \textbf{Finucane MM}, \textbf{Sharpton TJ}, \textbf{Laurent TJ},
@@
 \textbf{9}:e84689.
-doi:\url{http://doi.org/10.1371/journal.pone.0084689}.
+doi:\href{https://doi.org/10.1371/journal.pone.0084689}{10.1371/journal.pone.0084689}.
 
-\hyperdef{}{ref-turnbaughux5fobesity-associatedux5f2006}{\label{ref-turnbaughux5fobesity-associatedux5f2006}}
+\hypertarget{ref-turnbaughux5fobesity-associatedux5f2006}{}
 10. \textbf{Turnbaugh PJ}, \textbf{Ley RE}, \textbf{Mahowald MA},
@@
 harvest. Nature \textbf{444}:1027--31.
-doi:\url{http://doi.org/10.1038/nature05414}.
+doi:\href{https://doi.org/10.1038/nature05414}{10.1038/nature05414}.
 
-\hyperdef{}{ref-Koren2012}{\label{ref-Koren2012}}
+\hypertarget{ref-Koren2012}{}
 11. \textbf{Koren O}, \textbf{Goodrich JK}, \textbf{Cullender TC},
@@
 \textbf{150}:470--480.
-doi:\url{http://doi.org/10.1016/j.cell.2012.07.008}.
+doi:\href{https://doi.org/10.1016/j.cell.2012.07.008}{10.1016/j.cell.2012.07.008}.
 
-\hyperdef{}{ref-Cox2014}{\label{ref-Cox2014}}
+\hypertarget{ref-Cox2014}{}
 12. \textbf{Cox LM}, \textbf{Yamanishi S}, \textbf{Sohn J},
@@
 \textbf{158}:705--721.
-doi:\url{http://doi.org/10.1016/j.cell.2014.05.052}.
+doi:\href{https://doi.org/10.1016/j.cell.2014.05.052}{10.1016/j.cell.2014.05.052}.
 
-\hyperdef{}{ref-Mahana2016}{\label{ref-Mahana2016}}
+\hypertarget{ref-Mahana2016}{}
 13. \textbf{Mahana D}, \textbf{Trent CM}, \textbf{Kurtz ZD},
@@
 high-fat diet. Genome Medicine \textbf{8}.
-doi:\url{http://doi.org/10.1186/s13073-016-0297-9}.
+doi:\href{https://doi.org/10.1186/s13073-016-0297-9}{10.1186/s13073-016-0297-9}.
 
-\hyperdef{}{ref-rossux5f16sux5f2015}{\label{ref-rossux5f16sux5f2015}}
+\hypertarget{ref-rossux5f16sux5f2015}{}
 14. \textbf{Ross MC}, \textbf{Muzny DM}, \textbf{McCormick JB},
@@
 16S gut community of the Cameron County Hispanic Cohort. Microbiome
-\textbf{3}:7. doi:\url{http://doi.org/10.1186/s40168-015-0072-y}.
+\textbf{3}:7.
+doi:\href{https://doi.org/10.1186/s40168-015-0072-y}{10.1186/s40168-015-0072-y}.
 
-\hyperdef{}{ref-zupancicux5fanalysisux5f2012}{\label{ref-zupancicux5fanalysisux5f2012}}
+\hypertarget{ref-zupancicux5fanalysisux5f2012}{}
 15. \textbf{Zupancic ML}, \textbf{Cantarel BL}, \textbf{Liu Z},
@@
 syndrome. PloS One \textbf{7}:e43052.
-doi:\url{http://doi.org/10.1371/journal.pone.0043052}.
+doi:\href{https://doi.org/10.1371/journal.pone.0043052}{10.1371/journal.pone.0043052}.
 
-\hyperdef{}{ref-namux5fcomparativeux5f2011}{\label{ref-namux5fcomparativeux5f2011}}
-16. \textbf{Nam Y-D}, \textbf{Jung M-J}, \textbf{Roh SW}, \textbf{Kim
-M-S}, \textbf{Bae J-W}. 2011. Comparative analysis of Korean human gut
-microbiota by barcoded pyrosequencing. PloS One \textbf{6}:e22109.
-doi:\url{http://doi.org/10.1371/journal.pone.0022109}.
-
-\hyperdef{}{ref-arumugamux5fenterotypesux5f2011}{\label{ref-arumugamux5fenterotypesux5f2011}}
-17. \textbf{Arumugam M}, \textbf{Raes J}, \textbf{Pelletier E},
-\textbf{Le Paslier D}, \textbf{Yamada T}, \textbf{Mende DR},
-\textbf{Fernandes GR}, \textbf{Tap J}, \textbf{Bruls T}, \textbf{Batto
-J-M}, \textbf{Bertalan M}, \textbf{Borruel N}, \textbf{Casellas F},
-\textbf{Fernandez L}, \textbf{Gautier L}, \textbf{Hansen T},
-\textbf{Hattori M}, \textbf{Hayashi T}, \textbf{Kleerebezem M},
-\textbf{Kurokawa K}, \textbf{Leclerc M}, \textbf{Levenez F},
-\textbf{Manichanh C}, \textbf{Nielsen HB}, \textbf{Nielsen T},
-\textbf{Pons N}, \textbf{Poulain J}, \textbf{Qin J},
-\textbf{Sicheritz-Ponten T}, \textbf{Tims S}, \textbf{Torrents D},
-\textbf{Ugarte E}, \textbf{Zoetendal EG}, \textbf{Wang J},
-\textbf{Guarner F}, \textbf{Pedersen O}, \textbf{Vos WM de},
-\textbf{Brunak S}, \textbf{Doré J}, \textbf{MetaHIT Consortium},
-\textbf{Antolín M}, \textbf{Artiguenave F}, \textbf{Blottiere HM},
-\textbf{Almeida M}, \textbf{Brechot C}, \textbf{Cara C},
-\textbf{Chervaux C}, \textbf{Cultrone A}, \textbf{Delorme C},
-\textbf{Denariaz G}, \textbf{Dervyn R}, \textbf{Foerstner KU},
-\textbf{Friss C}, \textbf{Guchte M van de}, \textbf{Guedon E},
-\textbf{Haimet F}, \textbf{Huber W}, \textbf{Hylckama-Vlieg J van},
-\textbf{Jamet A}, \textbf{Juste C}, \textbf{Kaci G}, \textbf{Knol J},
-\textbf{Lakhdari O}, \textbf{Layec S}, \textbf{Le Roux K},
-\textbf{Maguin E}, \textbf{Mérieux A}, \textbf{Melo Minardi R},
-\textbf{M'rini C}, \textbf{Muller J}, \textbf{Oozeer R},
-\textbf{Parkhill J}, \textbf{Renault P}, \textbf{Rescigno M},
-\textbf{Sanchez N}, \textbf{Sunagawa S}, \textbf{Torrejon A},
-\textbf{Turner K}, \textbf{Vandemeulebrouck G}, \textbf{Varela E},
-\textbf{Winogradsky Y}, \textbf{Zeller G}, \textbf{Weissenbach J},
-\textbf{Ehrlich SD}, \textbf{Bork P}. 2011. Enterotypes of the human gut
-microbiome. Nature \textbf{473}:174--180.
-doi:\url{http://doi.org/10.1038/nature09944}.
-
-\hyperdef{}{ref-goodrichux5fhumanux5f2014}{\label{ref-goodrichux5fhumanux5f2014}}
-18. \textbf{Goodrich JK}, \textbf{Waters JL}, \textbf{Poole AC},
+\hypertarget{ref-goodrichux5fhumanux5f2014}{}
+16. \textbf{Goodrich JK}, \textbf{Waters JL}, \textbf{Poole AC},
 \textbf{Sutter JL}, \textbf{Koren O}, \textbf{Blekhman R},
@@
 \textbf{159}:789--799.
-doi:\url{http://doi.org/10.1016/j.cell.2014.09.053}.
+doi:\href{https://doi.org/10.1016/j.cell.2014.09.053}{10.1016/j.cell.2014.09.053}.
 
-\hyperdef{}{ref-wuux5flinkingux5f2011}{\label{ref-wuux5flinkingux5f2011}}
-19. \textbf{Wu GD}, \textbf{Chen J}, \textbf{Hoffmann C},
+\hypertarget{ref-wuux5flinkingux5f2011}{}
+17. \textbf{Wu GD}, \textbf{Chen J}, \textbf{Hoffmann C},
 \textbf{Bittinger K}, \textbf{Chen Y-Y}, \textbf{Keilbaugh SA},
@@
 patterns with gut microbial enterotypes. Science (New York, NY)
-\textbf{334}:105--108. doi:\url{http://doi.org/10.1126/science.1208344}.
+\textbf{334}:105--108.
+doi:\href{https://doi.org/10.1126/science.1208344}{10.1126/science.1208344}.
 
-\hyperdef{}{ref-escobarux5fgutux5f2014}{\label{ref-escobarux5fgutux5f2014}}
-20. \textbf{Escobar JS}, \textbf{Klotz B}, \textbf{Valdes BE},
+\hypertarget{ref-escobarux5fgutux5f2014}{}
+18. \textbf{Escobar JS}, \textbf{Klotz B}, \textbf{Valdes BE},
 \textbf{Agudelo GM}. 2014. The gut microbiota of Colombians differs from
 that of Americans, Europeans and Asians. BMC microbiology
-\textbf{14}:311. doi:\url{http://doi.org/10.1186/s12866-014-0311-6}.
+\textbf{14}:311.
+doi:\href{https://doi.org/10.1186/s12866-014-0311-6}{10.1186/s12866-014-0311-6}.
 
-\hyperdef{}{ref-baxterux5fmicrobiota-basedux5f2016}{\label{ref-baxterux5fmicrobiota-basedux5f2016}}
-21. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\hypertarget{ref-baxterux5fmicrobiota-basedux5f2016}{}
+19. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 Genome Medicine \textbf{8}:37.
-doi:\url{http://doi.org/10.1186/s13073-016-0290-3}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
 
-\hyperdef{}{ref-schubertux5fmicrobiomeux5f2014}{\label{ref-schubertux5fmicrobiomeux5f2014}}
-22. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
+\hypertarget{ref-schubertux5fmicrobiomeux5f2014}{}
+20. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
 \textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
@@
 \textbf{5}:e01021--01014.
-doi:\url{http://doi.org/10.1128/mBio.01021-14}.
+doi:\href{https://doi.org/10.1128/mBio.01021-14}{10.1128/mBio.01021-14}.
 
-\hyperdef{}{ref-moherux5fpreferredux5f2010}{\label{ref-moherux5fpreferredux5f2010}}
-23. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
-\textbf{Altman DG}, \textbf{PRISMA Group}. 2010. Preferred reporting
-items for systematic reviews and meta-analyses: The PRISMA statement.
-International Journal of Surgery (London, England) \textbf{8}:336--341.
-doi:\url{http://doi.org/10.1016/j.ijsu.2010.02.007}.
-
-\hyperdef{}{ref-zeeviux5fpersonalizedux5f2015}{\label{ref-zeeviux5fpersonalizedux5f2015}}
-24. \textbf{Zeevi D}, \textbf{Korem T}, \textbf{Zmora N},
+\hypertarget{ref-zeeviux5fpersonalizedux5f2015}{}
+21. \textbf{Zeevi D}, \textbf{Korem T}, \textbf{Zmora N},
 \textbf{Israeli D}, \textbf{Rothschild D}, \textbf{Weinberger A},
@@
 Prediction of Glycemic Responses. Cell \textbf{163}:1079--1094.
-doi:\url{http://doi.org/10.1016/j.cell.2015.11.001}.
+doi:\href{https://doi.org/10.1016/j.cell.2015.11.001}{10.1016/j.cell.2015.11.001}.
 
-\hyperdef{}{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{\label{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}}
-25. \textbf{Human Microbiome Project Consortium}. 2012. Structure,
+\hypertarget{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{}
+22. \textbf{Human Microbiome Project Consortium}. 2012. Structure,
 function and diversity of the healthy human microbiome. Nature
-\textbf{486}:207--214. doi:\url{http://doi.org/10.1038/nature11234}.
+\textbf{486}:207--214.
+doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
 
-\hyperdef{}{ref-Kozichux5f2013}{\label{ref-Kozichux5f2013}}
-26. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
-\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
-dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq Illumina sequencing platform.
-Applied and environmental microbiology \textbf{79}:5112--5120.
+\hypertarget{ref-Ioannidis2005}{}
+23. \textbf{Ioannidis JPA}. 2005. Why most published research findings
+are false. PLoS Med \textbf{2}:e124.
+doi:\href{https://doi.org/10.1371/journal.pmed.0020124}{10.1371/journal.pmed.0020124}.
 
-\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
-27. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
-Reducing the effects of PCR amplification and sequencing artifacts on
-16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
-doi:\url{http://doi.org/10.1371/journal.pone.0027310}.
-
-\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
-28. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
-\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
-and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
-doi:\url{http://doi.org/10.1093/bioinformatics/btr381}.
-
-\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
-29. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
-methods outperform reference-based methods for assigning 16S rRNA gene
-sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\url{http://doi.org/10.7717/peerj.1487}.
-
-\hyperdef{}{ref-mothur}{\label{ref-mothur}}
-30. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{others}. 2009. Introducing mothur:
-open-source, platform-independent, community-supported software for
-describing and comparing microbial communities. Applied and
-environmental microbiology \textbf{75}:7537--7541.
-
-\hyperdef{}{ref-Ioannidis2005}{\label{ref-Ioannidis2005}}
-31. \textbf{Ioannidis JPA}. 2005. Why most published research findings
-are false. PLoS Med \textbf{2}:e124.
-doi:\url{http://doi.org/10.1371/journal.pmed.0020124}.
-
-\hyperdef{}{ref-Qin2010}{\label{ref-Qin2010}}
-32. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
+\hypertarget{ref-Qin2010}{}
+24. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
 \textbf{Burgdorf KS}, \textbf{Manichanh C}, \textbf{Nielsen T},
@@
 catalogue established by metagenomic sequencing. Nature
-\textbf{464}:59--65. doi:\url{http://doi.org/10.1038/nature08821}.
+\textbf{464}:59--65.
+doi:\href{https://doi.org/10.1038/nature08821}{10.1038/nature08821}.
 
-\hyperdef{}{ref-yatsunenkoux5fhumanux5f2012}{\label{ref-yatsunenkoux5fhumanux5f2012}}
-33. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
+\hypertarget{ref-yatsunenkoux5fhumanux5f2012}{}
+25. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
 \textbf{Trehan I}, \textbf{Dominguez-Bello MG}, \textbf{Contreras M},
@@
 gut microbiome viewed across age and geography. Nature
-\textbf{486}:222--227. doi:\url{http://doi.org/10.1038/nature11053}.
+\textbf{486}:222--227.
+doi:\href{https://doi.org/10.1038/nature11053}{10.1038/nature11053}.
 
-\hyperdef{}{ref-Moher2009}{\label{ref-Moher2009}}
-34. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
+\hypertarget{ref-moherux5fpreferredux5f2010}{}
+26. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
+\textbf{Altman DG}, \textbf{PRISMA Group}. 2010. Preferred reporting
+items for systematic reviews and meta-analyses: The PRISMA statement.
+International Journal of Surgery (London, England) \textbf{8}:336--341.
+doi:\href{https://doi.org/10.1016/j.ijsu.2010.02.007}{10.1016/j.ijsu.2010.02.007}.
+
+\hypertarget{ref-Kozichux5f2013}{}
+27. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq Illumina sequencing platform.
+Applied and environmental microbiology \textbf{79}:5112--5120.
+
+\hypertarget{ref-Schloss2011}{}
+28. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
+Reducing the effects of PCR amplification and sequencing artifacts on
+16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
+doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
+
+\hypertarget{ref-Edgar2011}{}
+29. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
+\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
+and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
+doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
+
+\hypertarget{ref-Westcott2015}{}
+30. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
+methods outperform reference-based methods for assigning 16S rRNA gene
+sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
+doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
+
+\hypertarget{ref-mothur}{}
+31. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{others}. 2009. Introducing mothur:
+open-source, platform-independent, community-supported software for
+describing and comparing microbial communities. Applied and
+environmental microbiology \textbf{75}:7537--7541.
+
+\hypertarget{ref-Anderson2001}{}
+32. \textbf{Anderson MJ}. 2001. A new method for non-parametric
+multivariate analysis of variance. Austral Ecology \textbf{26}:32--46.
+doi:\href{https://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}{10.1111/j.1442-9993.2001.01070.pp.x}.
+
+\hypertarget{ref-Moher2009}{}
+33. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
 \textbf{Altman DG}. 2009. Preferred reporting items for systematic
@@
 \textbf{6}:e1000097.
-doi:\url{http://doi.org/10.1371/journal.pmed.1000097}.
+doi:\href{https://doi.org/10.1371/journal.pmed.1000097}{10.1371/journal.pmed.1000097}.
 

### 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
Run make write.paper

--- 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
@@
 significant associations were observed between Shannon diversity, number
-of observed OTUs, and Shannon evenness and obesity status They were not
+of observed OTUs, and Shannon evenness and obesity status. They were not
 observed for the ratio of \emph{Bacteroidetes} and \emph{Firmicutes} or
@@
 non-obese and obese individuals was 2.07\%. A power analysis
-demonstrated that one of the studies had sufficient power to detect a
-5\% difference in diversity. When models trained on one dataset were
-then tested using the other 9 datasets, the median accuracy varied
-between 33.01 and 64.77\% (median=56.67\%). Although there is
-statistical support for a relationship between the microbial communities
-found in human feces and obesity status, this association is relatively
-weak and its detection is confounded by large interpersonal variation
-and insufficient sample sizes.
+demonstrated that only one of the studies had sufficient power to detect
+a 5\% difference in diversity. When Random Forest machine learning
+models were trained on one dataset and then tested using the other 9
+datasets, the median accuracy varied between 33.01 and 64.77\%
+(median=56.67\%). Although there was support for a relationship between
+the microbial communities found in human feces and obesity status, this
+association is relatively weak and its detection is confounded by large
+interpersonal variation and insufficient sample sizes.
 
@@
 The first such study used genetically obese mice and observed the ratio
-of the relative abundances of Bacteroidetes to Firmicutes (B:F) was
-lower in obese mice than lean mice (5). Translation of this result to
-humans by the same researchers did not observe this effect, but did find
-that obese individuals had a lower diversity than lean individuals (6).
-They also showed that the relative abundance of Bacteroidetes and
-Firmicutes increased and decreased, respectively, as obese individuals
-lost weight while on a fat or carbohydrate restricted diet (7). Two
-re-analysis studies interrogated previously published microbiome and
-obesity data and concluded that the previously reported differences in
-community diversity and B:F among non-obese and obese individuals could
-not be generalized (8, 9). Regardless of the results using human
-populations, mechanistic studies using animal models that were
-manipulated with antibiotics or colonization with varied communities
-appears to support the association since direct manipulation of the
-communities yielded variation in animal weight (10--13). The purported
-association between the differences in the microbiome and obesity have
-been widely repeated with little attention given to the lack of a clear
-signal in human cohort studies.
+of the relative abundances of \emph{Bacteroidetes} to \emph{Firmicutes}
+(B:F) was lower in obese mice than lean mice (5). Translation of this
+result to humans by the same researchers did not observe this effect,
+but did find that obese individuals had a lower diversity than lean
+individuals (6). They also showed that the relative abundance of
+\emph{Bacteroidetes} and \emph{Firmicutes} increased and decreased,
+respectively, as obese individuals lost weight while on a fat or
+carbohydrate restricted diet (7). Two re-analysis studies interrogated
+previously published microbiome and obesity data and concluded that the
+previously reported differences in community diversity and B:F among
+non-obese and obese individuals could not be generalized (8, 9).
+Regardless of the results using human populations, mechanistic studies
+using animal models that were manipulated with antibiotics or
+colonization with varied communities appears to support the association
+since these manipulation yielded variation in animal weight (10--13).
+The purported association between the differences in the microbiome and
+obesity have been widely repeated with little attention given to the
+lack of a clear signal in human cohort studies.
 
@@
 
-\subsection{Methods}\label{methods}
+\subsection{Results}\label{results}
 
@@
 (PRISMA) guidelines to identify studies to include in our meta-analysis
-(23). A detailed description of our selection process and the exact
-search terms are provided in the Supplement and in figure 1. Briefly, we
-searched PubMed for original research studies that involved studying
-obesity and the human microbiome. The initial search yielded 187
-studies. We identified number\_string{[}n{]} additional studies that
+(14). A detailed description of our selection process and the exact
+search terms are provided in the Supplemental Text and in Figure 1.
+Briefly, we searched PubMed for original research studies that involved
+studying obesity and the human microbiome. The initial search yielded
+187 studies. We identified number\_string{[}n{]} additional studies that
 were not designed to explicitly test for an association between the
@@
 because no individuals in the study had a BMI over 30. Among the final
-10 studies, 3 were from identified from our PubMed search (6, 15, 20), 5
-were originally identified from the 9 studies that did not explicitly
-investigate obesity but included BMI data (14, 18, 19, 24, 25), and two
-datasets were used (21, 22) because these publications did not
-specifically look for any metabolic or obesity conditions but had
-control populations and enabled us to help mitigate against publication
-biases associated with the bacterial microbiome and obesity. The
-number\_string{[}n{]} studies are summarized in Tables 1 and 2.
-
-\textbf{\emph{Sequence Analysis Pipeline.}} All sequence data were
-publicly available and were downloaded from the NCBI Sequence Read
-Archive, the European Nucleotide Archive, or the investigators' personal
-website
-(\url{https://gordonlab.wustl.edu/TurnbaughSE/_10/_09/STM/_2009.html}).
-In total seven studies used 454 (6, 14, 15, 19, 20, 22, 25) and three
-studies used Illumina sequencing (18, 21, 24). All of these studies used
-amplification-based 16S rRNA gene sequencing. Among the studies that
-sequenced the 16S rRNA gene, the researchers targeted the V1-V2 (19),
-V1-V3 (14, 15, 20), V3-V5 (22, 25), V4 {[}(18); (21); {]}, and V3-4 (24)
-regions. For those studies where multiple regions were sequenced, we
-selected the region that corresponded to the largest number of subjects
-(6, 25). We processed the 16S rRNA gene sequence data using a
-standardized mothur pipeline. Briefly, our pipelines attempted to follow
-previously recommended approaches for 454 and Illumina sequencing data
-(26, 27). All sequences were screened for chimeras using UCHIME and
-assigned to operational taxonomic units (OTUs) using the average
-neighbor algorithm using a 3\% distance threshold (28, 29). All sequence
-processing was performed using mothur (v.1.37.0) (30).
-
-\textbf{\emph{Data Analysis.}} We split the overall meta-analysis into
-three general strategies using R (3.3.0). First, we followed the
-approach employed by Finucane et al (9) and Walters et al (8) where each
-study was re-analyzed separately to identify associations between BMI
-and the relative abundance of Bacteroidetes and Firmicutes, the ratio of
-Bacteroidetes and Firmicutes relative abundances (B:F), Shannon
-diversity, observed richness, and Shannon evenness. After each variable
-was transformed to fit a normal distribution a two-tailed t-test was
-performed for comparison of non-obese and obese individuals (i.e.~BMI
-\textgreater{} 35.0). We performed a pooled analysis on these measured
-variables using linear random effect models to correct for study effect
-to asses differences on the combined dataset between non-obese and obese
-groups using the lme4 (v.1.1-12) R package. Next, we compared the
-community structure from non-obese and obese individuals using PERMANOVA
-analysis of Bray-Curtis distance matrices. This analysis was performed
-using the vegan (v.2.3-5) R package. For both analyses, the datasets
-were rarefied (N=1000) so that each study within a study had the same
-number of sequences. Second, for each study we partitioned the subjects
-into a low or high group depending on whether their alpha diversity
-metrics were below or above the median value for the study. The relative
-risk (RR) was then calculated as the ratio of the number of obese
-individuals in the low group to the number of obese individuals in the
-high group. We then performed a Fisher exact-test to investigate whether
-the RR was significantly different from 1.0 within each study and across
-all of the studies using the epiR (0.9-77) and metafor (1.9-8) packages.
-Third, we used the AUCRF (1.1) R package to generate Random Forest
-models. For each study we developed models using either OTUs or
-genus-level phylotypes. The quality of each model was assessed by
-measuring the area under the curve (AUC) of the Receive Operating
-Characteristic (ROC) using ten-fold cross validation. Because the
-genus-level phylotype models were developed using a common reference, it
-was possible to use one study's model (i.e.~the training set) to
-classify the samples from the other studies (i.e.~the testing sets). The
-optimum threshold for the training set was set as the probability
-threshold that had the highest combined sensitivity and specificity.
-This threshold was then used to calculate the accuracy of the model
-applied to the test studies. To generate Receiver Operator
-Characteristic (ROC) curves and calculate the accuracy of the models we
-used the pROC (1.8) R package. Finally, we performed power and sample
-number simulations for different effect sizes for each study using the
-pwr (1.1-3) R package and base R functions. We also calculated the
-actual sample size needed based on the effect size of each individual
-study.
-
-\textbf{\emph{Reproducible methods.}} A detailed and reproducible
-description of how the data were processed and analyzed can be found at
-\url{https://github.com/SchlossLab/Sze_Obesity_mBio_2016/}.
-
-\subsection{Results}\label{results}
+10 studies, 3 were from identified from our PubMed search (10, 15, 16),
+5 were originally identified from the 9 studies that did not explicitly
+investigate obesity but included BMI data (17--21), and two datasets
+were used (22, 23) because these publications did not specifically look
+for any metabolic or obesity conditions but had control populations and
+enabled us to help mitigate against publication biases associated with
+the bacterial microbiome and obesity. The ten studies are summarized in
+Table 1.
 
@@
 studies. The B:F and the relative abundance of \emph{Firmicutes} were
-not significantly associated with obesity in any study. We identified 7
-P-values less than 0.05: three studies indicated obese individuals had a
-lower richness, two studies indicated a significantly lower diversity,
-one study indicated a significantly lower evenness, and one study
-indicated a significantly higher relative abundance of
+not significantly associated with obesity in any study. We identified
+seven P-values that were less than 0.05: three studies indicated obese
+individuals had a lower richness, two studies indicated a significantly
+lower diversity, one study indicated a significantly lower evenness, and
+one study indicated a significantly higher relative abundance of
 \emph{Bacteroidetes} (Figures 2 and S1). These results largely match
 those of the Walters and Finucane re-analysis studies. Interestingly,
-although only two of the 10 studies observed the previously reported
+although only two of the ten studies observed the previously reported
 association between lower diversity and obesity, the other studies
@@
 among obese individuals (all P\textless{}0.011). Although there was a
-significant relationship between these metrics and diversity and obesity
-status, the effect size was quite small. The obese individuals averaged
-7.47\% lower richness, 0.88\% lower evenness, and 2.07\% lower
-diversity. There were no significant associations when we pooled the
-phylum-level metrics across studies. These results indicate that obese
-individuals do have a statistically significant lower diversity than
-non-obese individuals; however, it is questionable whether the
-difference is biologically significant.
+significant relationship between these metrics and obesity status, the
+effect size was quite small. The obese individuals averaged 7.47\% lower
+richness, 0.88\% lower evenness, and 2.07\% lower diversity. There were
+no significant associations when we pooled the phylum-level metrics
+across studies. These results indicate that obese individuals do have a
+statistically significant lower diversity than non-obese individuals;
+however, it is questionable whether the difference is biologically
+significant.
 
 \textbf{\emph{Relative risk.}} Building upon the alpha diversity
-analysis we calculated the relative risk of being obese based on whether
-an individual's alpha diversity metrics were below or above the median
-metric for that study. The results using relative risk largely matched
-those of using the untransformed alpha diversity data. Across the
-number\_string{[}n{]} studies and six metrics, the only significant
-relative risk values were the richness, evenness, and diversity values
-from the Goodrich study (Figures 3 and S2). Again, although the relative
-risk values were not significant for other studies, the values tended to
-be above one. When we pooled the data using a random effects model, the
-relative risk associated with having a richness, evenness, or diversity
-below the median for the population was significantly associated with
-obesity (all P\textless{}0.0044). The relative risks associated with
-alpha diversity were small. The relative risk of having a low richness
-was 1.30 (1.13-1.49), low evenness was 1.20 (1.06-1.37), and low
-diversity was 1.27 (1.09-1.48). There were no significant difference in
-the phylum-level metrics. Again, the relative risk results indicate that
+analysis we calculated the relative risk of being obese based on an
+individual's alpha diversity metrics relative to the median metric for
+that study. The results using relative risk largely matched those of
+using the untransformed alpha diversity data. Across the ten studies and
+six metrics, the only significant relative risk values were the
+richness, evenness, and diversity values from the Goodrich study
+(Figures 3 and S2). Again, although the relative risk values were not
+significant for other studies, the values tended to be above one. When
+we pooled the data using a random effects model, the relative risk
+associated with having a richness, evenness, or diversity below the
+median for the population was significantly associated with obesity (all
+P\textless{}0.0044). The relative risks associated with alpha diversity
+were small. The relative risk of having a low richness was 1.30
+(1.13-1.49), low evenness was 1.20 (1.06-1.37), and low diversity was
+1.27 (1.09-1.48). There were no significant difference in the
+phylum-level metrics. Again, the relative risk results indicate that
 individuals with a lower richness, evenness, or diversity are at
@@
 the membership and structure of the individuals from each study. We then
-used PERMANOVA to test for a significant differences between the
-structure of non-obese and obese individuals. The Escobar, Goodrich, and
+used AMOVA to test for a significant differences between the structure
+of non-obese and obese individuals (Table 1). The Escobar, Goodrich, and
 Turnbaugh datasets indicated a significant difference in community
@@
 The Walters re-analysis study suggested that it was possible to classify
-individuals as being obese or non-obese based on the composition of
+individuals as being non-obese or obese based on the composition of
 their microbiota. We repeated this analysis with additional datasets
@@
 (Figure 4A). So that we could test models on other datasets, we trained
-models using genus-level phylotype data for each dataset. The the
-observed AUC values for the models applied to the training datasets
-varied between 0.51 and 0.65, again indicating a relatively poor ability
-to classify individuals from the original dataset (Figure 4B). For each
-model we identified the probability where the sum of the sensitivity and
-specificity was the highest. We then used this to identify a threshold
-to calculate the accuracy of the models when applied to the other
-number\_string{[}n-1{]} datasets (Figure 5). Although there considerable
-variation in accuracy values for each model, the median accuracy for
-each model varied between 0.33 (Turnbaugh) and 0.65 (HMP) (median=0.57).
-When we considered the number of samples, balance of non-obese and obese
-individuals, and region within the 16S rRNA gene it was not possible to
-identify factors that predictably affected model performance. The
-ability to predict obesity status using the relative abundance of OTUs
-and genera in the communities is only marginally better than random.
-These results suggest that given the large diversity of microbiome
-compositions it is difficult to identify a taxonomic signal that can be
-associated with obesity.
+models using genus-level phylotype data for each dataset. The
+cross-validated AUC values for the models applied to the training
+datasets varied between 0.51 and 0.65, again indicating a relatively
+poor ability to classify individuals from the original dataset (Figure
+4B). For each model we identified the probability where the sum of the
+sensitivity and specificity was the highest. We then used this
+probability to define a threshold for calculating the accuracy of the
+models when applied to the other nine datasets (Figure 5). Although
+there considerable variation in accuracy values for each model, the
+median accuracy for each model varied between 0.33 (Turnbaugh) and 0.65
+(HMP) (median=0.57). When we considered the number of samples, balance
+of non-obese and obese individuals, and region within the 16S rRNA gene
+it was not possible to identify factors that predictably affected model
+performance. The ability to predict obesity status using the relative
+abundance of OTUs and genera in the communities is only marginally
+better than random. These results suggest that given the large diversity
+of microbiome compositions it is difficult to identify a taxonomic
+signal that can be associated with obesity.
 
@@
 assuming the observed interpersonal variation from each study and
-balanced sampling between the two groups. To detect a 1, 5, 10, and 15\%
-difference in Shannon index, the median sampling effort per group was
-approximately 3,400, 140, 35, and 16 individuals, respectively. To
-detect a 1, 5, 10, and 15\% difference in B:F values, the median
-sampling effort per group was approximately 160,000, 6,300, 1,600, and
-700 individuals, respectively. To detect a 1, 5, 10, and 15\% difference
-in relative risk values using Shannon diversity, the median sampling
-effort per group was approximately 39,000, 1,500, 380, and 170
-individuals, respectively. These estimates indicate that most microbiome
-studies are underpowered to detect modest effect sizes using either
-metric. In the case of obesity, the studies were underpowered to detect
-the \texttt{signif(range\_effect\_size,\ 1){[}1{]}} to 6\% difference in
-diversity that was observed across the studies.
+balanced sampling between the two groups. To detect a 1, 5, 10, or 15\%
+difference in Shannon index, the median required sampling effort per
+group was approximately 3,400, 140, 35, or 16 individuals, respectively.
+To detect a 1, 5, 10, and 15\% difference in B:F values, the median
+required sampling effort per group was approximately 160,000, 6,300,
+1,600, or 700 individuals, respectively. To detect a 1, 5, 10, and 15\%
+difference in relative risk values using Shannon diversity, the median
+required sampling effort per group was approximately 39,000, 1,500, 380,
+or 170 individuals, respectively. These estimates indicate that most
+microbiome studies are underpowered to detect modest effect sizes using
+either metric. In the case of obesity, the studies were underpowered to
+detect the \texttt{signif(range\_effect\_size,\ 1){[}1{]}} to 6\%
+difference in diversity that was observed across the studies.
 
@@
 associated with obesity. We performed an extensive literature review of
-the existing studies on the microbiome and obesity and perform a
+the existing studies on the microbiome and obesity and performed a
 meta-analysis on the studies that remained based on our inclusion and
@@
 reliably classify individuals as being obese based on the composition of
-their microbiome is limited. Finally, we assessed the ability of each
+their microbiome was limited. Finally, we assessed the ability of each
 study to detect defined differences in alpha diversity and observed that
@@
 rate of a significant finding in an underpowered study is small leading
-to results that cannot be reproduced (31). Future microbiome studies
+to results that cannot be reproduced (24). Future microbiome studies
 should articulate the basis for their experimental design.
@@
 studies whereas the Finucane study used 16S rRNA gene sequence data from
-3 studies (7, 10, 25) and a metagenomic study (32) and the Walters study
-used 16S rRNA gene sequence data from 5 studies (10, 15, 19, 25, 33);
-two studies were included in both analyses (10, 25). Our analysis
-included 4 of these studies (10, 15, 19, 25) and excluded 3 of the
-studies because they were too small (7), only utilized metagenomic data
-(32), or used short single read Illumina HiSeq data that has a high
-error rate making it untractable for \emph{de novo} OTU clustering (33).
-The additional seven datasets were published after the two reviews were
-performed and include datasets with more samples than were found in the
-original studies. Our collection of ten studies allowed us to largely
-use the same sequence analysis pipeline for all datasets and relied
-heavily on the availability of public data and access to metadata that
-included variables beyond the needs of the original study. To execute
-this analysis, we created an automated data analysis pipeline, which can
-be easily updated to add additional studies as they become available.
-Similarly, it would be possible to adapt this pipeline to other body
-sites and treatment or variables (e.g.~subject's sex or age).
+three studies (7, 10, 21) and a metagenomic study (25) and the Walters
+study used 16S rRNA gene sequence data from five studies (10, 15, 20,
+21, 26); two studies were included in both analyses (10, 21). Our
+analysis included four of these studies (10, 15, 20, 21) and excluded
+three of the studies because they were too small (7), only utilized
+metagenomic data (25), or used short single read Illumina HiSeq data
+that has a high error rate making it untractable for \emph{de novo} OTU
+clustering (26). The additional seven datasets were published after the
+two reviews were performed and include datasets with more samples than
+were found in the original studies. Our collection of ten studies
+allowed us to largely use the same sequence analysis pipeline for all
+datasets and relied heavily on the availability of public data and
+access to metadata that included variables beyond the needs of the
+original study. To execute this analysis, we created an automated data
+analysis pipeline, which can be easily updated to add additional studies
+as they become available. Similarly, it would be possible to adapt this
+pipeline to other body sites and treatment or variables (e.g.~subject's
+sex or age).
 
-Similar to our study, the Walters analysis (8), the authors generated
-Random Forest machine learning models to differentiate between non-obese
-and obese individuals. They obtained similar AUC values to our analysis;
-however, they did not attempt to test these models on the other studies
-in their analysis. When we performed the inter-dataset cross validation
-the median accuracy across datasets was only 56.67\% indicating that the
+Similar to our study, the Walters et al (8) generated Random Forest
+machine learning models to differentiate between non-obese and obese
+individuals. They obtained similar AUC values to our analysis; however,
+they did not attempt to test these models on the other studies in their
+analysis. When we performed the inter-dataset cross validation the
+median accuracy across datasets was only 56.67\% indicating that the
 models did a poor job when applied to other datasets. This could be due
@@
 
-Although we failed to find an effect it is not realistic to necessarily
-state that there is no microbiome impact on obesity. There is strong
-evidence in murine models of obesity that the microbiome and level of
-adiposity can be manipulated via genetic manipulation of the animal and
+Although we failed to find an effect this does not necessarily mean that
+there is no role for the microbiome in obesity. There is strong evidence
+in murine models of obesity that the microbiome and level of adiposity
+can be manipulated via genetic manipulation of the animal and
 manipulation of the community through antibiotics or colonizing germ
@@
 microbiome.
+
+\subsection{Methods}\label{methods}
+
+\textbf{\emph{Sequence Analysis Pipeline.}} All sequence data were
+publicly available and were downloaded from the NCBI Sequence Read
+Archive, the European Nucleotide Archive, or the investigators' personal
+website
+(\url{https://gordonlab.wustl.edu/TurnbaughSE/_10/_09/STM/_2009.html}).
+In total seven studies used 454 (6, 15, 16, 18, 20--22) and three
+studies used Illumina sequencing (17, 19, 23). All of these studies used
+amplification-based 16S rRNA gene sequencing. Among the studies that
+sequenced the 16S rRNA gene, the researchers targeted the V1-V2 (20),
+V1-V3 (15, 16, 18), V3-V5 (21, 22), V4 {[}(19); (23); {]}, and V3-4 (17)
+regions. For those studies where multiple regions were sequenced, we
+selected the region that corresponded to the largest number of subjects
+(6, 21). We processed the 16S rRNA gene sequence data using a
+standardized mothur pipeline. Briefly, our pipelines attempted to follow
+previously recommended approaches for 454 and Illumina sequencing data
+(27, 28). All sequences were screened for chimeras using UCHIME and
+assigned to operational taxonomic units (OTUs) using the average
+neighbor algorithm using a 3\% distance threshold (29, 30). All sequence
+processing was performed using mothur (v.1.37.0) (31).
+
+\textbf{\emph{Data Analysis.}} We split the overall meta-analysis into
+three general strategies using R (3.3.0). First, we followed the
+approach employed by Finucane et al (9) and Walters et al (8) where each
+study was re-analyzed separately to identify associations between BMI
+and the relative abundance of \emph{Bacteroidetes} and Firmicutes, the
+ratio of \emph{Bacteroidetes} and \emph{Firmicutes} relative abundances
+(B:F), Shannon diversity, observed richness, and Shannon evenness. After
+each variable was transformed to fit a normal distribution a two-tailed
+t-test was performed for comparison of non-obese and obese individuals
+(i.e.~BMI \textgreater{} 35.0). We performed a pooled analysis on these
+measured variables using linear random effect models to correct for
+study effect to asses differences on the combined dataset between
+non-obese and obese groups using the lme4 (v.1.1-12) R package. Next, we
+compared the community structure from non-obese and obese individuals
+using analysis of molecular variance (AMOVA) with Bray-Curtis distance
+matrices (32). This analysis was performed using the vegan (v.2.3-5) R
+package. For both analyses, the datasets were rarefied (N=1000) so that
+each study within a study had the same number of sequences. Second, for
+each study we partitioned the subjects into a low or high group
+depending on whether their alpha diversity metrics were below or above
+the median value for the study. The relative risk (RR) was then
+calculated as the ratio of the number of obese individuals in the low
+group to the number of obese individuals in the high group. We then
+performed a Fisher exact-test to investigate whether the RR was
+significantly different from 1.0 within each study and across all of the
+studies using the epiR (0.9-77) and metafor (1.9-8) packages. Third, we
+used the AUCRF (1.1) R package to generate Random Forest models. For
+each study we developed models using either OTUs or genus-level
+phylotypes. The quality of each model was assessed by measuring the area
+under the curve (AUC) of the Receive Operating Characteristic (ROC)
+using ten-fold cross validation. Because the genus-level phylotype
+models were developed using a common reference, it was possible to use
+one study's model (i.e.~the training set) to classify the samples from
+the other studies (i.e.~the testing sets). The optimum threshold for the
+training set was set as the probability threshold that had the highest
+combined sensitivity and specificity. This threshold was then used to
+calculate the accuracy of the model applied to the test studies. To
+generate Receiver Operator Characteristic (ROC) curves and calculate the
+accuracy of the models we used the pROC (1.8) R package. Finally, we
+performed power and sample number simulations for different effect sizes
+for each study using the pwr (1.1-3) R package and base R functions. We
+also calculated the actual sample size needed based on the effect size
+of each individual study.
+
+\textbf{\emph{Reproducible methods.}} A detailed and reproducible
+description of how the data were processed and analyzed can be found at
+\url{https://github.com/SchlossLab/Sze_Obesity_mBio_2016/}.
 
@@
 
-\textbf{Table 1. Summary Demographics of Individuals used in the
-Meta-analysis.}
-
-\begin{verbatim}
-## 
-## > capwords <- function(s, strict = FALSE) {
-## +     s <- as.character(s)
-## +     cap <- function(s) paste(toupper(substring(s, 1, 1)), {
-## +         s <- su .... [TRUNCATED] 
-## 
-## > make_study_label <- function(dataset) {
-## +     study <- ifelse(dataset == "hmp", "HMP", capwords(dataset))
-## + }
-## 
-## > format_p <- function(p_value) {
-## +     char_p <- format(round(p_value, 3), nsmall = 3)
-## +     if (p_value < 0.001) {
-## +         char_p <- "<0.001"
-## +    .... [TRUNCATED] 
-## 
-## > get_study_summary <- function(study, beta = beta_summary) {
-## +     metadata_file <- paste0("data/", study, "/", study, ".metadata")
-## +     metadata <- .... [TRUNCATED] 
-## 
-## > beta_summary <- read.table("data/process/beta_tests.summary", 
-## +     header = T, row.names = 1)
-## 
-## > datasets <- sort(rownames(beta_summary))
-## 
-## > study_summary <- t(sapply(datasets, get_study_summary))
-## 
-## > kable(study_summary, row.names = FALSE, col.names = c("Study", 
-## +     "Subjects (N)", "Obese (%)", "Average BMI (Min-Max)", "Female (%)", 
-## +     "Av ..." ... [TRUNCATED] 
-## 
-## 
-## Study        Subjects (N)    Obese (%)    Average BMI (Min-Max)    Female (%)    Average Age (Min-Max)    Non-Hispanic White (%)    AMOVA (P-value) 
-## ----------  --------------  -----------  -----------------------  ------------  -----------------------  ------------------------  -----------------
-## Baxter           172           27.3         27.0 (17.5-46.9)          64.5         54.3 (29.0-80.0)                87.8                  0.078      
-## Escobar           30           33.3         27.4 (19.5-37.6)          46.7         38.1 (21.0-60.0)                 NA                   0.047      
-## Goodrich         982           19.7         26.3 (16.2-52.4)          98.9         61.0 (23.0-86.0)                 NA                  <0.001      
-## Hmp              287           10.8         24.3 (19.0-34.0)          49.1         26.3 (18.0-40.0)                81.5                  0.322      
-## Ross              63           60.3         31.6 (22.1-47.9)          76.2         57.0 (33.0-81.0)                0.0                   0.845      
-## Schubert         104           32.7         28.2 (18.5-62.5)          66.3         52.8 (19.0-88.0)                82.7                  0.180      
-## Turnbaugh        146           67.8                NA                  NA                 NA                       51.4                  0.040      
-## Wu                64            7.8         24.3 (14.0-41.3)          53.1         26.3 (2.16-50.0)                 NA                   0.577      
-## Zeevi            731            NA          26.4 (16.4-47.0)           NA          43.4 (18.0-70.0)                 NA                   0.135      
-## Zupancic         207           36.2         28.2 (18.2-127.0)         57.0         46.7 (20.0-79.0)               100.0                  0.206
-\end{verbatim}
-
-\newpage
-
-\textbf{Figure 1: PRISMA flow diagram of total records searched (34).}
+\textbf{Figure 1: PRISMA flow diagram of total records searched (33).}
 
@@
 Non-Obese groups Based on Evenness (A), Richness (B), or the Relative
-Abundance of Bacteroidetes (C) and Firmictues (D).}
+Abundance of \emph{Bacteroidetes} (C) and Firmictues (D).}
 
@@
 on Evenness (A), Richness (B), or the Relative Abundance of
-Bacteroidetes (C) and Firmictues (D).}
+\emph{Bacteroidetes} (C) and Firmictues (D).}
 
@@
 \textbf{Figure S6: Power (A) and sample size simulations (B) for the
-relative abundance of Bacteroidetes for differentiating between
+relative abundance of \emph{Bacteroidetes} for differentiating between
 non-obese versus obese for effect sizes of 1, 5, 10, and 15\%.} Power
@@
 \textbf{Figure S7: Power (A) and sample size simulations (B) for the
-relative abundance of Firmicutes for differentiating between non-obese
-versus obese for effect sizes of 1, 5, 10, and 15\%.} Power calculations
-use the sampling distribution from the original studies and the sample
-size estimations assume an equal amount of sampling from each treatment
-group.
+relative abundance of \emph{Firmicutes} for differentiating between
+non-obese versus obese for effect sizes of 1, 5, 10, and 15\%.} Power
+calculations use the sampling distribution from the original studies and
+the sample size estimations assume an equal amount of sampling from each
+treatment group.
 
@@
 
-\hyperdef{}{ref-rossux5f16sux5f2015}{\label{ref-rossux5f16sux5f2015}}
-14. \textbf{Ross MC}, \textbf{Muzny DM}, \textbf{McCormick JB},
-\textbf{Gibbs RA}, \textbf{Fisher-Hoch SP}, \textbf{Petrosino JF}. 2015.
-16S gut community of the Cameron County Hispanic Cohort. Microbiome
-\textbf{3}:7. doi:\url{http://doi.org/10.1186/s40168-015-0072-y}.
+\hyperdef{}{ref-moherux5fpreferredux5f2010}{\label{ref-moherux5fpreferredux5f2010}}
+14. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
+\textbf{Altman DG}, \textbf{PRISMA Group}. 2010. Preferred reporting
+items for systematic reviews and meta-analyses: The PRISMA statement.
+International Journal of Surgery (London, England) \textbf{8}:336--341.
+doi:\url{http://doi.org/10.1016/j.ijsu.2010.02.007}.
 
@@
 
-\hyperdef{}{ref-namux5fcomparativeux5f2011}{\label{ref-namux5fcomparativeux5f2011}}
-16. \textbf{Nam Y-D}, \textbf{Jung M-J}, \textbf{Roh SW}, \textbf{Kim
-M-S}, \textbf{Bae J-W}. 2011. Comparative analysis of Korean human gut
-microbiota by barcoded pyrosequencing. PloS One \textbf{6}:e22109.
-doi:\url{http://doi.org/10.1371/journal.pone.0022109}.
+\hyperdef{}{ref-escobarux5fgutux5f2014}{\label{ref-escobarux5fgutux5f2014}}
+16. \textbf{Escobar JS}, \textbf{Klotz B}, \textbf{Valdes BE},
+\textbf{Agudelo GM}. 2014. The gut microbiota of Colombians differs from
+that of Americans, Europeans and Asians. BMC microbiology
+\textbf{14}:311. doi:\url{http://doi.org/10.1186/s12866-014-0311-6}.
 
-\hyperdef{}{ref-arumugamux5fenterotypesux5f2011}{\label{ref-arumugamux5fenterotypesux5f2011}}
-17. \textbf{Arumugam M}, \textbf{Raes J}, \textbf{Pelletier E},
-\textbf{Le Paslier D}, \textbf{Yamada T}, \textbf{Mende DR},
-\textbf{Fernandes GR}, \textbf{Tap J}, \textbf{Bruls T}, \textbf{Batto
-J-M}, \textbf{Bertalan M}, \textbf{Borruel N}, \textbf{Casellas F},
-\textbf{Fernandez L}, \textbf{Gautier L}, \textbf{Hansen T},
-\textbf{Hattori M}, \textbf{Hayashi T}, \textbf{Kleerebezem M},
-\textbf{Kurokawa K}, \textbf{Leclerc M}, \textbf{Levenez F},
-\textbf{Manichanh C}, \textbf{Nielsen HB}, \textbf{Nielsen T},
-\textbf{Pons N}, \textbf{Poulain J}, \textbf{Qin J},
-\textbf{Sicheritz-Ponten T}, \textbf{Tims S}, \textbf{Torrents D},
-\textbf{Ugarte E}, \textbf{Zoetendal EG}, \textbf{Wang J},
-\textbf{Guarner F}, \textbf{Pedersen O}, \textbf{Vos WM de},
-\textbf{Brunak S}, \textbf{Doré J}, \textbf{MetaHIT Consortium},
-\textbf{Antolín M}, \textbf{Artiguenave F}, \textbf{Blottiere HM},
-\textbf{Almeida M}, \textbf{Brechot C}, \textbf{Cara C},
-\textbf{Chervaux C}, \textbf{Cultrone A}, \textbf{Delorme C},
-\textbf{Denariaz G}, \textbf{Dervyn R}, \textbf{Foerstner KU},
-\textbf{Friss C}, \textbf{Guchte M van de}, \textbf{Guedon E},
-\textbf{Haimet F}, \textbf{Huber W}, \textbf{Hylckama-Vlieg J van},
-\textbf{Jamet A}, \textbf{Juste C}, \textbf{Kaci G}, \textbf{Knol J},
-\textbf{Lakhdari O}, \textbf{Layec S}, \textbf{Le Roux K},
-\textbf{Maguin E}, \textbf{Mérieux A}, \textbf{Melo Minardi R},
-\textbf{M'rini C}, \textbf{Muller J}, \textbf{Oozeer R},
-\textbf{Parkhill J}, \textbf{Renault P}, \textbf{Rescigno M},
-\textbf{Sanchez N}, \textbf{Sunagawa S}, \textbf{Torrejon A},
-\textbf{Turner K}, \textbf{Vandemeulebrouck G}, \textbf{Varela E},
-\textbf{Winogradsky Y}, \textbf{Zeller G}, \textbf{Weissenbach J},
-\textbf{Ehrlich SD}, \textbf{Bork P}. 2011. Enterotypes of the human gut
-microbiome. Nature \textbf{473}:174--180.
-doi:\url{http://doi.org/10.1038/nature09944}.
+\hyperdef{}{ref-zeeviux5fpersonalizedux5f2015}{\label{ref-zeeviux5fpersonalizedux5f2015}}
+17. \textbf{Zeevi D}, \textbf{Korem T}, \textbf{Zmora N},
+\textbf{Israeli D}, \textbf{Rothschild D}, \textbf{Weinberger A},
+\textbf{Ben-Yacov O}, \textbf{Lador D}, \textbf{Avnit-Sagi T},
+\textbf{Lotan-Pompan M}, \textbf{Suez J}, \textbf{Mahdi JA},
+\textbf{Matot E}, \textbf{Malka G}, \textbf{Kosower N}, \textbf{Rein M},
+\textbf{Zilberman-Schapira G}, \textbf{Dohnalová L},
+\textbf{Pevsner-Fischer M}, \textbf{Bikovsky R}, \textbf{Halpern Z},
+\textbf{Elinav E}, \textbf{Segal E}. 2015. Personalized Nutrition by
+Prediction of Glycemic Responses. Cell \textbf{163}:1079--1094.
+doi:\url{http://doi.org/10.1016/j.cell.2015.11.001}.
+
+\hyperdef{}{ref-rossux5f16sux5f2015}{\label{ref-rossux5f16sux5f2015}}
+18. \textbf{Ross MC}, \textbf{Muzny DM}, \textbf{McCormick JB},
+\textbf{Gibbs RA}, \textbf{Fisher-Hoch SP}, \textbf{Petrosino JF}. 2015.
+16S gut community of the Cameron County Hispanic Cohort. Microbiome
+\textbf{3}:7. doi:\url{http://doi.org/10.1186/s40168-015-0072-y}.
 
 \hyperdef{}{ref-goodrichux5fhumanux5f2014}{\label{ref-goodrichux5fhumanux5f2014}}
-18. \textbf{Goodrich JK}, \textbf{Waters JL}, \textbf{Poole AC},
+19. \textbf{Goodrich JK}, \textbf{Waters JL}, \textbf{Poole AC},
 \textbf{Sutter JL}, \textbf{Koren O}, \textbf{Blekhman R},
@@
 \hyperdef{}{ref-wuux5flinkingux5f2011}{\label{ref-wuux5flinkingux5f2011}}
-19. \textbf{Wu GD}, \textbf{Chen J}, \textbf{Hoffmann C},
+20. \textbf{Wu GD}, \textbf{Chen J}, \textbf{Hoffmann C},
 \textbf{Bittinger K}, \textbf{Chen Y-Y}, \textbf{Keilbaugh SA},
@@
 
-\hyperdef{}{ref-escobarux5fgutux5f2014}{\label{ref-escobarux5fgutux5f2014}}
-20. \textbf{Escobar JS}, \textbf{Klotz B}, \textbf{Valdes BE},
-\textbf{Agudelo GM}. 2014. The gut microbiota of Colombians differs from
-that of Americans, Europeans and Asians. BMC microbiology
-\textbf{14}:311. doi:\url{http://doi.org/10.1186/s12866-014-0311-6}.
-
-\hyperdef{}{ref-baxterux5fmicrobiota-basedux5f2016}{\label{ref-baxterux5fmicrobiota-basedux5f2016}}
-21. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}:37.
-doi:\url{http://doi.org/10.1186/s13073-016-0290-3}.
+\hyperdef{}{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{\label{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}}
+21. \textbf{Human Microbiome Project Consortium}. 2012. Structure,
+function and diversity of the healthy human microbiome. Nature
+\textbf{486}:207--214. doi:\url{http://doi.org/10.1038/nature11234}.
 
@@
 
-\hyperdef{}{ref-moherux5fpreferredux5f2010}{\label{ref-moherux5fpreferredux5f2010}}
-23. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
-\textbf{Altman DG}, \textbf{PRISMA Group}. 2010. Preferred reporting
-items for systematic reviews and meta-analyses: The PRISMA statement.
-International Journal of Surgery (London, England) \textbf{8}:336--341.
-doi:\url{http://doi.org/10.1016/j.ijsu.2010.02.007}.
-
-\hyperdef{}{ref-zeeviux5fpersonalizedux5f2015}{\label{ref-zeeviux5fpersonalizedux5f2015}}
-24. \textbf{Zeevi D}, \textbf{Korem T}, \textbf{Zmora N},
-\textbf{Israeli D}, \textbf{Rothschild D}, \textbf{Weinberger A},
-\textbf{Ben-Yacov O}, \textbf{Lador D}, \textbf{Avnit-Sagi T},
-\textbf{Lotan-Pompan M}, \textbf{Suez J}, \textbf{Mahdi JA},
-\textbf{Matot E}, \textbf{Malka G}, \textbf{Kosower N}, \textbf{Rein M},
-\textbf{Zilberman-Schapira G}, \textbf{Dohnalová L},
-\textbf{Pevsner-Fischer M}, \textbf{Bikovsky R}, \textbf{Halpern Z},
-\textbf{Elinav E}, \textbf{Segal E}. 2015. Personalized Nutrition by
-Prediction of Glycemic Responses. Cell \textbf{163}:1079--1094.
-doi:\url{http://doi.org/10.1016/j.cell.2015.11.001}.
-
-\hyperdef{}{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{\label{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}}
-25. \textbf{Human Microbiome Project Consortium}. 2012. Structure,
-function and diversity of the healthy human microbiome. Nature
-\textbf{486}:207--214. doi:\url{http://doi.org/10.1038/nature11234}.
-
-\hyperdef{}{ref-Kozichux5f2013}{\label{ref-Kozichux5f2013}}
-26. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
-\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
-dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq Illumina sequencing platform.
-Applied and environmental microbiology \textbf{79}:5112--5120.
-
-\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
-27. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
-Reducing the effects of PCR amplification and sequencing artifacts on
-16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
-doi:\url{http://doi.org/10.1371/journal.pone.0027310}.
-
-\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
-28. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
-\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
-and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
-doi:\url{http://doi.org/10.1093/bioinformatics/btr381}.
-
-\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
-29. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
-methods outperform reference-based methods for assigning 16S rRNA gene
-sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\url{http://doi.org/10.7717/peerj.1487}.
-
-\hyperdef{}{ref-mothur}{\label{ref-mothur}}
-30. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{others}. 2009. Introducing mothur:
-open-source, platform-independent, community-supported software for
-describing and comparing microbial communities. Applied and
-environmental microbiology \textbf{75}:7537--7541.
+\hyperdef{}{ref-baxterux5fmicrobiota-basedux5f2016}{\label{ref-baxterux5fmicrobiota-basedux5f2016}}
+23. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\url{http://doi.org/10.1186/s13073-016-0290-3}.
 
 \hyperdef{}{ref-Ioannidis2005}{\label{ref-Ioannidis2005}}
-31. \textbf{Ioannidis JPA}. 2005. Why most published research findings
+24. \textbf{Ioannidis JPA}. 2005. Why most published research findings
 are false. PLoS Med \textbf{2}:e124.
@@
 \hyperdef{}{ref-Qin2010}{\label{ref-Qin2010}}
-32. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
+25. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
 \textbf{Burgdorf KS}, \textbf{Manichanh C}, \textbf{Nielsen T},
@@
 \hyperdef{}{ref-yatsunenkoux5fhumanux5f2012}{\label{ref-yatsunenkoux5fhumanux5f2012}}
-33. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
+26. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
 \textbf{Trehan I}, \textbf{Dominguez-Bello MG}, \textbf{Contreras M},
@@
 
+\hyperdef{}{ref-Kozichux5f2013}{\label{ref-Kozichux5f2013}}
+27. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq Illumina sequencing platform.
+Applied and environmental microbiology \textbf{79}:5112--5120.
+
+\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
+28. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
+Reducing the effects of PCR amplification and sequencing artifacts on
+16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
+doi:\url{http://doi.org/10.1371/journal.pone.0027310}.
+
+\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
+29. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
+\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
+and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
+doi:\url{http://doi.org/10.1093/bioinformatics/btr381}.
+
+\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
+30. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
+methods outperform reference-based methods for assigning 16S rRNA gene
+sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
+doi:\url{http://doi.org/10.7717/peerj.1487}.
+
+\hyperdef{}{ref-mothur}{\label{ref-mothur}}
+31. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{others}. 2009. Introducing mothur:
+open-source, platform-independent, community-supported software for
+describing and comparing microbial communities. Applied and
+environmental microbiology \textbf{75}:7537--7541.
+
+\hyperdef{}{ref-Anderson2001}{\label{ref-Anderson2001}}
+32. \textbf{Anderson MJ}. 2001. A new method for non-parametric
+multivariate analysis of variance. Austral Ecology \textbf{26}:32--46.
+doi:\url{http://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}.
+
 \hyperdef{}{ref-Moher2009}{\label{ref-Moher2009}}
-34. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
+33. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
 \textbf{Altman DG}. 2009. Preferred reporting items for systematic

### 2016-06-06 submission/Sze_Obesity_mBio_2016.Rmd
Incorporate Marc's edits

@@
 
-Two recent studies have re-analyzed published data and found that when datasets are analyzed independently there was limited support the widely accepted hypothesis that changes in the microbiome are associated with obesity.
-This hypothesis was reconsidered by increasing the number of data sets and pooling the results across the individual datasets.
+Two recent studies have re-analyzed published data and found that when datasets are analyzed independently there was limited support for the widely accepted hypothesis that changes in the microbiome are associated with obesity.
+This hypothesis was reconsidered by increasing the number of datasets and pooling the results across the individual datasets.
 The Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines were applied to identify 10 studies for an updated and more synthetic analysis.
@@
 This number increases to approximately 35% in adults (aged 20 or older) and these statistics have seen little change since 2003 [@ogden_prevalence_2014].
-Traditionally the body mass index (BMI) has been used as the traditional method of classifying individuals as non-obese or obese [@lichtash_body_2013].
+Traditionally, the body mass index (BMI) has been used to classify individuals as non-obese or obese
+ [@lichtash_body_2013].
 Recently, there has been increased interest in the role of the microbiome in modulating obesity [@brahe_can_2016; @dror_microbiota_2016].
@@
 Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since these manipulation yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016].
+Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities
+were manipulated with antibiotics or underwent colonization with varied  communities appears to support the association since these manipulation yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016].
 The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.
@@
 The initial search yielded 187 studies.
-We identified number_string[n] additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
+We identified `r number_string[n]` additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
 We then manually curated the 196 studies to select those studies that included BMI and sequence data.
@@
 An additional study was removed from our analysis because no individuals in the study had a BMI over 30.
-Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_obesity-associated_2006; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity.
+Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_obesity-associated_2006; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 10 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity.
 The `r number_string[n]` studies are summarized in Table 1.
@@
 ***Alpha diversity analysis.***
-We calculated the Shannon diversity index, richness, and Shannon evenness,  the relative abundance of *Bacteroidetes* and *Firmicutes*, and the ratio of their relative abundance (B:F) for each sample.
+We calculated the Shannon diversity index, observed richness, and Shannon evenness, the relative abundance of *Bacteroidetes* and *Firmicutes*, and the ratio of their relative abundance (B:F) for each sample.
 Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the `r number_string[n]` studies.
@@
 Following the approach used by the Walters and Finucane re-analysis studies, for each dataset we calculated a Bray-Curtis distance matrix to measure the difference in the membership and structure of the individuals from each study.
-We then used AMOVA to test for a significant differences between the structure of non-obese and obese individuals (Table 1).
+We then used AMOVA to test for significant differences between the structure of non-obese and obese individuals (Table 1).
 The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05).
@@
 We then used this probability to define a threshold for calculating the accuracy of the models when applied to the other `r number_string[n-1]` datasets (Figure 5).
-Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`).
+Although there was considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`).
 When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance.
@@
 These estimates indicate that most microbiome studies are underpowered to detect modest effect sizes using either metric.
-In the case of obesity, the studies were underpowered to detect the `signif(range_effect_size, 1)[1]` to `r signif(range_effect_size, 1)[2]`% difference in diversity that was observed across the studies.
+In the case of obesity, the studies were underpowered to detect the `r signif(range_effect_size, 1)[1]` to `r signif(range_effect_size, 1)[2]`% difference in diversity that was observed across the studies.
 
@@
 
-Our meta-analysis helps to provide a clarity to the ongoing debate of whether or not there are specific microbiome-based markers that can be associated with obesity.
+Our meta-analysis helps to provide clarity to the ongoing debate of whether or not there are specific microbiome-based markers that can be associated with obesity.
 We performed an extensive literature review of the existing studies on the microbiome and obesity and performed a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.
@@
 Finally, we assessed the ability of each study to detect defined differences in alpha diversity and observed that most studies were underpowered to detect modest effect sizes.
-Considering these datasets are among the largest published, it appears that most microbiome studies are underpowered to detect differences in alpha diversity.
+Considering these datasets are among the largest published, it appears that most human microbiome studies are underpowered to detect differences in alpha diversity.
 
@@
 This was also true of those studies that originally sought to identify associations with obesity.
-Even with a balanced design, we showed that it was necessary to obtain approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)` and `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)` sequences per sample to detect a 5% difference in Shannon diversity or B:F, respectively.
-It was interesting that these sample sizes agreed across studies regardless of their sequencing method, region with in the 16S rRNA gene, or subject population (Figure 6).
+Even with a balanced design, we showed that it was necessary to obtain approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)` and `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)` samples per group to detect a 5% difference in Shannon diversity or B:F, respectively.
+It was interesting that these sample sizes agreed across studies regardless of their sequencing method, region within the 16S rRNA gene, or subject population (Figure 6).
 This suggests that regardless of the treatment or category, these sample sizes represent a good starting point for subject recruitment when using stool samples.
@@
 Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that there was not a consistent association between alpha diversity and obesity; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference.
-Our analysis also used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from three studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from five studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012].
+Additionally, our analysis used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from three studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from five studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012].
 Our analysis included four of these studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011] and excluded three of the studies because they were too small [@ley_microbial_2006], only utilized metagenomic data [@Qin2010], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering [@yatsunenko_human_2012].
@@
 For each study we developed models using either OTUs or genus-level phylotypes.
-The quality of each model was assessed by measuring the area under the curve (AUC) of the Receive Operating Characteristic (ROC) using ten-fold cross validation.
+The quality of each model was assessed by measuring the area under the curve (AUC) of the Receiver Operating Characteristic (ROC) using ten-fold cross validation.
 Because the genus-level phylotype models were developed using a common reference, it was possible to use one study's model (i.e.
@@
 This threshold was then used to calculate the accuracy of the model applied to the test studies.
-To generate Receiver Operator Characteristic (ROC) curves and calculate the accuracy of the models we used the pROC (`r packageDescription("pROC")$Version`) R package.
+To generate ROC curves and calculate the accuracy of the models we used the pROC (`r packageDescription("pROC")$Version`) R package.
 Finally, we performed power and sample number simulations for different effect sizes for each study using the pwr (`r packageDescription("pwr")$Version`) R package and base R functions.

### 2016-06-06 submission/Sze_Obesity_mBio_2016.tex
Run make write.paper with Marc's edits

@@
 Two recent studies have re-analyzed published data and found that when
-datasets are analyzed independently there was limited support the widely
-accepted hypothesis that changes in the microbiome are associated with
-obesity. This hypothesis was reconsidered by increasing the number of
-data sets and pooling the results across the individual datasets. The
+datasets are analyzed independently there was limited support for the
+widely accepted hypothesis that changes in the microbiome are associated
+with obesity. This hypothesis was reconsidered by increasing the number
+of datasets and pooling the results across the individual datasets. The
 Preferred Reporting Items for Systematic Reviews and Meta-Analyses
@@
 20 or older) and these statistics have seen little change since 2003
-(1). Traditionally the body mass index (BMI) has been used as the
-traditional method of classifying individuals as non-obese or obese (2).
-Recently, there has been increased interest in the role of the
-microbiome in modulating obesity (3, 4). If the microbiome does affect
-obesity status, then manipulating the microbiome could have a
-significant role in the future treatment of obesity and in helping to
-stem the current epidemic.
+(1). Traditionally, the body mass index (BMI) has been used to classify
+individuals as non-obese or obese (2). Recently, there has been
+increased interest in the role of the microbiome in modulating obesity
+(3, 4). If the microbiome does affect obesity status, then manipulating
+the microbiome could have a significant role in the future treatment of
+obesity and in helping to stem the current epidemic.
 
@@
 using animal models that were manipulated with antibiotics or
-colonization with varied communities appears to support the association
-since these manipulation yielded variation in animal weight (10--13).
-The purported association between the differences in the microbiome and
-obesity have been widely repeated with little attention given to the
-lack of a clear signal in human cohort studies.
+colonization with varied communities were manipulated with antibiotics
+or underwent colonization with varied communities appears to support the
+association since these manipulation yielded variation in animal weight
+(10--13). The purported association between the differences in the
+microbiome and obesity have been widely repeated with little attention
+given to the lack of a clear signal in human cohort studies.
 
@@
 studying obesity and the human microbiome. The initial search yielded
-187 studies. We identified number\_string{[}n{]} additional studies that
-were not designed to explicitly test for an association between the
-microbiome and obesity. We then manually curated the 196 studies to
-select those studies that included BMI and sequence data. This yielded
-10 eligible studies. An additional study was removed from our analysis
-because no individuals in the study had a BMI over 30. Among the final
-10 studies, 3 were from identified from our PubMed search (10, 15, 16),
-5 were originally identified from the 9 studies that did not explicitly
+187 studies. We identified ten additional studies that were not designed
+to explicitly test for an association between the microbiome and
+obesity. We then manually curated the 196 studies to select those
+studies that included BMI and sequence data. This yielded 10 eligible
+studies. An additional study was removed from our analysis because no
+individuals in the study had a BMI over 30. Among the final 10 studies,
+3 were from identified from our PubMed search (10, 15, 16), 5 were
+originally identified from the 10 studies that did not explicitly
 investigate obesity but included BMI data (17--21), and two datasets
@@
 \textbf{\emph{Alpha diversity analysis.}} We calculated the Shannon
-diversity index, richness, and Shannon evenness, the relative abundance
-of \emph{Bacteroidetes} and \emph{Firmicutes}, and the ratio of their
-relative abundance (B:F) for each sample. Once we transformed each of
-the six alpha diversity metrics to make them normally distributed, we
-used a t-test to identify significant associations between the alpha
-diversity metric and whether an individual was obese for each of the ten
-studies. The B:F and the relative abundance of \emph{Firmicutes} were
-not significantly associated with obesity in any study. We identified
-seven P-values that were less than 0.05: three studies indicated obese
-individuals had a lower richness, two studies indicated a significantly
-lower diversity, one study indicated a significantly lower evenness, and
-one study indicated a significantly higher relative abundance of
-\emph{Bacteroidetes} (Figures 2 and S1). These results largely match
-those of the Walters and Finucane re-analysis studies. Interestingly,
-although only two of the ten studies observed the previously reported
-association between lower diversity and obesity, the other studies
-appeared to have the same trend, albeit the differences were not
-statistically significant. We used a random effects linear model to
-combine the studies using the study as the random effect and found
-statistical support for decreased richness, evenness, and diversity
-among obese individuals (all P\textless{}0.011). Although there was a
-significant relationship between these metrics and obesity status, the
-effect size was quite small. The obese individuals averaged 7.47\% lower
-richness, 0.88\% lower evenness, and 2.07\% lower diversity. There were
-no significant associations when we pooled the phylum-level metrics
-across studies. These results indicate that obese individuals do have a
-statistically significant lower diversity than non-obese individuals;
-however, it is questionable whether the difference is biologically
-significant.
+diversity index, observed richness, and Shannon evenness, the relative
+abundance of \emph{Bacteroidetes} and \emph{Firmicutes}, and the ratio
+of their relative abundance (B:F) for each sample. Once we transformed
+each of the six alpha diversity metrics to make them normally
+distributed, we used a t-test to identify significant associations
+between the alpha diversity metric and whether an individual was obese
+for each of the ten studies. The B:F and the relative abundance of
+\emph{Firmicutes} were not significantly associated with obesity in any
+study. We identified seven P-values that were less than 0.05: three
+studies indicated obese individuals had a lower richness, two studies
+indicated a significantly lower diversity, one study indicated a
+significantly lower evenness, and one study indicated a significantly
+higher relative abundance of \emph{Bacteroidetes} (Figures 2 and S1).
+These results largely match those of the Walters and Finucane
+re-analysis studies. Interestingly, although only two of the ten studies
+observed the previously reported association between lower diversity and
+obesity, the other studies appeared to have the same trend, albeit the
+differences were not statistically significant. We used a random effects
+linear model to combine the studies using the study as the random effect
+and found statistical support for decreased richness, evenness, and
+diversity among obese individuals (all P\textless{}0.011). Although
+there was a significant relationship between these metrics and obesity
+status, the effect size was quite small. The obese individuals averaged
+7.47\% lower richness, 0.88\% lower evenness, and 2.07\% lower
+diversity. There were no significant associations when we pooled the
+phylum-level metrics across studies. These results indicate that obese
+individuals do have a statistically significant lower diversity than
+non-obese individuals; however, it is questionable whether the
+difference is biologically significant.
 
@@
 the membership and structure of the individuals from each study. We then
-used AMOVA to test for a significant differences between the structure
-of non-obese and obese individuals (Table 1). The Escobar, Goodrich, and
+used AMOVA to test for significant differences between the structure of
+non-obese and obese individuals (Table 1). The Escobar, Goodrich, and
 Turnbaugh datasets indicated a significant difference in community
@@
 models when applied to the other nine datasets (Figure 5). Although
-there considerable variation in accuracy values for each model, the
+there was considerable variation in accuracy values for each model, the
 median accuracy for each model varied between 0.33 (Turnbaugh) and 0.65
@@
 either metric. In the case of obesity, the studies were underpowered to
-detect the \texttt{signif(range\_effect\_size,\ 1){[}1{]}} to 6\%
-difference in diversity that was observed across the studies.
+detect the 0.90 to 6\% difference in diversity that was observed across
+the studies.
 
@@
 
-Our meta-analysis helps to provide a clarity to the ongoing debate of
+Our meta-analysis helps to provide clarity to the ongoing debate of
 whether or not there are specific microbiome-based markers that can be
@@
 Considering these datasets are among the largest published, it appears
-that most microbiome studies are underpowered to detect differences in
-alpha diversity.
+that most human microbiome studies are underpowered to detect
+differences in alpha diversity.
 
@@
 showed that it was necessary to obtain approximately 140 and 6,300
-sequences per sample to detect a 5\% difference in Shannon diversity or
+samples per group to detect a 5\% difference in Shannon diversity or
 B:F, respectively. It was interesting that these sample sizes agreed
-across studies regardless of their sequencing method, region with in the
+across studies regardless of their sequencing method, region within the
 16S rRNA gene, or subject population (Figure 6). This suggests that
@@
 assess whether the studies were sufficiently powered to detect a
-difference. Our analysis also used 16S rRNA gene sequence data from ten
-studies whereas the Finucane study used 16S rRNA gene sequence data from
-three studies (7, 10, 21) and a metagenomic study (25) and the Walters
-study used 16S rRNA gene sequence data from five studies (10, 15, 20,
-21, 26); two studies were included in both analyses (10, 21). Our
-analysis included four of these studies (10, 15, 20, 21) and excluded
-three of the studies because they were too small (7), only utilized
-metagenomic data (25), or used short single read Illumina HiSeq data
-that has a high error rate making it untractable for \emph{de novo} OTU
-clustering (26). The additional seven datasets were published after the
-two reviews were performed and include datasets with more samples than
-were found in the original studies. Our collection of ten studies
+difference. Additionally, our analysis used 16S rRNA gene sequence data
+from ten studies whereas the Finucane study used 16S rRNA gene sequence
+data from three studies (7, 10, 21) and a metagenomic study (25) and the
+Walters study used 16S rRNA gene sequence data from five studies (10,
+15, 20, 21, 26); two studies were included in both analyses (10, 21).
+Our analysis included four of these studies (10, 15, 20, 21) and
+excluded three of the studies because they were too small (7), only
+utilized metagenomic data (25), or used short single read Illumina HiSeq
+data that has a high error rate making it untractable for \emph{de novo}
+OTU clustering (26). The additional seven datasets were published after
+the two reviews were performed and include datasets with more samples
+than were found in the original studies. Our collection of ten studies
 allowed us to largely use the same sequence analysis pipeline for all
@@
 phylotypes. The quality of each model was assessed by measuring the area
-under the curve (AUC) of the Receive Operating Characteristic (ROC)
+under the curve (AUC) of the Receiver Operating Characteristic (ROC)
 using ten-fold cross validation. Because the genus-level phylotype
@@
 calculate the accuracy of the model applied to the test studies. To
-generate Receiver Operator Characteristic (ROC) curves and calculate the
-accuracy of the models we used the pROC (1.8) R package. Finally, we
-performed power and sample number simulations for different effect sizes
-for each study using the pwr (1.1-3) R package and base R functions. We
-also calculated the actual sample size needed based on the effect size
-of each individual study.
+generate ROC curves and calculate the accuracy of the models we used the
+pROC (1.8) R package. Finally, we performed power and sample number
+simulations for different effect sizes for each study using the pwr
+(1.1-3) R package and base R functions. We also calculated the actual
+sample size needed based on the effect size of each individual study.
 
@@
 high-fat diet. Genome Medicine \textbf{8}.
-doi:\url{http://doi.org/10.1186/s13073-016-0297-9}.
+doi:\href{https://doi.org/10.1186/s13073-016-0297-9}{10.1186/s13073-016-0297-9}.
 
-\hyperdef{}{ref-moherux5fpreferredux5f2010}{\label{ref-moherux5fpreferredux5f2010}}
+\hypertarget{ref-moherux5fpreferredux5f2010}{}
 14. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
@@
 International Journal of Surgery (London, England) \textbf{8}:336--341.
-doi:\url{http://doi.org/10.1016/j.ijsu.2010.02.007}.
+doi:\href{https://doi.org/10.1016/j.ijsu.2010.02.007}{10.1016/j.ijsu.2010.02.007}.
 
-\hyperdef{}{ref-zupancicux5fanalysisux5f2012}{\label{ref-zupancicux5fanalysisux5f2012}}
+\hypertarget{ref-zupancicux5fanalysisux5f2012}{}
 15. \textbf{Zupancic ML}, \textbf{Cantarel BL}, \textbf{Liu Z},
@@
 syndrome. PloS One \textbf{7}:e43052.
-doi:\url{http://doi.org/10.1371/journal.pone.0043052}.
+doi:\href{https://doi.org/10.1371/journal.pone.0043052}{10.1371/journal.pone.0043052}.
 
-\hyperdef{}{ref-escobarux5fgutux5f2014}{\label{ref-escobarux5fgutux5f2014}}
+\hypertarget{ref-escobarux5fgutux5f2014}{}
 16. \textbf{Escobar JS}, \textbf{Klotz B}, \textbf{Valdes BE},
@@
 that of Americans, Europeans and Asians. BMC microbiology
-\textbf{14}:311. doi:\url{http://doi.org/10.1186/s12866-014-0311-6}.
+\textbf{14}:311.
+doi:\href{https://doi.org/10.1186/s12866-014-0311-6}{10.1186/s12866-014-0311-6}.
 
-\hyperdef{}{ref-zeeviux5fpersonalizedux5f2015}{\label{ref-zeeviux5fpersonalizedux5f2015}}
+\hypertarget{ref-zeeviux5fpersonalizedux5f2015}{}
 17. \textbf{Zeevi D}, \textbf{Korem T}, \textbf{Zmora N},
@@
 Prediction of Glycemic Responses. Cell \textbf{163}:1079--1094.
-doi:\url{http://doi.org/10.1016/j.cell.2015.11.001}.
+doi:\href{https://doi.org/10.1016/j.cell.2015.11.001}{10.1016/j.cell.2015.11.001}.
 
-\hyperdef{}{ref-rossux5f16sux5f2015}{\label{ref-rossux5f16sux5f2015}}
+\hypertarget{ref-rossux5f16sux5f2015}{}
 18. \textbf{Ross MC}, \textbf{Muzny DM}, \textbf{McCormick JB},
@@
 16S gut community of the Cameron County Hispanic Cohort. Microbiome
-\textbf{3}:7. doi:\url{http://doi.org/10.1186/s40168-015-0072-y}.
+\textbf{3}:7.
+doi:\href{https://doi.org/10.1186/s40168-015-0072-y}{10.1186/s40168-015-0072-y}.
 
-\hyperdef{}{ref-goodrichux5fhumanux5f2014}{\label{ref-goodrichux5fhumanux5f2014}}
+\hypertarget{ref-goodrichux5fhumanux5f2014}{}
 19. \textbf{Goodrich JK}, \textbf{Waters JL}, \textbf{Poole AC},
@@
 \textbf{159}:789--799.
-doi:\url{http://doi.org/10.1016/j.cell.2014.09.053}.
+doi:\href{https://doi.org/10.1016/j.cell.2014.09.053}{10.1016/j.cell.2014.09.053}.
 
-\hyperdef{}{ref-wuux5flinkingux5f2011}{\label{ref-wuux5flinkingux5f2011}}
+\hypertarget{ref-wuux5flinkingux5f2011}{}
 20. \textbf{Wu GD}, \textbf{Chen J}, \textbf{Hoffmann C},
@@
 patterns with gut microbial enterotypes. Science (New York, NY)
-\textbf{334}:105--108. doi:\url{http://doi.org/10.1126/science.1208344}.
+\textbf{334}:105--108.
+doi:\href{https://doi.org/10.1126/science.1208344}{10.1126/science.1208344}.
 
-\hyperdef{}{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{\label{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}}
+\hypertarget{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{}
 21. \textbf{Human Microbiome Project Consortium}. 2012. Structure,
 function and diversity of the healthy human microbiome. Nature
-\textbf{486}:207--214. doi:\url{http://doi.org/10.1038/nature11234}.
+\textbf{486}:207--214.
+doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
 
-\hyperdef{}{ref-schubertux5fmicrobiomeux5f2014}{\label{ref-schubertux5fmicrobiomeux5f2014}}
+\hypertarget{ref-schubertux5fmicrobiomeux5f2014}{}
 22. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
@@
 \textbf{5}:e01021--01014.
-doi:\url{http://doi.org/10.1128/mBio.01021-14}.
+doi:\href{https://doi.org/10.1128/mBio.01021-14}{10.1128/mBio.01021-14}.
 
-\hyperdef{}{ref-baxterux5fmicrobiota-basedux5f2016}{\label{ref-baxterux5fmicrobiota-basedux5f2016}}
+\hypertarget{ref-baxterux5fmicrobiota-basedux5f2016}{}
 23. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
@@
 Genome Medicine \textbf{8}:37.
-doi:\url{http://doi.org/10.1186/s13073-016-0290-3}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
 
-\hyperdef{}{ref-Ioannidis2005}{\label{ref-Ioannidis2005}}
+\hypertarget{ref-Ioannidis2005}{}
 24. \textbf{Ioannidis JPA}. 2005. Why most published research findings
 are false. PLoS Med \textbf{2}:e124.
-doi:\url{http://doi.org/10.1371/journal.pmed.0020124}.
+doi:\href{https://doi.org/10.1371/journal.pmed.0020124}{10.1371/journal.pmed.0020124}.
 
-\hyperdef{}{ref-Qin2010}{\label{ref-Qin2010}}
+\hypertarget{ref-Qin2010}{}
 25. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
@@
 catalogue established by metagenomic sequencing. Nature
-\textbf{464}:59--65. doi:\url{http://doi.org/10.1038/nature08821}.
+\textbf{464}:59--65.
+doi:\href{https://doi.org/10.1038/nature08821}{10.1038/nature08821}.
 
-\hyperdef{}{ref-yatsunenkoux5fhumanux5f2012}{\label{ref-yatsunenkoux5fhumanux5f2012}}
+\hypertarget{ref-yatsunenkoux5fhumanux5f2012}{}
 26. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
@@
 gut microbiome viewed across age and geography. Nature
-\textbf{486}:222--227. doi:\url{http://doi.org/10.1038/nature11053}.
+\textbf{486}:222--227.
+doi:\href{https://doi.org/10.1038/nature11053}{10.1038/nature11053}.
 
-\hyperdef{}{ref-Kozichux5f2013}{\label{ref-Kozichux5f2013}}
+\hypertarget{ref-Kozichux5f2013}{}
 27. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
@@
 
-\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
+\hypertarget{ref-Schloss2011}{}
 28. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
@@
 16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
-doi:\url{http://doi.org/10.1371/journal.pone.0027310}.
+doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
 
-\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
+\hypertarget{ref-Edgar2011}{}
 29. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
@@
 and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
-doi:\url{http://doi.org/10.1093/bioinformatics/btr381}.
+doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
 
-\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
+\hypertarget{ref-Westcott2015}{}
 30. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
@@
 sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\url{http://doi.org/10.7717/peerj.1487}.
+doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
 
-\hyperdef{}{ref-mothur}{\label{ref-mothur}}
+\hypertarget{ref-mothur}{}
 31. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
@@
 
-\hyperdef{}{ref-Anderson2001}{\label{ref-Anderson2001}}
+\hypertarget{ref-Anderson2001}{}
 32. \textbf{Anderson MJ}. 2001. A new method for non-parametric
 multivariate analysis of variance. Austral Ecology \textbf{26}:32--46.
-doi:\url{http://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}.
+doi:\href{https://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}{10.1111/j.1442-9993.2001.01070.pp.x}.
 
-\hyperdef{}{ref-Moher2009}{\label{ref-Moher2009}}
+\hypertarget{ref-Moher2009}{}
 33. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},

### 2016-06-06 submission/Sze_Obesity_mBio_2016.Rmd
Fix title

@@
 ---
-title: "Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome"
+title: ""
 bibliography: reference.bib
@@
 
-# Lack of power and small effect size confounds the ability to differentiate non-obese and obese individuals using gut microbiome data
+# Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome
 

### 2016-06-06 submission/Sze_Obesity_mBio_2016.tex
Fix title

@@
 \hypersetup{unicode=true,
-            pdftitle={Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome}
-  \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
+  \title{}
+  \pretitle{\vspace{\droptitle}}
+  \posttitle{}
   \author{}
@@
 \begin{document}
-\maketitle
 
-\section{Lack of power and small effect size confounds the ability to
-differentiate non-obese and obese individuals using gut microbiome
-data}\label{lack-of-power-and-small-effect-size-confounds-the-ability-to-differentiate-non-obese-and-obese-individuals-using-gut-microbiome-data}
+\section{Looking for a Signal in the Noise: Revisiting Obesity and the
+Microbiome}\label{looking-for-a-signal-in-the-noise-revisiting-obesity-and-the-microbiome}
 

### 2016-06-22 results/tables/table_1.Rmd
Ignore NAs when calculating percent obese
Fix Schubert entry in Table 1

@@
 	ross = "V1-V3",
-	schubert = "V4",
+	schubert = "V3-V5",
 	turnbaugh = "V2",
@@
 	ross = "454",
-	schubert = "MiSeq",
+	schubert = "454",
 	turnbaugh = "454",
@@
 
-	per_obese <- format(round(100 * mean(metadata$obese), 1), nsmall=1L)
+	per_obese <- format(round(100 * mean(metadata$obese, na.rm=T), 1), nsmall=1L)
 

### 2016-06-22 code/run_aucrf_genus.R
Generalize code to analyze any taxonomic level

@@ -2,40 +2,58 @@ source('code/utilities.R')
 source('code/cross_validate.R')
 get_dependencies(c('AUCRF', 'pROC'))
 
-get_genus_shared <- function(dataset){
+
+tax_name <- c("kingdom", "phylum", "class", "order", "family", "genus")
+
+
+get_tax_substring <- function(tax, tax_level){
+	substring <- unlist(strsplit(tax, ";"))[1:tax_level]
+	paste(substring, collapse='.')
+}
+
+get_tax_name <- function(tax_file, tax_level){
+
+	tax_file <- paste0('data/', dataset, '/', dataset, '.taxonomy')
+
+	tax_data <- read.table(file=tax_file, header=T, stringsAsFactors=F)
+	taxonomy <- tax_data$Taxonomy
+	taxonomy <- gsub("\\(\\d*\\)", "", taxonomy)
+	taxonomy <- gsub('"', '', taxonomy)
+
+	tax_substring <- sapply(taxonomy, get_tax_substring, tax_level)
+
+	names(tax_substring) <- tax_data$OTU
+
+	tax_substring
+}
+
+get_tax_level_shared <- function(dataset, tax_level){
 
 	shared_file <- paste0('data/', dataset, '/', dataset, '.0.03.subsample.shared')
 	shared_otus <- read.table(file=shared_file, header=T, stringsAsFactors=F, row.names=2)[,-c(1,2)]
 	is_present <- apply(shared_otus, 2, sum) > 0
 	shared <- shared_otus[,is_present]
 
-	tax_file <- paste0('data/', dataset, '/', dataset, '.taxonomy')
-	tax_data <- read.table(file=tax_file, header=T, stringsAsFactors=F)
-	taxonomy <- tax_data$Taxonomy
-	taxonomy <- gsub("\\(\\d*\\)", "", taxonomy)
-	taxonomy <- gsub('"', '', taxonomy)
-	taxonomy <- gsub(';', '.', taxonomy)
-	names(taxonomy) <- tax_data$OTU
+	taxonomy <- get_tax_name(tax_file, tax_level)
 	taxonomy <- taxonomy[colnames(shared)]
-
 	unique_taxa <- levels(as.factor(taxonomy))
 
-	shared_genus <- NULL
+	shared_tax_level <- NULL
 
 	for(ut in unique_taxa){
 		otus <- names(taxonomy[taxonomy %in% ut])
[…]

### 2016-06-23 submission/Sze_Obesity_mBio_2016.Rmd
Create funnel plot diagram and incorporate into text

@@
 Building upon the alpha diversity analysis we calculated the relative risk of being obese based on an individual's alpha diversity metrics relative to the median metric for that study.
-The results using relative risk largely matched those of using the untransformed alpha diversity data.
-Across the `r number_string[n]` studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2).
+Inspection of funnel plots for each of the metrics suggested that the studies included in our analysis were not biased (Figure S2).
+The results using relative risk largely matched those of using the raw alpha diversity data.
+Across the `r number_string[n]` studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S3).
 Again, although the relative risk values were not significant for other studies, the values tended to be above one.
@@
 The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals.
-To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S3-S8).
+To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S4-S9).
 Although there is no biological rationale for these effect sizes, they represent a range that is plausible.
 Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6).
-None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S5).
+None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S4).
 In fact, the maximum power among any of the studies to detect a 15% difference in B:F values was `r max(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"])`.
@@
 
-**Figure S2: Meta Analysis of the Relative Risk of Obesity Based on Evenness (A), Richness (B), or the Relative Abundance of *Bacteroidetes* (C) and Firmictues (D).**
+**Figure S2: Funnel plots depicting the general lack of bias in the selection of datasets included in the analysis.**
 
-**Figure S3: Power (A) and sample size simulations (B) for B:F for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S3: Meta Analysis of the Relative Risk of Obesity Based on Evenness (A), Richness (B), or the Relative Abundance of *Bacteroidetes* (C) and Firmictues (D).**
 
-**Figure S4: Power (A) and sample size simulations (B) for richness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S4: Power (A) and sample size simulations (B) for B:F for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure S5: Power (A) and sample size simulations (B) for evenness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S5: Power (A) and sample size simulations (B) for richness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure S6: Power (A) and sample size simulations (B) for the relative abundance of *Bacteroidetes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S6: Power (A) and sample size simulations (B) for evenness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure S7: Power (A) and sample size simulations (B) for the relative abundance of *Firmicutes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S7: Power (A) and sample size simulations (B) for the relative abundance of *Bacteroidetes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure S8: Power (A) and sample size simulations (B) for relative risk of obesity based on Shannon diversity.**  Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S8: Power (A) and sample size simulations (B) for the relative abundance of *Firmicutes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+
+**Figure S9: Power (A) and sample size simulations (B) for relative risk of obesity based on Shannon diversity.**  Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 

### 2016-06-23 code/plot_funnels.R

@@ -1,0 +1,58 @@
+pretty_metric <- c("shannon" = "Shannon", "sobs" = "Richness",
+									"shannoneven" = "Evenness",
+									"bf_ratio" = "Bacteroidetes:Firmicutes",
+									"bacteroidetes" = "Bacteroidetes",
+									"firmicutes" = "Firmicutes")
+
+pch <- c(baxter=21, escobar=21, goodrich=21, hmp=21, ross=21,
+				schubert=21, turnbaugh=21, wu=21, zeevi=21, zupancic=21)
+col <- c(baxter="black", escobar="red", goodrich="green", hmp="blue",
+ 				ross="orange", schubert="black", turnbaugh="red", wu="green",
+				zeevi="blue", zupancic="orange")
+bg <- c(baxter="white", escobar="white", goodrich="white", hmp="white",
+ 				ross="white", schubert="black", turnbaugh="red", wu="green",
+				zeevi="blue", zupancic="orange")
+names <- c(baxter="Baxter", escobar="Escobar", goodrich="Goodrich", hmp="HMP",
+ 				ross="Ross", schubert="Schubert", turnbaugh="Turnbaugh", wu="Wu",
+				zeevi="Zeevi", zupancic="Zupancic")
+
+my_funnel <- function(metric, rr){
+	metric_frame <- rr[rr$metric == metric, ]
+
+	metric_frame$se <- (log(metric_frame$upper) - log(metric_frame$est)) / 1.96
+
+	overall <- mean(log(metric_frame$est))
+	se <- seq(0,1,0.01)
+	l <- overall - 1.96 * se
+	u <- overall + 1.96 * se
+
+	par(mar=c(2.5,2,1.5,0.5))
+	plot(metric_frame$se~log(metric_frame$est), ylim=c(1,0), ylab="Standard Error", pch=pch[metric_frame$dataset], col=col[metric_frame$dataset], bg=bg[metric_frame$dataset])
+	abline(v=overall)
+
+	points(se~l, type="l")
+	points(se~u, type="l")
+	text(x=par()$usr[1], y=-0.1, label=pretty_metric[metric], xpd=TRUE, font=2, cex=1.3, pos=4)
+}
+
+rr <- read.table(file="data/process/relative_risk.summary", header=T)
+
[…]

### 2016-07-11 submission/Response_to_reviewers.md
Initial commit of formatted reviewers' comments

**Reviewer #1 (Comments for the Author):

Major Comments
• P.5 (line 81): The bioinformatics behind the statistical analyses are important. Include a brief summary in the main text. Specifically, it is important to clarify that you focused on studies that collected targeted 16S sequence data (not shotgun metagenomes). Also briefly mention that you did OTU clustering from scratch (per study and pooled?). Here or in the methods, clarify how alpha and beta diversity statistics were estimated. **

**• P.7 (lines 134-136): To overcome challenges with non-overlapping 16S regions, might it be possible to predict taxonomy of OTUs (which it appears you have already done for phylum level analyses) and then do a pooled community structure analysis on the taxa rather than OTUs? Alternatively, perhaps tools designed for identifying OTUs from non-overlapping shotgun marker gene reads might work. **

**• P.8 (line 162): What variability was used in the calculations (presumably that observed in the studies)? For designing future studies, it would be great to investigate the sensitivity of the results to the amount of between-sample variance, since variance does differ between studies (e.g., is higher in rural populations and is lower between moms and their newborns than between unrelated individuals). **

**• P.10 (line 204): When discussing biological relevance, it would be useful to also mention that relative abundance may not be the relevant factor for the host. Imagine an OTU (A) that produces a metabolite with a physiological effect on obesity (or whatever host trait one is studying). Suppose the number of cells and the amount of metabolite produced remains constant, but the number of cells for some other OTU (B) goes up by an order of magnitude. The relative abundance of A in the community has gone down, but this may be irrelevant to the molecular/cellular mechanism affecting obesity, especially if B has no relationship to A or obesity (e.g., it occupies a distinct niche in the gut). The same argument probably holds for alpha diversity metrics based on relative abundances - the host likely cares about specific components of the diversity not diversity per se. **

**• P.12 (line 244): Where can the pipeline be obtained? This info is in the methods but should probably be highlighted.**

**• P.12 (lines 265-267): Elaborate further on this point. The right biomarker may be a protein or pathway. Even if this is taxonomically restricted, it may be hard to detect with community-level statistics or even OTU relative abundances. In addition to the issue noted above regarding relative abundance, species-level OTUs can still miss functional variability due to large differences in gene content among strains with identical or nearly identical 16S sequences. **

**Minor Comments
• P.4 (lines 53-56): repeated text and typos: "mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities were manipulated with antibiotics or underwent colonization with varied communities appears to support the association since these manipulation yielded" **

**P.6 (line 100) and elsewhere: When first referring to a study by name (e.g., Walters, Finucane, Goodrich) provide the citation. This info is in Table 1, but it would be helpful to provide in the text as well. **

**P.6 (line 103): In the random effects model does study have a random intercept or slope (or both)? **

**P.6 (line 88): What is the overlap of studies with the Walters and Finucane analyses (could move or copy this from the discussion)? **

**P.6 (line 88): What is the range of sample size per study (in Table 1, but would be helpful to give this overview statistic in the text)? **

**P.7 (lines 122-123): Clarify if ranges on relative risk are 95% confidence intervals or some other statistic. **

**P.7 (line 123): Fix plural: "were no significant difference" **

**P.7 (lines 132-133): Explain why directionality cannot be inferred. **

**P.13 (line 281): Is this algorithm the same as agglomerative hierarchical clustering with average linkage? Also, provide some more detail on how the read counts for OTUs were used to estimate the various alpha and beta diversity statistics. **

**P.14 (line 297): Typo: "each study within a study" **

**Reviewer #2 (Comments for the Author):

General comments:
Overall this paper is a useful meta analysis of obesity and the microbiome. I have some miscellaneous comments that are below. In addition I have one more important concern that I believe needs to be addressed.

More important concern:
Given that their analysis depends on them compiling data from multiple studies and reanalyzing that data, I believe it is necessary for these authors to make sure that all components of this analysis are reproducible. I commend the authors for their commitment to sharing code for all the work. However, I have a few concerns about the data. First, some of the data is from a personal web site. I would recommend that the authors here either repost that data to a more sustainable site (e.g., Figshare), get the original producers of the data to post it to a more sustainable site, or come up with some other solution. Again, I commend the authors for their efforts in openness and reproducibility, I am just concerned that some aspects of this paper will not be reproducible. **

**Additional concerns and comments:
L47 "individuals had a lower diversity than lean individuals (6)." Specify what kind of diversity **

**L54 "with antibiotics or colonization with varied communities were manipulated with antibiotics" Typo? Should this be "that were manipulated?" **

**L73 "Literature Review and Study Inclusion. We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis (14)." Would be good to say something about whether this is considered an important approach in meta-analysis **

**L77 "we searched PubMed for original research studies that involved studying obesity and the..." Given the importance of covering the literature well in meta-analyses it would be good to say how Pubmed was searched.**

**L78 "We identified ten additional studies" How?**

**L80 "and obesity. We then manually curated the 196 studies to select." Numbers don't add up - 187 + 10 = 197. I looked through the Figure and I am not sure this explains all the numbers either.**

**L102 "other studies appeared to have the same trend, albeit the differences were not statistically significant." Quibble about wording here. If the result is not significant - is it a trend?**

**L110 "significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant." Some justification for this statement is needed.**

**L126 "obese, it is questionable whether that risk is biologically or clinically relevant." Again, Some justification for this statement is needed.**

**L164 "each of the studies (Figures 6, S3-S8). Although there is no biological rationale for these effect sizes, they represent a range that is plausible." It would help to clarify what is meant here by "range that is plausible"?**

**L193 "datasets. This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome was limited" I think it would be good here to clarify that this is really the ability to classify based solely on microbiome data. It might be possible to do this much better if individuals were subclassified (e.g., by genotype, age, gender, etc)

**Some references could / should be added: lme4 (v.1.1-12) R package, vegan (v.2.3-5) R package, AUCRF (1.1) R package, pROC (1.8) R package, pwr (1.1-3) R package**

**Table 1 - would be good to add the references for the studies**

**Figure 1 - some of the boxes in the flow chart are cutoff.**

### 2016-07-11 submission/Response_to_reviewers.md
Respond to reviewer's comments and edit text accordingly

@@
 Major Comments
-• P.5 (line 81): The bioinformatics behind the statistical analyses are important.
+P.5 (line 81): The bioinformatics behind the statistical analyses are important.
 Include a brief summary in the main text.
@@
 Also briefly mention that you did OTU clustering from scratch (per study and pooled?).
-Here or in the methods, clarify how alpha and beta diversity statistics were estimated.
-**
+Here or in the methods, clarify how alpha and beta diversity statistics were estimated.**
 
-**• P.7 (lines 134-136): To overcome challenges with non-overlapping 16S regions, might it be possible to predict taxonomy of OTUs (which it appears you have already done for phylum level analyses) and then do a pooled community structure analysis on the taxa rather than OTUs?
-Alternatively, perhaps tools designed for identifying OTUs from non-overlapping shotgun marker gene reads might work.
-**
+We have added text to address these comments (LXXX).
 
-**• P.8 (line 162): What variability was used in the calculations (presumably that observed in the studies)?
-For designing future studies, it would be great to investigate the sensitivity of the results to the amount of between-sample variance, since variance does differ between studies (e.g., is higher in rural populations and is lower between moms and their newborns than between unrelated individuals).
-**
+**P.7 (lines 134-136): To overcome challenges with non-overlapping 16S regions, might it be possible to predict taxonomy of OTUs (which it appears you have already done for phylum level analyses) and then do a pooled community structure analysis on the taxa rather than OTUs?
+Alternatively, perhaps tools designed for identifying OTUs from non-overlapping shotgun marker gene reads might work.**
 
-**• P.10 (line 204): When discussing biological relevance, it would be useful to also mention that relative abundance may not be the relevant factor for the host.
+We did this in the random forest analysis at the genus level and did not see a difference in the results relative to the OTU-based analysis.
+
+ - need to add extra text for RF analysis at all of the levels.
+
+**P.8 (line 162): What variability was used in the calculations (presumably that observed in the studies)?
+For designing future studies, it would be great to investigate the sensitivity of the results to the amount of between-sample variance, since variance does differ between studies (e.g., is higher in rural populations and is lower between moms and their newborns than between unrelated individuals).**
+
+We used the observed variation in the data from each study. Added a comment in discussion to address cohorts that would result in different levels of interpersonal variation.
+
+**P.10 (line 204): When discussing biological relevance, it would be useful to also mention that relative abundance may not be the relevant factor for the host.
 Imagine an OTU (A) that produces a metabolite with a physiological effect on obesity (or whatever host trait one is studying).
@@
 The relative abundance of A in the community has gone down, but this may be irrelevant to the molecular/cellular mechanism affecting obesity, especially if B has no relationship to A or obesity (e.g., it occupies a distinct niche in the gut).
-The same argument probably holds for alpha diversity metrics based on relative abundances - the host likely cares about specific components of the diversity not diversity per se.
-**
+The same argument probably holds for alpha diversity metrics based on relative abundances - the host likely cares about specific components of the diversity not diversity per se.**
 
-**• P.12 (line 244): Where can the pipeline be obtained? This info is in the methods but should probably be highlighted.**
+We agree with this sentiment and included text at L258-267 to point out that there are clear examples where changing the microbiome changes BMI profile.
+We have expanded on this point as suggested by the reviewer below.
 
-**• P.12 (lines 265-267): Elaborate further on this point.
+**P.12 (line 244): Where can the pipeline be obtained? This info is in the methods but should probably be highlighted.**
+
+We have added a link to the GitHub repository that contains the pipeline for hte analysis and this manuscript.
+
+**P.12 (lines 265-267): Elaborate further on this point.
 The right biomarker may be a protein or pathway.
 Even if this is taxonomically restricted, it may be hard to detect with community-level statistics or even OTU relative abundances.
-In addition to the issue noted above regarding relative abundance, species-level OTUs can still miss functional variability due to large differences in gene content among strains with identical or nearly identical 16S sequences.
-**
+In addition to the issue noted above regarding relative abundance, species-level OTUs can still miss functional variability due to large differences in gene content among strains with identical or nearly identical 16S sequences.**
+
+We have added text to this paragraph to expand upon this thought.
 
 **Minor Comments
-• P.4 (lines 53-56): repeated text and typos: "mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities were manipulated with antibiotics or underwent colonization with varied communities appears to support the association since these manipulation yielded" **
+P.4 (lines 53-56): repeated text and typos: "mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities were manipulated with antibiotics or underwent colonization with varied communities appears to support the association since these manipulation yielded"**
+
+This has been fixed.
 
 **P.6 (line 100) and elsewhere: When first referring to a study by name (e.g., Walters, Finucane, Goodrich) provide the citation.
-This info is in Table 1, but it would be helpful to provide in the text as well.
-**
+This info is in Table 1, but it would be helpful to provide in the text as well.**
 
-**P.6 (line 103): In the random effects model does study have a random intercept or slope (or both)? **
+We have added an explicit reference to the first time we mention the Walters and Finucane studies at LXXX.
+Also, to respond to a comment from Reviewer 2, we have added the reference numbers to Table 1.
+This should make the connection between the studies and specific references more clear.
 
-**P.6 (line 88): What is the overlap of studies with the Walters and Finucane analyses (could move or copy this from the discussion)? **
+**P.6 (line 103): In the random effects model does study have a random intercept or slope (or both)?**
 
-**P.6 (line 88): What is the range of sample size per study (in Table 1, but would be helpful to give this overview statistic in the text)? **
+Each study has a random intercept and slope.
 
-**P.7 (lines 122-123): Clarify if ranges on relative risk are 95% confidence intervals or some other statistic. **
+**P.6 (line 88): What is the overlap of studies with the Walters and Finucane analyses (could move or copy this from the discussion)?**
 
-**P.7 (line 123): Fix plural: "were no significant difference" **
+We have added text at LXX to indicate which of our studies overlapped with those from the Walters and Finucane.
 
-**P.7 (lines 132-133): Explain why directionality cannot be inferred. **
+**P.6 (line 88): What is the range of sample size per study (in Table 1, but would be helpful to give this overview statistic in the text)?**
+
+Given the broad variation in the number of subjects per study, it does not seem appropriate to reduce that variation to a single parameter or a range of values.
+The table lists the sample sizes for the ten studies.
+
+**P.7 (lines 122-123): Clarify if ranges on relative risk are 95% confidence intervals or some other statistic.**
+
+Added "95% CI" to each set of parentheses to indicate that these are the 95% confidence interval.
+
+**P.7 (line 123): Fix plural: "were no significant difference"**
+
+This has been fixed.
+
+**P.7 (lines 132-133): Explain why directionality cannot be inferred.**
+
+We have added text to clarify the point.
+Basically, in a univariate analysis, we can say that a group of people are taller or shorter than another, but we cannot do the analogous comparison to community structures since their relationship is in an no-dimensional space.
 
 **P.13 (line 281): Is this algorithm the same as agglomerative hierarchical clustering with average linkage?
-Also, provide some more detail on how the read counts for OTUs were used to estimate the various alpha and beta diversity statistics.
-**
+Also, provide some more detail on how the read counts for OTUs were used to estimate the various alpha and beta diversity statistics.**
 
-**P.14 (line 297): Typo: "each study within a study" **
+Yes it is.
+There are multiple names for the same algorithm.
+A citation is provided to make it clear what we are referring to.
+The number of reads that clustered within each OTU or taxonomic group were used to calculate the various alpha and beta-diversity metrics.
+We have added a citation to Magurran (2003) that describes the forumlas that were used to calculate the various parameters.
+
+**P.14 (line 297): Typo: "each study within a study"**
+
+This has been fixed.
 
@@
 I would recommend that the authors here either repost that data to a more sustainable site (e.g., Figshare), get the original producers of the data to post it to a more sustainable site, or come up with some other solution.
-Again, I commend the authors for their efforts in openness and reproducibility, I am just concerned that some aspects of this paper will not be reproducible.
-**
+Again, I commend the authors for their efforts in openness and reproducibility, I am just concerned that some aspects of this paper will not be reproducible.**
+
+We appreciate the reviewer's concern regarding the openness of the underlying data used in this study.
+As far as we can tell all of the sequence data is available either through the ENA, SRA, or dbGaP except for the Turnbaugh and Schubert datasets.
+Given the age of this dataset and the lack of access to the original sff files (we've asked previously) it is unlikely that continued requests to the Gordon lab will get these data to be made any more public.
+Furthermore, this is a keystone study in this research area because it was the original study to claim a relationship.
+The Schubert dataset is now available on the SRA.
+The reviewer is correct that some of the metadata is not publicly available (i.e.
+Wu and Zeevi).
+These were obtained from the original authors.
+We have encouraged the authors to make these data more available.
+Regardless of the outcome of that encouragement, we have posted the metadata for each study in the GitHub repository for this study.
+This is obviously a complex problem and we can only do our best to encourage our colleagues and follow best practices ourselves.
+Ultimately, we do not think it is necessary to censor datasets from the study because their data are not publicly available.
 
 **Additional concerns and comments:
-L47 "individuals had a lower diversity than lean individuals (6)." Specify what kind of diversity **
+L47 "individuals had a lower diversity than lean individuals (6)." Specify what kind of diversity**
 
-**L54 "with antibiotics or colonization with varied communities were manipulated with antibiotics" Typo? Should this be "that were manipulated?" **
+We have specified that it was alpha-diversity.
+It is actually unclear what type of diversity the original authors intended.
+Given that richness and diversity are often used interchangeably, it is difficult to know with certainty what was meant.
+
+**L54 "with antibiotics or colonization with varied communities were manipulated with antibiotics" Typo? Should this be "that were manipulated?"**
+
+We corrected this sentence.
 
 **L73 "Literature Review and Study Inclusion.
-We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis (14)." Would be good to say something about whether this is considered an important approach in meta-analysis **
+We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis (14)." Would be good to say something about whether this is considered an important approach in meta-analysis**
+
+We have edited the sentence to make it clear that following the PRISMA guidelines is necessary to prevent inclusion bias and to be considered a robust study.
 
@@
 
+The exact approach we used has been clarified in the Supplemental Materials.
+
 **L80 "and obesity. We then manually curated the 196 studies to select." Numbers don't add up - 187 + 10 = 197. I looked through the Figure and I am not sure this explains all the numbers either.**
+
+We corrected this error in the new review.
 
@@
 
-**L110 "significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant." Some justification for this statement is needed.**
+We appreciate the reviewers' concern, but alternative wording (e.g.
+effect sizes had the same sign) was clunky and effectively said the same as "trend".
+We do not think that a trend has to be significant to be a trend.
+We are also critical of others that claim pseudo-significance from things that are "trending", but feel we are being cautious here since we then do the aggregate test to show that the trending is in fact significant when considered across all of the studies.
 
-**L126 "obese, it is questionable whether that risk is biologically or clinically relevant." Again, Some justification for this statement is needed.**
+**L110 "significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant." Some justification for this statement is needed.
+
+L126 "obese, it is questionable whether that risk is biologically or clinically relevant." Again, Some justification for this statement is needed.**
+
+We expand upon these statements at L204 where we discuss the need to develop guidelines as to what is a biologically significant difference.
+There are likely some researchers who would contend that any statistically significant difference in diversity is also biologically significant; however, it is also possible for a study to be overpowered so that it could yield statistically significant differences that are not biologically significant.
+Our discussion at L204 comments on this distinction and the need to develop better theory around what is biologically significant for microbiome studies.
 
@@
 
+We have clarified that by plausible, we mean that the range includes values that most microbiome researchers consider to be biologically meaningful - even if they do not know what it means.
+Furthermore, most of the studies were sufficiently powered to detect effect sizes of 15%.
+
 **L193 "datasets.
 This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome was limited" I think it would be good here to clarify that this is really the ability to classify based solely on microbiome data.
-It might be possible to do this much better if individuals were subclassified (e.g., by genotype, age, gender, etc)
+It might be possible to do this much better if individuals were subclassified (e.g., by genotype, age, gender, etc)**
+
+We have highlighted the text to indicate that the analysis was based solely on microbiome  data.
 
@@
 
+We have added references for lme4, pROC, and AUCRF. The others have not been published.
+
 **Table 1 - would be good to add the references for the studies**
 
+We have added the citation number for each study to Table 1.
+
 **Figure 1 - some of the boxes in the flow chart are cutoff.**
+
+We have added whitespace around the boxes on the right side of the figure.

### 2016-07-11 submission/Sze_Obesity_mBio_2016.Rmd
Respond to reviewer's comments and edit text accordingly

@@
 The first such study used genetically obese mice and observed the ratio of the relative abundances of *Bacteroidetes* to *Firmicutes* (B:F) was lower in obese mice than lean mice [@ley_obesity_2005].
-Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009].
+Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower alpha-diversity than lean individuals [@turnbaugh_core_2009].
 They also showed that the relative abundance of *Bacteroidetes* and *Firmicutes* increased and decreased, respectively, as obese individuals lost weight while on a fat or carbohydrate restricted diet [@ley_microbial_2006].
-Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014].
-Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities
-were manipulated with antibiotics or underwent colonization with varied  communities appears to support the association since these manipulation yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016].
+Two re-analysis studies by Walters et al.
+[@walters_meta-analyses_2014] and Finucane et al.
+[@finucane_taxonomic_2014] interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized.
+Regardless of the results using human populations, studies using animal models where the community was manipulated with antibiotics or established by colonizing germ-free animals  with varied communities appear to support the association since these manipulations yielded differences in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016].
 The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.
@@
 Alternatively, diversity metrics may mask the appropriate signal and it is necessary to measure the association at the level of microbial populations.
-Walters et al.
-[-@walters_meta-analyses_2014] demonstrated that Random Forest machine learning models were capable of predicting obesity status within a single cohort, but did not attempt to test the models on other cohorts.
+The Walters re-analysis study demonstrated that Random Forest machine learning models were capable of predicting obesity status within a single cohort, but did not attempt to test the models on other cohorts.
 The purpose of this study was to perform a meta-analysis of the association between differences in the microbiome and obesity status by analyzing and applying a more systematic and synthetic approach than was used previously.
@@
 ***Literature Review and Study Inclusion.***
-We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010].
+To perform a robust meta-analysis and limit inclusion bias, we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify the studies that we analyzed [@moher_preferred_2010].
 A detailed description of our selection process and the exact search terms are provided in the Supplemental Text and in Figure 1.
@@
 We identified `r number_string[n]` additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
-We then manually curated the 196 studies to select those studies that included BMI and sequence data.
+We then manually curated the 196 studies to select those studies that included BMI and 16S rRNA gene sequence data.
 This yielded 10 eligible studies.
@@
 The `r number_string[n]` studies are summarized in Table 1.
+For comparison, two of these studies were included in the Finucane re-analysis study [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012] and four of these studies were included in the Walters re-analysis study [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011].
+The 16S rRNA gene sequence data from each study was re-analyzed using a similar approach based on previously described methods for reducing the number of chimeric sequences and sequencing errors for 454 and Illumina MiSeq data [@Kozich_2013; @Schloss2011].
+The sequences were either clustered into operational taxonomic units (OTUs) using the average neighbor approach [@Westcott2015] or taxonomic groupings based on their classification using a naive Bayesian classifier [@Wang2007].
 
@@
 
-	paste0(format(data$rr, digits=2, nsmall=2L), " (", format(data$ci_lb, digits=2, nsmall=2L), "-", format(data$ci_ub, digits=2, nsmall=2L), ")")
+	paste0(format(data$rr, digits=2, nsmall=2L), " (95% CI: ", format(data$ci_lb, digits=2, nsmall=2L), "-", format(data$ci_ub, digits=2, nsmall=2L), ")")
 
@@
 The relative risk of having a low richness was `r concatenate_rr(relative_composite["sobs",])`, low evenness was `r concatenate_rr(relative_composite["shannoneven",])`, and low diversity was `r concatenate_rr(relative_composite["shannon",])`.
-There were no significant difference in the phylum-level metrics.
+There were no significant differences in the phylum-level metrics.
 Again, the relative risk results indicate that individuals with a lower richness, evenness, or diversity are at statistically significant increased risk of being obese, it is questionable whether that risk is biologically or clinically relevant.
@@
 The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05).
-Because it was not possible to ascertain the directionality of the difference in community structure nor perform a pooled analysis using studies that had non-overlapping 16S rRNA gene sequence regions it is unclear whether these differences reflect a broader, but perhaps small, shift in community structure between non-obese and obese individuals.
+Because it was not possible to ascertain the directionality of the difference in community structure because the samples are arrayed in a non-dimensional space.
+nor perform a pooled analysis using studies that had non-overlapping 16S rRNA gene sequence regions it is unclear whether these differences reflect a broader, but perhaps small, shift in community structure between non-obese and obese individuals.
 
@@
 To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S4-S9).
-Although there is no biological rationale for these effect sizes, they represent a range that is plausible.
+Although there is no biological rationale for these effect sizes, they represent a range that includes effect sizes that would be generally considered to be biologically significant.
 Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6).
@@
 We also generated Random Forest machine learning models trained on each dataset and tested on the remaining datasets.
-This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome was limited.
+This analysis demonstrated that the ability to reliably classify individuals as being obese based solely on the composition of their microbiome was limited.
 Finally, we assessed the ability of each study to detect defined differences in alpha diversity and observed that most studies were underpowered to detect modest effect sizes.
@@
 There is wide interpersonal variation in the diversity and structure of the human microbiome.
+Some factors such as relationship between subjects could potentially decrease the amount variation [@turnbaugh_core_2009] and other factors such as whether one lives in a rural environment could increase the amount of variation [@yatsunenko_human_2012].
+It was interesting that although there were ranges in variation across the studies included in our analysis, the sample size calculations yielded similar results.
 In addition, the common experimental designs limit their power.
@@
 
-Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that there was not a consistent association between alpha diversity and obesity; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference.
+Two previous re-analysis studies have stated that there was not a consistent association between alpha diversity and obesity [@walters_meta-analyses_2014; @finucane_taxonomic_2014]; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference.
 Additionally, our analysis used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from three studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from five studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012].
@@
 Our collection of `r number_string[n]` studies allowed us to largely use the same sequence analysis pipeline for all datasets and relied heavily on the availability of public data and access to metadata that included variables beyond the needs of the original study.
-To execute this analysis, we created an automated data analysis pipeline, which can be easily updated to add additional studies as they become available.
+To execute this analysis, we created an automated data analysis pipeline, which can be easily updated to add additional studies as they become available (https://github.com/SchlossLab/Sze_Obesity_mBio_2016/).
 Similarly, it would be possible to adapt this pipeline to other body sites and treatment or variables (e.g.
@@
 
-Similar to our study, the Walters et al [@walters_meta-analyses_2014] generated Random Forest machine learning models to differentiate between non-obese and obese individuals.
+Similar to our study, the Walters study generated Random Forest machine learning models to differentiate between non-obese and obese individuals [@walters_meta-analyses_2014].
 They obtained similar AUC values to our analysis; however, they did not attempt to test these models on the other studies in their analysis.
@@
 Recalling the large interpersonal variation in the structure of the microbiome, it is possible that each individual has their own signatures of obesity.
-Alternatively, it could be that the involvement of the microbiome in obesity is at the level of a common set of metabolites that can be produced from different structures of the microbiome.
+Alternatively, it could be that the involvement of the microbiome in obesity is not apparent based on the taxonomic information provided by 16S rRNA gene sequence data.
+Rather, the differences could become more apparent at the level of a common set of gene transcripts or metabolites that can be produced from different structures of the microbiome.
 
@@
 We split the overall meta-analysis into three general strategies using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`).
-First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of *Bacteroidetes* and Firmicutes, the ratio of *Bacteroidetes* and *Firmicutes* relative abundances (B:F), Shannon diversity, observed richness, and Shannon evenness.
+First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of *Bacteroidetes* and Firmicutes, the ratio of *Bacteroidetes* and *Firmicutes* relative abundances (B:F), Shannon diversity, observed richness, and Shannon evenness [@Magurran2003].
 After each variable was transformed to fit a normal distribution a two-tailed t-test was performed for comparison of non-obese and obese individuals (i.e.
 BMI > 35.0).
-We performed a pooled analysis on these measured variables using linear random effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups using the lme4 (v.`r packageDescription("lme4")$Version`) R package.
+We performed a pooled analysis on these measured variables using linear random effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups using the lme4 (v.`r packageDescription("lme4")$Version`) R package [@lme4_2015].
 Next, we compared the community structure from non-obese and obese individuals using analysis of molecular variance (AMOVA) with Bray-Curtis distance matrices [@Anderson2001].
 This analysis was performed using the vegan (v.`r packageDescription("vegan")$Version`) R package.
-For both analyses, the datasets were rarefied (N=1000) so that each study within a study had the same number of sequences.
+For both analyses, the datasets were rarefied (N=1000) so that each study had the same number of sequences.
 Second, for each study we partitioned the subjects into a low or high group depending on whether their alpha diversity metrics were below or above the median value for the study.
@@
 We then performed a Fisher exact-test to investigate whether the RR was significantly different from 1.0 within each study and across all of the studies using the epiR (`r packageDescription("epiR")$Version`) and metafor (`r packageDescription("metafor")$Version`) packages.
-Third, we used the AUCRF (`r packageDescription("AUCRF")$Version`) R package to generate Random Forest models.
+Third, we used the AUCRF (`r packageDescription("AUCRF")$Version`) R package to generate Random Forest models [@Calle2011].
 For each study we developed models using either OTUs or genus-level phylotypes.
@@
 This threshold was then used to calculate the accuracy of the model applied to the test studies.
-To generate ROC curves and calculate the accuracy of the models we used the pROC (`r packageDescription("pROC")$Version`) R package.
+To generate ROC curves and calculate the accuracy of the models we used the pROC (`r packageDescription("pROC")$Version`) R package [@pROC_2011].
 Finally, we performed power and sample number simulations for different effect sizes for each study using the pwr (`r packageDescription("pwr")$Version`) R package and base R functions.

### 2016-07-11 submission/supp_text.Rmd
Respond to reviewer's comments and edit text accordingly

@@
 This search yielded a total of 187 manuscripts.
-From two previous other reviews of obesity and the bacterial microbiome along with knowledge of two other published papers that investigated obesity but were missed by the database search we obtained a total of 7 more articles.
+From two previous other reviews of obesity and the bacterial microbiome along with knowledge of two other published papers that investigated obesity but were missed by the database search we obtained a total of 10 more articles.
 We also had access to normal healthy individuals from an unpublished dataset.
-This brought our total number of records to 196.
+This brought our total number of records to 197.
 

### 2016-07-18 submission/Response_to_reviewers.md
Odds and ends

@@
 
-We have added a link to the GitHub repository that contains the pipeline for hte analysis and this manuscript.
+We have added a link to the GitHub repository that contains the pipeline for the analysis and this manuscript.
 
@@
 Furthermore, this is a keystone study in this research area because it was the original study to claim a relationship.
-The Schubert dataset is now available on the SRA.
+The Schubert dataset is now available on the SRA (accession SRP078489).
 The reviewer is correct that some of the metadata is not publicly available (i.e.

### 2016-07-18 submission/Response_to_reviewers.md
Add rules for response to reveiewers and track changes

@@
-**Reviewer #1 (Comments for the Author):
+---
+title: ""
+output:
+  pdf_document:
+    includes:
+      in_header: header.tex
+fontsize: 11pt
+geometry: margin=1.0in
+---
 
-Major Comments
+**Reviewer #1 (Comments for the Author):**
+
+**Major Comments
 P.5 (line 81): The bioinformatics behind the statistical analyses are important.
@@
 
-We have added text to address these comments (LXXX).
+> We have added text to address these comments (LXXX).
 
@@
 
-We did this in the random forest analysis at the genus level and did not see a difference in the results relative to the OTU-based analysis.
+> We did this in the random forest analysis at the genus level and did not see a difference in the results relative to the OTU-based analysis.
 
@@
 
-We used the observed variation in the data from each study. Added a comment in discussion to address cohorts that would result in different levels of interpersonal variation.
+> We used the observed variation in the data from each study. Added a comment in discussion to address cohorts that would result in different levels of interpersonal variation.
 
@@
 
-We agree with this sentiment and included text at L258-267 to point out that there are clear examples where changing the microbiome changes BMI profile.
+> We agree with this sentiment and included text at L258-267 to point out that there are clear examples where changing the microbiome changes BMI profile.
 We have expanded on this point as suggested by the reviewer below.
@@
 
-We have added a link to the GitHub repository that contains the pipeline for the analysis and this manuscript.
+> We have added a link to the GitHub repository that contains the pipeline for the analysis and this manuscript.
 
@@
 
-We have added text to this paragraph to expand upon this thought.
+> We have added text to this paragraph to expand upon this thought.
 
@@
 
-This has been fixed.
+> This has been fixed.
 
@@
 
-We have added an explicit reference to the first time we mention the Walters and Finucane studies at LXXX.
+> We have added an explicit reference to the first time we mention the Walters and Finucane studies at LXXX.
 Also, to respond to a comment from Reviewer 2, we have added the reference numbers to Table 1.
@@
 
-Each study has a random intercept and slope.
+> Each study has a random intercept and slope.
 
@@
 
-We have added text at LXX to indicate which of our studies overlapped with those from the Walters and Finucane.
+> We have added text at LXX to indicate which of our studies overlapped with those from the Walters and Finucane.
 
@@
 
-Given the broad variation in the number of subjects per study, it does not seem appropriate to reduce that variation to a single parameter or a range of values.
+> Given the broad variation in the number of subjects per study, it does not seem appropriate to reduce that variation to a single parameter or a range of values.
 The table lists the sample sizes for the ten studies.
@@
 
-Added "95% CI" to each set of parentheses to indicate that these are the 95% confidence interval.
+> Added "95% CI" to each set of parentheses to indicate that these are the 95% confidence interval.
 
@@
 
-We have added text to clarify the point.
+> We have added text to clarify the point.
 Basically, in a univariate analysis, we can say that a group of people are taller or shorter than another, but we cannot do the analogous comparison to community structures since their relationship is in an no-dimensional space.
@@
 
-Yes it is.
+> Yes it is.
 There are multiple names for the same algorithm.
@@
 
-This has been fixed.
+> This has been fixed.
 
-**Reviewer #2 (Comments for the Author):
+**Reviewer #2 (Comments for the Author):**
 
-General comments:
+**General comments:
 Overall this paper is a useful meta analysis of obesity and the microbiome.
 I have some miscellaneous comments that are below.
-In addition I have one more important concern that I believe needs to be addressed.
+In addition I have one more important concern that I believe needs to be addressed.**
 
-More important concern:
+**More important concern:
 Given that their analysis depends on them compiling data from multiple studies and reanalyzing that data, I believe it is necessary for these authors to make sure that all components of this analysis are reproducible.
@@
 
-We appreciate the reviewer's concern regarding the openness of the underlying data used in this study.
+> We appreciate the reviewer's concern regarding the openness of the underlying data used in this study.
 As far as we can tell all of the sequence data is available either through the ENA, SRA, or dbGaP except for the Turnbaugh and Schubert datasets.
@@
 
-We have specified that it was alpha-diversity.
+> We have specified that it was alpha-diversity.
 It is actually unclear what type of diversity the original authors intended.
@@
 
-We corrected this sentence.
+> We corrected this sentence.
 
@@
 
-We have edited the sentence to make it clear that following the PRISMA guidelines is necessary to prevent inclusion bias and to be considered a robust study.
+> We have edited the sentence to make it clear that following the PRISMA guidelines is necessary to prevent inclusion bias and to be considered a robust study.
 
@@
 
-The exact approach we used has been clarified in the Supplemental Materials.
+> The exact approach we used has been clarified in the Supplemental Materials.
 
@@
 
-We corrected this error in the new review.
+> We corrected this error in the new review.
 
@@
 
-We appreciate the reviewers' concern, but alternative wording (e.g.
+> We appreciate the reviewers' concern, but alternative wording (e.g.
 effect sizes had the same sign) was clunky and effectively said the same as "trend".
@@
 
-**L110 "significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant." Some justification for this statement is needed.
+**L110 "significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant." Some justification for this statement is needed.**
 
-L126 "obese, it is questionable whether that risk is biologically or clinically relevant." Again, Some justification for this statement is needed.**
+**L126 "obese, it is questionable whether that risk is biologically or clinically relevant." Again, Some justification for this statement is needed.**
 
-We expand upon these statements at L204 where we discuss the need to develop guidelines as to what is a biologically significant difference.
+> We expand upon these statements at L204 where we discuss the need to develop guidelines as to what is a biologically significant difference.
 There are likely some researchers who would contend that any statistically significant difference in diversity is also biologically significant; however, it is also possible for a study to be overpowered so that it could yield statistically significant differences that are not biologically significant.
@@
 
-We have clarified that by plausible, we mean that the range includes values that most microbiome researchers consider to be biologically meaningful - even if they do not know what it means.
+> We have clarified that by plausible, we mean that the range includes values that most microbiome researchers consider to be biologically meaningful - even if they do not know what it means.
 Furthermore, most of the studies were sufficiently powered to detect effect sizes of 15%.
@@
 
-We have highlighted the text to indicate that the analysis was based solely on microbiome  data.
+> We have highlighted the text to indicate that the analysis was based solely on microbiome  data.
 
@@
 
-We have added references for lme4, pROC, and AUCRF. The others have not been published.
+> We have added references for lme4, pROC, and AUCRF. The others have not been published.
 
@@
 
-We have added the citation number for each study to Table 1.
+> We have added the citation number for each study to Table 1.
 
@@
 
-We have added whitespace around the boxes on the right side of the figure.
+> We have added whitespace around the boxes on the right side of the figure.

### 2016-07-19 submission/Sze_Obesity_mBio_2016.tex
Re-run make write.paper

--- 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
@@
             pdfauthor={},
-            pdftitle={Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome},
+            pdftitle={},
             colorlinks=true,
@@
 
-\section{Lack of power and small effect size confounds the ability to
-differentiate non-obese and obese individuals using gut microbiome
-data}\label{lack-of-power-and-small-effect-size-confounds-the-ability-to-differentiate-non-obese-and-obese-individuals-using-gut-microbiome-data}
+\section{Looking for a Signal in the Noise: Revisiting Obesity and the
+Microbiome}\label{looking-for-a-signal-in-the-noise-revisiting-obesity-and-the-microbiome}
 
@@
 datasets, the median accuracy varied between 33.01 and 64.77\%
-(median=56.67\%). Although there was support for a relationship between
+(median=56.68\%). Although there was support for a relationship between
 the microbial communities found in human feces and obesity status, this
@@
 result to humans by the same researchers did not observe this effect,
-but did find that obese individuals had a lower diversity than lean
-individuals (6). They also showed that the relative abundance of
+but did find that obese individuals had a lower alpha-diversity than
+lean individuals (6). They also showed that the relative abundance of
 \emph{Bacteroidetes} and \emph{Firmicutes} increased and decreased,
 respectively, as obese individuals lost weight while on a fat or
-carbohydrate restricted diet (7). Two re-analysis studies interrogated
-previously published microbiome and obesity data and concluded that the
-previously reported differences in community diversity and B:F among
-non-obese and obese individuals could not be generalized (8, 9).
-Regardless of the results using human populations, mechanistic studies
-using animal models that were manipulated with antibiotics or
-colonization with varied communities appears to support the association
-since these manipulation yielded variation in animal weight (10--13).
-The purported association between the differences in the microbiome and
+carbohydrate restricted diet (7). Two re-analysis studies by Walters et
+al. (8) and Finucane et al. (9) interrogated previously published
+microbiome and obesity data and concluded that the previously reported
+differences in community diversity and B:F among non-obese and obese
+individuals could not be generalized. Regardless of the results using
+human populations, studies using animal models where the community was
+manipulated with antibiotics or established by colonizing germ-free
+animals with varied communities appear to support the association since
+these manipulations yielded differences in animal weight (10--13). The
+purported association between the differences in the microbiome and
 obesity have been widely repeated with little attention given to the
@@
 mask the appropriate signal and it is necessary to measure the
-association at the level of microbial populations. Walters et al. (8)
-demonstrated that Random Forest machine learning models were capable of
-predicting obesity status within a single cohort, but did not attempt to
-test the models on other cohorts. The purpose of this study was to
-perform a meta-analysis of the association between differences in the
-microbiome and obesity status by analyzing and applying a more
-systematic and synthetic approach than was used previously.
+association at the level of microbial populations. The Walters
+re-analysis study demonstrated that Random Forest machine learning
+models were capable of predicting obesity status within a single cohort,
+but did not attempt to test the models on other cohorts. The purpose of
+this study was to perform a meta-analysis of the association between
+differences in the microbiome and obesity status by analyzing and
+applying a more systematic and synthetic approach than was used
+previously.
 
@@
 
-\textbf{\emph{Literature Review and Study Inclusion.}} We followed the
-Preferred Reporting Items for Systematic Reviews and Meta-Analyses
-(PRISMA) guidelines to identify studies to include in our meta-analysis
-(14). A detailed description of our selection process and the exact
-search terms are provided in the Supplemental Text and in Figure 1.
-Briefly, we searched PubMed for original research studies that involved
-studying obesity and the human microbiome. The initial search yielded
-187 studies. We identified number\_string{[}n{]} additional studies that
-were not designed to explicitly test for an association between the
-microbiome and obesity. We then manually curated the 196 studies to
-select those studies that included BMI and sequence data. This yielded
-10 eligible studies. An additional study was removed from our analysis
-because no individuals in the study had a BMI over 30. Among the final
-10 studies, 3 were from identified from our PubMed search (10, 15, 16),
-5 were originally identified from the 9 studies that did not explicitly
-investigate obesity but included BMI data (17--21), and two datasets
-were used (22, 23) because these publications did not specifically look
-for any metabolic or obesity conditions but had control populations and
-enabled us to help mitigate against publication biases associated with
-the bacterial microbiome and obesity. The ten studies are summarized in
-Table 1.
+\textbf{\emph{Literature Review and Study Inclusion.}} To perform a
+robust meta-analysis and limit inclusion bias, we followed the Preferred
+Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA)
+guidelines to identify the studies that we analyzed (14). A detailed
+description of our selection process and the exact search terms are
+provided in the Supplemental Text and in Figure 1. Briefly, we searched
+PubMed for original research studies that involved studying obesity and
+the human microbiome. The initial search yielded 187 studies. We
+identified ten additional studies that were not designed to explicitly
+test for an association between the microbiome and obesity. We then
+manually curated the 196 studies to select those studies that included
+BMI and 16S rRNA gene sequence data. This yielded 10 eligible studies.
+An additional study was removed from our analysis because no individuals
+in the study had a BMI over 30. Among the final 10 studies, 3 were from
+identified from our PubMed search (10, 15, 16), 5 were originally
+identified from the 10 studies that did not explicitly investigate
+obesity but included BMI data (17--21), and two datasets were used (22,
+23) because these publications did not specifically look for any
+metabolic or obesity conditions but had control populations and enabled
+us to help mitigate against publication biases associated with the
+bacterial microbiome and obesity. The ten studies are summarized in
+Table 1. For comparison, two of these studies were included in the
+Finucane re-analysis study (10, 21) and four of these studies were
+included in the Walters re-analysis study (10, 15, 20, 21). The 16S rRNA
+gene sequence data from each study was re-analyzed using a similar
+approach based on previously described methods for reducing the number
+of chimeric sequences and sequencing errors for 454 and Illumina MiSeq
+data (24, 25). The sequences were either clustered into operational
+taxonomic units (OTUs) using the average neighbor approach (26) or
+taxonomic groupings based on their classification using a naive Bayesian
+classifier (27).
 
 \textbf{\emph{Alpha diversity analysis.}} We calculated the Shannon
-diversity index, richness, and Shannon evenness, the relative abundance
-of \emph{Bacteroidetes} and \emph{Firmicutes}, and the ratio of their
-relative abundance (B:F) for each sample. Once we transformed each of
-the six alpha diversity metrics to make them normally distributed, we
-used a t-test to identify significant associations between the alpha
-diversity metric and whether an individual was obese for each of the ten
-studies. The B:F and the relative abundance of \emph{Firmicutes} were
-not significantly associated with obesity in any study. We identified
-seven P-values that were less than 0.05: three studies indicated obese
-individuals had a lower richness, two studies indicated a significantly
-lower diversity, one study indicated a significantly lower evenness, and
-one study indicated a significantly higher relative abundance of
-\emph{Bacteroidetes} (Figures 2 and S1). These results largely match
-those of the Walters and Finucane re-analysis studies. Interestingly,
-although only two of the ten studies observed the previously reported
-association between lower diversity and obesity, the other studies
-appeared to have the same trend, albeit the differences were not
-statistically significant. We used a random effects linear model to
-combine the studies using the study as the random effect and found
-statistical support for decreased richness, evenness, and diversity
-among obese individuals (all P\textless{}0.011). Although there was a
-significant relationship between these metrics and obesity status, the
-effect size was quite small. The obese individuals averaged 7.47\% lower
-richness, 0.88\% lower evenness, and 2.07\% lower diversity. There were
-no significant associations when we pooled the phylum-level metrics
-across studies. These results indicate that obese individuals do have a
-statistically significant lower diversity than non-obese individuals;
-however, it is questionable whether the difference is biologically
-significant.
+diversity index, observed richness, and Shannon evenness, the relative
+abundance of \emph{Bacteroidetes} and \emph{Firmicutes}, and the ratio
+of their relative abundance (B:F) for each sample. Once we transformed
+each of the six alpha diversity metrics to make them normally
+distributed, we used a t-test to identify significant associations
+between the alpha diversity metric and whether an individual was obese
+for each of the ten studies. The B:F and the relative abundance of
+\emph{Firmicutes} were not significantly associated with obesity in any
+study. We identified seven P-values that were less than 0.05: three
+studies indicated obese individuals had a lower richness, two studies
+indicated a significantly lower diversity, one study indicated a
+significantly lower evenness, and one study indicated a significantly
+higher relative abundance of \emph{Bacteroidetes} (Figures 2 and S1).
+These results largely match those of the Walters and Finucane
+re-analysis studies. Interestingly, although only two of the ten studies
+observed the previously reported association between lower diversity and
+obesity, the other studies appeared to have the same trend, albeit the
+differences were not statistically significant. We used a random effects
+linear model to combine the studies using the study as the random effect
+and found statistical support for decreased richness, evenness, and
+diversity among obese individuals (all P\textless{}0.011). Although
+there was a significant relationship between these metrics and obesity
+status, the effect size was quite small. The obese individuals averaged
+7.47\% lower richness, 0.88\% lower evenness, and 2.07\% lower
+diversity. There were no significant associations when we pooled the
+phylum-level metrics across studies. These results indicate that obese
+individuals do have a statistically significant lower diversity than
+non-obese individuals; however, it is questionable whether the
+difference is biologically significant.
 
@@
 individual's alpha diversity metrics relative to the median metric for
-that study. The results using relative risk largely matched those of
-using the untransformed alpha diversity data. Across the ten studies and
-six metrics, the only significant relative risk values were the
-richness, evenness, and diversity values from the Goodrich study
-(Figures 3 and S2). Again, although the relative risk values were not
-significant for other studies, the values tended to be above one. When
-we pooled the data using a random effects model, the relative risk
-associated with having a richness, evenness, or diversity below the
-median for the population was significantly associated with obesity (all
-P\textless{}0.0044). The relative risks associated with alpha diversity
-were small. The relative risk of having a low richness was 1.30
-(1.13-1.49), low evenness was 1.20 (1.06-1.37), and low diversity was
-1.27 (1.09-1.48). There were no significant difference in the
+that study. Inspection of funnel plots for each of the metrics suggested
+that the studies included in our analysis were not biased (Figure S2).
+The results using relative risk largely matched those of using the raw
+alpha diversity data. Across the ten studies and six metrics, the only
+significant relative risk values were the richness, evenness, and
+diversity values from the Goodrich study (Figures 3 and S3). Again,
+although the relative risk values were not significant for other
+studies, the values tended to be above one. When we pooled the data
+using a random effects model, the relative risk associated with having a
+richness, evenness, or diversity below the median for the population was
+significantly associated with obesity (all P\textless{}0.0044). The
+relative risks associated with alpha diversity were small. The relative
+risk of having a low richness was 1.30 (95\% CI: 1.13-1.49), low
+evenness was 1.20 (95\% CI: 1.06-1.37), and low diversity was 1.27 (95\%
+CI: 1.09-1.48). There were no significant differences in the
 phylum-level metrics. Again, the relative risk results indicate that
@@
 ascertain the directionality of the difference in community structure
-nor perform a pooled analysis using studies that had non-overlapping 16S
-rRNA gene sequence regions it is unclear whether these differences
-reflect a broader, but perhaps small, shift in community structure
-between non-obese and obese individuals.
+because the samples are arrayed in a non-dimensional space. nor perform
+a pooled analysis using studies that had non-overlapping 16S rRNA gene
+sequence regions it is unclear whether these differences reflect a
+broader, but perhaps small, shift in community structure between
+non-obese and obese individuals.
 
@@
 of the alpha diversity metrics using the sample sizes used in each of
-the studies (Figures 6, S3-S8). Although there is no biological
-rationale for these effect sizes, they represent a range that is
-plausible. Only the Goodrich study had power greater than 0.80 to detect
-a 5\% difference in Shannon diversity and six of the studies had enough
-power to detect a 10\% difference (Figure 6). None of the studies had
-sufficient power to detect a 15\% difference between B:F values (Figure
-S5). In fact, the maximum power among any of the studies to detect a
-15\% difference in B:F values was 0.25. Among the tests for relative
-risk, none of the studies had sufficient power to detect a Cohen's d of
-0.10 and only two studies had sufficient power to detect a Cohen's d of
-0.15. We next estimated how many individuals would need to have been
-sampled to have sufficient power to detect the four effect sizes
-assuming the observed interpersonal variation from each study and
+the studies (Figures 6, S4-S9). Although there is no biological
+rationale for these effect sizes, they represent a range that includes
+effect sizes that would be generally considered to be biologically
+significant. Only the Goodrich study had power greater than 0.80 to
+detect a 5\% difference in Shannon diversity and six of the studies had
+enough power to detect a 10\% difference (Figure 6). None of the studies
+had sufficient power to detect a 15\% difference between B:F values
+(Figure S4). In fact, the maximum power among any of the studies to
+detect a 15\% difference in B:F values was 0.25. Among the tests for
+relative risk, none of the studies had sufficient power to detect a
+Cohen's d of 0.10 and only two studies had sufficient power to detect a
+Cohen's d of 0.15. We next estimated how many individuals would need to
+have been sampled to have sufficient power to detect the four effect
+sizes assuming the observed interpersonal variation from each study and
 balanced sampling between the two groups. To detect a 1, 5, 10, or 15\%
@@
 remaining datasets. This analysis demonstrated that the ability to
-reliably classify individuals as being obese based on the composition of
-their microbiome was limited. Finally, we assessed the ability of each
-study to detect defined differences in alpha diversity and observed that
-most studies were underpowered to detect modest effect sizes.
-Considering these datasets are among the largest published, it appears
-that most microbiome studies are underpowered to detect differences in
-alpha diversity.
+reliably classify individuals as being obese based solely on the
+composition of their microbiome was limited. Finally, we assessed the
+ability of each study to detect defined differences in alpha diversity
+and observed that most studies were underpowered to detect modest effect
+sizes. Considering these datasets are among the largest published, it
+appears that most human microbiome studies are underpowered to detect
+differences in alpha diversity.
 
@@
 studies. There is wide interpersonal variation in the diversity and
-structure of the human microbiome. In addition, the common experimental
-designs limit their power. As we observed, most of the studies included
-in our analysis were unbalanced for the variable that we were interested
-in. This was also true of those studies that originally sought to
-identify associations with obesity. Even with a balanced design, we
-showed that it was necessary to obtain approximately 140 and 6,300
-sequences per sample to detect a 5\% difference in Shannon diversity or
-B:F, respectively. It was interesting that these sample sizes agreed
-across studies regardless of their sequencing method, region with in the
-16S rRNA gene, or subject population (Figure 6). This suggests that
-regardless of the treatment or category, these sample sizes represent a
-good starting point for subject recruitment when using stool samples.
-Unfortunately, few studies have been published with this level of
-subject recruitment. This is troubling since the positive predictive
-rate of a significant finding in an underpowered study is small leading
-to results that cannot be reproduced (24). Future microbiome studies
-should articulate the basis for their experimental design.
+structure of the human microbiome. Some factors such as relationship
+between subjects could potentially decrease the amount variation (6) and
+other factors such as whether one lives in a rural environment could
+increase the amount of variation (28). It was interesting that although
+there were ranges in variation across the studies included in our
+analysis, the sample size calculations yielded similar results. In
+addition, the common experimental designs limit their power. As we
+observed, most of the studies included in our analysis were unbalanced
+for the variable that we were interested in. This was also true of those
+studies that originally sought to identify associations with obesity.
+Even with a balanced design, we showed that it was necessary to obtain
+approximately 140 and 6,300 samples per group to detect a 5\% difference
+in Shannon diversity or B:F, respectively. It was interesting that these
+sample sizes agreed across studies regardless of their sequencing
+method, region within the 16S rRNA gene, or subject population (Figure
+6). This suggests that regardless of the treatment or category, these
+sample sizes represent a good starting point for subject recruitment
+when using stool samples. Unfortunately, few studies have been published
+with this level of subject recruitment. This is troubling since the
+positive predictive rate of a significant finding in an underpowered
+study is small leading to results that cannot be reproduced (29). Future
+microbiome studies should articulate the basis for their experimental
+design.
 
-Two previous reviews (8, 9) have stated that there was not a consistent
-association between alpha diversity and obesity; however, neither of
-these studies made an attempt to pool the existing data together to try
-and harness the additional power that this would give and they did not
-assess whether the studies were sufficiently powered to detect a
-difference. Our analysis also used 16S rRNA gene sequence data from ten
-studies whereas the Finucane study used 16S rRNA gene sequence data from
-three studies (7, 10, 21) and a metagenomic study (25) and the Walters
-study used 16S rRNA gene sequence data from five studies (10, 15, 20,
-21, 26); two studies were included in both analyses (10, 21). Our
-analysis included four of these studies (10, 15, 20, 21) and excluded
-three of the studies because they were too small (7), only utilized
-metagenomic data (25), or used short single read Illumina HiSeq data
-that has a high error rate making it untractable for \emph{de novo} OTU
-clustering (26). The additional seven datasets were published after the
-two reviews were performed and include datasets with more samples than
-were found in the original studies. Our collection of ten studies
-allowed us to largely use the same sequence analysis pipeline for all
-datasets and relied heavily on the availability of public data and
-access to metadata that included variables beyond the needs of the
-original study. To execute this analysis, we created an automated data
-analysis pipeline, which can be easily updated to add additional studies
-as they become available. Similarly, it would be possible to adapt this
-pipeline to other body sites and treatment or variables (e.g.~subject's
-sex or age).
+Two previous re-analysis studies have stated that there was not a
+consistent association between alpha diversity and obesity (8, 9);
+however, neither of these studies made an attempt to pool the existing
+data together to try and harness the additional power that this would
+give and they did not assess whether the studies were sufficiently
+powered to detect a difference. Additionally, our analysis used 16S rRNA
+gene sequence data from ten studies whereas the Finucane study used 16S
+rRNA gene sequence data from three studies (7, 10, 21) and a metagenomic
+study (30) and the Walters study used 16S rRNA gene sequence data from
+five studies (10, 15, 20, 21, 28); two studies were included in both
+analyses (10, 21). Our analysis included four of these studies (10, 15,
+20, 21) and excluded three of the studies because they were too small
+(7), only utilized metagenomic data (30), or used short single read
+Illumina HiSeq data that has a high error rate making it untractable for
+\emph{de novo} OTU clustering (28). The additional seven datasets were
+published after the two reviews were performed and include datasets with
+more samples than were found in the original studies. Our collection of
+ten studies allowed us to largely use the same sequence analysis
+pipeline for all datasets and relied heavily on the availability of
+public data and access to metadata that included variables beyond the
+needs of the original study. To execute this analysis, we created an
+automated data analysis pipeline, which can be easily updated to add
+additional studies as they become available
+(\url{https://github.com/SchlossLab/Sze_Obesity_mBio_2016/}). Similarly,
+it would be possible to adapt this pipeline to other body sites and
+treatment or variables (e.g.~subject's sex or age).
 
-Similar to our study, the Walters et al (8) generated Random Forest
-machine learning models to differentiate between non-obese and obese
-individuals. They obtained similar AUC values to our analysis; however,
-they did not attempt to test these models on the other studies in their
-analysis. When we performed the inter-dataset cross validation the
-median accuracy across datasets was only 56.67\% indicating that the
-models did a poor job when applied to other datasets. This could be due
-to differences in subject populations and methods. Considering the
-median AUC for models trained and tested on the same data with ten-fold
-cross validation only varied between 0.51 and 0.65 and there was not a
-strong signal in the alpha diversity data, we suspect that there is
-insufficient signal to reliably classify individuals.
+Similar to our study, the Walters study generated Random Forest machine
+learning models to differentiate between non-obese and obese individuals
+(8). They obtained similar AUC values to our analysis; however, they did
+not attempt to test these models on the other studies in their analysis.
+When we performed the inter-dataset cross validation the median accuracy
+across datasets was only 56.68\% indicating that the models did a poor
+job when applied to other datasets. This could be due to differences in
+subject populations and methods. Considering the median AUC for models
+trained and tested on the same data with ten-fold cross validation only
+varied between 0.51 and 0.65 and there was not a strong signal in the
+alpha diversity data, we suspect that there is insufficient signal to
+reliably classify individuals.
 
@@
 signatures of obesity. Alternatively, it could be that the involvement
-of the microbiome in obesity is at the level of a common set of
-metabolites that can be produced from different structures of the
-microbiome.
+of the microbiome in obesity is not apparent based on the taxonomic
+information provided by 16S rRNA gene sequence data. Rather, the
+differences could become more apparent at the level of a common set of
+gene transcripts or metabolites that can be produced from different
+structures of the microbiome.
 
@@
 previously recommended approaches for 454 and Illumina sequencing data
-(27, 28). All sequences were screened for chimeras using UCHIME and
+(24, 25). All sequences were screened for chimeras using UCHIME and
 assigned to operational taxonomic units (OTUs) using the average
-neighbor algorithm using a 3\% distance threshold (29, 30). All sequence
-processing was performed using mothur (v.1.37.0) (31).
+neighbor algorithm using a 3\% distance threshold (26, 31). All sequence
+processing was performed using mothur (v.1.37.0) (32).
 
@@
 ratio of \emph{Bacteroidetes} and \emph{Firmicutes} relative abundances
-(B:F), Shannon diversity, observed richness, and Shannon evenness. After
-each variable was transformed to fit a normal distribution a two-tailed
-t-test was performed for comparison of non-obese and obese individuals
-(i.e.~BMI \textgreater{} 35.0). We performed a pooled analysis on these
-measured variables using linear random effect models to correct for
-study effect to asses differences on the combined dataset between
-non-obese and obese groups using the lme4 (v.1.1-12) R package. Next, we
-compared the community structure from non-obese and obese individuals
-using analysis of molecular variance (AMOVA) with Bray-Curtis distance
-matrices (32). This analysis was performed using the vegan (v.2.3-5) R
-package. For both analyses, the datasets were rarefied (N=1000) so that
-each study within a study had the same number of sequences. Second, for
-each study we partitioned the subjects into a low or high group
-depending on whether their alpha diversity metrics were below or above
-the median value for the study. The relative risk (RR) was then
+(B:F), Shannon diversity, observed richness, and Shannon evenness (33).
+After each variable was transformed to fit a normal distribution a
+two-tailed t-test was performed for comparison of non-obese and obese
+individuals (i.e.~BMI \textgreater{} 35.0). We performed a pooled
+analysis on these measured variables using linear random effect models
+to correct for study effect to asses differences on the combined dataset
+between non-obese and obese groups using the lme4 (v.1.1-12) R package
+(34). Next, we compared the community structure from non-obese and obese
+individuals using analysis of molecular variance (AMOVA) with
+Bray-Curtis distance matrices (35). This analysis was performed using
+the vegan (v.2.3-5) R package. For both analyses, the datasets were
+rarefied (N=1000) so that each study had the same number of sequences.
+Second, for each study we partitioned the subjects into a low or high
+group depending on whether their alpha diversity metrics were below or
+above the median value for the study. The relative risk (RR) was then
 calculated as the ratio of the number of obese individuals in the low
@@
 studies using the epiR (0.9-77) and metafor (1.9-8) packages. Third, we
-used the AUCRF (1.1) R package to generate Random Forest models. For
-each study we developed models using either OTUs or genus-level
+used the AUCRF (1.1) R package to generate Random Forest models (36).
+For each study we developed models using either OTUs or genus-level
 phylotypes. The quality of each model was assessed by measuring the area
-under the curve (AUC) of the Receive Operating Characteristic (ROC)
+under the curve (AUC) of the Receiver Operating Characteristic (ROC)
 using ten-fold cross validation. Because the genus-level phylotype
@@
 calculate the accuracy of the model applied to the test studies. To
-generate Receiver Operator Characteristic (ROC) curves and calculate the
-accuracy of the models we used the pROC (1.8) R package. Finally, we
-performed power and sample number simulations for different effect sizes
-for each study using the pwr (1.1-3) R package and base R functions. We
-also calculated the actual sample size needed based on the effect size
-of each individual study.
+generate ROC curves and calculate the accuracy of the models we used the
+pROC (1.8) R package (37). Finally, we performed power and sample number
+simulations for different effect sizes for each study using the pwr
+(1.1-3) R package and base R functions. We also calculated the actual
+sample size needed based on the effect size of each individual study.
 
@@
 
-\textbf{Figure 1: PRISMA flow diagram of total records searched (33).}
+\textbf{Figure 1: PRISMA flow diagram of total records searched (38).}
 
@@
 
-\textbf{Figure S2: Meta Analysis of the Relative Risk of Obesity Based
+\textbf{Figure S2: Funnel plots depicting the general lack of bias in
+the selection of datasets included in the analysis.}
+
+\textbf{Figure S3: Meta Analysis of the Relative Risk of Obesity Based
 on Evenness (A), Richness (B), or the Relative Abundance of
@@
 
-\textbf{Figure S3: Power (A) and sample size simulations (B) for B:F for
+\textbf{Figure S4: Power (A) and sample size simulations (B) for B:F for
 differentiating between non-obese versus obese for effect sizes of 1, 5,
@@
 
-\textbf{Figure S4: Power (A) and sample size simulations (B) for
+\textbf{Figure S5: Power (A) and sample size simulations (B) for
 richness for differentiating between non-obese versus obese for effect
@@
 
-\textbf{Figure S5: Power (A) and sample size simulations (B) for
+\textbf{Figure S6: Power (A) and sample size simulations (B) for
 evenness for differentiating between non-obese versus obese for effect
@@
 
-\textbf{Figure S6: Power (A) and sample size simulations (B) for the
+\textbf{Figure S7: Power (A) and sample size simulations (B) for the
 relative abundance of \emph{Bacteroidetes} for differentiating between
@@
 
-\textbf{Figure S7: Power (A) and sample size simulations (B) for the
+\textbf{Figure S8: Power (A) and sample size simulations (B) for the
 relative abundance of \emph{Firmicutes} for differentiating between
@@
 
-\textbf{Figure S8: Power (A) and sample size simulations (B) for
+\textbf{Figure S9: Power (A) and sample size simulations (B) for
 relative risk of obesity based on Shannon diversity.} Power calculations
@@
 
+\hyperdef{}{ref-Kozichux5f2013}{\label{ref-Kozichux5f2013}}
+24. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq Illumina sequencing platform.
+Applied and environmental microbiology \textbf{79}:5112--5120.
+
+\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
+25. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
+Reducing the effects of PCR amplification and sequencing artifacts on
+16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
+doi:\url{http://doi.org/10.1371/journal.pone.0027310}.
+
+\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
+26. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
+methods outperform reference-based methods for assigning 16S rRNA gene
+sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
+doi:\url{http://doi.org/10.7717/peerj.1487}.
+
+\hyperdef{}{ref-Wang2007}{\label{ref-Wang2007}}
+27. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
+of rRNA sequences into the new bacterial taxonomy. Applied and
+Environmental Microbiology \textbf{73}:5261--5267.
+doi:\url{http://doi.org/10.1128/aem.00062-07}.
+
+\hyperdef{}{ref-yatsunenkoux5fhumanux5f2012}{\label{ref-yatsunenkoux5fhumanux5f2012}}
+28. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
+\textbf{Trehan I}, \textbf{Dominguez-Bello MG}, \textbf{Contreras M},
+\textbf{Magris M}, \textbf{Hidalgo G}, \textbf{Baldassano RN},
+\textbf{Anokhin AP}, \textbf{Heath AC}, \textbf{Warner B},
+\textbf{Reeder J}, \textbf{Kuczynski J}, \textbf{Caporaso JG},
+\textbf{Lozupone CA}, \textbf{Lauber C}, \textbf{Clemente JC},
+\textbf{Knights D}, \textbf{Knight R}, \textbf{Gordon JI}. 2012. Human
+gut microbiome viewed across age and geography. Nature
+\textbf{486}:222--227. doi:\url{http://doi.org/10.1038/nature11053}.
+
 \hyperdef{}{ref-Ioannidis2005}{\label{ref-Ioannidis2005}}
-24. \textbf{Ioannidis JPA}. 2005. Why most published research findings
+29. \textbf{Ioannidis JPA}. 2005. Why most published research findings
 are false. PLoS Med \textbf{2}:e124.
@@
 \hyperdef{}{ref-Qin2010}{\label{ref-Qin2010}}
-25. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
+30. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
 \textbf{Burgdorf KS}, \textbf{Manichanh C}, \textbf{Nielsen T},
@@
 
-\hyperdef{}{ref-yatsunenkoux5fhumanux5f2012}{\label{ref-yatsunenkoux5fhumanux5f2012}}
-26. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
-\textbf{Trehan I}, \textbf{Dominguez-Bello MG}, \textbf{Contreras M},
-\textbf{Magris M}, \textbf{Hidalgo G}, \textbf{Baldassano RN},
-\textbf{Anokhin AP}, \textbf{Heath AC}, \textbf{Warner B},
-\textbf{Reeder J}, \textbf{Kuczynski J}, \textbf{Caporaso JG},
-\textbf{Lozupone CA}, \textbf{Lauber C}, \textbf{Clemente JC},
-\textbf{Knights D}, \textbf{Knight R}, \textbf{Gordon JI}. 2012. Human
-gut microbiome viewed across age and geography. Nature
-\textbf{486}:222--227. doi:\url{http://doi.org/10.1038/nature11053}.
-
-\hyperdef{}{ref-Kozichux5f2013}{\label{ref-Kozichux5f2013}}
-27. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
-\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
-dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq Illumina sequencing platform.
-Applied and environmental microbiology \textbf{79}:5112--5120.
-
-\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
-28. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
-Reducing the effects of PCR amplification and sequencing artifacts on
-16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
-doi:\url{http://doi.org/10.1371/journal.pone.0027310}.
-
 \hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
-29. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
+31. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
 \textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
@@
 
-\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
-30. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
-methods outperform reference-based methods for assigning 16S rRNA gene
-sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\url{http://doi.org/10.7717/peerj.1487}.
-
 \hyperdef{}{ref-mothur}{\label{ref-mothur}}
-31. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+32. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 
+\hyperdef{}{ref-Magurran2003}{\label{ref-Magurran2003}}
+33. \textbf{Magurran AE}. 2003. Measuring biological diversity 264.
+
+\hyperdef{}{ref-lme4ux5f2015}{\label{ref-lme4ux5f2015}}
+34. \textbf{Bates D}, \textbf{Mächler M}, \textbf{Bolker B},
+\textbf{Walker S}. 2015. Fitting linear mixed-effects models using lme4.
+Journal of Statistical Software \textbf{67}:1--48.
+doi:\url{http://doi.org/10.18637/jss.v067.i01}.
+
 \hyperdef{}{ref-Anderson2001}{\label{ref-Anderson2001}}
-32. \textbf{Anderson MJ}. 2001. A new method for non-parametric
+35. \textbf{Anderson MJ}. 2001. A new method for non-parametric
 multivariate analysis of variance. Austral Ecology \textbf{26}:32--46.
@@
 
+\hyperdef{}{ref-Calle2011}{\label{ref-Calle2011}}
+36. \textbf{Calle ML}, \textbf{Urrea V}, \textbf{Boulesteix A-L},
+\textbf{Malats N}. 2011. AUC-RF: A new strategy for genomic profiling
+with random forest. Human Heredity \textbf{72}:121--132.
+doi:\url{http://doi.org/10.1159/000330778}.
+
+\hyperdef{}{ref-pROCux5f2011}{\label{ref-pROCux5f2011}}
+37. \textbf{Robin X}, \textbf{Turck N}, \textbf{Hainard A},
+\textbf{Tiberti N}, \textbf{Lisacek F}, \textbf{Sanchez J-C},
+\textbf{Müller M}. 2011. PROC: An open-source package for r and s+ to
+analyze and compare rOC curves. BMC Bioinformatics \textbf{12}:77.
+
 \hyperdef{}{ref-Moher2009}{\label{ref-Moher2009}}
-33. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
+38. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
 \textbf{Altman DG}. 2009. Preferred reporting items for systematic

### 2016-07-19 submission/Sze_Obesity_mBio_2016.Rmd
Add expanded Random Forest analysis; Closes Issue #18

@@
 For each study we developed a Random Forest machine learning model to classify individuals.
-Using ten-fold cross validation, the observed AUC values varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A).
-So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset.
+Using ten-fold cross validation, the observed AUC values for the OTU-based models varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A).
+To test models on other datasets, we trained models using genus-level phylotype data for each dataset.
 The cross-validated AUC values for the models applied to the training datasets varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B).
@@
 Although there was considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`).
+We built similar models using taxonomic representation based on phylum, class, order, and family assignments and saw no variation in the results (Figure S4).
+We also attempted to predict individual BMI values as continuous variables based on the relative abundance of OTUs and genera.
+The median percent of the variance explained with the resulting models was `r median(otu$regression_auc, na.rm=T)` for the OTU-based models and `r median(genus$regression_auc, na.rm=T)` for the genus-based models.
 When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance.
-The ability to predict obesity status using the relative abundance of OTUs and genera in the communities is only marginally better than random.
+The ability to predict obesity status using relative abundance data from the communities is only marginally better than random.
 These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.
@@
 The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals.
-To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S4-S9).
+To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S5-S10).
 Although there is no biological rationale for these effect sizes, they represent a range that includes effect sizes that would be generally considered to be biologically significant.
 Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6).
-None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S4).
+None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S5).
 In fact, the maximum power among any of the studies to detect a 15% difference in B:F values was `r max(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"])`.
@@
 This could be due to differences in subject populations and methods.
-Considering the median AUC for models trained and tested on the same data with ten-fold cross validation only varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)` and there was not a strong signal in the alpha diversity data, we suspect that there is insufficient signal to reliably classify individuals.
+Furthermore, others have reported improved classification at broader taxonomic levels [@Sun2010]; we did not find this to be the case across the studies in our analysis (Figure S4).
+Considering the median AUC for models trained and tested on the same data with ten-fold cross validation only varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)` and that there was not a strong signal in the alpha diversity data, we suspect that there is insufficient signal to reliably classify individuals to a BMI category based on their microbiota.
 
@@
 
-**Figure S4: Power (A) and sample size simulations (B) for B:F for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S4: Overall accuracy of each study to predict non-obese and obese individuals based on that study's Random Forest machine learning model applied to each of the other studies when trained using relative abundance of each phylum, class, order, family, or genus.** The cross-validated AUC values for the training model are provided for each study and taxonomic level.
 
-**Figure S5: Power (A) and sample size simulations (B) for richness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S5: Power (A) and sample size simulations (B) for B:F for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure S6: Power (A) and sample size simulations (B) for evenness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S6: Power (A) and sample size simulations (B) for richness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure S7: Power (A) and sample size simulations (B) for the relative abundance of *Bacteroidetes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S7: Power (A) and sample size simulations (B) for evenness for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure S8: Power (A) and sample size simulations (B) for the relative abundance of *Firmicutes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S8: Power (A) and sample size simulations (B) for the relative abundance of *Bacteroidetes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 
-**Figure S9: Power (A) and sample size simulations (B) for relative risk of obesity based on Shannon diversity.**  Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+**Figure S9: Power (A) and sample size simulations (B) for the relative abundance of *Firmicutes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
+
+**Figure S10: Power (A) and sample size simulations (B) for relative risk of obesity based on Shannon diversity.**  Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
 

### 2016-07-19 code/plot_roc_data.R

@@ -27,8 +27,8 @@ plot_roc_data <- function(data_type, double=TRUE){
 
 	N <- ifelse(data_type == "otu", "OTUs", "Genera")
 
-	rf_summary <- rf_summary[order(rf_summary$auc_cv, decreasing=T),]
-	datasets <- sort(rf_summary$dataset)
+	rf_summary <- rf_summary[order(rf_summary$dataset, decreasing=F),]
+	datasets <- rf_summary$dataset
 
 	study_name <- capwords(datasets)
 	study_name[study_name == "Hmp"] <- "HMP"

### 2016-07-19 code/plot_train_test_supp.R

@@ -1,0 +1,79 @@
+pch <- c(baxter=21, escobar=21, goodrich=21, hmp=21, ross=21,
+				schubert=21, turnbaugh=21, wu=21, zeevi=21, zupancic=21)
+col <- c(baxter="black", escobar="red", goodrich="green", hmp="blue",
+ 				ross="orange", schubert="black", turnbaugh="red", wu="green",
+				zeevi="blue", zupancic="orange")
+bg <- c(baxter="white", escobar="white", goodrich="white", hmp="white",
+ 				ross="white", schubert="black", turnbaugh="red", wu="green",
+				zeevi="blue", zupancic="orange")
+names <- c(baxter="Baxter", escobar="Escobar", goodrich="Goodrich", hmp="HMP",
+ 				ross="Ross", schubert="Schubert", turnbaugh="Turnbaugh", wu="Wu",
+				zeevi="Zeevi", zupancic="Zupancic")
+
+
+
+plot_train_test <- function(rank){
+
+	train_test_file <- paste0("data/process/random_forest.", tolower(rank),
+													".train_test")
+	z <- read.table(file=train_test_file, header=T, stringsAsFactors=F)
+
+	o <- order(z$train, z$test)
+	z <- z[o,]
+
+	par(mar=c(0.5,1,1.5,0.5))
+	plot(NA, xlim=c(0.9,10.1), ylim=c(0,1), ylab="", xlab="", axes=F)
+	abline(h=0.5, col="gray", lwd=2, lty=2)
+	abline(v=seq(1.5,9.5,1))
+	studies <- sort(unique(z$train))
+	n_studies <- length(studies)
+
+	stagger <- seq(-0.3,0.3,length.out=10)
+	for(s in 1:n_studies){
+		study <- z[z$test==studies[s],]
+	#	segments(x0=s+stagger, y0=study$accuracy_lci, y1=study$accuracy_uci)
+		points(x=s+stagger, y=study$accuracy, col=col[studies], bg=bg[studies],
+						lwd=2, pch=pch[studies])
+	}
+	box()
+	axis(2, las=2)
+	mtext(side=3, text=rank, at=0.5, adj=0, font=2)
+
+
+	summary_file <- paste0("data/process/random_forest.", tolower(rank),
+													".summary")
+	s <- read.table(file=summary_file, header=T, stringsAsFactors=F)
[…]

### 2016-07-19 results/tables/table_1.Rmd
Update references and add to Table 1

@@
 	study <- ifelse(dataset=="hmp", "HMP", capwords(dataset))
+	paste0(study, " (", citation[dataset], ")")
 }
@@
 
+citation <-  c(
+	baxter = "23",#
+	escobar = "16",#
+	hmp = "21",#
+	ross = "18",#
+	schubert = "22",#
+	turnbaugh = "6",#
+	wu = "20",#
+	zupancic = "15",#
+	goodrich = "19",#
+	zeevi = "17"#
+)
+
 get_study_summary <- function(study, beta = beta_summary){
@@
 
-	list(study = capwords(study),
+	list(study = make_study_label(study),
 				n_subjects = as.character(n_subjects),
@@
 #colnames(study_summary) <- LETTERS[1:8]
-colnames(study_summary) <- c("Study", "\\bigcell{c}{Subjects \\\\ (N)}",
+colnames(study_summary) <- c("Study (Ref.)", "\\bigcell{c}{Subjects \\\\ (N)}",
 													"\\bigcell{c}{Obese \\\\ (\\%)}",

### 2016-07-19 submission/Sze_Obesity_mBio_2016.tex
Update references and add to Table 1

@@
 \usepackage[margin=1.0in]{geometry}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
+\usepackage{hyperref}
+\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
+\hypersetup{unicode=true,
+            pdfborder={0 0 0},
+            breaklinks=true}
 \urlstyle{same}  % don't use monospace font for urls
@@
 \begin{document}
-\maketitle
 
@@
 Random Forest machine learning model to classify individuals. Using
-ten-fold cross validation, the observed AUC values varied between 0.52
-and 0.69 indicating a relatively poor ability to classify individuals
-(Figure 4A). So that we could test models on other datasets, we trained
-models using genus-level phylotype data for each dataset. The
+ten-fold cross validation, the observed AUC values for the OTU-based
+models varied between 0.52 and 0.69 indicating a relatively poor ability
+to classify individuals (Figure 4A). To test models on other datasets,
+we trained models using genus-level phylotype data for each dataset. The
 cross-validated AUC values for the models applied to the training
@@
 median accuracy for each model varied between 0.33 (Turnbaugh) and 0.65
-(HMP) (median=0.57). When we considered the number of samples, balance
-of non-obese and obese individuals, and region within the 16S rRNA gene
-it was not possible to identify factors that predictably affected model
-performance. The ability to predict obesity status using the relative
-abundance of OTUs and genera in the communities is only marginally
-better than random. These results suggest that given the large diversity
-of microbiome compositions it is difficult to identify a taxonomic
-signal that can be associated with obesity.
+(HMP) (median=0.57). We built similar models using taxonomic
+representation based on phylum, class, order, and family assignments and
+saw no variation in the results (Figure S4). We also attempted to
+predict individual BMI values as continuous variables based on the
+relative abundance of OTUs and genera. The median percent of the
+variance explained with the resulting models was 0.13 for the OTU-based
+models and 0.082 for the genus-based models. When we considered the
+number of samples, balance of non-obese and obese individuals, and
+region within the 16S rRNA gene it was not possible to identify factors
+that predictably affected model performance. The ability to predict
+obesity status using relative abundance data from the communities is
+only marginally better than random. These results suggest that given the
+large diversity of microbiome compositions it is difficult to identify a
+taxonomic signal that can be associated with obesity.
 
@@
 of the alpha diversity metrics using the sample sizes used in each of
-the studies (Figures 6, S4-S9). Although there is no biological
+the studies (Figures 6, S5-S10). Although there is no biological
 rationale for these effect sizes, they represent a range that includes
@@
 had sufficient power to detect a 15\% difference between B:F values
-(Figure S4). In fact, the maximum power among any of the studies to
+(Figure S5). In fact, the maximum power among any of the studies to
 detect a 15\% difference in B:F values was 0.25. Among the tests for
@@
 job when applied to other datasets. This could be due to differences in
-subject populations and methods. Considering the median AUC for models
-trained and tested on the same data with ten-fold cross validation only
-varied between 0.51 and 0.65 and there was not a strong signal in the
-alpha diversity data, we suspect that there is insufficient signal to
-reliably classify individuals.
+subject populations and methods. Furthermore, others have reported
+improved classification at broader taxonomic levels (31); we did not
+find this to be the case across the studies in our analysis (Figure S4).
+Considering the median AUC for models trained and tested on the same
+data with ten-fold cross validation only varied between 0.51 and 0.65
+and that there was not a strong signal in the alpha diversity data, we
+suspect that there is insufficient signal to reliably classify
+individuals to a BMI category based on their microbiota.
 
@@
 assigned to operational taxonomic units (OTUs) using the average
-neighbor algorithm using a 3\% distance threshold (26, 31). All sequence
-processing was performed using mothur (v.1.37.0) (32).
+neighbor algorithm using a 3\% distance threshold (26, 32). All sequence
+processing was performed using mothur (v.1.37.0) (33).
 
@@
 ratio of \emph{Bacteroidetes} and \emph{Firmicutes} relative abundances
-(B:F), Shannon diversity, observed richness, and Shannon evenness (33).
+(B:F), Shannon diversity, observed richness, and Shannon evenness (34).
 After each variable was transformed to fit a normal distribution a
@@
 between non-obese and obese groups using the lme4 (v.1.1-12) R package
-(34). Next, we compared the community structure from non-obese and obese
+(35). Next, we compared the community structure from non-obese and obese
 individuals using analysis of molecular variance (AMOVA) with
-Bray-Curtis distance matrices (35). This analysis was performed using
+Bray-Curtis distance matrices (36). This analysis was performed using
 the vegan (v.2.3-5) R package. For both analyses, the datasets were
@@
 studies using the epiR (0.9-77) and metafor (1.9-8) packages. Third, we
-used the AUCRF (1.1) R package to generate Random Forest models (36).
+used the AUCRF (1.1) R package to generate Random Forest models (37).
 For each study we developed models using either OTUs or genus-level
@@
 generate ROC curves and calculate the accuracy of the models we used the
-pROC (1.8) R package (37). Finally, we performed power and sample number
+pROC (1.8) R package (38). Finally, we performed power and sample number
 simulations for different effect sizes for each study using the pwr
@@
 
-\textbf{Figure 1: PRISMA flow diagram of total records searched (38).}
+\textbf{Figure 1: PRISMA flow diagram of total records searched (39).}
 
@@
 
-\textbf{Figure S4: Power (A) and sample size simulations (B) for B:F for
+\textbf{Figure S4: Overall accuracy of each study to predict non-obese
+and obese individuals based on that study's Random Forest machine
+learning model applied to each of the other studies when trained using
+relative abundance of each phylum, class, order, family, or genus.} The
+cross-validated AUC values for the training model are provided for each
+study and taxonomic level.
+
+\textbf{Figure S5: Power (A) and sample size simulations (B) for B:F for
 differentiating between non-obese versus obese for effect sizes of 1, 5,
@@
 
-\textbf{Figure S5: Power (A) and sample size simulations (B) for
+\textbf{Figure S6: Power (A) and sample size simulations (B) for
 richness for differentiating between non-obese versus obese for effect
@@
 
-\textbf{Figure S6: Power (A) and sample size simulations (B) for
+\textbf{Figure S7: Power (A) and sample size simulations (B) for
 evenness for differentiating between non-obese versus obese for effect
@@
 
-\textbf{Figure S7: Power (A) and sample size simulations (B) for the
+\textbf{Figure S8: Power (A) and sample size simulations (B) for the
 relative abundance of \emph{Bacteroidetes} for differentiating between
@@
 
-\textbf{Figure S8: Power (A) and sample size simulations (B) for the
+\textbf{Figure S9: Power (A) and sample size simulations (B) for the
 relative abundance of \emph{Firmicutes} for differentiating between
@@
 
-\textbf{Figure S9: Power (A) and sample size simulations (B) for
+\textbf{Figure S10: Power (A) and sample size simulations (B) for
 relative risk of obesity based on Shannon diversity.} Power calculations
@@
 Genome Medicine \textbf{8}:37.
-doi:\url{http://doi.org/10.1186/s13073-016-0290-3}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
 
-\hyperdef{}{ref-Kozichux5f2013}{\label{ref-Kozichux5f2013}}
+\hypertarget{ref-Kozichux5f2013}{}
 24. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
@@
 
-\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
+\hypertarget{ref-Schloss2011}{}
 25. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
@@
 16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
-doi:\url{http://doi.org/10.1371/journal.pone.0027310}.
+doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
 
-\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
+\hypertarget{ref-Westcott2015}{}
 26. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
@@
 sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\url{http://doi.org/10.7717/peerj.1487}.
+doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
 
-\hyperdef{}{ref-Wang2007}{\label{ref-Wang2007}}
+\hypertarget{ref-Wang2007}{}
 27. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
@@
 Environmental Microbiology \textbf{73}:5261--5267.
-doi:\url{http://doi.org/10.1128/aem.00062-07}.
+doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
 
-\hyperdef{}{ref-yatsunenkoux5fhumanux5f2012}{\label{ref-yatsunenkoux5fhumanux5f2012}}
+\hypertarget{ref-yatsunenkoux5fhumanux5f2012}{}
 28. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
@@
 gut microbiome viewed across age and geography. Nature
-\textbf{486}:222--227. doi:\url{http://doi.org/10.1038/nature11053}.
+\textbf{486}:222--227.
+doi:\href{https://doi.org/10.1038/nature11053}{10.1038/nature11053}.
 
-\hyperdef{}{ref-Ioannidis2005}{\label{ref-Ioannidis2005}}
+\hypertarget{ref-Ioannidis2005}{}
 29. \textbf{Ioannidis JPA}. 2005. Why most published research findings
 are false. PLoS Med \textbf{2}:e124.
-doi:\url{http://doi.org/10.1371/journal.pmed.0020124}.
+doi:\href{https://doi.org/10.1371/journal.pmed.0020124}{10.1371/journal.pmed.0020124}.
 
-\hyperdef{}{ref-Qin2010}{\label{ref-Qin2010}}
+\hypertarget{ref-Qin2010}{}
 30. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
@@
 catalogue established by metagenomic sequencing. Nature
-\textbf{464}:59--65. doi:\url{http://doi.org/10.1038/nature08821}.
+\textbf{464}:59--65.
+doi:\href{https://doi.org/10.1038/nature08821}{10.1038/nature08821}.
 
-\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
-31. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
+\hypertarget{ref-Sun2010}{}
+31. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Mai V}, \textbf{Farmerie W},
+\textbf{Yu F}, \textbf{Li J}, \textbf{Goodison S}. 2010. Advanced
+computational algorithms for microbial community analysis using massive
+16S rRNA sequence data. Nucleic Acids Research \textbf{38}:e205--e205.
+doi:\href{https://doi.org/10.1093/nar/gkq872}{10.1093/nar/gkq872}.
+
+\hypertarget{ref-Edgar2011}{}
+32. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
 \textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
 and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
-doi:\url{http://doi.org/10.1093/bioinformatics/btr381}.
+doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
 
-\hyperdef{}{ref-mothur}{\label{ref-mothur}}
-32. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\hypertarget{ref-mothur}{}
+33. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 
-\hyperdef{}{ref-Magurran2003}{\label{ref-Magurran2003}}
-33. \textbf{Magurran AE}. 2003. Measuring biological diversity 264.
+\hypertarget{ref-Magurran2003}{}
+34. \textbf{Magurran AE}. 2003. Measuring biological diversity 264.
 
-\hyperdef{}{ref-lme4ux5f2015}{\label{ref-lme4ux5f2015}}
-34. \textbf{Bates D}, \textbf{Mächler M}, \textbf{Bolker B},
+\hypertarget{ref-lme4ux5f2015}{}
+35. \textbf{Bates D}, \textbf{Mächler M}, \textbf{Bolker B},
 \textbf{Walker S}. 2015. Fitting linear mixed-effects models using lme4.
 Journal of Statistical Software \textbf{67}:1--48.
-doi:\url{http://doi.org/10.18637/jss.v067.i01}.
+doi:\href{https://doi.org/10.18637/jss.v067.i01}{10.18637/jss.v067.i01}.
 
-\hyperdef{}{ref-Anderson2001}{\label{ref-Anderson2001}}
-35. \textbf{Anderson MJ}. 2001. A new method for non-parametric
+\hypertarget{ref-Anderson2001}{}
+36. \textbf{Anderson MJ}. 2001. A new method for non-parametric
 multivariate analysis of variance. Austral Ecology \textbf{26}:32--46.
-doi:\url{http://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}.
+doi:\href{https://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}{10.1111/j.1442-9993.2001.01070.pp.x}.
 
-\hyperdef{}{ref-Calle2011}{\label{ref-Calle2011}}
-36. \textbf{Calle ML}, \textbf{Urrea V}, \textbf{Boulesteix A-L},
+\hypertarget{ref-Calle2011}{}
+37. \textbf{Calle ML}, \textbf{Urrea V}, \textbf{Boulesteix A-L},
 \textbf{Malats N}. 2011. AUC-RF: A new strategy for genomic profiling
 with random forest. Human Heredity \textbf{72}:121--132.
-doi:\url{http://doi.org/10.1159/000330778}.
+doi:\href{https://doi.org/10.1159/000330778}{10.1159/000330778}.
 
-\hyperdef{}{ref-pROCux5f2011}{\label{ref-pROCux5f2011}}
-37. \textbf{Robin X}, \textbf{Turck N}, \textbf{Hainard A},
+\hypertarget{ref-pROCux5f2011}{}
+38. \textbf{Robin X}, \textbf{Turck N}, \textbf{Hainard A},
 \textbf{Tiberti N}, \textbf{Lisacek F}, \textbf{Sanchez J-C},
@@
 
-\hyperdef{}{ref-Moher2009}{\label{ref-Moher2009}}
-38. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
+\hypertarget{ref-Moher2009}{}
+39. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
 \textbf{Altman DG}. 2009. Preferred reporting items for systematic

### 2016-07-19 submission/Sze_Obesity_mBio_2016.tex
Re-run make write.paper

@@
 \usepackage[margin=1.0in]{geometry}
-\usepackage{hyperref}
-\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
 \urlstyle{same}  % don't use monospace font for urls
@@
 \begin{document}
+\maketitle
 
@@
 the United States, 2011-2012. JAMA \textbf{311}:806--814.
-doi:\href{https://doi.org/10.1001/jama.2014.732}{10.1001/jama.2014.732}.
-
-\hypertarget{ref-lichtashux5fbodyux5f2013}{}
+doi:\url{http://doi.org/10.1001/jama.2014.732}.
+
+\hyperdef{}{ref-lichtashux5fbodyux5f2013}{\label{ref-lichtashux5fbodyux5f2013}}
 2. \textbf{Lichtash CT}, \textbf{Cui J}, \textbf{Guo X}, \textbf{Chen
@@
 PloS One \textbf{8}:e65954.
-doi:\href{https://doi.org/10.1371/journal.pone.0065954}{10.1371/journal.pone.0065954}.
-
-\hypertarget{ref-braheux5fcanux5f2016}{}
+doi:\url{http://doi.org/10.1371/journal.pone.0065954}.
+
+\hyperdef{}{ref-braheux5fcanux5f2016}{\label{ref-braheux5fcanux5f2016}}
 3. \textbf{Brahe LK}, \textbf{Astrup A}, \textbf{Larsen LH}. 2016. Can
@@
 the Gut Microbiota? Advances in Nutrition (Bethesda, Md)
-\textbf{7}:90--101.
-doi:\href{https://doi.org/10.3945/an.115.010587}{10.3945/an.115.010587}.
-
-\hypertarget{ref-drorux5fmicrobiotaux5f2016}{}
+\textbf{7}:90--101. doi:\url{http://doi.org/10.3945/an.115.010587}.
+
+\hyperdef{}{ref-drorux5fmicrobiotaux5f2016}{\label{ref-drorux5fmicrobiotaux5f2016}}
 4. \textbf{Dror T}, \textbf{Dickstein Y}, \textbf{Dubourg G},
@@
 Microbial Pathogenesis.
-doi:\href{https://doi.org/10.1016/j.micpath.2016.01.002}{10.1016/j.micpath.2016.01.002}.
-
-\hypertarget{ref-leyux5fobesityux5f2005}{}
+doi:\url{http://doi.org/10.1016/j.micpath.2016.01.002}.
+
+\hyperdef{}{ref-leyux5fobesityux5f2005}{\label{ref-leyux5fobesityux5f2005}}
 5. \textbf{Ley RE}, \textbf{Bäckhed F}, \textbf{Turnbaugh P},
@@
 \textbf{102}:11070--11075.
-doi:\href{https://doi.org/10.1073/pnas.0504978102}{10.1073/pnas.0504978102}.
-
-\hypertarget{ref-turnbaughux5fcoreux5f2009}{}
+doi:\url{http://doi.org/10.1073/pnas.0504978102}.
+
+\hyperdef{}{ref-turnbaughux5fcoreux5f2009}{\label{ref-turnbaughux5fcoreux5f2009}}
 6. \textbf{Turnbaugh PJ}, \textbf{Hamady M}, \textbf{Yatsunenko T},
@@
 obese and lean twins. Nature \textbf{457}:480--484.
-doi:\href{https://doi.org/10.1038/nature07540}{10.1038/nature07540}.
-
-\hypertarget{ref-leyux5fmicrobialux5f2006}{}
+doi:\url{http://doi.org/10.1038/nature07540}.
+
+\hyperdef{}{ref-leyux5fmicrobialux5f2006}{\label{ref-leyux5fmicrobialux5f2006}}
 7. \textbf{Ley RE}, \textbf{Turnbaugh PJ}, \textbf{Klein S},
@@
 associated with obesity. Nature \textbf{444}:1022--1023.
-doi:\href{https://doi.org/10.1038/4441022a}{10.1038/4441022a}.
-
-\hypertarget{ref-waltersux5fmeta-analysesux5f2014}{}
+doi:\url{http://doi.org/10.1038/4441022a}.
+
+\hyperdef{}{ref-waltersux5fmeta-analysesux5f2014}{\label{ref-waltersux5fmeta-analysesux5f2014}}
 8. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
@@
 FEBS letters \textbf{588}:4223--4233.
-doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.
-
-\hypertarget{ref-finucaneux5ftaxonomicux5f2014}{}
+doi:\url{http://doi.org/10.1016/j.febslet.2014.09.039}.
+
+\hyperdef{}{ref-finucaneux5ftaxonomicux5f2014}{\label{ref-finucaneux5ftaxonomicux5f2014}}
 9. \textbf{Finucane MM}, \textbf{Sharpton TJ}, \textbf{Laurent TJ},
@@
 \textbf{9}:e84689.
-doi:\href{https://doi.org/10.1371/journal.pone.0084689}{10.1371/journal.pone.0084689}.
-
-\hypertarget{ref-turnbaughux5fobesity-associatedux5f2006}{}
+doi:\url{http://doi.org/10.1371/journal.pone.0084689}.
+
+\hyperdef{}{ref-turnbaughux5fobesity-associatedux5f2006}{\label{ref-turnbaughux5fobesity-associatedux5f2006}}
 10. \textbf{Turnbaugh PJ}, \textbf{Ley RE}, \textbf{Mahowald MA},
@@
 harvest. Nature \textbf{444}:1027--31.
-doi:\href{https://doi.org/10.1038/nature05414}{10.1038/nature05414}.
-
-\hypertarget{ref-Koren2012}{}
+doi:\url{http://doi.org/10.1038/nature05414}.
+
+\hyperdef{}{ref-Koren2012}{\label{ref-Koren2012}}
 11. \textbf{Koren O}, \textbf{Goodrich JK}, \textbf{Cullender TC},
@@
 \textbf{150}:470--480.
-doi:\href{https://doi.org/10.1016/j.cell.2012.07.008}{10.1016/j.cell.2012.07.008}.
-
-\hypertarget{ref-Cox2014}{}
+doi:\url{http://doi.org/10.1016/j.cell.2012.07.008}.
+
+\hyperdef{}{ref-Cox2014}{\label{ref-Cox2014}}
 12. \textbf{Cox LM}, \textbf{Yamanishi S}, \textbf{Sohn J},
@@
 \textbf{158}:705--721.
-doi:\href{https://doi.org/10.1016/j.cell.2014.05.052}{10.1016/j.cell.2014.05.052}.
-
-\hypertarget{ref-Mahana2016}{}
+doi:\url{http://doi.org/10.1016/j.cell.2014.05.052}.
+
+\hyperdef{}{ref-Mahana2016}{\label{ref-Mahana2016}}
 13. \textbf{Mahana D}, \textbf{Trent CM}, \textbf{Kurtz ZD},
@@
 high-fat diet. Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0297-9}{10.1186/s13073-016-0297-9}.
-
-\hypertarget{ref-moherux5fpreferredux5f2010}{}
+doi:\url{http://doi.org/10.1186/s13073-016-0297-9}.
+
+\hyperdef{}{ref-moherux5fpreferredux5f2010}{\label{ref-moherux5fpreferredux5f2010}}
 14. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},
@@
 International Journal of Surgery (London, England) \textbf{8}:336--341.
-doi:\href{https://doi.org/10.1016/j.ijsu.2010.02.007}{10.1016/j.ijsu.2010.02.007}.
-
-\hypertarget{ref-zupancicux5fanalysisux5f2012}{}
+doi:\url{http://doi.org/10.1016/j.ijsu.2010.02.007}.
+
+\hyperdef{}{ref-zupancicux5fanalysisux5f2012}{\label{ref-zupancicux5fanalysisux5f2012}}
 15. \textbf{Zupancic ML}, \textbf{Cantarel BL}, \textbf{Liu Z},
@@
 syndrome. PloS One \textbf{7}:e43052.
-doi:\href{https://doi.org/10.1371/journal.pone.0043052}{10.1371/journal.pone.0043052}.
-
-\hypertarget{ref-escobarux5fgutux5f2014}{}
+doi:\url{http://doi.org/10.1371/journal.pone.0043052}.
+
+\hyperdef{}{ref-escobarux5fgutux5f2014}{\label{ref-escobarux5fgutux5f2014}}
 16. \textbf{Escobar JS}, \textbf{Klotz B}, \textbf{Valdes BE},
@@
 that of Americans, Europeans and Asians. BMC microbiology
-\textbf{14}:311.
-doi:\href{https://doi.org/10.1186/s12866-014-0311-6}{10.1186/s12866-014-0311-6}.
-
-\hypertarget{ref-zeeviux5fpersonalizedux5f2015}{}
+\textbf{14}:311. doi:\url{http://doi.org/10.1186/s12866-014-0311-6}.
+
+\hyperdef{}{ref-zeeviux5fpersonalizedux5f2015}{\label{ref-zeeviux5fpersonalizedux5f2015}}
 17. \textbf{Zeevi D}, \textbf{Korem T}, \textbf{Zmora N},
@@
 Prediction of Glycemic Responses. Cell \textbf{163}:1079--1094.
-doi:\href{https://doi.org/10.1016/j.cell.2015.11.001}{10.1016/j.cell.2015.11.001}.
-
-\hypertarget{ref-rossux5f16sux5f2015}{}
+doi:\url{http://doi.org/10.1016/j.cell.2015.11.001}.
+
+\hyperdef{}{ref-rossux5f16sux5f2015}{\label{ref-rossux5f16sux5f2015}}
 18. \textbf{Ross MC}, \textbf{Muzny DM}, \textbf{McCormick JB},
@@
 16S gut community of the Cameron County Hispanic Cohort. Microbiome
-\textbf{3}:7.
-doi:\href{https://doi.org/10.1186/s40168-015-0072-y}{10.1186/s40168-015-0072-y}.
-
-\hypertarget{ref-goodrichux5fhumanux5f2014}{}
+\textbf{3}:7. doi:\url{http://doi.org/10.1186/s40168-015-0072-y}.
+
+\hyperdef{}{ref-goodrichux5fhumanux5f2014}{\label{ref-goodrichux5fhumanux5f2014}}
 19. \textbf{Goodrich JK}, \textbf{Waters JL}, \textbf{Poole AC},
@@
 \textbf{159}:789--799.
-doi:\href{https://doi.org/10.1016/j.cell.2014.09.053}{10.1016/j.cell.2014.09.053}.
-
-\hypertarget{ref-wuux5flinkingux5f2011}{}
+doi:\url{http://doi.org/10.1016/j.cell.2014.09.053}.
+
+\hyperdef{}{ref-wuux5flinkingux5f2011}{\label{ref-wuux5flinkingux5f2011}}
 20. \textbf{Wu GD}, \textbf{Chen J}, \textbf{Hoffmann C},
@@
 patterns with gut microbial enterotypes. Science (New York, NY)
-\textbf{334}:105--108.
-doi:\href{https://doi.org/10.1126/science.1208344}{10.1126/science.1208344}.
-
-\hypertarget{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{}
+\textbf{334}:105--108. doi:\url{http://doi.org/10.1126/science.1208344}.
+
+\hyperdef{}{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}{\label{ref-humanux5fmicrobiomeux5fprojectux5fconsortiumux5fstructureux5f2012}}
 21. \textbf{Human Microbiome Project Consortium}. 2012. Structure,
 function and diversity of the healthy human microbiome. Nature
-\textbf{486}:207--214.
-doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
-
-\hypertarget{ref-schubertux5fmicrobiomeux5f2014}{}
+\textbf{486}:207--214. doi:\url{http://doi.org/10.1038/nature11234}.
+
+\hyperdef{}{ref-schubertux5fmicrobiomeux5f2014}{\label{ref-schubertux5fmicrobiomeux5f2014}}
 22. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
@@
 \textbf{5}:e01021--01014.
-doi:\href{https://doi.org/10.1128/mBio.01021-14}{10.1128/mBio.01021-14}.
-
-\hypertarget{ref-baxterux5fmicrobiota-basedux5f2016}{}
+doi:\url{http://doi.org/10.1128/mBio.01021-14}.
+
+\hyperdef{}{ref-baxterux5fmicrobiota-basedux5f2016}{\label{ref-baxterux5fmicrobiota-basedux5f2016}}
 23. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
@@
 Genome Medicine \textbf{8}:37.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
-
-\hypertarget{ref-Kozichux5f2013}{}
+doi:\url{http://doi.org/10.1186/s13073-016-0290-3}.
+
+\hyperdef{}{ref-Kozichux5f2013}{\label{ref-Kozichux5f2013}}
 24. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
@@
 
-\hypertarget{ref-Schloss2011}{}
+\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
 25. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
@@
 16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
-doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
-
-\hypertarget{ref-Westcott2015}{}
+doi:\url{http://doi.org/10.1371/journal.pone.0027310}.
+
+\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
 26. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
@@
 sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
-
-\hypertarget{ref-Wang2007}{}
+doi:\url{http://doi.org/10.7717/peerj.1487}.
+
+\hyperdef{}{ref-Wang2007}{\label{ref-Wang2007}}
 27. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
@@
 Environmental Microbiology \textbf{73}:5261--5267.
-doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
-
-\hypertarget{ref-yatsunenkoux5fhumanux5f2012}{}
+doi:\url{http://doi.org/10.1128/aem.00062-07}.
+
+\hyperdef{}{ref-yatsunenkoux5fhumanux5f2012}{\label{ref-yatsunenkoux5fhumanux5f2012}}
 28. \textbf{Yatsunenko T}, \textbf{Rey FE}, \textbf{Manary MJ},
@@
 gut microbiome viewed across age and geography. Nature
-\textbf{486}:222--227.
-doi:\href{https://doi.org/10.1038/nature11053}{10.1038/nature11053}.
-
-\hypertarget{ref-Ioannidis2005}{}
+\textbf{486}:222--227. doi:\url{http://doi.org/10.1038/nature11053}.
+
+\hyperdef{}{ref-Ioannidis2005}{\label{ref-Ioannidis2005}}
 29. \textbf{Ioannidis JPA}. 2005. Why most published research findings
 are false. PLoS Med \textbf{2}:e124.
-doi:\href{https://doi.org/10.1371/journal.pmed.0020124}{10.1371/journal.pmed.0020124}.
-
-\hypertarget{ref-Qin2010}{}
+doi:\url{http://doi.org/10.1371/journal.pmed.0020124}.
+
+\hyperdef{}{ref-Qin2010}{\label{ref-Qin2010}}
 30. \textbf{Qin J}, \textbf{Li R}, \textbf{Raes J}, \textbf{Arumugam M},
@@
 catalogue established by metagenomic sequencing. Nature
-\textbf{464}:59--65.
-doi:\href{https://doi.org/10.1038/nature08821}{10.1038/nature08821}.
-
-\hypertarget{ref-Sun2010}{}
+\textbf{464}:59--65. doi:\url{http://doi.org/10.1038/nature08821}.
+
+\hyperdef{}{ref-Sun2010}{\label{ref-Sun2010}}
 31. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Mai V}, \textbf{Farmerie W},
@@
 16S rRNA sequence data. Nucleic Acids Research \textbf{38}:e205--e205.
-doi:\href{https://doi.org/10.1093/nar/gkq872}{10.1093/nar/gkq872}.
-
-\hypertarget{ref-Edgar2011}{}
+doi:\url{http://doi.org/10.1093/nar/gkq872}.
+
+\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
 32. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
@@
 and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
-doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
-
-\hypertarget{ref-mothur}{}
+doi:\url{http://doi.org/10.1093/bioinformatics/btr381}.
+
+\hyperdef{}{ref-mothur}{\label{ref-mothur}}
 33. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
@@
 
-\hypertarget{ref-Magurran2003}{}
+\hyperdef{}{ref-Magurran2003}{\label{ref-Magurran2003}}
 34. \textbf{Magurran AE}. 2003. Measuring biological diversity 264.
 
-\hypertarget{ref-lme4ux5f2015}{}
+\hyperdef{}{ref-lme4ux5f2015}{\label{ref-lme4ux5f2015}}
 35. \textbf{Bates D}, \textbf{Mächler M}, \textbf{Bolker B},
@@
 Journal of Statistical Software \textbf{67}:1--48.
-doi:\href{https://doi.org/10.18637/jss.v067.i01}{10.18637/jss.v067.i01}.
-
-\hypertarget{ref-Anderson2001}{}
+doi:\url{http://doi.org/10.18637/jss.v067.i01}.
+
+\hyperdef{}{ref-Anderson2001}{\label{ref-Anderson2001}}
 36. \textbf{Anderson MJ}. 2001. A new method for non-parametric
 multivariate analysis of variance. Austral Ecology \textbf{26}:32--46.
-doi:\href{https://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}{10.1111/j.1442-9993.2001.01070.pp.x}.
-
-\hypertarget{ref-Calle2011}{}
+doi:\url{http://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}.
+
+\hyperdef{}{ref-Calle2011}{\label{ref-Calle2011}}
 37. \textbf{Calle ML}, \textbf{Urrea V}, \textbf{Boulesteix A-L},
@@
 with random forest. Human Heredity \textbf{72}:121--132.
-doi:\href{https://doi.org/10.1159/000330778}{10.1159/000330778}.
-
-\hypertarget{ref-pROCux5f2011}{}
+doi:\url{http://doi.org/10.1159/000330778}.
+
+\hyperdef{}{ref-pROCux5f2011}{\label{ref-pROCux5f2011}}
 38. \textbf{Robin X}, \textbf{Turck N}, \textbf{Hainard A},
@@
 
-\hypertarget{ref-Moher2009}{}
+\hyperdef{}{ref-Moher2009}{\label{ref-Moher2009}}
 39. \textbf{Moher D}, \textbf{Liberati A}, \textbf{Tetzlaff J},

### 2016-07-20 submission/Response_to_reviewers.md
Incorporated @marcsze's comments (Issue #20) and final editing

@@
 
-> We have added text to address these comments (LXXX).
+> We have added text to address these comments (L81 and L91).
 
@@
 
-> We did this in the random forest analysis at the genus level and did not see a difference in the results relative to the OTU-based analysis.
-
- - need to add extra text for RF analysis at all of the levels.
+> We did this in the random forest analysis at the phylum, class, order, family and genus levels and did not see a difference in the results relative to the OTU-based analysis.
+We have added text at L161 to L165
 
@@
 
-> We have added an explicit reference to the first time we mention the Walters and Finucane studies at LXXX.
+> We have added an explicit reference to the first time we mention the Walters and Finucane studies at L50.
 Also, to respond to a comment from Reviewer 2, we have added the reference numbers to Table 1.
@@
 
-> We have added text at LXX to indicate which of our studies overlapped with those from the Walters and Finucane.
+> We have added text at L89 to L91 to indicate which of our studies overlapped with those from the Walters and Finucane.
 

### 2016-07-20 submission/Sze_Obesity_mBio_2016.Rmd
Incorporated @marcsze's comments (Issue #20) and final editing

@@
 When Random Forest machine learning models were trained on one dataset and then tested using the other 9 datasets, the median accuracy varied between `r 100*range(test_summary$accuracy)[1]` and `r 100*range(test_summary$accuracy)[2]`% (median=`r 100*median(test_summary$accuracy)`%).
-Although there was support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.
+Although there was support for a relationship between the microbial communities found in human feces and obesity status, this association was relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.
 
@@
 The initial search yielded 187 studies.
-We identified `r number_string[n]` additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
-We then manually curated the 196 studies to select those studies that included BMI and 16S rRNA gene sequence data.
-This yielded 10 eligible studies.
+We identified 10 additional studies that were not designed to explicitly test for an association between the microbiome and obesity.
+We then manually curated the 197 studies to select those studies that included BMI and 16S rRNA gene sequence data.
+This yielded 11 eligible studies.
 An additional study was removed from our analysis because no individuals in the study had a BMI over 30.
-Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_obesity-associated_2006; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 10 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity.
+Among the final 10 studies, 3 were identified from our PubMed search [@turnbaugh_obesity-associated_2006; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 10 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity.
 The `r number_string[n]` studies are summarized in Table 1.
@@
 The 16S rRNA gene sequence data from each study was re-analyzed using a similar approach based on previously described methods for reducing the number of chimeric sequences and sequencing errors for 454 and Illumina MiSeq data [@Kozich_2013; @Schloss2011].
-The sequences were either clustered into operational taxonomic units (OTUs) using the average neighbor approach [@Westcott2015] or taxonomic groupings based on their classification using a naive Bayesian classifier [@Wang2007].
+The sequences were clustered into operational taxonomic units (OTUs) using the average neighbor approach [@Westcott2015] and into taxonomic groupings based on their classification using a naive Bayesian classifier [@Wang2007].
 
@@
 The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05).
-Because it was not possible to ascertain the directionality of the difference in community structure because the samples are arrayed in a non-dimensional space.
-nor perform a pooled analysis using studies that had non-overlapping 16S rRNA gene sequence regions it is unclear whether these differences reflect a broader, but perhaps small, shift in community structure between non-obese and obese individuals.
+Because it was not possible to ascertain the directionality of the difference in community structure because the samples are arrayed in a non-dimensional space or perform a pooled analysis using studies that had non-overlapping 16S rRNA gene sequence regions, it is unclear whether these differences reflect a broader, but perhaps small, shift in community structure between non-obese and obese individuals.
 
@@
 For each study we developed a Random Forest machine learning model to classify individuals.
-Using ten-fold cross validation, the observed AUC values for the OTU-based models varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A).
+Using ten-fold cross validation, the cross-validated AUC values for the OTU-based models varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A).
 To test models on other datasets, we trained models using genus-level phylotype data for each dataset.
@@
 Although there was considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`).
-We built similar models using taxonomic representation based on phylum, class, order, and family assignments and saw no variation in the results (Figure S4).
+We built similar models using taxonomic representation based on phylum, class, order, and family assignments and saw no improvement in the results (Figure S4).
 We also attempted to predict individual BMI values as continuous variables based on the relative abundance of OTUs and genera.
-The median percent of the variance explained with the resulting models was `r median(otu$regression_auc, na.rm=T)` for the OTU-based models and `r median(genus$regression_auc, na.rm=T)` for the genus-based models.
-When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance.
-The ability to predict obesity status using relative abundance data from the communities is only marginally better than random.
+The median percent of the variance explained with the resulting models was `r format(round(100*median(otu$regression_auc, na.rm=T), digits=1), nsmall=1L)`% for the OTU-based models and `r format(round(100*median(genus$regression_auc, na.rm=T), digits=1), nsmall=1L)`% for the genus-based models.
+When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene for each study it was not possible to identify factors that predictably affected model performance.
+The ability to predict obesity status using relative abundance data from the communities was only marginally better than random.
 These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.
@@
 ***Power and Sample Size Estimate Simulations.***
-The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals.
-To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S5-S10).
+The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, or unbalanced sampling of non-obese and obese individuals.
+To assess these factors, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S5-S10).
 Although there is no biological rationale for these effect sizes, they represent a range that includes effect sizes that would be generally considered to be biologically significant.
-Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6).
+Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6A).
 None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S5).
@@
 Among the tests for relative risk, none of the studies had sufficient power to detect a Cohen's d of 0.10 and only `r number_string[sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.15,"power"] > 0.80)]` studies had sufficient power to detect a Cohen's d of 0.15.
-We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups.
-To detect a 1, 5, 10, or 15% difference in Shannon index, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively.
+We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups (Figure 6B).
+To detect a 1, 5, 10, or 15% difference in Shannon diversity, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively.
 To detect a 1, 5, 10, and 15% difference in B:F values, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively.
@@
 Alpha diversity metrics are attractive because they distill a complex dataset to a single value.
-For example, diversity is a measure of the entropy in a community and integrates richness and evenness information.
+For example, Shannon diversity is a measure of the entropy in a community and integrates richness and evenness information.
 Two communities with little taxonomic similarity can have the same diversity.
@@
 Some factors such as relationship between subjects could potentially decrease the amount variation [@turnbaugh_core_2009] and other factors such as whether one lives in a rural environment could increase the amount of variation [@yatsunenko_human_2012].
-It was interesting that although there were ranges in variation across the studies included in our analysis, the sample size calculations yielded similar results.
 In addition, the common experimental designs limit their power.
@@
 Additionally, our analysis used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from three studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from five studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012].
-Our analysis included four of these studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011] and excluded three of the studies because they were too small [@ley_microbial_2006], only utilized metagenomic data [@Qin2010], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering [@yatsunenko_human_2012].
+Our analysis included four of these studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011] and excluded three of the studies because they were too small [@ley_microbial_2006], only utilized metagenomic data [@Qin2010], or used short single read Illumina HiSeq data that has a high error rate making it intractable for *de novo* OTU clustering [@yatsunenko_human_2012].
 The additional `r number_string[n-3]` datasets were published after the two reviews were performed and include datasets with more samples than were found in the original studies.

### 2016-07-20 submission/Sze_Obesity_mBio_2016.tex
Incorporated @marcsze's comments (Issue #20) and final editing

--- 2016-07-19 submission/Sze_Obesity_mBio_2016.tex
@@
 the microbial communities found in human feces and obesity status, this
-association is relatively weak and its detection is confounded by large
+association was relatively weak and its detection is confounded by large
 interpersonal variation and insufficient sample sizes.
@@
 the human microbiome. The initial search yielded 187 studies. We
-identified ten additional studies that were not designed to explicitly
+identified 10 additional studies that were not designed to explicitly
 test for an association between the microbiome and obesity. We then
-manually curated the 196 studies to select those studies that included
-BMI and 16S rRNA gene sequence data. This yielded 10 eligible studies.
+manually curated the 197 studies to select those studies that included
+BMI and 16S rRNA gene sequence data. This yielded 11 eligible studies.
 An additional study was removed from our analysis because no individuals
-in the study had a BMI over 30. Among the final 10 studies, 3 were from
+in the study had a BMI over 30. Among the final 10 studies, 3 were
 identified from our PubMed search (10, 15, 16), 5 were originally
@@
 of chimeric sequences and sequencing errors for 454 and Illumina MiSeq
-data (24, 25). The sequences were either clustered into operational
-taxonomic units (OTUs) using the average neighbor approach (26) or
-taxonomic groupings based on their classification using a naive Bayesian
+data (24, 25). The sequences were clustered into operational taxonomic
+units (OTUs) using the average neighbor approach (26) and into taxonomic
+groupings based on their classification using a naive Bayesian
 classifier (27).
@@
 ascertain the directionality of the difference in community structure
-because the samples are arrayed in a non-dimensional space. nor perform
-a pooled analysis using studies that had non-overlapping 16S rRNA gene
-sequence regions it is unclear whether these differences reflect a
+because the samples are arrayed in a non-dimensional space or perform a
+pooled analysis using studies that had non-overlapping 16S rRNA gene
+sequence regions, it is unclear whether these differences reflect a
 broader, but perhaps small, shift in community structure between
@@
 Random Forest machine learning model to classify individuals. Using
-ten-fold cross validation, the observed AUC values for the OTU-based
-models varied between 0.52 and 0.69 indicating a relatively poor ability
-to classify individuals (Figure 4A). To test models on other datasets,
-we trained models using genus-level phylotype data for each dataset. The
-cross-validated AUC values for the models applied to the training
-datasets varied between 0.51 and 0.65, again indicating a relatively
-poor ability to classify individuals from the original dataset (Figure
-4B). For each model we identified the probability where the sum of the
-sensitivity and specificity was the highest. We then used this
-probability to define a threshold for calculating the accuracy of the
-models when applied to the other nine datasets (Figure 5). Although
+ten-fold cross validation, the cross-validated AUC values for the
+OTU-based models varied between 0.52 and 0.69 indicating a relatively
+poor ability to classify individuals (Figure 4A). To test models on
+other datasets, we trained models using genus-level phylotype data for
+each dataset. The cross-validated AUC values for the models applied to
+the training datasets varied between 0.51 and 0.65, again indicating a
+relatively poor ability to classify individuals from the original
+dataset (Figure 4B). For each model we identified the probability where
+the sum of the sensitivity and specificity was the highest. We then used
+this probability to define a threshold for calculating the accuracy of
+the models when applied to the other nine datasets (Figure 5). Although
 there was considerable variation in accuracy values for each model, the
@@
 representation based on phylum, class, order, and family assignments and
-saw no variation in the results (Figure S4). We also attempted to
+saw no improvement in the results (Figure S4). We also attempted to
 predict individual BMI values as continuous variables based on the
 relative abundance of OTUs and genera. The median percent of the
-variance explained with the resulting models was 0.13 for the OTU-based
-models and 0.082 for the genus-based models. When we considered the
-number of samples, balance of non-obese and obese individuals, and
-region within the 16S rRNA gene it was not possible to identify factors
-that predictably affected model performance. The ability to predict
-obesity status using relative abundance data from the communities is
-only marginally better than random. These results suggest that given the
-large diversity of microbiome compositions it is difficult to identify a
-taxonomic signal that can be associated with obesity.
+variance explained with the resulting models was 12.9\% for the
+OTU-based models and 8.2\% for the genus-based models. When we
+considered the number of samples, balance of non-obese and obese
+individuals, and region within the 16S rRNA gene for each study it was
+not possible to identify factors that predictably affected model
+performance. The ability to predict obesity status using relative
+abundance data from the communities was only marginally better than
+random. These results suggest that given the large diversity of
+microbiome compositions it is difficult to identify a taxonomic signal
+that can be associated with obesity.
 
@@
 insufficient statistical power to detect a difference because of
-insufficient sampling, large interpersonal variation, and unbalanced
-sampling of non-obese and obese individuals. To assess this, we
+insufficient sampling, large interpersonal variation, or unbalanced
+sampling of non-obese and obese individuals. To assess these factors, we
 calculated the power to detect differences of 1, 5, 10, and 15\% in each
@@
 detect a 5\% difference in Shannon diversity and six of the studies had
-enough power to detect a 10\% difference (Figure 6). None of the studies
-had sufficient power to detect a 15\% difference between B:F values
-(Figure S5). In fact, the maximum power among any of the studies to
-detect a 15\% difference in B:F values was 0.25. Among the tests for
+enough power to detect a 10\% difference (Figure 6A). None of the
+studies had sufficient power to detect a 15\% difference between B:F
+values (Figure S5). In fact, the maximum power among any of the studies
+to detect a 15\% difference in B:F values was 0.25. Among the tests for
 relative risk, none of the studies had sufficient power to detect a
@@
 sizes assuming the observed interpersonal variation from each study and
-balanced sampling between the two groups. To detect a 1, 5, 10, or 15\%
-difference in Shannon index, the median required sampling effort per
-group was approximately 3,400, 140, 35, or 16 individuals, respectively.
-To detect a 1, 5, 10, and 15\% difference in B:F values, the median
-required sampling effort per group was approximately 160,000, 6,300,
-1,600, or 700 individuals, respectively. To detect a 1, 5, 10, and 15\%
-difference in relative risk values using Shannon diversity, the median
-required sampling effort per group was approximately 39,000, 1,500, 380,
-or 170 individuals, respectively. These estimates indicate that most
-microbiome studies are underpowered to detect modest effect sizes using
-either metric. In the case of obesity, the studies were underpowered to
-detect the 0.90 to 6\% difference in diversity that was observed across
-the studies.
+balanced sampling between the two groups (Figure 6B). To detect a 1, 5,
+10, or 15\% difference in Shannon diversity, the median required
+sampling effort per group was approximately 3,400, 140, 35, or 16
+individuals, respectively. To detect a 1, 5, 10, and 15\% difference in
+B:F values, the median required sampling effort per group was
+approximately 160,000, 6,300, 1,600, or 700 individuals, respectively.
+To detect a 1, 5, 10, and 15\% difference in relative risk values using
+Shannon diversity, the median required sampling effort per group was
+approximately 39,000, 1,500, 380, or 170 individuals, respectively.
+These estimates indicate that most microbiome studies are underpowered
+to detect modest effect sizes using either metric. In the case of
+obesity, the studies were underpowered to detect the 0.90 to 6\%
+difference in diversity that was observed across the studies.
 
@@
 Alpha diversity metrics are attractive because they distill a complex
-dataset to a single value. For example, diversity is a measure of the
-entropy in a community and integrates richness and evenness information.
-Two communities with little taxonomic similarity can have the same
-diversity. Among ecologists the relevance of these metrics is questioned
-because it is difficult to ascribe a mechanistic interpretation to their
-relationship with stability or disease. Regardless, the concept of a
-biologically significant effect size needs to be developed among
-microbiome researchers. Alternative metrics could include the ability to
-detect a defined difference in the relative abundance of an OTU
-representing a defined relative abundance. What makes for a biologically
-significant difference or relative abundance is an important point that
-has yet to be discussed in the microbiome field. The use of
-operationally defined effect sizes should be adequate until it is
-possible to decide upon an accepted practice.
+dataset to a single value. For example, Shannon diversity is a measure
+of the entropy in a community and integrates richness and evenness
+information. Two communities with little taxonomic similarity can have
+the same diversity. Among ecologists the relevance of these metrics is
+questioned because it is difficult to ascribe a mechanistic
+interpretation to their relationship with stability or disease.
+Regardless, the concept of a biologically significant effect size needs
+to be developed among microbiome researchers. Alternative metrics could
+include the ability to detect a defined difference in the relative
+abundance of an OTU representing a defined relative abundance. What
+makes for a biologically significant difference or relative abundance is
+an important point that has yet to be discussed in the microbiome field.
+The use of operationally defined effect sizes should be adequate until
+it is possible to decide upon an accepted practice.
 
@@
 other factors such as whether one lives in a rural environment could
-increase the amount of variation (28). It was interesting that although
-there were ranges in variation across the studies included in our
-analysis, the sample size calculations yielded similar results. In
-addition, the common experimental designs limit their power. As we
-observed, most of the studies included in our analysis were unbalanced
-for the variable that we were interested in. This was also true of those
-studies that originally sought to identify associations with obesity.
-Even with a balanced design, we showed that it was necessary to obtain
-approximately 140 and 6,300 samples per group to detect a 5\% difference
-in Shannon diversity or B:F, respectively. It was interesting that these
-sample sizes agreed across studies regardless of their sequencing
-method, region within the 16S rRNA gene, or subject population (Figure
-6). This suggests that regardless of the treatment or category, these
-sample sizes represent a good starting point for subject recruitment
-when using stool samples. Unfortunately, few studies have been published
-with this level of subject recruitment. This is troubling since the
-positive predictive rate of a significant finding in an underpowered
-study is small leading to results that cannot be reproduced (29). Future
+increase the amount of variation (28). In addition, the common
+experimental designs limit their power. As we observed, most of the
+studies included in our analysis were unbalanced for the variable that
+we were interested in. This was also true of those studies that
+originally sought to identify associations with obesity. Even with a
+balanced design, we showed that it was necessary to obtain approximately
+140 and 6,300 samples per group to detect a 5\% difference in Shannon
+diversity or B:F, respectively. It was interesting that these sample
+sizes agreed across studies regardless of their sequencing method,
+region within the 16S rRNA gene, or subject population (Figure 6). This
+suggests that regardless of the treatment or category, these sample
+sizes represent a good starting point for subject recruitment when using
+stool samples. Unfortunately, few studies have been published with this
+level of subject recruitment. This is troubling since the positive
+predictive rate of a significant finding in an underpowered study is
+small leading to results that cannot be reproduced (29). Future
 microbiome studies should articulate the basis for their experimental
@@
 (7), only utilized metagenomic data (30), or used short single read
-Illumina HiSeq data that has a high error rate making it untractable for
+Illumina HiSeq data that has a high error rate making it intractable for
 \emph{de novo} OTU clustering (28). The additional seven datasets were

### 2016-07-20 submission/Sze_Obesity_mBio_2016.Rmd
Move Supplemental Text to Methods

@@
 To perform a robust meta-analysis and limit inclusion bias, we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify the studies that we analyzed [@moher_preferred_2010].
-A detailed description of our selection process and the exact search terms are provided in the Supplemental Text and in Figure 1.
+A detailed description of our selection process and the exact search terms are provided in the Methods section and in Figure 1.
 Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome.
@@
 
+***In-Depth Overview of Search Strategy.***
+The initial search strategy included looking for all papers that initially fit under the below NCBI PubMed advanced search criteria: "(((((((((Bacterial Microbiome) AND (Obesity or bmi or body mass index or BMI or obesity) AND "last 10 years"[PDat] AND Humans[Mesh])) NOT review[ptyp]) AND "last 10 years"[PDat] AND Humans[Mesh])) AND "last 10 years"[PDat] AND Humans[Mesh])) AND "last 10 years"[PDat] AND Humans[Mesh])".
+The terms included in this criteria were that the manuscript had to have "Bacterial Microbiome" and "Obesity, BMI, bmi, obesity" in their manuscript criteria, it was not published more than 10 years ago, they were not review articles, and it contained research on humans only.
+This search yielded a total of 187 manuscripts.
+From the Finucane and Walters re-analysis studies along with knowledge of other published papers that included BMI information in their sequence metadata we obtained 10 additional studies.
+This brought our total number of records to 197.
+We browsed abstracts of these studies and included studies that mentioned stool or feces examination, did not involve children, was not a clinical trial for probiotics or other diet related treatments, did not have participants with inflammatory bowel disease, were in English, and were based on the analysis of sequence data from the 454 or MiSeq platforms.
+This ultimately excluded all but a total of 11 studies.
+Because we decided *a priori* to use the standard definition for BMI group classification one study was excluded because it did not have any individuals who were obese.
+
 ***Sequence Analysis Pipeline.***

### 2016-07-20 submission/Sze_Obesity_mBio_2016.tex
Run rules for submission on axiom

@@
 description of our selection process and the exact search terms are
-provided in the Supplemental Text and in Figure 1. Briefly, we searched
+provided in the Methods section and in Figure 1. Briefly, we searched
 PubMed for original research studies that involved studying obesity and
@@
 \subsection{Methods}\label{methods}
+
+\textbf{\emph{In-Depth Overview of Search Strategy.}} The initial search
+strategy included looking for all papers that initially fit under the
+below NCBI PubMed advanced search criteria: ``(((((((((Bacterial
+Microbiome) AND (Obesity or bmi or body mass index or BMI or obesity)
+AND''last 10 years``{[}PDat{]} AND Humans{[}Mesh{]})) NOT
+review{[}ptyp{]}) AND''last 10 years``{[}PDat{]} AND Humans{[}Mesh{]}))
+AND''last 10 years``{[}PDat{]} AND Humans{[}Mesh{]})) AND''last 10
+years``{[}PDat{]} AND Humans{[}Mesh{]})''. The terms included in this
+criteria were that the manuscript had to have ``Bacterial Microbiome''
+and ``Obesity, BMI, bmi, obesity'' in their manuscript criteria, it was
+not published more than 10 years ago, they were not review articles, and
+it contained research on humans only. This search yielded a total of 187
+manuscripts. From the Finucane and Walters re-analysis studies along
+with knowledge of other published papers that included BMI information
+in their sequence metadata we obtained 10 additional studies. This
+brought our total number of records to 197. We browsed abstracts of
+these studies and included studies that mentioned stool or feces
+examination, did not involve children, was not a clinical trial for
+probiotics or other diet related treatments, did not have participants
+with inflammatory bowel disease, were in English, and were based on the
+analysis of sequence data from the 454 or MiSeq platforms. This
+ultimately excluded all but a total of 11 studies. Because we decided
+\emph{a priori} to use the standard definition for BMI group
+classification one study was excluded because it did not have any
+individuals who were obese.
 
